### 2017-10-02 papers/Physica2016/Romansky_Physica2016.tex
initial

\documentclass[a4paper]{jpconf}
\bibliographystyle{iopart-num}
\usepackage{amsmath}
\usepackage{citesort}
\usepackage{subfigure}
\usepackage{graphicx}
\graphicspath{{fig/}}
\usepackage{ifpdf}
\ifpdf\usepackage{epstopdf}\fi
\usepackage[export]{adjustbox}

%----------------------------------------------------- 
%\usepackage{soul,ulem,color,xspace,bm}
% Andrei's commands
% Suggest to remove
%\newcommand{\asrm}[1]{{\color{magenta}\sout{#1}}}
% Suggest to insert
%\newcommand{\as}[1]{\color{cyan}#1\xspace\color{black}}
% Suggest to replace
%\newcommand{\asrp}[2]{\asrm{#1} \as{#2}}
% Comment
%\newcommand{\ascm}[1]{{\color{green}\;AS: #1}}
%------------------------------------------------------

\begin{document}
\title{Particle-in-cell simulation of shock waves with different kinds of ions.}

\author{V I Romanky$^{1,2}$, A M Bykov$^1$, S M Osipov$^1$ and P E Gladilin$^1$}

\address{$^1$ Ioffe Institute, 26 Politekhnicheskaya st., St. Petersburg 194021, Russia}
\address{$^2$ 
Sternberg Astronomical Institute, Moscow State University
 Universitetsky pr., 13, Moscow 119234, Russia}

\ead{romanskyvadim@gmail.com}

\begin{abstract}
The problem of Cosmic Ray (CR) origin and acceleration is a widely discussed problem of
modern astrophysics. It is considered that most likely CR sources are supernova remnant
shock waves. One of the methods to study particle injection and acceleration by shock
waves is a particle-in-cell computer simulation. In this paper we present results of shock
wave simulations based on our self-developed implicit particle-in-cell code. First, we test the code
with the well-known two stream and Weibel instabilities and second, simulate shock waves in
plasma with different particle composition. Using the obtained results we study the
process of charged particle acceleration at the shock front.
\end{abstract}
\input{intro.tex}
\input{model.tex}
\input{test.tex}
\input{shock.tex}
\input{conclusions.tex}
\input{biblio.tex}
\end{document}

### 2017-10-02 papers/Physica2016/biblio.tex
initial

\section*{References}
\begin{thebibliography}{20}
\bibitem{Bell1978} Bell A R 1978 \textit{MNRAS} \textbf{182} 147
\bibitem{Blandford1978} Blandford R D and Ostriker J P 1978 \textit{Astrophys. J.} \textbf{221} L29 
\bibitem{Berezhko2003} Berezhko E G, Ksenofontov L T and V{\"o}lk H J  2003 \textit{A}{\&}\textit{A} \textbf{412} L11
\bibitem{Uchiyama2007} Uchiyama Y, Aharonian F A, Tanaka T, Takahashi T and Maeda Y 2007 \textit{Nature} \textbf{449} 576
\bibitem{Lapenta2006} Lapenta G, Brackbill J U and Ricci P 2006 \textit{Phys. Plasmas} \textbf{13} 055904
\bibitem{Noguchi2007} Noguchi K, Tronci C, Zuccaro G and Lapenta G 2007 \textit{Phys. Plasmas} \textbf{14} 042308
\bibitem{Weibel1959} Weibel E S 1959 \textit{Phys. Rev. Lett.} \textbf{2} 83
\bibitem{Yoon1987} Yoon P H and Davidson R C 1987 \textit{Phys. Rev. A} \textbf{35} 2718
\bibitem{Park2015} Park J, Caprioli D and Spitkovsky A 2015 \textit{Phys. Rev. Lett.} \textbf{114} 085003
\end{thebibliography}

### 2017-10-02 papers/Physica2016/intro.tex
initial

\section{Introduction}
One of the possible sources of high-energy cosmic rays are shock waves in astrophysical collisionless plasma due to the diffusive shock acceleration (DSA) mechanism, see \cite{Bell1978}, \cite{Blandford1978}. Strong magnetic fields are necessary for effective acceleration of charged particles and the observations of the synchrotron emission from supernova remnants (SNRs) show, that the magnetic field there is about 100 times stronger then the interstellar field (see for example \cite{Berezhko2003},\cite{Uchiyama2007}). So strong fields can be produced due to instabilities in anisotropic plasma. One of the most powerful methods to explore processes in collisionless plasma is the Particle-in-Cell simulation. In this work we explore instabilities in the plasma using this method.

### 2017-10-02 papers/Physica2016/model.tex
initial

\section{Model of implicit particle-in-cell}
We developed the implicit particle-in-cell (PIC) code, based on the scheme suggested by Lapenta et al.~\cite{Lapenta2006} and improved for the relativistic case by Noguchi et al.\cite{Noguchi2007}. This scheme is described in details in these papers and below we present only a short description. The common assumption for all particle-in-cell codes is the consideration of the plasma as a superposition of a large number of finite elements, called super-particles. Distribution function can be represented as the sum of members for all super-particles: 
\begin{equation}
F \left( \textbf{x},\textbf{p},t \right) =\sum _{p=1}^{N_{s}}S \left( x-x_{{p}} \right) 
S \left( y-y_{{p}} \right) S \left( z-z_{{p}} \right) \delta \left( \textbf{p}-\textbf{p}_{{p}} \right), 
\end{equation}
where $\textbf{x}$ and $\textbf{p}$ are coordinates and the momentum of a particle , respectively, $S$ is a shape-function, and $\delta$ is the Dirac's delta-function. The shape-function is chosen in form of b spline. Also we introduce interpolation function $W$ which represents the part of a particle which is contained in the grid cell:
\begin{equation}
W \left(x_{{c}} - x_{{p}} \right)= \int\limits_{-\infty}^{+\infty}S(x-x_{{p}})\phi(x)dx,
\end{equation} 
where $c$ and $p$ are indexes of the cell and the particle, respectively, and $\phi$ equals 1 inside the cell and 0 outside. Using this function we can determine electric and magnetic field for each particle as
\begin{equation}
\begin{aligned}
&\textbf{E}_{p} = \sum_{c}\textbf{E}_c W \left(x_{{c}} - x_{{p}} \right),
\\
&\textbf{B}_{p} = \sum_{c}\textbf{B}_c W \left(x_{{c}} - x_{{p}} \right).
\end{aligned}
\end{equation}

Then we use particle fields to solve equations of motion for particles:
\begin{equation}
\begin{aligned}
&{\frac {d\textbf{x}_p}{dt}}=\textbf{v}_p,
\\
&{\frac {d\textbf{p}_p}{dt}}=q_p \left(\textbf{E}_{p}+\frac{\textbf{v}_p\times\textbf{B}_{p}}{c}\right),
\end{aligned}
\end{equation}
where $q_p$ is the particle charge and $\textbf{p}_p$ is the momentum.

At the same time, using interpolation functions we can determine macroscopic plasma parameters, such as the electric flux and the electric density :
\begin{equation}
\begin{aligned}\label{rhoj}
&\rho_c=\sum_p q_p W \left(x_{{c}} - x_{{p}} \right),
\\
&\textbf{J}_c=\sum_p q_p \textbf{v}_p W \left(x_{{c}} - x_{{p}} \right).
\end{aligned}
\end{equation}

The difficulty comes from the fact that particles coordinates and fields are coupled and we can not separate the equations of motion and Maxwell equations. The first possible solution is the explicit method: equations of fields use particle velocities on the previous time step and vice versa. Explicit approach is rather simple, but it has strong restrictions on stability. The other possible method is the implicit approach which is more complicated, but at the same time more stable. The main idea of it is to predict particle velocity at the next time step, using the field at the next time step and then use velocity in the implicit scheme for the field.

In the implicit approach we should introduce intermediate values $F^{n+\theta}=\theta F^{n+1} + \left(1-\theta\right) F^n$, where $F^n$ is the value at the n-th time step and $F^{n+\theta}$ the value at the intermediate time step. $\theta$ is the parameter of the scheme and should be greater or equal to $\frac{1}{2}$ for stability. Using this denotations we write time discretisation of the Maxwell equations as:
\begin{equation}
\begin{aligned}\label{maxwell}
&\nabla\times\textbf{E}^{n+\theta}+\frac{\textbf{B}^{n+1}-\textbf{B}^n}{c\Delta t}=0,
\\
&\nabla\times\textbf{B}^{n+\theta}-\frac{\textbf{E}^{n+1}-\textbf{E}^n}{c\Delta t}=\frac{4\pi}{c}\textbf{J}^{n+\theta},
\\
&\Delta\cdot\textbf{E}^{n+\theta}=4\pi\rho^{n+\theta},
\\
&\Delta\cdot\textbf{B}^n=0.
\end{aligned}
\end{equation}

Time discretization of equations of particle motion reads as
\begin{equation}\label{particlemoving}
\begin{aligned}
&\textbf{x}_p^{n+1}=\overline{\textbf{v}}_p\Delta t,
\\
&\textbf{p}_p^{n+1}=\textbf{p}_p^{n}+q_p\Delta t
\left(\textbf{E}_p^{n+\theta}\left(\textbf{x}_p^{n+\theta}\right)+\frac{\overline{\textbf{v}}_p\times\textbf{B}_p^n\left(\textbf{x}_p^{n+\theta}\right)}{c}\right),
\end{aligned}
\end{equation}
where $\overline{\textbf{v}}_p$ is the average velocity. It should be noted that the electric field is evaluated at the moment $n+\theta$, while the magnetic field at the moment $n$. If $\theta = \frac{1}{2}$ the scheme has the second order accuracy in $\Delta t$.  Finally we can express average velocity explicitly using $\textbf{v}_p^{n}$ and $\textbf{E}_p^{n+\theta}$: 
\begin{equation}\label{averagev}
\overline{\textbf{v}}_p = \widehat{\textbf{v}}_p+\frac{q_p\Delta t}{2m_p\Gamma_p}\alpha_p^{n}\textbf{E}_p^{n+\theta}\left(\textbf{x}_p^{n+\theta}\right),
\end{equation}
where we use following notations:
\begin{equation}
\widehat{\textbf{v}}_p=\alpha_p^{n}\gamma_p^n\textbf{v}_p^n,
\end{equation} 
\begin{equation}
\alpha_p^n=\frac{1}{\Gamma_p\left(1+\left(\frac{q_p\Delta t\textbf{B}_p^n}{2m_p\Gamma_p}\right)^2\right)}\left(I-\frac{q_p\Delta t}{2m_p\Gamma_p}I\times\textbf{B}_p^n+\left(\frac{q_p\Delta t}{2m_p\Gamma_p}\right)^2\textbf{B}_p^n\textbf{B}_p^n\right),
\end{equation}
\begin{equation}
\Gamma_p=\frac{q_p\Delta t\textbf{B}_p^n}{2m_p}\textbf{E}_p^n\cdot\textbf{v}_p^n+\gamma_p^n,
\end{equation}
where $I$ is the identity tensor and $\gamma_p^n$ is the gamma-factor of the particle at the moment $n$. Using expression(\ref{averagev}) for the average velocity we can expand in series expressions for the electric flux and density (\ref{rhoj}) by term $\textbf{x}_p^n-\textbf{x}_p^{n+\theta}$ and substitute them into the Maxwell equations (\ref{maxwell}). After solving algebraic equations, combining all terms and excluding magnetic field we finally obtain the implicit second-order elliptic equation for $\textbf{E}^{n+\theta}$:
\begin{eqnarray}\label{electricfield}
\left(c\theta\Delta t\right)^2 \left(-\nabla^2\textbf{E}^{n+\theta}-\nabla\nabla\cdot\left(\mu^n\cdot\textbf{E}^{n+\theta}\right)\right)+\epsilon^n\textbf{E}^{n+\theta}=
\nonumber\\
\textbf{E}^{n}-c\theta\Delta t\left(\frac{4\pi\widehat{\textbf{J}}}{c}-\frac{\Delta t}{2}\nabla\cdot\widehat{\Pi}-\nabla\times\textbf{B}^n\right)
-\left(c\theta\Delta t\right)^2 4\pi \nabla\widehat{\rho},
\end{eqnarray}
where we used following notations:
\begin{equation}
 \epsilon^n=I+\mu^n ,
\end{equation}
\begin{equation}
\mu^n=-\sum_p\frac{2\pi q_p^2\theta\Delta t^2}{m_p}\alpha_p^n W\left(x-x_p^n\right),
\end{equation}
\begin{equation}
\widehat{\Pi}=\sum_p q_p\widehat{\textbf{v}}_p\widehat{\textbf{v}}_p W\left(x-x_p^n\right),
\end{equation}
\begin{equation}
\widehat{\textbf{J}}=\sum_p q_p \widehat{\textbf{v}}_p W\left(x-x_p^n\right),
\end{equation}
\begin{equation}
\widehat{\rho}=\rho^n - \theta \Delta t \nabla\cdot\left(\widehat{\textbf{J}} -\frac{\Delta t}{2}\nabla\cdot\widehat{\Pi} \right).
\end{equation}

The spacial discretization of the equation of the electric field (\ref{electricfield}) is performed with a finite-volume scheme and  then the system of linear equations is solved using general minimal residual algorithm. After that we can explicitly evaluate magnetic field using the first Maxwell equation (\ref{maxwell}) and finally, when fields computed, we can solve equations of motion (\ref{particlemoving}) using iterative non-linear solver.

Our code is fully three dimensional and parallelized with MPI technology, which is adapted for distributed computing and can be executed on a wide class of computers.

### 2017-10-02 papers/Physica2016/test.tex
initial

\section{Verification of numerical model}
We chose two well-known instabilities for testing our code: the two stream instability and the Weibel instability. Electromagnetic two stream instability is caused by two fluxes of electrons moving in opposite directions. Modes with the wave vector, normal to electron velocities, grow with the increment $\frac {v}{c}\frac{\omega_{p}}{\sqrt{\gamma}}$, where $\omega_{p}$ is the electron plasma frequency and $\gamma$ is the gamma-factor of both electron fluxes. We study the case with electron velocity $v=0.99c$. As shown in Figure \ref{twostream} magnetic energy growth is very similar to linear growth with analytically computed increment. When the fraction of the magnetic energy becomes close to 10 percent of the full initial energy, the instability reaches its saturation limit. The total energy deviation at the end of the simulation does not exceed 5 percent of the instability energy.
\begin{figure}[h!]
	\centering
	\begin{minipage}{0.49\textwidth}
		\center{\includegraphics[width=0.99\linewidth]{twostream.eps}}
		\caption{Time dependence of the ratio of the magnetic energy to the full initial energy  in case of the two stream instability.}
		\label{twostream}
	\end{minipage}\hfill
	\begin{minipage}{0.49\textwidth}
		\center{\includegraphics[width=0.99\linewidth]{weibel.eps}}
		\caption{Time dependence of the ratio of the magnetic energy to the full initial energy  in case of the Weibel instability.}
		\label{weibel}
	\end{minipage}
\end{figure}

The second test of the code is the Weibel instability \cite{Weibel1959}, which occurs when the temperature in one spatial direction is smaller then in the other two. This instability has many different cases, we chose the one which was studied by Yoon et al. \cite{Yoon1987}. In this case electrons have delta distribution in transverse momentum and uniform one in parallel. Electron distribution function is:
\begin{equation}
F\left(p_\perp^2, p_\parallel\right)=\frac{1}{2\pi p_{0\perp}}\delta\left(p_\perp - p_{0\perp}\right)\frac{1}{2p_{0\parallel}}H(p_{0\parallel}^2-p_\parallel^2),
\end{equation}
where $\delta$ is the delta-function, $H$ is the Heaviside function, $p_\perp, p_\parallel$ are momentum of particles in perpendicular and parallel directions , respectively, and $p_{0\perp}, p_{0\parallel}$ are constants which determine dispersion of distribution. Dispersion relation for this case can be directly solved, while the expression for the increment is rather cumbersome and will not be presented here for brevity. As shown in Figure \ref{weibel} the numerical simulation with parameters $p_{0\perp} = 20m_e c$ and $p_{0\parallel} = 2 m_e c$ is in good correspondence with the theory.

These results show that our PIC code gives consistent results in the simulation of plasma instabilities and can be applied to more complicated cases.

### 2017-10-02 FieldSolver1d/util.cpp

@@ -2,6 +2,7 @@
 #include <stdlib.h>
 #include <math.h>
 #include "util.h"
+#include "../../../../../../../Program Files/Microsoft SDKs/MPI/Include/mpi.h"
 
 double power(double v, double p){
 	return exp(p*log(v));
@@ -51,6 +52,7 @@ void alertNaNOrInfinity(double value, const char* s){
 	if(value != value || 0*value != 0*value){
 		printf("%s", s);
 		printf("\n");
+		MPI_Finalize();
 		exit(0);
 		//Sleep(1000);
 	}
@@ -60,6 +62,7 @@ void alertNotPositive(double value, const char* s){
 	if(value <= 0){
 		printf("%s", s);
 		printf("\n");
+		MPI_Finalize();
 		exit(0);
 	}
 }
@@ -68,6 +71,7 @@ void alertNegative(double value, const char* s){
 	if(value < 0){
 		printf("%s", s);
 		printf("\n");
+		MPI_Finalize();
 		exit(0);
 	}
 }

### 2017-10-02 Fourier/constants.h

@@ -1,0 +1,44 @@
+#ifndef CONSTANTS_H
+#define CONSTANTS_H
+
+//for plytech hpc
+//const char* const outputDirectory  = "/home/ipntsr/romansky/PIC3d1/output/";
+
+//for MPI on mvs
+//const char* const outputDirectory = "/home2/ioffe2/romansky/PIC3d1/output/";
+
+//for MPI on h1 or h2
+//const char* const outputDirectory = "/home/vadim/PIC3d1/output/";
+
+//for MPI on alpha
+//const char* const outputDirectory = "/home/uikam/romansky/PIC3d3/output/";
+
+//for MPI on Ubuntu
+//const char* const outputDirectory = "/home/vadim/PIC++/trunk/PIC++/output/";
+ 
+//for windows 32
+//char* const outputDirectory = "C:/users/Vadik/Documents/Visual Studio 2010/Projects/v4/PIC++/output/";
+
+//for windows 64 home
+//const char* const outputDirectory = "C:/users/Vadik/Documents/Visual Studio 2010/Projects/v4/Fourier/output/";
+
+//for windows 64
+const char* const outputDirectory = "C:/users/Vadim/Documents/Visual Studio 2010/Projects/v4/Fourier/output/";
+
+const int randomSeed = 1024;
+const double pi = 3.1415926535897932384626433832795028841971693993751;
+
+const int xnumberGeneral = 200;
+const int ynumberGeneral = 200;
+const int znumberGeneral = 200;
+
+const double xsizeGeneral = 200;
+const double ysizeGeneral = 200;
+const double zsizeGeneral = 200;
+
+const int MPI_nx = 1;
+const int MPI_ny = 1;
+const int MPI_nz = 1;
+const int MPI_dim = 3;
+
+#endif

### 2017-10-02 Fourier/main.cpp

@@ -1,0 +1,39 @@
+#include "stdio.h"
+#include <stdlib.h>
+#include <time.h>
+//#include <crtdbg.h>
+#include "mpi.h"
+
+#include "constants.h"
+#include "complex.h"
+#include "simulation.h"
+
+int main(int argc, char **argv){
+	int size;
+	int rank;
+	MPI_Init(&argc, &argv);
+
+	int dims[MPI_dim];
+	int period[MPI_dim];
+	period[0] = 0;
+	period[1] = 1;
+	period[2] = 1;
+	dims[0] = MPI_nx;
+	dims[1] = MPI_ny;
+	dims[2] = MPI_nz;
+
+	MPI_Comm_size(MPI_COMM_WORLD, &size);
+	MPI_Comm_rank(MPI_COMM_WORLD, &rank);
+	MPI_Comm comm;
+	MPI_Cart_create(MPI_COMM_WORLD, MPI_dim, dims, period, 0, &comm);
+
+	MPI_Comm_size(comm, &size);
+	MPI_Comm_rank(comm, &rank);
+
+	Simulation* simulation = new Simulation(comm);
+	//simulation->randomSimulation();
+	simulation->poissonSolving();
+
+	MPI_Finalize();
+	return 0;
+}

### 2017-10-02 Fourier/output.h

@@ -1,0 +1,8 @@
+#ifndef OUTPUT_H
+#define OUTPUT_H
+
+class Complex;
+
+void outputArray(const char* outputFileName, Complex*** array, int xnumber, int ynumber, int znumber, int xnumberGeneral, int ynumberGeneral, int znumberGeneral, MPI_Comm& cartComm, int* cartCoord, int* cartDim);
+
+#endif

### 2017-10-02 Fourier/simulation.cpp

@@ -1,0 +1,362 @@
+#include <stdio.h>
+#include <stdlib.h>
+#include <string>
+#include <time.h>
+#include <mpi.h>
+
+#include "simulation.h"
+#include "complex.h"
+#include "output.h"
+#include "fourier.h"
+
+Simulation::Simulation(MPI_Comm& comm) {
+
+	cartComm = comm;
+	int periods[MPI_dim];
+	MPI_Cart_get(cartComm, MPI_dim, cartDim, periods, cartCoord);
+	MPI_Comm_rank(cartComm, &rank);
+	MPI_Comm_size(cartComm, &nprocs);
+
+	int tempCoord[3];
+	for (int i = 0; i < 3; ++i) {
+		tempCoord[i] = cartCoord[i];
+	}
+	tempCoord[0] -= 1;
+	if (tempCoord[0] < 0) {
+		tempCoord[0] = cartDim[0] - 1;
+	}
+	MPI_Cart_rank(cartComm, tempCoord, &leftRank);
+	tempCoord[0] = cartCoord[0] + 1;
+	if (tempCoord[0] >= cartDim[0]) {
+		tempCoord[0] = 0;
+	}
+	MPI_Cart_rank(cartComm, tempCoord, &rightRank);
+	tempCoord[0] = cartCoord[0];
+	tempCoord[1] -= 1;
+	if (tempCoord[1] < 0) {
+		tempCoord[1] = cartDim[1] - 1;
+	}
+	MPI_Cart_rank(cartComm, tempCoord, &frontRank);
+	tempCoord[1] = cartCoord[1] + 1;
+	if (tempCoord[1] >= cartDim[1]) {
+		tempCoord[1] = 0;
+	}
+	MPI_Cart_rank(cartComm, tempCoord, &backRank);
+	tempCoord[1] = cartCoord[1];
+	tempCoord[2] -= 1;
+	if (tempCoord[2] < 0) {
+		tempCoord[2] = cartDim[2] - 1;
+	}
+	MPI_Cart_rank(cartComm, tempCoord, &bottomRank);
+	tempCoord[2] = cartCoord[2] + 1;
+	if (tempCoord[2] >= cartDim[2]) {
+		tempCoord[2] = 0;
+	}
+	MPI_Cart_rank(cartComm, tempCoord, &topRank);
+	setSpaceForProc();
+	createArrays();
+	initializeArrays();
+	
+	if(rank == 0){
+		createFiles();
[…]

### 2017-10-02 OutputParser/main.cpp

@@ -2,6 +2,14 @@
 #include <math.h>
 #include <string>
 
+int min2(int a, int b){
+    if(a <= b){
+        return a;
+    } else {
+        return b;
+    }
+}
+
 int countNumbers(FILE* file){
     int result = 0;
     double a;
@@ -33,64 +41,55 @@ int countNumbersLine(FILE* file, int numbersPerLine) {
 
 int main() {
     const int timePointsCount = 3; //must be >= 2
-    const int Np = 200;
-    std::string inputDir = "../../PIC++/output/";
-    //std::string inputDir = "../../../../iPic3D/data/";
-    FILE* XFile = fopen((inputDir + "Xfile.dat").c_str(),"r");
-    FILE* YFile = fopen((inputDir + "Yfile.dat").c_str(),"r");
-    FILE* ZFile = fopen((inputDir + "Zfile.dat").c_str(),"r");
-    FILE* BFile = fopen((inputDir + "Bfield.dat").c_str(),"r");
-    FILE* EFile = fopen((inputDir + "Efield.dat").c_str(), "r");
-    FILE* concentrationFile = fopen((inputDir + "concentrations.dat").c_str(), "r");
-    FILE* velocityElectronFile = fopen((inputDir + "velocity_electron.dat").c_str(), "r");
-    FILE* velocityProtonFile = fopen((inputDir + "velocity.dat").c_str(), "r");
-    FILE* fluxFile = fopen((inputDir + "flux.dat").c_str(), "r");
-    FILE* trajectoryProtonFile = fopen((inputDir + "trajectory_proton.dat").c_str(), "r");
-    FILE* trajectoryElectronFile = fopen((inputDir + "trajectory_electron.dat").c_str(), "r");
-    FILE* protonsFile = fopen((inputDir + "protons.dat").c_str(), "r");
-    FILE* electronsFile = fopen((inputDir + "electrons.dat").c_str(), "r");
-    FILE* positronsFile = fopen((inputDir + "positrons.dat").c_str(), "r");
-    FILE* alphasFile = fopen((inputDir + "alphas.dat").c_str(), "r");
-    FILE* divergenceFile = fopen((inputDir + "divergence_error.dat").c_str(), "r");
-    FILE* distributionProtonsFile = fopen((inputDir + "distribution_protons.dat").c_str(), "r");
[…]

### 2017-10-02 PIC++/boundaryFieldEvaluator.cpp

@@ -1,0 +1,56 @@
+#include "stdlib.h"
+//#include <crtdbg.h>
+
+//#include "memory_debug.h"
+#include "boundaryFieldEvaluator.h"
+
+BaseBoundaryFieldEvaluator::BaseBoundaryFieldEvaluator() {
+	
+}
+
+BaseBoundaryFieldEvaluator::~BaseBoundaryFieldEvaluator() {
+	
+}
+
+ConstantBoundaryFieldEvaluator::ConstantBoundaryFieldEvaluator(Vector3d& E0, Vector3d& B0) {
+	E=  E0;
+	B=  B0;
+}
+
+Vector3d ConstantBoundaryFieldEvaluator::evaluateEfield(double& t) {
+	return E;
+}
+
+Vector3d ConstantBoundaryFieldEvaluator::evaluateBfield(double& t) {
+	return B;
+}
+
+int StripeBoundaryFieldEvaluator::halfPeriodNumber(double& t) {
+	return ((t*ux*2 + shiftX)/lambda);
+}
+
+StripeBoundaryFieldEvaluator::StripeBoundaryFieldEvaluator(Vector3d& E0, Vector3d& B0, double l, double u, double shift) {
+	E = E0;
+	B = B0;
+	lambda = 0;
+	ux = ux;
+	shiftX = shift;
+}
+
+Vector3d StripeBoundaryFieldEvaluator::evaluateEfield(double& t) {
+	int n = halfPeriodNumber(t);
+	return E*n;
+}
+
+Vector3d StripeBoundaryFieldEvaluator::evaluateBfield(double& t) {
+	int n = halfPeriodNumber(t);
+	return B*n;
+}
+
+
+
+
+
+
+
+

### 2017-10-02 PIC++/dichotomousSolver.h

@@ -7,7 +7,10 @@
 
 class BaseDichotomousSolver{
 public:
-    double maxErrorX;
+	virtual ~BaseDichotomousSolver() {
+	}
+
+	double maxErrorX;
     double maxErrorY;
     BaseDichotomousSolver();
     virtual double function(double x);
@@ -20,7 +23,7 @@ public:
     double rightPart;
 
     TemperatureRelativisticMaxwellSolver(double alphaNormalValue, double rightPartValue);
-    double function(double x);
+    virtual double function(double x);
 };
 
 #endif //PIC_DICHOTOMOUSSOLVER_H

### 2017-10-02 PIC++/divergencecleanup.cpp

@@ -1,3 +1,4 @@
+#include <mpi.h>
 #include <time.h>
 //#include <crtdbg.h>
 
@@ -11,96 +12,38 @@
 #include "particle.h"
 #include "complex.h"
 #include "simulation.h"
+#include "mpi_util.h"
+#include "fourier.h"
 
 
-void Simulation::cleanupDivergence() {
+void Simulation::cleanupDivergence(Vector3d*** field, double*** density) {
 	double procTime = 0;
-	if(timing && (currentIteration % writeParameter == 0)) {
+	if (timing && (rank == 0) && (currentIteration % writeParameter == 0)) {
 		procTime = clock();
 	}
-	if((verbosity > 0)) printf("cleaning up divergence\n");
+	if ((rank == 0) && (verbosity > 0)) printf("cleaning up divergence\n");
 	fflush(stdout);
 
-	//substractMeanDensity();
-
-	if(ynumber == 1 && znumber == 1){
+	/*if (ynumberGeneral == 1 && znumberGeneral == 1) {
 		cleanupDivergence1d();
-		if(timing  && (currentIteration % writeParameter == 0)) {
+		MPI_Barrier(cartComm);
+		if (timing && (rank == 0) && (currentIteration % writeParameter == 0)) {
 			procTime = clock() - procTime;
-			printf("cleaning divergence time = %g sec\n", procTime/CLOCKS_PER_SEC);
+			printf("cleaning divergence time = %g sec\n", procTime / CLOCKS_PER_SEC);
 		}
 		return;
-	}
-
-
-	int matrixDimension = xnumber;
-
-	double fullDensity = 0;
-	for (int i = 0; i < xnumber; ++i) {
-		for (int j = 0; j < ynumber; ++j) {
-			for (int k = 0; k < znumber; ++k) {
-				fullDensity += chargeDensity[i][j][k] * volumeB(i, j, k);
-			}
-		}
-	}
-	fullDensity /= (xsize*ysize*zsize);
-	if((verbosity > 1)) printf("full density = %22.15g\n", fullDensity);
-	if((verbosity > 1)) printf("density[0][0][0] = %22.15g\n", chargeDensity[0][0][0]);
-	fflush(stdout);
-
-
-	/*for (int i = 0; i < xnumber; ++i) {
-		for (int j = 0; j < ynumber; ++j) {
-			for (int k = 0; k < znumber; ++k) {
[…]

### 2017-10-02 PIC++/divergencecleanup_buneman.cpp

@@ -1,0 +1,785 @@
+#include <mpi.h>
+#include <time.h>
+//#include <crtdbg.h>
+
+//#include "memory_debug.h"
+#include "constants.h"
+#include "specialmath.h"
+#include "util.h"
+#include "matrixElement.h"
+#include "vector3d.h"
+#include "matrix3d.h"
+#include "particle.h"
+#include "complex.h"
+#include "simulation.h"
+#include "mpi_util.h"
+
+void Simulation::cleanupDivergenceBuneman() {
+	double procTime = 0;
+	if (timing && (rank == 0) && (currentIteration % writeParameter == 0)) {
+		procTime = clock();
+	}
+	if ((rank == 0) && (verbosity > 0)) printf("cleaning up divergence\n");
+	fflush(stdout);
+
+	for (int i = 0; i < xnumberAdded; ++i) {
+		for (int j = 0; j < ynumberAdded; ++j) {
+			for (int k = 0; k < znumberAdded; ++k) {
+				divergenceCleanUpMatrix[i][j][k][0].clear();
+				divergenceCleanUpMatrix[i][j][k][1].clear();
+				divergenceCleanUpMatrix[i][j][k][2].clear();
+				divergenceCleanUpRightPart[i][j][k][0] = 0;
+				divergenceCleanUpRightPart[i][j][k][1] = 0;
+				divergenceCleanUpRightPart[i][j][k][2] = 0;
+				divergenceCleaningPotential[i][j][k][0] = 0;
+				divergenceCleaningField[i][j][k][0] = 0;
+				divergenceCleaningField[i][j][k][1] = 0;
+				divergenceCleaningField[i][j][k][2] = 0;
+			}
+		}
+	}
+
+	for(int i = 0; i < xnumberAdded + 1; ++i){
+		for(int j = 0; j < ynumberAdded + 1; ++j){
+			for(int k = 0; k < znumberAdded + 1; ++k){
+				bunemanDivergenceCleaningPotential[i][j][k][0] = 0;
+			}
+		}
+	}
+
+	for (int i = 0; i < xnumberAdded; ++i) {
+		for (int j = 0; j < ynumberAdded; ++j) {
+			for (int k = 0; k < znumberAdded; ++k) {
+				for (int l = 0; l < maxwellEquationMatrixSize; ++l) {
+					gmresOutput[i][j][k][l] = 0;
+				}
+			}
+		}
+	}
+
+	for(int  i = 0; i < xnumberAdded; ++i){
[…]

### 2017-10-02 PIC++/fields_exchange.cpp

@@ -1,0 +1,2902 @@
+#include "stdlib.h"
+#include "stdio.h"
+#include <cmath>
+#include <mpi.h>
+#include <time.h>
+//#include <crtdbg.h>
+
+//#include "memory_debug.h"
+#include "specialmath.h"
+#include "util.h"
+#include "matrixElement.h"
+#include "constants.h"
+#include "matrix3d.h"
+#include "vector3d.h"
+#include "particle.h"
+#include "simulation.h"
+#include "mpi_util.h"
+#include "output.h"
+void Simulation::exchangeEfield() {
+	exchangeGeneralEfield(Efield);
+	exchangeGeneralEfield(tempEfield);
+	exchangeGeneralEfield(newEfield);
+}
+
+void Simulation::exchangeGeneralEfield(Vector3d*** field) {
+	exchangeGeneralEfieldX(field);
+	exchangeGeneralEfieldY(field);
+	exchangeGeneralEfieldZ(field);
+}
+
+void Simulation::exchangeGeneralEfieldX(Vector3d*** field) {
+	if (cartDim[0] > 1) {
+		if (verbosity > 2) printf("start sending general field rank = %d\n", rank);
+		int bcount = 0;
+
+		for (int i = 0; i < 2 + additionalBinNumber; ++i) {
+			for (int j = 0; j < ynumberAdded + 1; ++j) {
+				for (int k = 0; k < znumberAdded + 1; ++k) {
+					for (int l = 0; l < 3; ++l) {
+						rightOutVectorNodeBuffer[bcount] = field[xnumberAdded - additionalBinNumber - 1 - i][j][k][l];
+						bcount++;
+					}
+				}
+			}
+		}
+		int number = (2 + additionalBinNumber) * 3 * (ynumberAdded + 1) * (znumberAdded + 1);
+		//int numberLeft = (1 + additionalBinNumber) * 3 * (ynumberAdded + 1) * (znumberAdded + 1);
+		if ((cartCoord[0] < cartDim[0] - 1 && cartCoord[0] > 0) || (boundaryConditionType == PERIODIC)) {
+			if (verbosity > 2) printf("before send general field right from %d to %d additionalBinNumber = %d, xnumber = %d, ynumber = %d, znumber = %d\n", rank, rightRank, additionalBinNumber, xnumberAdded, ynumberAdded, znumberAdded);
+			MPI_Status status;
[…]

### 2017-10-02 PIC++/fourier.h

@@ -1,0 +1,18 @@
+#ifndef FOURIER_H
+#define FOURIER_H
+
+#include "mpi.h"
+
+class Complex;
+
+void fourierTranslationXonePoint(Complex*** input, Complex*** output, bool direct, int xnumber, int j, int k, int xnumberGeneral, MPI_Comm& cartComm, int* xabsoluteIndex);
+Complex fourierTranslationXoneHarmonic(Complex*** input, bool direct, int xnumber, int j, int k, int xnumberGeneral, MPI_Comm& reducedCartComm, int knumber, int* xabsoluteIndex);
+void fourierTranslationYonePoint(Complex*** input, Complex*** output, bool direct, int ynumber, int i, int k, int ynumberGeneral, MPI_Comm& cartComm, int* yabsoluteIndex);
+Complex fourierTranslationYoneHarmonic(Complex*** input, bool direct, int ynumber, int i, int k, int ynumberGeneral, MPI_Comm& reducedCartComm, int knumber, int* yabsoluteIndex);
+void fourierTranslationZonePoint(Complex*** input, Complex*** output, bool direct, int znumber, int i, int j, int znumberGeneral, MPI_Comm& cartComm, int* zabsoluteIndex);
+Complex fourierTranslationZoneHarmonic(Complex*** input, bool direct, int znumber, int i, int j, int znumberGeneral, MPI_Comm& reducedCartComm, int knumber, int* zabsoluteIndex);
+
+
+void fourierTranslation(Complex*** input, Complex*** output, bool direct, int xnumber, int ynumber, int znumber, int xnumberGeneral, int ynumberGeneral, int znumberGeneral, MPI_Comm& cartComm, int* xabsoluteIndex, int* yabsoluteIndex, int* zabsoluteIndex, int* cartCoord, int* cartDim);
+
+#endif

### 2017-10-02 PIC++/largeVectorBasis.cpp

@@ -1,0 +1,119 @@
+#include "stdlib.h"
+#include "stdio.h"
+#include "mpi.h"
+//#include <crtdbg.h>
+
+//#include "memory_debug.h"
+#include "largeVectorBasis.h"
+
+//remember that we need xnumber + 1 size!!!
+LargeVectorBasis::LargeVectorBasis(int capacityv, int xnumberv, int ynumberv, int znumberv, int lnumberv) {
+	size = 0;
+	xnumber = xnumberv;
+	ynumber = ynumberv;
+	znumber = znumberv;
+	lnumber = lnumberv;
+	capacity = capacityv;
+
+	array = new double**** [capacity];
+	for(int m = 0; m < capacity; ++m) {
+		array[m] = new double*** [xnumber];
+		for(int i = 0; i < xnumber; ++i) {
+			array[m][i] = new double** [ynumber];
+			for(int j = 0; j < ynumber; ++j) {
+				array[m][i][j] = new double* [znumber];
+				for(int k = 0; k < znumber; ++k) {
+					array[m][i][j][k] = new double[lnumber];
+					for(int l = 0; l < lnumber; ++l) {
+						array[m][i][j][k][l] = 0;
+					}
+				}
+			}
+		}
+	}
+}
+
+LargeVectorBasis::~LargeVectorBasis() {
+	for(int m = 0; m < capacity; ++m) {
+		for(int i = 0; i < xnumber; ++i) {
+			for(int j = 0; j < ynumber; ++j) {
+				for(int k = 0; k < znumber; ++k) {
+					delete[] array[m][i][j][k];
+				}
+				delete[] array[m][i][j];
+			}
+			delete[] array[m][i];
+		}
+		delete[] array[m];
+	}	
+	delete[] array;
+}
+
+void LargeVectorBasis::resize(int capacityv) {
+	if(capacityv > capacity) {
+		double***** newArray = new double**** [capacityv];
+		for(int m = 0; m < capacity; ++m) {
+			newArray[m] = array[m];
+		}
+		for(int m = capacity; m < capacityv; ++m) {
+			newArray[m] = new double*** [xnumber];
+			for(int i = 0; i < xnumber; ++i) {
+				newArray[m][i] = new double** [ynumber];
+				for(int j = 0; j < ynumber; ++j) {
+					newArray[m][i][j] = new double* [znumber];
+					for(int k = 0; k < znumber; ++k) {
[…]

### 2017-10-02 PIC++/massMatrix.h

@@ -1,0 +1,14 @@
+#ifndef MASS_MATRIX_H
+#define MASS_MATRIX_H
+
+#include "constants.h"
+#include "matrix3d.h"
+
+struct MassMatrix{
+	int xindex[2*splineOrder+3];
+	int yindex[2*splineOrder+3];
+	int zindex[2*splineOrder+3];
+	Matrix3d matrix[2*splineOrder+3][2*splineOrder+3][2*splineOrder+3];
+};
+
+#endif

### 2017-10-02 PIC++/matrix3d.cpp

@@ -10,35 +10,30 @@
 #include "constants.h"
 
 Matrix3d::Matrix3d() {
-	//matrix = new double* [3];
-	for (int i = 0; i < 3; ++i) {
-		//matrix[i] = new double [3];
-		for (int j = 0; j < 3; ++j) {
-			if (i == j) {
-				matrix[i][j] = 1;
-			} else {
-				matrix[i][j] = 0;
-			}
-		}
-	}
+	matrix[0][0] = 0;
+	matrix[0][1] = 0;
+	matrix[0][2] = 0;
+	matrix[1][0] = 0;
+	matrix[1][1] = 0;
+	matrix[1][2] = 0;
+	matrix[2][0] = 0;
+	matrix[2][1] = 0;
+	matrix[2][2] = 0;
 }
 
 Matrix3d::Matrix3d(const Matrix3d& m) {
-	int i, j;
-	//matrix = new double* [3];
-	for (i = 0; i < 3; ++i) {
-		//matrix[i] = new double [3];
-		for (j = 0; j < 3; ++j) {
-			matrix[i][j] = m.matrix[i][j];
-		}
-	}
+	matrix[0][0] = m.matrix[0][0];
+	matrix[0][1] = m.matrix[0][1];
+	matrix[0][2] = m.matrix[0][2];
+	matrix[1][0] = m.matrix[1][0];
+	matrix[1][1] = m.matrix[1][1];
+	matrix[1][2] = m.matrix[1][2];
+	matrix[2][0] = m.matrix[2][0];
+	matrix[2][1] = m.matrix[2][1];
+	matrix[2][2] = m.matrix[2][2];
 }
 
 Matrix3d::Matrix3d(double m11, double m12, double m13, double m21, double m22, double m23, double m31, double m32, double m33) {
-	/*matrix = new double* [3];
-	for (int i = 0; i < 3; ++i) {
-		matrix[i] = new double [3];
-	}*/
 	matrix[0][0] = m11;
 	matrix[0][1] = m12;
 	matrix[0][2] = m13;
@@ -51,11 +46,6 @@ Matrix3d::Matrix3d(double m11, double m12, double m13, double m21, double m22, d
 }
 
 Matrix3d::~Matrix3d() {
-	/*int i;
-	for (i = 0; i < 3; ++i) {
-		delete[] matrix[i];
-	}
-	delete[] matrix;*/
 }
 
[…]

### 2017-10-02 PIC++/memory_util.h

@@ -1,0 +1,4 @@
+#ifndef MEMORY_UTIL_H
+#define MEMORY_UTIL_H
+void process_mem_usage(double& vm_usage, double& resident_set);
+#endif

### 2017-10-02 PIC++/momentum_collecting.cpp

@@ -1,0 +1,1087 @@
+#include "stdlib.h"
+#include "stdio.h"
+#include <cmath>
+#include <omp.h>
+#include <mpi.h>
+#include <time.h>
+//#include <crtdbg.h>
+
+//#include "memory_debug.h"
+#include "mpi_util.h"
+#include "util.h"
+#include "output.h"
+#include "matrix3d.h"
+#include "vector3d.h"
+#include "particle.h"
+#include "random.h"
+#include "simulation.h"
+
+void Simulation::updateElectroMagneticParameters() {
+	double procTime = 0;
+	if (timing && (rank == 0) && (currentIteration % writeParameter == 0)) {
+		procTime = clock();
+	}
+	if(solverType == BUNEMAN){
+		updateBunemanFlux();
+	} else {
+	//MPI_Barrier(cartComm);
+	if ((rank == 0) && (verbosity > 0)) printf("updating flux, density snd dielectric tensor\n");
+	if ((rank == 0) && (verbosity > 0)) printLog("updating flux, density and dielectric tensor\n");
+
+	int particlePartsCount = 0;
+	for (int i = 0; i < xnumberAdded + 1; ++i) {
+		for (int j = 0; j < ynumberAdded + 1; ++j) {
+			for (int k = 0; k < znumberAdded + 1; ++k) {
+				electricFlux[i][j][k] = Vector3d(0, 0, 0);
+				electricFluxMinus[i][j][k] = Vector3d(0, 0, 0);
+				dielectricTensor[i][j][k] = Matrix3d(0, 0, 0, 0, 0, 0, 0, 0, 0);
+				divPressureTensor[i][j][k] = Vector3d(0, 0, 0);
+				divPressureTensorMinus[i][j][k] = Vector3d(0, 0, 0);
+
+				for(int tempI = 0; tempI < 2*splineOrder + 3; ++tempI){
+					for(int tempJ = 0; tempJ < 2*splineOrder + 3; ++tempJ){
+						for(int tempK = 0; tempK < 2*splineOrder + 3; ++tempK){
+							for(int curI = 0; curI < 3; ++curI){
+								for(int curJ = 0; curJ < 3; ++curJ){
+									massMatrix[i][j][k].matrix[tempI][tempJ][tempK].matrix[curI][curJ] = 0;
+									tempMassMatrix[i][j][k].matrix[tempI][tempJ][tempK].matrix[curI][curJ] = 0;
+								}
+							}
[…]

### 2017-10-02 PIC++/mover.cpp

@@ -2,6 +2,7 @@
 #include "stdio.h"
 #include "cmath"
 #include <omp.h>
+#include <mpi.h>
 #include <time.h>
 //#include <crtdbg.h>
 
@@ -14,21 +15,22 @@
 #include "vector3d.h"
 #include "random.h"
 #include "simulation.h"
+#include "mpi_util.h"
 
 void Simulation::moveParticles() {
 	double procTime = 0;
-	if (timing && (currentIteration % writeParameter == 0)) {
+	if (timing && (rank == 0) && (currentIteration % writeParameter == 0)) {
 		procTime = clock();
 	}
-	//MPI_Barrier(MPI_COMM_WORLD);
-	if ((verbosity > 0)) printf("moving particles\n");
-	if ((verbosity > 0)) printLog("moving particles\n");
+	//MPI_Barrier(cartComm);
+	if ((rank == 0) && (verbosity > 0)) printf("moving particles\n");
+	if ((rank == 0) && (verbosity > 0)) printLog("moving particles\n");
 	int i = 0;
 
 	for (i = 0; i < particles.size(); ++i) {
 		if (i % 1000 == 0) {
 			if (verbosity > 2) {
-				printf("move particle number %d\n", i);
+				printf("move particle number %d rank %d\n", i, rank);
 			}
 		}
 		moveParticle(particles[i]);
@@ -38,24 +40,110 @@ void Simulation::moveParticles() {
 	//if (boundaryConditionType == SUPER_CONDUCTOR_LEFT) {
 	//removeEscapedParticles();
 	//}
-	if ((verbosity > 0)) printf("end moving particles\n");
-	if ((verbosity > 0)) printLog("end moving particles\n");
-	if (timing && (currentIteration % writeParameter == 0)) {
+	if ((rank == 0) && (verbosity > 0)) printf("end moving particles\n");
+	if ((rank == 0) && (verbosity > 0)) printLog("end moving particles\n");
+	MPI_Barrier(cartComm);
+	if (timing && (rank == 0) && (currentIteration % writeParameter == 0)) {
 		procTime = clock() - procTime;
 		printf("moving particles time = %g sec\n", procTime / CLOCKS_PER_SEC);
 	}
 }
 
-void Simulation::removeEscapedParticles() {
-	double procTime = 0;
-	if (timing && (currentIteration % writeParameter == 0)) {
-		procTime = clock();
[…]

### 2017-10-02 PIC++/mpi_util.h

@@ -1,0 +1,237 @@
+//
+// Created by vadim on 19.06.16.
+//
+
+#ifndef PIC_MPI_UTIL_H
+#define PIC_MPI_UTIL_H
+
+#include <vector>
+#include "particle.h"
+
+class Simulation;
+class Vector3d;
+class Matrix3d;
+class Particle;
+class MassMatrix;
+
+enum MPI_TAGS{MPI_INPUT_INTEGER_TAG = 1, MPI_INPUT_DOUBLE_TAG = 2, MPI_BFIELD_LEFT = 3, MPI_BFIELD_RIGHT = 4,
+    MPI_EFILDZ_RIGHT = 5, MPI_EFIELDX_LEFT = 6, MPI_EFIELDY_LEFT = 7, MPI_EFILDZ_LEFT = 8,
+    MPI_TEMPVECTOR_RIGHT = 9, MPI_TEMPVECTOR_LEFT = 10, MPI_SEND_DOUBLE_NUMBER_ALL = 11, MPI_SEND_GENERAL_PARAMETERS = 12,
+    MPI_CELL_PARAMETERS_LEFT = 14, MPI_CELL_PARAMETERS_RIGHT = 15, MPI_NODE_PARAMETERS_LEFT = 16, MPI_NODE_PARAMETERS_RIGHT = 17,
+    MPI_SEND_INTEGER_NUMBER_LEFT = 18, MPI_SEND_INTEGER_NUMBER_RIGHT = 19, MPI_SEND_DOUBLE_NUMBER_LEFT = 20, MPI_SEND_DOUBLE_NUMBER_RIGHT = 21,
+    MPI_SEND_VECTOR_LEFT = 22, MPI_SEND_VECTOR_FIRST_TO_ALL = 23, MPI_SEND_VECTOR_ALL_TO_FIRST = 24, MPI_EFIELD_LEFT = 25, MPI_EFIELD_RIGHT = 26, MPI_SEND_DOUBLE_FIRST_TO_ALL = 27, MPI_SEND_DOUBLE_ALL_TO_FIRST = 30, MPI_SEND_INTEGER_ALL_TO_FIRST = 31, MPI_SEND_INTEGER_FIRST_TO_ALL = 125};
+
+void sendInput(Simulation &simulation, int nprocs);
+Simulation recieveInput(MPI_Comm comm);
+
[…]

### 2017-10-02 PIC++/output/plot_B_derivative.m

@@ -1,0 +1,64 @@
+clear;
+load rotEFile.dat;
+load Xfile.dat;
+load Yfile.dat;
+load Zfile.dat;
+
+Nx = size(Xfile, 1)-1;
+Ny = size(Yfile, 1)-1;
+Nz = size(Zfile, 1)-1;
+
+NE = Nx*Ny*Nz;
+Nt = (size(rotEFile, 1)/NE)-1;
+
+ynumber = 1;
+znumber = 1;
+
+a = 0;
+b = fix(Nt/2);
+c = fix(Nt) - 1;
+
+
+rotEx(1:Nx, 1:3) = 0;
+rotEy(1:Nx, 1:3) = 0;
+rotEz(1:Nx, 1:3) = 0;
+
+
+
+
+
+for i=1:Nx,
+   
+   rotEx(i,1) = rotEFile((Nz)*(Ny)*(i-1) + (Nz)*(ynumber-1) + znumber + a*NE, 1);
+   rotEx(i,2) = rotEFile((Nz)*(Ny)*(i-1) + (Nz)*(ynumber-1) + znumber + b*NE, 1);
+   rotEx(i,3) = rotEFile((Nz)*(Ny)*(i-1) + (Nz)*(ynumber-1) + znumber + c*NE, 1);
+   rotEy(i,1) = rotEFile((Nz)*(Ny)*(i-1) + (Nz)*(ynumber-1) + znumber + a*NE, 2);
+   rotEy(i,2) = rotEFile((Nz)*(Ny)*(i-1) + (Nz)*(ynumber-1) + znumber + b*NE, 2);
+   rotEy(i,3) = rotEFile((Nz)*(Ny)*(i-1) + (Nz)*(ynumber-1) + znumber + c*NE, 2);
+   rotEz(i,1) = rotEFile((Nz)*(Ny)*(i-1) + (Nz)*(ynumber-1) + znumber + a*NE, 3);
+   rotEz(i,2) = rotEFile((Nz)*(Ny)*(i-1) + (Nz)*(ynumber-1) + znumber + b*NE, 3);
+   rotEz(i,3) = rotEFile((Nz)*(Ny)*(i-1) + (Nz)*(ynumber-1) + znumber + c*NE, 3);
+   
+end;
+
+figure(1);
[…]

### 2017-10-02 PIC++/output/plot_E_derivative.m

@@ -13,12 +13,12 @@ Nz = size(Zfile, 1);
 NE = Nx*Ny*Nz;
 Nt = (size(flux, 1)/NE)-1;
 
-ynumber = 1;
-znumber = 1;
+ynumber = 2;
+znumber = 2;
 
 a = 0;
 b = fix(Nt/2);
-c = Nt - 1;
+c = fix(Nt);
 
 Jx(1:Nx, 1:3) = 0;
 Jy(1:Nx, 1:3) = 0;

### 2017-10-02 PIC++/output/plot_E_derivative_z.m

@@ -1,0 +1,132 @@
+clear;
+load rotBFile.dat;
+load EderivativeFile.dat;
+load flux.dat
+load Zfile.dat;
+load Yfile.dat;
+load Zfile.dat;
+
+Nx = size(Xfile, 1);
+Ny = size(Yfile, 1);
+Nz = size(Zfile, 1);
+
+NE = Nz*Ny*Nz;
+Nt = (size(flux, 1)/NE)-1;
+
+ynumber = 2;
+xnumber = 2;
+
+a = 0;
+b = fix(Nt/2);
+c = fix(Nt);
+
+Jx(1:Nz, 1:3) = 0;
+Jy(1:Nz, 1:3) = 0;
+Jz(1:Nz, 1:3) = 0;
+
+rotBx(1:Nz, 1:3) = 0;
+rotBy(1:Nz, 1:3) = 0;
+rotBz(1:Nz, 1:3) = 0;
+
+derEx(1:Nz, 1:3) = 0;
+derEy(1:Nz, 1:3) = 0;
+derEz(1:Nz, 1:3) = 0;
+
+
+
+
+for i=1:Nz,
+   Jx(i,1) = 4*3.14*flux((Nz)*(Ny)*(xnumber-1) + (Nz)*(ynumber-1) + i + a*NE, 1);
+   Jx(i,2) = 4*3.14*flux((Nz)*(Ny)*(xnumber-1) + (Nz)*(ynumber-1) + i + b*NE, 1);
+   Jx(i,3) = 4*3.14*flux((Nz)*(Ny)*(xnumber-1) + (Nz)*(ynumber-1) + i + c*NE, 1);
+   Jy(i,1) = 4*3.14*flux((Nz)*(Ny)*(xnumber-1) + (Nz)*(ynumber-1) + i + a*NE, 2);
+   Jy(i,2) = 4*3.14*flux((Nz)*(Ny)*(xnumber-1) + (Nz)*(ynumber-1) + i + b*NE, 2);
+   Jy(i,3) = 4*3.14*flux((Nz)*(Ny)*(xnumber-1) + (Nz)*(ynumber-1) + i + c*NE, 2);
+   Jz(i,1) = 4*3.14*flux((Nz)*(Ny)*(xnumber-1) + (Nz)*(ynumber-1) + i + a*NE, 3);
+   Jz(i,2) = 4*3.14*flux((Nz)*(Ny)*(xnumber-1) + (Nz)*(ynumber-1) + i + b*NE, 3);
[…]

### 2017-10-02 PIC++/output/plot_concentrations_3d_xz.m

@@ -1,0 +1,83 @@
+clear;
+load concentrations.dat;
+load particleTypes.dat;
+load Xfile.dat;
+load Yfile.dat;
+load Zfile.dat;
+
+Nx = size(Xfile, 1)-1;
+Ny = size(Yfile, 1)-1;
+Nz = size(Zfile, 1)-1;
+
+N = Nx*Ny*Nz;
+
+Nt = size(concentrations, 1)/N;
+Ntypes = size(particleTypes, 1);
+
+a = 0;
+b = fix(Nt/2);
+c = Nt - 1;
+
+charge_density(1:Ny, 1:Nz) = 0;
+charge_density_hat(1:Ny, 1:Nz) = 0;
+particle_concentrations(1:Ny, 1:Nz, 1:Ntypes) = 0;
+
+xnumber = 3;
+
+middleY(1:Ny) = 0;
+middleZ(1:Nz) = 0;
+
+for i=1:Ny,  
+   middleY(i) = 0.5*(Yfile(i) + Yfile(i+1));
+   for j = 1:Nz,
+       charge_density(i, j) = concentrations(Nz*Ny*(xnumber-1) + Nz*(i-1) + j + c*N, 1);
+       charge_density_hat(i, j) = concentrations(Nz*Ny*(xnumber-1) + Nz*(i-1) + j + c*N, 2);
+       for t = 1:Ntypes,
+           particle_concentrations(i, j, t) = concentrations(Nz*Ny*(xnumber-1) + Nz*(i-1) + j + c*N, 2 + t);
+       end;
+   end;
+end;
+
+for j = 1:Nz,
+    middleZ(j) = 0.5*(Zfile(j) + Zfile(j+1));
+end;
+
+cons(1:Ny, 1:Nz) = 0;
+for t = 1:Ntypes,
+    if(particleTypes(t) > 0)
+        for i = 1:Ny,
+            for j = 1:Nz,
+                cons(i,j)=particle_concentrations(i, j, t);
+            end;
+        end;
+        figure(t);
+        [Y, Z] = meshgrid(middleZ, middleY);
+        surf(Y, Z, cons);
+        shading interp;
+        title ('concentration');
+        xlabel ('z');
+        ylabel ('y');
+        zlabel ('n cm^-3');
+        grid ;
+    end;
+end;
+
+figure(Ntypes + 1);
+[Y, Z] = meshgrid(middleZ, middleY);
+surf(Y, Z, charge_density);
[…]

### 2017-10-02 PIC++/output/plot_concentrations_3d_yz.m

@@ -1,0 +1,83 @@
+clear;
+load concentrations.dat;
+load particleTypes.dat;
+load Xfile.dat;
+load Yfile.dat;
+load Zfile.dat;
+
+Nx = size(Xfile, 1)-1;
+Ny = size(Yfile, 1)-1;
+Nz = size(Zfile, 1)-1;
+
+N = Nx*Ny*Nz;
+
+Nt = size(concentrations, 1)/N;
+Ntypes = size(particleTypes, 1);
+
+a = 0;
+b = fix(Nt/2);
+c = Nt - 1;
+
+charge_density(1:Ny, 1:Nz) = 0;
+charge_density_hat(1:Ny, 1:Nz) = 0;
+particle_concentrations(1:Ny, 1:Nz, 1:Ntypes) = 0;
+
+xnumber = 500;
+
+middleY(1:Ny) = 0;
+middleZ(1:Nz) = 0;
+
+for i=1:Ny,  
+   middleY(i) = 0.5*(Yfile(i) + Yfile(i+1));
+   for j = 1:Nz,
+       charge_density(i, j) = concentrations(Nz*Ny*(xnumber-1) + Nz*(i-1) + j + c*N, 1);
+       charge_density_hat(i, j) = concentrations(Nz*Ny*(xnumber-1) + Nz*(i-1) + j + c*N, 2);
+       for t = 1:Ntypes,
+           particle_concentrations(i, j, t) = concentrations(Nz*Ny*(xnumber-1) + Nz*(i-1) + j + c*N, 2 + t);
+       end;
+   end;
+end;
+
+for j = 1:Nz,
+    middleZ(j) = 0.5*(Zfile(j) + Zfile(j+1));
+end;
+
+cons(1:Ny, 1:Nz) = 0;
+for t = 1:Ntypes,
+    if(particleTypes(t) > 0)
+        for i = 1:Ny,
+            for j = 1:Nz,
+                cons(i,j)=particle_concentrations(i, j, t);
+            end;
+        end;
+        figure(t);
+        [Y, Z] = meshgrid(middleZ, middleY);
+        surf(Y, Z, cons);
+        shading interp;
+        title ('concentration');
+        xlabel ('z');
+        ylabel ('y');
+        zlabel ('n cm^-3');
+        grid ;
+    end;
+end;
+
+figure(Ntypes + 1);
+[Y, Z] = meshgrid(middleZ, middleY);
+surf(Y, Z, charge_density);
[…]

### 2017-10-02 PIC++/output/plot_distribution.m

@@ -2,6 +2,8 @@ clear;
 load distribution_protons.dat;
 load distribution_electrons.dat;
 load distribution_alphas.dat;
+load distribution_positrons.dat;
+load initialParameters.dat;
 
 Np = 500;
 
@@ -16,12 +18,14 @@ c = Nt - 1;
 Fp(1:Np, 1:3) = 0;
 Fe(1:Np, 1:3) = 0;
 Fa(1:Np, 1:3) = 0;
+Fpos(1:Np, 1:3) = 0;
 
 Pp(1:Np, 1:3) = 0;
 Pe(1:Np, 1:3) = 0;
 Pa(1:Np, 1:3) = 0;
+Ppos(1:Np, 1:3) = 0;
 
-me = 9.1*10^-26;
+me = initialParameters(36);
 v=3*10^10;
 
 for i=1:Np,
@@ -48,6 +52,14 @@ for i=1:Np,
    Fa(i,1) = distribution_alphas(i + a*Np, 2)*Pa(i,1)*Pa(i,1);
    Fa(i,2) = distribution_alphas(i + b*Np, 2)*Pa(i,2)*Pa(i,2);
    Fa(i,3) = distribution_alphas(i + c*Np, 2)*Pa(i,3)*Pa(i,3);
+   
+   Ppos(i,1) = distribution_positrons(i + a*Np,1);
+   Ppos(i,2) = distribution_positrons(i + b*Np,1);
+   Ppos(i,3) = distribution_positrons(i + c*Np,1);
+   
+   Fpos(i,1) = distribution_positrons(i + a*Np, 2)*Ppos(i,1)*Ppos(i,1);
+   Fpos(i,2) = distribution_positrons(i + b*Np, 2)*Ppos(i,2)*Ppos(i,2);
+   Fpos(i,3) = distribution_positrons(i + c*Np, 2)*Ppos(i,3)*Ppos(i,3);
 end;
 set(0,'DefaultAxesFontSize',14,'DefaultAxesFontName','Times New Roman');
 set(0,'DefaultTextFontSize',20,'DefaultTextFontName','Times New Roman'); 
@@ -72,5 +84,13 @@ plot (Pa(1:Np,1)/(me*v),Fa(1:Np,1), 'red',Pa(1:Np,2)/(me*v),Fa(1:Np,2), 'green',
 title ('alphas distribution function');
 xlabel ('p/{m_e c}');
 ylabel ('F_\alpha(p) p^4');
[…]

### 2017-10-02 PIC++/output/plot_distribution_downstream.m

@@ -1,0 +1,49 @@
+clear;
+load distribution_protons_downstream.dat;
+load distribution_electrons_downstream.dat;
+
+Np = 500;
+
+Nt = size(distribution_electrons_downstream, 1)/Np;
+
+a = 0;
+b = fix(Nt/2);
+c = Nt - 1;
+
+Fp(1:Np, 1:3) = 0;
+Fe(1:Np, 1:3) = 0;
+
+Pp(1:Np, 1:3) = 0;
+Pe(1:Np, 1:3) = 0;
+
+for i=1:Np,
+   Pp(i,1) = distribution_protons_downstream(i + a*Np,1);
+   Pp(i,2) = distribution_protons_downstream(i + b*Np,1);
+   Pp(i,3) = distribution_protons_downstream(i + c*Np,1);
+   
+   Fp(i,1) = distribution_protons_downstream(i + a*Np, 2)*Pp(i,1)*Pp(i,1);
+   Fp(i,2) = distribution_protons_downstream(i + b*Np, 2)*Pp(i,2)*Pp(i,2);
+   Fp(i,3) = distribution_protons_downstream(i + c*Np, 2)*Pp(i,3)*Pp(i,3);
+   
+   Pe(i,1) = distribution_electrons_downstream(i + a*Np,1);
+   Pe(i,2) = distribution_electrons_downstream(i + b*Np,1);
+   Pe(i,3) = distribution_electrons_downstream(i + c*Np,1);
+   
+   Fe(i,1) = distribution_electrons_downstream(i + a*Np, 2)*Pe(i,1)*Pe(i,1);
+   Fe(i,2) = distribution_electrons_downstream(i + b*Np, 2)*Pe(i,2)*Pe(i,2);
+   Fe(i,3) = distribution_electrons_downstream(i + c*Np, 2)*Pe(i,3)*Pe(i,3);
+end;
+
+figure(1);
+plot (Pp(1:Np,1),Fp(1:Np,1), 'red',Pp(1:Np,2),Fp(1:Np,2), 'green',Pp(1:Np,3),Fp(1:Np,3), 'blue');
+title ('protons distinution function');
+xlabel ('p');
+ylabel ('Fp');
+grid ;
+
+figure(2);
+plot (Pe(1:Np,1),Fe(1:Np,1), 'red',Pe(1:Np,2),Fe(1:Np,2), 'green',Pe(1:Np,3),Fe(1:Np,3), 'blue');
[…]

### 2017-10-02 PIC++/output/plot_divergence.m

@@ -1,5 +1,5 @@
 clear;
-load concentrations.dat;
+%load concentrations.dat;
 load Xfile.dat;
 load Yfile.dat;
 load Zfile.dat;
@@ -10,10 +10,10 @@ Ny = size(Yfile, 1)-1;
 Nz = size(Zfile, 1)-1;
 
 N = (Nx)*(Ny)*(Nz);
-Nt = fix(size(concentrations, 1)/N)-1;
+Nt = fix(size(divergence_error, 1)/N)-1;
 
-ynumber = 1;
-znumber = 1;
+ynumber = 3;
+znumber = 2;
 
 a = 0;
 b = fix(Nt/2);
@@ -22,6 +22,7 @@ c = Nt;
 divergenceError(1:Nx, 1:3) = 0;
 divergence(1:Nx, 1:3) = 0;
 density(1:Nx, 1:3) = 0;
+divergenceB(1:Nx, 1:3) = 0;
 
 middleX(1:Nx)=0;
 
@@ -37,6 +38,9 @@ for i=1:Nx,
    density(i,1) = divergence_error((Nz*Ny*(i-1) + Nz*(ynumber-1) + znumber) + a*N, 3);
    density(i,2) = divergence_error((Nz*Ny*(i-1) + Nz*(ynumber-1) + znumber) + b*N, 3);
    density(i,3) = divergence_error((Nz*Ny*(i-1) + Nz*(ynumber-1) + znumber) + c*N, 3);
+   divergenceB(i, 1) = divergence_error((Nz*Ny*(i-1) + Nz*(ynumber-1) + znumber) + a*N, 4);
+   divergenceB(i, 2) = divergence_error((Nz*Ny*(i-1) + Nz*(ynumber-1) + znumber) + b*N, 4);
+   divergenceB(i, 3) = divergence_error((Nz*Ny*(i-1) + Nz*(ynumber-1) + znumber) + c*N, 4);
 end;
 
 figure(1);
@@ -56,6 +60,13 @@ grid ;
 figure(3);
 plot (middleX(1:Nx),density(1:Nx, 1), 'red', middleX(1:Nx), density(1:Nx, 2), 'green', middleX(1:Nx), density(1:Nx, 3), 'blue');
 title ('4*pi*density');
+xlabel ('x cm');
+ylabel ('rho sgs*cm^-3');
+grid ;
+
+figure(4);
[…]

### 2017-10-02 PIC++/output/plot_divergence_3d.m

@@ -1,0 +1,82 @@
+clear;
+%load concentrations.dat;
+load Xfile.dat;
+load Yfile.dat;
+load Zfile.dat;
+load divergence_error.dat;
+load initialParameters.dat;
+
+Nx = size(Xfile, 1)-1;
+Ny = size(Yfile, 1)-1;
+Nz = size(Zfile, 1)-1;
+
+N = (Nx)*(Ny)*(Nz);
+Nt = fix(size(divergence_error, 1)/N)-1;
+
+znumber = 2;
+
+a = 0;
+b = fix(Nt/2);
+c = Nt;
+
+divergenceError(1:Nx, 1:Ny) = 0;
+divergence(1:Nx, 1:Ny) = 0;
+density(1:Nx, 1:Ny) = 0;
+divergenceB(1:Nx, 1:Ny) = 0;
+
+middleX(1:Nx)=0;
+middleY(1:Ny)=0;
+
+cv = initialParameters(10);
+omega = initialParameters(21);
+omegaElectron = initialParameters(20);
+
+for i=1:Nx,
+   middleX(i) = ((Xfile(i)+Xfile(i+1))/2 - Xfile(2))*omegaElectron/cv;
+   for j =1:Ny,
+      divergenceError(i, j) = divergence_error((Nz*Ny*(i-1) + Nz*(j-1) + znumber) + c*N, 1);
+      divergence(i, j) = divergence_error((Nz*Ny*(i-1) + Nz*(j-1) + znumber) + c*N, 2);
+      density(i,j) = divergence_error((Nz*Ny*(i-1) + Nz*(j-1) + znumber) + c*N, 3);
+      divergenceB(i, j) = divergence_error((Nz*Ny*(i-1) + Nz*(j-1) + znumber) + c*N, 4);
+   end;
+end;
+
+for j = 1:Ny,
+    middleY(j) = (0.5*(Yfile(j) + Yfile(j+1)) - Yfile(2))*omegaElectron/cv;
+end;
+
+figure(1)
+[X, Y] = meshgrid(middleY, middleX);
+surf(X, Y, divergenceError);
+shading interp;
+title ('divergence error');
+xlabel ('y\omega_e/c');
+ylabel ('x\omega_e/c');
+zlabel ('divergence error e cgs/cm^3');
+
+figure(2)
+[X, Y] = meshgrid(middleY, middleX);
+surf(X, Y, divergence);
+shading interp;
+title ('divergence');
[…]

### 2017-10-02 PIC++/output/plot_fields.m

@@ -13,12 +13,14 @@ Nz = size(Zfile, 1);
 NE = Nx*Ny*Nz;
 NB = (Nx-1)*(Ny-1)*(Nz-1);
 Nt = (size(Efield, 1)/NE);
+%Nt=10;
+%Nt = 23;
 NtB = (size(Bfield, 1)/NB);
-ynumber = 1;
-znumber = 1;
+ynumber = 2;
+znumber = 2;
 a = 0;
 b = fix(Nt/2);
-c = Nt-1;
+c = fix(Nt)-1;
 
 Ex(1:Nx, 1:3) = 0;
 Ey(1:Nx, 1:3) = 0;
@@ -34,12 +36,13 @@ B0=initialParameters(19);
 middleX(1:Nx-1) = 0;
 Xgrid(1:Nx) = 0;
 cv = initialParameters(10);
-omega = initialParameters(20);
+omega = initialParameters(21);
+omegaElectron = initialParameters(20);
 
 
 for i=1:Nx,
-   %Xgrid(i) = (Xfile(i) - Xfile(2))*omega/cv;
-   Xgrid(i) = (Xfile(i) - Xfile(2));
+   Xgrid(i) = (Xfile(i) - Xfile(2))*omegaElectron/cv;
+   %Xgrid(i) = (Xfile(i) - Xfile(2));
    Ex(i,1) = Efield((Nz)*(Ny)*(i-1) + (Nz)*(ynumber-1) + znumber + a*NE, 1);
    Ex(i,2) = Efield((Nz)*(Ny)*(i-1) + (Nz)*(ynumber-1) + znumber + b*NE, 1);
    Ex(i,3) = Efield((Nz)*(Ny)*(i-1) + (Nz)*(ynumber-1) + znumber + c*NE, 1);
@@ -53,8 +56,8 @@ end;
 %ynumber = 1;
 %znumber = 1;
 for i = 1:Nx-1,
-   %middleX(i) = (0.5*(Xfile(i) + Xfile(i+1)) - Xfile(2))*omega/cv;
-   middleX(i) = (0.5*(Xfile(i) + Xfile(i+1)) - Xfile(2));
+   middleX(i) = (0.5*(Xfile(i) + Xfile(i+1)) - Xfile(2))*omegaElectron/cv;
+   %middleX(i) = (0.5*(Xfile(i) + Xfile(i+1)) - Xfile(2));
[…]

### 2017-10-02 PIC++/output/plot_fields_3d_yz.m

@@ -12,11 +12,11 @@ Nz = size(Zfile, 1);
 NE = Nx*Ny*Nz;
 NB = (Nx-1)*(Ny-1)*(Nz-1);
 Nt = fix(size(Efield, 1)/NE);
-xnumber = fix(Nx/2);
+xnumber = 2;
 
 a = 0;
 b = fix(Nt/2);
-c = Nt - 1;
+c = fix(Nt)-2;
 
 Ex(1:Ny, 1:Nz) = 0;
 Ey(1:Ny, 1:Nz) = 0;
@@ -54,6 +54,7 @@ end;
 figure(1)
 [Y, Z] = meshgrid(Zfile, Yfile);
 surf(Y, Z, Ex);
+shading interp;
 title ('Ex');
 xlabel ('z');
 ylabel ('y');
@@ -63,6 +64,7 @@ grid ;
 figure(2)
 [Y, Z] = meshgrid(Zfile, Yfile);
 surf(Y, Z, Ey);
+shading interp;
 title ('Ey');
 xlabel ('z');
 ylabel ('y');
@@ -72,6 +74,7 @@ grid ;
 figure(3)
 [Y, Z] = meshgrid(Zfile, Yfile);
 surf(Y, Z, Ez);
+shading interp;
 title ('Ez');
 xlabel ('z');
 ylabel ('y');
@@ -81,6 +84,7 @@ grid ;
 figure(4)
 [Y, Z] = meshgrid(middleZ, middleY);
 surf(Y, Z, Bx);
+shading interp;
 title ('Bx');
 xlabel ('z');
 ylabel ('x');
@@ -90,6 +94,7 @@ grid ;
 figure(5)
 [Y, Z] = meshgrid(middleZ, middleY);
 surf(Y, Z, By);
+shading interp;
 title ('By');
 xlabel ('z');
 ylabel ('y');
@@ -99,6 +104,7 @@ grid ;
 figure(6)
 [Y, Z] = meshgrid(middleZ, middleY);
 surf(Y, Z, Bz);
+shading interp;
 title ('Bz');
 xlabel ('z');
 ylabel ('y');

### 2017-10-02 PIC++/output/plot_fields_section_xy.m

@@ -1,0 +1,124 @@
+clear;
+load EfieldXY.dat;
+load BfieldXY.dat;
+load Xfile.dat;
+load Yfile.dat;
+load Zfile.dat;
+load initialParameters.dat;
+
+Nx = size(Xfile, 1);
+Ny = size(Yfile, 1);
+Nz = size(Zfile, 1);
+
+NE = Nx*Ny;
+NB = (Nx-1)*(Ny-1);
+Nt = fix(size(EfieldXY, 1)/NE);
+NtB = (size(BfieldXY, 1)/NB);
+
+a = 0;
+b = fix(Nt/2);
+c = fix(Nt)-1;
+%c = 0;
+
+Ex(1:Nx, 1:Ny) = 0;
+Ey(1:Nx, 1:Ny) = 0;
+Ez(1:Nx, 1:Ny) = 0;
+
+Bx(1:Nx-1, 1:Ny-1) = 0;
+By(1:Nx-1, 1:Ny-1) = 0;
+Bz(1:Nx-1, 1:Ny-1) = 0;
+
+middleX(1:Nx-1) = 0;
+middleY(1:Ny-1) = 0;
+Xgrid(1:Nx) = 0;
+Ygrid(1:Ny) = 0;
+cv = initialParameters(10);
+omega = initialParameters(21);
+omegaElectron = initialParameters(20);
+
+
+
+for i=1:Nx,
+    Xgrid(i) = (Xfile(i) - Xfile(2))*omegaElectron/cv;
+   %Xgrid(i) = (Xfile(i) - Xfile(2));
+    for j = 1:Ny,
+        Ex(i,j) = EfieldXY((Ny)*(i-1) + j + c*NE, 1);
+        Ey(i,j) = EfieldXY((Ny)*(i-1) + j + c*NE, 2);
+        Ez(i,j) = EfieldXY((Ny)*(i-1) + j + c*NE, 3);
+    end;
+end;
+
+for j = 1:Ny,
+    Ygrid(j) = (Yfile(j) - Yfile(2))*omegaElectron/cv;
+   %Ygrid(i) = (Yfile(i) - Yfile(2));
+end;
+
+for i = 1:Nx-1,
+   middleX(i) = (0.5*(Xfile(i) + Xfile(i+1)) - Xfile(2))*omegaElectron/cv;
+   %middleX(i) = (0.5*(Xfile(i) + Xfile(i+1)) - Xfile(2));
[…]

### 2017-10-02 PIC++/output/plot_fields_section_xy_time.m

@@ -1,0 +1,260 @@
+clear;
+load EfieldXY.dat;
+load BfieldXY.dat;
+load Xfile.dat;
+load Yfile.dat;
+load Zfile.dat;
+load initialParameters.dat;
+
+Nx = size(Xfile, 1);
+Ny = size(Yfile, 1);
+Nz = size(Zfile, 1);
+
+NE = Nx*Ny;
+NB = (Nx-1)*(Ny-1);
+Nt = fix(size(EfieldXY, 1)/NE);
+NtB = (size(BfieldXY, 1)/NB);
+
+a = 0;
+b = fix(Nt/2);
+c = fix(Nt)-1;
+%c = 0;
+
+Ex1(1:Nx, 1:Ny) = 0;
+Ey1(1:Nx, 1:Ny) = 0;
+Ez1(1:Nx, 1:Ny) = 0;
+
+Ex2(1:Nx, 1:Ny) = 0;
+Ey2(1:Nx, 1:Ny) = 0;
+Ez2(1:Nx, 1:Ny) = 0;
+
+Ex3(1:Nx, 1:Ny) = 0;
+Ey3(1:Nx, 1:Ny) = 0;
+Ez3(1:Nx, 1:Ny) = 0;
+
+Bx1(1:Nx-1, 1:Ny-1) = 0;
+By1(1:Nx-1, 1:Ny-1) = 0;
+Bz1(1:Nx-1, 1:Ny-1) = 0;
+
+Bx2(1:Nx-1, 1:Ny-1) = 0;
+By2(1:Nx-1, 1:Ny-1) = 0;
+Bz2(1:Nx-1, 1:Ny-1) = 0;
+
+Bx3(1:Nx-1, 1:Ny-1) = 0;
+By3(1:Nx-1, 1:Ny-1) = 0;
+Bz3(1:Nx-1, 1:Ny-1) = 0;
+
+middleX(1:Nx-1) = 0;
+middleY(1:Ny-1) = 0;
+Xgrid(1:Nx) = 0;
+Ygrid(1:Ny) = 0;
+cv = initialParameters(10);
+omega = initialParameters(21);
+omegaElectron = initialParameters(20);
+
+
+
+for i=1:Nx,
+    Xgrid(i) = (Xfile(i) - Xfile(2))*omegaElectron/cv;
+   %Xgrid(i) = (Xfile(i) - Xfile(2));
+    for j = 1:Ny,
[…]

### 2017-10-02 PIC++/output/plot_fields_section_xz_time.m

@@ -1,0 +1,259 @@
+clear;
+load EfieldXZ.dat;
+load BfieldXZ.dat;
+load Xfile.dat;
+load Yfile.dat;
+load Zfile.dat;
+
+Nx = size(Xfile, 1);
+Ny = size(Yfile, 1);
+Nz = size(Zfile, 1);
+
+NE = Nx*Nz;
+NB = (Nx-1)*(Nz-1);
+Nt = fix(size(EfieldXZ, 1)/NE);
+
+a = 0;
+b = fix(Nt/2);
+c = fix(Nt)-1;
+
+Ex1(1:Nx, 1:Nz) = 0;
+Ey1(1:Nx, 1:Nz) = 0;
+Ez1(1:Nx, 1:Nz) = 0;
+
+Ex2(1:Nx, 1:Nz) = 0;
+Ey2(1:Nx, 1:Nz) = 0;
+Ez2(1:Nx, 1:Nz) = 0;
+
+Ex3(1:Nx, 1:Nz) = 0;
+Ey3(1:Nx, 1:Nz) = 0;
+Ez3(1:Nx, 1:Nz) = 0;
+
+Bx1(1:Nx-1, 1:Nz-1) = 0;
+By1(1:Nx-1, 1:Nz-1) = 0;
+Bz1(1:Nx-1, 1:Nz-1) = 0;
+
+Bx2(1:Nx-1, 1:Nz-1) = 0;
+By2(1:Nx-1, 1:Nz-1) = 0;
+Bz2(1:Nx-1, 1:Nz-1) = 0;
+
+Bx3(1:Nx-1, 1:Nz-1) = 0;
+By3(1:Nx-1, 1:Nz-1) = 0;
+Bz3(1:Nx-1, 1:Nz-1) = 0;
+
+middleX(1:Nx-1) = 0;
+middleZ(1:Nz-1) = 0;
+
+
+for i=1:Nx,
+    for j = 1:Nz,
+        Ex1(i,j) = EfieldXZ((Nz)*(i-1) + j + a*NE, 1);
+        Ey1(i,j) = EfieldXZ((Nz)*(i-1) + j + a*NE, 2);
+        Ez1(i,j) = EfieldXZ((Nz)*(i-1) + j + a*NE, 3);
+        Ex2(i,j) = EfieldXZ((Nz)*(i-1) + j + b*NE, 1);
+        Ey2(i,j) = EfieldXZ((Nz)*(i-1) + j + b*NE, 2);
[…]

### 2017-10-02 PIC++/output/plot_flux_y.m

@@ -1,0 +1,70 @@
+clear;
+load flux.dat;
+load Xfile.dat;
+load Yfile.dat;
+load Zfile.dat;
+
+Nx = size(Xfile, 1);
+Ny = size(Yfile, 1);
+Nz = size(Zfile, 1);
+
+NE = Nx*Ny*Nz;
+Nt = fix(size(flux,1)/NE);
+
+xnumber = 2;
+znumber = 2;
+
+a = 0;
+b = fix(Nt/2);
+c = Nt-1;
+
+Jx(1:Ny, 1:3) = 0;
+Jy(1:Ny, 1:3) = 0;
+Jz(1:Ny, 1:3) = 0;
+
+%extJx(1:Nx, 1:3) = 0;
+%extJy(1:Nx, 1:3) = 0;
+%extJz(1:Nx, 1:3) = 0;
+
+
+for i=1:Ny,
+   Jx(i,1) = flux(Nz*Ny*(xnumber-1) + Nz*(i-1) + znumber + a*NE, 1);
+   Jx(i,2) = flux(Nz*Ny*(xnumber-1) + Nz*(i-1) + znumber + b*NE, 1);
+   Jx(i,3) = flux(Nz*Ny*(xnumber-1) + Nz*(i-1) + znumber + c*NE, 1);
+   Jy(i,1) = flux(Nz*Ny*(xnumber-1) + Nz*(i-1) + znumber + a*NE, 2);
+   Jy(i,2) = flux(Nz*Ny*(xnumber-1) + Nz*(i-1) + znumber + b*NE, 2);
+   Jy(i,3) = flux(Nz*Ny*(xnumber-1) + Nz*(i-1) + znumber + c*NE, 2);
+   Jz(i,1) = flux(Nz*Ny*(xnumber-1) + Nz*(i-1) + znumber + a*NE, 3);
+   Jz(i,2) = flux(Nz*Ny*(xnumber-1) + Nz*(i-1) + znumber + b*NE, 3);
+   Jz(i,3) = flux(Nz*Ny*(xnumber-1) + Nz*(i-1) + znumber + c*NE, 3);
+ 
+   %extJx(i,1) = flux(Nz*Ny*(i-1) + Nz*(ynumber-1) + znumber + a*NE, 4);
+   %extJx(i,2) = flux(Nz*Ny*(i-1) + Nz*(ynumber-1) + znumber + b*NE, 4);
[…]

### 2017-10-02 PIC++/output/plot_general_anisotropy.m

@@ -1,0 +1,45 @@
+clear;
+load generalAnisotropy.dat;
+load particleTypes.dat;
+N1=1;
+N2=size(generalAnisotropy,1);
+Ntypes = size(particleTypes,1);
+linearAnisotropy(1:N2, 1:Ntypes) = 0;
+
+%omega_plasma = 4.432550293*10^10;
+omega_plasma = 2*3.14159*generalAnisotropy(2,2)/generalAnisotropy(2,3);
+%omega_plasma = 4.37*10^11;
+omega_gyro_a = 1.982193107*10^8;
+%gamma = 0.01*omega_gyro_a;
+gamma = 0.01*2*3.14159*omega_gyro_a/omega_plasma;
+
+for i = 1:Ntypes,
+    if(particleTypes(i) > 0)
+        linearAnisotropy(1,i) = generalAnisotropy(1, 4 + 3*(i-1));
+        for j = 2:N2,
+            linearAnisotropy(j,i) = linearAnisotropy(1, i)*exp(-2*gamma*generalAnisotropy(j,2));
+        end
+        figure(3*(i-1) +1);
+        plot (generalAnisotropy(1:N2,2), generalAnisotropy(1:N2,4 + 3*(i-1)), 'red');
+        %if (i == 6)
+        %   plot (generalAnisotropy(1:N2,2), generalAnisotropy(1:N2,3 + i), 'red',generalAnisotropy(1:N2,2), linearAnisotropy(1:N2,i), 'blue');
+        %end;
+        title ('anisotropy');
+        xlabel ('{{t w_p}/{2\pi}}');
+        ylabel ('T1/T2 - T2/T1');
+        grid ;      
+        figure(3*(i-1) + 2);
+        plot (generalAnisotropy(1:N2,2), generalAnisotropy(1:N2,5 + 3*(i-1)), 'red');
+        title ('temperature parallel');
+        xlabel ('{{t w_p}/{2\pi}}');
+        ylabel ('T2');
+        grid ;     
+        figure(3*(i-1) + 3);
+        plot (generalAnisotropy(1:N2,2), generalAnisotropy(1:N2,6 + 3*(i-1)), 'red');
+        title ('temperature normal');
+        xlabel ('{{t w_p}/{2\pi}}');
+        ylabel ('T1');
+        grid ;  
+    end;
+end;
+

### 2017-10-02 PIC++/output/plot_particles.m

@@ -7,6 +7,14 @@ load deuterium.dat;
 load helium3.dat;
 load oxygen_plus_3.dat;
 load silicon_plus_1.dat;
+load initialParameters.dat;
+load Xfile.dat;
+load Yfile.dat;
+load Zfile.dat;
+
+cv = initialParameters(10);
+omega = initialParameters(21);
+omegaElectron = initialParameters(20);
 
 N1=size(electron,1);
 N2=size(proton,1);
@@ -18,14 +26,14 @@ N7 = size(oxygen_plus_3,1);
 N8 = size(silicon_plus_1,1);
 if(N1 > 0)
     figure(1);
-    plot (electron(1:N1,1), electron(1:N1,2),'blue');
+    plot ((electron(1:N1,1) - Xfile(2))*omegaElectron/cv, electron(1:N1,7),'blue');
     title ('electrons');
     xlabel ('x cm');
     ylabel ('P g*cm/s');
     %legend('electrons','Location','northeast');
     grid ; 
     figure(2);
-    plot (electron(1:N1,1), electron(1:N1,3), 'blue');
+    plot ((electron(1:N1,1) - Xfile(2))*omegaElectron/cv, electron(1:N1,2), 'blue');
     title ('electrons');
     xlabel ('x cm');
     ylabel ('Px g*cm/s');
@@ -35,14 +43,14 @@ end;
 
 if(N2 > 0)
     figure(3);
-    plot (proton(1:N2,1), proton(1:N2,2),'blue');
+    plot ((proton(1:N2,1) - Xfile(2))*omegaElectron/cv, proton(1:N2,7),'blue');
     title ('protons');
     xlabel ('x cm');
     ylabel ('P g*cm/s');
     %legend('protons','Location','northeast');
     grid ; 
     figure(4);
-    plot (proton(1:N2,1), proton(1:N2,3), 'blue');
+    plot ((proton(1:N2,1) - Xfile(2))*omegaElectron/cv, proton(1:N2,2), 'blue');
     title ('protons');
     xlabel ('x cm');
     ylabel ('Px g*cm/s');
@@ -52,14 +60,14 @@ end;
 
 if(N3 > 0)
     figure(5);
-    plot (positron(1:N3,1), positro(1:N3,2),'blue');
+    plot ((positron(1:N3,1) - Xfile(2))*omegaElectron/cv, positron(1:N3,7),'blue');
[…]

### 2017-10-02 PIC++/output/plot_right_part.m

@@ -1,0 +1,73 @@
+clear;
+load rightPart.dat;
+load Xfile.dat;
+load Yfile.dat;
+load Zfile.dat;
+load initialParameters.dat;
+
+Nx = size(Xfile, 1);
+Ny = size(Yfile, 1);
+Nz = size(Zfile, 1);
+
+NB = (Nx-1)*(Ny-1)*(Nz-1);
+Nt = (size(rightPart, 1)/NB);
+ynumber = 3;
+znumber = 3;
+a = 0;
+b = fix(Nt/2);
+c = fix(Nt)-1;
+
+
+Bx(1:Nx-1, 1:3) = 0;
+By(1:Nx-1, 1:3) = 0;
+Bz(1:Nx-1, 1:3) = 0;
+
+B0=initialParameters(19);
+
+middleX(1:Nx-1) = 0;
+Xgrid(1:Nx) = 0;
+cv = initialParameters(10);
+omega = initialParameters(21);
+omegaElectron = initialParameters(20);
+
+
+
+
+for i = 1:Nx-1,
+   middleX(i) = (0.5*(Xfile(i) + Xfile(i+1)) - Xfile(2))*omegaElectron/cv;
+   %middleX(i) = (0.5*(Xfile(i) + Xfile(i+1)) - Xfile(2));
+   Bx(i, 1) = rightPart(((Nz-1)*(Ny-1)*(i-1) + (Nz-1)*(ynumber-1) + znumber) + a*NB, 1);
+   Bx(i, 2) = rightPart(((Nz-1)*(Ny-1)*(i-1) + (Nz-1)*(ynumber-1) + znumber) + b*NB, 1);
+   Bx(i, 3) = rightPart(((Nz-1)*(Ny-1)*(i-1) + (Nz-1)*(ynumber-1) + znumber) + c*NB, 1);
+   By(i, 1) = rightPart(((Nz-1)*(Ny-1)*(i-1) + (Nz-1)*(ynumber-1) + znumber) + a*NB, 2);
+   By(i, 2) = rightPart(((Nz-1)*(Ny-1)*(i-1) + (Nz-1)*(ynumber-1) + znumber) + b*NB, 2);
+   By(i, 3) = rightPart(((Nz-1)*(Ny-1)*(i-1) + (Nz-1)*(ynumber-1) + znumber) + c*NB, 2);
+   Bz(i, 1) = rightPart(((Nz-1)*(Ny-1)*(i-1) + (Nz-1)*(ynumber-1) + znumber) + a*NB, 3);
[…]

### 2017-10-02 PIC++/output/plot_traectory_electron_2.m

@@ -1,0 +1,52 @@
+clear;
+load trajectory_electron_2.dat;
+N1=1;
+N2=size(trajectory_electron_2,1);
+
+xteoretic(1:N2) = 0;
+Efield = 1E-6;
+q = -4.803529695E-10;
+m = 0.910938291E-27;
+
+for i = 1:N2;
+    xteoretic(i) = trajectory_electron_2(1,3) + (q*Efield*0.5/m)*(trajectory_electron_2(i,2))^2;
+end;
+
+
+figure(1);
+%plot (traectory_electron(1:N2,1),traectory_electron(1:N2,3),'red',traectory_electron(1:N2,1),traectory_electron(1:N2,4),'green',traectory_electron(1:N2,1),traectory_electron(1:N2,5),'blue', traectory_electron(1:N2,1), xteoretic(1:N2), 'yellow');
+plot (trajectory_electron_2(1:N2,1),trajectory_electron_2(1:N2,3),'red',trajectory_electron_2(1:N2,1),trajectory_electron_2(1:N2,4),'green',trajectory_electron_2(1:N2,1),trajectory_electron_2(1:N2,5),'blue');
+title ('r');
+xlabel ('t');
+ylabel ('r');
+legend('x', 'y','z','Location','southwest');
+grid ;
+
+figure(2);
+plot (trajectory_electron_2(1:N2,1),trajectory_electron_2(1:N2,6),'red',trajectory_electron_2(1:N2,1),trajectory_electron_2(1:N2,7),'green',trajectory_electron_2(1:N2,1),trajectory_electron_2(1:N2,8),'blue', trajectory_electron_2(1:N2,2),trajectory_electron_2(1:N2,9),'black');
+title ('p');
+xlabel ('t');
+ylabel ('p');
+legend('px', 'py','pz', 'p','Location','southwest');
+grid ;
+
+figure(3);
+plot (trajectory_electron_2(1:N2,4),trajectory_electron_2(1:N2,5));
+title ('r');
+xlabel ('y');
+ylabel ('z');
+grid ;
+
+figure(4);
+plot (trajectory_electron_2(1:N2,3),trajectory_electron_2(1:N2,9));
+title ('p(x)');
+xlabel ('x');
+ylabel ('p');
+grid ;
+
+figure(5);
+plot (trajectory_electron_2(1:N2,3),trajectory_electron_2(1:N2,6));
+title ('p_x(x)');
+xlabel ('x');
+ylabel ('p_x');
+grid ;

### 2017-10-02 PIC++/output/plot_traectory_electron_4.m

@@ -1,0 +1,52 @@
+clear;
+load trajectory_electron_4.dat;
+N1=1;
+N2=size(trajectory_electron_4,1);
+
+xteoretic(1:N2) = 0;
+Efield = 1E-6;
+q = -4.803529695E-10;
+m = 0.910938291E-27;
+
+for i = 1:N2;
+    xteoretic(i) = trajectory_electron_4(1,3) + (q*Efield*0.5/m)*(trajectory_electron_4(i,2))^2;
+end;
+
+
+figure(1);
+%plot (traectory_electron(1:N2,1),traectory_electron(1:N2,3),'red',traectory_electron(1:N2,1),traectory_electron(1:N2,4),'green',traectory_electron(1:N2,1),traectory_electron(1:N2,5),'blue', traectory_electron(1:N2,1), xteoretic(1:N2), 'yellow');
+plot (trajectory_electron_4(1:N2,1),trajectory_electron_4(1:N2,3),'red',trajectory_electron_4(1:N2,1),trajectory_electron_4(1:N2,4),'green',trajectory_electron_4(1:N2,1),trajectory_electron_4(1:N2,5),'blue');
+title ('r');
+xlabel ('t');
+ylabel ('r');
+legend('x', 'y','z','Location','southwest');
+grid ;
+
+figure(2);
+plot (trajectory_electron_4(1:N2,1),trajectory_electron_4(1:N2,6),'red',trajectory_electron_4(1:N2,1),trajectory_electron_4(1:N2,7),'green',trajectory_electron_4(1:N2,1),trajectory_electron_4(1:N2,8),'blue', trajectory_electron_4(1:N2,2),trajectory_electron_4(1:N2,9),'black');
+title ('p');
+xlabel ('t');
+ylabel ('p');
+legend('px', 'py','pz', 'p','Location','southwest');
+grid ;
+
+figure(3);
+plot (trajectory_electron_4(1:N2,4),trajectory_electron_4(1:N2,5));
+title ('r');
+xlabel ('y');
+ylabel ('z');
+grid ;
+
+figure(4);
+plot (trajectory_electron_4(1:N2,3),trajectory_electron_4(1:N2,9));
+title ('p(x)');
+xlabel ('x');
+ylabel ('p');
+grid ;
+
+figure(5);
+plot (trajectory_electron_4(1:N2,3),trajectory_electron_4(1:N2,6));
+title ('p_x(x)');
+xlabel ('x');
+ylabel ('p_x');
+grid ;

### 2017-10-02 PIC++/output/plot_traectory_electron_5.m

@@ -1,0 +1,52 @@
+clear;
+load trajectory_electron_5.dat;
+N1=1;
+N2=size(trajectory_electron_5,1);
+
+xteoretic(1:N2) = 0;
+Efield = 1E-6;
+q = -4.803529695E-10;
+m = 0.910938291E-27;
+
+for i = 1:N2;
+    xteoretic(i) = trajectory_electron_5(1,3) + (q*Efield*0.5/m)*(trajectory_electron_5(i,2))^2;
+end;
+
+
+figure(1);
+%plot (traectory_electron(1:N2,1),traectory_electron(1:N2,3),'red',traectory_electron(1:N2,1),traectory_electron(1:N2,4),'green',traectory_electron(1:N2,1),traectory_electron(1:N2,5),'blue', traectory_electron(1:N2,1), xteoretic(1:N2), 'yellow');
+plot (trajectory_electron_5(1:N2,1),trajectory_electron_5(1:N2,3),'red',trajectory_electron_5(1:N2,1),trajectory_electron_5(1:N2,4),'green',trajectory_electron_5(1:N2,1),trajectory_electron_5(1:N2,5),'blue');
+title ('r');
+xlabel ('t');
+ylabel ('r');
+legend('x', 'y','z','Location','southwest');
+grid ;
+
+figure(2);
+plot (trajectory_electron_5(1:N2,1),trajectory_electron_5(1:N2,6),'red',trajectory_electron_5(1:N2,1),trajectory_electron_5(1:N2,7),'green',trajectory_electron_5(1:N2,1),trajectory_electron_5(1:N2,8),'blue', trajectory_electron_5(1:N2,2),trajectory_electron_5(1:N2,9),'black');
+title ('p');
+xlabel ('t');
+ylabel ('p');
+legend('px', 'py','pz', 'p','Location','southwest');
+grid ;
+
+figure(3);
+plot (trajectory_electron_5(1:N2,4),trajectory_electron_5(1:N2,5));
+title ('r');
+xlabel ('y');
+ylabel ('z');
+grid ;
+
+figure(4);
+plot (trajectory_electron_5(1:N2,3),trajectory_electron_5(1:N2,9));
+title ('p(x)');
+xlabel ('x');
+ylabel ('p');
+grid ;
+
+figure(5);
+plot (trajectory_electron_5(1:N2,3),trajectory_electron_5(1:N2,6));
+title ('p_x(x)');
+xlabel ('x');
+ylabel ('p_x');
+grid ;

### 2017-10-02 PIC++/output/plot_traectory_proton_1.m

@@ -1,0 +1,43 @@
+clear;
+load trajectory_proton_1.dat;
+N1=1;
+N2=size(trajectory_proton_1,1);
+
+
+figure(1);
+plot (trajectory_proton_1(1:N2,1),trajectory_proton_1(1:N2,3),'red',trajectory_proton_1(1:N2,1),trajectory_proton_1(1:N2,4),'green',trajectory_proton_1(1:N2,1),trajectory_proton_1(1:N2,5),'blue');
+title ('r');
+xlabel ('t');
+ylabel (')');
+legend(4, 'x', 'y','z');
+grid ;
+
+figure(2);
+plot (trajectory_proton_1(1:N2,1),trajectory_proton_1(1:N2,6),'red',trajectory_proton_1(1:N2,1),trajectory_proton_1(1:N2,7),'green',trajectory_proton_1(1:N2,1),trajectory_proton_1(1:N2,8),'blue', trajectory_proton_1(1:N2,1),trajectory_proton_1(1:N2,9),'black');
+title ('p');
+xlabel ('t');
+ylabel ('p');
+legend(4, 'px', 'py','pz', 'p');
+grid ;
+
+figure(3);
+plot (trajectory_proton_1(1:N2,4),trajectory_proton_1(1:N2,5));
+title ('r');
+xlabel ('y');
+ylabel ('z');
+grid ;
+
+figure(4);
+plot (trajectory_proton_1(1:N2,3),trajectory_proton_1(1:N2,9));
+title ('p(x)');
+xlabel ('x');
+ylabel ('p');
+grid ;
+
+figure(5);
+plot (trajectory_proton_1(1:N2,3),trajectory_proton_1(1:N2,6));
+title ('p_x(x)');
+xlabel ('x');
+ylabel ('p_x');
+grid ;
+

### 2017-10-02 PIC++/output/plot_traectory_proton_3.m

@@ -1,0 +1,43 @@
+clear;
+load trajectory_proton_3.dat;
+N1=1;
+N2=size(trajectory_proton_3,1);
+
+
+figure(1);
+plot (trajectory_proton_3(1:N2,1),trajectory_proton_3(1:N2,3),'red',trajectory_proton_3(1:N2,1),trajectory_proton_3(1:N2,4),'green',trajectory_proton_3(1:N2,1),trajectory_proton_3(1:N2,5),'blue');
+title ('r');
+xlabel ('t');
+ylabel (')');
+legend(4, 'x', 'y','z');
+grid ;
+
+figure(2);
+plot (trajectory_proton_3(1:N2,1),trajectory_proton_3(1:N2,6),'red',trajectory_proton_3(1:N2,1),trajectory_proton_3(1:N2,7),'green',trajectory_proton_3(1:N2,1),trajectory_proton_3(1:N2,8),'blue', trajectory_proton_3(1:N2,1),trajectory_proton_3(1:N2,9),'black');
+title ('p');
+xlabel ('t');
+ylabel ('p');
+legend(4, 'px', 'py','pz', 'p');
+grid ;
+
+figure(3);
+plot (trajectory_proton_3(1:N2,4),trajectory_proton_3(1:N2,5));
+title ('r');
+xlabel ('y');
+ylabel ('z');
+grid ;
+
+figure(4);
+plot (trajectory_proton_3(1:N2,3),trajectory_proton_3(1:N2,9));
+title ('p(x)');
+xlabel ('x');
+ylabel ('p');
+grid ;
+
+figure(5);
+plot (trajectory_proton_3(1:N2,3),trajectory_proton_3(1:N2,6));
+title ('p_x(x)');
+xlabel ('x');
+ylabel ('p_x');
+grid ;
+

### 2017-10-02 PIC++/output/plot_traectory_proton_6.m

@@ -1,0 +1,43 @@
+clear;
+load trajectory_proton_6.dat;
+N1=1;
+N2=size(trajectory_proton_6,1);
+
+
+figure(1);
+plot (trajectory_proton_6(1:N2,1),trajectory_proton_6(1:N2,3),'red',trajectory_proton_6(1:N2,1),trajectory_proton_6(1:N2,4),'green',trajectory_proton_6(1:N2,1),trajectory_proton_6(1:N2,5),'blue');
+title ('r');
+xlabel ('t');
+ylabel (')');
+legend(4, 'x', 'y','z');
+grid ;
+
+figure(2);
+plot (trajectory_proton_6(1:N2,1),trajectory_proton_6(1:N2,6),'red',trajectory_proton_6(1:N2,1),trajectory_proton_6(1:N2,7),'green',trajectory_proton_6(1:N2,1),trajectory_proton_6(1:N2,8),'blue', trajectory_proton_6(1:N2,1),trajectory_proton_6(1:N2,9),'black');
+title ('p')
+xlabel ('t');
+ylabel ('p');
+legend(4, 'px', 'py','pz', 'p');
+grid ;
+
+figure(3);
+plot (trajectory_proton_6(1:N2,4),trajectory_proton_6(1:N2,5));
+title ('r');
+xlabel ('y');
+ylabel ('z');
+grid ;
+
+figure(4);
+plot (trajectory_proton_6(1:N2,3),trajectory_proton_6(1:N2,9));
+title ('p(x)');
+xlabel ('x');
+ylabel ('p');
+grid ;
+
+figure(5);
+plot (trajectory_proton_6(1:N2,3),trajectory_proton_6(1:N2,6));
+title ('p_x(x)');
+xlabel ('x');
+ylabel ('p_x');
+grid ;
+

### 2017-10-02 PIC++/output/plot_velocity_3d.m

@@ -1,0 +1,95 @@
+clear;
+load velocity.dat;
+load particleTypes.dat;
+load Xfile.dat;
+load Yfile.dat;
+load Zfile.dat;
+
+Nx = size(Xfile, 1)-1;
+Ny = size(Yfile, 1)-1;
+Nz = size(Zfile, 1)-1;
+
+N = Nx*Ny*Nz;
+
+Nt = size(velocity, 1)/N;
+Ntypes = size(particleTypes,1);
+
+znumber = 1;
+
+a = 0;
+b = fix(Nt/2);
+c = Nt - 1;
+
+particleVelocitiesX(1:Nx, 1:Ny, 1:Ntypes) = 0;
+particleVelocitiesY(1:Nx, 1:Ny, 1:Ntypes) = 0;
+particleVelocitiesZ(1:Nx, 1:Ny, 1:Ntypes) = 0;
+
+middleX(1:Nx) = 0;
+middleY(1:Ny) = 0;
+
+for i=1:Nx,
+   middleX(i) = 0.5*(Xfile(i) + Xfile(i+1));
+   for j = 1:Ny,
+     for t = 1:Ntypes,
+       particleVelocitiesX(i, j, t) = velocity(Nz*Ny*(i-1) + Nz*(j-1) + znumber + c*N, 1 + 3*(t-1));
+
+       particleVelocitiesY(i, j, t) = velocity(Nz*Ny*(i-1) + Nz*(j-1) + znumber + c*N, 2 + 3*(t-1));
+       
+       particleVelocitiesZ(i, j, t) = velocity(Nz*Ny*(i-1) + Nz*(j-1) + znumber + c*N, 3 + 3*(t-1));
+     end;
+   end;
+end;
+
+for j = 1:Ny,
+    middleY(j) = 0.5*(Yfile(j) + Yfile(j+1));
+end;
+
+vel(1:Nx,1:Ny) = 0;
+for t = 1:Ntypes,
+    if (particleTypes(t) > 0)
+        for i = 1:Nx,
+            for j = 1:Ny,
+                vel(i,j)=particleVelocitiesX(i, j, t);
+            end;
+        end;
+        figure(1 + 3*(t-1));
+        [X, Y] = meshgrid(middleY, middleX);
+        surf(X, Y, vel);
+        shading interp;
+        title ('velocity x');
+        xlabel ('y');
+        ylabel ('x');
+        zlabel ('v cm/s');
+        grid ;  
+    
+        for i = 1:Nx,
[…]

### 2017-10-02 PIC++/output/plot_velocity_y.m

@@ -1,6 +1,6 @@
 clear;
 load velocity.dat;
-load velocity_electron.dat;
+load particleTypes.dat;
 load Xfile.dat;
 load Yfile.dat;
 load Zfile.dat;
@@ -11,85 +11,58 @@ Nz = size(Zfile, 1)-1;
 
 N = Nx*Ny*Nz;
 Nt = size(velocity, 1)/N;
-xnumber = 1;
-ynumber = 1;
-znumber = 1;
+Ntypes = size(particleTypes,1);
+xnumber = 10;
+znumber = 2;
 
 a = 0;
 b = fix(Nt/2);
 c = Nt - 1;
 
-Vx(1:Ny, 1:3) = 0;
-Vy(1:Ny, 1:3) = 0;
-Vz(1:Ny, 1:3) = 0;
+particleVelocitiesX(1:Ny, 1:(3*Ntypes)) = 0;
+particleVelocitiesY(1:Ny, 1:(3*Ntypes)) = 0;
+particleVelocitiesZ(1:Ny, 1:(3*Ntypes)) = 0;
 
-Velectronx(1:Ny, 1:3) = 0;
-Velectrony(1:Ny, 1:3) = 0;
-Velectronz(1:Ny, 1:3) = 0;
-
-middleY(1:Ny-1) = 0;
+middleY(1:Ny) = 0;
 
 for i=1:Ny,
    middleY(i) = 0.5*(Yfile(i) + Yfile(i+1));
-   Vx(i,1) = velocity((Nz)*(Ny)*(xnumber-1) + (Nz)*(i-1) + znumber + a*N, 1);
-   Vx(i,2) = velocity((Nz)*(Ny)*(xnumber-1) + (Nz)*(i-1) + znumber + b*N, 1);
-   Vx(i,3) = velocity((Nz)*(Ny)*(xnumber-1) + (Nz)*(i-1) + znumber + c*N, 1);
-   Vy(i,1) = velocity((Nz)*(Ny)*(xnumber-1) + (Nz)*(i-1) + znumber + a*N, 2);
-   Vy(i,2) = velocity((Nz)*(Ny)*(xnumber-1) + (Nz)*(i-1) + znumber + b*N, 2);
-   Vy(i,3) = velocity((Nz)*(Ny)*(xnumber-1) + (Nz)*(i-1) + znumber + c*N, 2);
-   Vz(i,1) = velocity((Nz)*(Ny)*(xnumber-1) + (Nz)*(i-1) + znumber + a*N, 3);
[…]

### 2017-10-02 PIC++/simulation.h

@@ -3,15 +3,17 @@
 
 #include <string>
 #include "stdlib.h"
+#include "constants.h"
 #include "stdio.h"
 #include "vector"
 #include "list"
 #include "largeVectorBasis.h"
+#include "massMatrix.h"
 
 class MatrixElement;
 class Complex;
 
-enum SolverType {EXPLICIT, IMPLICIT};
+enum SolverType {EXPLICIT, IMPLICIT, IMPLICIT_EC, BUNEMAN};
 
 enum InputType {CGS, Theoretical};
 
@@ -27,19 +29,44 @@ class Simulation {
 private:
 	bool arrayCreated;
 public:
+	int rank;
+	int cartCoord[MPI_dim];
+	int cartDim[MPI_dim];
+	MPI_Comm cartComm;
+	int nprocs;
+	int leftRank;
+	int rightRank;
+	int frontRank;
+	int backRank;
+	int bottomRank;
+	int topRank;
 
 	std::string outputDir;
+	std::string inputDir;
+	std::string reducedOutputDir;
 	int xnumberGeneral;
 	int ynumberGeneral;
 	int znumberGeneral;
 	int firstAbsoluteXindex;
+	int firstAbsoluteYindex;
+	int firstAbsoluteZindex;
 	double leftX;
 	double rightX;
+	double leftY;
+	double rightY;
+	double leftZ;
+	double rightZ;
 
 	int xnumber;
 	int ynumber;
 	int znumber;
+	int xnumberAdded;
+	int ynumberAdded;
+	int znumberAdded;
 	int particlesNumber;
+
+	int resistiveLayerWidth;
+	double fakeCondactivity;
 
 	double massProton;
 	double massElectron;
@@ -83,6 +110,8 @@ public:
 	double deltaY2;
 	double deltaZ2;
 
+	double cellVolume;
+
 	double theta;
 	double eta;
 
@@ -106,12 +135,20 @@ public:
 
 	double particleEnergy;
 	double electricFieldEnergy;
+	double electricFieldEnergyX;
+	double electricFieldEnergyY;
+	double electricFieldEnergyZ;
 	double magneticFieldEnergy;
+	double magneticFieldEnergyX;
+	double magneticFieldEnergyY;
+	double magneticFieldEnergyZ;
 	double energy;
 
 	double theoreticalEnergy;
 	double generalTheoreticalEnergy;
 	Vector3d theoreticalMomentum;
[…]

### 2017-10-02 PIC++/specialmath.cpp

@@ -1,5 +1,6 @@
 #include <stdio.h>
 #include <math.h>
+#include <mpi.h>
 //#include <crtdbg.h>
 
 //#include "memory_debug.h"
@@ -11,24 +12,7 @@
 #include "matrixElement.h"
 #include "random.h"
 #include "specialmath.h"
-
-void updatePeriodicBoundaries(double**** tempVector, int xnumber, int ynumber, int znumber,
-                              int lnumber) {
-	for (int j = 0; j < ynumber; ++j) {
-		for (int k = 0; k < znumber; ++k) {
-			for (int l = 0; l < lnumber; ++l) {
-				tempVector[xnumber][j][k][l] = tempVector[1][j][k][l];
-			}
-		}
-	}
-	for (int j = 0; j < ynumber; ++j) {
-		for (int k = 0; k < znumber; ++k) {
-			for (int l = 0; l < lnumber; ++l) {
-				tempVector[0][j][k][l] = tempVector[xnumber - 1][j][k][l];
-			}
-		}
-	}
-}
+#include "mpi_util.h"
 
 double evaluateError(double** hessenbergMatrix, double* vector, double beta, int n) {
 	double* resVector = new double[n + 1];
@@ -53,25 +37,48 @@ double evaluateError(double** hessenbergMatrix, double* vector, double beta, int
 	return sqrt(norm);
 }
 
-double**** multiplySpecialMatrixVector(std::vector<MatrixElement>**** matrix, Vector3d*** vector, int xnumber, int ynumber, int znumber, int lnumber, bool periodic) {
-	double**** result = new double***[xnumber + 1];
-	int i = 0;
+double**** multiplySpecialMatrixVector(std::vector<MatrixElement>**** matrix, Vector3d*** vector, int xnumberAdded, int ynumberAdded, int znumberAdded, int additionalBinNumber, int lnumber, bool periodic, int rank, int nprocs, MPI_Comm& cartComm, int* cartCoord, int* cartDim) {
+	double**** result = new double***[xnumberAdded];
 
-	for (i = 0; i < xnumber + 1; ++i) {
+	int minI = 1 + additionalBinNumber;
+	if (cartCoord[0] == 0 && !periodic) {
+		minI = 0;
+	}
+	int maxI = xnumberAdded - 2 - additionalBinNumber;
+	if (cartCoord[0] == cartDim[0] - 1 && !periodic) {
+		maxI = xnumberAdded - 1;
+	}
[…]

### 2017-10-02 PIC++/vector3d.cpp

@@ -1,3 +1,4 @@
+#include <mpi.h>
 #include "math.h"
 #include "stdlib.h"
 #include "stdio.h"
@@ -97,6 +98,21 @@ Matrix3d Vector3d::tensorMult(const Vector3d& vector) const {
 	return result;
 }
 
+Matrix3d Vector3d::selfTensorMult() const {
+	Matrix3d result;
+	result.matrix[0][0] = x * x;
+	result.matrix[0][1] = x * y;
+	result.matrix[0][2] = x * z;
+	result.matrix[1][0] = result.matrix[0][1];
+	result.matrix[1][1] = y * y;
+	result.matrix[1][2] = y * z;
+	result.matrix[2][0] = result.matrix[0][2];
+	result.matrix[2][1] = result.matrix[1][2];
+	result.matrix[2][2] = z * z;
+
+	return result;
+}
+
 double& Vector3d::operator[](int i){
 	switch (i) {
 	case 0:
@@ -107,6 +123,7 @@ double& Vector3d::operator[](int i){
 		return z;
 	default:
 		printf("i must be 0 < i < 3\n");
+			MPI_Finalize();
 		exit(0);
 	}
 }

### 2017-10-02 PIC++1d/simulation.cpp

@@ -16,10 +16,10 @@ void Simulation::simulate() {
 		createArrays();
 		createFiles();
 		initialize();
-		//initializeTwoStream();
+		initializeTwoStream();
 		//initializeExternalFluxInstability();
 		//initializeAlfvenWave(1, 0.01);
-		initializeFluxFromRight();
+		//initializeFluxFromRight();
 		//initializeShockWave();
 		//initializeSimpleElectroMagneticWave();
 		//initializeLangmuirWave();

### 2017-10-02 test/specialmath.cpp

@@ -1,11 +1,15 @@
 #include <stdio.h>
 #include <stdlib.h>
 #include <math.h>
+#include <mpi.h>
 
 #include "specialmath.h"
 #include "complex.h"
+#include "mpi_util.h"
 
-double** arnoldiIterations(double** matrix, double** outHessenbergMatrix, int n, double** prevBasis, double** prevHessenbergMatrix){
+
+double** arnoldiIterations(double** matrix, double** outHessenbergMatrix, int n, double** prevBasis, double** prevHessenbergMatrix, int number, bool periodic, int rank, int nprocs){
+	MPI_Barrier(MPI_COMM_WORLD);
 	double** resultBasis = new double*[n];
 	for (int m = 0; m < n - 1; ++m) {
 		resultBasis[m] = prevBasis[m];
@@ -25,39 +29,85 @@ double** arnoldiIterations(double** matrix, double** outHessenbergMatrix, int n,
 		}
 	}
 	delete[] prevHessenbergMatrix;
+	//printf("update hessenberg\n");
+	double* tempVector = multiplyMatrixVector(matrix, resultBasis[n - 2], number, periodic, rank, nprocs);
+	//printf("mult special matrix");
+	double* bufferRightSend = new double[1];
+	double* bufferRightRecv = new double[1];
+	double* bufferLeftSend = new double[1];
+	double* bufferLeftRecv = new double[1];
 
-	double* tempVector = multiplyMatrixVector(matrix, resultBasis[n - 2]);
-	double b = sqrt(scalarMultiplyLargeVectors(tempVector, tempVector));
+	//printf("start exchange\n");
+
+	MPI_Barrier(MPI_COMM_WORLD);
+
+	if( (periodic) || (rank > 0)) sendLargeVectorToLeft(tempVector, bufferLeftSend, number);
+	if( (periodic) || (rank < nprocs-1)) sendLargeVectorToRight(tempVector, bufferRightSend, number);
+
+	MPI_Barrier(MPI_COMM_WORLD);
+
+	//printf("finish sending\n");
+
+	if( (periodic) || (rank < nprocs-1)) receiveLargeVectorFromRight(tempVector, bufferRightRecv, number);
+	if( (periodic) || (rank > 0)) receiveLargeVectorFromLeft(tempVector, bufferLeftRecv, number);
+
+
[…]

### 2017-10-02 test/specialmath.h

@@ -3,14 +3,19 @@
 
 #include "complex.h"
 
-const int number = 7;
 
 const double pi = 3.1415926535897932384626433832795028841971693993751;
 
-double** arnoldiIterations(double** matrix, double** outHessenbergMatrix, int n, double** prevBasis, double** prevHessenbergMatrix);
-double* generalizedMinimalResidualMethod(double** matrix, double* rightPart);
-double scalarMultiplyLargeVectors(double* a, double* b);
-double* multiplyMatrixVector(double** matrix, double* vector);
+double** arnoldiIterations(double** matrix, double** outHessenbergMatrix, int n, double** prevBasis, double** prevHessenbergMatrix, int number, bool periodic, int rank, int nprocs);
+double* generalizedMinimalResidualMethod(double** matrix, double* rightPart, int number, bool periodic, int rank, int nprocs);
+double scalarMultiplyLargeVectors(double* a, double* b, int number, bool periodic, int rank, int nprocs);
+double* multiplyMatrixVector(double** matrix, double* vector, int number, bool periodic, int rank, int nprocs);
+void multiplyMatrixVector(double* result, double** matrix, double* vector, int number, bool periodic, int rank, int nprocs);
+
+void conjugateGradientMethod(double** matrix, double* rightPart, double* outVector, int number, double precision, int maxIterationr, bool periodic, int rank, int nprocs);
+void biconjugateGradientMethod(double** matrix, double* rightPart, double* outVector, int number, double precision, int maxIterationr, bool periodic, int rank, int nprocs);
+void biconjugateStabilizedGradientMethod(double** matrix, double* rightPart, double* outVector, int number, double precision, int maxIterationr, bool periodic, int rank, int nprocs);
+void transposeMatrix(double** result, double** matrix, int number);
 
 double solve4orderEquation(double a4, double a3, double a2, double a1, double a0, double startX);
 double polynomValue(double a4, double a3, double a2, double a1, double a0, double x);
@@ -18,5 +23,14 @@ double polynomDerivativeValue(double a4, double a3, double a2, double a1, double
 
 Complex*** evaluateFourierTranslation(double*** a, int xnumber, int ynumber, int znumber);
 double*** evaluateReverceFourierTranslation(Complex*** a, int xnumber, int ynumber, int znumber);
+
+Complex* evaluateFourierTranslation(double* a, int number);
[…]

### 2017-10-02 test/test.cpp

@@ -3,12 +3,201 @@
 
 #include <stdio.h>
 #include <stdlib.h>
+#include <time.h> 
 #include <math.h>
+#include <mpi.h>
 
 #include "specialmath.h"
 #include "complex.h"
 
+void testFastFourier() {
+	const int number = 2*2*2*2*2;
+	double* a = new double[number];
+	for(int i = 0; i < number; ++i) {
+		a[i] = cos(2*pi*i/number) + cos(16*pi*i/number);
+		//a[i] = rand()%number;
+		//a[i] = i;
+	}
+	//sortInputFastFourier(a, number);
+	time_t time0;
+	time_t time1;
+	time_t time2;
+
+	printf("evaluating fourier\n");
+	time(&time0);
+	Complex* result =evaluateFourierTranslation(a, number);
+	time(&time1);
+	time1 = time1 - time0;
+	printf("evaluating fast fourier\n");
+	time(&time0);
+	Complex* resultFast = fastFourierTransition(a, number);
+	time(&time2);
+	time2 = time2 - time0;
+
+	for(int i = 0; i < number; ++i) {
+		printf("%g %g             %g %g\n", result[i].re, result[i].im, resultFast[i].re, resultFast[i].im);
+	}
+
+	printf("time fourier = %lld\n", time1);
+	printf("time fast fourier = %lld\n", time2);
+
+	printf("evaluating reverce fourier\n");
+	time(&time0);
+	double* b = evaluateReverceFourierTranslation(result, number);
+	time(&time1);
+	time1 = time1 - time0;
+	printf("evaluating reverce fast fourier\n");
+	time(&time0);
+	double* c = fastFourierReverceTransition(resultFast, number);
+	time(&time2);
+	time2 = time2 - time0;
+
+	for(int i = 0; i < number; ++i) {
+		printf("%g %g %g\n", a[i], b[i], c[i]);
+	}
+
+	printf("time fourier = %lld\n", time1);
+	printf("time fast fourier = %lld\n", time2);
+
+	delete[] result;
+	delete[] resultFast;
+
+	delete[] a;
+	delete[] b;
+	delete[] c;
+}
+
 void testGMRES(){
+	int rank;
+	int nprocs;
+	MPI_Comm_size(MPI_COMM_WORLD, &nprocs);
+	MPI_Comm_rank(MPI_COMM_WORLD, &rank);
[…]

### 2017-10-02 testMPI/complex.cpp

@@ -1,0 +1,84 @@
+#include "math.h"
+#include "stdlib.h"
+#include "stdio.h"
+
+#include "complex.h"
+
+Complex::Complex(){
+	re = 0;
+	im = 0;
+}
+
+Complex::Complex(double a, double b){
+	re = a;
+	im = b;
+}
+
+Complex::Complex(const Complex& a){
+	re = a.re;
+	im = a.im;
+}
+
+Complex& Complex::operator=(const Complex& a) {
+	re = a.re;
+	im = a.im;
+
+	return *this;
+}
+
+double Complex::module(){
+	return sqrt(re*re + im*im);
+}
+
+double Complex::phase(){
+	return atan2(im, re);
+}
+
+Complex Complex::conjugate(){
+	return Complex(re, -im);
+}
+
+Complex Complex::operator-(const Complex& a){
+	return Complex(re - a.re, im - a.im);
+}
+
+Complex Complex::operator+(const Complex& a){
+	return Complex(re + a.re, im + a.im);
+}
+
+Complex& Complex::operator+=(const Complex& a){
+	re += a.re;
+	im += a.im;
+	return *this;
+}
+
+Complex& Complex::operator-=(const Complex& a){
+	re -= a.re;
+	im -= a.im;
+	return *this;
+}
+
+Complex Complex::operator*(const double& a){
+	return Complex(re*a, im*a);
+}
+
+Complex Complex::operator/(const double& a){
+	return Complex(re/a, im/a);
+}
+
+Complex Complex::operator*(const Complex& a){
+	double newRe = re*a.re - im*a.im;
+	double newIm = re*a.im + im*a.re;
+	return Complex(newRe, newIm);
+}
+
+Complex Complex::operator/(const Complex& a){
+	double mod2 = a.re*a.re + a.im*a.im;
+	double newRe = (re*a.re + im*a.im)/mod2;
+	double newIm = (im*a.re - re*a.im)/mod2;
+	return Complex(newRe, newIm);
+}
+
+Complex complexExp(double phase){
+	return Complex(cos(phase), sin(phase));
+}

### 2017-10-02 testMPI/mpi_util.cpp

@@ -1,0 +1,99 @@
+//
+// Created by vadim on 19.06.16.
+//
+#include <mpi.h>
+#include "mpi_util.h"
+
+
+
+void sendGMRESTempVectorToRight(double *tempVector, double *buffer, int number) {
+
+    buffer[0] = tempVector[number - 2];
+    int rank;
+    int size;
+    MPI_Comm_size(MPI_COMM_WORLD, &size);
+    MPI_Comm_rank(MPI_COMM_WORLD, &rank);
+    int rightRank = rank+1;
+    if(rightRank >= size){
+        rightRank = 0;
+    }
+    if(rightRank != rank) {
+        printf("sent from %d to %d\n", rank, rightRank);
+        MPI_Send(buffer, 1, MPI_DOUBLE, rightRank, 0, MPI_COMM_WORLD);
+    }
+
+}
+
+void sendGMRESTempVectorToLeft(double *tempVector, double *buffer, int number) {
+    buffer[0] = tempVector[1];
+
+    int rank;
+    int size;
+    MPI_Comm_size(MPI_COMM_WORLD, &size);
+    MPI_Comm_rank(MPI_COMM_WORLD, &rank);
+    int leftRank = rank-1;
+    if(leftRank < 0){
+        leftRank = size-1;
+    }
+
+    if(leftRank != rank)
+        printf("sent from %d to %d\n", rank, leftRank);
+        MPI_Send(buffer, 1, MPI_DOUBLE, leftRank, 0, MPI_COMM_WORLD);
+
+
+}
+
+void receiveGMRESTempVectorFromRight(double *tempVector, double* buffer, int number) {
+    int rank;
+    int size;
+    MPI_Comm_size(MPI_COMM_WORLD, &size);
+    MPI_Comm_rank(MPI_COMM_WORLD, &rank);
+    int rightRank = rank+1;
+    if(rightRank >= size){
+        rightRank = 0;
+    }
+    if(rightRank != rank) {
+        MPI_Status status;
+        printf("recieving from %d to %d\n", rightRank, rank);
+        MPI_Recv(buffer, 1, MPI_DOUBLE, rightRank, 0, MPI_COMM_WORLD,
+                 &status);
+        printf("recieved from %d to %d\n", rightRank, rank);
+
+        tempVector[number-1] = buffer[0];
+    } else {
+        tempVector[number-1] = tempVector[1];
+    }
+}
+
[…]

### 2017-10-02 testMPI/mpi_util.h

@@ -1,0 +1,23 @@
+//
+// Created by vadim on 19.06.16.
+//
+
+#ifndef PIC_MPI_UTIL_H
+#define PIC_MPI_UTIL_H
+
+#include <vector>
+
+enum MPI_TAGS{MPI_INPUT_INTEGER_TAG = 1, MPI_INPUT_DOUBLE_TAG = 2, MPI_EFIELDX_RIGHT = 3, MPI_EFIELDY_RIGHT = 4,
+    MPI_EFILDZ_RIGHT = 5, MPI_EFIELDX_LEFT = 6, MPI_EFIELDY_LEFT = 7, MPI_EFILDZ_LEFT = 8,
+    MPI_TEMPVECTOR_RIGHT = 9, MPI_TEMPVECTOR_LEFT = 10, MPI_SEND_DOUBLE_NUMBER_ALL = 11, MPI_SEND_GENERAL_PARAMETERS = 12,
+    MPI_CELL_PARAMETERS_LEFT = 14, MPI_CELL_PARAMETERS_RIGHT = 15, MPI_NODE_PARAMETERS_LEFT = 16, MPI_NODE_PARAMETERS_RIGHT = 17,
+    MPI_SEND_INTEGER_NUMBER_LEFT = 18, MPI_SEND_INTEGER_NUMBER_RIGHT = 19, MPI_SEND_DOUBLE_NUMBER_LEFT = 20, MPI_SEND_DOUBLE_NUMBER_RIGHT = 21};
+
+
+void sendGMRESTempVectorToRight(double *tempVector, double *buffer, int number);
+void sendGMRESTempVectorToLeft(double *tempVector, double *buffer, int number);
+void receiveGMRESTempVectorFromRight(double *tempVector, double *buffer, int number);
+void receiveGMRESTempVectorFromLeft(double *tempVector, double *buffer, int number);
+
+
+#endif //PIC_MPI_UTIL_H

### 2017-10-02 testMPI/specialmath.cpp

@@ -1,0 +1,448 @@
+#include <stdio.h>
+#include <stdlib.h>
+#include <math.h>
+#include <mpi.h>
+
+#include "specialmath.h"
+#include "complex.h"
+#include "mpi_util.h"
+
+double** arnoldiIterations(double** matrix, double** outHessenbergMatrix, int n, double** prevBasis, double** prevHessenbergMatrix, int number){
+	int rank;
+	int size;
+	MPI_Comm_size(MPI_COMM_WORLD, &size);
+	MPI_Comm_rank(MPI_COMM_WORLD, &rank);
+
+	MPI_Barrier(MPI_COMM_WORLD);
+
+	double** resultBasis = new double*[n];
+	for (int m = 0; m < n - 1; ++m) {
+		resultBasis[m] = prevBasis[m];
+	}
+	delete[] prevBasis;
+
+	for (int i = 0; i < n; ++i) {
+		if (i < n - 1) {
+			for (int j = 0; j < n - 2; ++j) {
+				outHessenbergMatrix[i][j] = prevHessenbergMatrix[i][j];
+			}
+			delete[] prevHessenbergMatrix[i];
+		} else {
+			for (int j = 0; j < n - 2; ++j) {
+				outHessenbergMatrix[i][j] = 0;
+			}
+		}
+	}
+	delete[] prevHessenbergMatrix;
+
+	double* tempVector = multiplyMatrixVector(matrix, resultBasis[n - 2], number);
+
+
+
+	double bufferRightSend[1];
+	double bufferRightRecv[1];
+	double bufferLeftSend[1];
+	double bufferLeftRecv[1];
+
+	//printf("start exchange\n");
+
+
+	if(rank > 0) sendGMRESTempVectorToLeft(tempVector, bufferLeftSend, number);
+	if(rank < size-1) sendGMRESTempVectorToRight(tempVector, bufferRightSend, number);
+
+
+	//printf("finish sending\n");
+	MPI_Barrier(MPI_COMM_WORLD);
+
+	if(rank < size -1 )receiveGMRESTempVectorFromRight(tempVector, bufferRightRecv, number);
+	if(rank > 0) receiveGMRESTempVectorFromLeft(tempVector, bufferLeftRecv, number);
+
+
+
+	MPI_Barrier(MPI_COMM_WORLD);
+
+
+	for (int m = 0; m < n - 1; ++m) {
[…]

### 2017-10-02 testMPI/test.cpp

@@ -1,0 +1,141 @@
+// test.cpp : Defines the entry point for the console application.
+//
+
+#include <stdio.h>
+#include <stdlib.h>
+#include <time.h> 
+#include <math.h>
+#include <mpi.h>
+
+#include "specialmath.h"
+
+
+
+void testGMRES(){
+	const int generalNumber = 8;
+	double** generalMatrix = new double*[generalNumber];
+	double* generalRightPart = new double[generalNumber];
+
+	///only 3-diagonal generalMatrix!!!
+
+	for(int i = 0; i < generalNumber; ++i){
+		generalMatrix[i] = new double[generalNumber];
+		for(int j = 0; j < generalNumber; ++j){
+			if(i == j){
+				generalMatrix[i][j] = 1;
+			} else {
+				generalMatrix[i][j] = 0;
+			}
+		}
+	}
+
+    generalMatrix[1][2] = 0.2;
+    generalMatrix[2][3] = 0.1;
+    generalMatrix[3][1] = 0.3;
+    generalMatrix[3][4] = 0.1;
+    generalMatrix[4][5] = -0.5;
+    generalMatrix[5][6] = 0.1;
+    generalMatrix[6][5] = 0.1;
+
+	int rank;
+	int nprocs;
+	MPI_Comm_size(MPI_COMM_WORLD, &nprocs);
+	MPI_Comm_rank(MPI_COMM_WORLD, &rank);
+
+	const int localNumber = ((generalNumber-2)/nprocs) + 2;
+
+
+	for(int i = 0; i < generalNumber; ++i){
+		generalRightPart[i] = 0;
+	}
+    generalRightPart[1] = 1;
+    generalRightPart[2] = 1;
+    generalRightPart[5] = 1;
+    generalRightPart[6] = 1;
+	MPI_Barrier(MPI_COMM_WORLD);
+
+	if(rank == 0) {
+		for (int i = 0; i < generalNumber; ++i) {
+			for (int j = 0; j < generalNumber; ++j) {
+				printf("%lf     ", generalMatrix[i][j]);
+			}
+			printf("   %lf\n", generalRightPart[i]);
+		}
+	}
+	MPI_Barrier(MPI_COMM_WORLD);
+
+	double** localMatrix = new double*[localNumber];
+	double* localRightPart = new double[localNumber];
+
+	for(int i = 0; i < localNumber; ++i){
[…]

### 2017-10-31 PIC++/fourier.h
left fourier

@@ -19,6 +19,13 @@ Complex fourierTranslationZoneHarmonicRight(Complex*** input, bool direct, int z
 
 void fourierTranslationRight(Complex*** input, Complex*** output, Complex*** tempResultX, Complex*** tempResultXY, bool direct, int xnumber, int ynumber, int znumber, int xnumberGeneral, int ynumberGeneral, int znumberGeneral, int startAbsoluteIndex, MPI_Comm& cartComm, MPI_Comm& cartCommX, MPI_Comm& cartCommY, MPI_Comm& cartCommZ, int* xabsoluteIndex, int* yabsoluteIndex, int* zabsoluteIndex, Complex* localFactorX, Complex* localFactorY, Complex* localFactorZ, int* cartCoord, int* cartDim);
 
+Complex fourierTranslationXoneHarmonicLeft(Complex*** input, bool direct, int xnumber, int j, int k, int xnumberGeneral, MPI_Comm& reducedCartComm, int knumber, int* xabsoluteIndex, int startAbsoluteIndex, Complex* localFactor, int* cartCoord, int* cartDim);
+Complex fourierTranslationYoneHarmonicLeft(Complex*** input, bool direct, int ynumber, int i, int k, int ynumberGeneral, MPI_Comm& reducedCartComm, int knumber, int* yabsoluteIndex, int startAbsoluteIndex, Complex* localFactor, int* cartCoord, int* cartDim);
+Complex fourierTranslationZoneHarmonicLeft(Complex*** input, bool direct, int znumber, int i, int j, int znumberGeneral, MPI_Comm& reducedCartComm, int knumber, int* zabsoluteIndex, int startAbsoluteIndex, Complex* localFactor, int* cartCoord, int* cartDim);
+
+
+void fourierTranslationLeft(Complex*** input, Complex*** output, Complex*** tempResultX, Complex*** tempResultXY, bool direct, int xnumber, int ynumber, int znumber, int xnumberGeneral, int ynumberGeneral, int znumberGeneral, int startAbsoluteIndex, MPI_Comm& cartComm, MPI_Comm& cartCommX, MPI_Comm& cartCommY, MPI_Comm& cartCommZ, int* xabsoluteIndex, int* yabsoluteIndex, int* zabsoluteIndex, Complex* localFactorX, Complex* localFactorY, Complex* localFactorZ, int* cartCoord, int* cartDim);
+
[…]

### 2017-10-31 PIC++/initialization.cpp

@@ -227,10 +227,10 @@ Simulation::Simulation(int xn, int yn, int zn, double xsizev, double ysizev, dou
 	debugMode = false;
 	newlyStarted = true;
 	preserveChargeGlobal = true;
-	//solverType = IMPLICIT; //не явный
+	solverType = IMPLICIT; //не явный
 	//solverType = IMPLICIT_EC; //не явный с сохранением энергии
 	//solverType = EXPLICIT; //явный
-	solverType = BUNEMAN;
+	//solverType = BUNEMAN;
 	boundaryConditionType = PERIODIC;
 	//boundaryConditionType = SUPER_CONDUCTOR_LEFT;
 	maxwellEquationMatrixSize = 3;

### 2017-10-31 PIC++/simulation.cpp

@@ -28,7 +28,7 @@ void Simulation::simulate() {
 		createFiles();
 		//initializeTwoStream();
 		//initializeExternalFluxInstability();
-		initializeAlfvenWaveX(1, 0.01);
+		//initializeAlfvenWaveX(1, 0.01);
 		//initializeAlfvenWaveY(1, 0.01);
 		//initializeAlfvenWaveZ(1, 0.01);
 		//initializeRotatedAlfvenWave(1, 1, 0, 0.01);
@@ -36,7 +36,7 @@ void Simulation::simulate() {
 		//initializeAnisotropicSilicon();
 		//initializeWeibel();
 		//initializeRingWeibel();
-		//initializeFluxFromRight();
+		initializeFluxFromRight();
 		//initializeBell();
 		//initializeSimpleElectroMagneticWave();
 		//initializeSimpleElectroMagneticWaveY();
@@ -272,7 +272,7 @@ void Simulation::simulate() {
 				//cleanupDivergenceBuneman();
 				//cleanupDivergenceBunemanMagnetic();
 			} else {
-				//cleanupDivergence(newEfield, chargeDensity);
+				cleanupDivergence(newEfield, chargeDensity);
 				//cleanupDivergenceMagnetic();
 			}
 		}
@@ -283,7 +283,7 @@ void Simulation::simulate() {
 				//updateMaxEderivativePoint();
 			}
 			//if(currentIteration > 150){
-			//filterFields(5);
+			filterFields(5);
 			//}
 			//filterFieldsLocal(5);
 		}

### 2017-10-31 PIC++/simulation.h

@@ -932,6 +932,7 @@ public:
 	void filterFields(int cutWaveNumber);
 	void filterFieldGeneral(Vector3d*** field, int cutWaveNumber);
 	void filterFieldGeneralRight(Vector3d*** field, int cutWaveNumber, int startIndex);
+	void filterFieldGeneralLeft(Vector3d*** field, int cutWaveNumber, int startIndex);
 	void filterFieldGeneralRightMirror(Vector3d*** field, int cutWaveNumber, int startIndex);
 
 	void filterFieldsLocal(int cutWaveNumber);

### 2017-11-10 papers/Physica2017/Romansky_Physica2017.tex
left fourier
abstract

--- 2017-10-02 papers/Physica2016/Romansky_Physica2016.tex
@@
 \begin{document}
-\title{Particle-in-cell simulation of shock waves with different kinds of ions.}
+\title{Particle-in-cell simulation of ion and electron acceleration in trans-relativistic shocks.}
 
-\author{V I Romanky$^{1,2}$, A M Bykov$^1$, S M Osipov$^1$ and P E Gladilin$^1$}
+\author{V I Romansky$^{1,2}$, A M Bykov$^1$, S M Osipov$^1$ and P E Gladilin$^1$}
 
@@
 waves is a particle-in-cell computer simulation. In this paper we present results of shock
-wave simulations based on our self-developed implicit particle-in-cell code. First, we test the code
-with the well-known two stream and Weibel instabilities and second, simulate shock waves in
-plasma with different particle composition. Using the obtained results we study the
-process of charged particle acceleration at the shock front.
+wave simulations based on our self-developed implicit particle-in-cell code. 
 \end{abstract}

### 2017-11-10 PIC++/constants.h

@@ -41,7 +41,7 @@ const double initialTheta =  0.5;
 const double smoothingParameter = 0.001;
 const double particleSplitLevel = 3;
 const double particleSplitAngle = 0.02;
-const double speed_of_light_correction = 1;
+const double speed_of_light_correction = 1.0;
 const double speed_of_light_correction_sqr = speed_of_light_correction*speed_of_light_correction;
 const int splitParticlesParameter = 100;
 const int pnumber = 500;

### 2017-11-10 PIC++/mover.cpp

@@ -34,8 +34,8 @@ void Simulation::moveParticles() {
 				printf("move particle number %d rank %d\n", i, rank);
 			}
 		}
-		//moveParticle(particles[i]);
-		moveParticle(particles[i], 0, 10);
+		moveParticle(particles[i]);
+		//moveParticle(particles[i], 0, 10);
 		//particles[i]->momentum.x = 0;
 	}
 
@@ -220,7 +220,7 @@ void Simulation::moveParticle(Particle* particle) {
 		return;
 	}
 	//
-	tempParticle.addMomentum((E + (velocity.vectorMult(B) / speed_of_light_normalized)) * particle->charge * deltaT);
+	tempParticle.addMomentum((E + (velocity.vectorMult(B) / (speed_of_light_normalized*speed_of_light_correction))) * particle->charge * deltaT);
 	alertNaNOrInfinity(E.x, "E.x = Nan in move particle\n");
 
 	newVelocity = tempParticle.getVelocity(speed_of_light_normalized);
@@ -316,7 +316,7 @@ void Simulation::moveParticle(Particle* particle) {
 		//E = E0;
 		//B = B0;
 
-		tempParticle.addMomentum((E + (middleVelocity.vectorMult(B) / speed_of_light_normalized)) * (particle->charge * deltaT));
+		tempParticle.addMomentum((E + (middleVelocity.vectorMult(B) / (speed_of_light_normalized*speed_of_light_correction))) * (particle->charge * deltaT));
 		newVelocity = tempParticle.getVelocity(speed_of_light_normalized);
 		newMomentum = tempParticle.getMomentum();
 		//error = (prevVelocity - newVelocity).norm();
@@ -691,16 +691,16 @@ Matrix3d Simulation::evaluateAlphaRotationTensor(double beta, Vector3d& velocity
 
 	double G = ((beta * (EField.scalarMult(velocity)) / speed_of_light_normalized_sqr) + gamma);
 	beta = beta / G;
-	double beta2c = beta * beta / speed_of_light_normalized_sqr;
+	double beta2c = beta * beta / (speed_of_light_normalized_sqr * speed_of_light_correction_sqr);
 	double denominator = G * (1 + beta2c * BField.scalarMult(BField));
 
 	for (int i = 0; i < 3; i++) {
 		for (int j = 0; j < 3; j++) {
[…]

### 2017-11-10 PIC++/output/plot_fields_line_x.m

@@ -15,7 +15,7 @@ NE = Nx;
 NB = (Nx-1);
 Nt = (size(EfieldX, 1)/NE);
 %Nt=3;
-%Nt = 7;
+Nt = 6;
 NtB = (size(BfieldX, 1)/NB);
 ynumber = 2;
 znumber = 2;

### 2017-11-10 papers/Physica2017/Romansky_Physica2017.tex
fz

@@
 The problem of Cosmic Ray (CR) origin and acceleration is a widely discussed problem of
-modern astrophysics. It is considered that most likely CR sources are supernova remnant
-shock waves. One of the methods to study particle injection and acceleration by shock
-waves is a particle-in-cell computer simulation. In this paper we present results of shock
-wave simulations based on our self-developed implicit particle-in-cell code. 
+modern astrophysics. It is considered that most likely CR sources are 
+shock waves in supernova remnants or jets. One of the methods to study particle injection and acceleration by shock
+waves is a particle-in-cell computer simulation. In this paper we present results of
+simulations of relativistic shock waves with low lorenz-factor, based on our self-developed implicit particle-in-cell code. 
 \end{abstract}
 \input{intro.tex}
-\input{model.tex}
-\input{test.tex}
 \input{shock.tex}

### 2017-11-10 papers/Physica2017/conclusions.tex
fz

\section{Conclusions}
In this paper we presented the developed 3D implicit particle-in-cell code, which can be applied for various problems in the collisionless plasma. We have tested it and obtained results, that are consistent with theoretical predictions. We studied the influence of admixture of heavy ions on the evolution of the shock wave and simulations have shown that the presence of heavy ions increases the number of accelerated particles. Further research is necessary in order to study this phenomena in detail.
\ack
A M Bykov, P E  Gladilin and S M  Osipov have developed the implicit fields solver module, which provides conservation of the divergenceless magnetic field. 
V I Romansky acknowledges support from RSF grant 16-12-10519 which was used to develop the
particle mover, parallelization and perform the testing of the code.

The results of the work were obtained using computational resources of Peter the Great Saint-Petersburg Polytechnic University Supercomputing Center (http://www.spbstu.ru).

### 2017-11-10 papers/Physica2017/intro.tex
fz

--- 2017-10-02 papers/Physica2016/intro.tex
@@
 In this work we explore instabilities in the plasma using this method.
+We developed the implicit particle-in-cell (PIC) code, based on the scheme suggested by Lapenta et al.~\cite{Lapenta2006} and improved for the relativistic case by Noguchi et al.\cite{Noguchi2007}.
+Our code is fully three dimensional and parallelized with MPI technology, which is adapted for distributed computing and can be executed on a wide class of computers.

### 2017-11-10 papers/Physica2017/shock.tex
fz

\section{Shock wave simulation}
We studied the influence of fractions of heavy ions on the shock wave, in particular, the spectrum of particles and the shape of the shock wave. We present results of two simulations with different compositions of the plasma : pure protons and electrons in the first case and with the admixture of alpha particles in the second case. Initially the homogeneous flux is  moving from the right free boundary to the left. On the left side there is a reflecting super-conducting wall, which causes formation of shock wave. Simulations are one-dimensional and have following parameters: initial velocity $v = 0.2c$, number densities $n_e = 10^{-4} \rm{cm}^{-3}$, $n_p = 10^{-4} \rm{cm}^{-3}$ in the first simulation and $n_e = 10^{-4} \rm{cm}^{-3}$, $n_p = 0.6\cdot10^{-4} \rm{cm}^{-3}$, $n_\alpha = 0.2\cdot10^{-4} \rm{cm}^{-3}$ in the second, temperature $5\cdot10^8 \rm{K}$, magnetic field $B_\parallel = 10^{-4} \rm{G}, B_\perp= 0.7\cdot10^{-4} \rm{G}$, the full size of the box $L = 1\cdot10^{12} \rm{cm}$, the number of cells $N=2\cdot10^4$. Electron mass is reduced to $m_e = \frac{m_p}{20}$. The full time of simulation is $T = 2000 {\omega_p}^{-1}$.
\begin{figure}[h!]
	\centering
	\begin{minipage}{0.49\textwidth}
		\centering
		\includegraphics[width=0.98\textwidth]{fig/protons.eps} 
		\caption{Distribution of protons in the simulation without alpha particles.}
		\label{protons}
	\end{minipage}\hfill
	\begin{minipage}{0.49\textwidth}
		\centering
		\includegraphics[width=0.98\textwidth]{fig/protons_with_He.eps} 
		\caption{Distribution of protons in the simulation with alpha particles.}
		\label{protons_with_alpha}
	\end{minipage}
\end{figure}
\begin{figure}[h!]
	\centering
	\begin{minipage}{0.49\textwidth}
		\centering
		\includegraphics[width=0.98\textwidth]{fig/electrons.eps} 
		\caption{Distribution of electrons in the simulation without alpha particles.}
		\label{electrons}
	\end{minipage}\hfill
	\begin{minipage}{0.49\textwidth}
		\centering
		\includegraphics[width=0.98\textwidth]{fig/electrons_with_He.eps} 
		\caption{Distribution of electrons in the simulation with alpha particles.}
		\label{electrons_with_alpha}
	\end{minipage}
\end{figure}

The results presented in Figures \ref{protons}-\ref{electrons_with_alpha} show that the values of $F(p)p^4$ for high energies are much greater for protons, than for electrons in both simulations. Also electrons need much more time to be accelerated and to form spectrum $\propto p^{-4}$. It means that protons are injected into the acceleration process more efficiently, and it is consistent with the work of Park et al. {\cite{Park2015}}. Also we have shown that minority of heavy ions increases the spectrum of protons and do not have influence on the spectrum of electrons. It can be explained by the fact, that heavy ions form the large scale turbulence and protons can efficiently scatter on this turbulence. Otherwise for electrons turbulence produced by protons is already enough large-scale  and the influence of heavy ions is neglectable.

### 2017-11-11 papers/Physica2017/biblio.tex
paper

--- 2017-10-02 papers/Physica2016/biblio.tex
@@
 \bibitem{Bell1978} Bell A R 1978 \textit{MNRAS} \textbf{182} 147
-\bibitem{Blandford1978} Blandford R D and Ostriker J P 1978 \textit{Astrophys. J.} \textbf{221} L29 
+\bibitem{Blandford1978} Blandford R D and Ostriker J P 1978 \textit{ApJ} \textbf{221} L29 
+\bibitem{Bykov2014} Bykov A M, Ellison D C, Osipov S M and Vladimirov A E 2014 \textit{ApJ} \textbf{789} Issue 2, 137
 \bibitem{Berezhko2003} Berezhko E G, Ksenofontov L T and V{\"o}lk H J  2003 \textit{A}{\&}\textit{A} \textbf{412} L11
 \bibitem{Uchiyama2007} Uchiyama Y, Aharonian F A, Tanaka T, Takahashi T and Maeda Y 2007 \textit{Nature} \textbf{449} 576
+\bibitem{Romansky2016} Romansky V I, Bykov A M, Osipov S M and Gladilin P E \textit{Journal of Physics: CS}
 \bibitem{Lapenta2006} Lapenta G, Brackbill J U and Ricci P 2006 \textit{Phys. Plasmas} \textbf{13} 055904
 \bibitem{Noguchi2007} Noguchi K, Tronci C, Zuccaro G and Lapenta G 2007 \textit{Phys. Plasmas} \textbf{14} 042308
-\bibitem{Weibel1959} Weibel E S 1959 \textit{Phys. Rev. Lett.} \textbf{2} 83
-\bibitem{Yoon1987} Yoon P H and Davidson R C 1987 \textit{Phys. Rev. A} \textbf{35} 2718
-\bibitem{Park2015} Park J, Caprioli D and Spitkovsky A 2015 \textit{Phys. Rev. Lett.} \textbf{114} 085003
+\bibitem{Ellison2013} Ellison D C, Warren D C and Bykov A M 2013 \textit{ApJ} \textbf{776} Issue 1, 46
+\bibitem{Sironi2011} Sironi L and Spitkovsky A 2011 \textit{ApJ} \textbf{741} Issue 1, 39
 \end{thebibliography}

### 2017-11-11 papers/Physica2017/shock.tex
paper

@@
 \section{Shock wave simulation}
-We studied the influence of fractions of heavy ions on the shock wave, in particular, the spectrum of particles and the shape of the shock wave.
-We present results of two simulations with different compositions of the plasma : pure protons and electrons in the first case and with the admixture of alpha particles in the second case.
-Initially the homogeneous flux is  moving from the right free boundary to the left.
-On the left side there is a reflecting super-conducting wall, which causes formation of shock wave.
-Simulations are one-dimensional and have following parameters: initial velocity $v = 0.2c$, number densities $n_e = 10^{-4} \rm{cm}^{-3}$, $n_p = 10^{-4} \rm{cm}^{-3}$ in the first simulation and $n_e = 10^{-4} \rm{cm}^{-3}$, $n_p = 0.6\cdot10^{-4} \rm{cm}^{-3}$, $n_\alpha = 0.2\cdot10^{-4} \rm{cm}^{-3}$ in the second, temperature $5\cdot10^8 \rm{K}$, magnetic field $B_\parallel = 10^{-4} \rm{G}, B_\perp= 0.7\cdot10^{-4} \rm{G}$, the full size of the box $L = 1\cdot10^{12} \rm{cm}$, the number of cells $N=2\cdot10^4$.
-Electron mass is reduced to $m_e = \frac{m_p}{20}$.
-The full time of simulation is $T = 2000 {\omega_p}^{-1}$.
+We studied the evolution of trans-relativistic shocks and spectum of accelerated particles.
+We chose relastivistic shocks with low lorenz-factor because the maximum energy of produced cosmic rays increases with shock wave lorenz-factor, but the efficiency of acceleration decreases, because it is more difficult for particle to cross fast moving front many times \cite{Ellison2013}.
+So we assume that intermediate case of trans-relativistic shocks provides the most effective acceleration.
+
+We developed the implicit particle-in-cell (PIC) code, presented in our previous paper \cite{Romansky2016}, based on the scheme suggested by Lapenta et al.~\cite{Lapenta2006} and improved for the relativistic case by Noguchi et al.\cite{Noguchi2007}.
+Our code is fully three dimensional and parallelized with MPI technology, which is adapted for distributed computing and can be executed on a wide class of computers.
+
+In simulation setup the homogeneous plasma flows in through the right boundary
+and collides with the reflecting superconducting wall on the left boundary, causing the shock wave
+
 \begin{figure}[h!]
 	\centering
-	\begin{minipage}{0.49\textwidth}
-		\centering
-		\includegraphics[width=0.98\textwidth]{fig/protons.eps} 
-		\caption{Distribution of protons in the simulation without alpha particles.}
+	\includegraphics[width=0.86\textwidth]{fig/protons.eps} 
+	\caption{Distribution of protons.}
 	\label{protons}
-	\end{minipage}\hfill
-	\begin{minipage}{0.49\textwidth}
-		\centering
-		\includegraphics[width=0.98\textwidth]{fig/protons_with_He.eps} 
-		\caption{Distribution of protons in the simulation with alpha particles.}
-		\label{protons_with_alpha}
-	\end{minipage}
 \end{figure}
@@
 	\centering
-	\begin{minipage}{0.49\textwidth}
-		\centering
-		\includegraphics[width=0.98\textwidth]{fig/electrons.eps} 
-		\caption{Distribution of electrons in the simulation without alpha particles.}
+	\includegraphics[width=0.86\textwidth]{fig/electrons.eps} 
+	\caption{Distribution of electrons.}
 	\label{electrons}
-	\end{minipage}\hfill
-	\begin{minipage}{0.49\textwidth}
-		\centering
-		\includegraphics[width=0.98\textwidth]{fig/electrons_with_He.eps} 
-		\caption{Distribution of electrons in the simulation with alpha particles.}
-		\label{electrons_with_alpha}
-	\end{minipage}
 \end{figure}
 
-The results presented in Figures \ref{protons}-\ref{electrons_with_alpha} show that the values of $F(p)p^4$ for high energies are much greater for protons, than for electrons in both simulations.
-Also electrons need much more time to be accelerated and to form spectrum $\propto p^{-4}$.
-It means that protons are injected into the acceleration process more efficiently, and it is consistent with the work of Park et al.
-{\cite{Park2015}}.
-Also we have shown that minority of heavy ions increases the spectrum of protons and do not have influence on the spectrum of electrons.
-It can be explained by the fact, that heavy ions form the large scale turbulence and protons can efficiently scatter on this turbulence.
-Otherwise for electrons turbulence produced by protons is already enough large-scale  and the influence of heavy ions is neglectable.
+Simulations are one-dimensional and have following parameters: initial flow lorenz-factor $\gamma = 1.5$, number densities $n_e = 10^{-4} \rm{cm}^{-3}$, $n_p = 10^{-4} \rm{cm}^{-3}$ , temperature $5\cdot10^8 \rm{K}$, magnetic field $B = 10^{-4} \rm{G}$, the full size of the box $L = 2\cdot10^{12} \rm{cm}$, the number of cells $N=2\cdot10^4$.
+Electron mass is reduced to $m_e = \frac{m_p}{20}$.
+The full time of simulation is $T = 5000 {\omega_p}^{-1}$.
+$\theta$ is angle between flow velocity and magnetic field.
+We present results of particle spectrum in several simulations with different $\theta$.
+
+One can see from Figures \ref{protons} and \ref{electrons} that spectrum of accelerated particles strongly depends on angle theta.
+If $\theta$ is less then critical value, defined by equation $c\cdot cos(\theta_{crit})=v_{shock}$, where all values are measured in upstream rest frame, particles can escape from the front and cross it several times to gain more energy.
+It is difficult to evaluate critical angle apriori, because we don't know compression relation of shock wave which will be created in simulation, but estimation for case of strong wave (comression relation $\sigma=4^{\circ}$) gets $\theta_{crit}=45$ in downstream frame.
+And it is consistent with results of our simulations.

### 2017-11-11 PIC++/distributions/compare_two_distribution.m

@@ -1,0 +1,62 @@
+clear;
+load distribution_protons1.dat;
+load distribution_electrons1.dat;
+load distribution_protons2.dat;
+load distribution_electrons2.dat;
+
+Np = 500;
+
+Nt(1:2) = 11;
+
+Nt(1) = fix(size(distribution_protons1,1)/Np) - 1;
+Nt(2) = fix(size(distribution_protons2,1)/Np) - 1;
+
+
+Fp(1:Np, 1:2) = 0;
+Fe(1:Np, 1:2) = 0;
+
+
+Pp(1:Np, 1:2) = 0;
+Pe(1:Np, 1:2) = 0;
+
+me = 0.91*10^-25;
+mp = 1.6*10^-24;
+v=2.998*10^10;
+
+%Color = {'r','green','blue','black','yellow','cyan','magenta','purple','grey'};
+Color = {'red','blue'};
+
+for i=1:Np,
+   Pp(i,1) = distribution_protons1(i + Nt(1)*Np,1);
+   Pp(i,2) = distribution_protons2(i + Nt(2)*Np,1);
+   
+   Fp(i,1) = distribution_protons1(i + Nt(1)*Np, 2)*Pp(i,1)*Pp(i,1);
+   Fp(i,2) = distribution_protons2(i + Nt(2)*Np, 2)*Pp(i,2)*Pp(i,2);
+   
+   Pe(i,1) = distribution_electrons1(i + Nt(1)*Np,1);
+   Pe(i,2) = distribution_electrons2(i + Nt(2)*Np,1);
+   
+   Fe(i,1) = distribution_electrons1(i + Nt(1)*Np, 2)*Pe(i,1)*Pe(i,1);
+   Fe(i,2) = distribution_electrons2(i + Nt(2)*Np, 2)*Pe(i,2)*Pe(i,2);
+end;
+set(0,'DefaultAxesFontSize',14,'DefaultAxesFontName','Times New Roman');
+set(0,'DefaultTextFontSize',20,'DefaultTextFontName','Times New Roman'); 
+figure(1);
+hold on;
+for j=1:2,
+    plot (Pp(1:Np,j)/(mp*v),Fp(1:Np,j),'color',Color{j});
+end;
+xlabel ('p/{m_p c}');
+ylabel ('F_p(p) p^4');
+legend('{\theta} = 10','{\theta} = 20','Location','southeast');
+grid ;
+
+figure(2);
+hold on;
[…]

### 2017-11-11 PIC++/simulation.cpp

@@ -1195,7 +1195,7 @@ void Simulation::updateTheoreticalEnergy(){
 	//if (currentIteration > 0) {
 		//if (boundaryConditionType != PERIODIC) {
 
-			/*for (int i = 0; i < escapedParticlesLeft.size(); ++i) {
+			for (int i = 0; i < escapedParticlesLeft.size(); ++i) {
 				Particle* particle = escapedParticlesLeft[i];
 				theoreticalEnergy -= particle->energy(speed_of_light_normalized) * particle->weight *
 					sqr(scaleFactor / plasma_period);
@@ -1232,7 +1232,7 @@ void Simulation::updateTheoreticalEnergy(){
 				theoreticalEnergy -= particle->energy(speed_of_light_normalized) * particle->weight *
 					sqr(scaleFactor / plasma_period);
 				theoreticalMomentum -= particle->getMomentum() * particle->weight * scaleFactor / plasma_period;
-			}*/
+			}
 
 			for (int j = minJ; j < maxJ; ++j) {
 				for (int k = minK; k < maxK; ++k) {

### 2017-11-14 papers/Physica2017/intro.tex
paper

--- 2017-10-02 papers/Physica2016/intro.tex
@@
 One of the possible sources of high-energy cosmic rays are shock waves in astrophysical collisionless plasma due to the diffusive shock acceleration (DSA) mechanism, see \cite{Bell1978}, \cite{Blandford1978}.
-Strong magnetic fields are necessary for effective acceleration of charged particles and the observations of the synchrotron emission from supernova remnants (SNRs) show, that the magnetic field there is about 100 times stronger then the interstellar field (see for example \cite{Berezhko2003},\cite{Uchiyama2007}).
-So strong fields can be produced due to instabilities in anisotropic plasma.
+The particle acceleration on the shock front is sophisticated and nonlinear problem, because cosmic rays have strongly influence on the shock wave.
+The pressure of cosmic rays is comparable with pressure of background plasma and significatly modifies the front (see for example \cite{Bykov2014}).
+Also there are a instabilities in anisotropic plasma, which lead to strong field amplification and the observations of the synchrotron emission from supernova remnants (SNRs) show, that the magnetic field there is about 100 times stronger then the interstellar field (see for example \cite{Berezhko2003},\cite{Uchiyama2007}).
 One of the most powerful methods to explore processes in collisionless plasma is the Particle-in-Cell simulation.
-In this work we explore instabilities in the plasma using this method.

### 2017-11-14 papers/Physica2017/shock.tex
paper

@@
 
-We developed the implicit particle-in-cell (PIC) code, presented in our previous paper \cite{Romansky2016}, based on the scheme suggested by Lapenta et al.~\cite{Lapenta2006} and improved for the relativistic case by Noguchi et al.\cite{Noguchi2007}.
+We developed the implicit particle-in-cell (PIC) code, based on the scheme suggested by Lapenta et al.~\cite{Lapenta2006} and improved for the relativistic case by Noguchi et al.\cite{Noguchi2007}.
 Our code is fully three dimensional and parallelized with MPI technology, which is adapted for distributed computing and can be executed on a wide class of computers.
 
-In simulation setup the homogeneous plasma flows in through the right boundary
-and collides with the reflecting superconducting wall on the left boundary, causing the shock wave
+In simulation setup the homogeneous plasma, which flows in through the right boundary
+and collides with the reflecting superconducting wall on the left boundary.
 
+ Simulations are one-dimensional and have following parameters: initial flow lorenz-factor $\gamma = 1.5$, number densities $n_e = 10^{-4} \rm{cm}^{-3}$, $n_p = 10^{-4} \rm{cm}^{-3}$ , temperature $5\cdot10^8 \rm{K}$, magnetic field $B = 10^{-4} \rm{G}$, the full size of the box $L = 4\cdot10^{12} \rm{cm}$, the number of cells $N=4\cdot10^4$.
+Electron mass is reduced to $m_e = \frac{m_p}{20}$.
+The full time of simulation is $T = 5000 {\omega_p}^{-1}$.
+$\theta$ is angle between flow velocity and magnetic field.
+We present results of particle spectrum in several simulations with different $\theta$.
 \begin{figure}[h!]
 	\centering
-	\includegraphics[width=0.86\textwidth]{fig/protons.eps} 
+	\includegraphics[width=0.8\textwidth]{fig/protons.eps} 
 	\caption{Distribution of protons.}
@@
 	\centering
-	\includegraphics[width=0.86\textwidth]{fig/electrons.eps} 
+	\includegraphics[width=0.8\textwidth]{fig/electrons.eps} 
 	\caption{Distribution of electrons.}
@@
 
-Simulations are one-dimensional and have following parameters: initial flow lorenz-factor $\gamma = 1.5$, number densities $n_e = 10^{-4} \rm{cm}^{-3}$, $n_p = 10^{-4} \rm{cm}^{-3}$ , temperature $5\cdot10^8 \rm{K}$, magnetic field $B = 10^{-4} \rm{G}$, the full size of the box $L = 2\cdot10^{12} \rm{cm}$, the number of cells $N=2\cdot10^4$.
-Electron mass is reduced to $m_e = \frac{m_p}{20}$.
-The full time of simulation is $T = 5000 {\omega_p}^{-1}$.
-$\theta$ is angle between flow velocity and magnetic field.
-We present results of particle spectrum in several simulations with different $\theta$.
-
-One can see from Figures \ref{protons} and \ref{electrons} that spectrum of accelerated particles strongly depends on angle theta.
-If $\theta$ is less then critical value, defined by equation $c\cdot cos(\theta_{crit})=v_{shock}$, where all values are measured in upstream rest frame, particles can escape from the front and cross it several times to gain more energy.
-It is difficult to evaluate critical angle apriori, because we don't know compression relation of shock wave which will be created in simulation, but estimation for case of strong wave (comression relation $\sigma=4^{\circ}$) gets $\theta_{crit}=45$ in downstream frame.
-And it is consistent with results of our simulations.
+The results presented in Figures \ref{protons}-\ref{electrons_with_alpha} show that the values of $F(p)p^4$ for high energies are much greater for protons, than for electrons in both simulations.
+Also electrons need much more time to be accelerated and to form spectrum $\propto p^{-4}$.
+It means that protons are injected into the acceleration process more efficiently, and it is consistent with the work of Park et al.
+{\cite{Park2015}}.
+Also we have shown that minority of heavy ions increases the spectrum of protons and do not have influence on the spectrum of electrons.
+It can be explained by the fact, that heavy ions form the large scale turbulence and protons can efficiently scatter on this turbulence.
+Otherwise for electrons turbulence produced by protons is already enough large-scale  and the influence of heavy ions is neglectable.

### 2017-11-14 matlab/plot_density.m

@@ -1,79 +1,0 @@
-clear;
-load concentrations.dat;
-load Xfile.dat;
-load Yfile.dat;
-load Zfile.dat;
-load divergence_error.dat;
-
-Nx = size(Xfile, 1);
-Ny = size(Yfile, 1);
-Nz = size(Zfile, 1);
-
-N = Nx*Ny*Nz;
-Nt = size(concentrations, 1)/N;
-
-ynumber = 3;
-znumber = 3;
-
-a = 0;
-b = fix(Nt/2);
-c = Nt - 1;
-
-electronConcentration(1:Nx, 1:3) = 0;
-protonConcentration(1:Nx, 1:3) = 0;
-chargeDensity(1:Nx, 1:3) = 0;
-shiftChargeDensity(1:Nx, 1:3) = 0;
-divergenceError(1:Nx, 1:3) = 0;
-
-
-for i=1:Nx,   
-   electronConcentration(i, 1) = concentrations((Nz*Ny*(i-1) + Nz*(ynumber-1) + znumber) + a*N, 1);
-   electronConcentration(i, 2) = concentrations((Nz*Ny*(i-1) + Nz*(ynumber-1) + znumber) + b*N, 1);
-   electronConcentration(i, 3) = concentrations((Nz*Ny*(i-1) + Nz*(ynumber-1) + znumber) + c*N, 1);
-   protonConcentration(i, 1) = concentrations((Nz*Ny*(i-1) + Nz*(ynumber-1) + znumber) + a*N, 2);
-   protonConcentration(i, 2) = concentrations((Nz*Ny*(i-1) + Nz*(ynumber-1) + znumber) + b*N, 2);
-   protonConcentration(i, 3) = concentrations((Nz*Ny*(i-1) + Nz*(ynumber-1) + znumber) + c*N, 2);
-   chargeDensity(i, 1) = concentrations((Nz*Ny*(i-1) + Nz*(ynumber-1) + znumber) + a*N, 3);
-   chargeDensity(i, 2) = concentrations((Nz*Ny*(i-1) + Nz*(ynumber-1) + znumber) + b*N, 3);
-   chargeDensity(i, 3) = concentrations((Nz*Ny*(i-1) + Nz*(ynumber-1) + znumber) + c*N, 3);
-   shiftChargeDensity(i, 1) = concentrations((Nz*Ny*(i-1) + Nz*(ynumber-1) + znumber) + a*N, 4);
[…]

### 2017-11-14 matlab/plot_traectory_proton.m

@@ -1,28 +1,0 @@
-clear;
-load traectory_proton.dat;
-N1=1;
-N2=size(traectory_proton,1);
-
-
-figure(1);
-plot (traectory_proton(1:N2,1),traectory_proton(1:N2,2),'red',traectory_proton(1:N2,1),traectory_proton(1:N2,3),'green',traectory_proton(1:N2,1),traectory_proton(1:N2,4),'blue');
-title ('r');
-xlabel ('t');
-ylabel (')');
-legend(4, 'x', 'y','z');
-grid ;
-
-figure(2);
-plot (traectory_proton(1:N2,1),traectory_proton(1:N2,5),'red',traectory_proton(1:N2,1),traectory_proton(1:N2,6),'green',traectory_proton(1:N2,1),traectory_proton(1:N2,7),'blue', traectory_proton(1:N2,1),traectory_proton(1:N2,8),'black');
-title ('p');
-xlabel ('t');
-ylabel ('p');
-legend(4, 'px', 'py','pz', 'p');
-grid ;
-
-figure(3);
-plot (traectory_proton(1:N2,3),traectory_proton(1:N2,4));
-title ('r');
-xlabel ('y');
-ylabel ('z');
-grid ;

### 2017-11-15 papers/Physica2017/shock.tex
paper

@@
 The full time of simulation is $T = 5000 {\omega_p}^{-1}$.
+All values are measured in downstream frame.
 $\theta$ is angle between flow velocity and magnetic field.
@@
 
-The results presented in Figures \ref{protons}-\ref{electrons_with_alpha} show that the values of $F(p)p^4$ for high energies are much greater for protons, than for electrons in both simulations.
-Also electrons need much more time to be accelerated and to form spectrum $\propto p^{-4}$.
-It means that protons are injected into the acceleration process more efficiently, and it is consistent with the work of Park et al.
-{\cite{Park2015}}.
-Also we have shown that minority of heavy ions increases the spectrum of protons and do not have influence on the spectrum of electrons.
-It can be explained by the fact, that heavy ions form the large scale turbulence and protons can efficiently scatter on this turbulence.
-Otherwise for electrons turbulence produced by protons is already enough large-scale  and the influence of heavy ions is neglectable.
+One can see that spectrum strongly depends of angle $\theta$.
+If $\theta$ is less then critical value, defined as $c\cdot cos(\theta_{crit})=v_{shock}$, measured in shock rest frame, shock is subluminal and particles can escape from front and than cross it enough of times to e accelerated.

### 2017-11-15 PIC++/mover.cpp

@@ -208,7 +208,7 @@ void Simulation::moveParticle(Particle* particle) {
 	double gamma = particle->gammaFactor(speed_of_light_normalized);
 	double beta = 0.5 * particle->charge * deltaT / particle->mass;
 
-	int particleIterations = 50;
+	int particleIterations = 50*gamma;
 
 
 	Particle tempParticle = *particle;
@@ -336,13 +336,21 @@ void Simulation::moveParticle(Particle* particle) {
 
 	double deltaGammaTheor = particle->charge*deltaT*E.scalarMult(middleVelocity)/(particle->mass*speed_of_light_normalized_sqr);
 
+	/*if(i >= particleIterations){
+		printf("i >= particle iterations\n");
+		printf("error = %g\n", error);
+		printf("relative error = %g\n", error/momentumNorm);
+	}*/
+
 	if(fabs(newGamma - oldGamma) > 0.1){
 		printf("delta gamma > 0.1\n");
+		/*printf("particle iterations = %d\n", i);
 		printf("cartx = %d carty = %d cartz = %d\n", cartCoord[0], cartCoord[1], cartCoord[2]);
 		printf("oldGamma = %g newGamma = %g delta gamma = %g\n", oldGamma, newGamma, newGamma -oldGamma);
-		printf("theoretical delta gamm = %g\n", deltaGammaTheor);
+		printf("theoretical delta gamma = %g\n", deltaGammaTheor);
 		printf("particle number = %d\n", particle->number);
 		printf("x = %g y = %g z = %g\n", particle->coordinates.x, particle->coordinates.y, particle->coordinates.z);
+		printf("xindex = %d yindex = %d zindex = %d\n", particle->correlationMapCell.xindex[(2 + splineOrder)/2], particle->correlationMapCell.yindex[(2 + splineOrder)/2], particle->correlationMapCell.zindex[(2 + splineOrder)/2]);
 		printf("vx = %g vy = %g vz = %g\n", velocity.x, velocity.y, velocity.z);
 		printf("newvx = %g newvy = %g newvz = %g\n", newVelocity.x, newVelocity.y, newVelocity.z);
 		printf("Ex = %g Ey = %g Ez = %g\n", E.x, E.y, E.z);
@@ -351,7 +359,7 @@ void Simulation::moveParticle(Particle* particle) {
 			for(int j = 0; j < 3; ++j){
[…]

### 2017-11-17 papers/Physica2017/Romansky_Physica2017.tex
paper

@@
 
-\author{V I Romansky$^{1,2}$, A M Bykov$^1$, S M Osipov$^1$ and P E Gladilin$^1$}
+\author{V I Romansky$^{1,2}$, A M Bykov$^1$ and S M Osipov$^1$}
 
@@
 waves is a particle-in-cell computer simulation. In this paper we present results of
-simulations of relativistic shock waves with low lorenz-factor, based on our self-developed implicit particle-in-cell code. 
+simulations of relativistic shock waves with low lorentz-factor, based on our self-developed implicit particle-in-cell code. 
 \end{abstract}

### 2017-11-17 papers/Physica2017/conclusions.tex
paper

@@
 \section{Conclusions}
-In this paper we presented the developed 3D implicit particle-in-cell code, which can be applied for various problems in the collisionless plasma.
-We have tested it and obtained results, that are consistent with theoretical predictions.
-We studied the influence of admixture of heavy ions on the evolution of the shock wave and simulations have shown that the presence of heavy ions increases the number of accelerated particles.
+In this paper we studied trans-relativistic shock wave using our implicit particle-in-cell code.
+Simulations show, that particles injection in acceleration process highly depends on angle between flow bulk velocity and magnetic field.
+Also injection of protons is more effective than electrons for angles less then critical, but there is regime of more effective electrons acceleration when angle is close to critical.
 Further research is necessary in order to study this phenomena in detail.
 \ack
-A M Bykov, P E  Gladilin and S M  Osipov have developed the implicit fields solver module, which provides conservation of the divergenceless magnetic field. 
-V I Romansky acknowledges support from RSF grant 16-12-10519 which was used to develop the
-particle mover, parallelization and perform the testing of the code.
-
+V I Romansky acknowledges support from RSF grant 16-12-10519.
 The results of the work were obtained using computational resources of Peter the Great Saint-Petersburg Polytechnic University Supercomputing Center (http://www.spbstu.ru).

### 2017-11-17 papers/Physica2017/intro.tex
paper

@@
 One of the possible sources of high-energy cosmic rays are shock waves in astrophysical collisionless plasma due to the diffusive shock acceleration (DSA) mechanism, see \cite{Bell1978}, \cite{Blandford1978}.
-The particle acceleration on the shock front is sophisticated and nonlinear problem, because cosmic rays have strongly influence on the shock wave.
+The particle acceleration on the shock front is sophisticated and nonlinear problem, because cosmic rays have have strong influence on the shock wave.
 The pressure of cosmic rays is comparable with pressure of background plasma and significatly modifies the front (see for example \cite{Bykov2014}).
-Also there are a instabilities in anisotropic plasma, which lead to strong field amplification and the observations of the synchrotron emission from supernova remnants (SNRs) show, that the magnetic field there is about 100 times stronger then the interstellar field (see for example \cite{Berezhko2003},\cite{Uchiyama2007}).
+Also there are instabilities in anisotropic plasma, which lead to strong field amplification and the observations of the synchrotron emission from supernova remnants (SNRs) show, that the magnetic field there is about 100 times stronger then the interstellar field (see for example \cite{Berezhko2003},\cite{Uchiyama2007}).
 One of the most powerful methods to explore processes in collisionless plasma is the Particle-in-Cell simulation.

### 2017-11-17 papers/Physica2017/shock.tex
paper

--- 2017-11-11 papers/Physica2017/shock.tex
@@
 We studied the evolution of trans-relativistic shocks and spectum of accelerated particles.
-We chose relastivistic shocks with low lorenz-factor because the maximum energy of produced cosmic rays increases with shock wave lorenz-factor, but the efficiency of acceleration decreases, because it is more difficult for particle to cross fast moving front many times \cite{Ellison2013}.
+We chose relastivistic shocks with low lorentz-factor because the maximum energy of produced cosmic rays increases with shock wave lorentz-factor, but the efficiency of acceleration decreases, because it is more difficult for particle to cross fast moving front many times \cite{Ellison2013}.
 So we assume that intermediate case of trans-relativistic shocks provides the most effective acceleration.
@@
 In simulation setup the homogeneous plasma flows in through the right boundary
-and collides with the reflecting superconducting wall on the left boundary, causing the shock wave
+and collides with the reflecting superconducting wall on the left boundary, causing the shock wave.
+It is common way to initialize shock wave, because initialization using Rankine–Hugoniot conditions does not take in account microscopic distribution and shock wave, created that way, will probably fall apart for several discontinuities.
+
+  Simulations are one-dimensional and have following parameters: initial flow lorentz-factor $\gamma = 1.5$, number densities $n_e = 10^{-4} \rm{cm}^{-3}$, $n_p = 10^{-4} \rm{cm}^{-3}$ , temperature $5\cdot10^8 \rm{K}$, magnetic field $B = 10^{-4} \rm{G}$, the full size of the box $L = 2\cdot10^{12} \rm{cm}$, the number of cells $N=2\cdot10^4$.
+Electron mass is reduced to $m_e = \frac{m_p}{20}$.
+The full time of simulation is $T = 5000 {\omega_p}^{-1}$.
+$\theta$ is angle between flow velocity and magnetic field.
+We present results of particle spectrum in several simulations with different $\theta$.
 
@@
 	\centering
-	\includegraphics[width=0.86\textwidth]{fig/protons.eps} 
+	\includegraphics[width=0.95\textwidth]{fig/protons.eps} 
 	\caption{Distribution of protons.}
@@
 	\centering
-	\includegraphics[width=0.86\textwidth]{fig/electrons.eps} 
+	\includegraphics[width=0.95\textwidth]{fig/electrons.eps} 
 	\caption{Distribution of electrons.}
@@
 
-Simulations are one-dimensional and have following parameters: initial flow lorenz-factor $\gamma = 1.5$, number densities $n_e = 10^{-4} \rm{cm}^{-3}$, $n_p = 10^{-4} \rm{cm}^{-3}$ , temperature $5\cdot10^8 \rm{K}$, magnetic field $B = 10^{-4} \rm{G}$, the full size of the box $L = 2\cdot10^{12} \rm{cm}$, the number of cells $N=2\cdot10^4$.
-Electron mass is reduced to $m_e = \frac{m_p}{20}$.
-The full time of simulation is $T = 5000 {\omega_p}^{-1}$.
-$\theta$ is angle between flow velocity and magnetic field.
-We present results of particle spectrum in several simulations with different $\theta$.
+One can see on Figures \ref{protons} and \ref{electrons} spectrum of particles with different angle $\theta$.
+Spectrum consists of three parts - narrow peak of cold but fast moving initial flow, wide peak of hot downstream and non-thermal accelerated component.
+Figures show that spectrum of accelerated particles strongly depends on angle $\theta$.
+If $\theta$ is less then critical value, defined by equation $c\cdot cos(\theta_{crit})=v_{shock}$, where all values are measured in upstream rest frame, particles can escape from the front and cross it several times to gain more energy.
+It is difficult to evaluate critical angle apriori, because we don't know compression relation of shock wave which will be created in simulation, but estimation for case of strong wave (compression relation $\sigma=4$) gets $\theta_{crit}=45^{\circ}$ in downstream frame.
+And it is consistent with results of our simulations.
 
-One can see from Figures \ref{protons} and \ref{electrons} that spectrum of accelerated particles strongly depends on angle theta.
-If $\theta$ is less then critical value, defined by equation $c\cdot cos(\theta_{crit})=v_{shock}$, where all values are measured in upstream rest frame, particles can escape from the front and cross it several times to gain more energy.
-It is difficult to evaluate critical angle apriori, because we don't know compression relation of shock wave which will be created in simulation, but estimation for case of strong wave (comression relation $\sigma=4^{\circ}$) gets $\theta_{crit}=45$ in downstream frame.
-And it is consistent with results of our simulations.
+Also, one can see that for angles less then critical, particle spectrum are higher and longer for larger $\theta$.
+Spectrum of nonthermal component is about 10 times higher for protons than for electrons at the same energy.
+It is consistent with other Particle-in-cell simulations o shock waves {\cite{Sironi2011}}.
+
+Results show interesting effects for angles close to the critical value (black and yellow line)- electrons are still accelerating while protons are not.
+It can be explained by electrons accelerating in shortwave turbulence in precursor of shock wave.

### 2017-11-17 PIC++/distributions/compare_distribution.m

@@ -36,6 +36,13 @@ Nt(8) = fix(size(distribution_protons7,1)/Np) - 1;
 Nt(9) = fix(size(distribution_protons8,1)/Np) - 1;
 Nt(10) = fix(size(distribution_protons9,1)/Np) - 1;
 
+Nt(4) = 24;
+Nt(5) = 8;
+Nt(6) = 9;
+Nt(7) = 10;
+Nt(8) = 12;
+Nt(9) = 14;
+Nt(10) = 24;
 Fp(1:Np, 1:10) = 0;
 Fe(1:Np, 1:10) = 0;
 

### 2017-11-17 papers/Physica2017/Romansky_Physica2017.tex
paper

--- 2017-11-10 papers/Physica2017/Romansky_Physica2017.tex
@@
 \begin{document}
-\title{Particle-in-cell simulation of ion and electron acceleration in trans-relativistic shocks.}
+\title{Electron and ion acceleration by relativistic shocks: particle-in-cell model simulations}
 
-\author{V I Romansky$^{1,2}$, A M Bykov$^1$, S M Osipov$^1$ and P E Gladilin$^1$}
+\author{V I Romansky$^{1,2}$, A M Bykov$^{1,3}$ and S M Osipov$^1$}
 
@@
 	Universitetsky pr., 13, Moscow 119234, Russia}
+\address{$^3$ Peter the Great St.~Petersburg Polytechnic University, 29 Politekhnicheskaya st., St. Petersburg 195251, Russia}
 
@@
 \begin{abstract}
-The problem of Cosmic Ray (CR) origin and acceleration is a widely discussed problem of
-modern astrophysics. It is considered that most likely CR sources are supernova remnant
-shock waves. One of the methods to study particle injection and acceleration by shock
-waves is a particle-in-cell computer simulation. In this paper we present results of shock
-wave simulations based on our self-developed implicit particle-in-cell code. 
+	The problem of particle acceleration by the collisionless astrophysical shocks is of fundamental importance for the cosmic ray origin problem and the high energy astrophysics.
+The fast magnetohydrodynamical shocks produced by relativistic jets are considered as the very likely sites of the cosmic ray acceleration.
+	While cosmic ray acceleration by both non-relativistic and ultra relativistic shocks was studied in some details the shocks of a moderate Lorentz factors has attracted much less attention so far, while they are important for modelling the gamma-ray burst afterglows and a number of other interesting objects.
+In this paper we present the results of simulations of relativistic shock waves in the electron-proton plasmas obtained with an implicit particle-in-cell code for the initial study the particle injection efficiencies.
 \end{abstract}
 \input{intro.tex}
-\input{model.tex}
-\input{test.tex}
 \input{shock.tex}

### 2017-11-17 papers/Physica2017/conclusions.tex
paper

@@
 \section{Conclusions}
-In this paper we studied trans-relativistic shock wave using our implicit particle-in-cell code.
-Simulations show, that particles injection in acceleration process highly depends on angle between flow bulk velocity and magnetic field.
-Also injection of protons is more effective than electrons for angles less then critical, but there is regime of more effective electrons acceleration when angle is close to critical.
-Further research is necessary in order to study this phenomena in detail.
+In this paper we presented the first results of the PIC simulations of particle acceleration by the moderately relativistic shock waves.
+Simulations show, that the injection of particles into the acceleration process depends strongly on the angle between the flow bulk velocity and the magnetic field.
+The injection efficiency as a function of the inclination angle is studied.
+The injection of protons is more efficient than that of electrons for the shock inclination angles below the critical angle where the transition to the superluminal shock regime occurs.
+On the other hand the electron acceleration is more efficient in the shocks with the inclination angles which are close to the critical angle.
 \ack
 V I Romansky acknowledges support from RSF grant 16-12-10519.
-The results of the work were obtained using computational resources of Peter the Great Saint-Petersburg Polytechnic University Supercomputing Center (http://www.spbstu.ru).
+The results of the work were obtained using computational resources of Peter the Great Saint-Petersburg Polytechnic University Supercomputing Center (http://www.scc.spbstu.ru).

### 2017-11-17 papers/Physica2017/intro.tex
paper

--- 2017-11-14 papers/Physica2017/intro.tex
@@
 \section{Introduction}
-One of the possible sources of high-energy cosmic rays are shock waves in astrophysical collisionless plasma due to the diffusive shock acceleration (DSA) mechanism, see \cite{Bell1978}, \cite{Blandford1978}.
-The particle acceleration on the shock front is sophisticated and nonlinear problem, because cosmic rays have strongly influence on the shock wave.
-The pressure of cosmic rays is comparable with pressure of background plasma and significatly modifies the front (see for example \cite{Bykov2014}).
-Also there are a instabilities in anisotropic plasma, which lead to strong field amplification and the observations of the synchrotron emission from supernova remnants (SNRs) show, that the magnetic field there is about 100 times stronger then the interstellar field (see for example \cite{Berezhko2003},\cite{Uchiyama2007}).
-One of the most powerful methods to explore processes in collisionless plasma is the Particle-in-Cell simulation.
+One of the possible sources of the high-energy cosmic rays are the shock waves in astrophysical collisionless plasma due to the diffusive shock acceleration (DSA) mechanism, see \cite{Bell1978}, \cite{Blandford1978}.
+The particle acceleration on the shock front is a sophisticated and nonlinear problem, because the cosmic rays have a strong influence on the shock wave.
+The pressure of cosmic rays is comparable with the pressure of background plasma and significantly modifies the front (see, e.g., \cite{Bykov2014}).
+Also the instabilities in the anisotropic plasma can yield a strong field amplification, and according to the observations of synchrotron emission from supernova remnants (SNRs), the magnetic field there is about 100 times stronger than the interstellar field (see, e.g., \cite{Berezhko2003},\cite{Uchiyama2007}).
+One of the most powerful methods to explore the processes in collisionless plasma is the Particle-in-Cell simulation.

### 2017-11-17 papers/Physica2017/shock.tex
paper

--- 2017-11-15 papers/Physica2017/shock.tex
@@
 \section{Shock wave simulation}
-We studied the evolution of trans-relativistic shocks and spectum of accelerated particles.
-We chose relastivistic shocks with low lorenz-factor because the maximum energy of produced cosmic rays increases with shock wave lorenz-factor, but the efficiency of acceleration decreases, because it is more difficult for particle to cross fast moving front many times \cite{Ellison2013}.
-So we assume that intermediate case of trans-relativistic shocks provides the most effective acceleration.
+We studied the evolution of moderately relativistic shocks and spectrum of accelerated particles.
+We chose relativistic shocks with a low Lorentz factor because the maximum energy of the produced cosmic rays increases with the shock wave Lorentz factor, while the efficiency of acceleration decreases,
+because it is more difficult for particle to cross fast moving front many times 
+%
+\cite{Ellison2013}. So we assume that the intermediate case of moderately relativistic shocks provides the most efficient acceleration.
 
-We developed the implicit particle-in-cell (PIC) code, based on the scheme suggested by Lapenta et al.~\cite{Lapenta2006} and improved for the relativistic case by Noguchi et al.\cite{Noguchi2007}.
-Our code is fully three dimensional and parallelized with MPI technology, which is adapted for distributed computing and can be executed on a wide class of computers.
+We developed the implicit particle-in-cell (PIC) code, presented in our previous paper \cite{Romansky2016}, based on the scheme suggested by Lapenta et al.~\cite{Lapenta2006} and improved for the relativistic case by Noguchi et al.\cite{Noguchi2007}.
+Our code is fully three-dimensional and parallelized with MPI technology, which is adapted for distributed computing and can be executed on a wide class of computers. We use 
+%cleaning divergence of electric field 
+the electric field divergence correction
+after every several time steps and 
+%Fourier filtering of shortwave harmonics 
+the shortwave harmonics Fourier filtering
+to suppress the numerical Cherenkov instability and improve the energy conservation.
 
-In simulation setup the homogeneous plasma, which flows in through the right boundary
-and collides with the reflecting superconducting wall on the left boundary.
+In the modeling setup the homogeneous plasma flows in the simulation box through the right boundary
+and collides with the reflecting superconducting wall on the left boundary, launching the shock wave.
+It is a common way to initialize a shock wave, because the initialization using the Rankine–Hugoniot conditions does not take into account the microscopic distribution, and the shock wave, created in such way, will probably fall apart into several discontinuities.
 
- Simulations are one-dimensional and have following parameters: initial flow lorenz-factor $\gamma = 1.5$, number densities $n_e = 10^{-4} \rm{cm}^{-3}$, $n_p = 10^{-4} \rm{cm}^{-3}$ , temperature $5\cdot10^8 \rm{K}$, magnetic field $B = 10^{-4} \rm{G}$, the full size of the box $L = 4\cdot10^{12} \rm{cm}$, the number of cells $N=4\cdot10^4$.
-Electron mass is reduced to $m_e = \frac{m_p}{20}$.
-The full time of simulation is $T = 5000 {\omega_p}^{-1}$.
-All values are measured in downstream frame.
-$\theta$ is angle between flow velocity and magnetic field.
-We present results of particle spectrum in several simulations with different $\theta$.
 \begin{figure}[h!]
 	\centering
-	\includegraphics[width=0.8\textwidth]{fig/protons.eps} 
-	\caption{Distribution of protons.}
+	\includegraphics[width=1.0\textwidth]{fig/protons.eps} 
+	\caption{Distribution of protons in the relativistic shock wave with Lorentz factor 1.5 with different inclination angle.}
 	\label{protons}
 \end{figure}
+
+The simulations are one-dimensional and have the following parameters: the initial flow Lorentz factor $\gamma = 1.5$, the number densities $n_e = 10^{-4} \rm{cm}^{-3}$, $n_p = 10^{-4} \rm{cm}^{-3}$ , the temperature $5\cdot10^8 \rm{K}$, the magnetic field $B = 10^{-4} \rm{G}$, the full size of the box $L = 2\cdot10^{12} \rm{cm}$, the number of cells $N=2\cdot10^4$.
+The electron mass is reduced to $m_e = \frac{m_p}{20}$.
+The full time of simulation is $T = 5000 {\omega_p}^{-1}$.This values gives the dimensionless parameter magnetization $\sigma = \frac{B^2}{4\pi\gamma (n_p m_p + n_e m_e) c^2} = 0.003$.
+Inclination angle $\theta$ is the angle between the flow velocity and the magnetic field.
+We present the results for the particle spectrum in several simulations with different $\theta$.
+The stable shock waves formed in all simulations and the total energy deviation is less than $3\%$, and it proves that our code provides correct model of the shock waves.
+
 \begin{figure}[h!]
 	\centering
-	\includegraphics[width=0.8\textwidth]{fig/electrons.eps} 
-	\caption{Distribution of electrons.}
+	\includegraphics[width=1.0\textwidth]{fig/electrons.eps} 
+	\caption{Distribution of electrons in the relativistic shock wave with Lorentz factor 1.5 with different inclination angle.}
 	\label{electrons}
@@
 
-One can see that spectrum strongly depends of angle $\theta$.
-If $\theta$ is less then critical value, defined as $c\cdot cos(\theta_{crit})=v_{shock}$, measured in shock rest frame, shock is subluminal and particles can escape from front and than cross it enough of times to e accelerated.
+One can see in Figures \ref{protons} and \ref{electrons} the particle spectrum for different angles $\theta$.
+The spectra consist of three parts - the narrow peak of cold but fast moving initial flow, the wide peak of hot downstream and the non-thermal accelerated component.
+Figures show that the spectrum of accelerated particles strongly depends on angle $\theta$.
+If $\theta$ is less than critical value, defined by the equation $c\cdot \cos(\theta_{crit})=v_{shock}$, where all values are measured in the upstream rest frame, particles can escape from the front and cross it several times to gain more energy (see, e.g., \cite{Pelletier2017}).
+It is difficult to evaluate the critical angle a priori, because the compression relation of the shock wave which will be created in the simulation is unknown.
+However the estimation for case of a strong wave (the compression relation equals $4$) provides $\theta_{crit}=45^{\circ}$ in the downstream frame, which is consistent with the results of our simulations.
+
+Also, one can see that for the angles less then the critical one, the particle spectra are higher and longer for larger $\theta$.
+The spectrum of non-thermal component is about 10 times higher for protons than for electrons at the same energy.
+It is consistent with other Particle-in-cell simulations of the shock waves {\cite{Sironi2011}}.
+
+The results show some interesting effects for angles close to the critical value (black and yellow lines)- electrons are still accelerating while protons are not.
+It can be explained by electrons acceleration in the short-wavelength turbulence of the shock wave precursor.
+Due to the small gyroradius of electrons in initial flow, short-wavelength turbulence scatter them more efficiently than protons.
+However, further research is necessary in order to study this phenomena in detail.

### 2017-11-17 PIC++/distributions/compare_two_distribution.m

@@ -8,9 +8,13 @@ Np = 500;
 
 Nt(1:2) = 11;
 
+
 Nt(1) = fix(size(distribution_protons,1)/Np) - 1;
 Nt(2) = fix(size(distribution_protons_rel,1)/Np) - 1;
 
+
+%Nt(1)=3;
+%Nt(2)=5;
 
 Fp(1:Np, 1:2) = 0;
 Fe(1:Np, 1:2) = 0;

### 2018-01-11 papers/Physica2017/Romansky_Physica2017.tex
matlab

--- 2017-11-17 papers/Physica2017/Romansky_Physica2017.tex
@@
 \begin{document}
-\title{Particle-in-cell simulation of ion and electron acceleration in trans-relativistic shocks.}
+\title{Electron and ion acceleration by relativistic shocks: particle-in-cell simulations}
 
-\author{V I Romansky$^{1,2}$, A M Bykov$^1$ and S M Osipov$^1$}
+\author{V I Romansky$^{1,2}$, A M Bykov$^{1,3}$ and S M Osipov$^1$}
 
@@
 \begin{abstract}
-The problem of Cosmic Ray (CR) origin and acceleration is a widely discussed problem of
-modern astrophysics. It is considered that most likely CR sources are 
-shock waves in supernova remnants or jets. One of the methods to study particle injection and acceleration by shock
-waves is a particle-in-cell computer simulation. In this paper we present results of
-simulations of relativistic shock waves with low lorentz-factor, based on our self-developed implicit particle-in-cell code. 
+	The problem of particle acceleration by collisionless astrophysical shocks is of fundamental importance for the cosmic ray origin problem and  for the high energy astrophysics as a whole.
+Fast magnetohydrodynamical shocks produced by relativistic jets are considered as very likely sites of cosmic ray acceleration.
+	While cosmic ray acceleration by both non-relativistic and ultra relativistic shocks was studied in some details, the shocks of moderate Lorentz factors have attracted much less attention so far, while they are important for modeling gamma-ray burst afterglows and a number of other interesting objects.
+In this paper we present simulations of relativistic shock waves in electron-proton plasmas obtained with an implicit particle-in-cell code for the initial study the particle injection efficiencies.
 \end{abstract}
-\input{intro.tex}
-\input{shock.tex}
-\input{conclusions.tex}
-\input{biblio.tex}
+\section{Introduction}
+One of the possible sources of high-energy cosmic rays (CRs) are shock waves in astrophysical collisionless plasmas which are believed to accelerate particles via  the diffusive shock acceleration (DSA) mechanism, see, e.
+g.
+\cite{Bell1978}, \cite{Blandford1978}.
+The particle acceleration in the vicinity of shock front is a sophisticated and nonlinear problem, because the CRs strongly influence on the structure of the shock.
+Their pressure is comparable with the pressure of background plasma and significantly modifies the shock front (see, e.g., \cite{Bykov2014}).
+Also the instabilities in the anisotropic plasma can result to a strong magnetic field amplification.
+The spectra of synchrotron emission observed from supernova remnants (SNRs) indicate that the magnetic field in their shells is about 100 times stronger than the interstellar field (see, e.g., \cite{Berezhko2003},\cite{Uchiyama2007}).
+One of the most powerful methods to explore the processes in collisionless plasma is particle-in-cell (PIC) simulation.
+
+\section{Shock wave simulation}
+We studied the evolution of moderately relativistic shocks and spectra of accelerated particles.
+We chose relativistic shocks with a relatively low Lorentz factor ($\gamma=1.5$) because the maximum energy of the produced CRs cosmic rays increases with the shock Lorentz factor, while the efficiency of acceleration decreases,
+because it becomes more difficult for particle to cross fast moving front
+%
+\cite{Ellison2013}. So we assumed that an intermediate case of moderately relativistic shocks would provide the most efficient acceleration.
+
+To perform the simulations we developed the implicit PIC code \cite{Romansky2016}, based on the scheme suggested by Lapenta et al.~\cite{Lapenta2006} and improved for the relativistic case by Noguchi et al.\cite{Noguchi2007}.
+The code is fully three-dimensional and parallelized with MPI technology. The electric field divergence is corrected
+after every several time steps and 
+the shortwave harmonics Fourier filtering is applied
+to suppress the numerical Cherenkov instability and improve energy conservation.
+
+In the modeling setup the homogeneous plasma flows into the simulation box through its right boundary
+and collides with the reflecting superconducting wall at the left boundary thus launching a shock wave.
+This  is a common way to initialize a shock, because an initialization using the Rankine-Hugoniot conditions does not take into account the microscopic distribution, and a shock, created in such a way, would probably fall apart into several discontinuities.
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=1.0\textwidth]{fig/protons.eps} 
+	\caption{Distribution of protons in a relativistic shock wave with Lorentz factor 1.5 at various inclination angle $\theta$.}
+	\label{protons}
+\end{figure}
+
+The simulations are one-dimensional and have the following parameters of the initial flow: Lorentz factor $\gamma = 1.5$, the number densities $n_e = 10^{-4} \rm{cm}^{-3}$, $n_p = 10^{-4} \rm{cm}^{-3}$ , the temperature, $T_e = T_p = 5\cdot10^8 \rm{K}$, the magnetic field $B = 10^{-4} \rm{G}$, the full size of the box $L = 2\cdot10^{12} \rm{cm}$, the number of cells $N=2\cdot10^4$.
+The electron mass is increase to $m_e = \frac{m_p}{20}$ to enlarge electron scales of  the plasma, it is common approach in PIC simulations see e.g.
+{\cite{Sironi2011}} to reduce required computational recources.
+The full time of simulation is $T = 5000 {\omega_p}^{-1}$, where $\omega_p$ is electron plasma frequency.These values allow one to define the dimensionless parameter magnetization $\sigma = B^2/4\pi\gamma (n_p m_p + n_e m_e) c^2 = 0.003$.
+Inclination angle $\theta$ is the angle between the flow velocity and the magnetic field.
+We present results for the particle spectrum in several simulations at various $\theta$.
+Stable shocks formed in all simulations and the total energy deviation is less thanof the system conserved to better than $3\%$, which indicates that the code provides a correct model of the shock waves.
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=1.0\textwidth]{fig/electrons.eps} 
+	\caption{Distribution of electrons in a relativistic shock wave with Lorentz factor 1.5 at various inclination angle $\theta$.}
+	\label{electrons}
+\end{figure}
+
+One can see in Figures \ref{protons} and \ref{electrons} the particle spectra for various inclination angles $\theta$.
+The spectra consist of three parts - the narrow peak of cold but fast moving initial flow, the wide peak of hot downstream and the non-thermal accelerated component.
+Figures show that the spectrum of accelerated particles strongly depends on he inclination angle $\theta$.
+If $\theta$ is less than the critical value, defined by the equation $c\cdot \cos(\theta_{crit})=v_{shock}$, where all values are measured in the upstream rest frame, particles can escape from the front and cross it several times to gain more energy (see, e.g., \cite{Pelletier2017}), if $\theta$ is greater, shock becomes superluminal - particles need to move along the field lines faster than light to escape from the front, and acceleration is suppressed.
+It is difficult to evaluate the critical angle \textit{a priori}, because the compression relation of the shock which will be created in the simulation is unknown.
+However the estimate for the case of a strong wave (the compression relation equals $4$) provides $\theta_{crit}=45^{\circ}$ in the downstream frame, which is consistent with the results of our simulations.
+The small-scale oscilations of spectra are produced by numerical errors and should not be taken into account.
+
+Also, one can see that for the angles less then the critical one, concentrations of accelerated particles are higher and extend to higher enegies for larger $\theta$.
+The spectrum of non-thermal component is about 10 times higher for the protons than for the electrons at the same energy.
+It is consistent with other PIC simulations of the shock waves {\cite{Sironi2011}}.
+
+These results show some interesting effects for angles close to the critical value (black and yellow lines)- the electrons are still accelerating while the protons are not.
+It can be interpreted as electrons accelerate in the short-wavelength turbulence of the shock precursor.
+Due to the small gyroradii of the electrons in initial flow, the short-wavelength turbulence scatter them more efficiently than protons.
+However, further research is necessary to study this phenomenon in detail.
+
+\section{Conclusions}
+In this paper we presented results of the PIC simulations of particle acceleration by moderately relativistic shock waves in space plasmas.
+Simulations show, that injection of particles into the acceleration process strongly depends on the angle between the flow bulk velocity and the magnetic field.
+The injection efficiency as a function of the inclination angle has been studied.
+Injection of protons is more efficient than that of electrons for the inclination angles below the critical angle where the transition to the superluminal shock regime occurs.
+On  another hand the electron acceleration is more efficient in shocks with the inclination angles close to the critical angle.
+\ack
+V I Romansky acknowledges support from RSF grant 16-12-10519.
+The numerical results presented here were obtained using computational resources of Peter the Great Saint-Petersburg Polytechnic University Supercomputing Center (http://www.scc.spbstu.ru).
+
+\section*{References}
+\begin{thebibliography}{20}
+	\bibitem{Bell1978} Bell A R 1978 \textit{MNRAS} \textbf{182} 147
+	\bibitem{Blandford1978} Blandford R D and Ostriker J P 1978 \textit{ApJ} \textbf{221} L29 
+	\bibitem{Bykov2014} Bykov A M, Ellison D C, Osipov S M and Vladimirov A E 2014 \textit{ApJ} \textbf{789} Issue 2, 137
+	\bibitem{Berezhko2003} Berezhko E G, Ksenofontov L T and V{\"o}lk H J  2003 \textit{A}{\&}\textit{A} \textbf{412} L11
+	\bibitem{Uchiyama2007} Uchiyama Y, Aharonian F A, Tanaka T, Takahashi T and Maeda Y 2007 \textit{Nature} \textbf{449} 576
+	\bibitem{Romansky2016} Romansky V I, Bykov A M, Osipov S M and Gladilin P E \textit{Journal of Physics: CS}
+	\bibitem{Lapenta2006} Lapenta G, Brackbill J U and Ricci P 2006 \textit{Phys. Plasmas} \textbf{13} 055904
+	\bibitem{Noguchi2007} Noguchi K, Tronci C, Zuccaro G and Lapenta G 2007 \textit{Phys. Plasmas} \textbf{14} 042308
+	\bibitem{Ellison2013} Ellison D C, Warren D C and Bykov A M 2013 \textit{ApJ} \textbf{776} Issue 1, 46
+	\bibitem{Pelletier2017} Pelletier G, Bykov A M, Ellison D C and Lemoine M 2017 \textit{Space Science Reviews} \textbf{207} Issue 1-4, pp. 319-360
+	\bibitem{Sironi2011} Sironi L and Spitkovsky A 2011 \textit{ApJ} \textbf{741} Issue 1, 39
+\end{thebibliography}
 \end{document}

### 2018-01-11 PIC++/distributions/compare_distribution.m

@@ -47,16 +47,32 @@ Nt(10) = 24;
 Fp(1:Np, 1:10) = 0;
 Fe(1:Np, 1:10) = 0;
 
-
 Pp(1:Np, 1:10) = 0;
 Pe(1:Np, 1:10) = 0;
 
-me = 0.91*10^-25;
-mp = 1.6*10^-24;
-v=2.998*10^10;
+upstreamFp(1:Np) = 0;
+upstreamFe(1:Np) = 0;
+upstreamPp(1:Np) = 0;
+upstreamPe(1:Np) = 0;
+
+downstreamFp(1:Np) = 0;
+downstreamFe(1:Np) = 0;
+downstreamPp(1:Np) = 0;
+downstreamPe(1:Np) = 0;
+
+me = 0.910938356*10^-25;
+mp = 1.672621*10^-24;
+cv = 2.99792458*10^10;
+kBoltzman = 1.38064852*10^-16;
+T=5*10^8;
+
+gamma = 1.5;
+
+beta = sqrt(1 - 1/(gamma*gamma));
+v = beta*cv;
 
 %Color = {'r','green','blue','black','yellow','cyan','magenta','purple','grey'};
-Color = {[.7,.3,.3],'red','green','blue','black','yellow','cyan','magenta',[.5,.5,.5],[.3,.7,.3]};
+Color = {[.7,.3,.3],'red','green','blue','black','magenta','cyan','yellow',[.5,.5,.5],[.3,.7,.3],[1.0,.5,0],[.75,0.0,.7]};
 
 for i=1:Np,
    %Pp(i,1) = distribution_protons0(i + c*Np,1);
@@ -103,24 +119,67 @@ for i=1:Np,
    Fe(i,9) = distribution_electrons8(i + Nt(9)*Np, 2)*Pe(i,9)*Pe(i,9);
    Fe(i,10) = distribution_electrons9(i + Nt(10)*Np, 2)*Pe(i,10)*Pe(i,10);
 end;
+
+upstreamPp(1) = 0.99*beta*gamma*mp*cv;
+upstreamPe(1) = 0.99*beta*gamma*me*cv;
+for i=1:Np,
+    upstreamPp(i) = upstreamPp(1) + (i-1)*0.02*beta*gamma*mp*cv/Np;
+    upstreamPe(i) = upstreamPe(1) + (i-1)*0.02*beta*gamma*me*cv/Np;
+    downstreamPp(i) = Pp(i,2);
+    downstreamPe(i) = Pe(i,2);
+end;
+
[…]

### 2018-01-12 papers/Physica2017/Romansky_Physica2017.tex
paper

@@
 The simulations are one-dimensional and have the following parameters of the initial flow: Lorentz factor $\gamma = 1.5$, the number densities $n_e = 10^{-4} \rm{cm}^{-3}$, $n_p = 10^{-4} \rm{cm}^{-3}$ , the temperature, $T_e = T_p = 5\cdot10^8 \rm{K}$, the magnetic field $B = 10^{-4} \rm{G}$, the full size of the box $L = 2\cdot10^{12} \rm{cm}$, the number of cells $N=2\cdot10^4$.
-The electron mass is increase to $m_e = \frac{m_p}{20}$ to enlarge electron scales of  the plasma, it is common approach in PIC simulations see e.g.
+The electron mass $m_e$ is increase by $100$ times to enlarge electron scales of  the plasma, it is common approach in PIC simulations see e.g.
 {\cite{Sironi2011}} to reduce required computational recources.
@@
 We present results for the particle spectrum in several simulations at various $\theta$.
-Stable shocks formed in all simulations and the total energy deviation is less thanof the system conserved to better than $3\%$, which indicates that the code provides a correct model of the shock waves.
+Stable shocks formed in all simulations and the total energy of the system conserved to better than $3\%$, which indicates that the code provides a correct model of the shock waves.
 
@@
 One can see in Figures \ref{protons} and \ref{electrons} the particle spectra for various inclination angles $\theta$.
-The spectra consist of three parts - the narrow peak of cold but fast moving initial flow, the wide peak of hot downstream and the non-thermal accelerated component.
+The spectra consist of three parts - the narrow peak of cold but fast moving initial flow, approximated by Maxwell distribution, shifted on the initial momentum, the wide peak of hot downstream, approximated by Maxwell-J{\"u}ttner distribution, and the non-thermal accelerated component.
+The small-scale oscilations of spectra are produced by numerical errors and should not be taken into account.
 Figures show that the spectrum of accelerated particles strongly depends on he inclination angle $\theta$.
@@
 However the estimate for the case of a strong wave (the compression relation equals $4$) provides $\theta_{crit}=45^{\circ}$ in the downstream frame, which is consistent with the results of our simulations.
-The small-scale oscilations of spectra are produced by numerical errors and should not be taken into account.
 
@@
 	\bibitem{Uchiyama2007} Uchiyama Y, Aharonian F A, Tanaka T, Takahashi T and Maeda Y 2007 \textit{Nature} \textbf{449} 576
-	\bibitem{Romansky2016} Romansky V I, Bykov A M, Osipov S M and Gladilin P E \textit{Journal of Physics: CS}
+	\bibitem{Romansky2016} Romansky V I, Bykov A M, Osipov S M and Gladilin P E 2017 \textit{Journal of Physics: CS} \textbf{929} id 012014 
 	\bibitem{Lapenta2006} Lapenta G, Brackbill J U and Ricci P 2006 \textit{Phys. Plasmas} \textbf{13} 055904

### 2018-01-12 PIC++/distributions/compare_distribution.m

@@ -23,7 +23,10 @@ load distribution_electrons9.dat;
 
 Np = 500;
 
-Nt(1:11) = 0;
+Nt(1:10) = 0;
+sumP(1:10) = 0;
+sumE(1:10) = 0;
+
 
 %Nt(1) = fix(size(distribution_protons0,1)/Np) - 1;
 Nt(2) = fix(size(distribution_protons1,1)/Np) - 1;
@@ -59,6 +62,11 @@ downstreamFe(1:Np) = 0;
 downstreamPp(1:Np) = 0;
 downstreamPe(1:Np) = 0;
 
+sumUpstreamP = 0;
+sumUpstreamE = 0;
+sumDownstreamP = 0;
+sumDownstreamE = 0;
+
 me = 0.910938356*10^-25;
 mp = 1.672621*10^-24;
 cv = 2.99792458*10^10;
@@ -71,7 +79,7 @@ beta = sqrt(1 - 1/(gamma*gamma));
 v = beta*cv;
 
 %Color = {'r','green','blue','black','yellow','cyan','magenta','purple','grey'};
-Color = {[.7,.3,.3],'red','green','blue','black','cyan','magenta','yellow',[.5,.5,.5],[.3,.7,.3],[1.0,.5,0],[.75,0.0,.7]};
+Color = {[.7,.3,.3],'red','green','blue','black','yellow',[.5,.5,.5],'cyan',[.3,.7,.3], 'magenta',[1.0,.5,0],[.75,0.0,.7]};
 
 for i=1:Np,
    %Pp(i,1) = distribution_protons0(i + c*Np,1);
@@ -118,6 +126,14 @@ for i=1:Np,
    Fe(i,9) = distribution_electrons8(i + Nt(9)*Np, 2)*Pe(i,9)*Pe(i,9);
    Fe(i,10) = distribution_electrons9(i + Nt(10)*Np, 2)*Pe(i,10)*Pe(i,10);
 end;
+for i=1:Np-1,
+    for j=1:10,
+        p = Pp(i,j);
+        sumP(j) = sumP(j) + Fp(i,j)*p^-2*(Pp(i+1,j) - Pp(i,j));
+        p = Pe(i,j);
+        sumE(j) = sumE(j) + Fe(i,j)*p^-2*(Pe(i+1,j) - Pe(i,j));
+    end;
+end;
 
 weightP = 0.1;
 weightE = 0.1;
@@ -135,7 +151,7 @@ denomP = (2*pi*mp*kBoltzman*T)^-1.5;
[…]

### 2018-01-14 papers/Physica2016/shock.tex
matlab

\section{Shock wave simulation}
We studied the influence of fractions of heavy ions on the shock wave, in particular, the spectrum of particles and the shape of the shock wave. We present results of two simulations with different compositions of the plasma : pure protons and electrons in the first case and with the admixture of alpha particles in the second case. Initially the homogeneous flux is  moving from the right free boundary to the left. On the left side there is a reflecting super-conducting wall, which causes formation of shock wave. Simulations are one-dimensional and have following parameters: initial velocity $v = 0.2c$, number densities $n_e = 10^{-4} \rm{cm}^{-3}$, $n_p = 10^{-4} \rm{cm}^{-3}$ in the first simulation and $n_e = 10^{-4} \rm{cm}^{-3}$, $n_p = 0.6\cdot10^{-4} \rm{cm}^{-3}$, $n_\alpha = 0.2\cdot10^{-4} \rm{cm}^{-3}$ in the second, temperature $5\cdot10^8 \rm{K}$, magnetic field $B_\parallel = 10^{-4} \rm{G}, B_\perp= 0.7\cdot10^{-4} \rm{G}$, the full size of the box $L = 1\cdot10^{12} \rm{cm}$, the number of cells $N=2\cdot10^4$. Electron mass is reduced to $m_e = \frac{m_p}{20}$. The full time of simulation is $T = 2000 {\omega_p}^{-1}$.
\begin{figure}[h!]
	\centering
	\begin{minipage}{0.49\textwidth}
		\centering
		%\includegraphics[width=0.98\textwidth]{fig/protons.eps} 
		\caption{Distribution of protons in the simulation without alpha particles.}
		\label{protons}
	\end{minipage}\hfill
	\begin{minipage}{0.49\textwidth}
		\centering
		%\includegraphics[width=0.98\textwidth]{fig/protons_with_He.eps} 
		\caption{Distribution of protons in the simulation with alpha particles.}
		\label{protons_with_alpha}
	\end{minipage}
\end{figure}
\begin{figure}[h!]
	\centering
	\begin{minipage}{0.49\textwidth}
		\centering
		%\includegraphics[width=0.98\textwidth]{fig/electrons.eps} 
		\caption{Distribution of electrons in the simulation without alpha particles.}
		\label{electrons}
	\end{minipage}\hfill
	\begin{minipage}{0.49\textwidth}
		\centering
		%\includegraphics[width=0.98\textwidth]{fig/electrons_with_He.eps} 
		\caption{Distribution of electrons in the simulation with alpha particles.}
		\label{electrons_with_alpha}
	\end{minipage}
\end{figure}

The results presented in Figures \ref{protons}-\ref{electrons_with_alpha} show that the values of $F(p)p^4$ for high energies are much greater for protons, than for electrons in both simulations. Also electrons need much more time to be accelerated and to form spectrum $\propto p^{-4}$. It means that protons are injected into the acceleration process more efficiently, and it is consistent with the work of Park et al. {\cite{Park2015}}. Also we have shown that minority of heavy ions increases the spectrum of protons and do not have influence on the spectrum of electrons. It can be explained by the fact, that heavy ions form the large scale turbulence and protons can efficiently scatter on this turbulence. Otherwise for electrons turbulence produced by protons is already enough large-scale  and the influence of heavy ions is neglectable.

### 2018-01-14 PIC++/initialization.cpp

@@ -4504,7 +4504,7 @@ void Simulation::initializeLangmuirWave() {
 
 void Simulation::initializeFluxFromRight() {
 	boundaryConditionTypeX = SUPER_CONDUCTOR_LEFT;
-	//boundaryConditionTypeX = PERIODIC;
+	boundaryConditionTypeX = PERIODIC;
 	createParticles();
 	E0 = E0 - V0.vectorMult(B0) / (speed_of_light_normalized * speed_of_light_correction);
 	//initializeAlfvenWaveY(10, 1.0E-4);

### 2018-01-14 PIC++/paths.h

@@ -26,10 +26,10 @@
 //const char* const backupDirectory = "/home/uikam/romansky/PIC3d3/backup/";
 
 //for MPI on Ubuntu
-const char* const outputDirectory = "/home/vadim/PICpp/PIC++/output/";
-const char* const reducedOutputDirectory = "/home/vadim/PICpp/PIC++/reduced_output/";
-const char* const inputDirectory = "/home/vadim/PICpp/PIC++/input/";
-const char* const backupDirectory = "/home/vadim/PICpp/PIC++/backup/";
+//const char* const outputDirectory = "/home/vadim/PICpp/PIC++/output/";
+//const char* const reducedOutputDirectory = "/home/vadim/PICpp/PIC++/reduced_output/";
+//const char* const inputDirectory = "/home/vadim/PICpp/PIC++/input/";
+//const char* const backupDirectory = "/home/vadim/PICpp/PIC++/backup/";
  
 //for windows 32
 //const char* const outputDirectory = "C:/users/Vadik/Documents/Visual Studio 2010/Projects/v4/PIC++/output/";
@@ -38,10 +38,10 @@ const char* const backupDirectory = "/home/vadim/PICpp/PIC++/backup/";
 //const char* const backupDirectory = "C:/users/Vadik/Documents/Visual Studio 2010/Projects/v4/PIC++/backup/";
 
 //for windows 64 home
-//const char* const outputDirectory = "C:/users/Vadik/Documents/Visual Studio 2010/Projects/PICpp/PIC++/output/";
-//const char* const reducedOutputDirectory = "C:/users/Vadik/Documents/Visual Studio 2010/Projects/PICpp/PIC++/reduced_output/";
-//const char* const inputDirectory = "C:/users/Vadik/Documents/Visual Studio 2010/Projects/PICpp/PIC++/input/";
-//const char* const backupDirectory = "C:/users/Vadik/Documents/Visual Studio 2010/Projects/PICpp/PIC++/backup/";
+const char* const outputDirectory = "C:/users/Vadik/Documents/Visual Studio 2010/Projects/PICpp/PIC++/output/";
+const char* const reducedOutputDirectory = "C:/users/Vadik/Documents/Visual Studio 2010/Projects/PICpp/PIC++/reduced_output/";
+const char* const inputDirectory = "C:/users/Vadik/Documents/Visual Studio 2010/Projects/PICpp/PIC++/input/";
[…]

### 2018-01-15 papers/Physica2017/Romansky_Physica2017.tex
output parser

@@
 %\usepackage{soul,ulem,color,xspace,bm}
-% Andrei's commands
 % Suggest to remove
@@
 and collides with the reflecting superconducting wall at the left boundary thus launching a shock wave.
-This  is a common way to initialize a shock, because an initialization using the Rankine-Hugoniot conditions does not take into account the microscopic distribution, and a shock, created in such a way, would probably fall apart into several discontinuities.
+This  is a common way to initialize a shock (e.
+g.
+\cite{Sironi2011}, \cite{Iwamoto2017}), because an initialization using the Rankine-Hugoniot conditions does not take into account the microscopic distribution, and a shock, created in such a way, would probably fall apart into several discontinuities.
 
@@
 
-These results show some interesting effects for angles close to the critical value (black and yellow lines)- the electrons are still accelerating while the protons are not.
+These results show some interesting effects for angles close to the critical value (black and yellow solid lines)- the electrons are still accelerating while the protons are not.
 It can be interpreted as electrons accelerate in the short-wavelength turbulence of the shock precursor.
@@
 	\bibitem{Sironi2011} Sironi L and Spitkovsky A 2011 \textit{ApJ} \textbf{741} Issue 1, 39
+	\bibitem{Iwamoto2017} Iwamoto M, Amano T, Hoshino M and Matsumoto Y 2017 \textit{ApJ} \textbf{840} Issue 1 article id. 52, 14 pp.
 \end{thebibliography}

### 2018-01-15 PIC++/initialization.cpp

@@ -360,6 +360,7 @@ Simulation::Simulation(int xn, int yn, int zn, double xsizev, double ysizev, dou
 		deltaX2 /= scaleFactor * scaleFactor;
 		deltaY2 /= scaleFactor * scaleFactor;
 		deltaZ2 /= scaleFactor * scaleFactor;
+        cellVolume /= scaleFactor * scaleFactor * scaleFactor;
 
 		leftX /= scaleFactor;
 		rightX /= scaleFactor;

### 2018-01-23 papers/Physica2017/Romansky_Physica2017.tex
paper

@@
 The spectra of synchrotron emission observed from supernova remnants (SNRs) indicate that the magnetic field in their shells is about 100 times stronger than the interstellar field (see, e.g., \cite{Berezhko2003},\cite{Uchiyama2007}).
-One of the most powerful methods to explore the processes in collisionless plasma is particle-in-cell (PIC) simulation.
+One of the most powerful methods to explore the processes in collisionless plasma on microscopic scales is particle-in-cell (PIC) simulation.
 
@@
 One can see in Figures \ref{protons} and \ref{electrons} the particle spectra for various inclination angles $\theta$.
-The spectra consist of three parts - the narrow peak of cold but fast moving initial flow, approximated by Maxwell distribution, shifted on the initial momentum, the wide peak of hot downstream, approximated by Maxwell-J{\"u}ttner distribution, and the non-thermal accelerated component.
+These spectra consist of three parts - the narrow peak of cold but fast moving initial flow, the wide peak of hot downstream, and the non-thermal accelerated component.
+Initial flow spectra are compared with Maxwell distribution, shifted on the initial momentum, and hot downstream spectra are compared with Maxwell-J{\"u}ttner distribution.
 The small-scale oscilations of spectra are produced by numerical errors and should not be taken into account.
@@
 
-These results show some interesting effects for angles close to the critical value (black and yellow solid lines)- the electrons are still accelerating while the protons are not.
+These results show some new interesting effects for angles close to the critical value (black and yellow solid lines)- the electrons are still accelerating while the protons are not.
 It can be interpreted as electrons accelerate in the short-wavelength turbulence of the shock precursor.

### 2018-01-23 PIC++/initialization.cpp

@@ -334,8 +334,8 @@ Simulation::Simulation(int xn, int yn, int zn, double xsizev, double ysizev, dou
 		//}
 		scaleFactor = speed_of_light * plasma_period;
 
-		plasma_period = 1.0;
-		scaleFactor = 1.0;
+		//plasma_period = 1.0;
+		//scaleFactor = 1.0;
 
 		//scaleFactor = xsize;
 		//scaleFactor = xsizeGeneral;

### 2018-01-24 papers/Physica2017/Romansky_Physica2017.tex
paper

@@
 
-In the modeling setup the homogeneous plasma flows into the simulation box through its right boundary
+In the model setup the homogeneous plasma flows into the simulation box through its right boundary
 and collides with the reflecting superconducting wall at the left boundary thus launching a shock wave.
@@
 The electron mass $m_e$ is increase by $100$ times to enlarge electron scales of  the plasma, it is common approach in PIC simulations see e.g.
-{\cite{Sironi2011}} to reduce required computational recources.
+{\cite{Sironi2011}} to reduce required computational resources.
 The full time of simulation is $T = 5000 {\omega_p}^{-1}$, where $\omega_p$ is electron plasma frequency.These values allow one to define the dimensionless parameter magnetization $\sigma = B^2/4\pi\gamma (n_p m_p + n_e m_e) c^2 = 0.003$.
@@
 One can see in Figures \ref{protons} and \ref{electrons} the particle spectra for various inclination angles $\theta$.
-These spectra consist of three parts - the narrow peak of cold but fast moving initial flow, the wide peak of hot downstream, and the non-thermal accelerated component.
-Initial flow spectra are compared with Maxwell distribution, shifted on the initial momentum, and hot downstream spectra are compared with Maxwell-J{\"u}ttner distribution.
-The small-scale oscilations of spectra are produced by numerical errors and should not be taken into account.
+Each spectrum consists of three parts: the narrow peak of cold but moving fast initial upstream flow, the wide peak of hot downstream and the non-thermal accelerated component.
+The initial upstream flow spectrum is compared with the Maxwell distribution, shifted by the initial momentum, while the hot downstream spectrum is compared with the Maxwell-J{\"u}ttner distribution.
+The small-scale oscillations of spectra are produced by numerical errors and should not be taken into account.
 Figures show that the spectrum of accelerated particles strongly depends on he inclination angle $\theta$.
@@
 
-Also, one can see that for the angles less then the critical one, concentrations of accelerated particles are higher and extend to higher enegies for larger $\theta$.
+Also, one can see that for the angles less then the critical one, concentrations of accelerated particles are higher and extend to higher energies for larger $\theta$.
 The spectrum of non-thermal component is about 10 times higher for the protons than for the electrons at the same energy.

### 2018-01-24 PIC++/constants.h

@@ -30,7 +30,7 @@ const double epsilon = 1E-16;
 const double maxErrorLevel = 1E-8;
 const double maxCleanupErrorLevel = 1E-8;
 const double particleVelocityErrorLevel = 1E-14;
-const double timeEpsilonKourant = 0.4;
+const double timeEpsilonKourant = 0.2;
 const double timeEpsilonFields = 0.2;
 const double timeEpsilonPlasma = 0.2;
 const double timeEpsilonCapture = 0.2;

### 2019-03-22 papers/Physica2019/Romansky_Physica2019.tex
paper

--- 2018-01-15 papers/Physica2017/Romansky_Physica2017.tex
@@
 \begin{document}
-\title{Electron and ion acceleration by relativistic shocks: particle-in-cell simulations}
+\title{Modelling of electron acceleration in relativistic supernovae}
 
-\author{V I Romansky$^{1,2}$, A M Bykov$^{1,3}$ and S M Osipov$^1$}
+\author{V I Romansky$^{1}$, A M Bykov$^{1,2}$ and S M Osipov$^1$}
 
 \address{$^1$ Ioffe Institute, 26 Politekhnicheskaya st., St. Petersburg 194021, Russia}
-\address{$^2$
-	Sternberg Astronomical Institute, Moscow State University
-	Universitetsky pr., 13, Moscow 119234, Russia}
-\address{$^3$ Peter the Great St.~Petersburg Polytechnic University, 29 Politekhnicheskaya st., St. Petersburg 195251, Russia}
+\address{$^2$ Peter the Great St.~Petersburg Polytechnic University, 29 Politekhnicheskaya st., St. Petersburg 195251, Russia}
 
@@
 \begin{abstract}
-	The problem of particle acceleration by collisionless astrophysical shocks is of fundamental importance for the cosmic ray origin problem and  for the high energy astrophysics as a whole.
-Fast magnetohydrodynamical shocks produced by relativistic jets are considered as very likely sites of cosmic ray acceleration.
-	While cosmic ray acceleration by both non-relativistic and ultra relativistic shocks was studied in some details, the shocks of moderate Lorentz factors have attracted much less attention so far, while they are important for modeling gamma-ray burst afterglows and a number of other interesting objects.
-In this paper we present simulations of relativistic shock waves in electron-proton plasmas obtained with an implicit particle-in-cell code for the initial study the particle injection efficiencies.
+	Modern observations of supernova remnants radio in radio ???
+with high angle resolution show that there exists type of objects, expanding with velocities close to speed of light.
+Model of electron acceleration is necessary for reasonable interpretation of this observations.
+In this paper we present numerical Particle-in-Cell simulation of relativistic shock wave propagating in turbulent medium.
 \end{abstract}
 \section{Introduction}
-One of the possible sources of high-energy cosmic rays (CRs) are shock waves in astrophysical collisionless plasmas which are believed to accelerate particles via  the diffusive shock acceleration (DSA) mechanism, see, e.
-g.
-\cite{Bell1978}, \cite{Blandford1978}.
-The particle acceleration in the vicinity of shock front is a sophisticated and nonlinear problem, because the CRs strongly influence on the structure of the shock.
-Their pressure is comparable with the pressure of background plasma and significantly modifies the shock front (see, e.g., \cite{Bykov2014}).
-Also the instabilities in the anisotropic plasma can result to a strong magnetic field amplification.
-The spectra of synchrotron emission observed from supernova remnants (SNRs) indicate that the magnetic field in their shells is about 100 times stronger than the interstellar field (see, e.g., \cite{Berezhko2003},\cite{Uchiyama2007}).
-One of the most powerful methods to explore the processes in collisionless plasma is particle-in-cell (PIC) simulation.
 
-\section{Shock wave simulation}
-We studied the evolution of moderately relativistic shocks and spectra of accelerated particles.
-We chose relativistic shocks with a relatively low Lorentz factor ($\gamma=1.5$) because the maximum energy of the produced CRs cosmic rays increases with the shock Lorentz factor, while the efficiency of acceleration decreases,
-because it becomes more difficult for particle to cross fast moving front
-%
-\cite{Ellison2013}. So we assumed that an intermediate case of moderately relativistic shocks would provide the most efficient acceleration.
+\section{Numerical setup}
 
-To perform the simulations we developed the implicit PIC code \cite{Romansky2016}, based on the scheme suggested by Lapenta et al.~\cite{Lapenta2006} and improved for the relativistic case by Noguchi et al.\cite{Noguchi2007}.
-The code is fully three-dimensional and parallelized with MPI technology. The electric field divergence is corrected
-after every several time steps and 
-the shortwave harmonics Fourier filtering is applied
-to suppress the numerical Cherenkov instability and improve energy conservation.
-
-In the modeling setup the homogeneous plasma flows into the simulation box through its right boundary
-and collides with the reflecting superconducting wall at the left boundary thus launching a shock wave.
-This  is a common way to initialize a shock (e.
-g.
-\cite{Sironi2011}, \cite{Iwamoto2017}), because an initialization using the Rankine-Hugoniot conditions does not take into account the microscopic distribution, and a shock, created in such a way, would probably fall apart into several discontinuities.
-
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width=1.0\textwidth]{fig/protons.eps} 
-	\caption{Distribution of protons in a relativistic shock wave with Lorentz factor 1.5 at various inclination angle $\theta$.}
-	\label{protons}
-\end{figure}
-
-The simulations are one-dimensional and have the following parameters of the initial flow: Lorentz factor $\gamma = 1.5$, the number densities $n_e = 10^{-4} \rm{cm}^{-3}$, $n_p = 10^{-4} \rm{cm}^{-3}$ , the temperature, $T_e = T_p = 5\cdot10^8 \rm{K}$, the magnetic field $B = 10^{-4} \rm{G}$, the full size of the box $L = 2\cdot10^{12} \rm{cm}$, the number of cells $N=2\cdot10^4$.
-The electron mass $m_e$ is increase by $100$ times to enlarge electron scales of  the plasma, it is common approach in PIC simulations see e.g.
-{\cite{Sironi2011}} to reduce required computational recources.
-The full time of simulation is $T = 5000 {\omega_p}^{-1}$, where $\omega_p$ is electron plasma frequency.These values allow one to define the dimensionless parameter magnetization $\sigma = B^2/4\pi\gamma (n_p m_p + n_e m_e) c^2 = 0.003$.
-Inclination angle $\theta$ is the angle between the flow velocity and the magnetic field.
-We present results for the particle spectrum in several simulations at various $\theta$.
-Stable shocks formed in all simulations and the total energy of the system conserved to better than $3\%$, which indicates that the code provides a correct model of the shock waves.
-
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width=1.0\textwidth]{fig/electrons.eps} 
-	\caption{Distribution of electrons in a relativistic shock wave with Lorentz factor 1.5 at various inclination angle $\theta$.}
-	\label{electrons}
-\end{figure}
-
-One can see in Figures \ref{protons} and \ref{electrons} the particle spectra for various inclination angles $\theta$.
-The spectra consist of three parts - the narrow peak of cold but fast moving initial flow, approximated by Maxwell distribution, shifted on the initial momentum, the wide peak of hot downstream, approximated by Maxwell-J{\"u}ttner distribution, and the non-thermal accelerated component.
-The small-scale oscilations of spectra are produced by numerical errors and should not be taken into account.
-Figures show that the spectrum of accelerated particles strongly depends on he inclination angle $\theta$.
-If $\theta$ is less than the critical value, defined by the equation $c\cdot \cos(\theta_{crit})=v_{shock}$, where all values are measured in the upstream rest frame, particles can escape from the front and cross it several times to gain more energy (see, e.g., \cite{Pelletier2017}), if $\theta$ is greater, shock becomes superluminal - particles need to move along the field lines faster than light to escape from the front, and acceleration is suppressed.
-It is difficult to evaluate the critical angle \textit{a priori}, because the compression relation of the shock which will be created in the simulation is unknown.
-However the estimate for the case of a strong wave (the compression relation equals $4$) provides $\theta_{crit}=45^{\circ}$ in the downstream frame, which is consistent with the results of our simulations.
-
-Also, one can see that for the angles less then the critical one, concentrations of accelerated particles are higher and extend to higher enegies for larger $\theta$.
-The spectrum of non-thermal component is about 10 times higher for the protons than for the electrons at the same energy.
-It is consistent with other PIC simulations of the shock waves {\cite{Sironi2011}}.
-
-These results show some interesting effects for angles close to the critical value (black and yellow solid lines)- the electrons are still accelerating while the protons are not.
-It can be interpreted as electrons accelerate in the short-wavelength turbulence of the shock precursor.
-Due to the small gyroradii of the electrons in initial flow, the short-wavelength turbulence scatter them more efficiently than protons.
-However, further research is necessary to study this phenomenon in detail.
+\section{Results}
 
 \section{Conclusions}
-In this paper we presented results of the PIC simulations of particle acceleration by moderately relativistic shock waves in space plasmas.
-Simulations show, that injection of particles into the acceleration process strongly depends on the angle between the flow bulk velocity and the magnetic field.
-The injection efficiency as a function of the inclination angle has been studied.
-Injection of protons is more efficient than that of electrons for the inclination angles below the critical angle where the transition to the superluminal shock regime occurs.
-On  another hand the electron acceleration is more efficient in shocks with the inclination angles close to the critical angle.
+
 \ack
-V I Romansky acknowledges support from RSF grant 16-12-10519.
+
 The numerical results presented here were obtained using computational resources of Peter the Great Saint-Petersburg Polytechnic University Supercomputing Center (http://www.scc.spbstu.ru).

### 2019-03-29 papers/Physica2019/Romansky_Physica2019.tex
maple

@@
 \begin{abstract}
-	Modern observations of supernova remnants radio in radio ???
-with high angle resolution show that there exists type of objects, expanding with velocities close to speed of light.
+	Modern observations with high angle resolution of radio emission in supernova remnants show that there exists type of objects, expanding with velocities close to speed of light.
 Model of electron acceleration is necessary for reasonable interpretation of this observations.
@@
 \section{Numerical setup}
+We studied evolution of quasi-perpendicular trans-relativistic shock, propagating in turbulent medium, and spectra of accelerated particles.
+The simulation is two dimensional, with three dimensional velocities and fields.
+Plasma flows into the simulation region through the right boundary, and reflects on the super-conducting wall at the left boundary.
+We used following parameters for setup: the initial flow Lorentz factor $\gamma = 1.5$, magnetization $\sigma = \frac{B^2}{4\pi\gamma (n_p m_p + n_e m_e) c^2} = 0.04$, in turbulent case $B^2$ means mean square field.
+The temperature $T = 5\cdot10^8 \rm{K}$ and proton mass is reduced to $m_p = 25 m_e$.
+Size of simulation region along x axis is $Lx = 8000\frac{c}{\omega_p}$ and in transverse direction $Ly = 200\frac{c}{\omega_p}$, where $\omega_p$ is plasma frequency $\omega_p = \sqrt{\frac{4\pi q^2 n}{\gamma m_e}}$.
+This sizes correspond to $80000$ and $2000$ grid points respectively.
+Also, $2000$ grid points corresponds to approximately $10$ gyroradii of upstream protons.
+We initialize turbulent field via following formula: 
+\begin{equation}
+B_{turb} (x) = \sum_{i=0}^{n}\sum_{j=0}^{n}\sum_{p=1,2}B(k_x(i),k_y(j)) \textbf{e}_{p} sin(k_x(i) + k_y(j) + \phi (i,j,p))
+\end{equation}
+where $B(k_x(i),k_y(j))$ is amplitude of turbulent mode, usually we choose $B \propto k^{-5/3}$ and normalized to fixed fraction of total magnetic energy, $\eta$.
+$\textbf{e}_{p}$ are vectors corresponding to two different field polarizations and $\phi (i,j,p)$ is randomly generated phase for each mode.
+Wave vectors $k$ are distributed on uniform grid, $k_x = i \Delta k, k_y = j \Delta k$ and $\Delta k \approx \frac{2 \pi}{10 r_g}$ where $r_g$ is upstream proton gyroradius.
+And the maximum wavevector $k_{max} \approx \frac{2 \pi}{r_g}$.
+This equations are evaluated in plasma rest frame and then magnetic field is transformed to the downstream frame.
 
+In this work we used setups with different turbulence scales, fraction of turbulent energy and regular field orientation and studied how this parameters influence on accelerated particles spectra.
 \section{Results}

### 2019-04-05 papers/Physica2019/Romansky_Physica2019.tex
paper

@@
 \section{Introduction}
+Shock waves in supernova remnants are possible sources of cosmic rays.
+The process of acceleration on the shock front is nonlinear problem and it's efficiency depends on large number of factors.
+In this work we studied the trans-relativistic shocks because they are probably the most effective.
+Energy that particle gains when crosses the shock front increases with shock velocity, but in ultra-relativistic shocks only small fraction of particles can return from downstream and efficiency is relatively low.
+In trans-relativistis shocks, with Lorentz-factor about 1.5, energy is high and particles still can return from downstream and cross shock many times.
+Another problem is that due to Lorentz transformation perpendicular component of magnetic field growths and shock wave becomes quasi-perpendicular for majority of fields inclination angles.
+Quasi-perpendicular shock waves are very weak sources of accelerated particles as shown in .
+We introduce turbulent field in plasma medium to make shock wave quasi-parallel in some regions and make acceleration process more efficient.
 
@@
 
+Simulation shows that structure of shock wave and velocity of it's propagation depends on properties of turbulence in upstream flow.
+On Figures \ref{regularB}, \ref{turbulentB} one can see relation of magnetic field in shock wave to upstream regular field.
+In case of strong turbulence in upstream, in downstream also exists strong turbulence.
+Shock front moves slowly and seems to be coorugated.
+
+\begin{figure}[h!]
+	\centering
+	\begin{minipage}{0.49\textwidth}
+		\centering
+		\includegraphics[width=0.98\textwidth]{fig/regular_field.eps} 
+		\caption{Magnetic field in shock wave without turbulence.}
+		\label{regularB}
+	\end{minipage}\hfill
+	\begin{minipage}{0.49\textwidth}
+		\centering
+		\includegraphics[width=0.98\textwidth]{fig/turbulent_field.eps} 
+		\caption{Magnetic field in shock wave with 90\% turbulence.}
+		\label{turbulentB}
+	\end{minipage}
+\end{figure}
+Spectrum of accelerated electrons shows strong dependence on fraction of turbulent magnetic field, as shown on in Figure \ref{spectrum}.
+When half of energy is in turbulence, spectrum becomes significantly different from spectrum in regular field.
+Fraction of accelerated electrons highly increases with growth of turbulent energy fraction.
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=0.8\textwidth]{fig/spectrum.eps} 
+	\caption{Distribution of electrons in the relativistic shock wave with different turbulence fractions.}
+	\label{spectrum}
+\end{figure}
+
 \section{Conclusions}
@@
 	\bibitem{Iwamoto2017} Iwamoto M, Amano T, Hoshino M and Matsumoto Y 2017 \textit{ApJ} \textbf{840} Issue 1 article id. 52, 14 pp.
+	\bibitem{Margutti} Margutti R et al. 2014 \textit{ApJ}. \textbf{797} Issue 2, article id. 107 8 pp.
+	\bibitem{Buneman} Buneman O 1993 \textit{Computer Space Plasma Physics} Terra Scientific Tokyo  67 p.
+	\bibitem{Crumley} Crumley P, Caprioli D, Markoff S and Spitkovsky A \textit{MNRAS} \textbf{485} Issue 4, pp. 5105-5119
+
 \end{thebibliography}

### 2019-04-08 papers/Physica2019/Romansky_Physica2019.tex
paper

@@
 \begin{abstract}
-	Modern observations with high angle resolution of radio emission in supernova remnants show that there exists type of objects, expanding with velocities close to speed of light.
+	Modern observations with high angle resolution of radio emission in supernova remnants show that there exists type of objects, expanding with velocities close to speed of light \cite{Margutti}.
 Model of electron acceleration is necessary for reasonable interpretation of this observations.
@@
 \section{Introduction}
-Shock waves in supernova remnants are possible sources of cosmic rays.
+Shock waves in supernova remnants are possible sources of cosmic rays \cite{Bell1978}, \cite{Blandford1978}.
 The process of acceleration on the shock front is nonlinear problem and it's efficiency depends on large number of factors.
@@
 The simulation is two dimensional, with three dimensional velocities and fields.
-Plasma flows into the simulation region through the right boundary, and reflects on the super-conducting wall at the left boundary.
+For numerical simulation we use Tristan-mp code, working with explicit scheme, developed bu Buneman \cite{Buneman}, and improved by Spitkovsky and others \cite{Sironi2011}.
+
+Shock wave is initialized by plasma flowing into the simulation region through the right boundary and reflecting on the super-conducting wall at the left boundary.
 We used following parameters for setup: the initial flow Lorentz factor $\gamma = 1.5$, magnetization $\sigma = \frac{B^2}{4\pi\gamma (n_p m_p + n_e m_e) c^2} = 0.04$, in turbulent case $B^2$ means mean square field.
@@
 \begin{equation}
-B_{turb} (x) = \sum_{i=0}^{n}\sum_{j=0}^{n}\sum_{p=1,2}B(k_x(i),k_y(j)) \textbf{e}_{p} sin(k_x(i) + k_y(j) + \phi (i,j,p))
+B_{turb} (x,y) = \sum_{i=0}^{n}\sum_{j=0}^{n}\sum_{p=1,2}B(k_x(i),k_y(j)) \textbf{e}_{p} sin(k_x(i) x + k_y(j) y  + \phi (i,j,p))
 \end{equation}
@@
 Wave vectors $k$ are distributed on uniform grid, $k_x = i \Delta k, k_y = j \Delta k$ and $\Delta k \approx \frac{2 \pi}{10 r_g}$ where $r_g$ is upstream proton gyroradius.
-And the maximum wavevector $k_{max} \approx \frac{2 \pi}{r_g}$.
+And the maximum wave vector $k_{max} \approx \frac{2 \pi}{r_g}$.
 This equations are evaluated in plasma rest frame and then magnetic field is transformed to the downstream frame.
 
-In this work we used setups with different turbulence scales, fraction of turbulent energy and regular field orientation and studied how this parameters influence on accelerated particles spectra.
+In this work we used setups with different turbulence scales, fraction of turbulent energy  and studied how this parameters influence on accelerated particles spectra.
 \section{Results}
@@
 In case of strong turbulence in upstream, in downstream also exists strong turbulence.
-Shock front moves slowly and seems to be coorugated.
+Shock front moves slowly and seems to be corrugated.
 
@@
 
+Also, spectrum of accelerated electrons depends on scale of turbulence.
+We tested setups with different maximum wave length of turbulent modes, corresponding to minimum wave vector $k_{min} = 2 \pi / 5 r_g$, $2 \pi / 10 r_g$ and $ 2 \pi /20 r_g$.
+For smaller scales acceleration efficiency is less then for larger, but this dependence is weak and with further growths of scales, spectrum changes insignificantly.
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=0.8\textwidth]{fig/spectrum_length.eps} 
+	\caption{Distribution of electrons in the relativistic shock wave with different turbulence length.}
+	\label{spectrum_length}
+\end{figure}
+
 \section{Conclusions}
@@
 	\bibitem{Blandford1978} Blandford R D and Ostriker J P 1978 \textit{ApJ} \textbf{221} L29 
+	\bibitem{Buneman} Buneman O 1993 \textit{Computer Space Plasma Physics} Terra Scientific Tokyo  67 p.
 	\bibitem{Bykov2014} Bykov A M, Ellison D C, Osipov S M and Vladimirov A E 2014 \textit{ApJ} \textbf{789} Issue 2, 137
-	\bibitem{Berezhko2003} Berezhko E G, Ksenofontov L T and V{\"o}lk H J  2003 \textit{A}{\&}\textit{A} \textbf{412} L11
-	\bibitem{Uchiyama2007} Uchiyama Y, Aharonian F A, Tanaka T, Takahashi T and Maeda Y 2007 \textit{Nature} \textbf{449} 576
-	\bibitem{Romansky2016} Romansky V I, Bykov A M, Osipov S M and Gladilin P E 2017 \textit{Journal of Physics: CS} \textbf{929} id 012014 
-	\bibitem{Lapenta2006} Lapenta G, Brackbill J U and Ricci P 2006 \textit{Phys. Plasmas} \textbf{13} 055904
-	\bibitem{Noguchi2007} Noguchi K, Tronci C, Zuccaro G and Lapenta G 2007 \textit{Phys. Plasmas} \textbf{14} 042308
-	\bibitem{Ellison2013} Ellison D C, Warren D C and Bykov A M 2013 \textit{ApJ} \textbf{776} Issue 1, 46
-	\bibitem{Pelletier2017} Pelletier G, Bykov A M, Ellison D C and Lemoine M 2017 \textit{Space Science Reviews} \textbf{207} Issue 1-4, pp. 319-360
-	\bibitem{Sironi2011} Sironi L and Spitkovsky A 2011 \textit{ApJ} \textbf{741} Issue 1, 39
-	\bibitem{Iwamoto2017} Iwamoto M, Amano T, Hoshino M and Matsumoto Y 2017 \textit{ApJ} \textbf{840} Issue 1 article id. 52, 14 pp.
+	\bibitem{Crumley} Crumley P, Caprioli D, Markoff S and Spitkovsky A 2018 \textit{MNRAS} \textbf{485} Issue 4, pp. 5105-5119
 	\bibitem{Margutti} Margutti R et al. 2014 \textit{ApJ}. \textbf{797} Issue 2, article id. 107 8 pp.
-	\bibitem{Buneman} Buneman O 1993 \textit{Computer Space Plasma Physics} Terra Scientific Tokyo  67 p.
-	\bibitem{Crumley} Crumley P, Caprioli D, Markoff S and Spitkovsky A \textit{MNRAS} \textbf{485} Issue 4, pp. 5105-5119
+	\bibitem{Romansky} Romansky V I, Bykov A M and Osipov S M 2018 \textit{Journal of Physics: CS} \textbf{1038} Issue 1, article id. 012022
+	\bibitem{Sironi2011} Sironi L and Spitkovsky A 2011 \textit{ApJ} \textbf{726} Issue 2, 75
+	\bibitem{Spitkovsky} Spitkovsky A 2005 \textit{AIP Conf. Ser.} \textbf{801},  pp 345-350
 

### 2019-04-11 papers/Physica2019/Romansky_Physica2019.tex
bibl

--- 2019-03-29 papers/Physica2019/Romansky_Physica2019.tex
@@
 
+\def\apj{ApJ}
+\def\mnras{MNRAS}
+\def\nat{Nat}
+\def\prd{Phys. Rev. D}
+\def\araa{ARA\&A}                % "Ann. Rev. Astron. Astrophys."
+\def\aap{A\&A}                   % "Astron. Astrophys."
+\def\aaps{A\&AS}                 % "Astron. Astrophys. Suppl. Ser."
+\def\aj{AJ}                      % "Astron. J."
+\def\apjs{ApJS}                  % "Astrophys. J. Suppl. Ser."
+\def\pasp{PASP}                  % "Publ. Astron. Soc. Pac."
+\def\apjl{ApJ}                   % letter at ApJ
+\def\pasj{PASJ}
+\def\apss{Astroph. Space Sci.}
+\def\aplett{Astroph. Lett}
+\def\ssr{Space Sci. Rev.}
+\def\aapr{Astron. Astroph. Reviews}
+\def\physrep{Phys. Reports}
+\def\memsai{Mem. Societa Astronom. Italiana}
+\def\jgr{JGR}
+
 \begin{document}
@@
 \address{$^1$ Ioffe Institute, 26 Politekhnicheskaya st., St. Petersburg 194021, Russia}
-\address{$^2$ Peter the Great St.~Petersburg Polytechnic University, 29 Politekhnicheskaya st., St. Petersburg 195251, Russia}
+\address{$^2$ Peter the Great St. Petersburg Polytechnic University, 29 Politekhnicheskaya st., St. Petersburg 195251, Russia}
 
@@
 \begin{abstract}
-	Modern observations with high angle resolution of radio emission in supernova remnants show that there exists type of objects, expanding with velocities close to speed of light.
-Model of electron acceleration is necessary for reasonable interpretation of this observations.
-In this paper we present numerical Particle-in-Cell simulation of relativistic shock wave propagating in turbulent medium.
+Radio and X-ray observations revealed a rare but a very interesting class of supernovae (SNe) with a sizeable fraction of the kinetic  energy of ejecta moving with a trans-relativistic speed.
+These relativistic SNe are comprising a population of the objects intermediate between the numerous core collapse SNe expanding with non-relativistic velocities and the gamma-ray bursts with highly relativistic ejecta.
+An interpretation of the observed non-thermal emission from relativistic SNe requires a model of electron acceleration in trans-relativistic shocks.
+In this paper we present numerical Particle-in-Cell (PIC) simulation of electron spectra in trans-relativistic shock waves propagating in clumped stellar winds of the SN progenitors.
+It is shown here that the presence of background magnetic fluctuations has a drastical effect on the electron acceleration by the trans-relativistic shocks propagating transverse to the regular magnetic field in the clumped wind of a massive progenitor star.
 \end{abstract}
@@
 
+Shock waves of young galactic supernova remnants (SNRs) are known for long time as the efficient electron accelerators producing synchrotron radiation observed from the radio to X-rays \cite{GS64,Helder12}.
+While most of the SNe both thermonuclear and core collapsed are ejecting most of their kinetic energy in the shell moving with typical speed below 10,000 km/s, there are a few SNe
+where apparently a sizeable part of the kinetic energy was in the form of trans-relativistic outflow with the 4-speed $\gamma \beta \sim$ 1.5.
+Extragalactic supernova SN2009bb \cite{2010Natur.463..513S}  apparently representing the class of relativistic SNe.
+These SNe are of especial interest since they may provide a connection between the bulk population of supernovae and the gamma-ray bursts and shed a light on the nature of the central engine producing the relativistic outflows \cite{Margutti2014,2016ApJ...832..108M}.
+To model the observed non-thermal radiation from SNe  \cite{1998ApJ...499..810C} the knowledge
+of relativistic electrons distribution is needed.
+
+Moreover, the relativistic SNe we proposed as the possible sources of very high energy CRs \cite{2007PhRvD..76h3009W,2011NatCo...2E.175C,2008ApJ...673..928B,2013ApJ...776...46E,BEMO18}.
+Diffusive shock acceleration \cite{Bell78}, \cite{Blandford78} in SNRs is considered as the most likely mechanism of cosmic ray production in a wide range of energies below EeV \cite{BEMO18}.
+The process of acceleration at the shock front is nonlinear problem and it's efficiency depends on a number of factors.
+In this work we studied the trans-relativistic shocks because they are probably the most effective particle accelerators.
+Energy that particle gains when crosses the shock front increases with shock velocity, but in ultra-relativistic shocks only small fraction of particles can return from downstream and efficiency is relatively low.
+In trans-relativistic shocks, with Lorentz-factor about 1.5, energy, gained in every crossing of the the front, is large but the accelerated particles  can still return from the downstream and they cross shock a few times.
+This was demonstrated in the case of quasi-parallel trans-relativistic shocks in Monte Carlo simulations \cite{2013ApJ...776...46E,BEMO18}.
+However, the regular magnetic fields in the winds of rotating stars at the distances of interest are usually dominated by the azimuthal field component.
+Therefore the SN shock waves propagating through the progenitor star winds are expected to be quasi-perpendicular.
+The quasi-perpendicular shock waves are found to be weak sources of accelerated particles in both  relativistic PIC simulations \cite{Sironi2011} as well as in the trans-relativistic cases \cite{Romansky18,Crumley2019}.
+In this paper we study the effect of the presence of  background magnetic field fluctuations in the clumped stellar wind on the electron acceleration by trans-relativistic shock propagating transverse to the regular magnetic field.
+
 \section{Numerical setup}
-We studied evolution of quasi-perpendicular trans-relativistic shock, propagating in turbulent medium, and spectra of accelerated particles.
-The simulation is two dimensional, with three dimensional velocities and fields.
-Plasma flows into the simulation region through the right boundary, and reflects on the super-conducting wall at the left boundary.
+We simulated a structure of a quasi-perpendicular trans-relativistic shock propagating in a clumped (turbulent) wind and the particle distribution function.
+The simulation is two dimensional with three dimensional velocities and fields.
+For numerical simulation we used Tristan-mp code with the explicit numerical scheme developed by Buneman \cite{Buneman93} and  Spitkovsky\cite{Spitkovsky2005}.
+
+Shock wave is initialized by plasma flowing into the simulation region through the right boundary and reflecting on the super-conducting wall at the left boundary.
 We used following parameters for setup: the initial flow Lorentz factor $\gamma = 1.5$, magnetization $\sigma = \frac{B^2}{4\pi\gamma (n_p m_p + n_e m_e) c^2} = 0.04$, in turbulent case $B^2$ means mean square field.
-The temperature $T = 5\cdot10^8 \rm{K}$ and proton mass is reduced to $m_p = 25 m_e$.
-Size of simulation region along x axis is $Lx = 8000\frac{c}{\omega_p}$ and in transverse direction $Ly = 200\frac{c}{\omega_p}$, where $\omega_p$ is plasma frequency $\omega_p = \sqrt{\frac{4\pi q^2 n}{\gamma m_e}}$.
+The dimensionless thermal energy $\Delta \gamma = \frac{k T}{m_p c^2}$ is equal $10^{-4}$ and proton mass is reduced to $m_p = 25 m_e$.
+Size of simulation region along x axis is $L_x = 8000\frac{c}{\omega_p}$ and in transverse direction $L_y = 200\frac{c}{\omega_p}$, where $\omega_p$ is plasma frequency $\omega_p = \sqrt{\frac{4\pi q^2 n}{\gamma m_e}}$.
 This sizes correspond to $80000$ and $2000$ grid points respectively.
-Also, $2000$ grid points corresponds to approximately $10$ gyroradii of upstream protons.
+Also, $2000$ grid points corresponds to approximately $10$ gyroradius of upstream protons.
 We initialize turbulent field via following formula: 
 \begin{equation}
-B_{turb} (x) = \sum_{i=0}^{n}\sum_{j=0}^{n}\sum_{p=1,2}B(k_x(i),k_y(j)) \textbf{e}_{p} sin(k_x(i) + k_y(j) + \phi (i,j,p))
+B_{turb} (x,y) = \sum_{i=0}^{n}\sum_{j=0}^{n}\sum_{p=1,2}B(k_x(i),k_y(j)) \textbf{e}_{p} sin(k_x(i) x + k_y(j) y  + \phi (i,j,p))
 \end{equation}
-where $B(k_x(i),k_y(j))$ is amplitude of turbulent mode, usually we choose $B \propto k^{-5/3}$ and normalized to fixed fraction of total magnetic energy, $\eta$.
-$\textbf{e}_{p}$ are vectors corresponding to two different field polarizations and $\phi (i,j,p)$ is randomly generated phase for each mode.
-Wave vectors $k$ are distributed on uniform grid, $k_x = i \Delta k, k_y = j \Delta k$ and $\Delta k \approx \frac{2 \pi}{10 r_g}$ where $r_g$ is upstream proton gyroradius.
-And the maximum wavevector $k_{max} \approx \frac{2 \pi}{r_g}$.
+where $B(k_x(i),k_y(j))$ is amplitude of turbulent mode, usually we choose $B \propto k^{-11/6}$ to imitate Kolmogorov's spectrum, where $B^2 k^2 \propto k^{-5/3}$ and normalized to fixed fraction of total magnetic energy, $\eta$.
+$\textbf{e}_{p}$ are vectors perpendicular to wave vector $\textbf{k}$ corresponding to two different field polarizations and $\phi (i,j,p)$ is randomly generated phase for each mode.
+Wave vectors $\textbf{k}$ are distributed on uniform grid, $k_x = i \Delta k, k_y = j \Delta k$ and $\Delta k \approx \frac{2 \pi}{10 r_g}$ where $r_g$ is upstream proton gyroradius.
+And the maximum wave vector $k_{max} \approx \frac{2 \pi}{r_g}$.
 This equations are evaluated in plasma rest frame and then magnetic field is transformed to the downstream frame.
+Regular magnetic field is perpendicular to the plane of simulation.
+Such turbulent field structure is similar to so-called 2d-turbulence which is observed in solar wind\cite{Matthaeus1990}.
+Density and velocity of plasma flow is uniform and electric field is set to compensate Lorentz force in every point.
 
-In this work we used setups with different turbulence scales, fraction of turbulent energy and regular field orientation and studied how this parameters influence on accelerated particles spectra.
+In this work we used setups with different magnetic turbulence scales and also vary the partial fraction of the energy density in the magnetic fluctuations to study how these parameters influence on the accelerated particles spectra.
+
 \section{Results}
+Simulation shows that structure of shock wave and velocity of it's propagation depends on properties of turbulence in upstream flow.
+In Figures \ref{regularB}, \ref{turbulentB} one can see the relation of magnetic field in shock wave to upstream regular field.
+In the case of strong turbulence in upstream, in downstream also exists strong turbulence.
+Shock front moves more slowly and seems to be corrugated (cf \cite{2016ApJ...827...44L}).
+
+\begin{figure}[h!]
+		\centering
+		\includegraphics[width=0.7\textwidth]{fig/regular_field.eps} 
+		\caption{Magnetic field in shock wave without turbulence.}
+		\label{regularB}
+\end{figure}
+\begin{figure}[h!]
+		\centering
+		\includegraphics[width=0.7\textwidth]{fig/turbulent_field.eps} 
+		\caption{Magnetic field in shock wave with 90\% turbulence.}
+		\label{turbulentB}
+\end{figure}
+Spectrum of accelerated electrons shows strong dependence on the fraction of turbulent magnetic field energy density in the wind, as shown in Figure \ref{spectrum}.
+When half of the magnetic field energy density is in magnetic turbulence the particle spectrum becomes significantly different from the spectrum in regular field.
+Fraction of accelerated electrons highly increases with growth of turbulent energy fraction.
+The quantity showing how much upstream flow kinetic energy goes to the accelerated electrons $\epsilon_e$ is defined as in paper \cite{Crumley2019}
+\begin{equation}
+\epsilon_e = \frac{\int_{p_{inj}}^{\infty}E(p)F(p)dp}{\int_{0}^{\infty}E(p)F(p)dp }\frac{m_e(\langle \gamma \rangle - 1)}{m_p (\gamma_0 - 1)}
+\end{equation}
+where $F(p)$ is electrons distribution function, $E(p)$ energy of particle with momentum $p$ and $\langle \gamma \rangle$ is average Lorentz-factor.
+$p_{inj}$ is injection momentum, showing that particle is non-thermal and we consider it equal to $2 \gamma_0 \beta_0 m_p c$.
+For simulation with the strongest turbulence, $\epsilon_e$ is equal to $2\cdot10^{-3}$, and it is several times more then for quasi-parallel shock wave.
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=0.7\textwidth]{fig/spectrum.eps} 
+	\caption{Distribution of electrons in the relativistic shock wave with different turbulence fractions.}
+	\label{spectrum}
+\end{figure}
+
+Also, the spectrum of accelerated electrons depends on the scales of magnetic turbulence.
+We tested setups with different maximum wave length of the magnetic turbulent modes, corresponding to minimum wave vector $k_{min} = 2 \pi / 5 r_g$, $2 \pi / 10 r_g$ and $ 2 \pi /20 r_g$.
+As it is shown in Figure \ref{spectrum_length} the acceleration efficiency is somewhat higher for larger scales of fluctuations, but this dependence is weak and the spectrum changes not very significantly.
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=0.7\textwidth]{fig/spectrum_length.eps} 
+	\caption{Distribution of electrons in the relativistic shock wave with different turbulence length.}
+	\label{spectrum_length}
+\end{figure}
 
 \section{Conclusions}
+In this paper we presented results of particle-in-cell simulation of electron acceleration in trans-relativistic shock propagating in a clumped  wind of a massive star.
+Simulations show that the presence  of background  magnetic turbulence in the clumped wind has strong impact on particle acceleration and the acceleration is efficient for high enough level of the magnetic turbulence.
+The electron acceleration is highly suppressed in the case of  trans-relativistic shock propagating quasi-perpendicular to a regular magnetic field in the wind without the background magnetic turbulence.
+However, with the PIC simulations we demonstrated here that the presence of magnetic fluctuations in the background wind results in efficient electron acceleration by the quasi-perpendicular trans-relativistic shocks.
+We found that the scales and amplitudes of the magnetic turbulent field are important for this process.
+Further PIC simulations with a wide dynamical range will allow to study the effect of turbulent wind on particle acceleration in detail.
 
 \ack
+Authors acknowledges support from RSF grant 16-12-10225.
 
-The numerical results presented here were obtained using computational resources of Peter the Great Saint-Petersburg Polytechnic University Supercomputing Center (http://www.scc.spbstu.ru).
+%\newpage
 
 \section*{References}
-\begin{thebibliography}{20}
-	\bibitem{Bell1978} Bell A R 1978 \textit{MNRAS} \textbf{182} 147
-	\bibitem{Blandford1978} Blandford R D and Ostriker J P 1978 \textit{ApJ} \textbf{221} L29 
-	\bibitem{Bykov2014} Bykov A M, Ellison D C, Osipov S M and Vladimirov A E 2014 \textit{ApJ} \textbf{789} Issue 2, 137
-	\bibitem{Berezhko2003} Berezhko E G, Ksenofontov L T and V{\"o}lk H J  2003 \textit{A}{\&}\textit{A} \textbf{412} L11
-	\bibitem{Uchiyama2007} Uchiyama Y, Aharonian F A, Tanaka T, Takahashi T and Maeda Y 2007 \textit{Nature} \textbf{449} 576
-	\bibitem{Romansky2016} Romansky V I, Bykov A M, Osipov S M and Gladilin P E 2017 \textit{Journal of Physics: CS} \textbf{929} id 012014 
-	\bibitem{Lapenta2006} Lapenta G, Brackbill J U and Ricci P 2006 \textit{Phys. Plasmas} \textbf{13} 055904
-	\bibitem{Noguchi2007} Noguchi K, Tronci C, Zuccaro G and Lapenta G 2007 \textit{Phys. Plasmas} \textbf{14} 042308
-	\bibitem{Ellison2013} Ellison D C, Warren D C and Bykov A M 2013 \textit{ApJ} \textbf{776} Issue 1, 46
-	\bibitem{Pelletier2017} Pelletier G, Bykov A M, Ellison D C and Lemoine M 2017 \textit{Space Science Reviews} \textbf{207} Issue 1-4, pp. 319-360
-	\bibitem{Sironi2011} Sironi L and Spitkovsky A 2011 \textit{ApJ} \textbf{741} Issue 1, 39
-	\bibitem{Iwamoto2017} Iwamoto M, Amano T, Hoshino M and Matsumoto Y 2017 \textit{ApJ} \textbf{840} Issue 1 article id. 52, 14 pp.
-\end{thebibliography}
+
+\bibliographystyle{iopart-num}
+\bibliography{bibliogr}
 \end{document}

### 2019-04-12 papers/Physica2019/Romansky_Physica2019.tex
fz

@@
 
-Shock waves of young galactic supernova remnants (SNRs) are known for long time as the efficient electron accelerators producing synchrotron radiation observed from the radio to X-rays \cite{GS64,Helder12}.
+Shock waves of young galactic supernova remnants (SNRs) are known for a long time as the efficient electron accelerators producing synchrotron radiation observed from the radio to X-rays \cite{GS64,Helder12}.
 While most of the SNe both thermonuclear and core collapsed are ejecting most of their kinetic energy in the shell moving with typical speed below 10,000 km/s, there are a few SNe
 where apparently a sizeable part of the kinetic energy was in the form of trans-relativistic outflow with the 4-speed $\gamma \beta \sim$ 1.5.
-Extragalactic supernova SN2009bb \cite{2010Natur.463..513S}  apparently representing the class of relativistic SNe.
+Extragalactic supernova SN2009bb \cite{2010Natur.463..513S}  apparently represents the class of relativistic SNe.
 These SNe are of especial interest since they may provide a connection between the bulk population of supernovae and the gamma-ray bursts and shed a light on the nature of the central engine producing the relativistic outflows \cite{Margutti2014,2016ApJ...832..108M}.
@@
 Energy that particle gains when crosses the shock front increases with shock velocity, but in ultra-relativistic shocks only small fraction of particles can return from downstream and efficiency is relatively low.
-In trans-relativistic shocks, with Lorentz-factor about 1.5, energy, gained in every crossing of the the front, is large but the accelerated particles  can still return from the downstream and they cross shock a few times.
+In trans-relativistic shocks, with Lorentz-factor about 1.5, energy, gained in each crossing of the front, is large however the accelerated particles  can still return from the downstream and cross shock several times.
 This was demonstrated in the case of quasi-parallel trans-relativistic shocks in Monte Carlo simulations \cite{2013ApJ...776...46E,BEMO18}.
 However, the regular magnetic fields in the winds of rotating stars at the distances of interest are usually dominated by the azimuthal field component.
-Therefore the SN shock waves propagating through the progenitor star winds are expected to be quasi-perpendicular.
-The quasi-perpendicular shock waves are found to be weak sources of accelerated particles in both  relativistic PIC simulations \cite{Sironi2011} as well as in the trans-relativistic cases \cite{Romansky18,Crumley2019}.
-In this paper we study the effect of the presence of  background magnetic field fluctuations in the clumped stellar wind on the electron acceleration by trans-relativistic shock propagating transverse to the regular magnetic field.
+Therefore, the SN shock waves propagating through the progenitor star winds are expected to be quasi-perpendicular.
+The quasi-perpendicular shock waves are found to be weak sources of accelerated particles in both  relativistic PIC simulations \cite{Sironi2011} and trans-relativistic cases \cite{Romansky18,Crumley2019}.
+In this paper we study the effect of the presence of  background magnetic field fluctuations in the clumped stellar wind on the electron acceleration by trans-relativistic shock propagating transversally to the regular magnetic field.
 
 \section{Numerical setup}
-We simulated a structure of a quasi-perpendicular trans-relativistic shock propagating in a clumped (turbulent) wind and the particle distribution function.
+We simulate a structure of a quasi-perpendicular trans-relativistic shock propagating in a clumped (turbulent) wind and the particle distribution.
 The simulation is two dimensional with three dimensional velocities and fields.
-For numerical simulation we used Tristan-mp code with the explicit numerical scheme developed by Buneman \cite{Buneman93} and  Spitkovsky\cite{Spitkovsky2005}.
+For numerical simulation we use Tristan-mp code with the explicit numerical scheme developed by Buneman \cite{Buneman93} and  Spitkovsky\cite{Spitkovsky2005}.
 
 Shock wave is initialized by plasma flowing into the simulation region through the right boundary and reflecting on the super-conducting wall at the left boundary.
-We used following parameters for setup: the initial flow Lorentz factor $\gamma = 1.5$, magnetization $\sigma = \frac{B^2}{4\pi\gamma (n_p m_p + n_e m_e) c^2} = 0.04$, in turbulent case $B^2$ means mean square field.
-The dimensionless thermal energy $\Delta \gamma = \frac{k T}{m_p c^2}$ is equal $10^{-4}$ and proton mass is reduced to $m_p = 25 m_e$.
-Size of simulation region along x axis is $L_x = 8000\frac{c}{\omega_p}$ and in transverse direction $L_y = 200\frac{c}{\omega_p}$, where $\omega_p$ is plasma frequency $\omega_p = \sqrt{\frac{4\pi q^2 n}{\gamma m_e}}$.
-This sizes correspond to $80000$ and $2000$ grid points respectively.
-Also, $2000$ grid points corresponds to approximately $10$ gyroradius of upstream protons.
-We initialize turbulent field via following formula: 
-\begin{equation}
+We use the following parameters for setup: the initial flow Lorentz factor $\gamma = 1.5$, magnetization $\sigma = \frac{B^2}{4\pi\gamma (n_p m_p + n_e m_e) c^2} = 0.04$ (in turbulent case $B^2$ is the mean square field).
+The dimensionless thermal energy $\Delta \gamma = \frac{k T}{m_p c^2}$ is equal $10^{-4}$ and the proton mass is reduced to $m_p = 25 m_e$.
+Size of the simulation region along x axis is $L_x = 8000\frac{c}{\omega_p}$ and in transverse direction $L_y = 200\frac{c}{\omega_p}$, where $\omega_p$ is the plasma frequency $\omega_p = \sqrt{\frac{4\pi q^2 n}{\gamma m_e}}$.
+These sizes correspond to $80000$ and $2000$ grid points respectively.
+Also, $2000$ grid points correspond to approximately $10$ gyroradii of upstream protons.
+We initialize turbulent field via the following formula: 
+\begin{equation}\label{field}
 B_{turb} (x,y) = \sum_{i=0}^{n}\sum_{j=0}^{n}\sum_{p=1,2}B(k_x(i),k_y(j)) \textbf{e}_{p} sin(k_x(i) x + k_y(j) y  + \phi (i,j,p))
 \end{equation}
-where $B(k_x(i),k_y(j))$ is amplitude of turbulent mode, usually we choose $B \propto k^{-11/6}$ to imitate Kolmogorov's spectrum, where $B^2 k^2 \propto k^{-5/3}$ and normalized to fixed fraction of total magnetic energy, $\eta$.
-$\textbf{e}_{p}$ are vectors perpendicular to wave vector $\textbf{k}$ corresponding to two different field polarizations and $\phi (i,j,p)$ is randomly generated phase for each mode.
-Wave vectors $\textbf{k}$ are distributed on uniform grid, $k_x = i \Delta k, k_y = j \Delta k$ and $\Delta k \approx \frac{2 \pi}{10 r_g}$ where $r_g$ is upstream proton gyroradius.
-And the maximum wave vector $k_{max} \approx \frac{2 \pi}{r_g}$.
-This equations are evaluated in plasma rest frame and then magnetic field is transformed to the downstream frame.
+where $B(k_x(i),k_y(j))$ is the amplitude of turbulent mode, usually we choose $B \propto k^{-11/6}$ to imitate Kolmogorov's spectrum, where $B^2 k^2 \propto k^{-5/3}$.
+$B$ is normalized to fixed fraction of total magnetic energy $\eta$.
+In further research with more wide spectrum of turbulence, it's behaviour may be important.
+$\textbf{e}_{p}$ are vectors perpendicular to wavevector $\textbf{k}$ corresponding to two different field polarizations and $\phi (i,j,p)$ is randomly generated phase for each mode.
+Wavevectors $\textbf{k}$ are distributed on a uniform grid, $k_x = i \Delta k, k_y = j \Delta k$ and $\Delta k \approx \frac{2 \pi}{10 r_g}$ where $r_g$ is upstream proton gyroradius.
+The maximum wavevector $k_{max} \approx \frac{2 \pi}{r_g}$.
+These equations (\ref{field}) are evaluated in the plasma rest frame and then magnetic field is transformed to the downstream frame.
 Regular magnetic field is perpendicular to the plane of simulation.
 Such turbulent field structure is similar to so-called 2d-turbulence which is observed in solar wind\cite{Matthaeus1990}.
-Density and velocity of plasma flow is uniform and electric field is set to compensate Lorentz force in every point.
+Density and velocity of plasma flow is uniform and the electric field is set to compensate Lorentz force in every point.
 
-In this work we used setups with different magnetic turbulence scales and also vary the partial fraction of the energy density in the magnetic fluctuations to study how these parameters influence on the accelerated particles spectra.
+In this work we use setups with different magnetic turbulence scales and also vary the partial fraction of the energy density in the magnetic fluctuations to study how these parameters influence on the accelerated particles spectra.
 
 \section{Results}
-Simulation shows that structure of shock wave and velocity of it's propagation depends on properties of turbulence in upstream flow.
+Simulation shows that structure of shock wave and velocity of it's propagation depend on the properties of turbulence in upstream flow.
 In Figures \ref{regularB}, \ref{turbulentB} one can see the relation of magnetic field in shock wave to upstream regular field.
-In the case of strong turbulence in upstream, in downstream also exists strong turbulence.
+In the case of strong turbulence in upstream, strong turbulence also exists in downstream.
 Shock front moves more slowly and seems to be corrugated (cf \cite{2016ApJ...827...44L}).
@@
 Spectrum of accelerated electrons shows strong dependence on the fraction of turbulent magnetic field energy density in the wind, as shown in Figure \ref{spectrum}.
-When half of the magnetic field energy density is in magnetic turbulence the particle spectrum becomes significantly different from the spectrum in regular field.
-Fraction of accelerated electrons highly increases with growth of turbulent energy fraction.
-The quantity showing how much upstream flow kinetic energy goes to the accelerated electrons $\epsilon_e$ is defined as in paper \cite{Crumley2019}
+When half of the magnetic field energy density is in a magnetic turbulence the particle spectrum becomes significantly different from the spectrum in a regular field.
+Fraction of accelerated electrons highly increases with the growth of turbulent energy fraction.
+The quantity showing how much of upstream flow kinetic energy goes to the accelerated electrons $\epsilon_e$ is defined as in \cite{Crumley2019}
 \begin{equation}
@@
 \end{equation}
-where $F(p)$ is electrons distribution function, $E(p)$ energy of particle with momentum $p$ and $\langle \gamma \rangle$ is average Lorentz-factor.
+where $F(p)$ is the electron distribution function, $E(p)$ is the energy of particle with momentum $p$ and $\langle \gamma \rangle$ is average Lorentz-factor.
 $p_{inj}$ is injection momentum, showing that particle is non-thermal and we consider it equal to $2 \gamma_0 \beta_0 m_p c$.
-For simulation with the strongest turbulence, $\epsilon_e$ is equal to $2\cdot10^{-3}$, and it is several times more then for quasi-parallel shock wave.
+For simulation with the strongest turbulence, $\epsilon_e$ is equal to $2\cdot10^{-3}$, and it is several times higher then for the quasi-parallel shock wave.
 \begin{figure}[h!]
@@
 
-Also, the spectrum of accelerated electrons depends on the scales of magnetic turbulence.
-We tested setups with different maximum wave length of the magnetic turbulent modes, corresponding to minimum wave vector $k_{min} = 2 \pi / 5 r_g$, $2 \pi / 10 r_g$ and $ 2 \pi /20 r_g$.
+The spectrum of accelerated electrons also depends on the scales of magnetic turbulence.
+We tested setups with different maximum wave length of the magnetic turbulent modes, corresponding to minimum wavevector $k_{min} = 2 \pi / 5 r_g$, $2 \pi / 10 r_g$ and $ 2 \pi /20 r_g$.
 As it is shown in Figure \ref{spectrum_length} the acceleration efficiency is somewhat higher for larger scales of fluctuations, but this dependence is weak and the spectrum changes not very significantly.
@@
 \section{Conclusions}
-In this paper we presented results of particle-in-cell simulation of electron acceleration in trans-relativistic shock propagating in a clumped  wind of a massive star.
+In this paper we present the results of particle-in-cell simulation of electron acceleration in trans-relativistic shock propagating in a clumped  wind of a massive star.
 Simulations show that the presence  of background  magnetic turbulence in the clumped wind has strong impact on particle acceleration and the acceleration is efficient for high enough level of the magnetic turbulence.
 The electron acceleration is highly suppressed in the case of  trans-relativistic shock propagating quasi-perpendicular to a regular magnetic field in the wind without the background magnetic turbulence.
-However, with the PIC simulations we demonstrated here that the presence of magnetic fluctuations in the background wind results in efficient electron acceleration by the quasi-perpendicular trans-relativistic shocks.
-We found that the scales and amplitudes of the magnetic turbulent field are important for this process.
+However, with the PIC simulations we demonstrate here that the presence of magnetic fluctuations in the background wind results in efficient electron acceleration by the quasi-perpendicular trans-relativistic shocks.
+We find that the scales and amplitudes of the magnetic turbulent field are important for this process.
 Further PIC simulations with a wide dynamical range will allow to study the effect of turbulent wind on particle acceleration in detail.
@@
 \ack
-Authors acknowledges support from RSF grant 16-12-10225.
+Authors acknowledge support from RSF grant 16-12-10225.
 

### 2019-05-20 papers/Physica2019/Romansky_Physica2019.tex
paper

@@
 
-Moreover, the relativistic SNe we proposed as the possible sources of very high energy CRs \cite{2007PhRvD..76h3009W,2011NatCo...2E.175C,2008ApJ...673..928B,2013ApJ...776...46E,BEMO18}.
+Moreover, the relativistic SNe were proposed as the possible sources of very high energy CRs \cite{2007PhRvD..76h3009W,2011NatCo...2E.175C,2008ApJ...673..928B,2013ApJ...776...46E,BEMO18}.
 Diffusive shock acceleration \cite{Bell78}, \cite{Blandford78} in SNRs is considered as the most likely mechanism of cosmic ray production in a wide range of energies below EeV \cite{BEMO18}.
@@
 This was demonstrated in the case of quasi-parallel trans-relativistic shocks in Monte Carlo simulations \cite{2013ApJ...776...46E,BEMO18}.
-However, the regular magnetic fields in the winds of rotating stars at the distances of interest are usually dominated by the azimuthal field component.
+However, the regular magnetic fields in the winds of the rotating stars at the distances of interest are usually dominated by the azimuthal field component.
 Therefore, the SN shock waves propagating through the progenitor star winds are expected to be quasi-perpendicular.
-The quasi-perpendicular shock waves are found to be weak sources of accelerated particles in both  relativistic PIC simulations \cite{Sironi2011} and trans-relativistic cases \cite{Romansky18,Crumley2019}.
-In this paper we study the effect of the presence of  background magnetic field fluctuations in the clumped stellar wind on the electron acceleration by trans-relativistic shock propagating transversally to the regular magnetic field.
+The quasi-perpendicular shock waves are found to be weak sources of the accelerated particles in both  relativistic PIC simulations \cite{Sironi2011} and trans-relativistic cases \cite{Romansky18,Crumley2019}.
+In this paper we study the effect of the presence of  the background magnetic field fluctuations in the clumped stellar wind on the electron acceleration by trans-relativistic shock propagating transversally to the regular magnetic field.
 
 \section{Numerical setup}
-We simulate a structure of a quasi-perpendicular trans-relativistic shock propagating in a clumped (turbulent) wind and the particle distribution.
-The simulation is two dimensional with three dimensional velocities and fields.
-For numerical simulation we use Tristan-mp code with the explicit numerical scheme developed by Buneman \cite{Buneman93} and  Spitkovsky\cite{Spitkovsky2005}.
+We simulated a structure of a quasi-perpendicular trans-relativistic shock propagating in a clumped (turbulent) wind and derived the particle distribution. 
+The simulation is two dimensional with three dimensional velocities and fields. For the numerical simulation we used Tristan-mp code with the explicit numerical
+ scheme developed by Buneman \cite{Buneman93} and  Spitkovsky\cite{Spitkovsky2005}.
 
-Shock wave is initialized by plasma flowing into the simulation region through the right boundary and reflecting on the super-conducting wall at the left boundary.
-We use the following parameters for setup: the initial flow Lorentz factor $\gamma = 1.5$, magnetization $\sigma = \frac{B^2}{4\pi\gamma (n_p m_p + n_e m_e) c^2} = 0.04$ (in turbulent case $B^2$ is the mean square field).
+Shock wave is initialized by an electron-proton plasma which is flowing into the simulation region through the right boundary and then reflecting from 
+the super-conducting wall at the left boundary.
+We use the following parameters for the setups: the initial flow Lorentz factor $\gamma = 1.5$, magnetization $\sigma = \frac{B^2}{4\pi\gamma (n_p m_p + n_e m_e) c^2} = 0.04$ (in the turbulent case $B^2$ is the mean square field).
 The dimensionless thermal energy $\Delta \gamma = \frac{k T}{m_p c^2}$ is equal $10^{-4}$ and the proton mass is reduced to $m_p = 25 m_e$.
@@
 Also, $2000$ grid points correspond to approximately $10$ gyroradii of upstream protons.
-We initialize turbulent field via the following formula: 
+We initialized turbulent field model by the following equation: 
 \begin{equation}\label{field}
@@
 
-In this work we use setups with different magnetic turbulence scales and also vary the partial fraction of the energy density in the magnetic fluctuations to study how these parameters influence on the accelerated particles spectra.
+In this paper we report the results for a few setups with different magnetic turbulence scales and we also vary the partial fraction of the energy density 
+in the magnetic fluctuations to study how these parameters influence on the accelerated particles spectra.
 
 \section{Results}
-Simulation shows that structure of shock wave and velocity of it's propagation depend on the properties of turbulence in upstream flow.
-In Figures \ref{regularB}, \ref{turbulentB} one can see the relation of magnetic field in shock wave to upstream regular field.
-In the case of strong turbulence in upstream, strong turbulence also exists in downstream.
-Shock front moves more slowly and seems to be corrugated (cf \cite{2016ApJ...827...44L}).
+The PIC simulations presented here  revealed that both the shock structure  and  it's propagation speed depend on the properties of turbulence in upstream flow.
+In Figures \ref{regularB}, \ref{turbulentB} one can see the relation of the magnetic field in shock wave to upstream regular field.
+In the case of strong turbulence in the shock upstream,
+strong turbulence also exists in downstream. The shock front  it is apparently corrugated (cf \cite{2016ApJ...827...44L}) and moves somewhat slower than in 
+the regular magnetic field.
 
@@
 \end{figure}
-Spectrum of accelerated electrons shows strong dependence on the fraction of turbulent magnetic field energy density in the wind, as shown in Figure \ref{spectrum}.
-When half of the magnetic field energy density is in a magnetic turbulence the particle spectrum becomes significantly different from the spectrum in a regular field.
-Fraction of accelerated electrons highly increases with the growth of turbulent energy fraction.
-The quantity showing how much of upstream flow kinetic energy goes to the accelerated electrons $\epsilon_e$ is defined as in \cite{Crumley2019}
+The spectrum of accelerated electrons  shown in Figure \ref{spectrum}  demonstrated a strong dependence on the fraction of the turbulent magnetic 
+field energy density in the wind. 
+When a half of the magnetic field energy density is in the magnetic turbulence the particle spectrum becomes significantly 
+different from the spectrum in a homogeneous field. A fraction of the accelerated electrons highly increases with the growth of turbulent energy fraction. 
+The quantity which is showing how much of the upstream flow kinetic energy (or the ram pressure) goes to the accelerated electrons 
+$\epsilon_e$ can be defined as that in \cite{Crumley2019}
 \begin{equation}
@@
 \end{equation}
-where $F(p)$ is the electron distribution function, $E(p)$ is the energy of particle with momentum $p$ and $\langle \gamma \rangle$ is average Lorentz-factor.
+where $F(p)$ is the electron distribution function, $E(p)$ is the energy of a particle of momentum $p$ and $\langle \gamma \rangle$ is the averaged Lorentz-factor. 
 $p_{inj}$ is injection momentum, showing that particle is non-thermal and we consider it equal to $2 \gamma_0 \beta_0 m_p c$.
-For simulation with the strongest turbulence, $\epsilon_e$ is equal to $2\cdot10^{-3}$, and it is several times higher then for the quasi-parallel shock wave.
+For simulation with the strongest turbulence, $\epsilon_e$ is equal to $2\cdot10^{-3}$, and it is several times higher then that in a quasi-parallel shock wave.
 \begin{figure}[h!]
@@
 The spectrum of accelerated electrons also depends on the scales of magnetic turbulence.
-We tested setups with different maximum wave length of the magnetic turbulent modes, corresponding to minimum wavevector $k_{min} = 2 \pi / 5 r_g$, $2 \pi / 10 r_g$ and $ 2 \pi /20 r_g$.
-As it is shown in Figure \ref{spectrum_length} the acceleration efficiency is somewhat higher for larger scales of fluctuations, but this dependence is weak and the spectrum changes not very significantly.
+We tested a few setups with different maximal wave length of the magnetic turbulent modes, corresponding to the minimal wavenumber $k_{min} = 2 \pi / 5 r_g$, $2 \pi / 10 r_g$ and $ 2 \pi /20 r_g$.
+As it is shown in Figure \ref{spectrum_length} the acceleration efficiency is somewhat higher for larger scales of fluctuations, but this dependence is weak and the spectrum changes are not very significant.
 
@@
 	\includegraphics[width=0.7\textwidth]{fig/spectrum_length.eps} 
-	\caption{Distribution of electrons in the relativistic shock wave with different turbulence length.}
+	\caption{Distribution of electrons in the relativistic shock wave with different turbulence length, turbulence energy fraction is 90\%.}
 	\label{spectrum_length}
@@
 \section{Conclusions}
-In this paper we present the results of particle-in-cell simulation of electron acceleration in trans-relativistic shock propagating in a clumped  wind of a massive star.
+In this paper we presented the results of particle-in-cell simulation of electron acceleration in trans-relativistic shock propagating in a clumped  wind of a massive star.
 Simulations show that the presence  of background  magnetic turbulence in the clumped wind has strong impact on particle acceleration and the acceleration is efficient for high enough level of the magnetic turbulence.
 The electron acceleration is highly suppressed in the case of  trans-relativistic shock propagating quasi-perpendicular to a regular magnetic field in the wind without the background magnetic turbulence.
-However, with the PIC simulations we demonstrate here that the presence of magnetic fluctuations in the background wind results in efficient electron acceleration by the quasi-perpendicular trans-relativistic shocks.
+With the PIC simulations we demonstrated here that the presence of magnetic fluctuations in the background wind results in efficient electron acceleration by the quasi-perpendicular trans-relativistic shocks.
 We find that the scales and amplitudes of the magnetic turbulent field are important for this process.
-Further PIC simulations with a wide dynamical range will allow to study the effect of turbulent wind on particle acceleration in detail.
+Further PIC simulations with a wider dynamical range will allow to study the effect of turbulent wind on particle acceleration in detail.
 
 \ack
-Authors acknowledge support from RSF grant 16-12-10225.
+The authors acknowledge a support from RSF grant 16-12-10225.
 

### 2019-05-21 papers/Physica2019/Romansky_Physica2019.tex
paper

@@
 
+We tested setup with a longer transverse size of the simulation region, corresponding to approximately $40$ gyroradii of the upstream protons, with the magnetic turbulence energy fraction 90\%.
+The spectrum of the accelerated electrons has no qualitative differences from the setup with a smaller transverse size since the non-thermal part of the distribution function survives.
+We also studied the effect of the proton to electron  mass ratio within a setup with  $\frac{m_p}{m_e} = 50$.
+The spectrum obtained has a similar dependence on the turbulence energy fraction and scale.
+It is shown in Figure \ref{electrons_mass} that the electron spectrum in the higher mass ratio setup is just shifted to the higher Lorentz-factors, while  the fraction of the energy in accelerated electrons  $\epsilon_e = 2.5\cdot10^{-3}$ is almost the same as in the case with $\frac{m_p}{m_e} = 25$.
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=0.7\textwidth]{fig/electrons_mass.eps} 
+	\caption{Distribution of electrons in the relativistic shock wave with different mass relation, turbulence energy fraction is 90\%.}
+	\label{electrons_mass}
+\end{figure}
+
 \section{Conclusions}

### 2019-08-21 Radiation/main.cpp
fz

@@ -2,55 +2,64 @@
 #include <stdlib.h>
 #include <time.h>
 #include "math.h"
+#include <string>
 
 const double kBoltzman = 1.3806488E-16;
 const double speed_of_light = 2.99792458E10;
-const double speed_of_light2 = speed_of_light*speed_of_light;
-const double speed_of_light4 = speed_of_light2*speed_of_light2;
+const double speed_of_light2 = speed_of_light * speed_of_light;
+const double speed_of_light4 = speed_of_light2 * speed_of_light2;
 const double electron_charge = 4.803529695E-10;
-const double pi = 4*atan2(1.0,1.0);
+const double pi = 4 * atan2(1.0, 1.0);
 const double massElectronReal = 0.910938291E-27;
 const double massProtonReal = 1.67262177E-24;
 const double massRatio = 25;
 //const double massElectron = massProtonReal/massRatio;
 const double massElectron = massElectronReal;
 
-const int Napprox  = 40;
+const int Napprox = 40;
 
-const double McDonaldValue[Napprox] = {3.08E8, 2.11E7, 6.65E6, 4.55E5, 1.43E4, 9802, 3087, 670, 211, 107, 66.3, 33.6, 20.7, 14.1, 11.0, 10.3, 6.26, 4.2, 3.01, 2.25, 1.73, 1.37, 1.10, 0.737, 0.514, 0.368, 0.269, 0.2, 0.0994, 0.0518, 0.0278, 0.0152, 0.00846, 0.00475, 0.00154, 0.000511, 0.000172, 0.0000589, 0.0000203, 0.00000246};
+const double McDonaldValue[Napprox] = {
+	3.08E8, 2.11E7, 6.65E6, 4.55E5, 1.43E4, 9802, 3087, 670, 211, 107, 66.3, 33.6, 20.7, 14.1, 11.0, 10.3, 6.26, 4.2, 3.01, 2.25, 1.73, 1.37, 1.10,
+	0.737, 0.514, 0.368, 0.269, 0.2, 0.0994, 0.0518, 0.0278, 0.0152, 0.00846, 0.00475, 0.00154, 0.000511, 0.000172, 0.0000589, 0.0000203, 0.00000246
+};
 const double UvarovValue[Napprox] = {
[…]

### 2020-04-10 papers/Physica2020/Romansky_Physica2020.tex
approx works for 90 turb

--- 2019-05-21 papers/Physica2019/Romansky_Physica2019.tex
@@
 \begin{document}
-\title{Modelling of electron acceleration in relativistic supernovae}
+\title{Particle-in-Cell modelling of synchrotron emission from relativistic supernovae}
 
@@
 We use the following parameters for the setups: the initial flow Lorentz factor $\gamma = 1.5$, magnetization $\sigma = \frac{B^2}{4\pi\gamma (n_p m_p + n_e m_e) c^2} = 0.04$ (in the turbulent case $B^2$ is the mean square field).
-The dimensionless thermal energy $\Delta \gamma = \frac{k T}{m_p c^2}$ is equal $10^{-4}$ and the proton mass is reduced to $m_p = 25 m_e$.
-Size of the simulation region along x axis is $L_x = 8000\frac{c}{\omega_p}$ and in transverse direction $L_y = 200\frac{c}{\omega_p}$, where $\omega_p$ is the plasma frequency $\omega_p = \sqrt{\frac{4\pi q^2 n}{\gamma m_e}}$.
-These sizes correspond to $80000$ and $2000$ grid points respectively.
+The dimensionless thermal energy $\Delta \gamma = \frac{k T}{m_p c^2}$ is equal $10^{-4}$ and the proton mass is reduced to $m_p = 100 m_e$.
+Size of the simulation region along x axis is $L_x = 30000\frac{c}{\omega_p}$ and in transverse direction $L_y = 400\frac{c}{\omega_p}$, where $\omega_p$ is the plasma frequency $\omega_p = \sqrt{\frac{4\pi q^2 n}{\gamma m_e}}$.
+These sizes correspond to $150000$ and $2000$ grid points respectively.
 Also, $2000$ grid points correspond to approximately $10$ gyroradii of upstream protons.
@@
 \begin{equation}\label{field}
-B_{turb} (x,y) = \sum_{i=0}^{n}\sum_{j=0}^{n}\sum_{p=1,2}B(k_x(i),k_y(j)) \textbf{e}_{p} sin(k_x(i) x + k_y(j) y  + \phi (i,j,p))
+B_{turb} (\textbf{r}) = \sum_{\textbf{k}}\sum_{p=1,2}B(\textbf{k}) \textbf{e}_{p} sin(\textbf{k}\textbf{r}  + \phi (\textbf{k},p))
 \end{equation}
@@
 
+\section{Synchrotron spectrum evaluation}
+We evaluate spectrum of synchrotron emission using standart formula \cite{Syrovatsky} and take into account self-absorption\cite{Ghisellini1991}.
+Emissivity at the frequency $\nu$ of unit volume is given by
+\begin{equation}\label{emissivity}
+I_{\nu}={\int_{m_e c^2}^{\infty}}\frac{\sqrt{3} e^3}{m_e c^2}B_{\perp}F_e(E)\frac{\nu}{{\nu}_c}{\int_{\frac{\nu}{{\nu}_c}}^{\infty}}K_{\frac{5}{3}}({\eta})d{\eta} dE
+\end{equation}
+Where $B_{\perp}$ is component of magnetic field, normal to the observer direction, $F_e(E)$ is electrons' distribution function, normalized on concentration, ${\nu}_c$ is critical frequency defined as $\frac{3 e B_{\perp}}{4 \pi {\gamma}^2 m_e c}$, $\gamma$ is Lorentz-factor and $K_{\frac{5}{3}}$ is modified Bessel function of second kind.
+
+Self-absorption cross-section is defined as
+\begin{equation}\label{selfabsorption}
+{\sigma}(\nu, \gamma) = \frac{16 \pi^2}{3\sqrt{3}}\frac{e}{B_{\perp}}\frac{1}{\gamma^5}K_{\frac{5}{3}}(\frac{\nu}{{\nu}_c})
+\end{equation}
+
 \section{Results}
@@
 		\centering
-		\includegraphics[width=0.7\textwidth]{fig/regular_field.eps} 
+		%\includegraphics[width=0.7\textwidth]{fig/regular_field.eps} 
 		\caption{Magnetic field in shock wave without turbulence.}
@@
 		\centering
-		\includegraphics[width=0.7\textwidth]{fig/turbulent_field.eps} 
+		%\includegraphics[width=0.7\textwidth]{fig/turbulent_field.eps} 
 		\caption{Magnetic field in shock wave with 90\% turbulence.}
@@
 	\centering
-	\includegraphics[width=0.7\textwidth]{fig/spectrum.eps} 
+	%\includegraphics[width=0.7\textwidth]{fig/spectrum.eps} 
 	\caption{Distribution of electrons in the relativistic shock wave with different turbulence fractions.}
@@
 	\centering
-	\includegraphics[width=0.7\textwidth]{fig/spectrum_length.eps} 
+	%\includegraphics[width=0.7\textwidth]{fig/spectrum_length.eps} 
 	\caption{Distribution of electrons in the relativistic shock wave with different turbulence length, turbulence energy fraction is 90\%.}
@@
 	\centering
-	\includegraphics[width=0.7\textwidth]{fig/electrons_mass.eps} 
+	%\includegraphics[width=0.7\textwidth]{fig/electrons_mass.eps} 
 	\caption{Distribution of electrons in the relativistic shock wave with different mass relation, turbulence energy fraction is 90\%.}

### 2020-04-16 papers/Physica2020/Romansky_Physica2020.tex
PAPER

--- 2019-05-20 papers/Physica2019/Romansky_Physica2019.tex
@@
 \begin{document}
-\title{Modelling of electron acceleration in relativistic supernovae}
+\title{Evaluating MHD parameters of relativistic shock waves with Particle-in-Cell modelling}
 
@@
 \begin{abstract}
-Radio and X-ray observations revealed a rare but a very interesting class of supernovae (SNe) with a sizeable fraction of the kinetic  energy of ejecta moving with a trans-relativistic speed.
-These relativistic SNe are comprising a population of the objects intermediate between the numerous core collapse SNe expanding with non-relativistic velocities and the gamma-ray bursts with highly relativistic ejecta.
-An interpretation of the observed non-thermal emission from relativistic SNe requires a model of electron acceleration in trans-relativistic shocks.
-In this paper we present numerical Particle-in-Cell (PIC) simulation of electron spectra in trans-relativistic shock waves propagating in clumped stellar winds of the SN progenitors.
-It is shown here that the presence of background magnetic fluctuations has a drastical effect on the electron acceleration by the trans-relativistic shocks propagating transverse to the regular magnetic field in the clumped wind of a massive progenitor star.
+Reklativistic shock waves are widely discussed phenomenae of modern astrophysics.
+Studying of their properties is necessary for consistent models of supernova remnants, jets and other astrophysical objects.
+But magneto-hydrodinamical description of relativistic shock waves is not full, because it does not take into account electrons and a lot of relativistic effects such as displacement current density and changing of adiabatic index That's why we need more accurate methods to check and fix MHD models.
 \end{abstract}
 \section{Introduction}
+Relativistic shock jump (Rankine-Hugoniot) were given by Blandford and McKee \cite{Blandford76} and generalized to MHD by Emmering and Chevalier\cite{Emmering87}. Shock conditions are:
+\begin{equation}\label{hugoniot}
+[\![\rho u^r]\!] = [\![ (w + \frac{B^2}{4 \pi})u^r u^t]\!] = [\![(w + \frac{B^2}{4 \pi}){u^r}^2 + P + \frac{B^2}{8 \pi}]\!] = [\![\frac{B}{\rho}]\!] = 0
+\end{equation}
+where $u = (u^t,u^r)$ is velocity four-vector measured in the shock frame, $\rho$ is proper mass density, $B$ is magnetic field and $P$ is pressure in the fluid frame and $w = \rho c^2 + \frac{\hat{\gamma}}{\hat{\gamma} - 1} P$ is enthalpy, $\hat{\gamma}$ is adiabatic index and double brackets indicate the mismatches of quantities to both sides of the shock.
 
-Shock waves of young galactic supernova remnants (SNRs) are known for a long time as the efficient electron accelerators producing synchrotron radiation observed from the radio to X-rays \cite{GS64,Helder12}.
-While most of the SNe both thermonuclear and core collapsed are ejecting most of their kinetic energy in the shell moving with typical speed below 10,000 km/s, there are a few SNe
-where apparently a sizeable part of the kinetic energy was in the form of trans-relativistic outflow with the 4-speed $\gamma \beta \sim$ 1.5.
-Extragalactic supernova SN2009bb \cite{2010Natur.463..513S}  apparently represents the class of relativistic SNe.
-These SNe are of especial interest since they may provide a connection between the bulk population of supernovae and the gamma-ray bursts and shed a light on the nature of the central engine producing the relativistic outflows \cite{Margutti2014,2016ApJ...832..108M}.
-To model the observed non-thermal radiation from SNe  \cite{1998ApJ...499..810C} the knowledge
-of relativistic electrons distribution is needed.
+In more simple case of strictly perpendicular shock, shock velocity and wownstream temperature can be expressed\cite{Amato2006} as
+\begin{equation}\label{vshock}
+v_{sh} = \frac{c}{2}\frac{1}{1 + \sigma}\lbrace (\hat{\gamma}(1 + \frac{\sigma}{2})-1) + {\lbrack (\hat{\gamma}(1 + \frac{\sigma}{2})-1)^2 + 4\sigma(1+\sigma)(1 - \frac{\hat{\gamma}}{2})\rbrack}^{\frac{1}{2}} \rbrace
+\end{equation}
+\begin{equation}\label{temperature}
+T = m c^2 \Gamma \frac{v_{sh}}{c}(1 - \frac{\sigma}{2}\frac{c - v_{sh}}{v_sh})
+\end{equation}
+Here $\sigma$ is magnetisation and $\Gamma$ is upstream lorentz factor measured in downstream frame.
 
-Moreover, the relativistic SNe were proposed as the possible sources of very high energy CRs \cite{2007PhRvD..76h3009W,2011NatCo...2E.175C,2008ApJ...673..928B,2013ApJ...776...46E,BEMO18}.
-Diffusive shock acceleration \cite{Bell78}, \cite{Blandford78} in SNRs is considered as the most likely mechanism of cosmic ray production in a wide range of energies below EeV \cite{BEMO18}.
-The process of acceleration at the shock front is nonlinear problem and it's efficiency depends on a number of factors.
-In this work we studied the trans-relativistic shocks because they are probably the most effective particle accelerators.
-Energy that particle gains when crosses the shock front increases with shock velocity, but in ultra-relativistic shocks only small fraction of particles can return from downstream and efficiency is relatively low.
-In trans-relativistic shocks, with Lorentz-factor about 1.5, energy, gained in each crossing of the front, is large however the accelerated particles  can still return from the downstream and cross shock several times.
-This was demonstrated in the case of quasi-parallel trans-relativistic shocks in Monte Carlo simulations \cite{2013ApJ...776...46E,BEMO18}.
-However, the regular magnetic fields in the winds of the rotating stars at the distances of interest are usually dominated by the azimuthal field component.
-Therefore, the SN shock waves propagating through the progenitor star winds are expected to be quasi-perpendicular.
-The quasi-perpendicular shock waves are found to be weak sources of the accelerated particles in both  relativistic PIC simulations \cite{Sironi2011} and trans-relativistic cases \cite{Romansky18,Crumley2019}.
-In this paper we study the effect of the presence of  the background magnetic field fluctuations in the clumped stellar wind on the electron acceleration by trans-relativistic shock propagating transversally to the regular magnetic field.
+One weak point of this formalism is adiabatic index $\gamma$.
+For monoatomic gas it is well-defined in classical limit $\gamma = \frac{5}{3}$ and in ultra-relativistic linit $\gamma = \frac{4}{3}$.
+In the gap between them it is a non-trivial question which value of $\gamma$ to choose.
+Another problem is that one-fluid model doesn't take into account the difference between electrons' and protons' temperature.
+And two-fluid model also use some phenomenological methods to evaluate electrons' temperautre.
+
+In this work we used Particle-in-Cell method for modelling relativistic shock waves to study accuracy of MHD equations and to determine adiabatic index.
 
 \section{Numerical setup}
-We simulated a structure of a quasi-perpendicular trans-relativistic shock propagating in a clumped (turbulent) wind and derived the particle distribution. 
-The simulation is two dimensional with three dimensional velocities and fields. For the numerical simulation we used Tristan-mp code with the explicit numerical
- scheme developed by Buneman \cite{Buneman93} and  Spitkovsky\cite{Spitkovsky2005}.
+We simulated a structure of a perpendicular relativistic shock and varied magnetisation, lorentz-factor and fraction of turbulent field. 
+The simulation is two dimensional with three dimensional velocities and fields.
+For the numerical simulation we used Tristan-mp code with the explicit numerical scheme developed by Buneman \cite{Buneman93} and  Spitkovsky\cite{Spitkovsky2005}.
 
-Shock wave is initialized by an electron-proton plasma which is flowing into the simulation region through the right boundary and then reflecting from 
-the super-conducting wall at the left boundary.
-We use the following parameters for the setups: the initial flow Lorentz factor $\gamma = 1.5$, magnetization $\sigma = \frac{B^2}{4\pi\gamma (n_p m_p + n_e m_e) c^2} = 0.04$ (in the turbulent case $B^2$ is the mean square field).
-The dimensionless thermal energy $\Delta \gamma = \frac{k T}{m_p c^2}$ is equal $10^{-4}$ and the proton mass is reduced to $m_p = 25 m_e$.
-Size of the simulation region along x axis is $L_x = 8000\frac{c}{\omega_p}$ and in transverse direction $L_y = 200\frac{c}{\omega_p}$, where $\omega_p$ is the plasma frequency $\omega_p = \sqrt{\frac{4\pi q^2 n}{\gamma m_e}}$.
-These sizes correspond to $80000$ and $2000$ grid points respectively.
+Shock wave is initialized by an electron-proton plasma which is flowing into the simulation region through the right boundary and then reflecting from the super-conducting wall at the left boundary.
+We use the following parameters for the setups: the initial flow Lorentz factor $\Gamma = 1.5$, magnetization $\sigma = \frac{B^2}{4\pi\Gamma (n_p m_p + n_e m_e) c^2} = 0.04$ (in the turbulent case $B^2$ is the mean square field).
+The dimensionless thermal energy $\Delta \gamma = \frac{k T}{m_p c^2}$ is equal $10^{-4}$ and the proton mass is reduced to $m_p = 100 m_e$.
+Size of the simulation region along x axis is $L_x = 30000\frac{c}{\omega_p}$ and in transverse direction $L_y = 400\frac{c}{\omega_p}$, where $\omega_p$ is the plasma frequency $\omega_p = \sqrt{\frac{4\pi q^2 n}{\Gamma m_e}}$.
+These sizes correspond to $150000$ and $2000$ grid points respectively.
 Also, $2000$ grid points correspond to approximately $10$ gyroradii of upstream protons.
-We initialized turbulent field model by the following equation: 
-\begin{equation}\label{field}
-B_{turb} (x,y) = \sum_{i=0}^{n}\sum_{j=0}^{n}\sum_{p=1,2}B(k_x(i),k_y(j)) \textbf{e}_{p} sin(k_x(i) x + k_y(j) y  + \phi (i,j,p))
-\end{equation}
-where $B(k_x(i),k_y(j))$ is the amplitude of turbulent mode, usually we choose $B \propto k^{-11/6}$ to imitate Kolmogorov's spectrum, where $B^2 k^2 \propto k^{-5/3}$.
-$B$ is normalized to fixed fraction of total magnetic energy $\eta$.
-In further research with more wide spectrum of turbulence, it's behaviour may be important.
-$\textbf{e}_{p}$ are vectors perpendicular to wavevector $\textbf{k}$ corresponding to two different field polarizations and $\phi (i,j,p)$ is randomly generated phase for each mode.
-Wavevectors $\textbf{k}$ are distributed on a uniform grid, $k_x = i \Delta k, k_y = j \Delta k$ and $\Delta k \approx \frac{2 \pi}{10 r_g}$ where $r_g$ is upstream proton gyroradius.
-The maximum wavevector $k_{max} \approx \frac{2 \pi}{r_g}$.
-These equations (\ref{field}) are evaluated in the plasma rest frame and then magnetic field is transformed to the downstream frame.
-Regular magnetic field is perpendicular to the plane of simulation.
-Such turbulent field structure is similar to so-called 2d-turbulence which is observed in solar wind\cite{Matthaeus1990}.
-Density and velocity of plasma flow is uniform and the electric field is set to compensate Lorentz force in every point.
-
-In this paper we report the results for a few setups with different magnetic turbulence scales and we also vary the partial fraction of the energy density 
-in the magnetic fluctuations to study how these parameters influence on the accelerated particles spectra.
+In simulation with turbulent fields, we initialized it as summ of harmonic modes with Kolmogorov's spectum, and it is described in our previous paper in detail\cite{Romansky2019}.
+The energy density of turbulent field compared to the total magnetic energy density is $\eta$ and varying from 0 percent to 90 percent.
 
 \section{Results}
-The PIC simulations presented here  revealed that both the shock structure  and  it's propagation speed depend on the properties of turbulence in upstream flow.
-In Figures \ref{regularB}, \ref{turbulentB} one can see the relation of the magnetic field in shock wave to upstream regular field.
-In the case of strong turbulence in the shock upstream,
-strong turbulence also exists in downstream. The shock front  it is apparently corrugated (cf \cite{2016ApJ...827...44L}) and moves somewhat slower than in 
-the regular magnetic field.
+Using the density profiles of shock wave, we derived magnetohydrodynamic parameters of plasma, using formulae \ref{vshock} and \ref{temperature}.
+Density, averaged in transversal directions  has clear jump as shown on figures \ref{density_noturb} and \ref{density_turb}, indicating the coordinate of shock front.
 
@@
 	\centering
-		\includegraphics[width=0.7\textwidth]{fig/regular_field.eps} 
-		\caption{Magnetic field in shock wave without turbulence.}
-		\label{regularB}
-\end{figure}
-\begin{figure}[h!]
+	\begin{minipage}{0.5\textwidth}
 		\centering
-		\includegraphics[width=0.7\textwidth]{fig/turbulent_field.eps} 
-		\caption{Magnetic field in shock wave with 90\% turbulence.}
-		\label{turbulentB}
-\end{figure}
-The spectrum of accelerated electrons  shown in Figure \ref{spectrum}  demonstrated a strong dependence on the fraction of the turbulent magnetic 
-field energy density in the wind. 
-When a half of the magnetic field energy density is in the magnetic turbulence the particle spectrum becomes significantly 
-different from the spectrum in a homogeneous field. A fraction of the accelerated electrons highly increases with the growth of turbulent energy fraction. 
-The quantity which is showing how much of the upstream flow kinetic energy (or the ram pressure) goes to the accelerated electrons 
-$\epsilon_e$ can be defined as that in \cite{Crumley2019}
-\begin{equation}
-\epsilon_e = \frac{\int_{p_{inj}}^{\infty}E(p)F(p)dp}{\int_{0}^{\infty}E(p)F(p)dp }\frac{m_e(\langle \gamma \rangle - 1)}{m_p (\gamma_0 - 1)}
-\end{equation}
-where $F(p)$ is the electron distribution function, $E(p)$ is the energy of a particle of momentum $p$ and $\langle \gamma \rangle$ is the averaged Lorentz-factor. 
-$p_{inj}$ is injection momentum, showing that particle is non-thermal and we consider it equal to $2 \gamma_0 \beta_0 m_p c$.
-For simulation with the strongest turbulence, $\epsilon_e$ is equal to $2\cdot10^{-3}$, and it is several times higher then that in a quasi-parallel shock wave.
-\begin{figure}[h!]
+		\includegraphics[width=0.98\textwidth]{fig/density_noturb.png} 
+		\caption{Averaged by y dimension density profile of shock wave.}
+		\label{density_noturb}
+	\end{minipage}\hfill
+	\begin{minipage}{0.5\textwidth}
 		\centering
-	\includegraphics[width=0.7\textwidth]{fig/spectrum.eps} 
-	\caption{Distribution of electrons in the relativistic shock wave with different turbulence fractions.}
-	\label{spectrum}
+		\includegraphics[width=0.98\textwidth]{fig/density_turb.png} 
+		\caption{Averaged by y dimension density profile of shock wave in turbulent medium.}
+		\label{density_turb}
+	\end{minipage}
 \end{figure}
 
-The spectrum of accelerated electrons also depends on the scales of magnetic turbulence.
-We tested a few setups with different maximal wave length of the magnetic turbulent modes, corresponding to the minimal wavenumber $k_{min} = 2 \pi / 5 r_g$, $2 \pi / 10 r_g$ and $ 2 \pi /20 r_g$.
-As it is shown in Figure \ref{spectrum_length} the acceleration efficiency is somewhat higher for larger scales of fluctuations, but this dependence is weak and the spectrum changes are not very significant.
+Evalation of shock front coordinates at different time moments shows that shock front velocity is constant with high accuracy in the case without turbulent field and has small oscillations without turbulence, see figure\ref{shockx}.
+\begin{table}
+	\label{setups}
+	\begin{center}
+		\begin{tabular}{|c| c| c| c| c| c| c| c| c| c| c| c|}
+		\hline
+			Setup & Nx & Ny & Nz & $\sigma$ & $\Gamma$ & $\eta$ & $\beta$ & $\hat{\gamma}$ & $T  10^{12} K$ & $T_p  10^{12} K$ & $T_e  10^{12} K$\\
+			reg1 & 150000 & 2000 & 1 & 0.04 & 1.5 & 0 & 0.394 & 1.349 & 6.38 & 1 & 1\\
+			reg2 & 150000 & 2000 & 1 & 0.004 & 1.5 & 0 & 0.355 & 1.35 & 5.79 & 1 & 1\\
+			reg3 & 150000 & 2000 & 1 & 0.0004 & 1.5 & 0 & 0.339 & 1.338 & 5.53 &1 & 1\\
+			turb1 & 150000 & 2000 & 1 & 0.04 & 1.5 & 90 & 0.309 & 1.247 & 5.0 & 1 & 1\\
+			turb2 & 150000 & 2000 & 1 & 0.04 & 1.5 & 30 & 0.356 & 1.305 & 5.77 & 1 & 1\\
+			turb3 & 150000 & 2000 & 1 & 0.004 & 1.5 & 90 & 0.309 & 1.303 & 5.03 & 1 & 1\\
+			turb4 & 150000 & 2000 & 1 & 0.004 & 1.5 & 30 & 0.348 & 1.343 & 5.67 & 1 & 1\\
+			gam1 & 150000 & 2000 & 1 & 0.04 & 2 & 0 & 0.447 & 1.411 & 9.67 & 1 & 1\\
+			gam2 & 150000 & 2000 & 1 & 0.04 & 5 & 0 & 0.505 & 1.475 & 27.3 & 1 & 1\\
+			gam3 & 150000 & 2000 & 1 & 0.04 & 10 & 0 & 0.528 & 1.499 & 57.1 & 1 & 1\\
+			reg3d & 80000 & 40 & 40 & 0.04 & 1.5 & 0 & 0.447 & 1.43 & 1 & 1 & 1\\
+
+		\hline
+		\end{tabular}
+	\end{center}
+	\caption{Parameters of different setups}
+\end{table}
 
@@
 	\centering
-	\includegraphics[width=0.7\textwidth]{fig/spectrum_length.eps} 
-	\caption{Distribution of electrons in the relativistic shock wave with different turbulence length, turbulence energy fraction is 90\%.}
-	\label{spectrum_length}
+	\includegraphics[width=0.6\textwidth]{fig/shock_x.png} 
+	\caption{Coordinate of shock front for different setups.}
+	\label{shock_x}
 \end{figure}

### 2020-04-22 papers/Physica2020/Romansky_Physica2020.tex
spherical turbulence

@@
 \begin{document}
-\title{Evaluating MHD parameters of relativistic shock waves with Particle-in-Cell modelling}
+	\title{Evaluating MHD parameters of relativistic shock waves with particle-in-cell modeling}
 
-\author{V I Romansky$^{1}$, A M Bykov$^{1,2}$ and S M Osipov$^1$}
+	\author{V I Romansky, A M Bykov and S M Osipov}
 
-\address{$^1$ Ioffe Institute, 26 Politekhnicheskaya st., St. Petersburg 194021, Russia}
-\address{$^2$ Peter the Great St. Petersburg Polytechnic University, 29 Politekhnicheskaya st., St. Petersburg 195251, Russia}
+	\address{Ioffe Institute, 26 Politekhnicheskaya st., St. Petersburg 194021, Russia}
 
@@
 	\begin{abstract}
-Reklativistic shock waves are widely discussed phenomenae of modern astrophysics.
-Studying of their properties is necessary for consistent models of supernova remnants, jets and other astrophysical objects.
-But magneto-hydrodinamical description of relativistic shock waves is not full, because it does not take into account electrons and a lot of relativistic effects such as displacement current density and changing of adiabatic index That's why we need more accurate methods to check and fix MHD models.
+		Relativistic plasma outflows are observed in gamma-ray burst
+		sources, in jets of active galactic nuclei, in pulsar wind nebulae and
+		supernovae explosions. Magnetohydrodynamical (MHD) shock waves
+		inevitably result from interactions of such relativistic outflows with
+		the ambient interstellar matter. The widely used single-fluid MHD
+		description of relativistic shock waves is a useful tool to study the
+		global structure of such objects. However to interpret the observed
+		emission spectra of space objects with relativistic shocks, a kinetic
+		description of electrons, positrons, and ions is needed. The kinetic
+		structure of relativistic shocks can be modelled within the
+		particle-in-cell approach, which allows one to properly account for
+		the electron scale processes and for the density of the displacement
+		current. Here we discuss the effect of the micro-scale plasma processes on such
+		macroscopic parameters of relativistic shocks as the adiabatic index
+		of the relativistic fluid in the shock vicinity, with a relatively
+		long equilibration time. The adiabatic index is an important parameter of the
+		single-fluid MHD models commonly used for shock modeling at much
+		longer hydrodynamical scales.
 	\end{abstract}
+
 	\section{Introduction}
-Relativistic shock jump (Rankine-Hugoniot) were given by Blandford and McKee \cite{Blandford76} and generalized to MHD by Emmering and Chevalier\cite{Emmering87}. Shock conditions are:
+	Relativistic jets and winds are the generic outflows of the relativistic supernovae \cite{2010Natur.463..513S,2007ApJ...667..351W},the  accreting black holes, both suprmassive in the active galactic nucleis \cite{1984RvMP...56..255B} as well as the stellar masses in the gamma-ray burst sources, microquasars \cite{2019MmSAI..90...57M,1999PhR...314..575P,2014LNP...876.....R} and young pulsars \cite{2017SSRv..207....1B}.
+Relativistic shocks can be formed either inside the flow of low or moderate magnetization (the inner shocks) or can be driven by the outflow colliding with the ambient medium (the external shocks).
+In most cases the relatistivistic shocks in astrophysical objects are collisionless are associated with strongly non-equilibrium particle distributions
+	\cite{2012SSRv..173..309B,2015SSRv..191..519S,2017SSRv..207..319P}, and they can accelerate particles to ultra-high energies \cite{2009JCAP...11..009L}.
+The presence of non-maxwellian components with very long equilibration time requires a kinetic description of the shocked flows.
+On the other hand, available computer facilities do not alow one to provide a fully kinetic modelling of the macroscopic astrophysical flows and therefore fluid MHD description based on some macroscopic parameters like the effective adiabatic index of the flow is widely used.
+Here we discuss an approach allowing to derive the effective adiabatic indexes from the microscopical particle-in-cell modeling of the relativistic shocks.
+
+	Relativistic fluid shock dynamics was discussed by Lichnerowicz \cite{1967rhm..book.....L}, Blandford and McKee \cite{Blandford76} and by Emmering and Chevalier\cite{Emmering87}.
+The relativistic shock jump conditions, in the case of a perpendicular shock which is typical for relativisic shocks, can be written as:
 	\begin{equation}\label{hugoniot}
@@
 	\end{equation}
-where $u = (u^t,u^r)$ is velocity four-vector measured in the shock frame, $\rho$ is proper mass density, $B$ is magnetic field and $P$ is pressure in the fluid frame and $w = \rho c^2 + \frac{\hat{\gamma}}{\hat{\gamma} - 1} P$ is enthalpy, $\hat{\gamma}$ is adiabatic index and double brackets indicate the mismatches of quantities to both sides of the shock.
+	where $u = (u^t,u^r)$ is the velocity four-vector measured in the shock frame, $\rho$ is the proper mass density, $B$ is the magnetic field, $P$ is the pressure in the fluid frame, and $w = \rho c^2 + \frac{\hat{\gamma}}{\hat{\gamma} - 1} P$ is the enthalpy, while $\hat{\gamma}$ is the adiabatic index and the double brackets indicate the jump of the quantities through both sides of the shock.
 
-In more simple case of strictly perpendicular shock, shock velocity and wownstream temperature can be expressed\cite{Amato2006} as
+	In the ultra-relativistic limit, the shock velocity and downstream temperature can be expressed\cite{Amato2006,2011A&ARv..19...42B} as
 	\begin{equation}\label{vshock}
@@
 	\end{equation}
-Here $\sigma$ is magnetisation and $\Gamma$ is upstream lorentz factor measured in downstream frame.
+	Here $\sigma$ is magnetization and $\Gamma$ is upstream lorentz factor measured in downstream frame.
 
-One weak point of this formalism is adiabatic index $\gamma$.
-For monoatomic gas it is well-defined in classical limit $\gamma = \frac{5}{3}$ and in ultra-relativistic linit $\gamma = \frac{4}{3}$.
-In the gap between them it is a non-trivial question which value of $\gamma$ to choose.
-Another problem is that one-fluid model doesn't take into account the difference between electrons' and protons' temperature.
-And two-fluid model also use some phenomenological methods to evaluate electrons' temperautre.
-
-In this work we used Particle-in-Cell method for modelling relativistic shock waves to study accuracy of MHD equations and to determine adiabatic index.
+	A problem of the application of the single fluid formalism is the correct choice of the adiabatic index $\hat{\gamma}$.
+For a monoatomic gas it is well-defined: in the classical limit $\hat{\gamma} = \frac{5}{3}$ and in the ultra-relativistic linit $\hat{\gamma} = \frac{4}{3}$.
+However, between these limits it is a non-trivial question, which value of $\hat{\gamma}$ to choose.
+Another problem is that the one-fluid model doesn't take into account for the difference between the temperatures of electrons and protons while the Coulomb relaxation is slow in relativistic astrophysical plasmas.
+Two-fluid models requre  some phenomenological approach to account for the collisionless relaxation of the electron temperature.
+Here we used a particle-in-cell approach for modelling relativistic shock waves to address the accuracy of MHD equations and to determine the adiabatic index for various cases.
 
 	\section{Numerical setup}
-We simulated a structure of a perpendicular relativistic shock and varied magnetisation, lorentz-factor and fraction of turbulent field. 
-The simulation is two dimensional with three dimensional velocities and fields.
-For the numerical simulation we used Tristan-mp code with the explicit numerical scheme developed by Buneman \cite{Buneman93} and  Spitkovsky\cite{Spitkovsky2005}.
+	We simulated the structure of a perpendicular relativistic shock and varied the magnetization, Lorentz-factor and the fraction of the turbulent field. 
+	The simulation is two-dimensional with three-dimensional velocities and fields.
+To perform such a task, we employed the publicly available Tristan-mp code with the explicit numerical scheme developed by Buneman \cite{Buneman93} and  Spitkovsky\cite{Spitkovsky2005}.
 
-Shock wave is initialized by an electron-proton plasma which is flowing into the simulation region through the right boundary and then reflecting from the super-conducting wall at the left boundary.
-We use the following parameters for the setups: the initial flow Lorentz factor $\Gamma = 1.5$, magnetization $\sigma = \frac{B^2}{4\pi\Gamma (n_p m_p + n_e m_e) c^2} = 0.04$ (in the turbulent case $B^2$ is the mean square field).
-The dimensionless thermal energy $\Delta \gamma = \frac{k T}{m_p c^2}$ is equal $10^{-4}$ and the proton mass is reduced to $m_p = 100 m_e$.
-Size of the simulation region along x axis is $L_x = 30000\frac{c}{\omega_p}$ and in transverse direction $L_y = 400\frac{c}{\omega_p}$, where $\omega_p$ is the plasma frequency $\omega_p = \sqrt{\frac{4\pi q^2 n}{\Gamma m_e}}$.
-These sizes correspond to $150000$ and $2000$ grid points respectively.
-Also, $2000$ grid points correspond to approximately $10$ gyroradii of upstream protons.
-In simulation with turbulent fields, we initialized it as summ of harmonic modes with Kolmogorov's spectum, and it is described in our previous paper in detail\cite{Romansky2019}.
-The energy density of turbulent field compared to the total magnetic energy density is $\eta$ and varying from 0 percent to 90 percent.
+	A shock wave is initialized in an electron-proton plasma which is flowing into the simulation box through its right boundary and then reflecting from the super-conducting wall at the left boundary.
+The typical simulation parameters are: the initial flow Lorentz factor $\Gamma = 1.5$, magnetization $\sigma = \frac{B^2}{4\pi\Gamma (n_p m_p + n_e m_e) c^2} = 0.04$ (in the turbulent case $B^2$ is the mean square field).
+The dimensionless thermal energy $\Delta \gamma = \frac{k T}{m_p c^2}$ is equal to $10^{-4}$ and the electron mass is enlarged to $m_e = \frac{m_p}{100}$.
+The size of the simulation box along the $x$ axis is $L_x = 30000\frac{c}{\omega_p}$ and in the transverse direction $L_y = 400\frac{c}{\omega_p}$, where $\omega_p$ is the plasma frequency $\omega_p = \sqrt{\frac{4\pi q^2 n}{\Gamma m_e}}$.
+These sizes correspond to $150000$ and $2000$ grid points, respectively.
+Also, $2000$ grid points correspond to approximately $10$ gyroradii of the upstream protons.
+	In simulations with turbulent fields, we initialized it as a summ of harmonic modes with Kolmogorov's spectum, and it is described  in detail in\cite{Romansky2019}.
+The energy density of the turbulent field compared to the total magnetic energy density is $\eta$ , which can vary  from 0 percent to 90 percent.
 
 	\section{Results}
-Using the density profiles of shock wave, we derived magnetohydrodynamic parameters of plasma, using formulae \ref{vshock} and \ref{temperature}.
-Density, averaged in transversal directions  has clear jump as shown on figures \ref{density_noturb} and \ref{density_turb}, indicating the coordinate of shock front.
-
-\begin{figure}[h!]
-	\centering
-	\begin{minipage}{0.5\textwidth}
-		\centering
-		\includegraphics[width=0.98\textwidth]{fig/density_noturb.png} 
-		\caption{Averaged by y dimension density profile of shock wave.}
-		\label{density_noturb}
-	\end{minipage}\hfill
-	\begin{minipage}{0.5\textwidth}
-		\centering
-		\includegraphics[width=0.98\textwidth]{fig/density_turb.png} 
-		\caption{Averaged by y dimension density profile of shock wave in turbulent medium.}
-		\label{density_turb}
-	\end{minipage}
-\end{figure}
-
-Evalation of shock front coordinates at different time moments shows that shock front velocity is constant with high accuracy in the case without turbulent field and has small oscillations without turbulence, see figure\ref{shockx}.
-\begin{table}
+	We used several parameter sets, listed in table \ref{setups}, with different magnetization, lorentz-factors, fraction of turbulent field. Also we tested differences between 2d and 3d cases. 
+	\begin{table}[h!]
 		\label{setups}
 		\begin{center}
-		\begin{tabular}{|c| c| c| c| c| c| c| c| c| c| c| c|}
+			\begin{tabular}{|c | c| c| c| c| c| c| c| c| c| c| c|}
 				\hline
-			Setup & Nx & Ny & Nz & $\sigma$ & $\Gamma$ & $\eta$ & $\beta$ & $\hat{\gamma}$ & $T  10^{12} K$ & $T_p  10^{12} K$ & $T_e  10^{12} K$\\
-			reg1 & 150000 & 2000 & 1 & 0.04 & 1.5 & 0 & 0.394 & 1.349 & 6.38 & 1 & 1\\
-			reg2 & 150000 & 2000 & 1 & 0.004 & 1.5 & 0 & 0.355 & 1.35 & 5.79 & 1 & 1\\
-			reg3 & 150000 & 2000 & 1 & 0.0004 & 1.5 & 0 & 0.339 & 1.338 & 5.53 &1 & 1\\
-			turb1 & 150000 & 2000 & 1 & 0.04 & 1.5 & 90 & 0.309 & 1.247 & 5.0 & 1 & 1\\
-			turb2 & 150000 & 2000 & 1 & 0.04 & 1.5 & 30 & 0.356 & 1.305 & 5.77 & 1 & 1\\
-			turb3 & 150000 & 2000 & 1 & 0.004 & 1.5 & 90 & 0.309 & 1.303 & 5.03 & 1 & 1\\
-			turb4 & 150000 & 2000 & 1 & 0.004 & 1.5 & 30 & 0.348 & 1.343 & 5.67 & 1 & 1\\
-			gam1 & 150000 & 2000 & 1 & 0.04 & 2 & 0 & 0.447 & 1.411 & 9.67 & 1 & 1\\
-			gam2 & 150000 & 2000 & 1 & 0.04 & 5 & 0 & 0.505 & 1.475 & 27.3 & 1 & 1\\
-			gam3 & 150000 & 2000 & 1 & 0.04 & 10 & 0 & 0.528 & 1.499 & 57.1 & 1 & 1\\
-			reg3d & 80000 & 40 & 40 & 0.04 & 1.5 & 0 & 0.447 & 1.43 & 1 & 1 & 1\\
+				Setup & Nx & Ny & Nz & $\sigma$ & $\Gamma$ & $\eta$ & $\frac{v_{sh}}{c}$ & $\hat{\gamma}$ & $T \ 10^{12} K$ & $T_p\ 10^{12}K$ & $T_e \ 10^{12}K$\\
+				\hline
+				reg1 & 150000 & 2000 & 1 & 0.04 & 1.5 & 0 & 0.394 & 1.513 & 3.5 & 2.4 & 0.25\\
+				reg2 & 150000 & 2000 & 1 & 0.004 & 1.5 & 0 & 0.355 & 1.512 & 3.52 & 2.5 & 0.23\\
+				reg3 & 150000 & 2000 & 1 & 0.0004 & 1.5 & 0 & 0.339 & 1.514 & 3.46 &2.4 & 0.38\\
+				turb1 & 150000 & 2000 & 1 & 0.04 & 1.5 & 90 & 0.309 & 1.526 & 3.0 & 1.91 & 0.68\\
+				turb2 & 150000 & 2000 & 1 & 0.04 & 1.5 & 30 & 0.356 & 1.519 & 3.27 & 2.3 & 0.32\\
+				turb3 & 150000 & 2000 & 1 & 0.004 & 1.5 & 90 & 0.309 & 1.52 & 3.25 & 2.1 & 0.41\\
+				turb4 & 150000 & 2000 & 1 & 0.004 & 1.5 & 30 & 0.348 & 1.514 & 3.48 & 2.4 & 0.35\\
+				gam1 & 150000 & 2000 & 1 & 0.04 & 2 & 0 & 0.447 & 1.463 & 6.3 & 4.2 & 0.67\\
+				gam2 & 150000 & 2000 & 1 & 0.04 & 5 & 0 & 0.505 & 1.395 & 20.6 & 7.1 & 1.0\\
+				gam3 & 150000 & 2000 & 1 & 0.04 & 10 & 0 & 0.528 & 1.359 & 43.8 & 16.0 & 19.3\\
+				reg3d & 80000 & 40 & 40 & 0.04 & 1.5 & 0 & 0.349 & 1.52 & 3.23 & 2.9 & 0.21\\
 
@@
 
+	Using the density profiles of the simulated shock waves, we derived magnetohydrodynamic parameters of plasma, solving the jump conditions (\ref{hugoniot}).
+Density, averaged in transversal directions  has clear jump as shown on figures \ref{density_noturb} and \ref{density_turb}, indicating the position of the shock front.
+So we can evaluate shock velocity from simulations, and then solve jump conditions for adiabatic index and downstream temperature, using iterative method.
+
 	\begin{figure}[h!]
 		\centering
-	\includegraphics[width=0.6\textwidth]{fig/shock_x.png} 
-	\caption{Coordinate of shock front for different setups.}
+		\begin{minipage}{0.49\textwidth}
+			\centering
+			\includegraphics[width=0.98\textwidth]{fig/density_noturb.png} 
+			\caption{Density profile of a shock wave in a medium with regular field, averaged through the transversal dimensions, setup reg1.}
+			\label{density_noturb}
+		\end{minipage}\hfill
+		\begin{minipage}{0.5\textwidth}
+			\centering
+			\includegraphics[width=0.98\textwidth]{fig/density_turb.png} 
+			\caption{Density profile of a shock wave in a turbulent medium, averaged through the transversal dimensions, setup turb1.}
+			\label{density_turb}
+		\end{minipage}
+	\end{figure}
+
+	Evaluation of shock front coordinates, which we have determined as a point where  the density is two times larger than in the far upstream, at different time moments shows that the shock front velocity is constant in time to a high accuracy, see figure \ref{shock_x}.
+Also, it shows, that shock front velocity decreases with decreasing magnetization, and adiabatic index does not remain constant.The shock slows down with increasing turbulent field energy and it requires significant decreasing of adiabatic index.
+It can be explained by appearence of new degrees of freedom related to magnetic field oscillations.
+Obviously, the shock front is faster for higher values of the upstream Lorentz-factor, and the adiabatic index  converges to the value $4/3$ which is ultra-relativistic limit.
+For lorentz-factors less then 10, $\hat{\gamma}$ is significantly higher then limitary value.
+
+	\begin{figure}[h!]
+		\centering
+		\includegraphics[width=0.65\textwidth]{fig/shock_x.png} 
+		\caption{Evolution of the shock front position for various parameter sets.}
 		\label{shock_x}
@@
 
+	We derived the temperatures of protons and electrons by fitting their distibutions with Maxwell-Juttner function, via minimization of the functional $f(T) = \int_{0}^{p_{max}} (F(p) - F_{m}(p,T))^2dp$ by temperature, where $F(p)$ is the electron or proton distribution derived from the simulation, $F_{m}(p,T)$ is the Maxwell-Juttner function with temperature $T$ and $p_{max}$ is the momentum where the distribution becomes non-thermal.
+Such evaluation of the downstream temperature is possible even in the cases of strong non-thermal tail, as shown in figure \ref{temp_fit}.
+
+	\begin{figure}[h!]
+		\centering
+		\includegraphics[width=0.65\textwidth]{fig/temperature_fit.png} 
+		\caption{Fitting the distribution for turb1 parameter set.}
+		\label{temp_fit}
+	\end{figure}
+
+	An analysis of the distribution functions shows that the temperatures of protons and electrons are much smaller then predicted by jump conditions (\ref{hugoniot}).
+It means that magnetohydrodynamical formalism is not good enough for description of downstream temperatures of a relativistic shock wave.
+One can notice that relation between the temperatures of protons and electrons is not constant.
+It is close to $\sqrt{\frac{m_p}{m_e}}$ at low Lorenz-factors and goes to $1$ at higher ones.
+Turbulence also has a strong influence on the temperatures, as it makes electrons much hotter, while the protons become colder.
+
+	A comparision of 2d and 3d setups with same parameters (setups reg1 and reg3d) shows that the differences in result are noticable, but not crucial - temperature and shock velocity differes less than on 10 percent.
+This means that 2d-simulations are appropriate for modeling relativistic collisionless shock waves in megnetized plasmas.
+
 	\section{Conclusions}
-In this paper we presented the results of particle-in-cell simulation of electron acceleration in trans-relativistic shock propagating in a clumped  wind of a massive star.
-Simulations show that the presence  of background  magnetic turbulence in the clumped wind has strong impact on particle acceleration and the acceleration is efficient for high enough level of the magnetic turbulence.
-The electron acceleration is highly suppressed in the case of  trans-relativistic shock propagating quasi-perpendicular to a regular magnetic field in the wind without the background magnetic turbulence.
-With the PIC simulations we demonstrated here that the presence of magnetic fluctuations in the background wind results in efficient electron acceleration by the quasi-perpendicular trans-relativistic shocks.
-We find that the scales and amplitudes of the magnetic turbulent field are important for this process.
-Further PIC simulations with a wider dynamical range will allow to study the effect of turbulent wind on particle acceleration in detail.
+
+	Particle-in-cell simulations of relativistic have been used to derive effective macroscopic parameters of relativistic shocks in space plasmas.
+The effective adiabatic index and electron temperatures are two important parameters required to supplement large-scale singl-fluid magneto-hydrodynamic simulations of processes in astrophysical plasmas.
+Further modelling is needed to study the dependences of the two parameters  on the initial conditions which, reflect the presence of MHD large-scale turbulence in the shock upstream.
 
 	\ack
-The authors acknowledge a support from RSF grant 16-12-10225.
-
-%\newpage
+	Results of the work were obtained using computational resources of Peter the Great Saint-Petersburg Polytechnic University Supercomputing Center (http://scc.spbstu.ru)
 

### 2020-04-22 SphericalTurbulence/SphericalTurbulence.cpp

@@ -1,0 +1,543 @@
+// SphericalTurbulence.cpp : Defines the entry point for the console application.
+//
+#include <cstdlib>
+#include <math.h>
+#include "stdio.h"
+#include <time.h>
+#include <cmath>
+#include <string>
+
+const int randomParameter = 1024;
+const double atomicUnionMass = 1.66053904E-24;
+const double massProtonReal = 1.67262177E-24;
+const double massAlphaReal = 6.644656E-24;
+const double massElectron = 0.910938291E-27;
+//const double massElectronFactor = massProtonReal/(10.0*massElectronReal);
+//const double massElectronFactor = 100;
+const double massDeuteriumReal = 3.34449696893E-24;
+const double massHelium3Real = 5.00823792874E-24;
+const double massOxygenReal = 26.5601801672E-24;
+const double massSiliconReal = 46.4567787264E-24;
+const double kBoltzman = 1.3806488E-16;
+const double speed_of_light = 2.99792458E10;
+const double speed_of_light2 = speed_of_light * speed_of_light;
+const double speed_of_light4 = speed_of_light2 * speed_of_light2;
+const double electron_charge = 4.803529695E-10;
+const double pi = 4*atan2(1.0,1.0);
+const double four_pi = 4*pi;
+
+const int Nx = 20;
+const int Ny = 20;
+const int Nz = 20;
+const int Nk = 10;
+
+const int Napprox = 52;
+
+///      K5/3(x)
+const double McDonaldValue[Napprox] = { 1.43E15, 9.8E13, 3.09E13, 2.11E12, 6.65E11, 4.55E10, 1.43E10, 9.8E8,
+	3.08E8, 2.11E7, 6.65E6, 4.55E5, 1.43E4, 9802, 3087, 670, 211, 107, 66.3, 33.6, 20.7, 14.1, 11.0, 10.3, 6.26, 4.2, 3.01, 2.25, 1.73, 1.37, 1.10,
[…]

### 2020-08-22 papers/Physica2020/Romansky_Physica2020.tex
paper

--- 2020-04-16 papers/Physica2020/Romansky_Physica2020.tex
@@
 \documentclass[a4paper]{jpconf}
-\bibliographystyle{iopart-num}
+% \bibliographystyle{iopart-num}
 \usepackage{amsmath}
@@
 
+\usepackage{xspace}
+\usepackage[dvipsnames]{xcolor}
+\usepackage[normalem]{ulem}
+% Suggest to remove
+\newcommand{\asrm}[1]{{\color{OrangeRed}\sout{#1}}}
+% Suggest to insert
+\newcommand{\as}[1]{\color{RoyalBlue}#1\xspace\color{black}}
+% Suggest to replace
+\newcommand{\asrp}[2]{\asrm{#1} \as{#2}}
+% Comment
+\newcommand{\ascm}[1]{{\color{ForestGreen}#1}}
+
 \begin{document}
-\title{Evaluating MHD parameters of relativistic shock waves with Particle-in-Cell modelling}
+	\title{Evaluating MHD parameters of relativistic shock waves with particle-in-cell modeling}
 
-\author{V I Romansky$^{1}$, A M Bykov$^{1,2}$ and S M Osipov$^1$}
+	\author{V I Romansky$^{1}$, A M Bykov$^{1}$ and S M Osipov$^{1,2}$}
 
@@
 	\begin{abstract}
-Reklativistic shock waves are widely discussed phenomenae of modern astrophysics.
-Studying of their properties is necessary for consistent models of supernova remnants, jets and other astrophysical objects.
-But magneto-hydrodinamical description of relativistic shock waves is not full, because it does not take into account electrons and a lot of relativistic effects such as displacement current density and changing of adiabatic index That's why we need more accurate methods to check and fix MHD models.
+		Relativistic plasma outflows are observed in gamma-ray burst 
+		sources, jets of active galactic nuclei, pulsar wind nebulae and
+		supernovae explosions. Magnetohydrodynamical (MHD) shock waves
+		inevitably result from interactions of such relativistic outflows with
+		the ambient interstellar matter. The widely used single-fluid MHD
+		description of relativistic shock waves is a useful tool to study the
+		global structure of such objects. However, to interpret the observed
+		emission spectra of space objects with relativistic shocks, a kinetic
+		description of electrons, positrons, and ions is needed. The kinetic
+		structure of relativistic shocks can be modelled within the
+		particle-in-cell approach, which allows one to properly account for
+		the electron scale processes and for the density of the displacement
+		current. Here we discuss the effect of the micro-scale plasma processes on such
+		macroscopic parameters of relativistic shocks as the adiabatic index
+		of the relativistic fluid in the shock vicinity, with a relatively
+		long equilibration time. The adiabatic index is an important parameter of the
+		single-fluid MHD models commonly used for shock modeling at much
+		longer hydrodynamical scales.
 	\end{abstract}
+
 	\section{Introduction}
-Relativistic shock jump (Rankine-Hugoniot) were given by Blandford and McKee \cite{Blandford76} and generalized to MHD by Emmering and Chevalier\cite{Emmering87}. Shock conditions are:
+	Relativistic jets and winds are the generic outflows of the relativistic supernovae \cite{2010Natur.463..513S,2007ApJ...667..351W}, the accreting black holes, both suprmassive in the active galactic nucleis  \cite{1984RvMP...56..255B} as well as the stellar masses in the gamma-ray burst sources, microquasars \cite{2019MmSAI..90...57M,1999PhR...314..575P,2014LNP...876.....R} and young pulsars \cite{2017SSRv..207....1B}.
+Relativistic shocks can be formed either inside the flow of low or moderate magnetization (the internal shocks) or can be driven by the outflow colliding with the ambient medium (the external shocks).
+In most cases the relatistivistic shocks in astrophysical objects are collisionless and are associated with strongly non-equilibrium particle distributions
+	\cite{2012SSRv..173..309B,2015SSRv..191..519S,2017SSRv..207..319P}, and they can accelerate particles to ultra-high energies \cite{2009JCAP...11..009L}.
+On the one hand, the presence of non-maxwellian components with very long equilibration time requires a kinetic description of the shocked flows.
+On the other hand, available computer capabilities do not alow one to provide a fully kinetic modelling of the macroscopic astrophysical flows and, therefore, the fluid MHD description based on such macroscopic parameters as the effective adiabatic index of the flow is widely used.
+Here we discuss an approach allowing to determine the effective adiabatic indexes from the microscopical particle-in-cell modeling of the relativistic shocks.
+
+	Relativistic fluid shock dynamics was discussed by Lichnerowicz \cite{1967rhm..book.....L}, Blandford and McKee \cite{Blandford76} and by Emmering and Chevalier\cite{Emmering87}.
+The relativistic shock jump conditions in the case of a perpendicular shock which is typical for relativisic shocks can be written as:
 	\begin{equation}\label{hugoniot}
@@
 	\end{equation}
-where $u = (u^t,u^r)$ is velocity four-vector measured in the shock frame, $\rho$ is proper mass density, $B$ is magnetic field and $P$ is pressure in the fluid frame and $w = \rho c^2 + \frac{\hat{\gamma}}{\hat{\gamma} - 1} P$ is enthalpy, $\hat{\gamma}$ is adiabatic index and double brackets indicate the mismatches of quantities to both sides of the shock.
+	where $u = (u^t,u^r)$ is the velocity four-vector measured in the shock frame, $\rho$ is the proper mass density, $B$ is the magnetic field, $P$ is the pressure in the fluid frame, and $w = \rho c^2 + \frac{\hat{\gamma}}{\hat{\gamma} - 1} P$ is the enthalpy, while $\hat{\gamma}$ is the adiabatic index and the double brackets indicate the jump of the quantities between two sides of the shock.
 
-In more simple case of strictly perpendicular shock, shock velocity and wownstream temperature can be expressed\cite{Amato2006} as
+	In the ultra-relativistic limit, the shock velocity $v_{sh}$ and the downstream temperature $T$ can be expressed as, respectively, \cite{Amato2006,2011A&ARv..19...42B}:
 	\begin{equation}\label{vshock}
-v_{sh} = \frac{c}{2}\frac{1}{1 + \sigma}\lbrace (\hat{\gamma}(1 + \frac{\sigma}{2})-1) + {\lbrack (\hat{\gamma}(1 + \frac{\sigma}{2})-1)^2 + 4\sigma(1+\sigma)(1 - \frac{\hat{\gamma}}{2})\rbrack}^{\frac{1}{2}} \rbrace
+	v_{sh} = \frac{c}{2}\frac{1}{1 + \sigma}\lbrace (\hat{\gamma}(1 + \frac{\sigma}{2})-1) + {\lbrack (\hat{\gamma}(1 + \frac{\sigma}{2})-1)^2 + 4\sigma(1+\sigma)(1 - \frac{\hat{\gamma}}{2})\rbrack}^{\frac{1}{2}} \rbrace,
 	\end{equation}
 	\begin{equation}\label{temperature}
-T = m c^2 \Gamma \frac{v_{sh}}{c}(1 - \frac{\sigma}{2}\frac{c - v_{sh}}{v_sh})
+	T = m c^2 \Gamma \frac{v_{sh}}{c}(1 - \frac{\sigma}{2}\frac{c - v_{sh}}{v_sh}),
 	\end{equation}
-Here $\sigma$ is magnetisation and $\Gamma$ is upstream lorentz factor measured in downstream frame.
+	Where $\sigma$ is the magnetization and $\Gamma$ is the upstream Lorentz-factor measured in the downstream frame.
 
-One weak point of this formalism is adiabatic index $\gamma$.
-For monoatomic gas it is well-defined in classical limit $\gamma = \frac{5}{3}$ and in ultra-relativistic linit $\gamma = \frac{4}{3}$.
-In the gap between them it is a non-trivial question which value of $\gamma$ to choose.
-Another problem is that one-fluid model doesn't take into account the difference between electrons' and protons' temperature.
-And two-fluid model also use some phenomenological methods to evaluate electrons' temperautre.
-
-In this work we used Particle-in-Cell method for modelling relativistic shock waves to study accuracy of MHD equations and to determine adiabatic index.
+	The application of the single fluid formalism requires a correct choice of the adiabatic index $\hat{\gamma}$.
+For a monoatomic gas it is well-defined: in the classical limit $\hat{\gamma} = \frac{5}{3}$ and in the ultra-relativistic limit $\hat{\gamma} = \frac{4}{3}$.
+However, in the range between these limits the choice of the value of $\hat{\gamma}$ is not trivial.
+Moreover, the one-fluid model doesn't take into account for the difference between the temperatures of electrons and protons, while the Coulomb relaxation is slow in relativistic astrophysical plasmas.
+Two-fluid models require some phenomenological approach to account for the collisionless relaxation of the electron temperature.
+Therefore, in this study we used the particle-in-cell approach for modelling of relativistic shock waves in order to address the accuracy of MHD equations and to determine the adiabatic index for various cases.
 
 	\section{Numerical setup}
-We simulated a structure of a perpendicular relativistic shock and varied magnetisation, lorentz-factor and fraction of turbulent field. 
-The simulation is two dimensional with three dimensional velocities and fields.
-For the numerical simulation we used Tristan-mp code with the explicit numerical scheme developed by Buneman \cite{Buneman93} and  Spitkovsky\cite{Spitkovsky2005}.
+	In this study we performed simulations of a perpendicular relativistic shock  varying the magnetization, Lorentz-factor and the fraction of the turbulent field. 
+	The simulation domain is two-dimensional with three-dimensional velocities and fields.
+To perform such computation we employed the publicly available Tristan-mp code with the explicit numerical scheme developed by Buneman \cite{Buneman93} and  Spitkovsky\cite{Spitkovsky2005}.
 
-Shock wave is initialized by an electron-proton plasma which is flowing into the simulation region through the right boundary and then reflecting from the super-conducting wall at the left boundary.
-We use the following parameters for the setups: the initial flow Lorentz factor $\Gamma = 1.5$, magnetization $\sigma = \frac{B^2}{4\pi\Gamma (n_p m_p + n_e m_e) c^2} = 0.04$ (in the turbulent case $B^2$ is the mean square field).
-The dimensionless thermal energy $\Delta \gamma = \frac{k T}{m_p c^2}$ is equal $10^{-4}$ and the proton mass is reduced to $m_p = 100 m_e$.
-Size of the simulation region along x axis is $L_x = 30000\frac{c}{\omega_p}$ and in transverse direction $L_y = 400\frac{c}{\omega_p}$, where $\omega_p$ is the plasma frequency $\omega_p = \sqrt{\frac{4\pi q^2 n}{\Gamma m_e}}$.
-These sizes correspond to $150000$ and $2000$ grid points respectively.
-Also, $2000$ grid points correspond to approximately $10$ gyroradii of upstream protons.
-In simulation with turbulent fields, we initialized it as summ of harmonic modes with Kolmogorov's spectum, and it is described in our previous paper in detail\cite{Romansky2019}.
-The energy density of turbulent field compared to the total magnetic energy density is $\eta$ and varying from 0 percent to 90 percent.
+	A shock wave is initialized in an electron-proton plasma which is flowing into the simulation box through its right boundary and then is reflecting from the super-conducting wall at the left boundary.
+The typical simulation parameters are : the initial flow Lorentz factor $\Gamma = 1.5$, magnetization $\sigma = \frac{B^2}{4\pi\Gamma (n_p m_p + n_e m_e) c^2} = 0.004$ (in the turbulent case $B^2$ is the mean square field).
+The dimensionless thermal energy $\Delta \gamma = \frac{k T}{m_p c^2}$ is equal to $10^{-4}$ and the electron mass is increased up to $m_e = \frac{m_p}{100}$.
+The size of the simulation box along the $x$ axis is $L_x = 30000\frac{c}{\omega_p}$ and in the transverse direction $L_y = 400\frac{c}{\omega_p}$, where $\omega_p = \sqrt{\frac{4\pi q^2 n}{\Gamma m_e}}$ is the plasma frequency.
+These scales correspond to $150000$ and $2000$ grid points in $x$ and $y$ directions, respectively.
+Also, $2000$ grid points correspond to approximately $10$ gyroradii of the upstream protons.
+	In simulations with turbulent fields, we initialized it as  summ of harmonic modes with Kolmogorov's spectum, as described in detail in\cite{Romansky2019}.
+The ratio of the energy density of the turbulent field to the total magnetic energy density is denoted as $\eta$ , which varied from 0 percent to 90 percent in our simulations.
+Regular magnetic field is normal to the $x$ axis and is rotated by the angle $\phi$ about $x$ axis: the value $\phi = 0$ corresponds to the out-of-plane orientation of magnetic field and $\phi = 90$ corresponds to the in-plane orientation.
 
 	\section{Results}
-Using the density profiles of shock wave, we derived magnetohydrodynamic parameters of plasma, using formulae \ref{vshock} and \ref{temperature}.
-Density, averaged in transversal directions  has clear jump as shown on figures \ref{density_noturb} and \ref{density_turb}, indicating the coordinate of shock front.
+	We used several parameter sets, listed in table \ref{setups}, with different magnetization, Lorentz-factors and the fraction of turbulent field.
+Also, we investigated differences between 2d and 3d cases.
+	\begin{table}[h!]
+		\label{setups}
+		\begin{center}
+			\begin{tabular}{|c | c| c | c| c| c| c| c| c| c| c| c| c|}
+				\hline
+				Setup & Nx & Ny & Nz & $\phi$ & $\sigma$ & $\Gamma$ & $\eta$ & $\frac{v_{sh}}{c}$ & $\hat{\gamma}$ & $T \ 10^{12} K$ & $T_p\ 10^{12}K$ & $T_e \ 10^{12}K$\\
+				\hline
+				reg1 & 150000 & 2000 & 1 & 0 & 0.004 & 1.5 & 0 & 0.355 & 1.513 & 3.53 & 2.5 & 0.23\\
+				reg2 & 150000 & 2000 & 1 & 90 & 0.004 & 1.5 & 0 & 0.24 & 1.53 & 2.87 & 2.2 & 0.28\\
+				reg3 & 150000 & 2000 & 1 & 0 & 0.0004 & 1.5 & 0 & 0.339 & 1.515 & 3.43 &2.4 & 0.38\\
+				reg4 & 150000 & 2000 & 1 & 90 & 0.0004 & 1.5 & 0 & 0.228 & 1.532 & 2.81 &1.65 & 0.37\\
+				turb1 & 150000 & 2000 & 1 & 0 & 0.004 & 1.5 & 90 & 0.309 & 1.52 & 3.25 & 2.1 & 0.41\\
+				turb2 & 150000 & 2000 & 1 & 0 & 0.004 & 1.5 & 30 & 0.348 & 1.514 & 3.48 & 2.4 & 0.35\\
+				gam1 & 150000 & 2000 & 1 & 0 & 0.004 & 2 & 0 & 0.412 & 1.46 & 6.3 & 4.19 & 0.62\\
+				gam2 & 150000 & 2000 & 1 & 0 & 0.004 & 5 & 0 & 0.479 & 1.39 & 20.5 & 7.3 & 11.0\\
+				gam3 & 150000 & 2000 & 1 & 0 & 0.004 & 10 & 0 & 0.504 & 1.36 & 42.6 & 16.8 & 21.3\\
+				reg3d & 80000 & 40 & 40 & 0 & 0.004 & 1.5 & 0 & 0.313 & 1.519 & 3.27 & 3.0 & 0.21\\
+
+				\hline
+			\end{tabular}
+		\end{center}
+		\caption{Parameters of different setups. }
+	\end{table}
+
+	Using the density profiles of the simulated shock waves, we determined magnetohydrodynamic parameters of plasma, solving the jump conditions (\ref{hugoniot}).
+The density averaged in transversal directions has clear jump as shown in figures \ref{density_noturb} and \ref{density_turb}, which is indicating the position of the shock front.
+So we can evaluate the shock velocity from results of simulations, and then solve jump conditions for the adiabatic index and the downstream temperature, using the iterative method.
 
@@
 		\centering
-	\begin{minipage}{0.5\textwidth}
+		\begin{minipage}{0.49\textwidth}
 			\centering
 			\includegraphics[width=0.98\textwidth]{fig/density_noturb.png} 
-		\caption{Averaged by y dimension density profile of shock wave.}
+			\caption{Density profile of a shock wave in a medium with a regular field, averaged through the transversal dimensions, setup reg1.}
 			\label{density_noturb}
@@
 			\includegraphics[width=0.98\textwidth]{fig/density_turb.png} 
-		\caption{Averaged by y dimension density profile of shock wave in turbulent medium.}
+			\caption{Density profile of a shock wave in a turbulent medium, averaged through the transversal dimensions, setup turb1.}
 			\label{density_turb}
@@
 
-Evalation of shock front coordinates at different time moments shows that shock front velocity is constant with high accuracy in the case without turbulent field and has small oscillations without turbulence, see figure\ref{shockx}.
-\begin{table}
-	\label{setups}
-	\begin{center}
-		\begin{tabular}{|c| c| c| c| c| c| c| c| c| c| c| c|}
-		\hline
-			Setup & Nx & Ny & Nz & $\sigma$ & $\Gamma$ & $\eta$ & $\beta$ & $\hat{\gamma}$ & $T  10^{12} K$ & $T_p  10^{12} K$ & $T_e  10^{12} K$\\
-			reg1 & 150000 & 2000 & 1 & 0.04 & 1.5 & 0 & 0.394 & 1.349 & 6.38 & 1 & 1\\
-			reg2 & 150000 & 2000 & 1 & 0.004 & 1.5 & 0 & 0.355 & 1.35 & 5.79 & 1 & 1\\
-			reg3 & 150000 & 2000 & 1 & 0.0004 & 1.5 & 0 & 0.339 & 1.338 & 5.53 &1 & 1\\
-			turb1 & 150000 & 2000 & 1 & 0.04 & 1.5 & 90 & 0.309 & 1.247 & 5.0 & 1 & 1\\
-			turb2 & 150000 & 2000 & 1 & 0.04 & 1.5 & 30 & 0.356 & 1.305 & 5.77 & 1 & 1\\
-			turb3 & 150000 & 2000 & 1 & 0.004 & 1.5 & 90 & 0.309 & 1.303 & 5.03 & 1 & 1\\
-			turb4 & 150000 & 2000 & 1 & 0.004 & 1.5 & 30 & 0.348 & 1.343 & 5.67 & 1 & 1\\
-			gam1 & 150000 & 2000 & 1 & 0.04 & 2 & 0 & 0.447 & 1.411 & 9.67 & 1 & 1\\
-			gam2 & 150000 & 2000 & 1 & 0.04 & 5 & 0 & 0.505 & 1.475 & 27.3 & 1 & 1\\
-			gam3 & 150000 & 2000 & 1 & 0.04 & 10 & 0 & 0.528 & 1.499 & 57.1 & 1 & 1\\
-			reg3d & 80000 & 40 & 40 & 0.04 & 1.5 & 0 & 0.447 & 1.43 & 1 & 1 & 1\\
-
-		\hline
-		\end{tabular}
-	\end{center}
-	\caption{Parameters of different setups}
-\end{table}
+	Evaluation of shock front coordinates, which we have determined as a point where  the density is two times larger than the density observed far upstream, at different time steps shows that the shock front velocity is constant in time up to a high accuracy, see figure \ref{shock_x}.
+Also, it shows, that the shock front velocity decreases with the decreasing magnetization, and, consequently, the adiabatic index does not remain constant.
+The shock slows down with the increasing turbulent field energy and it requires significant decrease of the adiabatic index.
+It can be explained by the appearence of new degrees of freedom related to magnetic field oscillations.
+Obviously, the shock front is faster for higher values of the upstream Lorentz-factor, and the adiabatic index  converges to the value $4/3$ which is the ultra-relativistic limit.
+For Lorentz-factors less then 10, $\hat{\gamma}$ is significantly higher than the limiting value.
 
@@
 		\centering
-	\includegraphics[width=0.6\textwidth]{fig/shock_x.png} 
-	\caption{Coordinate of shock front for different setups.}
+		\includegraphics[width=0.7\textwidth]{fig/shock_x.png} 
+		\caption{Evolution of the shock front position for various parameter sets.}
 		\label{shock_x}
@@
 
+	Distributions of particles significantly vary through the x coordinate, and in figure \ref{points} we show the distribution of electrons at different distances from shock front, measured in proton gyroradius $r_g$.
+Positive distances correspond to upstream region and negative to downstream.
+As one can see, for distances more than $6 rg$ in the downstream distribution does not depend on coordinate.
+We obtained the temperatures of protons and electrons by fitting their distibutions in far downstream with Maxwell-Juttner function, via minimization of the functional $f(T) = \int_{0}^{p_{max}} (F(p) - F_{m}(p,T))^2dp$ by temperature, where $F(p)$ is the electron or proton distribution derived from the simulation, $F_{m}(p,T)$ is the Maxwell-Juttner function with temperature $T$ and $p_{max}$ is the momentum for which the distributioin becomes non-thermal.
+Obtained temperature has rather week dependence on choice of $p_{max}$.
+Such evaluation of the downstream temperature is possible even in the cases of strong non-thermal tail, as shown in figure \ref{temp_fit}.
+
+	\begin{figure}[h!]
+		\centering
+		\includegraphics[width=0.7\textwidth]{fig/electrons_at_points.png} 
+		\caption{Distributions of electrons at different regions of simulation box for turb1 parameter set}
+		\label{points}
+	\end{figure}
+
+	\begin{figure}[h!]
+		\centering
+		\includegraphics[width=0.7\textwidth]{fig/temperature_fit.png} 
+		\caption{Fitting the distribution for turb1 parameter set with Maxwell-Juttner distribution}
+		\label{temp_fit}
+	\end{figure}
+
+	An analysis of the distribution functions shows that the temperatures of protons and electrons are much smaller than predicted by jump conditions (\ref{hugoniot}).
+It means that magnetohydrodynamical formalism is not sufficient for the description of downstream temperatures of a relativistic shock wave.
+One can notice that the ratio of temperature of protons to the temperatures of electrons is not constant: it is close to $\sqrt{\frac{m_p}{m_e}}$ at low Lorenz-factors and decreases down to $1$ at higher ones.
+Turbulence also has a strong influence on the temperatures, as it makes electrons much hotter, while the protons become colder.
+
+	Finally, a comparison of the results obtained using 2d and 3d setups with the same parameters (setups reg1, reg2 and reg3d) reveals that the differences in results depend on orientation of magnetic field in 2d simulations.
+For out-of plane orientation these differences are noticeable, but not crucial - temperature and shock velocity differs in 2d and 3d setups by less than on 10 percent.
+This means that 2d simulations are appropriate for modeling of relativistic collisionless shock waves in magnetized plasmas.
+
 	\section{Conclusions}
-In this paper we presented the results of particle-in-cell simulation of electron acceleration in trans-relativistic shock propagating in a clumped  wind of a massive star.
-Simulations show that the presence  of background  magnetic turbulence in the clumped wind has strong impact on particle acceleration and the acceleration is efficient for high enough level of the magnetic turbulence.
-The electron acceleration is highly suppressed in the case of  trans-relativistic shock propagating quasi-perpendicular to a regular magnetic field in the wind without the background magnetic turbulence.
-With the PIC simulations we demonstrated here that the presence of magnetic fluctuations in the background wind results in efficient electron acceleration by the quasi-perpendicular trans-relativistic shocks.
-We find that the scales and amplitudes of the magnetic turbulent field are important for this process.
-Further PIC simulations with a wider dynamical range will allow to study the effect of turbulent wind on particle acceleration in detail.
+
+	Particle-in-cell simulations have been used to derive effective macroscopic parameters of relativistic shocks in space plasmas.
+The effective adiabatic index and electron temperatures are two important parameters required to supplement large-scale single-fluid magneto-hydrodynamic simulations of processes in astrophysical plasmas.
+Further modelling is needed to study the dependences of these two parameters  on the initial conditions which, reflect the presence of MHD large-scale turbulence in the shock upstream.
 
 	\ack
-The authors acknowledge a support from RSF grant 16-12-10225.
-
-%\newpage
+	V I Romansky and A M Bykov acknowledge a support from RSF grant 16-12-10225.
+	Results of the work were obtained using computational resources of Peter the Great Saint-Petersburg Polytechnic University Supercomputing Center (http://scc.spbstu.ru)
 
 	\section*{References}
+	% \bibliographystyle{apalike}
+	\bibliographystyle{iopart-num}
 
-\bibliographystyle{iopart-num}
 	\bibliography{bibliogr}

### 2020-08-31 papers/Physica2020/Romansky_Physica2020.tex
fz

@@
 \def\jgr{JGR}
+\def\jcap{Journal of Cosmology and Astroparticle Physics}
 
@@
 
-	\author{V I Romansky$^{1}$, A M Bykov$^{1}$ and S M Osipov$^{1,2}$}
+	\author{V I Romansky$^{1}$, A M Bykov$^{1,2}$ and S M Osipov$^{1}$}
 
@@
 		the ambient interstellar matter. The widely used single-fluid MHD
-		description of relativistic shock waves is a useful tool to study the
-		global structure of such objects. However, to interpret the observed
+		description of relativistic shock waves is the main tool to study the
+		global structure of such objects. However, to justify the validity of 
+		the global MHD models and to interpret the observed
 		emission spectra of space objects with relativistic shocks, a kinetic
-		description of electrons, positrons, and ions is needed. The kinetic
-		structure of relativistic shocks can be modelled within the
-		particle-in-cell approach, which allows one to properly account for
-		the electron scale processes and for the density of the displacement
-		current. Here we discuss the effect of the micro-scale plasma processes on such
-		macroscopic parameters of relativistic shocks as the adiabatic index
-		of the relativistic fluid in the shock vicinity, with a relatively
-		long equilibration time. The adiabatic index is an important parameter of the
+		description of electrons, positrons, and ions at microscales is needed. 
+		We model a plane relativistic shock propagating  transverse to a regular magnetic field in the electron-ion plasmas with imposed turbulent fluctuations in the shock upstream.   
+           Namely we study  the effect of the micro-scale plasma processes on 
+		macroscopic parameters of the mildly-relativistic shocks as the adiabatic index
+		of the relativistic fluid in the shock downstream. The adiabatic index is a macroscopic parameter of the
 		single-fluid MHD models commonly used for shock modeling at much
-		longer hydrodynamical scales.
+		longer hydrodynamical scales and it is especially important for the MHD modeling of the mildly-relativistic shocks.
 	\end{abstract}
@@
 	\section{Introduction}
-	Relativistic jets and winds are the generic outflows of the relativistic supernovae \cite{2010Natur.463..513S,2007ApJ...667..351W}, the accreting black holes, both suprmassive in the active galactic nucleis  \cite{1984RvMP...56..255B} as well as the stellar masses in the gamma-ray burst sources, microquasars \cite{2019MmSAI..90...57M,1999PhR...314..575P,2014LNP...876.....R} and young pulsars \cite{2017SSRv..207....1B}.
+	Relativistic jets and winds are the generic outflows of the relativistic supernovae \cite{2010Natur.463..513S,2007ApJ...667..351W}, the accreting black holes, both suprmassive in the active galactic nucleis  \cite{1984RvMP...56..255B} as well as the stellar masses in the gamma-ray burst sources, microquasars \cite{2019MmSAI..90...57M,1999PhR...314..575P,2014LNP...876.....R} and pulsar wind nebulae \cite{2019MNRAS.488.5690O,2017SSRv..207..235B,2017JPlPh..83e6301K,2019ApJ...876L...8B}.
 Relativistic shocks can be formed either inside the flow of low or moderate magnetization (the internal shocks) or can be driven by the outflow colliding with the ambient medium (the external shocks).
-In most cases the relatistivistic shocks in astrophysical objects are collisionless and are associated with strongly non-equilibrium particle distributions
-	\cite{2012SSRv..173..309B,2015SSRv..191..519S,2017SSRv..207..319P}, and they can accelerate particles to ultra-high energies \cite{2009JCAP...11..009L}.
+In many cases the relatistivistic shocks in astrophysical objects are collisionless and are often associated with non-equilibrium particle distributions \cite{2012SSRv..173..309B,2015SSRv..191..519S,2017SSRv..207..319P}.
+They can accelerate particles to ultra-high energies \cite{2009JCAP...11..009L}.
 On the one hand, the presence of non-maxwellian components with very long equilibration time requires a kinetic description of the shocked flows.
+The kinetic structure of relativistic shocks can be modelled within the particle-in-cell approach, which allows one to properly account for the electron scale processes and for the density of the displacement current which is important in relativistic plasma flows.
 On the other hand, available computer capabilities do not alow one to provide a fully kinetic modelling of the macroscopic astrophysical flows and, therefore, the fluid MHD description based on such macroscopic parameters as the effective adiabatic index of the flow is widely used.
-Here we discuss an approach allowing to determine the effective adiabatic indexes from the microscopical particle-in-cell modeling of the relativistic shocks.
+Here we discuss an approach allowing to determine the effective adiabatic indexes from the microscopical particle-in-cell modeling of the mildly-relativistic shocks typical for the relativistic supernovae and gamma-ray burst afterglows.
 
@@
 	A shock wave is initialized in an electron-proton plasma which is flowing into the simulation box through its right boundary and then is reflecting from the super-conducting wall at the left boundary.
-The typical simulation parameters are : the initial flow Lorentz factor $\Gamma = 1.5$, magnetization $\sigma = \frac{B^2}{4\pi\Gamma (n_p m_p + n_e m_e) c^2} = 0.004$ (in the turbulent case $B^2$ is the mean square field).
+The typical simulation parameters are : the initial flow Lorentz factor $\Gamma = 1.5$, the flow magnetization $\sigma = \frac{B^2}{4\pi\Gamma (n_p m_p + n_e m_e) c^2}$ was choosen to be  0.004 and 0.0004  (in the turbulent case $B^2$ is the mean square field).
 The dimensionless thermal energy $\Delta \gamma = \frac{k T}{m_p c^2}$ is equal to $10^{-4}$ and the electron mass is increased up to $m_e = \frac{m_p}{100}$.
@@
 The ratio of the energy density of the turbulent field to the total magnetic energy density is denoted as $\eta$ , which varied from 0 percent to 90 percent in our simulations.
-Regular magnetic field is normal to the $x$ axis and is rotated by the angle $\phi$ about $x$ axis: the value $\phi = 0$ corresponds to the out-of-plane orientation of magnetic field and $\phi = 90$ corresponds to the in-plane orientation.
+The regular magnetic field is normal to the $x$ axis and is rotated by the angle $\phi$ about $x$ axis: the value $\phi = 0$ corresponds to the out-of-plane orientation of magnetic field and $\phi = 90$ corresponds to the in-plane orientation.
 
@@
 	We used several parameter sets, listed in table \ref{setups}, with different magnetization, Lorentz-factors and the fraction of turbulent field.
-Also, we investigated differences between 2d and 3d cases.
+Also, we investigated differences between 2D and 3D cases.
 	\begin{table}[h!]
@@
 				\hline
-				reg1 & 150000 & 2000 & 1 & 0 & 0.004 & 1.5 & 0 & 0.355 & 1.513 & 3.53 & 2.5 & 0.23\\
-				reg2 & 150000 & 2000 & 1 & 90 & 0.004 & 1.5 & 0 & 0.24 & 1.53 & 2.87 & 2.2 & 0.28\\
-				reg3 & 150000 & 2000 & 1 & 0 & 0.0004 & 1.5 & 0 & 0.339 & 1.515 & 3.43 &2.4 & 0.38\\
-				reg4 & 150000 & 2000 & 1 & 90 & 0.0004 & 1.5 & 0 & 0.228 & 1.532 & 2.81 &1.65 & 0.37\\
-				turb1 & 150000 & 2000 & 1 & 0 & 0.004 & 1.5 & 90 & 0.309 & 1.52 & 3.25 & 2.1 & 0.41\\
-				turb2 & 150000 & 2000 & 1 & 0 & 0.004 & 1.5 & 30 & 0.348 & 1.514 & 3.48 & 2.4 & 0.35\\
-				gam1 & 150000 & 2000 & 1 & 0 & 0.004 & 2 & 0 & 0.412 & 1.46 & 6.3 & 4.19 & 0.62\\
+				reg1 & 150000 & 2000 & 1 & 0 & 0.004 & 1.5 & 0 & 0.355 & 1.51 & 3.5 & 2.5 & 0.23\\
+				reg2 & 150000 & 2000 & 1 & 90 & 0.004 & 1.5 & 0 & 0.240 & 1.53 & 2.9 & 2.2 & 0.28\\
+				reg3 & 150000 & 2000 & 1 & 0 & 0.0004 & 1.5 & 0 & 0.339 & 1.52 & 3.4 &2.4 & 0.38\\
+				reg4 & 150000 & 2000 & 1 & 90 & 0.0004 & 1.5 & 0 & 0.228 & 1.53 & 2.8 &1.7 & 0.37\\
+				turb1 & 150000 & 2000 & 1 & 0 & 0.004 & 1.5 & 90 & 0.309 & 1.52 & 3.3 & 2.1 & 0.41\\
+				turb2 & 150000 & 2000 & 1 & 0 & 0.004 & 1.5 & 30 & 0.348 & 1.51 & 3.5 & 2.4 & 0.35\\
+				gam1 & 150000 & 2000 & 1 & 0 & 0.004 & 2 & 0 & 0.412 & 1.46 & 6.3 & 4.2 & 0.62\\
 				gam2 & 150000 & 2000 & 1 & 0 & 0.004 & 5 & 0 & 0.479 & 1.39 & 20.5 & 7.3 & 11.0\\
 				gam3 & 150000 & 2000 & 1 & 0 & 0.004 & 10 & 0 & 0.504 & 1.36 & 42.6 & 16.8 & 21.3\\
-				reg3d & 80000 & 40 & 40 & 0 & 0.004 & 1.5 & 0 & 0.313 & 1.519 & 3.27 & 3.0 & 0.21\\
+				reg3d & 80000 & 40 & 40 & 0 & 0.004 & 1.5 & 0 & 0.313 & 1.52 & 3.3 & 3.0 & 0.21\\
 
@@
 		\end{minipage}\hfill
-		\begin{minipage}{0.5\textwidth}
+		\begin{minipage}{0.49\textwidth}
 			\centering
@@
 		\centering
-		\includegraphics[width=0.7\textwidth]{fig/shock_x.png} 
+		\includegraphics[width=0.65\textwidth]{fig/shock_x.png} 
 		\caption{Evolution of the shock front position for various parameter sets.}
@@
 Positive distances correspond to upstream region and negative to downstream.
-As one can see, for distances more than $6 rg$ in the downstream distribution does not depend on coordinate.
+As one can see, for distances more than $6 r_g$ in the downstream distribution does not depend on coordinate.
 We obtained the temperatures of protons and electrons by fitting their distibutions in far downstream with Maxwell-Juttner function, via minimization of the functional $f(T) = \int_{0}^{p_{max}} (F(p) - F_{m}(p,T))^2dp$ by temperature, where $F(p)$ is the electron or proton distribution derived from the simulation, $F_{m}(p,T)$ is the Maxwell-Juttner function with temperature $T$ and $p_{max}$ is the momentum for which the distributioin becomes non-thermal.
@@
 		\centering
-		\includegraphics[width=0.7\textwidth]{fig/electrons_at_points.png} 
-		\caption{Distributions of electrons at different regions of simulation box for turb1 parameter set}
+		\includegraphics[width=0.65\textwidth]{fig/electrons_at_points.png} 
+		\caption{The distribution functions of electrons at different positions in the shock vicinity (indicated at the insert) within the simulation box for the  turbulent parameter set (turb1).
+Note that the strong departures from the Maxwell-Juttner distributions and the formation of the non-thermal tails are due to the presence of the external turbulence in the shock upstream.}
 		\label{points}
@@
 		\centering
-		\includegraphics[width=0.7\textwidth]{fig/temperature_fit.png} 
-		\caption{Fitting the distribution for turb1 parameter set with Maxwell-Juttner distribution}
+		\includegraphics[width=0.65\textwidth]{fig/temperature_fit.png} 
+		\caption{The simulated distribution functions of electrons at far downstream.
+The red curve is for the shock propagating transverse to a regular magnetic field.
+The green curve is for the case of the turbulent upstream magnetic field (the turb1 parameter set).
+The Maxwell-Juttner fits for the both distributions are shown in the blue and black curves respectively.}
 		\label{temp_fit}
@@
 
-	Finally, a comparison of the results obtained using 2d and 3d setups with the same parameters (setups reg1, reg2 and reg3d) reveals that the differences in results depend on orientation of magnetic field in 2d simulations.
-For out-of plane orientation these differences are noticeable, but not crucial - temperature and shock velocity differs in 2d and 3d setups by less than on 10 percent.
-This means that 2d simulations are appropriate for modeling of relativistic collisionless shock waves in magnetized plasmas.
+	Finally, a comparison of the results obtained using 2D and 3D setups with the same parameters (setups reg1, reg2 and reg3d) reveals that the differences in results depend on orientation of magnetic field in 2D simulations.
+For out-of plane orientation these differences are noticeable, but not crucial - temperature and shock velocity differs in 2D and 3D setups by less than on 10 percent.
+This means that 2D simulations are appropriate for modeling of relativistic collisionless shock waves in magnetized plasmas.
 
@@
 The effective adiabatic index and electron temperatures are two important parameters required to supplement large-scale single-fluid magneto-hydrodynamic simulations of processes in astrophysical plasmas.
-Further modelling is needed to study the dependences of these two parameters  on the initial conditions which, reflect the presence of MHD large-scale turbulence in the shock upstream.
+The adiabatic indexes for the downstream flow derived with PiC method are summarized in Table 1 and can be used for MHD modeling of relativistic
+	supernovae.
+Further modelling is needed to study the dependence of these two parameters on the initial conditions which, reflect the presence of MHD large-scale turbulence in the stellar wind of the progenitor massive star upstream of the  mildly-relativistic shock of the supernova.
 

### 2020-09-06 papers/Physica2020/Romansky_Physica2020.tex
fz

@@
 These scales correspond to $150000$ and $2000$ grid points in $x$ and $y$ directions, respectively.
+For 3d setup transverse sizes are reduced to $40$ grid points.
 Also, $2000$ grid points correspond to approximately $10$ gyroradii of the upstream protons.
@@
 Also, we investigated differences between 2D and 3D cases.
+	\renewcommand{\arraystretch}{1.1}
 	\begin{table}[h!]
 		\label{setups}
+		\caption{Parameters of different setups. }
 		\begin{center}
-			\begin{tabular}{|c | c| c | c| c| c| c| c| c| c| c| c| c|}
+			\begin{tabular}{|c | c| c| c| c| c| c| c| c| c|}
 				\hline
-				Setup & Nx & Ny & Nz & $\phi$ & $\sigma$ & $\Gamma$ & $\eta$ & $\frac{v_{sh}}{c}$ & $\hat{\gamma}$ & $T \ 10^{12} K$ & $T_p\ 10^{12}K$ & $T_e \ 10^{12}K$\\
+				Setup & $\phi$ & $\sigma$ & $\Gamma$ & $\eta$ & $\frac{v_{sh}}{c}$ & $\hat{\gamma}$ & $T \ 10^{12} K$ & $T_p\ 10^{12}K$ & $T_e \ 10^{12}K$\\
 				\hline
-				reg1 & 150000 & 2000 & 1 & 0 & 0.004 & 1.5 & 0 & 0.355 & 1.51 & 3.5 & 2.5 & 0.23\\
-				reg2 & 150000 & 2000 & 1 & 90 & 0.004 & 1.5 & 0 & 0.240 & 1.53 & 2.9 & 2.2 & 0.28\\
-				reg3 & 150000 & 2000 & 1 & 0 & 0.0004 & 1.5 & 0 & 0.339 & 1.52 & 3.4 &2.4 & 0.38\\
-				reg4 & 150000 & 2000 & 1 & 90 & 0.0004 & 1.5 & 0 & 0.228 & 1.53 & 2.8 &1.7 & 0.37\\
-				turb1 & 150000 & 2000 & 1 & 0 & 0.004 & 1.5 & 90 & 0.309 & 1.52 & 3.3 & 2.1 & 0.41\\
-				turb2 & 150000 & 2000 & 1 & 0 & 0.004 & 1.5 & 30 & 0.348 & 1.51 & 3.5 & 2.4 & 0.35\\
-				gam1 & 150000 & 2000 & 1 & 0 & 0.004 & 2 & 0 & 0.412 & 1.46 & 6.3 & 4.2 & 0.62\\
-				gam2 & 150000 & 2000 & 1 & 0 & 0.004 & 5 & 0 & 0.479 & 1.39 & 20.5 & 7.3 & 11.0\\
-				gam3 & 150000 & 2000 & 1 & 0 & 0.004 & 10 & 0 & 0.504 & 1.36 & 42.6 & 16.8 & 21.3\\
-				reg3d & 80000 & 40 & 40 & 0 & 0.004 & 1.5 & 0 & 0.313 & 1.52 & 3.3 & 3.0 & 0.21\\
+				reg1  & 0 & 0.004 & 1.5 & 0 & 0.355 & 1.51 & 3.5 & 2.5 & 0.23\\
+				reg2  & 90 & 0.004 & 1.5 & 0 & 0.240 & 1.53 & 2.9 & 2.2 & 0.28\\
+				reg3 & 0 & 0.0004 & 1.5 & 0 & 0.339 & 1.52 & 3.4 &2.4 & 0.38\\
+				reg4  & 90 & 0.0004 & 1.5 & 0 & 0.228 & 1.53 & 2.8 &1.7 & 0.37\\
+				turb1  & 0 & 0.004 & 1.5 & 90 & 0.309 & 1.52 & 3.3 & 2.1 & 0.41\\
+				turb2  & 0 & 0.004 & 1.5 & 30 & 0.348 & 1.51 & 3.5 & 2.4 & 0.35\\
+				gam1  & 0 & 0.004 & 2 & 0 & 0.412 & 1.46 & 6.3 & 4.2 & 0.62\\
+				gam2  & 0 & 0.004 & 5 & 0 & 0.479 & 1.39 & 20.5 & 7.3 & 11.0\\
+				gam3  & 0 & 0.004 & 10 & 0 & 0.504 & 1.36 & 42.6 & 16.8 & 21.3\\
+				reg3d & 0 & 0.004 & 1.5 & 0 & 0.313 & 1.52 & 3.3 & 3.0 & 0.21\\
 
@@
 		\end{center}
-		\caption{Parameters of different setups. }
 	\end{table}
@@
 		\centering
-		\includegraphics[width=0.65\textwidth]{fig/shock_x.png} 
+		\includegraphics[width=0.64\textwidth]{fig/shock_x.png} 
 		\caption{Evolution of the shock front position for various parameter sets.}
@@
 		\centering
-		\includegraphics[width=0.65\textwidth]{fig/electrons_at_points.png} 
+		\includegraphics[width=0.64\textwidth]{fig/electrons_at_points.png} 
 		\caption{The distribution functions of electrons at different positions in the shock vicinity (indicated at the insert) within the simulation box for the  turbulent parameter set (turb1).
@@
 		\centering
-		\includegraphics[width=0.65\textwidth]{fig/temperature_fit.png} 
+		\includegraphics[width=0.64\textwidth]{fig/temperature_fit.png} 
 		\caption{The simulated distribution functions of electrons at far downstream.

### 2020-09-06 Adiabata/Adiabata.cpp

@@ -1,0 +1,135 @@
+// Adiabata.cpp : Defines the entry point for the console application.
+//
+#include "stdio.h"
+#include <stdlib.h>
+#include "math.h"
+
+double sqr(const double& a){
+	return a*a;
+}
+
+int main()
+{
+
+	int  i;
+	const int M=1000000;
+	const double two_pii = 6.2831853071796;
+	const double pii = 3.1415926536;
+	const double c_cgs=2.99792458E+10;
+	const double e_cgs=4.8032068E-10;
+	const double four_pii= 12.566370614359;
+	const double eight_pii= 25.132741228718; 
+	const double mp_cgs=1.6726231E-24;
+	const double sp_ht_rel=5.0/3.0;
+	const double kb_cgs=1.380658E-16;
+	const double c2_cgs       = c_cgs*c_cgs;   //! Speed of light squared, cm^2/s^2
+	const double mpc2_cgs     = mp_cgs*c2_cgs; //    ! Rest mass of proton, ergs 
+	const double mpc_cgs      = mp_cgs*c_cgs; 
+	const double Gam0=5.0/3.0; 
+	const double delta_W=2.0;
+	const double alpha_W=delta_W/(delta_W-1.0);
+	//const double InterstellarMagneticField=8.0E-3;
+	const double InterstellarMagneticField=0.0064;
+	//const double gamma_sh=1.5;
+	const double gamma0 = 10.0;
+	const double beta0 = sqrt(1.0 - 1.0/sqr(gamma0));
+	const double beta_sh = 0.504;
+	const double beta1 = (beta0 + beta_sh)/(1 + beta0*beta_sh);
+	const double gamma_sh= 1.0/sqrt(1.0 - beta1*beta1);
+	const double UpstreamNumberDensity=1.0/gamma_sh;
+	const double UpstreamFlowSpeed=c_cgs*sqrt((1.0-sqr(1.0/gamma_sh)));
[…]

### 2021-04-09 papers/Physica2021/Romansky_Physica2021.tex

--- 2020-09-06 papers/Physica2020/Romansky_Physica2020.tex
@@
 \begin{document}
-	\title{Evaluating MHD parameters of relativistic shock waves with particle-in-cell modeling}
+	\title{PIC simulation of instabilities in relativistic supernovae shocks}
 
@@
 	\begin{abstract}
-                 Relativistic plasma outflows are observed in gamma-ray burst 
-		sources, jets of active galactic nuclei, pulsar wind nebulae and
-		supernovae explosions. Magnetohydrodynamical (MHD) shock waves
-		inevitably result from interactions of such relativistic outflows with
-		the ambient interstellar matter. The widely used single-fluid MHD
-		description of relativistic shock waves is the main tool to study the
-		global structure of such objects. However, to justify the validity of 
-		the global MHD models and to interpret the observed
-		emission spectra of space objects with relativistic shocks, a kinetic
-		description of electrons, positrons, and ions at microscales is needed. 
-		We model a plane relativistic shock propagating  transverse to a regular magnetic field in the electron-ion plasmas with imposed turbulent fluctuations in the shock upstream.   
-           Namely we study  the effect of the micro-scale plasma processes on 
-		macroscopic parameters of the mildly-relativistic shocks as the adiabatic index
-		of the relativistic fluid in the shock downstream. The adiabatic index is a macroscopic parameter of the
-		single-fluid MHD models commonly used for shock modeling at much
-		longer hydrodynamical scales and it is especially important for the MHD modeling of the mildly-relativistic shocks.
+                 Modern observations show presence of relativistic supernovae - objects intermediate between regular supernovae and gamma-ray bursts.
+The typical lorentz-factors of plasma flows in relativistic supernovae are about $1.5$.
+The process of cosmic ray acceleration is very different from both non-relativistic and ultra-relativistic case.
+In this paper is presented particle-in-cell modeling of trans-relativistic collisionless shocks.
+It is shown, that development of magneto-hydrodynamic instabilities, particularly Bell's instability, has strong influence on cosmic ray acceleration.
+At times larger than $1500$ reverse ion plasma frequency magnetic field, generated by Bell's instability, significantly suppresses the electrons' spectrum.
+This effect should be taken into account for modeling of synchrothron spectrum from relativistic supernovae.
 	\end{abstract}

### 2021-04-12 papers/Physica2021/Romansky_Physica2021.tex
paper

@@
 It is shown, that development of magneto-hydrodynamic instabilities, particularly Bell's instability, has strong influence on cosmic ray acceleration.
-At times larger than $1500$ reverse ion plasma frequency magnetic field, generated by Bell's instability, significantly suppresses the electrons' spectrum.
+At times larger than $1500$ reverse ion plasma frequency magnetic field, generated by Bell's instability, significantly suppresses the electrons spectrum.
 This effect should be taken into account for modeling of synchrothron spectrum from relativistic supernovae.
@@
 
-	Relativistic fluid shock dynamics was discussed by Lichnerowicz \cite{1967rhm..book.....L}, Blandford and McKee \cite{Blandford76} and by Emmering and Chevalier\cite{Emmering87}.
-The relativistic shock jump conditions in the case of a perpendicular shock which is typical for relativisic shocks can be written as:
-	\begin{equation}\label{hugoniot}
-	[\![\rho u^r]\!] = [\![ (w + \frac{B^2}{4 \pi})u^r u^t]\!] = [\![(w + \frac{B^2}{4 \pi}){u^r}^2 + P + \frac{B^2}{8 \pi}]\!] = [\![\frac{B}{\rho}]\!] = 0
-	\end{equation}
-	where $u = (u^t,u^r)$ is the velocity four-vector measured in the shock frame, $\rho$ is the proper mass density, $B$ is the magnetic field, $P$ is the pressure in the fluid frame, and $w = \rho c^2 + \frac{\hat{\gamma}}{\hat{\gamma} - 1} P$ is the enthalpy, while $\hat{\gamma}$ is the adiabatic index and the double brackets indicate the jump of the quantities between two sides of the shock.
+	\section{Numerical setup}
+	In this work we use particle-in-cell code Smilei \cite{Smilei18} for modeling trans-relativistic collisionless shocks.
+We initialize two-dimensional space domain, with electron-proton plasma flowing into simulation box through the right boundary and reflecting wall on the left boundary.
+In the transversal direction boundary conditions are periodic.
 
-	In the ultra-relativistic limit, the shock velocity $v_{sh}$ and the downstream temperature $T$ can be expressed as, respectively, \cite{Amato2006,2011A&ARv..19...42B}:
-	\begin{equation}\label{vshock}
-	v_{sh} = \frac{c}{2}\frac{1}{1 + \sigma}\lbrace (\hat{\gamma}(1 + \frac{\sigma}{2})-1) + {\lbrack (\hat{\gamma}(1 + \frac{\sigma}{2})-1)^2 + 4\sigma(1+\sigma)(1 - \frac{\hat{\gamma}}{2})\rbrack}^{\frac{1}{2}} \rbrace,
-	\end{equation}
-	\begin{equation}\label{temperature}
-	T = m c^2 \Gamma \frac{v_{sh}}{c}(1 - \frac{\sigma}{2}\frac{c - v_{sh}}{v_sh}),
-	\end{equation}
-	Where $\sigma$ is the magnetization and $\Gamma$ is the upstream Lorentz-factor measured in the downstream frame.
+	The simulation parameters are : the initial flow Lorentz factor $\Gamma = 1.5$, the flow magnetization $\sigma = \frac{B^2}{4\pi\Gamma (n_p m_p + n_e m_e) c^2} = 0.004$.
+The dimensionless thermal energy $\Delta \gamma = \frac{k T}{m_p c^2}$ is equal to $10^{-4}$ and the electron mass is increased up to $m_e = \frac{m_p}{100}$.
+The size of the simulation box along the $x$ axis is $L_x = 30000\frac{c}{\omega_p}$ and in the transverse direction $L_y = 200\frac{c}{\omega_p}$, where $\omega_p = \sqrt{\frac{4\pi q^2 n}{\Gamma m_e}}$ is the plasma frequency.
+These scales correspond to $150000$ and $1000$ grid points in $x$ and $y$ directions, respectively.
 
-	The application of the single fluid formalism requires a correct choice of the adiabatic index $\hat{\gamma}$.
-For a monoatomic gas it is well-defined: in the classical limit $\hat{\gamma} = \frac{5}{3}$ and in the ultra-relativistic limit $\hat{\gamma} = \frac{4}{3}$.
-However, in the range between these limits the choice of the value of $\hat{\gamma}$ is not trivial.
-Moreover, the one-fluid model doesn't take into account for the difference between the temperatures of electrons and protons, while the Coulomb relaxation is slow in relativistic astrophysical plasmas.
-Two-fluid models require some phenomenological approach to account for the collisionless relaxation of the electron temperature.
-Therefore, in this study we used the particle-in-cell approach for modelling of relativistic shock waves in order to address the accuracy of MHD equations and to determine the adiabatic index for various cases.
-
-	\section{Numerical setup}
-	In this study we performed simulations of a perpendicular relativistic shock  varying the magnetization, Lorentz-factor and the fraction of the turbulent field. 
-	The simulation domain is two-dimensional with three-dimensional velocities and fields.
-To perform such computation we employed the publicly available Tristan-mp code with the explicit numerical scheme developed by Buneman \cite{Buneman93} and  Spitkovsky\cite{Spitkovsky2005}.
-
-	A shock wave is initialized in an electron-proton plasma which is flowing into the simulation box through its right boundary and then is reflecting from the super-conducting wall at the left boundary.
-The typical simulation parameters are : the initial flow Lorentz factor $\Gamma = 1.5$, the flow magnetization $\sigma = \frac{B^2}{4\pi\Gamma (n_p m_p + n_e m_e) c^2}$ was choosen to be  0.004 and 0.0004  (in the turbulent case $B^2$ is the mean square field).
-The dimensionless thermal energy $\Delta \gamma = \frac{k T}{m_p c^2}$ is equal to $10^{-4}$ and the electron mass is increased up to $m_e = \frac{m_p}{100}$.
-The size of the simulation box along the $x$ axis is $L_x = 30000\frac{c}{\omega_p}$ and in the transverse direction $L_y = 400\frac{c}{\omega_p}$, where $\omega_p = \sqrt{\frac{4\pi q^2 n}{\Gamma m_e}}$ is the plasma frequency.
-These scales correspond to $150000$ and $2000$ grid points in $x$ and $y$ directions, respectively.
-For 3d setup transverse sizes are reduced to $40$ grid points.
-Also, $2000$ grid points correspond to approximately $10$ gyroradii of the upstream protons.
-	In simulations with turbulent fields, we initialized it as  summ of harmonic modes with Kolmogorov's spectum, as described in detail in\cite{Romansky2019}.
-The ratio of the energy density of the turbulent field to the total magnetic energy density is denoted as $\eta$ , which varied from 0 percent to 90 percent in our simulations.
-The regular magnetic field is normal to the $x$ axis and is rotated by the angle $\phi$ about $x$ axis: the value $\phi = 0$ corresponds to the out-of-plane orientation of magnetic field and $\phi = 90$ corresponds to the in-plane orientation.
+	Inclination angle of magnetic field due to shock velocity is $\theta = 30$ degrees, which corresponds to the maximum efficiency of particle acceleration \cite{Romansky18} and magnetic field is lying in the simulation plane.
 
 	\section{Results}
-	We used several parameter sets, listed in table \ref{setups}, with different magnetization, Lorentz-factors and the fraction of turbulent field.
-Also, we investigated differences between 2D and 3D cases.
-	\renewcommand{\arraystretch}{1.1}
-	\begin{table}[h!]
-		\label{setups}
-		\caption{Parameters of different setups. }
-		\begin{center}
-			\begin{tabular}{|c | c| c| c| c| c| c| c| c| c|}
-				\hline
-				Setup & $\phi$ & $\sigma$ & $\Gamma$ & $\eta$ & $\frac{v_{sh}}{c}$ & $\hat{\gamma}$ & $T \ 10^{12} K$ & $T_p\ 10^{12}K$ & $T_e \ 10^{12}K$\\
-				\hline
-				reg1  & 0 & 0.004 & 1.5 & 0 & 0.355 & 1.51 & 3.5 & 2.5 & 0.23\\
-				reg2  & 90 & 0.004 & 1.5 & 0 & 0.240 & 1.53 & 2.9 & 2.2 & 0.28\\
-				reg3 & 0 & 0.0004 & 1.5 & 0 & 0.339 & 1.52 & 3.4 &2.4 & 0.38\\
-				reg4  & 90 & 0.0004 & 1.5 & 0 & 0.228 & 1.53 & 2.8 &1.7 & 0.37\\
-				turb1  & 0 & 0.004 & 1.5 & 90 & 0.309 & 1.52 & 3.3 & 2.1 & 0.41\\
-				turb2  & 0 & 0.004 & 1.5 & 30 & 0.348 & 1.51 & 3.5 & 2.4 & 0.35\\
-				gam1  & 0 & 0.004 & 2 & 0 & 0.412 & 1.46 & 6.3 & 4.2 & 0.62\\
-				gam2  & 0 & 0.004 & 5 & 0 & 0.479 & 1.39 & 20.5 & 7.3 & 11.0\\
-				gam3  & 0 & 0.004 & 10 & 0 & 0.504 & 1.36 & 42.6 & 16.8 & 21.3\\
-				reg3d & 0 & 0.004 & 1.5 & 0 & 0.313 & 1.52 & 3.3 & 3.0 & 0.21\\
 
-				\hline
-			\end{tabular}
-		\end{center}
-	\end{table}
-
-	Using the density profiles of the simulated shock waves, we determined magnetohydrodynamic parameters of plasma, solving the jump conditions (\ref{hugoniot}).
-The density averaged in transversal directions has clear jump as shown in figures \ref{density_noturb} and \ref{density_turb}, which is indicating the position of the shock front.
-So we can evaluate the shock velocity from results of simulations, and then solve jump conditions for the adiabatic index and the downstream temperature, using the iterative method.
+	Profiles of transversal magnetic field, obtained from numerical simulation, show significant field amplification in precursor of shock wave, see figure \ref{field}.
+It is result of development of instabilities in plasma, particularly Bell instability \cite{Bell04}, which grows rate is larger then grows rate of other instabilities.
 
@@
 		\centering
-		\begin{minipage}{0.49\textwidth}
-			\centering
-			\includegraphics[width=0.98\textwidth]{fig/density_noturb.png} 
-			\caption{Density profile of a shock wave in a medium with a regular field, averaged through the transversal dimensions, setup reg1.}
-			\label{density_noturb}
-		\end{minipage}\hfill
-		\begin{minipage}{0.49\textwidth}
-			\centering
-			\includegraphics[width=0.98\textwidth]{fig/density_turb.png} 
-			\caption{Density profile of a shock wave in a turbulent medium, averaged through the transversal dimensions, setup turb1.}
-			\label{density_turb}
-		\end{minipage}
+		\includegraphics[width=0.8\textwidth]{fig/Bnorm.png} 
+		\caption{Evolution of the magnetic field in time.}
+		\label{field}
 	\end{figure}
 
-	Evaluation of shock front coordinates, which we have determined as a point where  the density is two times larger than the density observed far upstream, at different time steps shows that the shock front velocity is constant in time up to a high accuracy, see figure \ref{shock_x}.
-Also, it shows, that the shock front velocity decreases with the decreasing magnetization, and, consequently, the adiabatic index does not remain constant.
-The shock slows down with the increasing turbulent field energy and it requires significant decrease of the adiabatic index.
-It can be explained by the appearence of new degrees of freedom related to magnetic field oscillations.
-Obviously, the shock front is faster for higher values of the upstream Lorentz-factor, and the adiabatic index  converges to the value $4/3$ which is the ultra-relativistic limit.
-For Lorentz-factors less then 10, $\hat{\gamma}$ is significantly higher than the limiting value.
+	Spectrum of electrons is shown in figure \ref{spectrum}.
+One can see, that electrons distribution function at high energies $(\gamma > 300)$ is maximal at times about 1500 inverse protons plasma frequencies and at the longer times spectrum decreases.
+It may by explained as consequence of magnetic field amplification.
+Instabilities increase the transverse magnetic field and also the angle between field and shock velocity.
+As a result shock becomes superluminal \cite{Sironi2011} and can not effectively accelerate particles.
 
@@
 		\centering
-		\includegraphics[width=0.64\textwidth]{fig/shock_x.png} 
-		\caption{Evolution of the shock front position for various parameter sets.}
-		\label{shock_x}
+		\includegraphics[width=0.8\textwidth]{fig/spectrum.png} 
+		\caption{Evolution of the electrons spectrum in time.}
+		\label{spectrum}
 	\end{figure}
-
-	Distributions of particles significantly vary through the x coordinate, and in figure \ref{points} we show the distribution of electrons at different distances from shock front, measured in proton gyroradius $r_g$.
-Positive distances correspond to upstream region and negative to downstream.
-As one can see, for distances more than $6 r_g$ in the downstream distribution does not depend on coordinate.
-We obtained the temperatures of protons and electrons by fitting their distibutions in far downstream with Maxwell-Juttner function, via minimization of the functional $f(T) = \int_{0}^{p_{max}} (F(p) - F_{m}(p,T))^2dp$ by temperature, where $F(p)$ is the electron or proton distribution derived from the simulation, $F_{m}(p,T)$ is the Maxwell-Juttner function with temperature $T$ and $p_{max}$ is the momentum for which the distributioin becomes non-thermal.
-Obtained temperature has rather week dependence on choice of $p_{max}$.
-Such evaluation of the downstream temperature is possible even in the cases of strong non-thermal tail, as shown in figure \ref{temp_fit}.
-
-	\begin{figure}[h!]
-		\centering
-		\includegraphics[width=0.64\textwidth]{fig/electrons_at_points.png} 
-		\caption{The distribution functions of electrons at different positions in the shock vicinity (indicated at the insert) within the simulation box for the  turbulent parameter set (turb1).
-Note that the strong departures from the Maxwell-Juttner distributions and the formation of the non-thermal tails are due to the presence of the external turbulence in the shock upstream.}
-		\label{points}
-	\end{figure}
-
-	\begin{figure}[h!]
-		\centering
-		\includegraphics[width=0.64\textwidth]{fig/temperature_fit.png} 
-		\caption{The simulated distribution functions of electrons at far downstream.
-The red curve is for the shock propagating transverse to a regular magnetic field.
-The green curve is for the case of the turbulent upstream magnetic field (the turb1 parameter set).
-The Maxwell-Juttner fits for the both distributions are shown in the blue and black curves respectively.}
-		\label{temp_fit}
-	\end{figure}
-
-	An analysis of the distribution functions shows that the temperatures of protons and electrons are much smaller than predicted by jump conditions (\ref{hugoniot}).
-It means that magnetohydrodynamical formalism is not sufficient for the description of downstream temperatures of a relativistic shock wave.
-One can notice that the ratio of temperature of protons to the temperatures of electrons is not constant: it is close to $\sqrt{\frac{m_p}{m_e}}$ at low Lorenz-factors and decreases down to $1$ at higher ones.
-Turbulence also has a strong influence on the temperatures, as it makes electrons much hotter, while the protons become colder.
-
-	Finally, a comparison of the results obtained using 2D and 3D setups with the same parameters (setups reg1, reg2 and reg3d) reveals that the differences in results depend on orientation of magnetic field in 2D simulations.
-For out-of plane orientation these differences are noticeable, but not crucial - temperature and shock velocity differs in 2D and 3D setups by less than on 10 percent.
-This means that 2D simulations are appropriate for modeling of relativistic collisionless shock waves in magnetized plasmas.
 

### 2021-04-14 papers/Physica2021/Romansky_Physica2021.tex
paper

@@
 In this paper is presented particle-in-cell modeling of trans-relativistic collisionless shocks.
-It is shown, that development of magneto-hydrodynamic instabilities, particularly Bell's instability, has strong influence on cosmic ray acceleration.
+It is shown, that development of magneto-hydrodynamic instabilities ii electron-ion relativistic shock, particularly Bell's instability, has strong influence on cosmic ray acceleration.
 At times larger than $1500$ reverse ion plasma frequency magnetic field, generated by Bell's instability, significantly suppresses the electrons spectrum.
@@
 	\section{Introduction}
-	Relativistic jets and winds are the generic outflows of the relativistic supernovae \cite{2010Natur.463..513S,2007ApJ...667..351W}, the accreting black holes, both suprmassive in the active galactic nucleis  \cite{1984RvMP...56..255B} as well as the stellar masses in the gamma-ray burst sources, microquasars \cite{2019MmSAI..90...57M,1999PhR...314..575P,2014LNP...876.....R} and pulsar wind nebulae \cite{2019MNRAS.488.5690O,2017SSRv..207..235B,2017JPlPh..83e6301K,2019ApJ...876L...8B}.
-Relativistic shocks can be formed either inside the flow of low or moderate magnetization (the internal shocks) or can be driven by the outflow colliding with the ambient medium (the external shocks).
-In many cases the relatistivistic shocks in astrophysical objects are collisionless and are often associated with non-equilibrium particle distributions \cite{2012SSRv..173..309B,2015SSRv..191..519S,2017SSRv..207..319P}.
-They can accelerate particles to ultra-high energies \cite{2009JCAP...11..009L}.
-On the one hand, the presence of non-maxwellian components with very long equilibration time requires a kinetic description of the shocked flows.
-The kinetic structure of relativistic shocks can be modelled within the particle-in-cell approach, which allows one to properly account for the electron scale processes and for the density of the displacement current which is important in relativistic plasma flows.
-On the other hand, available computer capabilities do not alow one to provide a fully kinetic modelling of the macroscopic astrophysical flows and, therefore, the fluid MHD description based on such macroscopic parameters as the effective adiabatic index of the flow is widely used.
-Here we discuss an approach allowing to determine the effective adiabatic indexes from the microscopical particle-in-cell modeling of the mildly-relativistic shocks typical for the relativistic supernovae and gamma-ray burst afterglows.
+	Relativistic shocks are widely discussed phenomena in modern astrophysicist.
+They can be generated in the supernovae \cite{2010Natur.463..513S,2007ApJ...667..351W}, the accreting supermassive black holes \cite{1984RvMP...56..255B}, the stellar masses black holes \cite{2019MmSAI..90...57M,1999PhR...314..575P,2014LNP...876.....R} and pulsar wind nebulae \cite{2019MNRAS.488.5690O,2017SSRv..207..235B,2017JPlPh..83e6301K,2019ApJ...876L...8B}.
+Relativistic shocks in astrophysical objects are sources of cosmic rays \cite{2012SSRv..173..309B,2015SSRv..191..519S,2017SSRv..207..319P} and can accelerate particles to ultra-high energies \cite{2009JCAP...11..009L}.
+
+	The presence of high energy non-thermal particles leads to development of various magneto-hydrodynamic instabilities.
+Bell instability \cite{Bell04} can be the most significant due to it's large growth rate.
+Instabilities modify the flow and has strong influence on process of particle acceleration.
+Below we present results of particle-in-cell modeling of relativistic shock and show that growth of magnetic field due to instabilities decreases efficiency of electron acceleration on shock front.
 
@@
 	Profiles of transversal magnetic field, obtained from numerical simulation, show significant field amplification in precursor of shock wave, see figure \ref{field}.
-It is result of development of instabilities in plasma, particularly Bell instability \cite{Bell04}, which grows rate is larger then grows rate of other instabilities.
+It is result of development of instabilities in plasma, particularly Bell instability \cite{Bell04}, which growth rate is larger then growth rate of other instabilities.
 
@@
 
-	Spectrum of electrons is shown in figure \ref{spectrum}.
+	Spectrum of electrons is shown in figure \ref{spectrume}.
 One can see, that electrons distribution function at high energies $(\gamma > 300)$ is maximal at times about 1500 inverse protons plasma frequencies and at the longer times spectrum decreases.
@@
 Instabilities increase the transverse magnetic field and also the angle between field and shock velocity.
-As a result shock becomes superluminal \cite{Sironi2011} and can not effectively accelerate particles.
+As a result shock becomes quasi-perpendicular and can not effectively accelerate particles \cite{Sironi2011}.
+On the other hand, protons distribution function constantly increases in time, as shown in figure \ref{spectrump}.
+We assume, that oscillating transverse field has the strongest influence on the smallest scales, and suppresses he electrons injection in acceleration process.
 
@@
 		\centering
-		\includegraphics[width=0.8\textwidth]{fig/spectrum.png} 
-		\caption{Evolution of the electrons spectrum in time.}
-		\label{spectrum}
+		\begin{minipage}{0.8\textwidth}
+			\centering
+			\includegraphics[width=0.98\textwidth]{fig/spectrum.png} 
+			\caption{Temporal evolution of electrons spectrum.}
+			\label{spectrume}
+		\end{minipage}\hfill
+		\begin{minipage}{0.8\textwidth}
+			\centering
+			\includegraphics[width=0.98\textwidth]{fig/spectrump.png} 
+			\caption{Temporal evolution of protons spectrum.}
+			\label{spectrump}
+		\end{minipage}
 	\end{figure}

### 2021-04-16 papers/Physica2021/Romansky_Physica2021.tex
paper

@@
 	\begin{abstract}
-                 Modern observations show presence of relativistic supernovae - objects intermediate between regular supernovae and gamma-ray bursts.
-The typical lorentz-factors of plasma flows in relativistic supernovae are about $1.5$.
-The process of cosmic ray acceleration is very different from both non-relativistic and ultra-relativistic case.
-In this paper is presented particle-in-cell modeling of trans-relativistic collisionless shocks.
-It is shown, that development of magneto-hydrodynamic instabilities ii electron-ion relativistic shock, particularly Bell's instability, has strong influence on cosmic ray acceleration.
-At times larger than $1500$ reverse ion plasma frequency magnetic field, generated by Bell's instability, significantly suppresses the electrons spectrum.
-This effect should be taken into account for modeling of synchrothron spectrum from relativistic supernovae.
+                 Radio observations revealed a presence of relativistic supernovae a class of objects intermediate between regular supernovae and gamma-ray bursts.
+The typical Lorentz-factors of plasma flows in relativistic radio-bright supernovae were estimated to be about 1.5.
+Mildly relativistic shocks in electron-ion plasmas are known to efficiently accelerate radio-emitting electrons if the shock is subluminous.
+The inclination angle of the subluminous shock to the ambient magnetic field should be below a critical angle which depends on the Mach number and the plasma magnetization parameter.
+In this paper we present particle-in-cell modeling of electron acceleration by a subluminal mildly-relativistic collisionless shock.
+It is shown that  a development of the ion scale Bell-type instability in electron-ion relativistic shock may have a strong influence on the electron injection and acceleration.
+In time period of about $1500 \omega_{pi}^{-1}$  ($\omega_{pi}$ is the ion plasma frequency) after the shock initialization the  magnetic field fluctuations  generated by Bell's instability may significantly suppresses the electrons spectrum even in a sub-luminous shock.
+This effect is important to account for modeling of synchrothron spectra from relativistic supernovae.
 	\end{abstract}
@@
 	\section{Introduction}
-	Relativistic shocks are widely discussed phenomena in modern astrophysicist.
+	Relativistic shocks are widely discussed phenomena in modern astrophysics.
 They can be generated in the supernovae \cite{2010Natur.463..513S,2007ApJ...667..351W}, the accreting supermassive black holes \cite{1984RvMP...56..255B}, the stellar masses black holes \cite{2019MmSAI..90...57M,1999PhR...314..575P,2014LNP...876.....R} and pulsar wind nebulae \cite{2019MNRAS.488.5690O,2017SSRv..207..235B,2017JPlPh..83e6301K,2019ApJ...876L...8B}.
@@
 Instabilities modify the flow and has strong influence on process of particle acceleration.
-Below we present results of particle-in-cell modeling of relativistic shock and show that growth of magnetic field due to instabilities decreases efficiency of electron acceleration on shock front.
+We present below results of particle-in-cell modeling of relativistic shock and show that growth of magnetic field due to instabilities decreases efficiency of electron acceleration on shock front.
 
@@
 
-	Inclination angle of magnetic field due to shock velocity is $\theta = 30$ degrees, which corresponds to the maximum efficiency of particle acceleration \cite{Romansky18} and magnetic field is lying in the simulation plane.
+	Inclination angle of magnetic field due to shock velocity varies in different setups from $\theta = 20$ degrees to $\theta = 40$.
+This values are chosen to contain critical angle, which is defined by the equation $c\cdot \cos(\theta_{crit})=v_{shock}$, where all values are measured in the upstream rest frame, and is equal approximately $30$ degrees.
+If inclination angle is larger then critical value, shock becomes superluminal - particle can not escape from it to upstream, even moving along the field lines with speed of light \cite{Pelletier2017,Sironi2011}.
+The maximum efficiency of particle acceleration is near the critical angle, less below it \cite{Romansky18}.
+Magnetic field is lying in the simulation plane.
 
@@
 	Profiles of transversal magnetic field, obtained from numerical simulation, show significant field amplification in precursor of shock wave, see figure \ref{field}.
-It is result of development of instabilities in plasma, particularly Bell instability \cite{Bell04}, which growth rate is larger then growth rate of other instabilities.
+It is result of development of instabilities in plasma, particularly Bell's instability \cite{Bell04}, which growth rate is larger then growth rate of other instabilities.
 
@@
 		\includegraphics[width=0.8\textwidth]{fig/Bnorm.png} 
-		\caption{Evolution of the magnetic field in time.}
+		\caption{Temporal evolution of the transversal magnetic field.}
 		\label{field}
@@
 
+	It is known, that development of Bell's instability can suppress electrons acceleration \cite{Crumley2019}.
+We study this phenomena with different inclination angles of magnetic field.
 Spectrum of electrons is shown in figure \ref{spectrume}.
-One can see, that electrons distribution function at high energies $(\gamma > 300)$ is maximal at times about 1500 inverse protons plasma frequencies and at the longer times spectrum decreases.
+One can see, that electrons distribution function at high energies $(\gamma > 300)$ is maximal at times about 1400 inverse protons plasma frequencies and at the longer times spectrum decreases.
 It may by explained as consequence of magnetic field amplification.
 Instabilities increase the transverse magnetic field and also the angle between field and shock velocity.
-As a result shock becomes quasi-perpendicular and can not effectively accelerate particles \cite{Sironi2011}.
+As a result shock becomes quasi-perpendicular and can not effectively accelerate particles \cite{Sironi2011,Romansky18}.
 On the other hand, protons distribution function constantly increases in time, as shown in figure \ref{spectrump}.
-We assume, that oscillating transverse field has the strongest influence on the smallest scales, and suppresses he electrons injection in acceleration process.
+We assume, that oscillating transverse field has the strongest influence on the smallest scales, and suppresses the electrons injection into acceleration process.
 
@@
 		\centering
-		\begin{minipage}{0.8\textwidth}
+		\begin{minipage}{0.49\textwidth}
 			\centering
-			\includegraphics[width=0.98\textwidth]{fig/spectrum.png} 
-			\caption{Temporal evolution of electrons spectrum.}
+			\includegraphics[width=0.95\textwidth]{fig/spectrum20.png} 
+		\end{minipage}\hfill
+		\begin{minipage}{0.49\textwidth}
+			\centering
+			\includegraphics[width=0.95\textwidth]{fig/spectrum30.png} 
+		\end{minipage}
+		\begin{minipage}{0.49\textwidth}
+			\centering
+			\includegraphics[width=0.95\textwidth]{fig/spectrum33.png} 
+		\end{minipage}
+		\begin{minipage}{0.49\textwidth}
+			\centering
+			\includegraphics[width=0.95\textwidth]{fig/spectrum40.png} 
+		\end{minipage}
+		\caption{Temporal evolution of electrons spectrum in shocks with different inclination angles.
+Top row: left $\theta = 20^\circ$, right $\theta = 25^\circ$, bottom row: left $\theta = 33^\circ$, right $\theta = 40^\circ$}
 		\label{spectrume}
+	\end{figure}
+
+\begin{figure}[h!]
+	\centering
+	\begin{minipage}{0.49\textwidth}
+		\centering
+		\includegraphics[width=0.95\textwidth]{fig/spectrump20.png} 
 	\end{minipage}\hfill
-		\begin{minipage}{0.8\textwidth}
+	\begin{minipage}{0.49\textwidth}
 		\centering
-			\includegraphics[width=0.98\textwidth]{fig/spectrump.png} 
-			\caption{Temporal evolution of protons spectrum.}
+		\includegraphics[width=0.95\textwidth]{fig/spectrump30.png} 
+	\end{minipage}
+	\begin{minipage}{0.49\textwidth}
+		\centering
+		\includegraphics[width=0.95\textwidth]{fig/spectrump33.png} 
+	\end{minipage}
+	\begin{minipage}{0.49\textwidth}
+		\centering
+		\includegraphics[width=0.95\textwidth]{fig/spectrump40.png} 
+	\end{minipage}
+	\caption{Temporal evolution of protons spectrum in shocks with different inclination angles.
+Top row: left $\theta = 20^\circ$, right $\theta = 25^\circ$, bottom row: left $\theta = 33^\circ$, right $\theta = 40^\circ$}
 	\label{spectrump}
-		\end{minipage}
 \end{figure}
@@
 
-	Particle-in-cell simulations have been used to derive effective macroscopic parameters of relativistic shocks in space plasmas.
-The effective adiabatic index and electron temperatures are two important parameters required to supplement large-scale single-fluid magneto-hydrodynamic simulations of processes in astrophysical plasmas.
-The adiabatic indexes for the downstream flow derived with PiC method are summarized in Table 1 and can be used for MHD modeling of relativistic
-	supernovae.
-Further modelling is needed to study the dependence of these two parameters on the initial conditions which, reflect the presence of MHD large-scale turbulence in the stellar wind of the progenitor massive star upstream of the  mildly-relativistic shock of the supernova.
+	Details of electrons acceleration are important for correct interpretations of observations synchrothron radiation from astrophysical sources.
+Particle-in-cell simulation has been used to demonstrate magnetic instabilities influence on electrons acceleration in relativistic electron-ion shock.
+It is shown, that amplification of magnetic field suppresses formation of electrons spectrum at high energies.This effect should be taken into account while deriving parameters of source of radiation via analysis of synchrothron spectrum.
+Further PIC simulations with a wider spatial and energetical ranges will allow to study the effect more precisely.
 

### 2021-04-22 papers/Physica2021/Romansky_Physica2021.tex
paper

@@
 \begin{document}
-	\title{PIC simulation of instabilities in relativistic supernovae shocks}
+	\title{On electron acceleration  by mildly-relativistic shocks: PIC simulations}
 
@@
 	\begin{abstract}
-                 Radio observations revealed a presence of relativistic supernovae a class of objects intermediate between regular supernovae and gamma-ray bursts.
-The typical Lorentz-factors of plasma flows in relativistic radio-bright supernovae were estimated to be about 1.5.
+                 Radio observations revealed a presence of relativistic supernovae - a class of objects intermediate between the regular supernovae and gamma-ray bursts.
+The typical Lorentz-factors of plasma flows in relativistic radio-bright supernovae were estimated to be about $1.5$.
 Mildly relativistic shocks in electron-ion plasmas are known to efficiently accelerate radio-emitting electrons if the shock is subluminous.
-The inclination angle of the subluminous shock to the ambient magnetic field should be below a critical angle which depends on the Mach number and the plasma magnetization parameter.
-In this paper we present particle-in-cell modeling of electron acceleration by a subluminal mildly-relativistic collisionless shock.
-It is shown that  a development of the ion scale Bell-type instability in electron-ion relativistic shock may have a strong influence on the electron injection and acceleration.
-In time period of about $1500 \omega_{pi}^{-1}$  ($\omega_{pi}$ is the ion plasma frequency) after the shock initialization the  magnetic field fluctuations  generated by Bell's instability may significantly suppresses the electrons spectrum even in a sub-luminous shock.
-This effect is important to account for modeling of synchrothron spectra from relativistic supernovae.
+The inclination angle of the velocity of subluminous shock to the ambient magnetic field should be below a critical angle which depends on the Mach number and the plasma magnetization parameter.
+In this paper we present particle-in-cell modeling of electron acceleration by mildly-relativistic collisionless shock of different obliquity in a plasma with magnetization parameter $\sigma \approx 0.004$ which is of interest for the relativistic supernovae modeling.
+It was shown earlier that  a development of the ion scale Bell-type instability in electron-ion relativistic shock may have a strong influence on the electron injection and acceleration.
+In the time period of about $1500 \omega_{pi}^{-1}$  ($\omega_{pi}$ is the ion plasma frequency) after the shock initialization the  magnetic field fluctuations  generated by Bell's instability may significantly suppresses the electrons spectrum even in a sub-luminous shock.
+We study here the evolution of the electron spectra of subluminous shocks of different obliquity.
+This is important to for modeling of synchrothron spectra from relativistic supernovae.
 	\end{abstract}
@@
 	\section{Introduction}
-	Relativistic shocks are widely discussed phenomena in modern astrophysics.
-They can be generated in the supernovae \cite{2010Natur.463..513S,2007ApJ...667..351W}, the accreting supermassive black holes \cite{1984RvMP...56..255B}, the stellar masses black holes \cite{2019MmSAI..90...57M,1999PhR...314..575P,2014LNP...876.....R} and pulsar wind nebulae \cite{2019MNRAS.488.5690O,2017SSRv..207..235B,2017JPlPh..83e6301K,2019ApJ...876L...8B}.
-Relativistic shocks in astrophysical objects are sources of cosmic rays \cite{2012SSRv..173..309B,2015SSRv..191..519S,2017SSRv..207..319P} and can accelerate particles to ultra-high energies \cite{2009JCAP...11..009L}.
+	Relativistic shocks are playing an important role  in modeling of luminous high energy objects in  astrophysics.
+They can be generated in the supernovae \cite{2010Natur.463..513S,2007ApJ...667..351W}, the accreting supermassive black holes \cite{1984RvMP...56..255B}, the stellar masses black holes \cite{2019MmSAI..90...57M,1999PhR...314..575P,2014LNP...876.....R} and pulsar wind nebulae \cite{Amato2006,2017SSRv..207..235B,2017JPlPh..83e6301K,2019ApJ...876L...8B}.
+Relativistic shocks in astrophysical objects are sources of cosmic rays \cite{2012SSRv..173..309B,2015SSRv..191..519S,Pelletier2017} and can accelerate particles to ultra-high energies \cite{Wang2007,2009JCAP...11..009L,BEMO18}.
 
-	The presence of high energy non-thermal particles leads to development of various magneto-hydrodynamic instabilities.
-Bell instability \cite{Bell04} can be the most significant due to it's large growth rate.
-Instabilities modify the flow and has strong influence on process of particle acceleration.
-We present below results of particle-in-cell modeling of relativistic shock and show that growth of magnetic field due to instabilities decreases efficiency of electron acceleration on shock front.
+	The presence of high energy non-thermal particles leads to development of various magneto-hydrodynamic instabilities (see e.g.
+\cite{MD2001,Amato09,Bykov2013}).
+Bell's instability \cite{Bell04} can be the most significant due to it's fast growth rate.
+Instabilities may modify the flow and has strong influence on the process of particle acceleration \cite{Crumley2019,2021MNRAS.501.4837L,2021MNRAS.502.5065L}.
+Namely, it was realized that Bell's instability is producing the strong ion-scale magnetic fluctuations which
+	provide on the electron scales some macroscopic size regions of the superluminal shock obliquity which locally suppress the electron injection.
+The shock rippling effect is important for the electron spectra formation in the mildly-relativistic shocks.
+	In the simulations \cite{Crumley2019} the electron spectra evolution was demonstrated in a mildly-relativistic shock   of an inclination angle 10 degrees the magnetic fluctuations produced by the Weibel instability initially efficiently reflect and energize both ions and electrons and after the time period of about $1500 \omega_{pi}^{-1}$  the efficiency of electron acceleration by the Bell-mediated shock drops down significantly.
+The proton spectrum in this simulations was not strongly affected in the Bell-mediated shock.
+Given the very serious practical interest of non-thermal particle acceleration by the subluminal mildly-relativistic shocks we present below results of particle-in-cell modeling of electron and proton acceleration by the mildly-relativistic shocks of different inclination angles.
 
 	\section{Numerical setup}
-	In this work we use particle-in-cell code Smilei \cite{Smilei18} for modeling trans-relativistic collisionless shocks.
+	In this work we use particle-in-cell code Smilei \cite{Smilei18} for modeling mildly-relativistic collisionless shocks.
 We initialize two-dimensional space domain, with electron-proton plasma flowing into simulation box through the right boundary and reflecting wall on the left boundary.
@@
 
-	Inclination angle of magnetic field due to shock velocity varies in different setups from $\theta = 20$ degrees to $\theta = 40$.
+	Inclination angle of magnetic field due to shock velocity varies in different setups from $\theta = 20^\circ$ to $\theta = 40^\circ$.
 This values are chosen to contain critical angle, which is defined by the equation $c\cdot \cos(\theta_{crit})=v_{shock}$, where all values are measured in the upstream rest frame, and is equal approximately $30$ degrees.
@@
 		\centering
-		\includegraphics[width=0.8\textwidth]{fig/Bnorm.png} 
+		\includegraphics[width=0.50\textwidth]{fig/Bnorm.png} 
 		\caption{Temporal evolution of the transversal magnetic field.}
@@
 
-	It is known, that development of Bell's instability can suppress electrons acceleration \cite{Crumley2019}.
+	It is known, that development of Bell's instability can suppress electrons acceleration \cite{Crumley2019,2021MNRAS.501.4837L,2021MNRAS.502.5065L}.
 We study this phenomena with different inclination angles of magnetic field.
 Spectrum of electrons is shown in figure \ref{spectrume}.
-One can see, that electrons distribution function at high energies $(\gamma > 300)$ is maximal at times about 1400 inverse protons plasma frequencies and at the longer times spectrum decreases.
+One can see, that electrons acceleration has strong dependence on inclination angle, and when it is more then critical (at $\theta = 40^\circ$), non-thermal component almost disappear.
+For smaller angles electrons distribution function at high energies $(\Gamma > 300)$ depends on time, and it is maximal at times about 1400 inverse protons plasma frequencies for angles $\theta=20^\circ, \theta=30^\circ$, and 2100 and inverse protons plasma frequencies for $\theta = 33^\circ$.
+At the longer times electron spectrum decreases.
 It may by explained as consequence of magnetic field amplification.
@@
 As a result shock becomes quasi-perpendicular and can not effectively accelerate particles \cite{Sironi2011,Romansky18}.
-On the other hand, protons distribution function constantly increases in time, as shown in figure \ref{spectrump}.
-We assume, that oscillating transverse field has the strongest influence on the smallest scales, and suppresses the electrons injection into acceleration process.
+On the other hand, protons distribution function constantly increases in time for sub-critical angles, and almost does not has non-thermal component for $\theta = 40^\circ$, as shown in figure \ref{spectrump}.
+We assume, that oscillating transverse field has the strongest influence on the smallest scales, and suppresses the electrons injection into acceleration process, while protons still can be accelerated.
+Also, one can see, that temporal evolutions of electrons and protons spectrums are correlated.
+In case of inclination angle $\theta = 33^\circ$ protons distribution function grows slower then for smaller angles.
+And electrons distribution in this case is maximal at larger time.
+This correlation is the confirmation of assumption, that instability suppressing electron acceleration is caused by accelerated protons.
 
@@
 			\centering
-			\includegraphics[width=0.95\textwidth]{fig/spectrum20.png} 
+			\includegraphics[width=0.98\textwidth]{fig/spectrum20.png} 
 		\end{minipage}\hfill
@@
 			\centering
-			\includegraphics[width=0.95\textwidth]{fig/spectrum30.png} 
+			\includegraphics[width=0.98\textwidth]{fig/spectrum30.png} 
 		\end{minipage}
@@
 			\centering
-			\includegraphics[width=0.95\textwidth]{fig/spectrum33.png} 
+			\includegraphics[width=0.98\textwidth]{fig/spectrum33.png} 
 		\end{minipage}
@@
 			\centering
-			\includegraphics[width=0.95\textwidth]{fig/spectrum40.png} 
+			\includegraphics[width=0.98\textwidth]{fig/spectrum40.png} 
 		\end{minipage}
-		\caption{Temporal evolution of electrons spectrum in shocks with different inclination angles.
-Top row: left $\theta = 20^\circ$, right $\theta = 25^\circ$, bottom row: left $\theta = 33^\circ$, right $\theta = 40^\circ$}
+		\caption{Temporal evolution of electrons spectrum in shocks with different magnetic field inclination angles.
+Top row: left $\theta = 20^\circ$, right $\theta = 30^\circ$, bottom row: left $\theta = 33^\circ$, right $\theta = 40^\circ$}
 		\label{spectrume}
@@
 		\centering
-		\includegraphics[width=0.95\textwidth]{fig/spectrump20.png} 
+		\includegraphics[width=0.98\textwidth]{fig/spectrump20.png} 
 	\end{minipage}\hfill
@@
 		\centering
-		\includegraphics[width=0.95\textwidth]{fig/spectrump30.png} 
+		\includegraphics[width=0.98\textwidth]{fig/spectrump30.png} 
 	\end{minipage}
@@
 		\centering
-		\includegraphics[width=0.95\textwidth]{fig/spectrump33.png} 
+		\includegraphics[width=0.98\textwidth]{fig/spectrump33.png} 
 	\end{minipage}
@@
 		\centering
-		\includegraphics[width=0.95\textwidth]{fig/spectrump40.png} 
+		\includegraphics[width=0.98\textwidth]{fig/spectrump40.png} 
 	\end{minipage}
-	\caption{Temporal evolution of protons spectrum in shocks with different inclination angles.
-Top row: left $\theta = 20^\circ$, right $\theta = 25^\circ$, bottom row: left $\theta = 33^\circ$, right $\theta = 40^\circ$}
+	\caption{Temporal evolution of protons spectrum in shocks with different magnetic field inclination angles.
+Top row: left $\theta = 20^\circ$, right $\theta = 30^\circ$, bottom row: left $\theta = 33^\circ$, right $\theta = 40^\circ$}
 	\label{spectrump}
@@
 	Details of electrons acceleration are important for correct interpretations of observations synchrothron radiation from astrophysical sources.
-Particle-in-cell simulation has been used to demonstrate magnetic instabilities influence on electrons acceleration in relativistic electron-ion shock.
-It is shown, that amplification of magnetic field suppresses formation of electrons spectrum at high energies.This effect should be taken into account while deriving parameters of source of radiation via analysis of synchrothron spectrum.
-Further PIC simulations with a wider spatial and energetical ranges will allow to study the effect more precisely.
+Particle-in-cell simulation has been used to demonstrate magnetic instabilities influence on electrons acceleration in mildly-relativistic electron-ion shock.
+It is shown, that amplification of magnetic field suppresses formation of electrons spectrum at high energies while protons spectrum can grow monotonously in time in subluminal shock.
+Also, we demonstrated that grows of proton spectrum is correlated with decreasing of electrons acceleration, and resulting  electrons spectrum strongly depends on magnetic field inclination angle.
+This effect should be taken into account for deriving parameters of source of synchrothron radiation via analysis of spectrum.
+Further particle-in-cell simulations with a wider spatial and energetical ranges will allow to study the influence of instabilities on electrons spectrum more precisely.
 
 	\ack
-	V I Romansky and A M Bykov acknowledge a support from RSF grant 16-12-10225.
-	Results of the work were obtained using computational resources of Peter the Great Saint-Petersburg Polytechnic University Supercomputing Center (http://scc.spbstu.ru)
+	The authors were supported by RSF grant 21-72-20020.
+	Some results of the work were obtained using computational resources of Peter the Great Saint-Petersburg Polytechnic University Supercomputing Center (http://scc.spbstu.ru)
 

### 2021-07-04 papers/Physica2021/Romansky_Physica2021.tex
paper

@@
 The inclination angle of the velocity of subluminous shock to the ambient magnetic field should be below a critical angle which depends on the Mach number and the plasma magnetization parameter.
-In this paper we present particle-in-cell modeling of electron acceleration by mildly-relativistic collisionless shock of different obliquity in a plasma with magnetization parameter $\sigma \approx 0.004$ which is of interest for the relativistic supernovae modeling.
+In this paper we present particle-in-cell modeling of electron acceleration by mildly-relativistic collisionless shock of different obliquity in a plasma with ratio of the magnetic energy to the bulk kinetic energy $\sigma \approx 0.004$ which is of interest for the relativistic supernovae modeling.
 It was shown earlier that  a development of the ion scale Bell-type instability in electron-ion relativistic shock may have a strong influence on the electron injection and acceleration.
-In the time period of about $1500 \omega_{pi}^{-1}$  ($\omega_{pi}$ is the ion plasma frequency) after the shock initialization the  magnetic field fluctuations  generated by Bell's instability may significantly suppresses the electrons spectrum even in a sub-luminous shock.
+In the time period of about $1500 \omega_{pi}^{-1}$  ($\omega_{pi}$ is the ion plasma frequency) after the shock initialization the  magnetic field fluctuations  generated by Bell's instability may significantly decreases number of accelerated electrons even in a sub-luminous shock.
 We study here the evolution of the electron spectra of subluminous shocks of different obliquity.
@@
 \cite{MD2001,Amato09,Bykov2013}).
-Bell's instability \cite{Bell04} can be the most significant due to it's fast growth rate.
+Bell's instability \cite{Bell04} can be the most significant due to it's large growth rate.
 Instabilities may modify the flow and has strong influence on the process of particle acceleration \cite{Crumley2019,2021MNRAS.501.4837L,2021MNRAS.502.5065L}.
@@
 The shock rippling effect is important for the electron spectra formation in the mildly-relativistic shocks.
-	In the simulations \cite{Crumley2019} the electron spectra evolution was demonstrated in a mildly-relativistic shock   of an inclination angle 10 degrees the magnetic fluctuations produced by the Weibel instability initially efficiently reflect and energize both ions and electrons and after the time period of about $1500 \omega_{pi}^{-1}$  the efficiency of electron acceleration by the Bell-mediated shock drops down significantly.
+	In the simulations \cite{Crumley2019} the electron spectra evolution was demonstrated in a mildly-relativistic shock   of an inclination angle 10 degrees.
+The magnetic fluctuations produced by the Weibel instability initially efficiently reflect and energize both ions and electrons.
+After the time period of about $1500 \omega_{pi}^{-1}$  the efficiency of electron acceleration by the shock drops down significantly because of developing of Bell's instability which increases the transversal magnetic field.
 The proton spectrum in this simulations was not strongly affected in the Bell-mediated shock.
-Given the very serious practical interest of non-thermal particle acceleration by the subluminal mildly-relativistic shocks we present below results of particle-in-cell modeling of electron and proton acceleration by the mildly-relativistic shocks of different inclination angles.
+Given the significant interest of non-thermal particle acceleration by the subluminal mildly-relativistic shocks we present below results of particle-in-cell modeling of electron and proton acceleration by the mildly-relativistic shocks of different inclination angles.
 
@@
 	Inclination angle of magnetic field due to shock velocity varies in different setups from $\theta = 20^\circ$ to $\theta = 40^\circ$.
-This values are chosen to contain critical angle, which is defined by the equation $c\cdot \cos(\theta_{crit})=v_{shock}$, where all values are measured in the upstream rest frame, and is equal approximately $30$ degrees.
+These values are chosen to contain critical angle, which is defined by the equation $c\cdot \cos(\theta_{crit})=v_{shock}$, where all values are measured in the upstream rest frame, and is equal approximately $30$ degrees.
 If inclination angle is larger then critical value, shock becomes superluminal - particle can not escape from it to upstream, even moving along the field lines with speed of light \cite{Pelletier2017,Sironi2011}.
-The maximum efficiency of particle acceleration is near the critical angle, less below it \cite{Romansky18}.
+The maximum efficiency of particle acceleration is near the critical angle, and is less below this angle \cite{Romansky18}.
 Magnetic field is lying in the simulation plane.
@@
 	Profiles of transversal magnetic field, obtained from numerical simulation, show significant field amplification in precursor of shock wave, see figure \ref{field}.
-It is result of development of instabilities in plasma, particularly Bell's instability \cite{Bell04}, which growth rate is larger then growth rate of other instabilities.
+It is result of development of instabilities in plasma, particularly Bell's instability \cite{Bell04}, which growth rate is larger than growth rate of other instabilities.
 
@@
 
-	It is known, that development of Bell's instability can suppress electrons acceleration \cite{Crumley2019,2021MNRAS.501.4837L,2021MNRAS.502.5065L}.
+	It is known, that development of Bell's instability can suppress electron acceleration \cite{Crumley2019,2021MNRAS.501.4837L,2021MNRAS.502.5065L}.
 We study this phenomena with different inclination angles of magnetic field.
 Spectrum of electrons is shown in figure \ref{spectrume}.
-One can see, that electrons acceleration has strong dependence on inclination angle, and when it is more then critical (at $\theta = 40^\circ$), non-thermal component almost disappear.
-For smaller angles electrons distribution function at high energies $(\Gamma > 300)$ depends on time, and it is maximal at times about 1400 inverse protons plasma frequencies for angles $\theta=20^\circ, \theta=30^\circ$, and 2100 and inverse protons plasma frequencies for $\theta = 33^\circ$.
-At the longer times electron spectrum decreases.
+One can see, that electron acceleration has strong dependence on inclination angle, and when it is more than critical value (at $\theta = 40^\circ$), non-thermal component almost disappears.
+For smaller angles electron distribution function at high energies $(\Gamma > 300)$ depends on time, and it maximizes at times about 1400 inverse proton plasma frequencies for angles $\theta=20^\circ, \theta=30^\circ$, and 2100 and inverse proton plasma frequencies for $\theta = 33^\circ$.
+After reaching the maximum value the electron distribution magnitude starts decreasing at high energies.
 It may by explained as consequence of magnetic field amplification.
@@
 As a result shock becomes quasi-perpendicular and can not effectively accelerate particles \cite{Sironi2011,Romansky18}.
-On the other hand, protons distribution function constantly increases in time for sub-critical angles, and almost does not has non-thermal component for $\theta = 40^\circ$, as shown in figure \ref{spectrump}.
+On the other hand, proton distribution function constantly increases in time for sub-critical angles, and almost does not have non-thermal component for $\theta = 40^\circ$, as shown in figure \ref{spectrump}.
 We assume, that oscillating transverse field has the strongest influence on the smallest scales, and suppresses the electrons injection into acceleration process, while protons still can be accelerated.
-Also, one can see, that temporal evolutions of electrons and protons spectrums are correlated.
-In case of inclination angle $\theta = 33^\circ$ protons distribution function grows slower then for smaller angles.
-And electrons distribution in this case is maximal at larger time.
-This correlation is the confirmation of assumption, that instability suppressing electron acceleration is caused by accelerated protons.
+Also, one can see, that temporal evolutions of electron and proton spectrums are correlated.
+In case of inclination angle $\theta = 33^\circ$ proton distribution function grows slower then for smaller angles.
+And electron distribution in this case is maximizes at larger time.
+This correlation confirms the assumption, that instability suppressing electron acceleration is caused by accelerated protons.
 
@@
 		\centering
-		\begin{minipage}{0.49\textwidth}
+		\begin{minipage}{0.48\textwidth}
 			\centering
@@
 		\end{minipage}\hfill
-		\begin{minipage}{0.49\textwidth}
+		\begin{minipage}{0.48\textwidth}
 			\centering
@@
 		\end{minipage}
-		\begin{minipage}{0.49\textwidth}
+		\begin{minipage}{0.48\textwidth}
 			\centering
@@
 		\end{minipage}
-		\caption{Temporal evolution of electrons spectrum in shocks with different magnetic field inclination angles.
+		\caption{Temporal evolution of electron spectrum in shocks with different magnetic field inclination angles.
 Top row: left $\theta = 20^\circ$, right $\theta = 30^\circ$, bottom row: left $\theta = 33^\circ$, right $\theta = 40^\circ$}
@@
 	\centering
-	\begin{minipage}{0.49\textwidth}
+	\begin{minipage}{0.48\textwidth}
 		\centering
@@
 	\end{minipage}\hfill
-	\begin{minipage}{0.49\textwidth}
+	\begin{minipage}{0.48\textwidth}
 		\centering
@@
 	\end{minipage}
-	\begin{minipage}{0.49\textwidth}
+	\begin{minipage}{0.48\textwidth}
 		\centering
@@
 	\end{minipage}
-	\caption{Temporal evolution of protons spectrum in shocks with different magnetic field inclination angles.
+	\caption{Temporal evolution of proton spectrum in shocks with different magnetic field inclination angles.
 Top row: left $\theta = 20^\circ$, right $\theta = 30^\circ$, bottom row: left $\theta = 33^\circ$, right $\theta = 40^\circ$}
@@
 
-	Details of electrons acceleration are important for correct interpretations of observations synchrothron radiation from astrophysical sources.
-Particle-in-cell simulation has been used to demonstrate magnetic instabilities influence on electrons acceleration in mildly-relativistic electron-ion shock.
-It is shown, that amplification of magnetic field suppresses formation of electrons spectrum at high energies while protons spectrum can grow monotonously in time in subluminal shock.
-Also, we demonstrated that grows of proton spectrum is correlated with decreasing of electrons acceleration, and resulting  electrons spectrum strongly depends on magnetic field inclination angle.
+	Details of electron acceleration are important for correct interpretations of observations of the  synchrothron radiation from astrophysical sources.
+Particle-in-cell simulation has been used to demonstrate magnetic instabilities influence on electron acceleration in mildly-relativistic electron-ion shock.
+It is shown, that amplification of magnetic field suppresses formation of electron spectrum at high energies while proton spectrum can grow monotonously in time in subluminal shock.
+Also, we demonstrated that groth of proton spectrum is correlated with decreasing of electron acceleration, and resulting  electron spectrum strongly depends on magnetic field inclination angle.
 This effect should be taken into account for deriving parameters of source of synchrothron radiation via analysis of spectrum.
-Further particle-in-cell simulations with a wider spatial and energetical ranges will allow to study the influence of instabilities on electrons spectrum more precisely.
+Further particle-in-cell simulations with a wider spatial and energetical ranges will allow to study the influence of instabilities on electron spectrum more precisely.
 

### 2021-07-06 papers/Physica2021/Romansky_Physica2021.tex
turbulence 1

@@
 		\centering
-		\includegraphics[width=0.50\textwidth]{fig/Bnorm.png} 
+		\includegraphics[width=0.48\textwidth]{fig/Bnorm.png} 
 		\caption{Temporal evolution of the transversal magnetic field.}
@@
 One can see, that electron acceleration has strong dependence on inclination angle, and when it is more than critical value (at $\theta = 40^\circ$), non-thermal component almost disappears.
-For smaller angles electron distribution function at high energies $(\Gamma > 300)$ depends on time, and it maximizes at times about 1400 inverse proton plasma frequencies for angles $\theta=20^\circ, \theta=30^\circ$, and 2100 and inverse proton plasma frequencies for $\theta = 33^\circ$.
+For smaller angles electron distribution function at high energies $(\Gamma > 300)$ depends on time, and it maximizes at times about $1400$ inverse proton plasma frequencies $\omega_{pi}$ for angles $\theta=20^\circ, \theta=30^\circ$, and $2100$ and inverse proton plasma frequencies for $\theta = 33^\circ$.
 After reaching the maximum value the electron distribution magnitude starts decreasing at high energies.
@@
 Also, one can see, that temporal evolutions of electron and proton spectrums are correlated.
-In case of inclination angle $\theta = 33^\circ$ proton distribution function grows slower then for smaller angles.
-And electron distribution in this case is maximizes at larger time.
+In case of inclination angle $\theta = 33^\circ$ proton distribution function grows slower than for smaller angles, at time less than $2000 {\omega_{pi}}^{-1}$  it has very weak non-thermal component.
+And electron distribution in this case maximizes at larger time than for smaller angles.
 This correlation confirms the assumption, that instability suppressing electron acceleration is caused by accelerated protons.
@@
 		\centering
-		\begin{minipage}{0.48\textwidth}
+		\begin{minipage}{0.47\textwidth}
 			\centering
@@
 		\end{minipage}\hfill
-		\begin{minipage}{0.48\textwidth}
+		\begin{minipage}{0.47\textwidth}
 			\centering
@@
 		\end{minipage}
-		\begin{minipage}{0.48\textwidth}
+		\begin{minipage}{0.47\textwidth}
 			\centering
@@
 	\centering
-	\begin{minipage}{0.48\textwidth}
+	\begin{minipage}{0.47\textwidth}
 		\centering
@@
 	\end{minipage}\hfill
-	\begin{minipage}{0.48\textwidth}
+	\begin{minipage}{0.47\textwidth}
 		\centering
@@
 	\end{minipage}
-	\begin{minipage}{0.48\textwidth}
+	\begin{minipage}{0.47\textwidth}
 		\centering
@@
 It is shown, that amplification of magnetic field suppresses formation of electron spectrum at high energies while proton spectrum can grow monotonously in time in subluminal shock.
-Also, we demonstrated that groth of proton spectrum is correlated with decreasing of electron acceleration, and resulting  electron spectrum strongly depends on magnetic field inclination angle.
+Also, we demonstrated that growth of proton spectrum is correlated with decreasing of electron acceleration, and resulting  electron spectrum strongly depends on magnetic field inclination angle.
 This effect should be taken into account for deriving parameters of source of synchrothron radiation via analysis of spectrum.

### 2021-07-06 ThinSphericalTurbulence/constants.h

@@ -153,9 +153,9 @@ const double aprmaxx = 6.50;
 const double aprmaxy = 19.3;
 
 const double minB = 0.0001;
-const double maxB = 0.2;
+const double maxB = 0.4;
 const double minN = 0.01;
-const double maxN = 200;
+const double maxN = 400;
 const double minFraction = 0.001;
 const double maxFraction = 0.2;
 const double maxSigma = 10.0;

### 2022-01-12 disser/introduction.tex
disser

\chapter*{Введение}					
\addcontentsline{toc}{chapter}{Введение}	%добавляем в оглавление

\section*{Быстрые голубые оптические транзиенты}
\addcontentsline{toc}{section}{Космические гамма-всплески}	%добавляем в оглавление

\section*{Актуальность темы диссертации}
\addcontentsline{toc}{section}{Актуальность темы диссертации}	%добавляем в оглавление

\section*{Цели работы}
\addcontentsline{toc}{section}{Цели работы}	%добавляем в оглавление

Для достижения поставленной цели решаются следующие задачи:
\begin{enumerate}
\item 
\end{enumerate}

\section*{Научная новизна}
\addcontentsline{toc}{section}{Научная новизна}	%добавляем в оглавление

Следующие основные результаты получены впервые:
\begin{enumerate}
\item 
\end{enumerate}

\section*{Научная и практическая значимость}
\addcontentsline{toc}{section}{Научная и практическая значимость}	%добавляем в оглавление

\begin{enumerate}
\item 
\end{enumerate}

\section*{Основные положения, выносимые на защиту}
\addcontentsline{toc}{section}{Основные положения, выносимые на защиту}

\begin{enumerate}
\item 
\end{enumerate}

\section*{Апробация работы и публикации}
Результаты, вошедшие в диссертацию, получены в период с 2014 по 2022~годы и 
опубликованы в 4~статьях в реферируемых журналах и в тезисах 5 конференций.

Статьи в рецензируемых изданиях:
\begin{enumerate}
\item %V.\,D.\,Pal'shin, K.\,Hurley, D.\,S.\,Svinkin et al. Interplanetary Network Localizations of
%Konus Short Gamma-Ray Bursts // Astrophys.~J.~Suppl. 2013. Vol.~207. id~38;

\end{enumerate}

Результаты докладывались на всероссийских и международных конференциях: 
\begin{enumerate}
\item 
\end{enumerate}
и на семинарах сектора теоретической астрофизики ФТИ~им.~А.~Ф.~Иоффе и ГАИШ МГУ.

\section*{Личный вклад}
\addcontentsline{toc}{section}{Личный вклад}

\section*{Структура диссертации}
\addcontentsline{toc}{section}{Структура диссертации}
Диссертация состоит из введения, 5 глав, заключения и библиографии.
Общий объем диссертации 155 страниц, включая 33 рисунка, 13 таблиц. 
Библиография включает 206 наименований на 18 страницах.

Во \textbf{введении} приведено 
Также сформулированы основные результаты и положения, выносимые на защиту, и приведен
список работ, в которых опубликованы основные результаты диссертации.

\textbf{Глава~\ref{FBOT}} посвящена

\enlargethispage{4\baselineskip}
\textbf{Заключение} содержит краткий обзор полученных в диссертации результатов.

\clearpage

### 2022-01-12 disser/part1.tex
disser

\chapter{Моделирование излучения FBOT}\label{FBOT}

Последние наблюдения открыли новый класс объектов - быстрые голубые оптические транзиенты (FBOT)\cite{Drout2014, Margutti2014, Coppejans2020, Ho2020}. Вместе со слабыми гамма-всплесками они возможно принадлежат к переходному классу объектов между нерелятивистскими сверхновыми и обычными гамма-всплесками. Как показано в работе Шевалье и Ирвина \cite{Chevalier2011} наблюдаемые характеристики вспышки определяются длительностью действия центрального источника энергии выброса и его взаимодействием со внешними слоями взрывающейся звезды. При различных параметрах вспышка будет относиться к разным классам объектов. Известные на данный момент быстрые голубые оптические транзиенты AT2018cow \cite{Margutti2014}, CSS161010 \cite{Coppejans2020}, ZTF2018abvkwla \cite{Ho2020} and AT2020xnd \cite{Ho2021, Bright2021} характеризуются большой светимостью $L > 10^43 эрг/с$, коротким характерным временным масштабом порядка нескольких дней, низкой выброшенной массой эжекты и высокими скоростями.
Рассматриваемы в данной главе объект CSS161010, расположенный в карликовой галактике на расстоянии 150 мегапарсек имеет по оценкам следующие\cite{Coppejans2020} характеристики: скорость эжекты 0.55 c на ранних этапах и около 0.3 c на поздних (порядка одного года), выброшенная масса порядка 0.01-0.1 солнечных масс. 
%почему четвертый был отдельно в статье?

В данной главе будет проведено Particle-in-Cell моедлирование ускорения частиц в транс-релятивистской ударной волне со скоростями, характерными для FBOT объектов. По имеющимся распределениям будет расчитано синхротронное излучение и с помощью фитирования наблюдательных данных будут подобраны такие параметры как магнитное поле, концентрация вещества и геометрические характеристики в ударной волне.

\section{1.1}

\section{1.2}

\subsection{1.2.1}\label{sec:Bg_rate}

\subsection{1.2.2}

\FloatBarrier
\section{Заключение}

Получены следующие результаты:
\begin{enumerate}
\item 
\item 
\end{enumerate}

\clearpage

### 2022-01-12 SimpleRadiation/main.cpp

@@ -959,7 +959,8 @@ int main()
 				Ee[j][i] = gamma*massElectron*speed_of_light2;
 				//maxEnergy = Ee[i];
 				Fe[j][i] = Fe[j][i] / (massElectron*speed_of_light2);
-				if(i > 137){
+				//if(i > 137){
+				if((i > 137)&&(u < 500)){
 					Fe[j][i] = Fe[j][137]*pow(Ee[j][i]/Ee[j][137], -3.5);
 				}
 				dFe[j][i] = (Fe[j][i] / (4*pi)) * (Ee[j][i] - Ee[j][i - 1]);

### 2022-01-13 disser/part1.tex
disser

@@
 Рассматриваемы в данной главе объект CSS161010, расположенный в карликовой галактике на расстоянии 150 мегапарсек имеет по оценкам следующие\cite{Coppejans2020} характеристики: скорость эжекты 0.55 c на ранних этапах и около 0.3 c на поздних (порядка одного года), выброшенная масса порядка 0.01-0.1 солнечных масс.
+Важнейшей особенностью этих объектов является вид их спектральной плотности излучения, имеющей выраженный максимум и степенные участки, говорящие о наличии синхротронного самопоглощения.
 %почему четвертый был отдельно в статье?
 
-В данной главе будет проведено Particle-in-Cell моедлирование ускорения частиц в транс-релятивистской ударной волне со скоростями, характерными для FBOT объектов.
+В данной главе будет проведено Particle-in-Cell моделирование ускорения частиц в транс-релятивистской ударной волне со скоростями, характерными для FBOT объектов.
 По имеющимся распределениям будет расчитано синхротронное излучение и с помощью фитирования наблюдательных данных будут подобраны такие параметры как магнитное поле, концентрация вещества и геометрические характеристики в ударной волне.
 
-\section{1.1}
+\section{Излучение объектов с синхротронным самопоглощением}
+Процесс синхротронного излучения хороши известен и описан в классических работах.
+Но с точки зрения квантовой электродинамки, любому процессу излучения можно так же сопоставить процесс поглощения.
+Сечение процесса синхротронного самопоглощения описано в работе Гизеллини и Свенсона \cite{Ghisellini1991}.
+Спектральная плотность мощности излучения единицы объема вещества определеяется формулой
+\begin{equation} \label{emission}
+I(\nu)=\int_{E_{min}}^{E_{max}} dE \frac {\sqrt {3}{e}^{3}n F(E) B \sin ( \phi)}{{m_e}{c}^{2}}
+\frac{\nu}{\nu_c}\int_{\frac {\nu}{\nu_c}}^{\infty }\it K_{5/3}(x)dx,
+\end{equation}
+где $\phi$ это угол межде вектором магнитного поля и лучом зрения, $\displaystyle\nu_{c}$ критическая частота, определяемая выражением $\displaystyle\nu_{c} = 3 e^{2} B \sin(\phi) E^{2}/4\pi {m_{e}}^{3} c^{5}$, и~$K_{5/3}$ - функция МакДональда.
+Коэффициент поглощения для фотонов, распростроняющихся вдоль луча зрения равен
+\begin{equation}\label{absorption}
+k(\nu)=\int_{E_{min}}^{E_{max}}dE\frac {\sqrt {3}{e}^{3}}{8\pi m_e \nu^2}\frac{n B\sin(\phi)}{E^2}
+\frac{d}{dE} E^2 F(E)\frac {\nu}{ \nu_c}\int_{\frac {\nu}{ \nu_c}}^{\infty }K_{5/3}(x) dx.
+\end{equation}
+Используя эти формулы Шевалье \cite{Chevalier1998} построил модель излучения плоского однородного диска, расположенного перпендикулярно к лучу зрения, с радиусом $R$, толщиной $s$, однородным магнитным полем $B$ и степенной функцией распределения электронов $N(E) = N_0 E^{-p}$.
+Вместо толщины обычно используется доля излучающего объема $f$, определенная так, что $\pi R^2 s = 4 \pi /3 R^3 f$.
+При таких предположениях \textcolor{red}{спектральная плотность излучения определяется следующими формулами}
+\begin{equation}
+I_{\nu}=S(\nu_1)J(\frac{\nu}{\nu_1},p)
+sin(\theta)}^{\frac{p+2}{p+4}},
+\end{equation}
+where
+\begin{equation}
+\nu_1 = 2 c_1 {(s c_6 N_0)}^{\frac{2}{p+4}}{B sin(\theta)}^{\frac{p+2}{p+4}}
+\end{equation}
+\begin{equation}
+S(\nu_1)=\frac{c_5}{c_6}{(B sin(\theta))}^{-1/2}{(\frac{\nu_1}{2 c_1})}^{5/2}
+\end{equation}
+\begin{equation}
+J(z,p)=z^{5/2}(1-exp(-z^{(p+4)/2}))
+\end{equation}
+\section{Particle-in-Cell моделирование }
 
-\section{1.2}
+\section{Расчёт излучения}
 
-\subsection{1.2.1}\label{sec:Bg_rate}
-
-\subsection{1.2.2}
+\section{Результаты}
 

### 2022-01-27 disser/part1.tex
bug with doppler

@@
 I_{\nu}=S(\nu_1)J(\frac{\nu}{\nu_1},p)
-sin(\theta)}^{\frac{p+2}{p+4}},
+{sin(\theta)}^{\frac{p+2}{p+4}},
 \end{equation}
-where
+где
 \begin{equation}
@@
 \end{equation}
+где $\theta$ - угол между магнитным полем и лучом зрения, $c_1, c_5, c_6$ - константы Пахольчика \cite{Pacholczyk}, зависящие от спектрального индекса электронов.
+
+Измерив спектральную плотность излучения и получив максимум в данный момент времени $F_{max}$, приходящийся на частоту $\nu_{max}$, модно получить размер и магнитное поле в источнике по формулам
+\begin{equation}\label{ChevR}
+R = {\left( \frac {6 \epsilon_B {c_6}^{p+5}{F_{max}}^{p+6}{
+			D}^{2p+12}}{\epsilon_e f \left( p-2 \right) {\pi}^{p+5}{{c_5}}^{p+6}
+		{E_1}^{p-2}} \right)} ^{ \frac{1}{p+13} } \frac{2 c_1}{\nu_{max}},
+\end{equation}
+\begin{equation}\label{ChevB}
+B = { \left(\frac {{\epsilon_B}^2 36 {\pi}^{3}{c_5}}{{
+			\epsilon_e}^{2}{f}^{2} \left( p-2 \right) ^{2}{{c_6}}^{3}{{E_1}}^{2 p-4}F_{max}{D}^{2}} \right)}
+^{\frac{2}{2p+13}}\frac{\nu_{max}}{2 c_1},
+\end{equation}
+где $D$ - расстояние до источника, $\epsilon_B, \epsilon_e$ - доли энергии в источнике, содержащиеся в магнитном поле и в электронах.
 \section{Particle-in-Cell моделирование }

### 2022-01-27 Radiation/constants.h

@@ -38,7 +38,7 @@ const std::string fileNameF = "Fs";
 const std::string logFileName = "log.dat";
 const std::string BFileName = "B.dat";
 
-const int Niterations = 20;
+const int Niterations = 10;
 const int Nmonth = 4;
 const int Nmontecarlo = 100000;
 

### 2022-01-27 Radiation/main.cpp

@@ -616,7 +616,8 @@ int main()
 				By3d[i][j][k] = 0.0;
 				Bz3d[i][j][k] = 0.0;
 				if(geometry == SPHERICAL) {
-					concentrations3d[i][j][k] = 1.0*sqr(rmax/r);
+					double tempBeta = 0.75;
+					concentrations3d[i][j][k] = 1.0*pow(1 - tempBeta, 7.0/4.0)/pow(1 - tempBeta*r/rmax, 7.0/4.0);
 				} else {
 					concentrations3d[i][j][k] = 1.0;
 				}
@@ -1176,7 +1177,7 @@ int main()
 	v = 0.67*speed_of_light;
 	sigma = 0.02;
 	//concentration = sqr(Bfactor)/(sigma*4*pi*massProtonReal*speed_of_light2);
-	bool optPar[8] = {true, true, true, true, true, true, true, true};
+	bool optPar[8] = {true, true, true, true, true, false, false, true};
 	double vector[8];
 	vector[0] = Bfactor/maxB;
 	vector[1] = concentration/maxN;
@@ -1198,15 +1199,15 @@ int main()
 	fprintf(logFile, "error = %lf\n", error);
 	const int Nbp = 8;
 	const int Nnp = 10;
-	const int Nfp = 4;
-	const int Nvp = 6;
+	const int Nfp = 6;
+	const int Nvp = 7;
 	const int Nr0 = 4;
 	const int Na = 3;
 	const int Nb1 = 3;
 	double Bpoints[Nbp] = {0.1, 0.2, 0.5, 1, 2, 5, 10, 20};
 	double npoints[Nnp] = {0.5, 1, 2, 5, 10, 20, 50, 100, 200, 500};
-	double fpoints[Nfp] = {0.1,0.3,0.4,0.5};
-	double vpoints[Nvp] = {0.5*speed_of_light, 0.6*speed_of_light, 0.65*speed_of_light, 0.7*speed_of_light, 0.75*speed_of_light, 0.8*speed_of_light};
+	double fpoints[Nfp] = {0.03, 0.06, 0.1,0.3,0.4,0.5};
+	double vpoints[Nvp] = {0.4*speed_of_light, 0.5*speed_of_light, 0.6*speed_of_light, 0.65*speed_of_light, 0.7*speed_of_light, 0.75*speed_of_light, 0.8*speed_of_light};
 	double rpoints[Nr0] = {0.1*maxR0, 0.2*maxR0, 0.4*maxR0, 0.5*maxR0};
[…]

### 2022-01-27 Radiation/spectrum.h

@@ -21,6 +21,7 @@ void evaluateImageSpherical(double*** image, double* nu, double**** nudoppler, d
 void evaluateImageFlat(double*** image, double* nu, double**** nudoppler, double**** Inu, double**** Anu, double rmax, int Nnu, double rfactor, double fractionLength, const double& beta);
 
 void evaluateNuDoppler(double***** NuDoppler, int Nmonth, int* Nnumonth, double** Numonth, const double& beta);
+void evaluateNuDoppler(double**** NuDoppler, int Nnu, double* nu, const double& beta);
 
 //for simple flat disk
 void evaluateSpectrumFlatSimple(double* nu, double* totalInu, double* Inu, double* Anu, int Nnu, double rmax, double fraction);

### 2022-01-27 Radiation/startparameters.h

@@ -8,13 +8,13 @@ const bool initialGridSearch = false;
 const bool optimization = true; //false for debug only
 
 enum Doppler {NO, INTEGER, DIFFERENTIAL};
-const Doppler doppler = Doppler::INTEGER;
+const Doppler doppler = Doppler::DIFFERENTIAL;
 
 enum Geometry {FLAT_SIMPLE, FLAT, SPHERICAL};
 const Geometry geometry = Geometry::SPHERICAL;
 
 enum Input {TRISTAN, SMILEI, MAXWELL, POWERLAW, COMBINED};
-const Input input = Input::SMILEI;
+const Input input = Input::COMBINED;
 
 enum Scale {LINEAR, LOG};
 const Scale scale = Scale::LINEAR;

### 2022-01-28 disser/part1.tex
fz

@@
 \end{equation}
-где $D$ - расстояние до источника, $\epsilon_B, \epsilon_e$ - доли энергии в источнике, содержащиеся в магнитном поле и в электронах.
+где $D$ - расстояние до источника, $\epsilon_B, \epsilon_e$ - доли энергии в источнике, содержащиеся в магнитном поле и в электронах, $E1$ - энергия, с которой начинается степенное распределение электронов.
 \section{Particle-in-Cell моделирование }

### 2022-01-28 Radiation/main.cpp

@@ -1267,14 +1267,14 @@ int main()
 		fscanf(initialFile, "%lf %lf %lf %lf %lf %lf %lf %lf", &Bfactor, &concentration, &fractionSize, &v, &r0, &a, &b, &fpower);
 		v = v*speed_of_light;
 		fclose(initialFile);
-		/*Bfactor = 0.2;
-		concentration = 10;
-		fractionSize = 0.4;
-		r0 = 1.1664e+16;
-		v = 0.5*speed_of_light;
+		Bfactor = 0.362992;
+		concentration = 501.865;
+		fractionSize = 0.112738;
+		r0 = 5.10995e+15;
+		v = 0.364802*speed_of_light;
 		a = 2.0;
-		b = 0.0;
-		fpower = 1.0;*/
+		b = 2.0;
+		fpower = 2.56063;
 	}
 
 	vector[0] = Bfactor/maxB;
@@ -1363,6 +1363,7 @@ int main()
 	rmin = r0 + v*timeMoments[0];
 	evaluateNuDoppler(tempNuDoppler, Nnu, nu, 0.75*v/speed_of_light); 
 	for(int l = 0; l < Nmonth; ++l){
+		printf("evaluate final spectrum month = %d\n", l);
 		fprintf(logFile, "evaluate final spectrum month = %d\n", l);
 		fflush(logFile);
 		double r = r0 + v*timeMoments[l];
@@ -1377,7 +1378,7 @@ int main()
 		//evaluateSpectrum(nu, tempTotalInu[l], Inu, Anu, area, length, Nnu, r, Rho, Phi);
 
 		evaluateAllEmissivityAndAbsorption(tempNuDoppler, tempInu, tempAnu, Nnu, weightedEe, weightedFe, Np, Ndist, B3d, sintheta3d, thetaIndex3d, concentrations3d, concentration, Bfactor, rfactor, a, b);
-		double tempFraction = fractionSize/pow(rfactor, fpower);
+		double tempFraction = fractionSize/pow(rfactor, fpower - 1.0);
 		double dopplerBeta = 0.75*v/speed_of_light;
 		if(l == 0) {
 			if(geometry == SPHERICAL){

### 2022-01-28 Radiation/spectrum.cpp

@@ -381,9 +381,9 @@ void evaluateSpectrumSpherical(double* nu, double**** nuDoppler, double* I, doub
 
 	double drho = tempRmax/Nrho;
 
-	double length[Nz];
-
+#pragma omp parallel for shared(nu, nuDoppler, I, Inu, Anu, rmax, Nnu, fractionLength, beta, tempNr, tempRmax, tempRmin, tempdr, dphi, dz, gamma)
 	for(int i = 0; i < tempNr; ++i){
+		double length[Nz];
 		double r = (i + 0.5)*tempdr;
 		double s = 0.5*dphi*(2*i + 1)*tempdr*tempdr;
 		double z1 = -sqrt(tempRmax*tempRmax -r*r);
@@ -522,9 +522,9 @@ void evaluateSpectrumSphericalAtNu(double nu, double*** nudoppler, double& I, do
 
 	double drho = tempRmax/Nrho;
 
-	double length[Nz];
-
+#pragma omp parallel for shared(nu, nudoppler, I, Inu, Anu, rmax,rfactor, fractionLength, d, beta, tempNr, tempRmax, tempRmin, tempdr, dphi, dz, gamma)
 	for(int i = 0; i < tempNr; ++i){
+		double length[Nz];
 		double r = (i + 0.5)*tempdr;
 		double s = 0.5*dphi*(2*i + 1)*tempdr*tempdr;
 		double z1 = -sqrt(tempRmax*tempRmax -r*r);
@@ -676,6 +676,7 @@ void evaluateSpectrumFlat(double* nu, double**** nudoppler, double* I, double***
 		}
 	}
 
+#pragma omp parallel for shared(nu, nudoppler, I, Inu, Anu, rmax, Nnu, fractionLength, beta, tempRmax, tempRmin, tempdr, dphi, dz, zmin, gamma, drho, length)
 	for(int i = 0; i < Nrho; ++i){
 		double s = 0.5*dphi*(2*i + 1)*tempdr*tempdr;
 		for(int j = 0; j < Nphi; ++j){
@@ -811,6 +812,7 @@ void evaluateSpectrumFlatAtNu(double nu, double*** nudoppler, double& I, double*
 		}
 	}
 
+#pragma omp parallel for shared(nu, nudoppler, I, Inu, Anu, rmax, rfactor, fractionLength, d, beta, tempRmax, tempRmin, tempdr, dphi, dz, zmin, gamma, drho, length)
 	for(int i = 0; i < Nrho; ++i){
[…]

### 2022-01-28 Radiation/startparameters.h

@@ -4,7 +4,7 @@
 const bool parker = false;
 const bool turbulence = false;
 
-const bool initialGridSearch = false;
+const bool initialGridSearch = true;
 const bool optimization = true; //false for debug only
 
 enum Doppler {NO, INTEGER, DIFFERENTIAL};

### 2022-02-22 disser/part1.tex
disser

@@
 При различных параметрах вспышка будет относиться к разным классам объектов.
-Известные на данный момент быстрые голубые оптические транзиенты AT2018cow \cite{Margutti2014}, CSS161010 \cite{Coppejans2020}, ZTF2018abvkwla \cite{Ho2020} and AT2020xnd \cite{Ho2021, Bright2021} характеризуются большой светимостью $L > 10^43 эрг/с$, коротким характерным временным масштабом порядка нескольких дней, низкой выброшенной массой эжекты и высокими скоростями.
+Известные на данный момент быстрые голубые оптические транзиенты AT2018cow \cite{Margutti2014}, CSS161010 \cite{Coppejans2020}, ZTF2018abvkwla \cite{Ho2020} и AT2020xnd \cite{Ho2021, Bright2021} характеризуются большой светимостью $L > 10^43 эрг/с$, коротким характерным временным масштабом порядка нескольких дней, низкой выброшенной массой эжекты и высокими скоростями.
 Рассматриваемы в данной главе объект CSS161010, расположенный в карликовой галактике на расстоянии 150 мегапарсек имеет по оценкам следующие\cite{Coppejans2020} характеристики: скорость эжекты 0.55 c на ранних этапах и около 0.3 c на поздних (порядка одного года), выброшенная масса порядка 0.01-0.1 солнечных масс.
-Важнейшей особенностью этих объектов является вид их спектральной плотности излучения, имеющей выраженный максимум и степенные участки, говорящие о наличии синхротронного самопоглощения.
+]textcolor{red}{Важнейшей особенностью этих объектов является вид их спектральной плотности излучения, имеющей выраженный максимум и степенные участки, говорящие о наличии синхротронного самопоглощения.}
 %почему четвертый был отдельно в статье?
@@
 
-Измерив спектральную плотность излучения и получив максимум в данный момент времени $F_{max}$, приходящийся на частоту $\nu_{max}$, модно получить размер и магнитное поле в источнике по формулам
+Измерив спектральную плотность излучения и получив максимум в данный момент времени $F_{max}$, приходящийся на частоту $\nu_{max}$, можно получить размер и магнитное поле в источнике по формулам
 \begin{equation}\label{ChevR}
@@
 где $D$ - расстояние до источника, $\epsilon_B, \epsilon_e$ - доли энергии в источнике, содержащиеся в магнитном поле и в электронах, $E1$ - энергия, с которой начинается степенное распределение электронов.
+Таким образом в работах  были оценены параметры источников AT2018cow \cite{Margutti2014}, CSS161010 \cite{Coppejans2020}, ZTF2018abvkwla \cite{Ho2020} и AT2020xnd \cite{Ho2021, Bright2021}.
+Так же, измеряя радиус источника $R$ в разные моменты времени можно оценить скорость распространения ударной волны.
 \section{Particle-in-Cell моделирование }
+Для более точного расчета синхротронного излучения оптических транзиентов необходимо знание функции распределения электронов.
+В перечисленных выше работах используется модель степенного распределения с произвольным параметром - долей энергии в электронах $\epsilon_e$, но как показано в работе \cite{Margalit2021}, тепловые электроны так же могут вносить значительный вклад в излучение.
+Мы для определения функции распределения электронов в субрелятивистской ударной волне использовали Particle-in-Cell расчеты, выполненные с помощью кода Smilei \cite{Smilei18}, разработанным Деруле и другими.
+Этот код основан на явной конечно-разностной схеме, с релятивистскими уравнениями движения и с алгоритмом сохранения заряда, предложенным Езиркеповым \cite{Esirkepov}.
+
+Для инициализации ударной волны мы используем распространенный подход - в двумерной симуляции на левой границе по оси $x$ установлена отражающая стенка, с которой сталкивается однородный поток плазмы.
+На правой границе использованы граничные условия сохраняющие поток.
+Вдоль перпендикулярного направления $y$ установлены периодические граничные условия.
+Все вектора, кроме координат частиц - поля и скорости, являются трехмерными, так называемая 2d3v схема.
+Число ячеек в продольном направлении
+$\displaystyle Nx = 204,800$, в поперечном $\displaystyle Ny = 400$.
+Размер ячейки $dx = dy = 0.2c/{\omega_e}$ , где $\displaystyle\omega_e$ - плазменная электронная частота $\displaystyle\omega_e = \sqrt{4 \pi n e^2/ m_e}$, $\displaystyle e$ - абсолютная величина заряда электрона, $\displaystyle n$ - концентрация (в коде SMILEI может быть выбрана произвольно, другие величины легко масштабируются), $\displaystyle m_e$ - масса электрона, увеличена для экономии численных ресурсов так, что отношение масс протона и электрона $m_p/m_e = 100$.
+Начальное магнитное поле $\displaystyle\vec{B}$ лежит в плоскости симуляции под углом $\displaystyle\theta$ к скорости потока.
+Электрическое поле инициализируется так, чтобы компенсировать силу Лоренца в лабораторной системе $\displaystyle\vec{E} = -\vec{v} \times
+\vec{B}/c$. Скорость потока плазмы $\displaystyle v$  is $0.3 c$ 
+что соответствует лоренц-фактору $\displaystyle\gamma = 1.05$.
+Такая скорость соответствует оценкам для выбранного нами транзиента CSS161010 из статьи \cite{Coppejans2020}.
+Замагниченность потока $\displaystyle\sigma = B^2/4 \pi n \gamma m_p c^2 = 10^{-4}$.
+С такими параметрами размер области симуляции вдоль оси $\displaystyle x$ соответствует $\displaystyle 500$
+гирорадиусам протонов в потоке плазмы $\displaystyle r_g = m_p v c \gamma/e B$, а вдоль оси $\displaystyle y$ одному гирорадиусу.
+Максимальное время симуляции составляет $\displaystyle 7\times 10^4~\omega_{e}^{-1}$ или около
+$\displaystyle 100$ обратных протонных гирочастот. Для концентраций порядка $\displaystyle n\approx
+1$~cm$^{-3}$, это время соответствует временам порядка одной секунды, что намного меньше характерного времени активного излучения быстрых оптических транзиентов, а значит такое квазистационарное моделирование применимо к рассматриваемому объекту в каждый отдельный момент времени.
+
+Эффективность ускорения частиц зависит от угла наклона магнитного поля $\displaystyle\theta$,
+особенно в случае релятивистских ударных волн \cite{SironiSpitkovsky2009pair, GuoSironi2014_1, Crumley2019, Romansky2018}.
+Чтобы быть вовлеченной в процесс ускорения Ферми первого рода, частице необходимо удалиться от фронта, и если
+она движется вдоль силовых линий магнитного поля, максимальная возможная скорость вдоль оси $\displaystyle x$ равна $\displaystyle c\cos(\theta')$, и для ухода от ударной волны она должна быть больше скорости самой ударной волны $\displaystyle v_{sh}'$ (все величины измерены в системе набегающего потока).
+Поэтому для квазипоперечных ударных волн, у которых $c \cos(\theta')$ меньше скорости ударной волны $\displaystyle v_{sh}'$ ускорение невозможно.
+При углах наклона меньших критического, ударная волна способна эффективно ускорять частицы, формировать степенной хвост их функции распределения, как показано на рисунке~\ref{compareelectrons}.
+\begin{figure}
+	\centering
+	\includegraphics[width=10.5 cm]{compareelectrons2.eps} 
+	\caption{Функция распределения электронов за фронтом ударной волны при следующих параметрах: скорость потока \mbox{$v = 0.3~c,$} , магнетизация $\sigma = 0.0002$ при нескольких углах наклона.}
+	\label{compareelectrons}
+\end{figure}
+
+Для дальнейшего моделирования излучения использовалась функция распределения, полученная для угла $\theta = 30^\circ$ градусов (квазипараллельная ударная волна).
+
+Из результатов моделирования можно извлечь параметры полученного распределения электронов.
+За фронтом ударной волны функция распределения электронов имеет сложный вид, но можно выделить две основных компоненты - тепловой пик и степенной хвост на высоких энергиях.
+На низких энергиях ($\displaystyle E < 5 m_e c^2$) мы аппроксимируем фунцию распределения с помощью распределения Максвелла-Юттнера, итеративным процессом минимизируя функционал $\displaystyle f(T) = \int_{m_e c^{2}}^{5 m_e c^{2}} (F(E) - F_{mj}(E,T))^{2}dE$, где $\displaystyle F(E)$ функция распределения полученная из моделирования, $\displaystyle F_{mj}(E,T)$ - распределение Максвелла-Юттнера.
+Таким образом определяется температура, соответствующая тепловому пику.
+Для высоких энергий ($\displaystyle 20~m_e c^{2} < E < 50~m_e c^{2}$) мы используем линейную регрессию в дважды логарифмических координатах для определения спектрального индекса $p$.
+Функция распределения электронов за фронтом ударной волны (от 2000 до 5000 ячеек за фронтом) в момент времени $\displaystyle t$ = 70,000~$\omega_{e}^{-1}$ при начальных параметрах $\displaystyle\sigma = 0.0002$, $\theta = 30^\circ$ и ее аппроксимации с \textcolor{red}{температурой $\displaystyle T_e = 5\times 10^{10}$~K} и спектральным индексом $\displaystyle p = 3.6$ изображены на рисунке ~\ref{electrons}.
+\begin{figure}
+		\centering
+		\includegraphics[width=10.5 cm]{electrons.eps} 
+		\caption{Функция распределения электронов за фронтом ударной волны при начальных параметрах $v = 0.3~c,$ $\sigma = 2\times10^{-4}$ и $\theta = 30^{\circ}$.}
+		\label{electrons}
+	\end{figure}
 

### 2022-02-22 Radiation/constants.h

@@ -55,9 +55,9 @@ const double distance = 40*3.08*1.0E24;
 //CSS161010
 //const double distance = 150*3.08*1.0E24;
 
-const int Nrho = 10;
+const int Nrho = 20;
 const int Nphi = 10;
-const int Nz = 10;
+const int Nz = 20;
 
 const int Nk = 10;
 
@@ -176,7 +176,7 @@ const double minBbound = 0;
 const double maxB = 50;
 const double minN = 0.01;
 const double minNbound = 0;
-const double maxN = 1000;
+const double maxN = 5000;
 const double minFraction = 0.001;
 const double maxFraction = 0.9;
 const double minFbound = 0;

### 2022-02-22 Radiation/main.cpp

@@ -1197,16 +1197,37 @@ int main()
 	const int Nfp = 6;
 	const int Nvp = 8;
 	const int Nr0 = 4;
-	const int Na = 3;
-	const int Nb1 = 3;
+	int Na = 3;
+	double* apoints;
+	if (parker) {
+		Na = 1;
+		apoints = new double[Na];
+		apoints[0] = 2.0;
+	} else {
+		Na = 3;
+		apoints = new double[Na];
+		apoints[0] = 1.0;
+		apoints[1] = 2.0;
+		apoints[2] = 3.0;
+	}
+	int Nb1 = 3;
+	double* bpoints;
+	if (parker) {
+		Nb1 = 1;
+		bpoints = new double[Nb1];
+		bpoints[0] = 3.0;
+	} else {
+		Nb1 = 2;
+		bpoints = new double[Nb1];
+		bpoints[0] = 1.0;
+		bpoints[1] = 3.0;
+	}
 	const int Nf1 = 2;
 	double Bpoints[Nbp] = {0.1, 0.2, 0.5, 1, 2, 5, 10, 20};
-	double npoints[Nnp] = {0.5, 1, 2, 5, 10, 20, 50, 100, 200, 500, 1000};
+	double npoints[Nnp] = {1, 2, 5, 10, 20, 50, 100, 200, 500, 1000, 2000};
 	double fpoints[Nfp] = {0.03, 0.06, 0.1,0.3,0.4,0.5};
 	double vpoints[Nvp] = {0.3 * speed_of_light, 0.4*speed_of_light, 0.5*speed_of_light, 0.6*speed_of_light, 0.65*speed_of_light, 0.7*speed_of_light, 0.75*speed_of_light, 0.8*speed_of_light};
 	double rpoints[Nr0] = {0.1*maxR0, 0.2*maxR0, 0.4*maxR0, 0.5*maxR0};
-	double apoints[Na] = {1,2,3};
-	double bpoints[Nb1] = {1,2,3};
 	double fppoints[Nf1] = {1,2};
 	if(initialGridSearch){
 		for(int i = 0; i < Nbp; ++i){
@@ -1275,6 +1296,8 @@ int main()
 		b = 2.0;
 		fpower = 2.56063;*/
 	}
+	delete[] bpoints;
+	delete[] apoints;
 
 	vector[0] = Bfactor/maxB;
 	vector[1] = concentration/maxN;
@@ -1336,13 +1359,13 @@ int main()
[…]

### 2022-02-22 Radiation/optimization.cpp

@@ -24,6 +24,7 @@ double evaluateOptimizationFunction5(double* vector, double* time, double** nu, 
 		double r = vector[4]*maxR0 + vector[3]*maxV*times[i];
 		//important 0 or Nmonth-1!
 		double rfactor = r/(vector[4]*maxR0 + vector[3]*maxV*times[0]);
+		updateConcentartions(concentrations, vector[3] * maxV / speed_of_light);
 		evaluateAllEmissivityAndAbsorption(nudoppler[i], Inu[i], Anu[i], Nnu[i], Ee, dFe, Np, Nd, Bn, sintheta, psi, thetaIndex, concentrations, vector[1]*maxN, vector[0]*maxB, rfactor, vector[5]*maxBpower, vector[6]*maxNpower, dopplerBeta);
 		//
 		//evaluateSpectrum(nu[i], totalInu, Inu[i], Anu[i], area, length, Nnu, rfactor);

### 2022-02-22 Radiation/spectrum.h

@@ -23,6 +23,8 @@ void evaluateImageFlat(double*** image, double* nu, double**** nudoppler, double
 void evaluateNuDoppler(double***** NuDoppler, int Nmonth, int* Nnumonth, double** Numonth, const double& beta);
 void evaluateNuDoppler(double**** NuDoppler, int Nnu, double* nu, const double& beta);
 
+void updateConcentartions(double*** concentrations3d, double beta);
+
 //for simple flat disk
 void evaluateSpectrumFlatSimple(double* nu, double* totalInu, double* Inu, double* Anu, int Nnu, double rmax, double fraction);
 

### 2022-03-01 disser/part1.tex
fz

@@
 Функция распределения электронов за фронтом ударной волны (от 2000 до 5000 ячеек за фронтом) в момент времени $\displaystyle t$ = 70,000~$\omega_{e}^{-1}$ при начальных параметрах $\displaystyle\sigma = 0.0002$, $\theta = 30^\circ$ и ее аппроксимации с \textcolor{red}{температурой $\displaystyle T_e = 5\times 10^{10}$~K} и спектральным индексом $\displaystyle p = 3.6$ изображены на рисунке ~\ref{electrons}.
+Так же с помощью Particle-in-Cell моделирования можно определить параметры $\epsilon_B, \epsilon_e$, определяющие связь между концентрацией и магнитным полем, которые необходимы для дальнейшего расчета излучения.
 \begin{figure}
@@
 
+Зная точную функцию распределения и предположив геометрию источника в виде плоского диска, можно рассчитать наблюдаемые потоки с помощью известных формул для синхротронного излучения и самопоглощения, описанных например в работе \cite{Ghisellini}.
+\textcolor{red}{Мощность излучения на единицу частоты излучаемая единичным объемом равна}
+\begin{equation} \label{emission}
+	I(\nu)=\int_{E_{min}}^{E_{max}} dE \frac {\sqrt {3}{e}^{3}n F(E) B \sin ( \phi)}{{m_e}{c}^{2}}
+	\frac{\nu}{\nu_c}\int_{\frac {\nu}{\nu_c}}^{\infty }\it K_{5/3}(x)dx,
+\end{equation}
+где $\phi$ угол между магнитным полем и лучом зрения, $\displaystyle\nu_{c}$ критическая частота, определяемая выражением
+$\displaystyle\nu_{c} = 3 e^{2} B \sin(\phi) E^{2}/4\pi {m_{e}}^{3} c^{5}$, и~$K_{5/3}$ функция МакДональда. Коэффициент поглощения равен
+\begin{equation}\label{absorption}
+	k(\nu)=\int_{E_{min}}^{E_{max}}dE\frac {\sqrt {3}{e}^{3}}{8\pi m_e \nu^2}\frac{n B\sin(\phi)}{E^2}
+	\frac{d}{dE} E^2 F(E)\frac {\nu}{ \nu_c}\int_{\frac {\nu}{ \nu_c}}^{\infty }K_{5/3}(x) dx.
+\end{equation}
+Проинтегрировав излучательную способность по объему с учетом самопоглощения и зная расстояние $\displaystyle D$ до источника, можно определить спектральную плотность потока излучения.
+
 \section{Результаты}
+Были проведены расчеты синхротронного излучения от оптического транзиента CSS161010 для момента времени $\displaystyle t = 357$ дней после взрыва, по данным работы \cite{Coppejans2020}.
+Параметры источника, определенные с помощью уравнений~(\ref{ChevR})
+и (\ref{ChevB}), в предположении равенства распределения энергий $\epsilon_e = \epsilon_B = 1/3$, получаются следующими: магнитное поле $\displaystyle B = 0.052$ Гс, радиус оболочки $\displaystyle R = 3.3\times10^{17}$~cm,
+концентрация $n = 1.9$~cm$^{-3}$, спектральный индекс электронов $s = 3.5$, который близок к полученному в PIC расчетах, и минимальна энергия электронов $\displaystyle E_{1} = 4~m_{e}c^{2}$.
+
+Расчеты были проведены для трех конфигураций источника - степенной функцией распределения и параметрами из статьи \cite{Coppejans2020}, с функцией распределения, полученной из PIC моделирования, и с функцией с экстраполированным степенным хвостом до энергий $\displaystyle 500~m_{e} c^{2}$.
+В последних двух случаях размер источника и магнитное поле подбирались минимизацией функционала $\displaystyle g(B,R) = \sum (F(\nu_{i}, B, R) - F_{obs}(\nu_{i}))^{2}$, где $\nu_i$
+наблюдаемые частоты, $\displaystyle F_{obs}(\nu_{i})$ наблюдаемые потоки
+$\displaystyle F(\nu_{i}, B, R)$---рассчитанные потоки.
+Мы использовали шесть измерений из ~\cite{Coppejans2020}: четыре сделанные VLA на 357-ой день
+на частотах 1.5, 3.0, 6.05 и 10~GHz и два сделанные GMRT на 350-ый день на частотах 0.33 и 0.61 GHz. Результаты расчетов излучения показаны на рисунке \ref{radiation}.
+Как можно видеть из рисунка, модель с функцией распределения полученной непосредственно из PIC расчетов не может объяснить наблюдаемое излучение.
+Но так как время расчетов намного порядков меньше времени наблюдения источника, мы считаем возможным экстраполировать степенной хвост функции распределения до более высоких энергий.
+Такая экстраполированная функция распределения, как и чисто степенная, хорошо объясняют наблюдения.
+Но полученные физические параметры источника оказываются существенно различными.
+Магнитное поле $\displaystyle B = 0.069$~G, и радиус источника $\displaystyle R = 3.0 \times
+10^{17}$~cm близки к значениям из статьи \cite{Coppejans2020}, в то время как полученное значение концентрации $\displaystyle n = 210$~cm$^{-3}$ оказывается значительно больше.
+\begin{figure}
+	\centering
+	\includegraphics[width=10.5 cm]{radiation.eps} 
+	\caption{Наблюдаемые 
+		(зеленые круги) и расчетные спектральные плотности излучения CSS161010  на 357-ой день для разных моделей распределения электронов.} 
+	\label{radiation} 
+\end{figure}
 

### 2022-10-11 papers/Physica2022/main.tex
paper

\documentclass[a4paper]{jpconf}
\bibliographystyle{iopart-num}
\usepackage{amsmath}
%\usepackage{citesort}
\usepackage{subfigure}
\usepackage{graphicx}
\graphicspath{{fig/}}
\usepackage{ifpdf}
\ifpdf\usepackage{epstopdf}\fi
\usepackage[export]{adjustbox}
\usepackage[numbers]{natbib}
%----------------------------------------------------- 
%\usepackage{soul,ulem,color,xspace,bm}
% Suggest to remove
%\newcommand{\asrm}[1]{{\color{magenta}\sout{#1}}}
% Suggest to insert
%\newcommand{\as}[1]{\color{cyan}#1\xspace\color{black}}
% Suggest to replace
%\newcommand{\asrp}[2]{\asrm{#1} \as{#2}}
% Comment
%\newcommand{\ascm}[1]{{\color{green}\;AS: #1}}
%------------------------------------------------------

\def\apj{ApJ}
\def\mnras{MNRAS}
\def\nat{Nat}
\def\prd{Phys. Rev. D}
\def\araa{ARA\&A}                % "Ann. Rev. Astron. Astrophys."
\def\aap{A\&A}                   % "Astron. Astrophys."
\def\aaps{A\&AS}                 % "Astron. Astrophys. Suppl. Ser."
\def\aj{AJ}                      % "Astron. J."
\def\apjs{ApJS}                  % "Astrophys. J. Suppl. Ser."
\def\pasp{PASP}                  % "Publ. Astron. Soc. Pac."
\def\apjl{ApJ}                   % letter at ApJ
\def\pasj{PASJ}
\def\apss{Astroph. Space Sci.}
\def\aplett{Astroph. Lett}
\def\ssr{Space Sci. Rev.}
\def\aapr{Astron. Astroph. Reviews}
\def\physrep{Phys. Reports}
\def\memsai{Mem. Societa Astronom. Italiana}
\def\jgr{JGR}
\def\jcap{Journal of Cosmology and Astroparticle Physics}
\def\nar{New Astronomy Reviews}

\usepackage{xspace}
\usepackage[dvipsnames]{xcolor}
\usepackage[normalem]{ulem}
% Suggest to remove
\newcommand{\asrm}[1]{{\color{OrangeRed}\sout{#1}}}
% Suggest to insert
\newcommand{\as}[1]{\color{RoyalBlue}#1\xspace\color{black}}
% Suggest to replace
\newcommand{\asrp}[2]{\asrm{#1} \as{#2}}
% Comment
\newcommand{\ascm}[1]{{\color{ForestGreen}#1}}

\begin{document}
	\title{Kinetic modeling of MHD parameters of mildly-relativistic shocks}

	\author{V I Romansky$^{1}$, A M Bykov$^{1,2}$ and S M Osipov$^{1}$}

	\address{$^1$ Ioffe Institute, 26 Politekhnicheskaya st., St. Petersburg 194021, Russia}
	\address{$^2$ Peter the Great St. Petersburg Polytechnic University, 29 Politekhnicheskaya st., St. Petersburg 195251, Russia}

	\ead{romanskyvadim@gmail.com}

	\begin{abstract}
    Mildly-relativistic outflows with shocks of  velocities  $0.1 - 0.7 c$ were deduced from multiwavelength observations of powerful fast transient sources. These outflows are associated with  merging relativistic objects, relativistic supernovae and fast blue optical transients. Relativistic  magneto-hydrodynamic (RMHD) models of this objects rely on the equation of state of the fluid which is a collisionless plasma with a contribution of non-thermal components. In this paper we present kinetic simulations of mildly-relativistic shocks with - Particle-in-Cell and Monte-Carlo techniques to derive the adiabatic index of plasma in shock downstream  directly from particle distributions which can be implemented into RMHD models. 
	\end{abstract}

\section{Introduction} 
Recent multiwavelength observations of fast energetic transient sources associated with some classes of supernova and neutron star mergers revealed a presence there of mildly-relativistic outflows with shock waves of velocities faster than 0.1c \citep{2016ApJ...819...35A,2019ApJ...872...18M,2020ApJ...895...49H,2021ApJ...912L...9N,2019LRR....23....1M,2021ARA&A..59..155M,2021ApJ...911..104K,2022ApJ...927L..17H}. 
Analysis of 42 day-timescale evolving transients detected with Zwicky Transient Facility (ZTF) \citep{ZTF_stat_Ho21} suggested that most of these are likely associated with core-collapse supernovae (SNe). The authors  distinguished a few sub-types of the typical events as   (i) subluminous SNe of Type Ib or IIb; (ii) luminous Type Ibn or hybrid IIn/Ibn SNe; and (iii) short-duration radio-loud luminous events which prototype is nearby AT2018cow event. While the subluminous SNe events of Type IIb are the most numerous the AT2018cow like events rate is less than 0.1\% of the local core-collapse SNe rate \citep{ZTF_stat_Ho21}. The multiwavelength data on fast SNe related transients can be generally understood assuming an action of a powerful central engine  in the collapsing stars which can launch a relativistic jet-type outflow. Relativistic hydrodynamical  simulations  performed in \citep{2012ApJ...750...68L,2021NewAR..9201614C} illustrated that depending on the time duration of the central engine power activity either GRB type jet source (for a long enough activity time) or, for a shorter power injection time, somewhat broader outflow and radio bright relativistic SN can be produced by a collapsing star. The different energy injection regimes by the central engine result in different energy versus the ejecta speed distributions. The powerful jet breaking through the stellar envelope can form  a mildly-relativistic expanding cocoon containing the energy $\sim$ 10$^{51}$ ergs (see e.g. \citep{2019ApJ...871L..25P}).    
The the cocoon interacting with the circumstellar winds may emit the synchrotron self-absorbed radio emission observed in fast optical transients \citep{2022MNRAS.tmp..970G}. The physical models of particle acceleration in fast transients based on particle in cell simulations of mildly-relativistic shocks in barionic plasma produced by the central engine activity in the fast transients were discussed in \cite{BRO22_Univ}. The mildly relativistic shocks are also determining the transition from the early highly-relativistic to the later time semi-relativistic likely barion-dominated outflows in the gamma-ray burst afterglows (see e.g. \citep{Piran04,Meszaros06,GRBafterglow_hydro12}) where the models can be used to model the non-thermal radiation. Moreover, mildly relativistic shocks in stellar mass transient sources can be considered as efficient accelerators of cosmic rays well above PeV regime (see \citep{2018SSRv..214...41B} and the references therein). The future  Large Synoptic Survey Telescope will detect a large amount of fast transients providing a good possibilities to of the follow up multiwavelength studies. Therefore there is clear need in detailed modeling of different appearances of the semi-relativistic outflows in the transient sources. While to model the global structure of the flows the RMHD simulations are widely used the collisionless shocks need the microscopic kinetic type of simulations since the collisionless shocks are producing non-thermal components which may influence the equation of state and the macroscopic parameters like the adiabatic indexes of the semi-relativistic plasma. Therefore we present below the results of a simulations of macroscopic plasma parameters which can be used in the global RMHD simulations.

\section{Particle-in-Cell simulation}
In this work we use particle-in-cell code Smilei \cite{Smilei18} for modeling collisionless shocks. Simulation domain is two-dimensional with reflective wall on the left boundary along the x axis and the plasma flowing in through the right boundary. Boundary conditions along the y axis are periodic.

The simulation parameters are : the initial flow Lorentz factor $\Gamma$ is in interval $1.05-1.5$ for different setups, the flow magnetization $\sigma = \frac{B^2}{4\pi\Gamma (n_0 m_p + n_0 m_e) c^2} = 0.0002$, where B is magnetic field, $m_p$ and $m_e$ are proton and electron masses, $c$ is speed of light. Upstream concentration $n_0$ is equal to $1$, but all quantities can be easily scaled to different value. The temperature $T = 0.02$ in units of the electron rest energy and the electron mass is increased up to $m_e = \frac{m_p}{100}$. We used setups with different velocities of plasma flow $v_u = \beta_u c$ and inclination angles of magnetic field: $\theta$ is angle between the magnetic field and the flow velocity and $\phi$ is angle of field rotation in perpendicular plane, $\phi=0$ corresponds to the magnetic field lying in the simulation plane. Setups and they parameters are listed in table \ref{setups}. The spatial grid step is $dx = 0.2 c/{\omega_p}$ and the time step $dt = 0.1/{\omega_p}$, where $\omega_p = \sqrt{\frac{4\pi q^2 n_0}{m_e}}$ is the plasma frequency, $\displaystyle q$ is the
absolute value of the electron charge. The size of the simulation box along the $x$ axis is $L_x = 40000\frac{c}{\omega_p}$ and in the transverse direction $L_y = 100\frac{c}{\omega_p}$. These scales correspond to $200000$ and $500$ grid points in $x$ and $y$ directions, respectively.

We obtain MHD parameters of the shock from the simulation. The position of the shock front is well defined as shown in figure \ref{density_noturb} so we can evaluate the shock velocity in the downstream frame $v^d_{sh}$  and $\beta^d_{sh} = {v^d}_{sh}/c$. The shock velocity in the upstream frame is defined as $\beta_sh = (\beta_u + \beta^d_{sh})/(1 + \beta_u \cdot\beta^d_{sh})$.
	\begin{figure}[h!]
		\centering
		\begin{minipage}{0.49\textwidth}
			\centering
			\includegraphics[width=0.98\textwidth]{concentrations.eps} 
			\caption{Time evolution of concentration normalized to the far upstream~concentration, setup B30.}
			\label{density_noturb}
		\end{minipage}\hfill
		\begin{minipage}{0.49\textwidth}
			\centering
			\includegraphics[width=0.98\textwidth]{electrons2.eps} 
			\caption{Fit of the electron distribution in the shock downstream with Maxwell-Juttner distribution, setup B30.}
			\label{density_turb}
		\end{minipage}
	\end{figure}
We can find temperature of every particle species behind the shock, minimizing the 
functional $\displaystyle f(T) = \int_{m c^{2}}^{5 m c^{2}} (F(E) - F_{mj}(E,T))^{2}dE$, where $\displaystyle F(E)$ is simulated distribution function and $\displaystyle F_{mj}(E,T)$ is Maxwell-Juttner distribution function. Also we can derive adiabatic index for every species of particles using definition $\gamma_i = 1 + P/E_k$ where $P$ is the pressure and  $E_k$ is the kinetic energy of particles in the plasma rest frame. $P = \int F(E) v_x(E,\theta) p_x(E,\theta) 2 \pi sin(\theta) d\theta dE$ and $E_k = \int F(E) (E - m c^2) 2 \pi sin(\theta) d\theta dE$. Also we define $\hat\gamma$ - the adiabatic index, evaluated with same formulae but using the Maxwell-Juttner particle distribution with the corresponding temperature.
	\begin{table}[h!]
	\label{setups}
	\caption{Parameters of different setups. }
	\begin{center}
		\begin{tabular}{|c | c| c| c| c| c| c| c| c| c| c| c|}
			\hline
			Setup & $\theta$ & $\phi$ & ${\beta}_u$ &  ${\beta^d}_{sh}$ & ${\beta}_{sh}$ & ${\gamma}_p$ & ${\gamma}_e$ & $T_p\ 10^{10}K$ & $T_e \ 10^{10}K$ & $\hat{\gamma}_p(T_p)$ & $\hat{\gamma}_e(T_e)$\\
			\hline
			A30 & 30 & 90 & 0.5 &  0.13 & 0.59 & 1.620 & 1.389 &75.5 & 19.0 & 1.617 & 1.389\\
			A80 & 80 & 90 & 0.5 & 0.16 & 0.61 & 1.601 & 1.396 &54.7 & 16.3 & 1.629 & 1.397\\	
			B30  & 30 & 90 & 0.3  & 0.088 & 0.38 & 1.647 & 1.499 & 14.5 & 4.0 & 1.656 & 1.502\\
			B50  & 50 & 90 & 0.3  & 0.098 & 0.39 & 1.648 & 1.468 & 11.0 & 5.9 & 1.658 & 1.469\\
			C30  & 30 & 0 & 0.3  & 0.088 & 0.38 & 1.647 & 1.502 & 14.4 & 3.9 & 1.656 & 1.504\\
			D30 & 30 & 90 & 0.1 &  0.035 & 0.135 & 1.665 & 1.574 &1.9 & 1.3 & 1.665 & 1.589\\
			D80 & 80 & 90 & 0.1 & 0.052 & 0.151 & 1.664 & 1.580 &2.6 & 1.5 & 1.665 & 1.579\\
			\hline
		\end{tabular}
	\end{center}
    \end{table}

One can see from the table that adiabatic index obtained from PIC simulation is smaller than from simple hydrodynamic theory. It should be taken into account in MHD simulations. This difference increases when particle acceleration is more efficient (case of quasi-parallel shock). Also, PIC simulation can not simulate long non-thermal tales of distributions because of it's high computational cost, and other methods, such as hybrid and Monte-Carlo simulation are needed for more precise modeling of MHD parameters.

\section{Monte Carlo simulation}	
Due to the limited computing power, PIC modeling can be performed only in a small area of the volume of real astrophysical objects. To describe astrophysical objects on their real scales, it is necessary to involve other numerical models. One of such models is Monte Carlo calculations, which, unlike PIC calculations, require the introduction of phenomenological laws, such as the mean free path of particles, the growth rates of plasma instabilities.

We develop a nonlinear numerical stationary plane-parallel relativistic Monte Carlo model of particle acceleration by longitudinal collisionless shocks \cite{BRO22_Univ}. Particle acceleration occurs by the first-order Fermi mechanism. The particles are scattered by magnetic fluctuations and repeatedly cross the shock front. In our model, on the basis of an iterative scheme, the conservation laws of energy and momentum fluxes near the shock are fulfilled. The model takes into account the modification of the upstream by the pressure of accelerated particles, the amplification of magnetic fluctuations due to plasma instabilities caused by the anisotropy of the accelerated particle distribution function in the upstream, the dissipation of turbulent modes, and the turbulent cascade. Particle propagation is organized on the basis of pitch-angle scattering. The particles are divided into acceleration and background. An accelerated particle is considered if it has crossed the front of the shock at least once from the downstream to the upstream.

The Table~\ref{MC_calculations} shows the results of Monte Carlo calculations for different shock velocities $u_{sh}$ ($\beta_{sh}=u_{sh}/c$). ${\gamma}_{th}$, ${\gamma}_{cr}$, ${\gamma}_{tot}$ are the adiabatic indices of the background plasma, the accelerated particle distribution and the total in the downstream, respectively. $T_{th}$ is the temperature of background protons in the downstream. Other model parameters except $u_{sh}$ are the same in all calculations. The free escape boundary of particles is at the coordinate $x_{FEB}=-5\cdot 10^{14}$ cm, the shock front corresponds to the coordinate x = 0. The number density of the background plasma $n_{0}=5\cdot 10^{5}$ cm$^{-3}$ in the far unperturbed upstream. The rms turbulent magnetic field $B_{st}\left(x_{FEB}\right)$ is equal to the constant magnetic field $B_{0}=3\cdot 10^{-3}$ G in the far unperturbed upstream. $B\left(x\right)=\sqrt{B_{st}^{2}\left(x\right)+B_{0}^{2}}$.

	\begin{table}[h!]
	\label{MC_calculations}
	\caption{Results of Monte Carlo calculations. }
	\begin{center}
		\begin{tabular}{|c | c| c| c| c| c| }
			\hline
			Setup  & ${\beta}_{sh}$ & ${\gamma}_{th}$ & ${\gamma}_{cr}$ & ${\gamma}_{tot}$ & $T_{th}\ 10^{10}K$ \\
			\hline
			MC1  & 0.1 & 1.66 & 1.43 & 1.49 & 0.59 \\
			MC2  & 0.3 & 1.66 & 1.42 & 1.49 & 7.11 \\
			MC3  & 0.5 & 1.64 & 1.41 & 1.49 & 26.5 \\
			MC4  & 0.7 & 1.61 & 1.40 & 1.48 & 82.4 \\

			\hline
		\end{tabular}
	\end{center}
    \end{table}

The Figure~\ref{u_B_eff_MC1} shows the profiles of the background plasma flow velocity and the magnetic field. The turbulent part of the magnetic field is amplified due to plasma instabilities and adiabatically. $r_{g0}=m_{p}cu_{sh}/eB_{0}$, where $e$ is elementary charge.

\begin{figure}[h!]
\begin{center}
\includegraphics[scale=0.5]{u_B_eff_MC1.eps}
\end{center}
\caption{Background plasma velocity profile (upper panel) and magnetic field profile (bottom
panel) for the Monte Carlo calculation MC1.}
\label{u_B_eff_MC1}
\end{figure}

The scales achievable in PIC modeling are of the order of several $r_{g0}$. That is, PIC modeling can describe a small area near the shock compared to Monte Carlo modeling (see Figure~\ref{u_B_eff_MC1}). Since the maximum energies of accelerated particles strongly depend on the size of the system, in Monte Carlo calculations, the maximum energies of particles are many orders of magnitude higher than in PIC calculations. The temperature of the background plasma in the downstream in Monte Carlo simulation is also affected by the amplification of modes by plasma instabilities associated with the anisotropy of the distribution of  high-energy accelerated particles and the energy dissipation of these modes at the upstream scales.

\section{Conclusions}
Hydrodynamic and RMHD models are constructed to interpret the light-curves and spectra of fast transients (e.g. \citep{1999ApJ...510..379M,2012ApJ...750...68L,2019LRR....23....1M,2020ApJ...903...66L,2021ApJS..256....8U,2022MNRAS.tmp..970G,2022arXiv220108432E} and the reference therein).
The parameters obtained in kinetic modeling of MHD shocks with non-thermal components can be used in the hydrodynamic modeling of  global structures of fast mildly-relativistic outflows.

\ack
	Some results of the work were obtained using computational resources of Peter the Great Saint-Petersburg Polytechnic University Supercomputing Center (http://scc.spbstu.ru)

%\section*{References}

%\providecommand{\newblock}{}

%\begin{thebibliography}{10}
%\expandafter\ifx\csname url\endcsname\relax
%  \def\url#1{{\tt #1}}\fi
%\expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL }\fi
%\providecommand{\eprint}[2][]{\url{#2}}
% Bibliography created with iopart-num v2.0
% /biblio/bibtex/contrib/iopart-num

% \bibliographystyle{apalike}
\bibliographystyle{iopart-num}

\bibliography{FBOT_bibl1}
%

\end{document}

### 2022-12-04 papers/distribution transform.tex
distribution transformation paper

\documentclass[a4paper,12pt]{jpconf}

\usepackage{cmap}
\usepackage[T2A]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[english,russian]{babel}

%\usepackage{makeidx}
\usepackage{amsmath,amssymb,amsthm,amsfonts,latexsym,mathtext}
\usepackage[pdftex]{graphicx}
\usepackage{float}
%\usepackage{citesort}
\usepackage{subfigure}
\graphicspath{{fig/}{pict3/}}
\usepackage{ifpdf}
\ifpdf\usepackage{epstopdf}\fi

\usepackage{bm,mathtools}
\usepackage{geometry}
\geometry{top=2cm}
\geometry{bottom=2.5cm}
\geometry{left=2.5cm}
\geometry{right=2cm}

\bibliographystyle{iopart-num}

\newcommand{\be}{\begin{equation}}
	\newcommand{\ee}{\end{equation}}
\newcommand{\ba}{\begin{aligned}}
	\newcommand{\ea}{\end{aligned}}

\def\lsim{\;\raise0.3ex\hbox{$<$\kern-0.75em\raise-1.1ex\hbox{$\sim$}}\;}
\def\gsim{\;\raise0.3ex\hbox{$>$\kern-0.75em\raise-1.1ex\hbox{$\sim$}}\;}
\begin{document}
	\title{Преобразование функции распределения}
	\section{Преобразование функции распределения фотонов}
	Функция распределения фотонов задана в сферических координатах $n_{ph}(\epsilon,\mu,\phi)$. Рассмотрим переход в систему отсчета, движущуюся в направлении оси $z$ с лоренц-фактором $\gamma = 1/\sqrt{1-\beta^2}$. Количество частиц в элементе фазового пространства $N$ - инвариант.
	\begin{equation}
		N = n_{ph}(\epsilon,\mu,\phi)d\epsilon d\mu d\phi dV = n'_{ph}(\epsilon',\mu',\phi')d\epsilon' d\mu' d\phi' dV'
	\end{equation}
	Рассмотрим преобразование вектора четырех-импульса. Поперечные компоненты не изменяются, а временная и продольная меняются следющим образом, учитывая что $p_z = \mu \epsilon$:
	\begin{equation}\label{lorentz_ph}
		\left(\begin{array}{c}
			\epsilon'\\
			\mu'\epsilon'
		\end{array}
		\right)
		= \left(
		\begin{array}{cc}
			\gamma & -\beta\gamma\\
			-\beta\gamma & \gamma
		\end{array}
		\right)
		\times
		\left(\begin{array}{c}
			\epsilon\\
			\mu\epsilon
		\end{array}
		\right)
	\end{equation}
	Из первой строчки матрицы получаем уравнение для допплеровского сдвига энергии
	\begin{equation}\label{doppler_ph}
		\epsilon'=\gamma(1-\mu\beta)\epsilon
	\end{equation}
	Вычислим производные новой энергии по старым координатам
	\begin{equation}
		\frac{d\epsilon'}{d\epsilon}=\gamma(1-\mu\beta)
	\end{equation}
	\begin{equation}
		\frac{d\epsilon'}{d\mu}=-\gamma\beta\epsilon
	\end{equation}
	Из второй строчки матрицы получаем $\mu'\epsilon'=-\beta\gamma\epsilon+\gamma\mu\epsilon$. Подставив значение $\epsilon'$ из \ref{doppler_ph} и сократив $\epsilon$ получим уравнение аберрации света
	\begin{equation}\label{aberration_ph}
		\mu'=\frac{\mu-\beta}{1-\mu\beta}
	\end{equation}
	Заметим, что угол наклона луча в новой системе не зависит от энергии в старой системе. Вычислим частноую производную $\frac{d\mu'}{d\mu}$
	\begin{equation}
		\frac{d\mu'}{d\mu}=\frac{d}{d\mu}\frac{1}{\beta}\frac{\beta\mu-1+1-\beta^2}{1-\mu\beta}=\frac{d}{d\mu}\frac{1}{\beta}\frac{1-\beta^2}{1-\mu\beta}=\frac{1-\beta^2}{(1-\mu\beta)^2}=\frac{1}{\gamma^2(1-\mu\beta)^2}
	\end{equation}
	Азиметальный угол не зависит от системы отсчета $\phi' = \phi$. Преобразование элемента объема описывается выражением $\frac{dV'}{dV} = \frac{\epsilon}{\epsilon'}$ см. ЛЛ Т2 параграф 10, вот только там используется переход в собственную систему. То есть
	\begin{equation}
		\frac{dV'}{dV} = \frac{1}{\gamma(1-\mu\beta)}
	\end{equation}
	Матрица якоби преобразования координат выглядит следующим образом
	\begin{equation}
		J=\left(
		\begin{array}{cccc}
			\frac{d\epsilon'}{d\epsilon} & \frac{d\epsilon'}{d\mu}& 0 & 0\\
			0 & \frac{d\mu'}{d\mu} & 0 & 0\\
			0 & 0 & 1 & 0\\
			0 & \frac{dV'}{d\mu} & 0 & \frac{dV'}{dV}
		\end{array}
		\right)
	\end{equation}
	При такой матрице якобиан, к счастью, равен произведению диагональных членов
	\begin{equation}\label{jacobian_ph}
		\frac{D(\epsilon',\mu',\phi',V')}{D(\epsilon,\mu,\phi,V)}=\frac{d\epsilon'}{d\epsilon}\frac{d\mu'}{d\mu}\frac{dV'}{dV}=\gamma(1-\mu\beta)\frac{1}{\gamma^2(1-\mu\beta)^2}\frac{1}{\gamma(1-\mu\beta)}=\frac{1}{\gamma^2(1-\mu\beta)^2}
	\end{equation}
	И в итоге функция распределения фотонов преобразуется с помощью деления на вычисленный якобиан
	\begin{equation}\label{distribution_ph}
		n'_{ph}(\epsilon',\mu',\phi') = \frac{n_{ph}(\epsilon,\mu,\phi)}{\frac{D(\epsilon',\mu',\phi',V')}{D(\epsilon,\mu,\phi,V)}}=\gamma^2(1-\mu\beta)^2 n_{ph}(\epsilon,\mu,\phi)
	\end{equation}
\end{document}

### 2022-12-04 Compton/main.cpp

@@ -201,10 +201,10 @@ int main()
 	const double Bfactor = 0.29;
 	const double epsilonB = 0.0012;
 	//const double electronConcentration = 10*Bfactor*Bfactor/(4*pi*massProtonReal*speed_of_light2*epsilonB);
-	const double electronConcentration = 25;
+	const double electronConcentration = 1E10;
 	const double photonConcentration = 1.0;
 	//double rmax = 0.3 * speed_of_light * 357 * 24 * 3600;
-	double rmax = 1.4E17;
+	double rmax = 1E14;
 
 
 	const int Nphi = 2;
@@ -268,13 +268,13 @@ int main()
 	double Tphotons5 = 7500;
 	double a1 = 1.0;
 	//double a1 = 15*L*cube(hplank*speed_of_light)/(32*pi*pi*pi*pi*pi*pi*speed_of_light*rmax*rmax*pow(kBoltzman*Tphotons1,4));
-	a1 = 0;
+	//a1 = 0;
 	double a2 = 4E-4;
 	a2 = 0;
 	double a3 = 4.0E-13;
 	a3 = 0;
 	double a4 = 1.65E-13;
-	//a4 = 0;
+	a4 = 0;
 	double a5 = 1E-14;
 	a5 = 0;
 	double* Fph = new double[Nphot];
@@ -296,7 +296,7 @@ int main()
 	dFph[0] = 0;
 	//todo check and normalize
 	//for intergalactic field
-	/*for (int i = 1; i < Nphot; ++i) {
+	for (int i = 1; i < Nphot; ++i) {
 		//todo pi or not to pi?
 		Fph[i] = 0;
 		dFph[i] = 0;
@@ -312,10 +312,10 @@ int main()
 		theta = Eph[i] / (kBoltzman * Tphotons5);
 		Fph[i] += a5 * (2 * Eph[i] * Eph[i] / cube(hplank * speed_of_light)) / (exp(theta) - 1.0);
 		dFph[i] = Fph[i]*(Eph[i] - Eph[i-1]);
-	}*/
+	}
 
 	//for cluster
-	double R = 1E18;
+	/*double R = 1E18;
 	double numax = 1E15;
 	double westerlundD = 4000 * 3E18;
 	double westerlundNuFNu = 1E14;
@@ -333,7 +333,25 @@ int main()
 		double theta = Eph[i] / (kBoltzman * temperature);
[…]

### 2022-12-04 Compton/startparameters.h

@@ -8,9 +8,9 @@ enum Geometry {FLAT_SIMPLE, FLAT, SPHERICAL};
 const Geometry geometry = Geometry::SPHERICAL;
 
 enum Input {TRISTAN, SMILEI, MAXWELL, POWERLAW, COMBINED, LONG_COMBINED};
-const Input input = Input::COMBINED;
+const Input input = Input::POWERLAW;
 
 enum Solver {DUBUS, UVAROV, THOMPSON};
-const Solver solver = Solver::UVAROV;
+const Solver solver = Solver::DUBUS;
 
 #endif

### 2022-12-04 papers/distribution transform.tex
fz

@@
 	\end{equation}
+\section{Комптоновское рассеяние}
+Рассмотрим рассеяние фотонов на одном электроне. Сечение Клейна-Нишины в системе покоя электрона равно
+\begin{equation}
+	\frac{d\sigma}{d\epsilon_1'd\Omega_1'}=\frac{{r_e}^2}{2}\left(\frac{\epsilon_1'}{\epsilon_0'}\right)^2\left(\frac{\epsilon_1'}{\epsilon_0'}+\frac{\epsilon_0'}{\epsilon_1'}-sin^2\Theta'\right)
+\end{equation}
+Где $r_e$ - классический радиус электрона, $\epsilon_0'$ и $\epsilon_1'$ - энергии начального и конечного фотона, соответственно, $\Theta'$ - угол между начальным и конечным фотоном.
+Штрихованные индексы относятся к системе отсчета электрона.
+Число фотонов, 
 \end{document}

### 2022-12-04 Compton/main.cpp

@@ -832,7 +832,7 @@ int main()
 									double photonInitialCosThetaPrimed = (photonInitialCosTheta - electronInitialBeta)/(1.0 - electronInitialBeta*photonInitialCosTheta);
 									double photonInitialSinThetaPrimed = sqrt(1.0 - photonInitialCosThetaPrimed*photonInitialCosThetaPrimed);
 
-									double coef = electronConcentration*concentrations3d[ir][itheta][iphi]*volume[ir][itheta][iphi]*(re2*speed_of_light*sqr(1.0 - electronInitialBeta*photonInitialCosTheta)/(2*electronInitialGamma*(1.0 - electronInitialBeta*photonFinalCosTheta)));
+									double coef = electronConcentration*concentrations3d[ir][itheta][iphi]*volume[ir][itheta][iphi]*(re2*speed_of_light*sqr(1.0 - electronInitialBeta*photonInitialCosTheta)/(2*sqr(electronInitialGamma)*(1.0 - electronInitialBeta*photonFinalCosTheta)));
 									double derivativeEE = electronInitialGamma-electronInitialGamma*electronInitialBeta*photonInitialCosTheta;
 									//printf("%g\n", derivativeEE);
 									//derivativeE = 1.0;
@@ -871,7 +871,7 @@ int main()
 										I[i] += coef*
 											(1.0 + cosXiPrimed*cosXiPrimed + sqr(photonFinalEnergyPrimed/(massElectron*speed_of_light2))*sqr(1.0 - cosXiPrimed)/(1.0 - (photonFinalEnergyPrimed/(massElectron*speed_of_light2))*(1.0 - cosXiPrimed)))*
 											//2*pi*dcosThetaFinal[j]*dphi*dcosThetaInitial*delectronEnergy*Fe[iangle][k]*evaluatePhotonDistribution(photonInitialEnergy, Nphot, Eph, Fph)/(derivativeE*derivativeEE - derivativeXi*(derivativeXiMu0*derivativeMu0E + derivativeXiMu1 * derivativeMu1E));
-											2*pi*dcosThetaFinal[j]*dphi*dcosThetaInitial*delectronEnergy*Fe[iangle][k]*evaluatePhotonDistribution(photonInitialEnergy, Nphot, Eph, Fph)/derivativeEE;
[…]

### 2023-04-08 papers/FBOTcompton/fbotcompton.tex
paper

%\documentclass[a4paper,12pt]{report}
\documentclass[a4paper,12pt]{extreport}

\bibliographystyle{iopart-num}
\usepackage{amsmath}
%\usepackage{citesort}
\usepackage{subfigure}
\usepackage{graphicx}
\graphicspath{{fig/}}
\usepackage{ifpdf}
\ifpdf\usepackage{epstopdf}\fi
\usepackage[export]{adjustbox}
\usepackage[numbers]{natbib}

\def\apj{ApJ}
\def\apjl{ApJ Lett.}
\def\mnras{MNRAS}
\def\nat{Nature}
\def\physrevB{Phys. Rev. B}
\def\prd{Phys. Rev. D}
\def\prl{Phys. Rev. Lett.}
\def\pre{Phys. Rev. E}
\def\araa{Ann. Rev. Astron. Astrophys.}                % "Ann. Rev. Astron. Astrophys."
\def\aap{Astron. Astrophys.}                  % "Astron. Astrophys."
\def\aaps{Astron. Astrophys. Suppl. Ser.}                 % "Astron. Astrophys. Suppl. Ser."
\def\aj{Astron. J.}                      % "Astron. J."
\def\apjs{Astrophys. J. Suppl. Ser.}                  % "Astrophys. J. Suppl. Ser."
\def\pasp{Publ. Astron. Soc. Pac.}                  % "Publ. Astron. Soc. Pac."
\def\apjl{ApJ Lett.}                   % letter at ApJ
\def\pasj{PASJ}
\def\apss{Astroph. Space Sci.}
\def\aplett{Astroph. Lett.}
\def\ssr{Space Sci. Rev.}
\def\jcap{J. Cosmol. Astropart. Phys.}
\def\apspr{Sov. Sci. Rev. Sect. E}
\def\nar{New Astron. Rev.}
\def\aapr{ Astron. Astrophys. Rev.}

\begin{document}
	\title{On X-ray emission from Fast Blue Optical Transients}

	\author{A M Bykov$^{1}$, V I Romansky$^{1}$}

Recent observations of Fast Blue Optical Transients \cite{Margutti2019,Ho2020,Coppejans2020} show high level of X-ray radiation. There are suggested different scenarios of  origins of this radiation: central engine or interaction of shock with external medium. In this paper we present model of interaction of shock with strong wind from close star, producing strong X-ray emission via inverse compton scattering on the photons of this star.

%написать почему не хватает комптона от всей оболочки

In our model we took two close stars, on a distance $1.4\cdot10^{17} cm$, wich correspondes to the radius of expanding shock of CSS161010 at 99 day after explosion. Such small distances are possible in young massive star clusters \cite{}. Also massive star clusters usually have dozens of bright stars with powerfull stellar winds (OB, Wolf-Rayet, and red supergiants). We will refer to them as first star - stable, generating wind, and the second - exploded as supernova. We assumed both stars have winds with velocity $v_w = ?$ and mass loss $\dot{M} = 10^{-5} M_\odot $ per year and luminocity $L=500000 L_\odot$ and surfece temperature $T = 2.5\cdot10^6 K$ which is typical for Wolf-Rayet stars. The magnetic fied of the wind is initialized using \cite{}, and for the first star  it's amplitude is $100 G$ at the radius $10$ while for the second we need much stronger field to satisfy synchrotron observational data, as it is shown in papers \cite{Coppejans2020, BykovUniverse}. So we took magnetic field $1000 G$ at the radius $10^12$, stars with such strong fields are described in \cite{}.

For hydrodinamcal simulation we use code PLUTO \cite{MignonePluto}. We used several nested setups on different scales. At first we performed 3d modeling of two stable stars with parameters described above until they winds form a stable configuration. Sizes of the simulation box are $Lx = $ with number of grid points respectively. The next step was 1d spherical modeling of second star explosion. We use ejecta mass $M_{ej} = 0.02 M_\odot$ and velocity $v_{ej} = 0.3 c$ and initial conditions as described in \cite{}. At the third step we use results of the previous steps as initial and boundary conditions in setup with smaller scale - with size $Lx = $ . And at the last step, in setup with the smallest scale $Lx = $, wich is necessary to resolve bow- and tremination shock of the wind of the first star, we use results of the third step as initial and boundary conditions. Scheme of setups is shown in Figure \ref{}.

\bibliographystyle{iopart-num}
\bibliography{fbotcompton}
\end{document}

### 2023-04-09 papers/FBOTcompton/fbotcompton.tex
paper

@@
 The magnetic fied of the wind is initialized using \cite{}, and for the first star  it's amplitude is $100 G$ at the radius $10$ while for the second we need much stronger field to satisfy synchrotron observational data, as it is shown in papers \cite{Coppejans2020, BykovUniverse}.
-So we took magnetic field $1000 G$ at the radius $10^12$, stars with such strong fields are described in \cite{}.
+So we took magnetic field $1000 G$ at the radius $10^{12}$, stars with such strong fields are described in \cite{}.
 
@@
 
+Simulations showed that ejecta of supernovae, moving in interstellar medium has profile with thin and dense peak, with parameters number density $n \approx 10^5$ and width $d = 1E^{15}$ which is consistent of results of modeling of synchrotron radiation.
+%вставить куда-то про расчет синхротрона
+
+The wind shock has radius $R_w \approx 0.5\times10^14 cm$ and number density $n_w \approx 10^9 cm^{-3}$.
+With this parameters, and luminocity and temperature of the star, one can estimate observed X-ray flux, produxed by inverse compton scattering.
+For evaluating inverse compton radiation we use our numerical code FAINA.
+We assume that source is a cylinder with radius $R_w$, and height also equals $R_w$, with axis oriendted along line of sight, and with uniform number density $n_w$.
+Electrons distribution function is
+%непонятно чт писать
+and photons distribution corresponds to the radiation of the first star at distance $R_w$.
+With this parameter modeled X-ray flux in the energy range $0.3-10 keV$ is $F_m = $ which is consistent with observed flux of CSS161010 at 99 day after explosion $F_{obs = }$ \cite{Coppejans2020}.
+
 \bibliographystyle{iopart-num}

### 2023-04-12 papers/FBOTcompton/fbotcompton.tex
paper

@@
 %написать почему не хватает комптона от всей оболочки
-
+\section{hydrodynamic}
 In our model we took two close stars, on a distance $1.4\cdot10^{17} cm$, wich correspondes to the radius of expanding shock of CSS161010 at 99 day after explosion.
@@
 
-Simulations showed that ejecta of supernovae, moving in interstellar medium has profile with thin and dense peak, with parameters number density $n \approx 10^5$ and width $d = 1E^{15}$ which is consistent of results of modeling of synchrotron radiation.
-%вставить куда-то про расчет синхротрона
+Simulations showed that ejecta of supernovae, moving in interstellar medium has profile with thin and dense peak, with parameters number density $n \approx 10^5$ and width $d = 1E^{15}$.
 
@@
 With this parameters, and luminocity and temperature of the star, one can estimate observed X-ray flux, produxed by inverse compton scattering.
-For evaluating inverse compton radiation we use our numerical code FAINA.
-We assume that source is a cylinder with radius $R_w$, and height also equals $R_w$, with axis oriendted along line of sight, and with uniform number density $n_w$.
-Electrons distribution function is
+\section{radiation}
+For evaluating radiation we use our numerical code FAINA, which allows to evaluate synchrotron and inverse compton radiation and also fit model to observational data.
+
+We assume that source of synchrotron radio flux, observed from CSS161010 is spherical shell with uniform density and magnetic field.
+Distribution function of emitting is evaluated using PIC simulation with code SMILEI \cite{Derouillat}.
+Setups for shock simulation is decribed in details in our previous work \cite{BykovUniverse}.
+Shock was initialized due to collision of uniform plasma flow with velocity $0.3 c$ with reflecting wall.
+Magnetiztion was $\sigma = B^2/4 pi m_p \gamma c^2 = 0.002$ and diffirent inclination angles of magnetic field were studied.
+Distributions for different inclination angles of magnetic field relative to local shock velocity are taken in different points of the source.
+With this assumptions we can fit our model to observational data of radio flux from CSS161010 at 98 day after explosion (the first observation when both radio and X-ray flux was detected), and derive such parameters as magnetic field, concentration and width of the spherical shell.
+Our best fit for radio flux and observational data are shown on Figure \ref{}.
+Evaluated parameters are magnetic field $B = 0.6 G$, number density $n = 50000{cm}^{-3}$ and shell width $l = 0.01\cdot R = 1.3E15 cm$
+
+We assume that source is a cylinder with radius $R_w$, and height also equals $R_w$, with axis oriendted along line of sight, and with uniform number density $n_w$. Electrons distribution function is 
 %непонятно чт писать

### 2023-04-12 SimpleRadiation/startparameters.h

@@ -5,7 +5,7 @@ const bool parker = false;
 const bool turbulence = false;
 
 enum Doppler { NO, INTEGER, DIFFERENTIAL };
-const Doppler doppler = Doppler::NO;
+const Doppler doppler = Doppler::INTEGER;
 
 enum Geometry {FLAT_SIMPLE, FLAT, SPHERICAL};
 const Geometry geometry = Geometry::SPHERICAL;

### 2023-04-23 papers/FBOTcompton/fbotcompton.tex
paper

@@
 Our best fit for radio flux and observational data are shown on Figure \ref{}.
-Evaluated parameters are magnetic field $B = 0.6 G$, number density $n = 50000{cm}^{-3}$ and shell width $l = 0.01\cdot R = 1.3E15 cm$
+Evaluated parameters are magnetic field $B = 0.6 G$, number density $n = 50000{cm}^{-3}$ and shell width $l = 0.01\cdot R = 1.3E15 cm$.
+This parameters are consistent with obtained from hydrodynamical simulations.
 
-We assume that source is a cylinder with radius $R_w$, and height also equals $R_w$, with axis oriendted along line of sight, and with uniform number density $n_w$. Electrons distribution function is 
+In our model X-ray flux is generated via inverse compton scattering in the region of iteraction between wind from the first star, and supernova ejecta from the second.
+Relativistic ejecta has huge bulk pressure and wind is compressed to small region with high number dencity.
+Also this region is close to high luminous star, and there are suitable conditions for strong inverse compton radiation.
+Electrons can be accelerated on the shock fronts or on the shearing flows.
+So electrons distribution function is
 %непонятно чт писать
+We assume that source is a cylinder with radius $R_w$, and height also equals $R_w$, with axis oriendted along line of sight, and with uniform number density $n_w$. 
 and photons distribution corresponds to the radiation of the first star at distance $R_w$.

### 2023-05-09 papers/FBOTcompton/fbotcompton.tex
paper

@@
 
-Recent observations of Fast Blue Optical Transients \cite{Margutti2019,Ho2020,Coppejans2020} show high level of X-ray radiation.
+Recent observations of Fast Blue Optical Transients \cite{Margutti2019, Ho2019cow,Ho2020koala,Coppejans2020, Ho2021at2020, YaoAt2020mrf} show high level of X-ray radiation.
 There are suggested different scenarios of  origins of this radiation: central engine or interaction of shock with external medium.
-In this paper we present model of interaction of shock with strong wind from close star, producing strong X-ray emission via inverse compton scattering on the photons of this star.
+In this paper we present model of interaction of shock with strong wind from close star, producing X-ray emission via inverse compton scattering on the photons of this star.
+This assumption is supported by studying of local environment of object AT2018cow \cite{SunAT2018environment} showing that there are two young massive stur clusters in vicinity of AT2018cow.
+Young massive star clusters (YMSC) have a lot of stars with strong winds and high luminocity in rather small volume (f.
+e.
+cluster Westerlund-1 has 6 yellow giants, 4 red supergiants, 24 Wolf-Rayet stars, and dozens of OB stars in the radius of 1 parsec \cite{Clark2005westerlund, Crowther2006westerlund, Negueruela2010westerlund}).
+So if supernova explosion happens in YMSC, it is probable that ejecta will interact with strong wind of neighbour star, placed at distance about $10^{17}  cm$.
+In this work we use object CSS161010, decribed in \cite{Coppejans2020}, because it has very high ejecta velocity ($0.3 c - 0.5 c$) and it makes Particle-in-Cell (PIC) modeling of particle acceleration, which we use to obtain distribution function of emitting electrons, more simple??.
+This paper is organized as follows: in section \ref{synchrotron} we evaluate synchrotron radio flux from CSS161010 and fit it to observation data, varying such hydrodynamical parameters as magnetic field and number density.
+In section \ref{hydrodynamic} we present hydrodinamic simulation of interaction between supernova ejecta and stellar wind, using parameters obtained on previous step.
+And in section \ref{compton} we evaluate Inverse Compton X-ray flux from the region beyond wind bow-shock.
 
+\section{synchrotron radiation}\label{synchrotron}
+Observations of radio flux from CSS16101 \cite{Coppejans2020} shows that it's spectrum is consistent with synchrotron self-absorption model.
+Method of analytical estimations of source parameters for this case was developed in works \cite{Chevalier1998, ChevalierFransson} and is commonly used to analyze spectrums of FBOTs \cite{Ho2019cow, Ho2020koala, Coppejans2020, Ho2021at2020}.
+But this approach needs an assumption of power-law distribution of electrons.
+In this work we use electrons distribution function obtained from PIC simulations of shock wave and evaluate synchrotron radiation using numerical integration of standard formulae for emissivity and absorption coefficient, described in \cite{Ginzburg1975, Ghisellini}.
+
+Distribution function of emitting electrons is evaluated using PIC simulation with code SMILEI \cite{Derouillat}.
+Setups for shock simulation is decribed in details in our previous work \cite{BykovUniverse}.
+Shock was initialized due to collision of uniform plasma flow with velocity $0.3 c$ with reflecting wall.
+Magnetiztion was $\sigma = B^2/4 \pi m_p \gamma c^2 = 0.002$ and diffirent inclination angles of magnetic field were studied.
+One can see the well-known effect, that efficiency of particle acceleration strongly depends on the magnetic obliquety to the shock \cite{SironiSpitkovsky2009pair, Crumley2019, GuoSironi2014_1, Romansky2018} on the figure \ref{distributions}, where electron distributions in the shock downstream for different inclination angles are shown.
+
+\begin{figure}[h!]
+	\centering
+	\begin{minipage}{0.48\textwidth}
+		\centering
+		\includegraphics[width=0.98\textwidth]{./fig/distributions.png} 
+		\caption{Electron distribution function in the downstream of shock with different inclination angles $\theta$}
+		\label{distributions}
+	\end{minipage}\hfill
+	\begin{minipage}{0.48\textwidth}
+		\centering
+		\includegraphics[width=0.98\textwidth]{./fig/sphericalSource.png} 
+		\caption{Scheme of numerical model of spherical radiation source}
+		\label{sphericalSource}
+	\end{minipage}
+\end{figure}
+
+For evaluating radiation we use our numerical code FAINA, which allows to evaluate synchrotron and inverse compton radiation and also fit model to observational data.
+Radiation source in the code is presented as spherical layer in cylindrical coordinates as schematically shown on figure \ref{sphericalSource}.
+Direction to observer is along z-axis.
+Colour shows the part of volume of cell is filled with emitting plasma.
+Each cell has it's own magnetic field, electron number density and distribution.
+This geometry allows to integrate emissivty along line of sight, taking into account self-absorption.
+
+Analytical analysis \cite{Coppejans2020} and numerical modeling \cite{BykovUniverse} shows that magnetic field in the system should be rather strong (about $0.3 G$ at distances about $10^{17} cm$).
+It does not seems to be possible that quasiuniform interstellar magnetic field can be so strong.
+Another option is to use magnetic field generated by the star, but on this distances it should be perpendicular to the shock velocity \cite{Parker} and particle can not be accelerated in quasiperpendicular shocks, as it was mentioned above.
+So we assumed presence of strong turbulence around the star, containing $90 \%$ of magnetic energy, wtih characteristic scale about radius of expanding ejecta, and Kolmogorov's energy spectum.
+Due to fluctuating magnetic field, shock become quasiparallel at some points, and can be a source of accelerated particles.
+Time of forming the powerlaw tail of electrons distribution in our PIC simulations is much smaller than time of significant changes of shock magnetic obliquety in this large-scale turbulence, and we assume that electron distribution become stationar in every cell of emitting envelope.
+We assume that source of synchrotron radio flux, observed from CSS161010 is spherical shell with uniform density and magnetic field.
+Distributions for different inclination angles of magnetic field relative to local shock velocity are taken in different points of the source.
+With this assumptions we can fit our model to observational data of radio flux from CSS161010 at 98 day after explosion (the first observation when both radio and X-ray flux was detected), and derive such parameters as magnetic field, concentration and width of the spherical shell.
+Our best fit for radio flux and observational data are shown on Figure \ref{}.
+Evaluated parameters are magnetic field $B = 0.6 G$, number density $n = 50000{cm}^{-3}$ and shell width $l = 0.01\cdot R = 1.3E15 cm$.
+This parameters are consistent with obtained from hydrodynamical simulations.
 %написать почему не хватает комптона от всей оболочки
-\section{hydrodynamic}
+\section{hydrodynamic}\label{hydrodynamic}
 In our model we took two close stars, on a distance $1.4\cdot10^{17} cm$, wich correspondes to the radius of expanding shock of CSS161010 at 99 day after explosion.
@@
 
-Simulations showed that ejecta of supernovae, moving in interstellar medium has profile with thin and dense peak, with parameters number density $n \approx 10^5$ and width $d = 1E^{15}$.
+Simulations showed that ejecta of supernovae, moving in interstellar medium has profile with thin and dense peak, with parameters number density $n \approx 10^5$ and width $d = 10^{15}$.
 
-The wind shock has radius $R_w \approx 0.5\times10^14 cm$ and number density $n_w \approx 10^9 cm^{-3}$.
+The wind shock has radius $R_w \approx 0.5\times10^{14} cm$ and number density $n_w \approx 10^9 cm^{-3}$.
 With this parameters, and luminocity and temperature of the star, one can estimate observed X-ray flux, produxed by inverse compton scattering.
-\section{radiation}
-For evaluating radiation we use our numerical code FAINA, which allows to evaluate synchrotron and inverse compton radiation and also fit model to observational data.
+\begin{figure}[h!]
+	\centering
+	\begin{minipage}{0.48\textwidth}
+		\centering
+		\includegraphics[width=0.98\textwidth]{./fig/density.png} 
+		\caption{Density profile of the system at different time moments}
+		\label{density}
+	\end{minipage}\hfill
+	\begin{minipage}{0.48\textwidth}
+		\centering
+		\includegraphics[width=0.98\textwidth]{./fig/Bfield.png} 
+		\caption{Magnetic field profile of the system at different time moments}
+		\label{Bfield}
+	\end{minipage}
+\end{figure}
 
-We assume that source of synchrotron radio flux, observed from CSS161010 is spherical shell with uniform density and magnetic field.
-Distribution function of emitting is evaluated using PIC simulation with code SMILEI \cite{Derouillat}.
-Setups for shock simulation is decribed in details in our previous work \cite{BykovUniverse}.
-Shock was initialized due to collision of uniform plasma flow with velocity $0.3 c$ with reflecting wall.
-Magnetiztion was $\sigma = B^2/4 pi m_p \gamma c^2 = 0.002$ and diffirent inclination angles of magnetic field were studied.
-Distributions for different inclination angles of magnetic field relative to local shock velocity are taken in different points of the source.
-With this assumptions we can fit our model to observational data of radio flux from CSS161010 at 98 day after explosion (the first observation when both radio and X-ray flux was detected), and derive such parameters as magnetic field, concentration and width of the spherical shell.
-Our best fit for radio flux and observational data are shown on Figure \ref{}.
-Evaluated parameters are magnetic field $B = 0.6 G$, number density $n = 50000{cm}^{-3}$ and shell width $l = 0.01\cdot R = 1.3E15 cm$.
-This parameters are consistent with obtained from hydrodynamical simulations.
+\section{Inverse compton radiation}\label{compton}
 

### 2023-05-09 papers/FBOTcompton/fbotcompton.tex
paper

@@
 Due to fluctuating magnetic field, shock become quasiparallel at some points, and can be a source of accelerated particles.
-Time of forming the powerlaw tail of electrons distribution in our PIC simulations is much smaller than time of significant changes of shock magnetic obliquety in this large-scale turbulence, and we assume that electron distribution become stationar in every cell of emitting envelope.
-We assume that source of synchrotron radio flux, observed from CSS161010 is spherical shell with uniform density and magnetic field.
-Distributions for different inclination angles of magnetic field relative to local shock velocity are taken in different points of the source.
-With this assumptions we can fit our model to observational data of radio flux from CSS161010 at 98 day after explosion (the first observation when both radio and X-ray flux was detected), and derive such parameters as magnetic field, concentration and width of the spherical shell.
-Our best fit for radio flux and observational data are shown on Figure \ref{}.
-Evaluated parameters are magnetic field $B = 0.6 G$, number density $n = 50000{cm}^{-3}$ and shell width $l = 0.01\cdot R = 1.3E15 cm$.
-This parameters are consistent with obtained from hydrodynamical simulations.
+Time of forming the powerlaw tail of electrons distribution in our PIC simulations is much smaller than time of significant changes of shock magnetic obliquety in this large-scale turbulence, and we assume that electron distribution becomes stationar in every cell of emitting envelope and corresponds to the inclination angle of magnetic field at this point.
+Number density of electrons is considered uniform inside the radiation source.
+
+With this model of radiation source we fit evaluated synchrotron radiation to observational data of radio flux from CSS161010 at 98 day after explosion (the first observation when both radio and X-ray flux was detected), and derive such parameters as mean magnetic field, concentration and width of the spherical shell.
+Our best fit for radio flux and observational data are shown on Figure \ref{synchrotron}.
+Evaluated parameters are magnetic field $B = 0.6 G$, number density $n = 50000{cm}^{-3}$ and shell width $l = 0.01\cdot R = 1.4E15 cm$.
+This parameters are consistent with results of other authors \cite{Coppejans2020}, only electrons number density is much higher because of not-powerlaw distribution.
+\begin{figure}
+	\centering
+	\includegraphics[width=0.5\textwidth]{./fig/radiation2.png} 
+	\caption{Observed and modeled radiospectrum of CSS161010 at 98 day after explosion.}
+	\label{synchrotron}
+\end{figure}
+
+And then we use obtained parameters for hydrodynamical modeling of interection between sub-relativistic ejecta and strong stellar wind from neighbour star.
 %написать почему не хватает комптона от всей оболочки
 \section{hydrodynamic}\label{hydrodynamic}
-In our model we took two close stars, on a distance $1.4\cdot10^{17} cm$, wich correspondes to the radius of expanding shock of CSS161010 at 99 day after explosion.
-Such small distances are possible in young massive star clusters \cite{}.
-Also massive star clusters usually have dozens of bright stars with powerfull stellar winds (OB, Wolf-Rayet, and red supergiants).
+In our model we took two close stars, on a distance $1.4\cdot10^{17} cm$, wich corresponds to the radius of expanding shock of CSS161010 at 98 day after explosion.
+It is  possible in young massive star clusters where are a lot of bright stars with powerfull stellar winds (OB, Wolf-Rayet, and red supergiants) in the small volume.
 We will refer to them as first star - stable, generating wind, and the second - exploded as supernova.
 We assumed both stars have winds with velocity $v_w = ?$ and mass loss $\dot{M} = 10^{-5} M_\odot $ per year and luminocity $L=500000 L_\odot$ and surfece temperature $T = 2.5\cdot10^6 K$ which is typical for Wolf-Rayet stars.
-The magnetic fied of the wind is initialized using \cite{}, and for the first star  it's amplitude is $100 G$ at the radius $10$ while for the second we need much stronger field to satisfy synchrotron observational data, as it is shown in papers \cite{Coppejans2020, BykovUniverse}.
+The magnetic fied of the wind is initialized using \cite{}, and for the first star  it's amplitude is $100 G$ at the radius $10$ while for the second we need much stronger field to satisfy synchrotron observational data.
 So we took magnetic field $1000 G$ at the radius $10^{12}$, stars with such strong fields are described in \cite{}.
@@
 At first we performed 3d modeling of two stable stars with parameters described above until they winds form a stable configuration.
-Sizes of the simulation box are $Lx = $ with number of grid points respectively.
+First star is placed at point with coordinates $x_1 = 0, y_1 = 0, z_1 = 0$, and the second at point with $x_2 = 0, y_2 = 0, z_2 = 1.4\cdot10^{17} cm$.
+Plane x-y is equatorial plane of winds of both stars.
+Sizes of the simulation box are $Lx = 6\cdot 10^{17} cm$ and $Ly = Lz = 3\cdot 10^{17}$ with number of grid points 400, 200 and 200 respectively.
 The next step was 1d spherical modeling of second star explosion.
-We use ejecta mass $M_{ej} = 0.02 M_\odot$ and velocity $v_{ej} = 0.3 c$ and initial conditions as described in \cite{}.
-At the third step we use results of the previous steps as initial and boundary conditions in setup with smaller scale - with size $Lx = $ .
-And at the last step, in setup with the smallest scale $Lx = $, wich is necessary to resolve bow- and tremination shock of the wind of the first star, we use results of the third step as initial and boundary conditions.
-Scheme of setups is shown in Figure \ref{}.
-
-Simulations showed that ejecta of supernovae, moving in interstellar medium has profile with thin and dense peak, with parameters number density $n \approx 10^5$ and width $d = 10^{15}$.
-
-The wind shock has radius $R_w \approx 0.5\times10^{14} cm$ and number density $n_w \approx 10^9 cm^{-3}$.
-With this parameters, and luminocity and temperature of the star, one can estimate observed X-ray flux, produxed by inverse compton scattering.
-\begin{figure}[h!]
+We use ejecta mass $M_{ej} = 0.1 M_\odot$ and velocity $v_{ej} = 0.3 c$ and initial conditions as described in \cite{}.
+At the third step we simulated setup with smaller scale - with size $Lx = 1.05\cdot 10^{17} cm$ and $Ly = Lz = 0.9\cdot 10^{16}$ and combined results of the previous steps as initial conditions.
+Also results of 1d modeling were used as time-dependent boundary condition on the right boundary.
+And at the last step, in setup with the smallest scale $Lx = 0.3\cdot 10^{17} cm$ and $Ly = Lz = 1.5\cdot 10^{15}$, wich is necessary to resolve bow- and tremination shock of the wind of the first star, we use results of the third step as initial and boundary conditions.
+Number density in different setups and their ierarchy scheme is shown on figure \ref{density}.
+On the panel a one can see iteraction of two winds in large-scale simulation at time before the explosion.
+On the panel b propagation of ejecta in middle-scale simulation at time $t_1 \approx 1.0\cdot10^7 s$ is shown.
+On the panel c one can see it in small-scale simulation at time $t_2 \approx 1.1\cdot10^7 s$.
+Ans on the panel d there is zoomed picture of wind bowshock in small-scale simulation at time $t_3 \approx 1.4\cdot10^7 s$ Magnetic field at the same time moments is shown on figure \ref{Bfield}.
+\begin{figure}
 	\centering
@@
 \end{figure}
+
+Simulations showed that ejecta of supernovae, moving in interstellar medium has profile with thin and dense peak, with parameters number density $n \approx 10^5$ and width $d = 10^{15}$.
+
+The wind shock has radius $R_w \approx 0.5\times10^{14} cm$ and number density $n_w \approx 10^9 cm^{-3}$.
+With this parameters, and luminocity and temperature of the star, one can estimate observed X-ray flux, produxed by inverse compton scattering.
 

### 2023-05-10 papers/FBOTcompton/fbotcompton.tex
paper

@@
 The next step was 1d spherical modeling of second star explosion.
-We use ejecta mass $M_{ej} = 0.1 M_\odot$ and velocity $v_{ej} = 0.3 c$ and initial conditions as described in \cite{}.
+We use ejecta mass $M_{ej} = 0.1 M_\odot$ and velocity $v_{ej} = 0.3 c$ and initial conditions as described in \cite{Chevalier1982, Petruk2021snr}.
+The simulation box is splitted into three regions - in the central region with radius $r_0 = ?$ density is constant and velocity is proportional to radius $v(r) = 0.3 c r/r_0$.
+In the next region between $r_0$ and $r_1 = 2 r_0$ number density depends on radius as $n(r) \propto r^{-9}$ and velocity is also proportional to radius.
+And in the last region with $r > r_1$ stellar wind with constant velocity $v_w$ and mass loss $\dot{M}$is initialized.
+Profiles of ejecta density at different time moments are shown on figure \ref{snr}.
+\begin{figure}
+	\centering
+	\includegraphics[width=0.5\textwidth]{./fig/snr.png} 
+	\caption{Density profile of supernova explosion.}
+	\label{snr}
+\end{figure}
+
 At the third step we simulated setup with smaller scale - with size $Lx = 1.05\cdot 10^{17} cm$ and $Ly = Lz = 0.9\cdot 10^{16}$ and combined results of the previous steps as initial conditions.
 Also results of 1d modeling were used as time-dependent boundary condition on the right boundary.
-And at the last step, in setup with the smallest scale $Lx = 0.3\cdot 10^{17} cm$ and $Ly = Lz = 1.5\cdot 10^{15}$, wich is necessary to resolve bow- and tremination shock of the wind of the first star, we use results of the third step as initial and boundary conditions.
+And at the last step, in setup with the smallest scale $Lx = 0.3\cdot 10^{17} cm$ and $Ly = Lz = 1.5\cdot 10^{15}$, which is necessary to resolve bow- and tremination shock of the wind of the first star, we use results of the third step as initial and boundary conditions.
 Number density in different setups and their ierarchy scheme is shown on figure \ref{density}.
@@
 
-Simulations showed that ejecta of supernovae, moving in interstellar medium has profile with thin and dense peak, with parameters number density $n \approx 10^5$ and width $d = 10^{15}$.
+Simulations showed that ejecta of supernovae, moving in interstellar medium has profile with thin and dense peak with number density $n \approx 10^5 cm^{-3}$ and width $d = 10^{16}$, and wide region with number density $\approx 10^4 cm^{-3}$ behind the peak.
 
-The wind shock has radius $R_w \approx 0.5\times10^{14} cm$ and number density $n_w \approx 10^9 cm^{-3}$.
-With this parameters, and luminocity and temperature of the star, one can estimate observed X-ray flux, produxed by inverse compton scattering.
+The wind bow-shock has radius $R_w \approx 2\times10^{14} cm$ and number density $n_w \approx 10^9 cm^{-3}$.
+With this parameters, and luminocity and temperature of the star, one can estimate observed X-ray flux, produced by inverse compton scattering.
 

### 2023-05-11 papers/FBOTcompton/fbotcompton.tex
paper

@@
 Each cell has it's own magnetic field, electron number density and distribution.
-This geometry allows to integrate emissivty along line of sight, taking into account self-absorption.
+This geometry allows to integrate emissivity along line of sight, taking into account self-absorption.
 
@@
 The next step was 1d spherical modeling of second star explosion.
-We use ejecta mass $M_{ej} = 0.1 M_\odot$ and velocity $v_{ej} = 0.3 c$ and initial conditions as described in \cite{Chevalier1982, Petruk2021snr}.
+We use ejecta mass $M_{ej} = 0.1 M_\odot$ and velocity $v_{ej} = 0.3 c$ and initial conditions as described in \cite{ChevalierLiang1989, Petruk2021snr}.
 The simulation box is splitted into three regions - in the central region with radius $r_0 = ?$ density is constant and velocity is proportional to radius $v(r) = 0.3 c r/r_0$.
@@
 
-At the third step we simulated setup with smaller scale - with size $Lx = 1.05\cdot 10^{17} cm$ and $Ly = Lz = 0.9\cdot 10^{16}$ and combined results of the previous steps as initial conditions.
+At the third step we simulated setup with smaller scale - with size $Lx = 1.05\cdot 10^{17} cm$ and $Ly = Lz = 0.9\cdot 10^{16} cm$ and combined results of the previous steps as initial conditions.
 Also results of 1d modeling were used as time-dependent boundary condition on the right boundary.
-And at the last step, in setup with the smallest scale $Lx = 0.3\cdot 10^{17} cm$ and $Ly = Lz = 1.5\cdot 10^{15}$, which is necessary to resolve bow- and tremination shock of the wind of the first star, we use results of the third step as initial and boundary conditions.
+And at the last step, in setup with the smallest scale $Lx = 0.3\cdot 10^{17} cm$ and $Ly = Lz = 1.5\cdot 10^{15} cm$, which is necessary to resolve bow- and tremination shock of the wind of the first star, we use results of the third step as initial and boundary conditions.
 Number density in different setups and their ierarchy scheme is shown on figure \ref{density}.
@@
 
-Simulations showed that ejecta of supernovae, moving in interstellar medium has profile with thin and dense peak with number density $n \approx 10^5 cm^{-3}$ and width $d = 10^{16}$, and wide region with number density $\approx 10^4 cm^{-3}$ behind the peak.
+Simulations showed that ejecta of supernovae, moving in interstellar medium has profile with thin and dense peak with number density $n \approx 10^5 cm^{-3}$ and width $d = 10^{16} cm$, and wide region with number density $\approx 10^4 cm^{-3}$ behind the peak.
 
@@
 \section{Inverse compton radiation}\label{compton}
+
+Let consider different models of origin of X-ray radiation, which observed flux in energy range $0.3 - 10\text{ }keV$ is $(1.33 \pm 0.76) \times 10^{-15}\text{ }erg\text{ }s^{-1} cm^{-2}$..
+At first, we assume that it is synchrotron radiation from the electron distribution, which is prolonged to much higher energies than it is possible to reach in PIC simulation.
+The necessary energy of electrons is determined by energy of emitted photon $E_e = (4\pi{m_e}^3 c^5 E_{ph}/0.29\cdot3 e B h)^{1/2} \approx 2\cdot10^6 m_e c^2$.
+The flux from source with parameters obtained in section \ref{synchrotron} and prolonged distribution is $what$.
+But on should take into account tht lifeime of electrons with this energy in magnetic field $B \approx 0.6 G$ is $\tau = {m_e}^4 c^7 / e^4 B^2 E_e \approx 10^3 s$ and distance that electron can move on this time is about $d = 3\cdot10^13 cm$.
+This means that emitting volume drasticaly decreases and we should evaluate radiation of spherical layer with width $d$.
 

### 2023-05-11 papers/FBOTcompton/fbotcompton.tex
paper

@@
 
-Let consider different models of origin of X-ray radiation, which observed flux in energy range $0.3 - 10\text{ }keV$ is $(1.33 \pm 0.76) \times 10^{-15}\text{ }erg\text{ }s^{-1} cm^{-2}$..
+Let consider different models of origin of X-ray radiation, which observed flux in energy range $0.3 - 10\text{ }keV$ is $(1.33 \pm 0.76) \times 10^{-15}\text{ }erg\text{ }s^{-1} cm^{-2}$.
+
 At first, we assume that it is synchrotron radiation from the electron distribution, which is prolonged to much higher energies than it is possible to reach in PIC simulation.
@@
 But on should take into account tht lifeime of electrons with this energy in magnetic field $B \approx 0.6 G$ is $\tau = {m_e}^4 c^7 / e^4 B^2 E_e \approx 10^3 s$ and distance that electron can move on this time is about $d = 3\cdot10^13 cm$.
-This means that emitting volume drasticaly decreases and we should evaluate radiation of spherical layer with width $d$.
+This means that emitting volume drasticaly decreases and we should evaluate radiation of spherical layer with width $d$, and the resulting synchrotron X-ray flux is $what$.
+
+Another option of X-ray source is inverse compton scattering of electrons in the ejecta on photons from the star.
 

### 2023-06-25 papers/FBOTcompton/fbotcompton.tex
electrons energy

@@
 	\author{A M Bykov$^{1}$, V I Romansky$^{1}$}
+
+\begin{equation}
+R_{ik}-\frac{1}{2} g_{ik}R=\frac{8\pi G}{c^4} T_{ik}
+\end{equation}
+\begin{equation}
+A(\gamma_{\alpha\gamma}\gamma_{\beta\delta} - \gamma_{\alpha\delta}\gamma_{\beta\gamma})
+\end{equation}
+\begin{equation}
+P^\gamma_{\alpha\gamma\beta}=2A\gamma_{\alpha\beta}
+\end{equation}
+\begin{equation}
+dl^2 = dx_1^2+dx_2^2+dx_3^2+\frac{(x_1 dx_1 + x_2 dx_2 +x_3 dx_3)^2}{a^2-x_1^2-x_2^2-x_3^2}
+\end{equation}
+\begin{equation}
+\gamma_{\alpha\beta}=\delta_{\alpha\beta}+x_\alpha x_\beta/a^2
+\end{equation}
+\begin{equation}
+dl^2=dr^2/(1-r^2/a^2)+r^2(sin^2 \theta d\phi^2+d\theta^2)
+\end{equation}
+\begin{equation}
+dl^2=a^2(d\chi^2 + sin^2\chi(sin^2 \theta d\phi^2+d\theta^2))
+\end{equation}
+\begin{equation}
+ds^2 = c^2 dt^2-dl^2 = c^2 dt^2 – a(t)^2(d\chi^2 + sin^2\chi(sin^2 \theta d\phi^2+d\theta^2))
+\end{equation}
+\begin{equation}
+ds^2 = c^2 dt^2-dl^2 = a(\eta)^2( d\eta^2 - d\chi^2 - sin^2\chi(sin^2 \theta d\phi^2+d\theta^2))
+\end{equation}
+\begin{equation}
+R^\chi_\chi = R^\theta_\theta = R^\phi_\phi= -(2 a^2 + (da/d\eta)^2 + a d^2a/d\eta^2)/a^4
+\end{equation}
+\begin{equation}
+R^\eta_\eta=3*((da/d\eta)^2-a da/d\eta)/a^4
+\end{equation}
+\begin{equation}
+R=-6(a+d^2a/d\eta^2)/a^3
+\end{equation}
+\begin{equation}
+3(a^2+d^2a/d\eta^2)/a^4 = 8 \pi G \epsilon/c^4
+\end{equation}
+\begin{equation}
+d\epsilon=-(\epsilon + p)dV/V
+\end{equation}
+\begin{equation}
+3 ln~a = -\int d\epsilon/(\epsilon + p)
+\end{equation}
+\begin{equation}
+a=a_0(1 – cos \eta)
+\end{equation}
+\begin{equation}
+t=\frac{a_0}{c} (\eta – sin \eta)
+\end{equation}
+\begin{equation}
+a_0 = 2GM/3\pi^2
+\end{equation}
+\begin{equation}
+R_{ik}-1/2 g_{ik}R=8\pi G/c^4 T_{ik}-\Lambda g_{ik}
+\end{equation}
 

### 2023-09-27 papers/FBOTcompton/fbotcompton.tex
fz

--- 2023-05-11 papers/FBOTcompton/fbotcompton.tex
@@
-%\documentclass[a4paper,12pt]{report}
-\documentclass[a4paper,12pt]{extreport}
-
-\bibliographystyle{iopart-num}
+\documentclass{article}
+\usepackage{graphicx} % Required for inserting images
 \usepackage{amsmath}
-%\usepackage{citesort}
-\usepackage{subfigure}
-\usepackage{graphicx}
-\graphicspath{{fig/}}
-\usepackage{ifpdf}
-\ifpdf\usepackage{epstopdf}\fi
-\usepackage[export]{adjustbox}
-\usepackage[numbers]{natbib}
 
@@
 
+\title{fbotCompton}
+
 \begin{document}
-	\title{On X-ray emission from Fast Blue Optical Transients}
 
-	\author{A M Bykov$^{1}$, V I Romansky$^{1}$}
-
-Recent observations of Fast Blue Optical Transients \cite{Margutti2019, Ho2019cow,Ho2020koala,Coppejans2020, Ho2021at2020, YaoAt2020mrf} show high level of X-ray radiation.
+	\section{Introduction}
+	Fast Blue Optical Tranients are highly luminous objects with very fast rise and decay 
+	phases discovered in optical  
+	Recent observations of Fast Blue Optical Transients 
+	%koala has not X-ray
+	\cite{Margutti2019, Ho2019cow,Ho2020koala,Coppejans2020, Ho2021at2020, YaoAt2020mrf, MatthewsAT2022tsd, CHrimesAT2023fhn},  show high level of X-ray radiation.
 There are suggested different scenarios of  origins of this radiation: central engine or interaction of shock with external medium.
-In this paper we present model of interaction of shock with strong wind from close star, producing X-ray emission via inverse compton scattering on the photons of this star.
-This assumption is supported by studying of local environment of object AT2018cow \cite{SunAT2018environment} showing that there are two young massive stur clusters in vicinity of AT2018cow.
-Young massive star clusters (YMSC) have a lot of stars with strong winds and high luminocity in rather small volume (f.
+In this paper we present model of interaction of shock with strong wind from close star, producing X-ray emission via inverse Compton scattering on the photons of this star.
+This assumption is supported by studying of local environment of object AT2018cow \cite{SunAT2018environment} showing that there are two young massive star clusters in vicinity of AT2018cow.
+Young massive star clusters (YMSC) have a lot of stars with strong winds and high luminosity in rather small volume (f.
 e.
 cluster Westerlund-1 has 6 yellow giants, 4 red supergiants, 24 Wolf-Rayet stars, and dozens of OB stars in the radius of 1 parsec \cite{Clark2005westerlund, Crowther2006westerlund, Negueruela2010westerlund}).
-So if supernova explosion happens in YMSC, it is probable that ejecta will interact with strong wind of neighbour star, placed at distance about $10^{17}  cm$.
-In this work we use object CSS161010, decribed in \cite{Coppejans2020}, because it has very high ejecta velocity ($0.3 c - 0.5 c$) and it makes Particle-in-Cell (PIC) modeling of particle acceleration, which we use to obtain distribution function of emitting electrons, more simple??.
-This paper is organized as follows: in section \ref{synchrotron} we evaluate synchrotron radio flux from CSS161010 and fit it to observation data, varying such hydrodynamical parameters as magnetic field and number density.
+So if supernova explosion happens in YMSC, it is probable that ejecta will interact with strong wind of neighbour star, placed at distance about $10^{17}~cm$.
+In this work we use object CSS161010, described in \cite{Coppejans2020}, because it has very high ejecta velocity ($0.3 c - 0.5 c$) and it makes Particle-in-Cell (PIC) modeling of particle acceleration, which we use to obtain distribution function of emitting electrons, more simple??.
+This paper is organized as follows: in section \ref{synchrotronChapter} we evaluate synchrotron radio flux from CSS161010 and fit it to observation data, varying such hydrodynamical parameters as magnetic field and number density.
 In section \ref{hydrodynamic} we present hydrodinamic simulation of interaction between supernova ejecta and stellar wind, using parameters obtained on previous step.
@@
 
-\section{synchrotron radiation}\label{synchrotron}
+	\section{The problem of X-ray}
+	While radio emission from CSS161010 can be explained with model of spherical ejecta with self-absorption \cite{Coppejans2020}, as is described in section \ref{synchrotronChapter}, the bright X-ray radiation detected on the 99 day after explosion (flux $F_{X} = (1.33 \pm 0.76)\times10^{-15} erg~s^{-1}~cm^{-2}$ and corresponding isotropic luminosity $L_{X}=(3.4\pm1.9)\times10^{39} erg~s^{-1}$ in range 0.3-10 keV)is not easy to explain.
+	At first one can evaluate X-ray part of synchrotron radiation, using parameters provided by authors of \cite{Coppejans2020}.
+Modeled synchrotron specturm and observational radio and X-ray data (assuming photon index 1.25) are shown on Figure \ref{synchrotronX}.
+	\begin{figure}
+		\centering
+		\includegraphics[width=0.7\textwidth]{./fig/what.png} 
+		\caption{Modeled synchrotron spectrum of CSS161010 and observational radio and X-ray data.}
+		\label{synchrotronX}
+	\end{figure}
+	Also one can assume that electrons spectrum becomes harder on higher energies, it follows from Monte-Carlo simulations \cite{BykovRomansky}.
+Using obtained in \cite{Coppejans2020} magnetic field $B \approx 0.6~G$ we can calculate the necessary energy of electrons, determined by energy of emitted photon $E_e = (4\pi{m_e}^3 c^5 E_{ph}/0.29\cdot3 e B h)^{1/2} \approx 2\cdot10^6 m_e c^2$.
+If we assume, that electrons spectrum has index 2 on energies higher than 100 MeV, synchrotron X-ray spectrum would increase.
+But one should take into account the lifetime of electrons with this energy is $\tau = {m_e}^4 c^7 / e^4 B^2 E_e \approx 10^3~s$ and distance that electron can move on this time is about $d = 3\cdot10^{13}~cm$.
+This means that emitting volume drastically decreases and we should evaluate radiation of spherical layer with width $d$, and the resulting synchrotron X-ray flux is $what$.
+
+	Another option of X-ray radiation is Inverse Compton scattering.
+The possible seed photons for it may origin from ambient Cosmic Microwave Background and mean galactic field or from the source itself.
+The ratio of luminocities of synchrotorn and inverse compton radiation of same electrons equals to the ratio of magnetic energy density and photons energy density $L_{IC}/L_{synch}=u_{ph}/u_{B}$.
+Ratio of radio and X-ray luminocities at 99 day after explosion is $L_{X}/L_{radio} \approx 10$ and the magnetic energy density in CSS161010 is $u_{B}=B^2/8\pi \approx 0.015 erg cm^{-3}$.
+So to explain X-ray flux as inverse Compton radiation of the same electrons that produce synchrotron radio flux, we need photons energy density about $0.15 erg cm^{-3}$.
+It is much more than energy density of CMB $4\cdot10^{-13}~erg~cm^{-3}$ and mean galactic photon field $10^{-12}~erg~cm^{-3}$ \cite{Mathis1983} and more than the most luminous stars $L \approx 10^{40}~erg~s^{-1}$ can provide on such distances $L/4\pi R^2 c \approx 10^{-6}~erg~s^{-1}$.
+
+	Another distribution of electrons on neighbour star
+
+	Estimation of energy
+
+	Estimation of synchrotron flux.
+
+	\section{Synchrotron radiation}\label{synchrotronChapter}
 	Observations of radio flux from CSS16101 \cite{Coppejans2020} shows that it's spectrum is consistent with synchrotron self-absorption model.
@@
 	Distribution function of emitting electrons is evaluated using PIC simulation with code SMILEI \cite{Derouillat}.
-Setups for shock simulation is decribed in details in our previous work \cite{BykovUniverse}.
-Shock was initialized due to collision of uniform plasma flow with velocity $0.3 c$ with reflecting wall.
-Magnetiztion was $\sigma = B^2/4 \pi m_p \gamma c^2 = 0.002$ and diffirent inclination angles of magnetic field were studied.
-One can see the well-known effect, that efficiency of particle acceleration strongly depends on the magnetic obliquety to the shock \cite{SironiSpitkovsky2009pair, Crumley2019, GuoSironi2014_1, Romansky2018} on the figure \ref{distributions}, where electron distributions in the shock downstream for different inclination angles are shown.
+Setups for shock simulation is described in details in our previous work \cite{BykovUniverse}.
+Shock was initialized due to collision of uniform plasma flow with velocity $0.3\,c$ with reflecting wall.
+Magnetization was $\sigma = B^2/4 \pi m_p \gamma c^2 = 0.0002$ and different inclination angles of magnetic field were studied.
+One can see the well-known effect, that efficiency of particle acceleration strongly depends on the magnetic obliquity to the shock \cite{SironiSpitkovsky2009pair, Crumley2019, GuoSironi2014_1, Romansky2018} on the figure \ref{distributions}, where electron distributions in the shock downstream for different inclination angles are shown.
 
-\begin{figure}[h!]
+	\begin{figure}
 		\centering
@@
 			\centering
-		\includegraphics[width=0.98\textwidth]{./fig/distributions.png} 
+			\includegraphics[width=0.98\textwidth]{./fig/electrons.png} 
 			\caption{Electron distribution function in the downstream of shock with different inclination angles $\theta$}
@@
 
-Analytical analysis \cite{Coppejans2020} and numerical modeling \cite{BykovUniverse} shows that magnetic field in the system should be rather strong (about $0.3 G$ at distances about $10^{17} cm$).
-It does not seems to be possible that quasiuniform interstellar magnetic field can be so strong.
-Another option is to use magnetic field generated by the star, but on this distances it should be perpendicular to the shock velocity \cite{Parker} and particle can not be accelerated in quasiperpendicular shocks, as it was mentioned above.
-So we assumed presence of strong turbulence around the star, containing $90 \%$ of magnetic energy, wtih characteristic scale about radius of expanding ejecta, and Kolmogorov's energy spectum.
-Due to fluctuating magnetic field, shock become quasiparallel at some points, and can be a source of accelerated particles.
-Time of forming the powerlaw tail of electrons distribution in our PIC simulations is much smaller than time of significant changes of shock magnetic obliquety in this large-scale turbulence, and we assume that electron distribution becomes stationar in every cell of emitting envelope and corresponds to the inclination angle of magnetic field at this point.
+	Analytical analysis \cite{Coppejans2020} and numerical modeling \cite{BykovUniverse} shows that magnetic field in the system should be rather strong (about $0.3~G$ at distances about $10^{17}~cm$).
+It does not seems to be possible that quasi-uniform interstellar magnetic field can be so strong.
+Another option is to use magnetic field generated by the star, but on this distances it should be perpendicular to the shock velocity \cite{Parker} and particle can not be accelerated in quasi-perpendicular shocks, as it was mentioned above.
+The presence of strong turbulence can increase efficiency of acceleration in quasi-perpendicular shocks, as shown with PIC simulations in works \cite{Demidem2023inhomogenousshock, Bresci2023turbulentchock, Romansky2019turbulence}.
+But here we use more simple model: PIC simulations are performed with constant initial field, but in evaluation of synchrotron radiation we assume strong largescale turbulence around the star, containing $90 \%$ of magnetic energy, with characteristic scale about radius of expanding ejecta, and Kolmogorov's energy spectrum.
+Due to fluctuating magnetic field, shock becomes quasi-parallel at some points, and can be a source of accelerated particles.
+Time of forming the powerlaw tail of electrons distribution in our PIC simulations is much smaller than time of significant changes of magnetic obliquity due the shock movement in this large-scale turbulence, and we consider that electron distribution becomes stationary in every cell of emitting envelope and corresponds to the inclination angle of magnetic field at this point.
 Number density of electrons is considered uniform inside the radiation source.
 
-With this model of radiation source we fit evaluated synchrotron radiation to observational data of radio flux from CSS161010 at 98 day after explosion (the first observation when both radio and X-ray flux was detected), and derive such parameters as mean magnetic field, concentration and width of the spherical shell.
-Our best fit for radio flux and observational data are shown on Figure \ref{synchrotron}.
-Evaluated parameters are magnetic field $B = 0.6 G$, number density $n = 50000{cm}^{-3}$ and shell width $l = 0.01\cdot R = 1.4E15 cm$.
-This parameters are consistent with results of other authors \cite{Coppejans2020}, only electrons number density is much higher because of not-powerlaw distribution.
+	We consider radio flux from CSS161010 at 98 day after explosion (the first observation when both radio and X-ray flux was detected).
+For defining parameters of the source: it's radius $R$, average magnetization $\sigma$, number density $n$, and width fraction of spherical layer $f$, we fitted modeled radiation to observational data and optimized residual, defined as $r(R, \sigma, n, f) = \sum (F(\nu_i, R, \sigma, n, f) - F_{obs}(\nu_i))^2/{\Delta_i}^2$, where summation goes through all observed data points, $\nu_i$ - i-th frequency, $F_obs$ - observed flux, $\Delta$ - uncertainty of observational data.
+Picture of this residual for fixed parameters $\sigma$ and $n$, and fitted $R$ and $f$ is shown on figure \ref{error}.
+One can see, that there is long valley of small values of residual and due to systematical uncertanties of our model we cannot peak one point with exact parameters for best fit.
+Also the global minimum of residual is located in the region of very high magnetization and low density, while shocks can exist only if $\sigma < 1$ \cite{?} and as shown in \cite{Sironi2011magnetization, Lemoine2014} electron acceleration is depressed for magnetization greater than $10^{-2}$.
+So for further modeling we fix value of $\sigma$ to value $0.0002$, which was used in PIC simulations.
+
+	With this assumption, obtained parameters of the source are: radius $R = 1.4\cdot10^{17}$, magnetic field $B = 0.09~G$, number density $n = 2000~{cm}^{-3}$ and shell width $l = 0.12\cdot R = 1.7\cdot10^{15}~cm$.
+The parameters are broadly consistent with the results of other authors \cite{Coppejans2020}, apart from the electrons number density which is much higher because of quasi-thermal distribution used.
+Modeled radio flux and observational data are shown in Figure \ref{synchrotron}.
+One can see, that peak of the spectral density is provided by thermal electrons, as predicted by \cite{Margalit2021}.
+
 	\begin{figure}
 		\centering
-	\includegraphics[width=0.5\textwidth]{./fig/radiation2.png} 
+		\begin{minipage}{0.48\textwidth}
+			\centering
+			\includegraphics[width=0.98\textwidth]{./fig/error.png} 
+			\caption{Profile of residual in plane $\sigma - n$ with other parameters optimized}
+			\label{error}
+		\end{minipage}\hfill
+		\begin{minipage}{0.48\textwidth}
+			\centering
+			\includegraphics[width=0.98\textwidth]{./fig/radiation2.png} 
 			\caption{Observed and modeled radiospectrum of CSS161010 at 98 day after explosion.}
 			\label{synchrotron}
+		\end{minipage}
 	\end{figure}
 
-And then we use obtained parameters for hydrodynamical modeling of interection between sub-relativistic ejecta and strong stellar wind from neighbour star.
-%написать почему не хватает комптона от всей оболочки
-\section{hydrodynamic}\label{hydrodynamic}
-In our model we took two close stars, on a distance $1.4\cdot10^{17} cm$, wich corresponds to the radius of expanding shock of CSS161010 at 98 day after explosion.
-It is  possible in young massive star clusters where are a lot of bright stars with powerfull stellar winds (OB, Wolf-Rayet, and red supergiants) in the small volume.
+	And then we use obtained parameters for hydrodinamical modeling of interaction between sub-relativistic ejecta and strong stellar wind from neighbour star.
+
+	\section{Hydrodynamic}\label{hydrodynamic}
+	In our model we took two close stars, on a distance $1.4\cdot10^{17}~cm$, which corresponds to the radius of expanding shock of CSS161010 at 98 day after explosion.
+It is  possible in young massive star clusters where are a lot of bright stars with powerful stellar winds (OB, Wolf-Rayet, and red supergiants) in the small volume.
 We will refer to them as first star - stable, generating wind, and the second - exploded as supernova.
-We assumed both stars have winds with velocity $v_w = ?$ and mass loss $\dot{M} = 10^{-5} M_\odot $ per year and luminocity $L=500000 L_\odot$ and surfece temperature $T = 2.5\cdot10^6 K$ which is typical for Wolf-Rayet stars.
-The magnetic fied of the wind is initialized using \cite{}, and for the first star  it's amplitude is $100 G$ at the radius $10$ while for the second we need much stronger field to satisfy synchrotron observational data.
-So we took magnetic field $1000 G$ at the radius $10^{12}$, stars with such strong fields are described in \cite{}.
+We assumed both stars have winds with velocity $v_w = ?$ and mass loss $\dot{M} = 10^{-5} M_\odot $ per year and luminosity $L=500000 L_\odot$ and surface temperature $T = 2.5\cdot10^6~K$ which is typical for Wolf-Rayet stars.
+The magnetic field of the wind is initialized using \cite{Bjorkman1993}, and for the first star  it's amplitude is $1~G$ at the radius $10~12 cm$ while for the second we need much stronger field to satisfy synchrotron observational data.
+So we took magnetic field $1000~G$ at the radius $10^{12}~cm$, stars with such strong fields are described in \cite{}.
 
@@
 At first we performed 3d modeling of two stable stars with parameters described above until they winds form a stable configuration.
-First star is placed at point with coordinates $x_1 = 0, y_1 = 0, z_1 = 0$, and the second at point with $x_2 = 0, y_2 = 0, z_2 = 1.4\cdot10^{17} cm$.
+First star is placed at point with coordinates $x_1 = 0, y_1 = 0, z_1 = 0$, and the second at point with $x_2 = 0, y_2 = 0, z_2 = 1.4\cdot10^{17}~cm$.
 Plane x-y is equatorial plane of winds of both stars.
-Sizes of the simulation box are $Lx = 6\cdot 10^{17} cm$ and $Ly = Lz = 3\cdot 10^{17}$ with number of grid points 400, 200 and 200 respectively.
+Sizes of the simulation box are $Lx = 6\cdot 10^{17}~cm$ and $Ly = Lz = 3\cdot 10^{17}~cm$ with number of grid points 400, 200 and 200 respectively.
 The next step was 1d spherical modeling of second star explosion.
-We use ejecta mass $M_{ej} = 0.1 M_\odot$ and velocity $v_{ej} = 0.3 c$ and initial conditions as described in \cite{ChevalierLiang1989, Petruk2021snr}.
-The simulation box is splitted into three regions - in the central region with radius $r_0 = ?$ density is constant and velocity is proportional to radius $v(r) = 0.3 c r/r_0$.
+We use ejecta mass $M_{ej} = 0.1 M_\odot$ and velocity $v_{ej} = 0.3\,c$ and initial conditions as described in \cite{ChevalierLiang1989, Petruk2021snr}.
+The simulation box is split into three regions - in the central region with radius $r_0 = ?$ density is constant and velocity is proportional to radius $v(r) = 0.3 c r/r_0$.
 In the next region between $r_0$ and $r_1 = 2 r_0$ number density depends on radius as $n(r) \propto r^{-9}$ and velocity is also proportional to radius.
-And in the last region with $r > r_1$ stellar wind with constant velocity $v_w$ and mass loss $\dot{M}$is initialized.
+And in the last region with $r > r_1$ stellar wind with constant velocity $v_w$ and mass loss $\dot{M}$ is initialized.
 Profiles of ejecta density at different time moments are shown on figure \ref{snr}.
@@
 		\centering
-	\includegraphics[width=0.5\textwidth]{./fig/snr.png} 
+		\includegraphics[width=0.7\textwidth]{./fig/snr.png} 
 		\caption{Density profile of supernova explosion.}
@@
 
-At the third step we simulated setup with smaller scale - with size $Lx = 1.05\cdot 10^{17} cm$ and $Ly = Lz = 0.9\cdot 10^{16} cm$ and combined results of the previous steps as initial conditions.
+	At the third step we simulated setup with smaller scale - with size $Lx = 1.05\cdot 10^{17}~cm$ and $Ly = Lz = 0.9\cdot 10^{16}~cm$ and combined results of the previous steps as initial conditions.
 Also results of 1d modeling were used as time-dependent boundary condition on the right boundary.
-And at the last step, in setup with the smallest scale $Lx = 0.3\cdot 10^{17} cm$ and $Ly = Lz = 1.5\cdot 10^{15} cm$, which is necessary to resolve bow- and tremination shock of the wind of the first star, we use results of the third step as initial and boundary conditions.
-Number density in different setups and their ierarchy scheme is shown on figure \ref{density}.
-On the panel a one can see iteraction of two winds in large-scale simulation at time before the explosion.
-On the panel b propagation of ejecta in middle-scale simulation at time $t_1 \approx 1.0\cdot10^7 s$ is shown.
-On the panel c one can see it in small-scale simulation at time $t_2 \approx 1.1\cdot10^7 s$.
-Ans on the panel d there is zoomed picture of wind bowshock in small-scale simulation at time $t_3 \approx 1.4\cdot10^7 s$ Magnetic field at the same time moments is shown on figure \ref{Bfield}.
+And at the last step, in setup with the smallest scale $Lx = 0.3\cdot 10^{17}~cm$ and $Ly = Lz = 1.5\cdot 10^{15}~cm$, which is necessary to resolve bow- and termination shock of the wind of the first star, we use results of the third step as initial and boundary conditions.
+Number density in different setups and their hierarchy scheme is shown on figure \ref{density}.
+On the panel a one can see interaction of two winds in large-scale simulation at time before the explosion.
+On the panel b propagation of ejecta in middle-scale simulation at time $t_1 \approx 1.0\cdot10^7~s$ is shown.
+On the panel c one can see it in small-scale simulation at time $t_2 \approx 1.1\cdot10^7~s$.
+Ans on the panel d there is zoomed picture of wind bowshock in small-scale simulation at time $t_3 \approx 1.4\cdot10^7~s$ Magnetic field at the same time moments is shown on figure \ref{Bfield}.
 	\begin{figure}
 		\centering
-	\begin{minipage}{0.48\textwidth}
+		\begin{minipage}{0.49\textwidth}
 			\centering
-		\includegraphics[width=0.98\textwidth]{./fig/density.png} 
+			\includegraphics[width=0.99\textwidth]{./fig/density.png} 
 			\caption{Density profile of the system at different time moments}
@@
 		\end{minipage}\hfill
-	\begin{minipage}{0.48\textwidth}
+		\begin{minipage}{0.49\textwidth}
 			\centering
-		\includegraphics[width=0.98\textwidth]{./fig/Bfield.png} 
+			\includegraphics[width=0.99\textwidth]{./fig/Bfield.png} 
 			\caption{Magnetic field profile of the system at different time moments}
@@
 
-Simulations showed that ejecta of supernovae, moving in interstellar medium has profile with thin and dense peak with number density $n \approx 10^5 cm^{-3}$ and width $d = 10^{16} cm$, and wide region with number density $\approx 10^4 cm^{-3}$ behind the peak.
+	Simulations showed that ejecta of supernovae, moving in interstellar medium has profile with thin and dense peak with number density $n \approx 10^5~cm^{-3}$ and width $d = 10^{16}~cm$, and wide region with number density $\approx 10^4~cm^{-3}$ behind the peak.
 
-The wind bow-shock has radius $R_w \approx 2\times10^{14} cm$ and number density $n_w \approx 10^9 cm^{-3}$.
+	The wind bow-shock has radius $R_w \approx 2\times10^{14}~cm$ and number density $n_w \approx 10^9~cm^{-3}$.
 With this parameters, and luminocity and temperature of the star, one can estimate observed X-ray flux, produced by inverse compton scattering.
@@
 
-Let consider different models of origin of X-ray radiation, which observed flux in energy range $0.3 - 10\text{ }keV$ is $(1.33 \pm 0.76) \times 10^{-15}\text{ }erg\text{ }s^{-1} cm^{-2}$..
+	Let consider different models of origin of X-ray radiation, which observed flux in energy range $0.3 - 10~keV$ is $F_{Xray} (1.33 \pm 0.76) \times 10^{-15}~erg~s^{-1}~cm^{-2}$.
+
 	At first, we assume that it is synchrotron radiation from the electron distribution, which is prolonged to much higher energies than it is possible to reach in PIC simulation.
-The necessary energy of electrons is determined by energy of emitted photon $E_e = (4\pi{m_e}^3 c^5 E_{ph}/0.29\cdot3 e B h)^{1/2} \approx 2\cdot10^6 m_e c^2$.
-The flux from source with parameters obtained in section \ref{synchrotron} and prolonged distribution is $what$.
-But on should take into account tht lifeime of electrons with this energy in magnetic field $B \approx 0.6 G$ is $\tau = {m_e}^4 c^7 / e^4 B^2 E_e \approx 10^3 s$ and distance that electron can move on this time is about $d = 3\cdot10^13 cm$.
-This means that emitting volume drasticaly decreases and we should evaluate radiation of spherical layer with width $d$.
 
-In our model X-ray flux is generated via inverse compton scattering in the region of iteraction between wind from the first star, and supernova ejecta from the second.
-Relativistic ejecta has huge bulk pressure and wind is compressed to small region with high number dencity.
+	Another option is inverse compton scattering of accelerated electrons.
+But there is well-known relation between synchrotron and  inverse compton radiation from same electrons \cite{Ghisellini} - total emitted powers relate to each other as energy densities of magnetic field and photon field.
+The energy density of magnetic field is $U_B = B^2/8\pi = 0.15 ~erg~cm^{-3}$, and total synchrotron flux is $F_s \approx 10^-16~erg~cm^{-2}~s^{-1}$.
+So to provide observed X-ray flux about $10^{-15}~erg~s^{-1}~cm^{-2}$ there should be energy density of photon field at least $1~erg~cm^{-3}$ which is much more than energy density of mean galactic photon field $10^{-12}~erg~cm^{-3}$ \cite{Mathis1983} and more than the most luminous stars $L \approx 10^{40}~erg~s^{-1}$ can provide on such distances $L/4\pi R^2 c \approx 10^{-7}~erg~s^{-1}$.
+If one choose a close distances, photon energy density may be higher, but part of volume of ejecta on this distance would decrease as square of $R$, and there would not be increase in inverse compton luminosity.
+Also one can consider inverse compton scattering on the self synchrotron radiation.
+The energy density of synchrotron radiation is easy to estimate via observed synchrotron flux $F_{synch} \approx 10^{-15}~erg~cm^{-2}~s^{-1}$.
+Synchrotron energy density $U_{synch} = F_{synch}/c \cdot (D/R)^2 \approx 10^{-7}~erg~cm^{-3}$, where $D$ is distance to the source $150~Mpc$, is also too low.
+So we consider that X-ray radiation is not emitted by the same electrons that produced synchrotron radiation.
+
+	In our model X-ray flux is generated via inverse compton scattering in the region of interaction between wind from the first star, and supernova ejecta from the second.
+Relativistic ejecta has huge bulk pressure and wind is compressed to small region with high number density.
 Also this region is close to high luminous star, and there are suitable conditions for strong inverse compton radiation.
 Electrons can be accelerated on the shock fronts or on the shearing flows.
-So electrons distribution function is
-%непонятно чт писать
-We assume that source is a cylinder with radius $R_w$, and height also equals $R_w$, with axis oriendted along line of sight, and with uniform number density $n_w$. 
+So electrons distribution function is ???
+
+	We assume that source is a cylinder with radius $R_w$, and height also equals $R_w$, with axis oriented along line of sight, and with uniform number density $n_w = 10^9~cm^{-3}$. 
 	and photons distribution corresponds to the radiation of the first star at distance $R_w$.
-With this parameter modeled X-ray flux in the energy range $0.3-10 keV$ is $F_m = $ which is consistent with observed flux of CSS161010 at 99 day after explosion $F_{obs = }$ \cite{Coppejans2020}.
+With this parameter modeled X-ray flux in the energy range $0.3-10~keV$ is $F_m = $ which is consistent with observed flux of CSS161010 at 99 day after explosion $F_{obs = }$ \cite{Coppejans2020}.
 
+	Also we estimate the synchrotron flux from this source.
+It should be much smaller than flux from the whole remnant, providing observed synchrotron flux.
+Once again we estimate it via relation between photons energy density and energy density of magnetic field.
+Total radio flux is about 10 times smaller than X-ray flux, so if we set limit to fraction of synchrotron flux, emitted from X-ray source as $0.1$ of total radio flux, we can obtain condition on magnetic field of second star ${B_2}^2/8\pi < 0.01 L_2/4\pi c {r_w}^2 \approx 1~G$ at distances about $10^{14}~cm$.
+It is much smaller than field of the first star, and our hydrodynamical simulation takes it into account/
+
+	\section{Conclusions}
+	In this work we presented model of origin of luminous X-ray radiation from FBOT CSS61010.
+
+	\ack{
+		The~authors  were supported by the RSF grant 21-72-20020.
+Some of the modeling was performed at the Joint Supercomputer Center JSCC RAS and some---at the Peter the Great Saint-Petersburg Polytechnic University Supercomputing Center.
+	}
 	\bibliographystyle{iopart-num}
 	\bibliography{fbotcompton}
+
 \end{document}

### 2023-12-12 papers/answer to reviewer.txt
fz

Thanks to the reviewers for important  remarks.

reviewer 1
1) line 35: Nonthermal X-ray luminosity is 10^39 erg/s as written later in the paper.
mistake is corrected
2) Fig.5  : it should be E^2F(E) on the vertical axis
F(E) is defined as energy flux spectrum density, not number of photons spectrum density, so no, in Fig. 4 it should be EF(E)

reviewer 2
1. It is not clear how the authors treat magnetic field amplification—which has bearing on the maximum energy and synchrotron luminosity. How does magnetic field amplification proceed in the Monte Carlo model? In principle, I would expect the non-resonant hybrid instability (Bell 2004) to dominate in this environment. In that case, the amplified field will depend on the acceleration efficiency (and thus the shock inclination) as well as the shock velocity. Regardless, a discussion of magnetic field amplification and its dependence on parameter choices is warranted here.
2. The authors obtain a best-fit shock velocity that is substantially faster than that reported in the literature (Coppejans et al., 2020) despite surprisingly similar inferred densities and magnetic fields. The authors should explain the origins of this discrepancy.
3. CSS161010 was observed at multiple epochs (Coppejans et al., 2020, see their Figure 4). While the most extensive multi-wavelenth SEDs were obtained at t=98 days post-explosion (the epoch considered in the manuscript), there are also detailed radio SEDs at t=68, 162, and 357 days. It would be useful to test whether the authors' model can fit the radio data at these epochs, as well as whether the inferred shock evolution is reasonable. In particular, are these observations consistent with the authors' assumption that CSS161010 does not slow down until at least a year after explosion (line 160)?
4. Related to Item 3: the radio SEDs shown in Coppejans et al., 2020 are far more detailed (>>4 data points) than what is shown in the manuscript. With the additional data points, does model B30 still produce a good fit?
5. While the authors argue that their calculated X-ray luminosity is consistent with observations, the normalization of the SED shown in their Figure 5 looks suspiciously small. I would like to see the X-ray and radio data over-plotted.
6. Section 5 (the jet-massive star interaction model) is interesting, but underdeveloped. I think it is important to make the point that the X-rays may come from another source (especially if, as I suspect, the shock-powered X-rays are a bit low). However, some specifics about the numerical simulations conducted are warranted (e.g., what code was used?). I would also like to see the resulting X-ray SED with data overplotted (both for the CSS161010 model and for the binary, AT2018-like scenario).

Minor points:
1. Given that the authors assumed a free-escape boundary that is half the size of the system, the 100 PeV maximum energy quoted in the abstract is a very optimistic upper limit and should be presented as such.
2. Please elaborate on the assumptions and limitations of the PIC simulations performed in Section 2 (e.g., how many dimensions were used? How realistic was the electron to proton mass ratio?).
3. When referring to the relationship between shock obliquity and acceleration efficiency, the authors should also cite Caprioli and Spitkovsky 2014.
4. Please specify the units of the y-axes in Figures 1-3, preferably in a manner that allows the reader to approximate an acceleration efficiency. Please also specify the obliquity considered in Figure 3.
5. In Figures 1-2, it might be useful to plot in terms of particle momentum (p^4f(E) vs p) order to make the supra-thermal bump more apparent and to eliminate features due to the transition from non-relativistic to relativistic protons.
6. Please clarify how the radio SED prediction would change if a sphere were assumed instead of cylinder (i.e., if there were regions of the source that were optically thicker/thinner at a given frequency)
7. It would be helpful to clarify early on that the magnetic field used to compute the synchrotron emission in Figures 4 and 5 comes directly from Monte Carlo simulations. Given assumptions of density, shock velocity, and magnetic inclination, acceleration efficiency and thus the amplified magnetic field are derived parameters. As such, the F(nu, B, n) notation starting around line 231 is confusing since, as far as I can tell, B is not a model input.
8. It would also be helpful to identify, either in Figure 5 or in the text, key features in the multiwavelenth SED. As I understand it, the first peak is nu_SSA, the dip corresponds to the hardening of the electron spectrum due to non-linear DSA, and the second peak is the cooling break.
9. Regarding the discussion of the strong inferred magnetic field starting around line 252: perhaps the origin of this magnetic field is simply CR-driven magnetic field amplification which, as I understand it, is already included in the Monte Carlo model?

### 2025-10-23 disser/part2.tex
disser

\chapter{Классификация гамма-всплесков, зарегистрированных в эксперименте Конус-Винд}\label{KW_GRB_classification}

\section{Введение}
Первым свидетельством наличия двух классов всплесков было обнаружение их бимодального 
распределения по длительности~\citep{Mazets_1981_part_1,Norris_1984,Kouveliotou_1993,Aptekar_1998}. 
В работе \citep{Kouveliotou_1993} были введены две меры длительности $T_{50}$ и $T_{90}$, 
которые соответствуют временам накопления 50\% и 90\% от суммарного числа отсчётов 
всплеска (интервалы от 25\% до 75\% и от 5\% до 95\% отсчётов, соответственно), 
и определена граница между длинными и короткими всплесками $T_{90}=2$~с 
на основе локального минимума бимодального распределения. 
В работе \citep{McBreen_1994} обсуждается, что распределение всплесков по 
длительности $T_{90}$ хорошо описывается суммой двух логнормальных компонент. 
Более детальный анализ распределения по $T_{90}$ всплесков из третьего каталога 
эксперимента BATSE~\citep{Horvath_2002} показал, что распределение 
хорошо аппроксимируется тремя логнормальными компонентами. 
Третья компонента располагалась между короткими и длинными всплесками, 
что свидетельствовало о наличие третьего класса всплесков с промежуточной длительностью.
Доля промежуточных всплесков составила 6\%.

Систематические ошибки величин $T_{50}$ и $T_{90}$, которые искажают наблюдаемое 
распределение всплесков, были рассмотрены в работе~\citep{Norris_and_Bonnel_2006ApJ} 
на примере всплесков, зарегистрированных в эксперименте \textit{CGRO}-BATSE~\citep{Fishman_1992NASCP3137}. 
Было показано, что как минимум четыре фактора влияют на измеренное значение длительности 
всплеска и размывают границы между классами коротких и длинных всплесков: 
(1)~различие в отношении сигнал-шум (S/N) между самым интенсивным и самым слабым 
всплеском  при условии, что эти всплески не отличаются по другим параметрам, 
приводит к различию их длительностей $T_{90}$ в $\sim 2$~раза~\citep{Bonnell_1997}; 
(2)~длительность космологически удалённых длинных всплесков (с красными смещениями источников 
$z \sim 2 \textrm{--}10$) увеличивается в 2--5 раз по сравнению со всплесками, 
пришедшими с $z=1$, в это же время поправка длительности 
для коротких всплесков не превышает двух раз, так как их источники в основном 
расположены на $z<1$ (этот фактор увеличивает обособленность распределений 
коротких и длинных всплесков); (3)~эффект зависимости длительности отдельных импульсов 
всплесков от энергии $\propto E^{-0.35}$~\citep{Fenimore_1995}, а так же модификация 
спектра за счёт космологического расширения приводит к зависимости длительности 
всплеска от энергетического диапазона наблюдений; (4)~распределение по длительностям 
обрезано со стороны коротких всплесков вследствие наблюдательной селекции триггерного 
алгоритма.

Кроме $T_{50}$ и $T_{90}$ была предложена и другая мера длительности~--- время 
эмиссии~\citep{Mitrofanov_1999}, которая вычисляется как суммарная длительность 
бинов временной истории, в которых интенсивность излучения превышает заданный 
уровень от пиковой интенсивности. Этот уровень задается таким образом, чтобы 
суммарное число отсчетов в бинах составляло заданную долю от общего числа отсчётов 
всплеска (использовались уровни 30\% и 50\%, соответствующие меры длительности 
были обозначены $\tau_{30}$ и $\tau_{50}$). Было показано, что данная мера 
длительности гораздо менее чувствительна и к отношению сигнал-шум и к количеству 
импульсов во всплеске, разделенных значительным промежутком фона, так как этот 
промежуток не входит, по определению, в определяемое время эмиссии. Ясно, однако, 
что эта мера чувствительна к временному разрешению. Распределения всплесков 
по $\tau_{30}$ и $\tau_{50}$  также являются бимодальными. Эти меры длительности 
не получили широкого распространения, поэтому классификация всплесков 
на их основе в данной работе не рассматривается.

Короткие и длинные всплески, в среднем, отличаются жёсткостью спектра.
В работе \citep{Kouveliotou_1993} было обнаружено, что короткие всплески BATSE, в 
среднем, имеют большую жёсткость ($\rmn{HR}_{32}$), чем длинные, где $\rmn{HR}_{32}$ определялась 
как отношение числа отсчетов, накопленных за $T_{90}$ в энергетических 
диапазонах 100--300~кэВ и 50--100~кэВ (диапазоны 3 и 2) эксперимента \textit{CGRO}-BATSE. 
Однако, в последующих работах было отмечено, что жёсткость не является обязательной чертой 
коротких всплесков~\citep[см., например,][]{Sakamoto_2006_proc, Norris_and_Bonnel_2006ApJ}.

На основе распределения на плоскости жёсткость-длительность также был обнаружен 
третий класс всплесков ~\citep{Mukherjee_1998, Hakkila_2000}. Анализ влияния 
систематических эффектов на наличие третьего класса всплесков был проведён 
в~\citep{Hakkila_2003}. Авторы показали, что третий класс всплесков не является 
физическим и может являться следствием селекции триггерного алгоритма. Также было 
показано, что сам по себе эффект зависимости длительности от яркости 
всплеска~\citep{Bonnell_1997} не может привести к появлению третьего класса всплесков. 
Однако, в последующей работе~\citep{Horvath_2006} были приведены доводы в пользу 
реальности третьего класса всплесков и исследовалось количество и значимость 
классов всплесков на основе параметров распределения на плоскости 
$\log T_{90}$--$\log \rmn{HR}_{32}$ для 1956 всплесков, зарегистрированных 
в эксперименте BATSE. При этом распределение аппроксимировалось 
суммой двумерных нормальных распределений. Было установлено, что распределение 
наилучшим образом описывается тремя компонентами (классами). При этом доля всплесков 
из класса с промежуточной длительностью составляет 10--15\%. Также было показано, что поправка набора 
всплесков на селекцию триггерного алгоритма сохраняет полученное разделение на классы. 
Аналогичный метод был применен к 222 всплескам \textit{Swift}-BAT~\citep{Horvath_2010}, 
при этом также было обнаружено три класса всплесков. При этом доля промежуточного 
класса всплесков составила~30\%.

Помимо жесткости и длительности, еще одним параметром для классификации всплесков 
может служить спектральная задержка. Этот параметр характеризует запаздывание 
излучения в более мягком диапазоне по отношению к излучению в более жестком диапазоне. 
Значение задержки ($\tau_\rmn{lag}$) соответствует положению  максимума 
кросскорреляционной функции временных историй в двух энергетических диапазонах~\citep{Norris_2000}. 
Анализ спектральных задержек коротких всплесков BATSE и \textit{Swift}-BAT показал, 
что они распределены вблизи нуля с разбросом 
порядка 25~мс \citep{Norris_and_Bonnel_2006ApJ, Norris_2011ApJ}. В то же время длинные 
всплески могут иметь значительные задержки. Идея о том, что различие спектральных 
задержек длинных и коротких всплесков отражает различие физических свойств источников 
всплесков, была выдвинута в работе~\citep{Gehrels_2006_Nature}.

%Существует несколько моделей, объясняющих появление спектральной задержки. 
%В работе~\citep{Ioka_and_Nakamura_2001} появление задержки объясняется, тем что 
%узкий джет наблюдается под разными углами для разных всплесков. Большая задержка 
%соответствует большему углу наблюдения относительно оси джета. 
%Авторы работы~\citep{Schaefer_2004_lag} считают, что спектральная задержка в 
%каждом импульсе всплеска характеризует время потери энергии излучающими частицами.

После начала эры многоволновых наблюдений гамма-всплесков стало возможным 
классифицировать всплески на основании параметров послесвечений и родительских галактик. 
В работах~\citep{Zhang_2006, Zhang_2007, Zhang_2009} была предложена схема классификации всплесков, 
основанная на этих параметрах, на физические типы: I и~II. Считается, что всплески типа~I, 
генерируются при слиянии компактных объектов, а всплески типа~II образуются 
при коллапсе молодых массивных звёзд. В работе~\citep{Zhang_2009} было показано, 
что всплески типа II на плоскости жёсткость-длительность располагаются в области 
длинных/мягких всплесков, при этом всплески типа~I~--- в области коротких/жестких событий. 
Авторы также отмечают, что всплески типа I обладают незначительной спектральной задержкой.

Неоднозначность соответствия коротких всплесков и всплесков типа~I связана, в том числе, с 
существованием коротких гамма-всплесков, сопровождающихся слабым сравнительно мягким 
излучением в гамма-диапазоне, длящимся десятки секунд после начального короткого импульса. 
Это явление получило название продленного излучения (EE).
Впервые этот класс всплесков был обнаружен в данных BATSE~\citep{Burenin_2000AstL} и 
Конус~\citep{Mazets_2002astro_ph}. В работе~\cite{Frederiks_2004ASPC} 
было обнаружено 11 всплесков с продлённым излучением среди 125 коротких гамма-всплесков KW. 
Для оставшихся всплесков был обнаружен избыток излучения на интервале 5--100~с в сумме временных 
профилей с вычтенным фоном. Такой же избыток был выявлен в данных 
BATSE~\citep{Lazzati_2001AandA, Connaughton_2002ApJ}, \textit{BeppoSAX}-GRBM~\citep{Montanari_2005ApJ} и 
\textit{INTEGRAL}-SPI-ACS~\citep{Minaev_2010AstL}. 
%написать про всплески с EE на BAT и GBM
Основную сложность представляет разделение коротких всплесков с EE 
и длинных всплесков. Авторы работы~\citep{Norris_and_Bonnel_2006ApJ} использовали визуальные признаки 
для выделения коротких всплесков с продлённым излучением: длительность начального 
импульса $T_{90}<2$~с, отсутствие заметной спектральной эволюции в начальном 
импульсе и EE. Авторы обнаружили в данных BATSE 8 всплесков, 
удовлетворяющих этим критериям, и показали, что их спектральные задержки близки к нулю. 
Также в работе отмечается наличие у некоторых всплесков интервала с практически нулевой интенсивностью излучения 
между начальным импульсом и продлённым излучением. Сравнение интенсивностей продлённого 
излучения для этих 8 всплесков и интенсивности излучения, обнаруженного в 
суммарном временном профиле в работе~\citep{Lazzati_2001AandA}, показывает, 
что динамический диапазон интенсивности продлённого излучения составляет $\sim 10^4$, 
так же как и динамический диапазон отношения интенсивности начального импульса и продлённого излучения.
В работе~\citep{Norris_2010ApJ} на основе 12 всплесков с продлённым излучением, зарегистрированных 
\textit{Swift}-BAT, был оценен физический порог отношения интенсивностей продлённого излучения 
и начального пика как единицы $\times 10^3$ и было показано, что для начальных импульсов 
всплесков без продлённого излучения характерны в $\sim 2\textrm{--}3$ раза меньшие длительности
 по сравнению с всплесками с EE, также всплески без продлённого излучения 
имеют меньшее число пиков в начальном импульсе.

Различие природы источников коротких гамма-всплесков с продлённым излучением и 
без него в настоящее время не подтверждено. В работе~\citep{Troja_2008MNRAS} 
было установлено, что смещения относительно центра родительской галактики 
коротких всплесков с продлённым излучением в среднем меньше, чем для всплесков без него: 
единицы и десятки килопарсек, соответственно. 
Исследование~\citep{Fong_2010ApJ} с использованием данных наблюдений 
телескопа \textit{Hubble Space Telescope} опровергло этот результат.

В данной главе анализируются временные истории всплесков, зарегистрированных в 
эксперименте Конус-Винд. Всплески классифицируются на основе длительности, жесткости и 
спектральной задержки, и определяется набор коротких всплесков для дальнейшей локализации и спектрального анализа. 
При этом обсуждается принадлежность исследуемых всплесков к типам~I и~II. 
В разделе~\ref{sec:GRB_sample} приводится описание использованного набора всплесков. 
В разделе~\ref{sec:Durations} рассматривается распределение всплесков по длительности, 
определяется граница между длинными и короткими всплесками на основе 
длительности $T_{50}$ и определяется итоговый набор коротких всплесков. 
В разделе~\ref{sec:Hardness} рассматриваются жесткости всплесков и производится 
классификация всплесков на основе соотношения жёсткость-длительность. 
В разделе~\ref{sec:Lags} анализируются спектральные задержки коротких всплесков. 
Сравнение классификации на физические типы~I и~II и классификации на основе жесткости, 
длительности и спектральной задержки приведено в разделе~\ref{sec:Phys_Classification}. 
Обобщение и обсуждение результатов приведено в разделе~\ref{sec:Conclision}.

\section{Набор всплесков}\label{sec:GRB_sample}
За период с ноября 1994~г. по декабрь 2010~г. KW зарегистрировал 2008 гамма-всплесков 
в триггерном режиме. Среди них 69 всплесков содержат сбои, представляющие собой 
промежутки в данных во время всплеска, или значительные флуктуации фона, 
накладывающиеся на всплеск. Всего 1939 всплесков без сбоев.

Для вычисления параметров временных историй использовалась 
сшивка фоновой и триггерной записи. Длительность всплесков, у которых значительная 
часть события лежит в бине фоновой записи, предшествующем триггерной записи, может быть 
искажена из-за низкого разрешения фоновой записи. Из дальнейшего рассмотрения 
было исключено 105 таких событий. Таким образом, для анализа использовались временные 
истории 1834 гамма-всплеска.

\section{Длительности}\label{sec:Durations}
Для вычисления длительностей мы использовали сумму временных историй всплеска 
в энергетических диапазонах G2 и G3. Такой выбор обоснован тем, что  пиковая 
энергия $E F_{E}$ спектра ($E_\rmn{p}$) абсолютного большинства всплесков лежит в 
диапазоне $E_\rmn{p}>80$~кэВ, и, следовательно, отсчёты, соответствующие основному энерговыделению 
во всплеске, регистрируются в диапазонах G2 и G3. Фон в G2 и G3 стабилен на более 
длительных интервалах времени, чем в G1. Это позволяет с гораздо меньшей вероятностью 
неверно определять начало и конец всплеска из-за флуктуаций фона. Также в диапазон G1 
может попадать начало рентгеновского послесвечения, учёт которого исказит 
длительность начального импульса всплеска.

Уровни фона для вычисления длительностей определялись по методике, описанной в 
разделе~\ref{sec:Bg_rate} Главы~\ref{KW_description}. 
В результате, из всего набора 1834 всплесков у 96\% всплесков длительность 
интервала для определения фона составила 750~с и только у $<1$\% всплесков длительность этого 
интервала составила 30--100~c.

\subsection{Автоматическая процедура определения длительности}
Основной задачей при вычислении любой меры длительности является определение момента 
начала и конца всплеска. В работе~\citep{Koshut_1996}, описывающей методику вычисления 
длительностей $T_{90}$ и $T_{50}$  для второго каталога BATSE, моменты начала и 
конца всплеска $t_{0}$ и $t_{100}$ определялись визуально на основании интеграла 
от временной истории с вычтенным фоном всплеска, где $t_{x}$ соответствует времени накопления заданной 
доли $x$  (0\%, 5\%, \dots, 100\%) от полного числа отсчётов всплеска. 
В работе~\citep{Bonnell_1997} использовалась автоматическая процедура поиска 
$t_{0}$ и $t_{100}$, которая заключалась в поиске положительной флуктуации числа 
отсчетов с вычетом фона со значимостью от $3\sigma$ до $6\sigma$ на временных 
масштабах от 0.5~с до 16~с в диапазоне~$>25$~кэВ.
% что для Swift-BAT и Fermi-GBM?

Для вычисления длительности всплесков KW использовалась автоматическая 
процедура поиска начала и конца всплеска. Поиск производился на интервале от 
$T_0-250$~с до $T_0+230$~с (конец триггерной записи), где $T_0$~--- время триггера KW. 
Для некоторых всплесков границы интервала вычисления длительности отличались от указанных, 
к примеру, в случае присутствия солнечной вспышки в данных. Для поиска начала 
всплеска производился поиск положительной флуктуации числа отсчетов с вычетом 
фона на заданном уровне значимости. При этом поиск превышения производился на 
интервале с длительностью от временного разрешения текущего участка записи до 100~с. 
Началом всплеска $t_0$ считается начало интервала, на котором было обнаружено превышение, 
если вероятность его случайного обнаружения была меньше пороговой. 
Иллюстрация алгоритма поиска начала всплеска приведена на рис.~\ref{fig:T100_calculation}.
Времена начала интервалов поиска превышения брались последовательно, начиная от $T_0-250$~с. 
Конец всплеска~--- $t_{100}$ определяется аналогичного, при этом поиск ведётся в 
обратном направлении по времени, начиная от $T_0 + 230$~с. Длительности всплесков 
вычислялись для порогов, соответствующих вероятности случайного превышения 
значений $4\sigma$, $5\sigma$ и $6\sigma$ для распределения Гаусса. При этом на 
интервалах, где число отсчётов от фона меньше 20, для вычисления вероятности 
случайного превышения использовалась статистика Пуассона. Значения $T_{90}$ и $T_{50}$ 
и их ошибки определялись методом, описанном в работе~\citep{Koshut_1996}.

\begin{figure}[h]
  \begin{minipage}[h]{0.5\textwidth}
    \center{\includegraphics[width=1.0\textwidth]{gDurStartRU.eps} \\ а)}
  \end{minipage}
  \hfill
  \begin{minipage}[h]{0.5\textwidth}
    \center{\includegraphics[width=1.0\textwidth]{gDurT5RU.eps} \\ б)}
  \end{minipage}  
  \caption{
  (а) Иллюстрация определения времени начала всплеска $t_0$ для GRB20031214\_T36655. 
  Сплошная линия~--- интегральное число отсчетов всплеска с вычетом фона. 
  Число отсчётов $N_\rmn{acc}$ (горизонтальные штриховые линии), 
  накопленное на интервале $\Delta T_\rmn{acc}$ (вертикальные штриховые линии), 
  соответствует превышению $5\sigma$ над фоном. Уровень $1\sigma$ относительно интегрального 
  числа отсчётов на момент $t_0$  изображен горизонтальной штрихпунктирной линией. 
  Горизонтальная точечная линия обозначает нулевой уровень.
  (б) Иллюстрация определения начала интервала $T_{90}$ ($t_5$) для того же всплеска.
  Число отсчётов, соответствующие 5\% от общего числа отсчётов $N_\rmn{tot}$ 
  обозначено горизонтальными штриховыми линиями, время начала всплеска $t_0$ и время $t_5$ 
  обозначены вертикальными штриховыми линиями. \label{fig:T100_calculation} }
\end{figure}

\subsection{Распределения по длительностям}
Для 1834 всплесков KW были построены гистограммы распределений всплесков по $T_{90}$ и $T_{50}$ 
для порогов значимости поиска начала и конца всплеска $4\sigma$, $5\sigma$ и $6\sigma$.

Распределения  аппроксимировались методом минимизации $\chi^2$ суммой двух логнормальных распределений
\begin{equation}
f(x) = \sum_{l=1}^{2} A_l f_l(x) \mbox{ ,}
\end{equation}
где
\begin{equation}
f_l(x) = \frac{1}{w_l \sqrt{\pi/2}} \exp\left(\frac{-2(x-x_{cl})^2}{w_l^2}\right) \mbox{ ,}
\end{equation}
с параметрами $A_l$~--- площадь под распределением, $w=2\sigma$~--- ширина, $x_{cl}$~--- среднее значение и 
$x=\log T$. При этом число всплесков в бине по модели равно  интегралу от аппроксимирующей функции в границах бина.

Доверительные интервалы параметров распределения на уровне 68\% вычислялись 
методом Mонте-Карло для 1000 реализаций распределения. При этом значение в каждом 
бине распределения разыгрывалось как пуассоновская случайной величиной со средним, равным 
числу отсчётов в соответствующем бине исходного распределения. Полученный набор 
значений каждого параметра распределения сортировался по возрастанию. В качестве нижней и верхней 
границы доверительного интервала на уровне 68\% из этого отсортированного массива 
параметров брались значения с индексами 160 и 840, соответственно. 
В качестве значения параметра бралась медиана распределения.

Из-за эффектов селекции число очень коротких всплесков ($T_{90} \lesssim 100$~мс) и 
число длинных всплесков ($T_{90} \gtrsim 200$~с) в наборе KW недооценено. Так же неверно 
считать ошибки числа отсчётов в бине гауссовыми при числе отсчётов $<10$, поэтому 
бины с $\log T_{90} \leq -1.2$ ($T_{90} \leq 0.063$~с) и $\log T_{90} \geq 2.4$ ($T_{90} \geq 251$~с) 
не учитывались при аппроксимации гистограммы распределения по $T_{90}$,  для аппроксимации 
использовалось 18 бинов. При аппроксимации гистограммы распределения по $T_{50}$ игнорировались 
бины c $\log T_{50} \leq -1.8$ ($T_{50} \leq 0.016$~с) и  $\log T_{50} \geq 2.0$ 
($T_{50} \geq 100$~с), для аппроксимации использовалось 19 бинов.

Параметры распределений представлены в таблицах~\ref{tab:T90_distr} и~\ref{tab:T50_distr}. 
Из представленных результатов видно, что при изменении порога от $4\sigma$ до $6\sigma$ 
параметры распределения по $T_{50}$ и $T_{90}$ для полного набора 1834 всплесков изменяются. 
Распределение длинных всплесков смещается в сторону меньших длительностей, такое же смещение, 
но в меньшей степени, наблюдается для распределение коротких всплесков. 
При этом изменение параметров распределения по $T_{90}$ гораздо существенней, 
чем для распределения по $T_{50}$, см. таб.~\ref{tab:Txx_vs_thd},
где $T_{\rmn{int}}$ соответствует пересечению распределений и 
$\delta T =|T_{4\sigma}-T_{6\sigma}| / T_{4\sigma}$. 
Так как вариации параметров распределения по $T_{50}$ существенно меньше, 
чем для распределения по $T_{90}$ для классификации всплесков на основе длительностей 
мы выбрали величину~$T_{50}$.

\input{tables/duration_vs_thd}

Вследствие наблюдательной селекции триггерной системы по отношению сигнал-шум 
наш набор всплесков неоднороден. Как видно из диаграммы $T_{50}$--$\rmn{S/N}$~(рис.~\ref{img:SNvsT50}) 
в наборе отсутствуют всплески с отношением сигнал-шум $\rmn{S/N} < 10$ и $T_{50} \lesssim 100$~мс. 
Здесь S/N определено как отношение числа отсчётов от источника к стандартному отклонению 
числа отсчётов фона на интервале 64~мс с максимальной скоростью счёта. 
Для дальнейшего рассмотрения был выбран однородный набор всплесков с $\rmn{S/N}\geq 10$, 
который содержит 1168 всплесков. 
Результаты аппроксимации распределений набора по $T_{50}$ и $T_{90}$ приведены в 
таблицах \ref{tab:T90_distr} и~\ref{tab:T50_distr}, и на рис.~\ref{img:T90andT50s5}. 
Относительные изменения параметров скорректированных распределений при изменении 
порога от $4\sigma$ до $6\sigma$ приведены в таб.~\ref{tab:Txx_vs_thd}.

В качестве порога значимости был выбран уровень $5\sigma$ так как было замечено, 
что при пороге $4\sigma$ в качестве начала и конца всплеска часто выбираются 
флуктуации фона или эпизоды излучения транзиентов, не относящиеся ко всплеску  
(установить отношение подобных слабых эпизодов к триггерному всплеску в 
большинстве случаев не представляется возможным). При пороге $6\sigma$ часто 
игнорируется <<хвост>>, непосредственно прилегающий к основному пику, тем самым полное 
число отсчетов во всплеске может недооцениваться на величину до $\sim 70$\%.
Такая ситуация возможна, к примеру, для коротких всплесков с продлённым излучением.

В качестве границы между длинными и короткими всплесками была выбрана точка пересечения 
двух логнормальных распределений по $T_{50}$ для порога значимости $5\sigma$ 
набора 1168 всплесков $T_{50\rmn{,int}} = 0.6$~с.

Для указанного распределения и границы между длинными и короткими всплесками 
можно оценить долю <<засорения>>\ набора коротких всплесков длинными и наоборот. 
Доля <<засорения>>\ набора коротких всплесков длинными равна отношению интеграла 
от распределения длинных всплесков и интеграла от распределения коротких всплесков 
на интервале $-\infty$ до $T_{50\rmn{int}}$) и составляет $\approx 7$\%. 
При этом доля <<засорения>>\ длинных всплесков короткими составляет~$\approx 2$\%.

Границы диапазонов KW менялись со временем, что могло внести систематический 
сдвиг в длительности всплесков, зарегистрированных в различные периоды, и 
привести к изменению со временем границы между длинными и короткими всплесками. 
Для анализа дрейфа границы между длинными и короткими всплесками набор 1168 
всплесков был разделён на три поднабора: 583 GRBs ноябрь 1994~-- май 2003; 
585 GRBs июнь 2003~-- декабрь 2010; 582 GRBs  май 1999~-- январь 2007. 
Полученные для поднаборов границы $T_{50\rmn{int}}$ равны: 
$0.41_{-0.09}^{+0.19}$~c, 
$0.74_{-0.16}^{+0.22}$~c и 
$0.62_{-0.17}^{+0.28}$~c, соответственно. 
Наблюдаемый дрейф границы лежит в пределах $1\sigma$ интервала для границы $T_{50\rmn{int}} = 0.6$~с, 
полученной с использованием полного набора всплесков.

\begin{figure} [h] 
  \center
  \includegraphics [width=0.75\textwidth] {gSNvsT50_ru}
  \caption[Соотношение S/N--$T_{50}$]{Соотношение S/N--$T_{50}$ для 1834 всплесков. 
  Пунктирная линия соответствует $\rmn{S/N}=10$. Штрихованная область наиболее подвержена 
  селекции триггерного алгоритма. Отношение сигнал-шум $\rmn{S/N}\geq 10$ имеют 1168 всплесков.} 
  \label{img:SNvsT50}  
\end{figure}

\begin{figure}[h]
  \begin{minipage}[h]{0.5\textwidth}
    \center{\includegraphics[width=1.0\textwidth]{dT90s5simple} \\ а)}
  \end{minipage}
  \hfill
  \begin{minipage}[h]{0.5\textwidth}
    \center{\includegraphics[width=1.0\textwidth]{dT90s5simple_SN10} \\ б)}
  \end{minipage}
  \vfill
  \begin{minipage}[h]{0.5\textwidth}
    \center{\includegraphics[width=1.0\textwidth]{dT50s5simple} \\ в)}
  \end{minipage}
  \hfill
  \begin{minipage}[h]{0.5\textwidth}
    \center{\includegraphics[width=1.0\textwidth]{dT50s5simple_SN10} \\ г)}
  \end{minipage}
  \caption[Распределения 1834 и 1168 всплесков с $\rmn{S/N}\geq 10$ по $T_{90}$ и~$T_{50}$]
  {Распределения 1834 и 1168 всплесков с $\rmn{S/N}\geq 10$ по $T_{90}$~(а и б, соответственно) и 
  $T_{50}$ (в и г, соответственно) для порога $5\sigma$ (непрерывная гистограмма), 
  игнорированные при аппроксимации бины гисторграммы показаны пунктиром. 
  Аппроксимация распределения суммой двух логнормальных распределений  показана 
  непрерывной линией, распределение коротких и длинных всплесков показаны 
  штрихпунктирными линиями.  Вертикальные штриховые лини обозначают средние значения 
  распределений коротких и длинных всплесков и точку их пересечения. 
  На нижней панели каждого рисунка показаны невязки аппроксимации.}
  \label{img:T90andT50s5}  
\end{figure}

\begin{landscape}
\input{tables/duration_tables_mc_diss}
\end{landscape}

\subsection{Сравнение длительностей определенных по данным BATSE и KW}
Среди 1939 всплесков KW без сбоев 267 всплесков наблюдались BATSE. 
Соотношение длительностей $T_{50}$ и $T_{90}$, вычисленных по данным BATSE~\citep{Paciesas_1999} 
и KW, изображено на рис.~\ref{img:T90andT50_KWvsBATSE}.

Большинство всплесков имеют б\'{о}льшую длительность по данным BATSE, 
в особенности $T_{90}$ по сравнению с KW, что связано с большей  эффективной площадью детекторов 
BATSE $\sim 10^3$~см$^2$ против $(0.80\textrm{--}1.6)\times 10^2$~см$^2$ у KW. 
Другим фактором является то, что длительность всплесков BATSE определяется в 
диапазоне $>25$~кэВ, что позволяет учитывать продолжительные мягкие хвосты всплесков. 
Более редкой ситуацией является существенное превышение длительностей по данным KW, 
наблюдаемое в 10 всплесках. Шесть всплесков имеют плавное спадание/нарастание интенсивности, 
которое не было учтено при вычислении  длительностей по данным BATSE, возможно, 
из-за недостаточно точной аппроксимации фона. В трех всплесках по данным KW
в интервал $T_{90}$ попадает слабый импульс, отделенный от основного пика. 
При этом эклиптические широты основного пика и слабого импульса согласуются, 
что позволяет отнести их к одному источнику. Во всплеске GRB~950904, 
$T_{0\rmn{,BATSE}} = 52777$~s~UT ($T_{90\rmn{,KW}} = 99 \pm 4$~с, $T_{90\rmn{,BATSE}} = 11.1\pm 1.8$~с) 
длительности BATSE даны для второго импульса, так как в~\citep{Hurley_2005} полагается, 
что два импульса являются различными всплесками.

Среди общих всплесков KW и BATSE 46 имеют $T_{50\rmn{KW}} \leq 0.6$~с, 
из них 9 входят в набор 105 всплесков с искаженной длительностью (см.~Раздел~\ref{sec:GRB_sample}). 
Из оставшихся 37, 8 имеют $T_{90\rmn{BATSE}} > 2$~с, 
из них один является короткими всплеском с продлённым излучением. Таким образом, 
засорение нашего списка длинными всплесками, связанное с низким S/N всплесков KW
по сравнению с BATSE, составляет 19\% (=7/37). % без учета всплесков с продлённым излучением. 
Ранее в работе~\citep{Ofek_2007ApJ} была получена доля 
<<засорения>> коротких всплесков KW длинными $\approx 60$\%. В этой работе
длительность всплесков в данных KW оценивалась визуально, что могло 
стать причиной получения высокой доли засорения.

\begin{figure}[h]
  \begin{minipage}[h]{0.5\textwidth}
    \center{\includegraphics[width=1.0\textwidth]{gT90_KWvsBATSE_ru} \\ а)}
  \end{minipage}
  \hfill
  \begin{minipage}[h]{0.5\textwidth}
    \center{\includegraphics[width=1.0\textwidth]{gT50_KWvsBATSE_ru} \\ б)}
  \end{minipage}
  \caption{Соотношение длительностей $T_{90}$~(a) и $T_{50}$~(б), определённых по 
  данным BATSE и KW для 267 гамма-всплесков.}
  \label{img:T90andT50_KWvsBATSE}  
\end{figure}

\subsection{Набор коротких всплесков}
Набор из 1834 всплесков KW содержит 277 событий с $T_{50} \leq 0.6$~с.    %<-- число 277 проверено
Среди них было обнаружено: пять очевидно длинных всплесков, в которых $T_{50}$ 
% (ID = 1122, 2196, 2922, 2184) и ID=1228 $T_{90} = 8$~с в данных SAX
определялось наиболее сильным импульсом, при этом полная длительность всплеска была 
значительно больше 2~с; и 12 кандидатов в короткие всплески с продлённым излучением.
Визуальный анализ показал, что среди 69 всплесков со сбоями пять имеют $T_{50} \leq 0.6$~с, 
эти всплески были включены в итоговый набор. 
% ID = 243, 704, 882 (солн. вспышка в данных), 934, 3319 (пропущен ранее)
Таким образом, набор коротких всплесков KW содержит 265 событий без продлённого излучения.

Кроме того, было обнаружено 19 кандидатов во всплески с продлённым излучением среди всплесков с $T_{50} > 0.6$~с. 
К этому типу всплесков были отнесены события, имеющие короткий начальный импульс с $T_{50} \leq 0.6$~с, 
за которым следует эпизод излучения, не содержащий импульсов с заметной спектральной эволюцией. 
В некоторых случаях начальный пик и продлённое излучение были разделены интервалом, 
на котором интенсивность излучения мала. Таким образом, полученный набор коротких всплесков с 
продлённым излучением насчитывает 31 событие.

Объединённый набор коротких всплесков с учётом всплесков с продлённым излучением 
содержит 296 событий.
\FloatBarrier

\section{Жесткости}\label{sec:Hardness}
Для анализа спектральных различий всплесков использовалось две величины: интегральная 
жёсткость ($\rmn{HR}_{32}$)~--- отношение числа отсчетов, накопленных в каналах G3 и G2 
за интервал $T_{100}$ и пиковая жёсткость ($\rmn{HR}_{32\rmn{pk}}$)~--- отношение числа отсчетов, 
накопленных за интервал 64~мс с пиковой скоростью счета. 
Отношение числа отсчётов в каналах G3 и G2 было выбрано, так как 
оно наименее чувствительно к углу падения на детектор и более чувствительно 
к пиковой энергии $E_\rmn{p}$.

При вычислении жесткости $\rmn{HR}_{32}$ был учтён временн\'{о}й дрейф границ диапазонов. 
Для этого трёхканальный спектр отсчётов аппроксимировался степенной функцией 
с экспоненциальным завалом $dN/dE \propto E^{\alpha} \exp(-E/E_0)$. 
С использованием полученной аппроксимации, вычислялось количество отсчётов в номинальных 
границах диапазонов: 13--50~кэВ~(G1), 50--200~кэВ~(G2) и 200--750~кэВ~(G3). 
Реальные границы диапазонов были вычислены с использованием калибровки, полученной 
по многоканальному спектру (см. Главу~\ref{KW_description}).

При рассмотрении интегральной жесткости использовался однородный набор из 1168 
всплесков с $\rmn{S/N} \geq 10$. Набор содержит восемь всплесков со сбоями в G1, 
для которых невозможно провести аппроксимацию трёхканального спектра. 
Из оставшихся всплесков, 17 имеют значительную ошибку $\rmn{HR}_{32}$ 
($\sigma\rmn{HR}_{32}/\rmn{HR}_{32} >0.3$), эти всплески были исключены из рассмотрения. 
В итоге для исследования соотношения жёсткость-длительность использовалось 1143 всплеска.

Распределение этого набора всплесков по $\rmn{HR}_{32}$ хорошо описывается суммой двух 
лог-нормальных распределений. Добавление третьей компоненты не даёт существенного 
улучшения аппроксимации.

Аналогично работе~\citep{Horvath_2006} распределения всплесков на 
плоскости $\log T_{50}$--$\log \rmn{HR}_{32}$ аппроксимировалось суммой двух и трёх 
гауссовых компонент, при помощи метода наибольшего правдоподобия. Каждая компонента имеет вид: 
\begin{equation}
p(x,y| l) = \frac{1}{2\pi \sigma_x \sigma_y \sqrt{1-r^2}} \times 
\exp\left[ -\frac{1}{2(1-r^2)}\left( \frac{(x-a_x)^2}{\sigma_x^2} + 
\frac{(y-a_y)^2}{\sigma_y^2} -\frac{C}{\sigma_x \sigma_y}\right)\right]\mbox{ ,}
\end{equation}
где
\begin{equation}
C = 2r(x-a_x)(y-a_y)\mbox{ ,} \nonumber
\end{equation}
где $x=\log T_{50}$ и $y=\log \mbox{HR}_{32}$,  $a_x$, $a_y$~--- средние, 
$\sigma_x$, $\sigma_y$~--- дисперсии, $r$~--- коэффициент корреляции, 
и $l$~--- номер компоненты. При этом функция правдоподобия определяется следующим образом:
\begin{equation}
L = \sum_i \ln p(x_i, y_i)\mbox{ ,}
\end{equation}
где
\begin{equation}
p(x,y) = \sum_l  p(x, y|l)p_l \nonumber
\end{equation}
Аппроксимация производилась с помощью алгоритма 
expectation maximization~\citep{Horvath_2006, Balazs_2003AA} реализованного в пакете 
Scikit-learn~\citep{scikit-learn}. 
Ошибки параметров вычислялись таким же методом, как и для распределения по длительностям. 
На каждой из 1000 итераций генерировалось распределение на плоскости  
$\log T_{50}$--$\log\rmn{HR}_{32}$ на основе ошибок $T_{50}$ и $\rmn{HR}_{32}$. 
Затем из полученного массива параметров в качестве центрального значения и ширины доверительного 
интервала бралась медиана и ширина на уровне 68\%. Значения параметров гауссовых компонент 
для случая двух и трёх компонент представлены в таблицах~\ref{tab:clusters2_HR} 
и~\ref{tab:clusters3_HR}, расположение компонент представлено на рис.~\ref{img:HRvsT50}.

Различие значений функций правдоподобия для двух компонент $L_2 = -1243$ и трёх 
компонент $L_3 = -1219$ ($\Delta L = 24$). Как показано в~\citep{Horvath_2006} 
в этом случае $2\Delta L$ распределено как $\chi^2_6$ и вероятность случайного 
отклонения $2\Delta L = 48$ равна $10^{-8}$. Функция правдоподобия для четырёх 
компонент $L_4 = -1211$, $2(L_4 - L_3) = 16$ вероятность получения этой величины случайным образом равна $0.01$. 
Это свидетельствует о наличии не более трёх классов всплесков. Третий класс всплесков 
значительно перекрывается с классом длинных/мягких всплесков (см. Рис.~\ref{img:HRvsT50}), поэтому нельзя утверждать, 
что он соответствует реальному физически выделенному классу всплесков, а не вызван эффектами селекции. 
Также, учитывая, что распределения по $T_{50}$ и $\rmn{HR}_{32}$ сами по себе описываются 
только двумя компонентами, для дальнейшего описания набора всплесков KW была выбрана
сумма только двух компонент, которые далее именуются короткие/жесткие и длинные/мягкие всплески.

Принадлежность всплеска к классу с номером $l$ определяется на основании значения индикаторной функции
\begin{equation}
I_l =\frac{p_l p(x,y|l)}{\sum_l  p(x, y|l)}
\end{equation}
Используя индикаторную функцию кластера коротких/жестких всплесков $I$, 
всплески были классифицированы следующим образом: 
$I \geq 0.9$~--- Тип~I, $0.1 \leq I < 0.9$~--- Тип~I/II (неопределённый тип), 
$I < 0.1$~--- Тип~II. Названия типов выбраны по аналогии с физической классификацией. 
Результаты классификации представлены на Рис~\ref{img:HRvsT50Types}.

В наборе 1143 всплесков доли всплесков разных типов составляют: Тип~I~--- 18\%,
Тип~I/II~--- 4\% и Тип~II~--- 78\%. Для всех всплесков типа~I длительность согласуется 
с ограничением $T_{50} \leq 0.6$~с. Доля всплесков типа~II среди коротких всплесков 
составляет 7\% (19\% если всплески типа~I/II относятся к типу~II), что согласуется с долей
<<засорения>> коротких всплесков длинными, вычисленной только на основе распределения 
по длительностям. Для временных историй коротких всплесков типа~II характерно 
наличие одного импульса с плавным нарастанием и спадом.
Из 31-го кандидата в короткие всплески с EE начальные импульсы 21-го события 
имеют Тип~I (далее Iee) и 10 имеют неопределённый тип (далее Iee/II).
При этом начальные импульсы двух событий, отнесённых к всплескам с EE, имеют тип~II.   % ID = 554, 2225

Также был произведён анализ распределения длительность-пиковая жёсткость. 
Для этого использовались 1104 GRBs с $\rmn{S/N} \geq 10$ и $\sigma \rmn{HR}_{32\rmn{pk}}/\rmn{HR}_{32\rmn{pk}} <0.3$. 
Результаты аппроксимации двумя и тремя гауссовыми компонентами приведены на рис.~\ref{img:HRpkvsT50}. 
Параметры распределений приведены в таблице~\ref{tab:tab_2clusters_HRpk} и~\ref{tab:tab_3clusters_HRpk}. 
Различие в жесткостях длинных и коротких всплесков составляет примерно 1.7 раза для пиковых 
жесткостей и 2.4 раза для интегральных.

\input{tables/cluster_tables_diss}

\begin{figure}[h]
  \begin{minipage}[h]{0.5\textwidth}
    \center{\includegraphics[width=1.0\textwidth]{list1143calHR_2} \\ а)}
  \end{minipage}
  \hfill
  \begin{minipage}[h]{0.5\textwidth}
    \center{\includegraphics[width=1.0\textwidth]{list1143calHR_3} \\ б)}
  \end{minipage}
  \caption[Аппроксимация распределения $\log T_{50}$--$\log \mbox{HR}_{32}$.]
  {Аппроксимация распределения $\log T_{50}$--$\log \mbox{HR}_{32}$ 
  для 1143 ярких ($\rmn{S/N} \geq 10$) всплесков двумя~(a) и тремя~(б) гауссовыми компонентами. 
  Эллипсами для каждого распределения отмечены области $1\sigma$ и~$3\sigma$. 
  Пунктирная вертикальная линия~--- $T_{50} = 0.6$~с. Кресты~--- короткие/жесткие всплески, 
  круги~--- длинные/мягкие всплески, треугольники~--- промежуточные всплески. 
  Изображенное разделение сделано на основе $I_l$, но отличается от указанного в тексте для большей наглядности.
  Всплеск отнесён к кластеру $l$ если значение $I_l$ для этого кластера, 
  превышает значения для других кластеров.
  \label{img:HRvsT50}}
\end{figure}

\begin{figure} [h] 
  \center
  \includegraphics [width=0.75\textwidth] {gHRT50_Types_ru}
  \caption{Классификация 1143 ярких всплесков KW.
  Черные точки~--- Тип~I, круги~--- Тип~I/II, треугольники~--- Тип~II,
  см. описание методики классификации в тексте.
  } 
  \label{img:HRvsT50Types}  
\end{figure}

\begin{figure}[h]
  \begin{minipage}[h]{0.5\textwidth}
    \center{\includegraphics[width=1.0\textwidth]{list1104calHRpcr_2} \\ а)}
  \end{minipage}
  \hfill
  \begin{minipage}[h]{0.5\textwidth}
    \center{\includegraphics[width=1.0\textwidth]{list1104calHRpcr_3} \\ б)}
  \end{minipage}
  \caption[Аппроксимация распределения $\log T_{50}\textrm{--}\log \mbox{HR}_{32\rmn{pk}}$.]
  {Аппроксимация распределения $\log T_{50}\textrm{--}\log \mbox{HR}_{32\rmn{pk}}$ 
  для 1104 ярких всплесков ($\rmn{S/N} \geq 10$) двумя~(a) и тремя~(б) гауссовыми компонентами. 
  Эллипсами для каждого распределения отмечены области $1\sigma$ и~$3\sigma$. 
  Пунктирная вертикальная линия~--- $T_{50} = 0.6$~с. Кресты~--- короткие/жесткие всплески, 
  круги~--- длинные/мягкие всплески, треугольники~--- промежуточные всплески.
  Изображенное разделение сделано на основе $I_l$, но отличается от указанного в тексте для большей наглядности.
  Всплеск отнесён к кластеру $l$ если значение $I_l$ для этого кластера, 
  превышает значения для других кластеров.
  }
  \label{img:HRpkvsT50}  
\end{figure}

\FloatBarrier

\section{Спектральные задержки}\label{sec:Lags}
Одной из характеристик спектральной эволюции гамма-всплесков является спектральная задержка.
Она представляет собой количественную характеристику запаздывания излучения в 
мягком спектральном диапазоне по сравнению с более жестким~\citep[см., например,][Глава~3]{Minaev_PhD}.
Обзор моделей, описывающих появление спектральной задержки, смотри там же.
В данном разделе спектральная задержка используется исключительно в роли дополнительного 
параметра классификации гамма-всплесков, без обсуждения физической причины её возникновения.

\subsection{Методика вычисления спектральных задержек для кривых блеска KW}
Для вычисления спектральных задержек всплесков KW использовался 
кросскорреляционный анализ временных историй~\citep{Band_1997ApJ, Norris_2000}. 
В этом методе задержка ($\tau_\rmn{lag}$) соответствует положению максимума 
кросс-корреляционной функции (ККФ) временных историй в различных каналах детектора.
Определение спектральной задержки между временными историями в двух диапазонах ($\tau_\rmn{lag}$) 
включало выбор разрешения временных историй, выбор интервала кросс-корреляции 
и вычисление $\tau_\rmn{lag}$ и её ошибки.

Временное разрешение выбралось таким образом, чтобы для временных историй 
в обоих диапазонах бин с наибольшей скоростью счета имел отношение сигнал-шум~$\rmn{S/N} \geq 8$. 
Порог $\rmn{S/N} \geq 8$ был выбран произвольно, чтобы исключить из рассмотрения 
слабые всплески с большими ошибками $\tau_\rmn{lag}$. Возможные значения разрешения 
временной истории составляли от 4~мс до 1024~мс.
В качестве интервала кросс-корреляции брался наиболее широкий интервал, ограниченный
бинами, в которых было обнаружено превышение $5\sigma$ над фоном хотя бы в одном из каналов.

Значение $\tau_\rmn{lag}$ и его ошибка определялись методом Монте-Карло на основе 
100 реализаций исходных кривых блеска. 
На каждой итерации генерировались искусственные временн\'{ы}е истории 
при помощи добавления пуассоновского шума к исходным данным.
На каждой итерации значения ККФ вычислялось аналогично формуле~(10) из~\citep{Band_1997ApJ}, 
ошибки ККФ вычислялись аналогично формуле~(5) из~\citep{Fenimore_1995}.
После построения ККФ производился поиск интервала, содержащего основной пик ККФ
при этом использовались только значения $>0.1$.  
После чего ККФ аппроксимировалась полиномом 4-й степени на выбранном интервале
и в качестве значения $\tau_\rmn{lag}$ брался максимум полинома.
Если $p$-значение аппроксимации (\textit{null hypothesis probability}) 
оказывалось меньше порогового значения 1\%, 
то из ККФ удалялись две крайние точки, и процедура повторялась до превышения порога. 
Если в результате уменьшения интервала аппроксимации число степеней свободы сокращалось до нуля,
то текущая итерация считалась неудачной. 
В случае, если итоговое число удачных итераций было больше $50$, то значение $\tau_\rmn{lag}$ 
вычислялось как среднее по всем удачным итерация, в качестве ошибки бралась дисперсия значений.
Иначе значение $\tau_\rmn{lag}$ считалось определенным ненадёжно и дальше не использовалось.
Примеры вычисления лага на одной из итераций представлены рис.~\ref{fig:lag_calculation}.

\begin{figure}[h]
  \center{\includegraphics[width=0.75\textwidth]{GRB20070512_T21972_lag_31}}
  \caption{Иллюстрация вычисления лага между кривыми блеска в G3 и G1 
   для короткого всплеска GRB20070512\_T21972 со значительной задержкой, $\tau_\rmn{lag}=277\pm16$~мс. 
  Вертикальные линии на нижней панели обозначают границы интервала кросс-корреляции. 
  Непрерывная линия на верхней панели показывает аппроксимацию ККФ полиномом 4-й степени. 
  Вертикальная пунктирная линия на верхней панели соответствует значению лага.
  \label{fig:lag_calculation}}  
\end{figure}

\subsection{Спектральные задержки коротких всплесков}
Спектральные задержки вычислялись между парами каналов G3--G1, G2--G1 и G3--G2, 
где $\tau_\rmn{lag}$ соответствует сдвигу временной истории в канале, указанном вторым
по отношению к каналу, указанному первым. Для коротких всплесков было получено
42, 66 и 158 задержек для пар каналов G3--G1, G2--G1 и G3--G2, соответственно.

На рис.~\ref{img:LagDistrs} представлены распределения коротких всплесков без EE и 
начальных импульсов коротких всплесков с EE по спектральным задержкам. 
Всплески Типа~I имеют, в основном, незначительные $\tau_\rmn{lag}\lesssim 25$~мс для 
пар каналов G3--G1 и G2--G1, при этом значительная часть всплесков типов~I/II и~II 
имеют задержки $\tau_\rmn{lag}\gtrsim 25$~мс для тех же пар каналов.

Среди коротких всплесков с продлённым излучением два события имеют существенные  % (ID=554, 1531)
спектральные задержки начального импульса в G3--G1 и были классифицированы Тип~Iee/II в предыдущем разделе. 
Таким образом, можно заключить, что значительная спектральная задержка подкрепляет классификацию 
на основе распределения на плоскости жёсткость-длительность.

\begin{figure}[h]
  \begin{minipage}[h]{0.5\textwidth}
    \center{\includegraphics[width=1.0\textwidth]{gDistLags} \\ а)}
  \end{minipage}
  \hfill
  \begin{minipage}[h]{0.5\textwidth}
    \center{\includegraphics[width=1.0\textwidth]{gDistLagsEE} \\ б)}
  \end{minipage}
  \caption{Распределение коротких всплесков без EE~(а) и начальных импульсов 
  всплесков с EE~(б) по спектральной задержке.
  На панели~(а), заполненная гистограмма~--- всплески Типа~I, заштрихованные 
  гистограммы~--- всплески Типа~I/II и~II.  На панели~(б), заполненная 
  гистограмма~--- всплески Типа~I, заштрихованные гистограммы~--- всплески Типа~Iee и~Iee/II.}
  \label{img:LagDistrs}  
\end{figure}

\FloatBarrier

\section{Сравнение классификаций на физические типы I и II всплесков KW 
с определенными красными смещениями}\label{sec:Phys_Classification}
Набор всплесков KW, зарегистрированных с ноября 1994~г. по июнь 2014~г., 
содержит 126 гамма-всплесков с измеренными красными смещениями~\citep{Tsvetkova_KW_GRBs_with_z}. 
Из них пять имеют сбои во временной истории, далее рассматривается набор из 121 всплесков. 
Указанный набор содержит 11 коротких и 110 длинных всплесков.

Короткие всплески KW имеют красные смещения в диапазоне 0.1--1.0, длинные~--- 0.1--5.0.
Для $\sim 50$\% длинных и практически всех коротких всплесков из набора были 
определены физические типы на основе данных о послесвечении и родительских 
галактиках~\citep{Zhang_2009, Kann_2010ApJ, Kann_2011ApJ}.
При этом стоит выделить два всплеска. 
Длинный всплеск GRB~060614 относят к коротким всплескам с продлённым излучением~\citep{Gehrels_2006_Nature}. 
Однако, по данным KW его нельзя отнести к этому классу, так как на основе полной 
длительности $T_{50} = 49 \pm 5$~с, длительности его начального импульса 
$T_{50} = 2.7\pm 0.3$~с и жесткости этот всплеск относится к типу~II.
Всплеск GRB~040924 можно отнести к коротким всплескам на основе длительности $T_{50} = 0.34\pm0.06$~c, 
при этом он имеет значительную спектральную задержку, $120 \pm 30$~мс между каналами G2 и G1, 
и является достаточно мягким. По данным KW этот всплеск классифицируется как Тип~II, 
так же как и по данным многоволновых наблюдений~\citep{Zhang_2009}.

Было также проанализировано распределение жёсткость-длительность в системе 
отсчёта источника всплеска. При этом $T_{50}$ уменьшается в $1+z$~раз, а жёсткость 
вычислялась в каналах, чьи границы соответствуют номинальным умноженным на $1/(1+z)$. 
Распределения на плоскости $\log T_{50}$--$\log \mbox{HR}_{32}$ в системе отсчёта наблюдателя 
и собственной системе отсчёта представлены на рис.~\ref{img:T50HRzCorr}. 
Для сравнения распределений по жесткости и длительности всплесков типов I и II
в системе наблюдателя и собственной системе был использован тест Колмогорова-Смирнова для двух наборов. 
Вероятности (p-значения) того, что жесткости или длительности 
121-го всплеска являются выборкой из общего распределения малы ($<1$\%) 
как в случае системы наблюдателя, так и источника. Хотя в последнем случае
граница между типами становится размытой, рассмотренные распределения существенно отличаются,
что свидетельствует различие параметров коротких/жестких и длинных/мягких всплесков как 
в системе наблюдателя, так и системе источника всплеска.

\begin{figure}[h]
  \begin{minipage}[h]{0.5\textwidth}
    \center{\includegraphics[width=1.0\textwidth]{gHRvsT50obs_ru} \\ а)}
  \end{minipage}
  \hfill
  \begin{minipage}[h]{0.5\textwidth}
    \center{\includegraphics[width=1.0\textwidth]{gHRvsT50src_ru} \\ б)}
  \end{minipage}
  \caption{Распределение всплесков на плоскости $\log T_{50}$--$\log \rmn{HR}_{32}$ 
  в системе отсчёта наблюдателя~(а) и источника~(б), с поправкой на космологическое красное смещение. 
  Круги~--- всплески типа~I, треугольники~--- всплески типа~II, 
  серые точки~--- набор 1143 ярких всплесков.
  }
  \label{img:T50HRzCorr}  
\end{figure}

\section{Заключение} \label{sec:Conclision}
В данной главе описана методика классификации всплесков, зарегистрированных в эксперименте
Конус-Винд и получены следующие результаты:

\begin{enumerate}
\item Для набора 1834 всплесков KW были вычислены длительности $T_{50}$ и $T_{90}$, жесткости 
и спектральные задержки. Показано, что распределения 
всплесков по $T_{50}$ и $T_{90}$ хорошо аппроксимируются двумя логнормальными 
распределениями. Обнаружено, что параметры аппроксимации распределения $T_{50}$ 
более устойчивы к выбору порога поиска начала и конца всплеска, поэтому длительность 
$T_{50}$ более предпочтительна для классификации всплесков. В качестве границы между 
длинными и короткими всплесками была выбрана точка пересечения логнормальных компонент 
для порога значимости $5\sigma$, $T_{50} = 0.6$~с.

\item Выбран набор коротких всплесков (без продлённого излучения), включающий 265 событий, 
что составляет 14\% от общего числа проанализированных всплесков. Для сравнения, 
доля коротких ($T_{90}<2$~с) всплесков в каталоге BATSE равна 24\%~\citep{Meegan_2001}, %(=500/2014)
в каталоге \textit{Fermi}-GBM 18\%~\citep{Paciesas_2012}, 
в каталоге \textit{Swift}-BAT 8\%~\citep{Sakamoto_2011ApJS}. 
Доля коротких всплесков, оцененная из площадей гауссовых компонент 
распределения $T_{50}$, равна 30\%. Соответствующая доля в наборе всплесков BATSE равна 32\%~\citep{Horvath_2002} 
и 8\% в наборе всплесков \textit{Swift}-BAT~\citep{Horvath_2008}.

\item Обнаружен 31 всплеск, который можно классифицировать как короткий 
всплеск с продлённым излучением. Таким образом, доля всплесков с продлённым 
излучением среди коротких всплесков составляет 10\%. Соответствующая доля 
в выборке \textit{Swift}-BAT равна 20\%~\citep{Sakamoto_2011ApJS}. 
Спектральный анализ продлённого излучения представлен в главе~\ref{sGRB_spectral_catalog}.

\item Аппроксимация распределения всплесков на плоскости $\log T_{50}$--$\log \rmn{HR}_{32}$ 
набором гауссовых компонент показала наличие двух классов всплесков, коротких/жестких 
и длинных/мягких. При этом доля коротких/жестких всплесков, определённая на основе 
аппроксимации, составляет 21\%. По данным  BATSE доля коротких/жестких всплесков 
составляет 28\%~\citep{Horvath_2002}.  
Добавление третьей компоненты даёт значимое улучшения аппроксимации, однако эта 
компонента существенно перекрывается с компонентой, описывающей длинные всплески, 
и не представляет физического смысла. Дополнительный довод в пользу использования
только двух классов всплесков связан с тем, что для описания распределений 
по $T_{50}$ и $\rmn{HR}_{32}$ достаточно только двух компонент.
На основании полученной аппроксимации 7\% всплесков с $T_{50} < 0.6$~с относятся к типу~II.  (0.07=18/260)
При этом к типу~I относятся только всплески с $T_{50} < 0.6$~с 
(исключение составляет жесткий всплеск GRB20061006\_T31422 с пограничной  длительностью $T_{50} = 0.620\pm 0.049$~с, 
который не входит в набор коротких всплесков). Неопределенную классификацию (I/II) имеют 
около 4\% всплесков с длительностями $T_{50}$ от 0.04~с до 1.65~с.

На основании аппроксимации распределения $\log T_{50}$--$\log \rmn{HR}_{32}$ 
для 1143 всплесков, был классифицирован набор 265 коротких всплесков без продлённого излучения. 
Набор включает $\sim 70$\% всплесков Типа~I, $\sim 8$\% Типа~II и $\sim 12$\%
всплесков неопределённого типа (I или~II). Доля коротких всплесков с продлённым 
излучением составляет $\sim 10$\%.
Среди начальных импульсов всплесков, отнесённых на основе морфологии временной 
истории к коротким всплескам с продлённым излучением, 21 (68\%) классифицированы как Тип~I 
7 как неопределённый тип (I/II) и 3 как Тип~II.

\item Анализ спектральных задержек коротких всплесков показал, что большинство коротких всплесков 
Типа~I имеют незначительную по абсолютному значению спектральную задержку $\tau_\rmn{lag} \lesssim 25$~мс, 
в то же время значительная доля всплески типов~II и I/II имеет $\tau_\rmn{lag} > 25$~мс. 
Два всплеска с продленным излучением имеют задержку начального импульса $> 100$~мс
и были классифицированы как тип~II, что свидетельствует о том, что
эти всплески относятся к популяции длинных всплесков.

\item Сравнение классификаций на физические типы~I и~II с классификацией на основе 
длительности, жесткости и спектральной задержки подтвердило, что всплески Типа~I 
относятся к коротким/жестким всплескам с малой спектральной задержкой, а всплески 
Типа~II, в основном,~--- длинные мягкие с заметной спектральной задержкой. 
Сравнение распределений $\log T_{50}$--$\log \rmn{HR}_{32}$ в системе отсчёта наблюдателя 
и в собственной системе отсчёта показывает, что различие в жесткости и длительности
всплесков типа~I и~II становится менее значимым, но сохраняется.
\end{enumerate}

Для дальнейшего анализа выбран полный набор 296 коротких всплесков. 
Различия между короткими всплесками разных типов будут подробно исследованы на
основе спектрального анализа всплесков в главе~\ref{sGRB_spectral_catalog}.

По материалам Главы~\ref{KW_GRB_classification} на защиту выносится следующее положение:
\begin{itemize}
\item Метод классификации гамма-всплесков по данным эксперимента Конус-Винд на основе
    длительности и жесткости излучения всплеска, а также величин спектральных задержек.
\end{itemize}

\clearpage

### 2025-10-23 disser/part3.tex
disser

\chapter{Локализация источников гамма-всплесков методом триангуляции} \label{IPN_catalog}
\section{Введение}
Для каждого всплеска из набора 296 коротких гамма-всплесков, рассмотренного в 
предыдущей главе, был произведён поиск детектирования на КА, входящих в межпланетную 
сеть Interplanetary Network (IPN). Было обнаружено, что 271 ($\sim 92$\%) коротких 
всплесков Конус-Винд (KW) были зарегистрированы по крайней мере одним КА IPN, 
что позволило получить их локализацию триангуляционным методом.

В период с ноября 1994~г. по декабрь 2010~г. IPN содержала от 3-х до 11 КА. 
За все время существования IPN туда, помимо KW, входили следующие аппараты
на большом удалении от Земли: 
\begin{itemize}
\item \textit{Ulysses}, находившийся на гелиоцентрической орбите расстоянии 
670--3180 световых секунды от Земли, с инструментом для изучения рентгеновского 
излучения Солнца и гамма-всплесков GRB~\citep{Hurley_1992AAS};
\item \textit{Near-Earth Asteroid Rendezvous} (\textit{NEAR}), находившийся 
на расстоянии до 3100 световых секунд от Земли, с рентгеновским/гамма-спектрометром XGRS~\citep{Trombka_1999NIMPA};
\item \textit{Mars Odyssey}, запущенный в апреле 2001~г и достигший орбиты вокруг 
Марса в октябре 2001~г на расстоянии до 1250 световых секунд от Земли~\citep{Saunders_2004SSRv}, 
КА оборудован гамма-спектрометром GRS, в состав которого входят два детектора 
с возможностью регистрировать гамма-всплески: гамма-детектор GSH и детектор 
высокоэнергичных нейтронов HEND~\citep{Boynton_2004SSRv, Hurley_2006ApJS_MO};
\item \textit{Mercury Surface, Space Environment, Geochemistry, and Ranging} (\textit{MESSENGER}) 
со спектрометром гамма-излучения и нейтронов GRNS~\citep{Goldsten_2007SSRv}, 
запущенный в августе 2004~г и вышедший на орбиту вокруг Меркурия в марте 2011~г 
на расстоянии до 700 световых секунд от Земли, полное функционирование КА 
началось в 2007~г~\citep{Gold_2001PSS,Solomon_2007SSRv};
\item \textit{International Gamma-Ray Astrophysics Laboratory} (\textit{INTEGRAL}), 
где в качестве детектора гамма излучения выступает защита (ACS) спектрометра 
SPI (SPI-ACS)~\citep{Rau_2005AA}, КА находится на вытянутой орбите с максимальным 
удалением до 0.5 световых секунд от Земли;
\end{itemize}

на околоземных орбитах:
\begin{itemize}
\item \textit{Compton Gamma-Ray Observatory} (\textit{CGRO}) с экспериментом 
Burst and Transient Source Experiment (BATSE)~\citep{Fishman_1992NASCP3137};
\item \textit{BeppoSAX} с экспериментом Gamma-Ray Burst Monitor (GRBM)~\citep{Frontera_1997AAS,Feroci_1997SPIE};
\item \textit{Reuven Ramaty High Energy Solar Spectroscopic Imager} (\textit{RHESSI})~\citep{Lin_2002SoPh, Smith_2002SoPh};
\item \textit{High Energy Transient Explorer} (\textit{HETE-2}) с телескопом 
French Gamma-Ray Telescope (FREGATE)~\citep{Ricker_2003AIPC, Atteia_2003AIPC};
\item \textit{Swift} с телескопом Burst Alert Telescope (BAT)~\citep{Barthelmy_2005SSRv,Gehrels_2004ApJ};
\item \textit{Suzaku} с телескопом Wide-band All-sky Monitor (WAM)~\citep{Yamaoka_2009PASJ,Takahashi_2007PASJ};
\item \textit{AGILE} с инструментами Mini-Calorimeter (MCAL) и Super-AGILE~\citep{Tavani_2009AA};
\item \textit{Fermi} с иструментом Gamma-Ray Burst Monitor (GBM)~\citep{Meegan_2009ApJ};
\item Солнечная обсерватория \textit{Коронас-Ф} с гамма-спектрометром Геликон~\citep{Oraevskii_2002PhyU};
\item КА \textit{Космос~2326} с гамма-спектрометром Конус-А~\citep{Aptekar_1998ApJ};
\item КА \textit{Космос~2367} с гамма-спектрометром Конус-А2;
\item КА \textit{Космос~2421} с гамма-спектрометром Конус-А3 и
\item Солнечная обсерватория \textit{Коронас-Фотон} с гамма-спектрометром Конус-РФ.
\end{itemize}

По крайней мере два других КА наблюдали гамма-всплески в рассматриваемый период:
\textit{Defense Meteorological Satellite Program}
(\textit{DMSP})~\citep{Terrell_1998AIPC,Terrell_1996AIPC,Terrell_2004AIPC} и 
\textit{Stretched Rohini Satellite Series} (\textit{SROSS})~\citep{Marar_1994AA}.
Однако они не использовались для триангуляции, поэтому они не относятся к IPN.

Далее представлена методика локализации и результаты, полученные для 271 короткого 
всплеска Конус-Винд, детектированных по крайней мере ещё одним КА IPN.

\section{Наблюдения}\label{sec:IPN_catalog_obserations}
Для каждого короткого всплеска Конус-Винд производился поиск в данных КА сети IPN. 
Для околоземных КА и \textit{INTEGRAL} временн\'{о}е окно поиска было центрировано 
на времени срабатывания триггера KW, ширина окна соответствовала расстоянию, 
немного превышающему расстояние от Земли до \textit{Wind}. Для КА в межпланетном 
пространстве ширина окна поиска соответствовала удвоенному расстоянию до КА, если 
направление прихода излучения было неизвестно, что имело место для большинства событий. 
Если направление прихода было известно хотя бы грубо, то вычислялось ожидаемое время 
прихода излучения на КА и производился поиск вблизи этого времени.

Временные интервалы существования различных КА в IPN и число коротких всплесков 
KW, зарегистрированных каждым КА/инструментом показаны на рис.~\ref{img:Mission_timeline}. 
Наибольшее число зарегистрированных всплесков 139, после KW, 
было зарегистрировано \textit{INTEGRAL}~(SPI-ACS).

\begin{figure}[h]
    \center{\includegraphics[width=1.0\textwidth]{Mission_timeline}}
    \caption[Время работы различных КА IPN]
  {Левая панель: сроки работы инструментов IPN с момента запуска \textit{Wind} в ноябре 1994 года 
  (имена инструментов приведены в скобках). Правая панель: число коротких всплесков KW,
  детектированных каждым инструментом (для KW~--- количество всплесков, 
  детектированных по меньшей мере одним другим КА IPN).}
  \label{img:Mission_timeline}  
\end{figure}

Данные о 271-м коротком всплеске KW представлены в таблице~3.1, размещённой 
на сайте ФТИ им. А.Ф.~Иоффе по 
ссылке~\url{http://www.ioffe.ru/LEA/shortGRBs/Catalog2/Data/tables/}.

В первой колонке приведено обозначение всплеска, <<GRBYYYYMMDD\_Tsssss>>, 
где YYYYMMDD~--- дата всплеска и sssss~--- время триггера KW в секундах UT,
округлённое до целых секунд (из-за значительного удаления \textit{Wind} от Земли
это время может отличаться на $\sim 6$~c от времени, поправленного на распространения 
сигнала от KW до центра Земли). 
Колонки (2)--(4) содержат год, месяц и число, когда произошел всплеск. 
Колонки~(5)--(7) содержат часы, минуты и секунды триггерного времени~KW.
Колонка (8) содержит Тип всплеска определенный в предыдущей главе. 
Типы имеют значения: I (результат слияния компактных объектов), 
II (результат коллапса массивной звезды), I/II (неопредёленный тип), Iee (тип I с продлённым излучением)
и Iee/II (тип не определён: Iee или~II). Колонка~(9) содержит время задержки на распространение
фронта излучения от  \textit{Wind} до центра Земли, колонки (10) и (11) содержат
$3\sigma$ ошибки этого значения (Задержки вычислены с использованием IPN локализаций приведённых ниже). 
Колонка (12) содержит названия миссий/инструментов, которые наблюдали всплеск
(также отмечены наблюдения инструментами, которые не являются часть IPN, а именно:
COMPTEL на \textit{CGRO}, DMSP, \textit{Fermi} LAT, MAXI (Monitor of All-sky X-ray Image) и SROSS)
Колонки (13)--(14) содержат полное число КА IPN и число дальних КА IPN, наблюдавших всплеск.
Последняя колонка содержит комментарий, если есть. Описание таблицы также приведено по указанной выше ссылке.

За рассматриваемый период четыре КА в межпланетном пространстве входили в состав IPN: 
\textit{Ulysses}, \textit{NEAR}, \textit{Mars Odyssey} и \textit{MESSENGER}. 
Из 271 всплеска, 30 наблюдались двумя из перечисленных КА, 102~--- одним, 
139~--- не наблюдались ни одним из указанных КА.

Семь коротких всплесков Конус-Винд были точно локализованы инструментами, 
способными строить изображения в рентгеновском или мягком гамма-диапазоне, 
а именно \textit{Swift}-BAT, \textit{HETE-2} (WXM/SXC) и \textit{INTEGRAL} (IBIS/ISGRI). 
Для большинства этих всплесков было зарегистрировано рентгеновское послесвечение; 
для некоторых из них было определено космологическое красное смещение источника 
по наблюдениям оптического послесвечения или спектроскопии родительской галактики. 
Эти всплески были использованы для проверки используемого метода триангуляции.
\FloatBarrier
\section{Методика триангуляции}
При регистрации всплеска на двух КА с временной задержкой $\delta T$, для него 
может быть построена область локализации в виде кольца на небесной сфере с углом 
раствора $\theta$ относительно вектора, соединяющего два КА. Значение угла $\theta$ определяется выражением
\begin{equation}
\cos \theta = \frac{c \delta T}{D} \mbox{ ,}
\end{equation}
где $c$~--- скорость света и $D$~--- расстояние между КА. Здесь подразумевается, 
что всплеск представляет собой плоскую волну, т.~е. расстояние до источника много больше $D$.

Измеряемая временная задержка имеет ошибку, которая в общем случае 
не симметричная $d_{\pm}(\delta T)$, т.~е. измеренная временная задержка имеет 
доверительный интервал от $\delta T + d_{-}(\delta T)$ 
до $\delta T + d_{+}(\delta T)$ ($d_{-}(\delta T)$~--- отрицательно) на заданном уровне значимости.

Полуширина кольца $d\theta_{\pm}$ определяется выражением
\begin{equation}\label{eq:CCWidth}
d\theta_{\pm} \equiv \theta_{\pm} -\theta = 
\arccos \left[ \frac{c (\delta T + d_{\mp}(\delta T))}{D} \right] - \arccos\left[ \frac{с \delta T}{D} \right]\mbox{ .}
\end{equation}

Следует отметить, что даже в случае симметричных ошибок $|d_{-}(\delta T)| = d_{+}(\delta T)$, 
кольцо может быть существенно несимметрично если $c (\delta T + d_{\pm}(\delta T))/D \sim 1$ 
(т.~е. направление на источник близк\'{о} к вектору, соединяющему два КА).

В случае $d_{\pm}(\delta T) \ll D/c$ уравнение \ref{eq:CCWidth} переходит в часто используемое выражение
\begin{equation}\label{eq:CCWidthRed}
d \theta_{\pm} = - \frac{c d_{\mp}(\delta T)}{D\sin \theta} \mbox{ .}
\end{equation}

Для вычисление наиболее вероятной временной задержки и её доверительного интервала 
был использован метод минимизации $\chi^2$, описанный в~\citep{Hurley_1999ApJSa} 
для триангуляции с дальними КА и этот же метод с некоторыми изменениями был 
использован для триангуляции с использованием KW и околоземных КА (или \textit{INTEGRAL}).

Наиболее вероятная временная задержка $\tau$ и её ошибка $d_{\pm}\tau$ между 
временными историями, записанными двумя инструментами вычислялась следующим образом. 
Пусть $n_{1,i} = n(t_{1,i})$, $n_{2,j} = n(t_{2,j})$ и 
$\sigma_{1,i}$, $\sigma_{2,j}$ обозначают числа отсчётов с вычетом фона 
и их ошибки в последовательных равномерных временных интервалах (бинах) 
$t_{1,i} = t_{1,0} + i\Delta_{1}$, $t_{2,j} = t_{2,0} + j\Delta_{2}$, 
где $i = 0,\dotsc,m_1$, $j = 0,\dotsc,m_2$; $\Delta_{1}$, 
$\Delta_{2}$~--- размеры бинов и $t_{1,0}$, $t_{2,0}$~--- времена привязки для каждого КА по всемирному времени (UT).

Для простоты будем считать, что $\Delta_1 = \Delta_2 = \Delta$ и что отсчёты детекторов 
подчиняются статистике Пуассона $\sigma_{1(2),i} = n_{\rmn{tot }1(2), i}^{1/2}$, 
где $n_{\rmn{tot }1(2), i}$~--- полное число отсчётов (источник + фон) в бине $i$. 
Предполагается, что обе временные истории содержат интересующий нас всплеск и интервалы до и после него 
(если эти интервалы отсутствуют они всегда могут быть заполнены нулями, 
а в качестве $\sigma_{1(2),i}$ взято стандартное отклонение числа отсчётов фона).
Также предполагается, что в первой временной истории $N+1$ бин, начиная 
с $i_\rmn{start}$ содержат всплеск или участок всплеска, 
который кросскоррелируется во второй временной истории. С учётом этих предположений 
можно сконструировать статистику:
\begin{equation}
R^2(\tau \equiv k\Delta) =  
\sum_{i=i_\rmn{start}}^{i=i_\rmn{start}+N} 
\frac{(n_{2,i} - s n_{1,i+k})^2}{(\sigma^2_{2,i} - s^2 \sigma_{1,i+k})} \mbox{ ,}
\end{equation}
где $s$~--- масштабный множитель являющийся отношением полного числа отсчётов, 
зарегистрированных инструментами $s = \sum_i n_{1,i} / \sum_j n_{2,j}$. 
Для идеального случая одинаковых детекторов с одинаковыми энергетическими диапазонами 
и углами падения излучения, и пуассоновской статистики отсчётов, $R^2$ распределена 
как $\chi^2$ с $N$ степенями свободы. В реальности существует несколько сложностей. 
Детекторы имеют разные энергетические диапазоны, разные аппаратные функции и работают 
в условиях с различным поведением фоновой скорости счёта (переменный фон на околоземных орбитах). 
Для коротких гамма-всплесков часть этих факторов оказывают незначительное влияние: 
вариации фона на малых временных масштабах малы, спектральная эволюция, которая 
приводит к значительному сдвигу между временными историями в различных диапазонах, 
практически отсутствует у коротких всплесков~\citep{Norris_2001grba}.

Для учёта всех отличий от идеального случая (учёта неопределённости вычисления $\tau$) 
был применён следующий метод: 
для заданного $N$ (числа бинов, используемых для построения $R^2$) вычислялось значение $\chi^2(N)$, 
соответствующие уровню значимости $3\sigma$ (соответствующие вероятности $Q(\chi^2|N)=2.7\times 10^{-3}$), 
на основе которого вычислялся соответствующий $3\sigma$ уровень значимости для приведённого $R^2_r(\equiv R^2/N)$, равный
\begin{equation}\label{eq:R3sigma}
R^2_{r,3\sigma} = \chi^2_{r,3\sigma} + R^2_{r,\rmn{min}} - 1 \mbox{ ,}
\end{equation}
где $R^2_{r,\rmn{min}}$~--- минимум $R^2_{r}(\tau)$. 
Вычитание 1 в формуле~\ref{eq:R3sigma} связано с тем, 
что $R^2_{r,\rmn{min}}\sim 1$ для идеального случая, однако, часто на практике $R^2_{r,\rmn{min}}> 1$ 
и следовательно $R^2_{r,3\sigma} > \chi^2_{r,3\sigma}$. Для определения $3\sigma$ 
доверительного интервала для $\tau$ используются ближайшие точки кривой  $R^2_{r}(\tau)$, 
лежащие выше уровня $3\sigma$ полученного из выражения~\ref{eq:R3sigma} (см. примеры на Рис.~\ref{img:CC_examples}). 
После определения кросскорреляционной задержки $\tau$ и её ошибок $d_{\pm}(\tau)$ 
может быть вычислена временная задержка $\delta T = t_{02} - t_{01} + \tau$; 
$d_{\pm}(\delta T) = d_{\pm}(\tau)$ (здесь предполагается, что абсолютные времена $t_{01}$ 
и $t_{02}$ определены точно). Далее для простоты будем называть $R^2$ как $\chi^2$.

\begin{figure}[h]
    \center{\includegraphics[width=0.75\textwidth]{gCCall}}
    \caption[Примеры кросскорреляций временных историй]
  {Примеры кросскорреляционных кривых $\chi^2_{r}(\tau)$. 
  Горизонтальные красные линии обозначают уровни $3\sigma$. 
  Вертикальные синие линии показывают наиболее вероятную кросскорреляционную задержку $\tau$ 
  (штрих-пунктирная линия) и её $3\sigma$ доверительный интервал (пунктирные линии).
  Вверху слева: GRB19971118\_T29008. Кросскорреляция временной истории KW (2~мс) с временной историей BATSE (64~мс),
  $\tau = 0.060 (-0.034,+0.012)$~с ($\rmn{dof} = 2$). 
  Вверху справа: GRB20070321\_T67937. Кросскорреляция временной истории KW (2~мс) с временной историей WAM (1/64~с),
  $\tau = 0.075 (-0.008,+0.016)$~с ($\rmn{dof} = 12$). 
  Внизу слева: GRB20090715\_T62736. Кросскорреляция временной истории KW (2~мс) с временной историей SPI-ACS (50~мс),
  $\tau = 0.120 (-0.016,+0,052)$~с ($\rmn{dof} = 7$). 
  Внизу справа: GRB20100206\_T48606. Кросскорреляция временной истории GBM (1~мс)с временной историей KW (2~мс),
  $\tau = -0.033 \pm 0.012$~с ($\rmn{dof} = 9$).
  }
  \label{img:CC_examples}  
\end{figure}
\FloatBarrier

\section{Триангуляционные кольца}
Используя приведённую выше методику для 271 короткого всплеска Конус-Винд было 
получено одно или более триангуляционное кольцо. Обсуждение деталей получения 
временных задержек для различных пар КА приведены в нижеследующих разделах.

\subsection{Кольца, полученные с использованием дальних КА}
Дальние КА (КА в межпланетном пространстве) играют важную роль в триангуляции 
гамма-всплесков. Их длинная база позволяет получать малые области локализации 
для большого числа всплесков. Однако детекторы на этих КА обычно меньше, чем на 
околоземных, при этом часть детекторов предназначены для планетарных исследований 
с возможностью регистрации гамма-всплесков. Эти детекторы могут иметь более 
грубое временное разрешение и меньшую чувствительность. Также часы на этих КА 
не всегда калиброваны по Всемирному координированному времени (UTC) настолько точно, 
насколько часы на околоземных КА (или их калибровка не может быть определена настолько аккуратно).

Для локализации использовались данные четырёх межпланетных КА: \textit{Ulysses}, 
\textit{NEAR}, \textit{Mars Odyssey} и \textit{MESSENGER}. Из них только \textit{Ulysses} 
имел эксперимент, посвященный гамма-всплескам. Временное разрешение этих четырёх 
экспериментов составляло от 32~мс (триггерный режим \textit{Ulysses}) 
до 1~с (\textit{MESSENGER}, \textit{NEAR}). При регистрации короткого всплеска 
детектором с разрешением, намного превышающим длительность всплеска, обычно 
наблюдается превышение скорости счёта в одном бине, при этом неопределённость 
временной задержки составляет половину от наибольшего временного разрешения. 
Точность часов КА определялась двумя способами. 
В случае \textit{Ulysses}, эксперименту, регистрирующему гамма-всплески, 
в точно известные моменты времени посылались определённые команды. 
Учитывая аберрационное время и задержки выполнения команд на борту КА, 
можно было уточнить временную привязку с точностью от нескольких миллисекунд 
до 125~мс. 
%(хотя точность привязки предполагалась равной нескольким миллисекундам, 
%зачастую технические сложности не давали возможность её проверить). 
Дополнительно точность временной привязки межпланетных КА может быть проверена 
триангуляцией известных источников, чьё положение хорошо известно из других измерений: 
это могут быть как источники мягких повторяющихся гамма-всплесков 
(другие употребляемые названия этих объектов: мягкие гамма-репитеры; SGRs), 
так и гамма-всплески локализованные \textit{Swift}-XRT 
или \textit{Swift}-UVOT. При вычислении кросскорреляционной задержки коротких всплесков
использовалась консервативная оценка ошибок, считалось, 
что ошибка $\tau$ на уровне $3\sigma$ не может быть меньше 125~мс.

Всего 132 коротких всплеска KW наблюдались дальними КА: 30 наблюдались 
двумя дальними КА и 102 одним дальним КА. Среди них девять были точно локализованы 
инструментами, способными строить изображения в рентгеновском или мягком гамма-диапазоне. 
Без учёта этих всплесков было получено 150 колец. Распределение $3\sigma$ полуширин 150 колец 
представлено на Рис.~\ref{img:dist_150}. Наименьшая полуширина $0\fdg0024$ (0\farcm14), 
наибольшая $2\fdg21$, средняя $0\fdg099$ ($5\farcm9$), геометрическое 
среднее $0\fdg028$ ($1\farcm7$).

\begin{figure}[h]
    \center{\includegraphics[width=0.5\textwidth]{dN_150_annuli_distant}}
    \caption[Распределение ширин колец, построенных с использованием дальних КА]
    {Распределение $3\sigma$ полуширин 150 триангуляционных колец, полученных с 
    использованием данных дальних КА.}
 \label{img:dist_150}  
\end{figure}
\FloatBarrier
\subsection{Кольца, полученные с использованием KW, \textit{INTEGRAL} и околоземных КА}
KW занимает особое место в IPN благодаря уникальному набору характеристик: 
непрерывному обзору всего неба двумя спектрометрами, положению в межпланетном 
пространстве в условиях исключительно стабильного фона, широкому энергетическому 
диапазону (10~кэВ--10~МэВ номинальный; $\sim 20$~кэВ--15~МэВ в 2010~г.) и достаточно 
высокой чувствительности ($\sim 10^{-7}$~эрг~см$^{-2}$). Доля времени наблюдения 
KW, отнесённая ко всему времени работы, составляет примерно 95\%. 
Эксперимент зарегистрировал большую часть событий IPN, являясь важным компонентом 
IPN на расстоянии $\simeq 1\textrm{--}7$ световых секунды (см. Рис.~\ref{img:Wind_distance}).

\begin{figure}[h]
    \center{\includegraphics[width=0.5\textwidth]{Wind_distance}}
    \caption[Расстояние \textit{Wind} от Земли]
    {Зависимость расстояния \textit{Wind} от Земли от времени.
     Максимальное расстояние составляло 7 световых секунд с января по май 2002 г., 
     когда КА находился на далекой орбите (distant prograde orbit, DPO).
     С 2004~г. \textit{Wind} находится на орбите Лиссажу вокруг точки либрации $L_1$
     в системе Солнце-Земля на расстоянии 5 световых секунд.}
 \label{img:Wind_distance}  
\end{figure}

В триггерном режиме временное разрешение KW составляет 2~мс на интервале 
от $T_0-0.512$~с до $T_0+0.512$~с ($T_0$~--- время срабатывания триггера), 
который покрывает, в большинстве случаев, весь короткий всплеск или, по крайней 
мере, его наиболее интенсивную часть, позволяя производить точную кросс-корреляцию 
с временными историями других инструментов. Погрешность часов KW составляет 
менее 1~мс и их точность была проверена триангуляцией всплесков от SGRs 
и точно локализованных гамма-всплесков.

Наибольшую точность кросс-корреляции с KW (наименьшие неопределённости 
времени задержки) дают околоземные КА с большими эффективными площадями, 
а именно: \textit{CGRO}~(BATSE), \textit{BeppoSAX}~(GRBM), \textit{INTEGRAL}~(SPI-ACS), 
\textit{Suzaku}~(WAM), \textit{Swift}~(BAT), и \textit{Fermi}~(GBM). 
В настоящее время кросс-корреляции с \textit{Fermi}~(GBM) обычно даёт наилучший 
результат (наиболее узкое кольцо) благодаря схожим детекторам (сцинтилляционные 
спектрометры на основе NaI(Tl)), большой эффективной площади GBM 
(несколько сотен см$^2$ при использовании нескольких детекторов) и временной 
привязке каждого фотона в 128 энергетических каналах, что позволяет получать 
временную историю GBM с любым временным разрешением и в том же спектральном диапазоне, 
что у KW.

Так как часы на большинстве околоземных КА очень точные, высокая статистика 
отсчётов в сумме с высоким временным разрешением дают ошибки временных задержек 
вплоть до нескольких миллисекунд. Таким образом, несмотря на достаточно небольшое 
расстояние между околоземными КА и KW в несколько световых секунд, 
получаемая относительная ошибка временной задержки ($c d_{\pm}(\delta T)/D$), 
которая определяет ширину кольца (см. Уравнения~\ref{eq:CCWidth} и~\ref{eq:CCWidthRed}) 
может быть сравнима или даже меньше, чем для кольца с дальним КА. Подобные малые ошибки, 
порядка нескольких миллисекунд, и следовательно узкие кольца, могут быть получены 
для коротких всплесков с острым пиком или быстрым нарастанием/спадом. 
С другой стороны, всплески с плавными импульсами дают достаточно большие ошибки 
времени задержки, и следовательно более широкие кольца.

Для KA \textit{Wind}, \textit{INTEGRAL} и околоземных аппаратов, неопределённости эфемерид 
незначительны по сравнению с ошибками времён задержки и поэтому они не учитываются 
при построении триангуляционных колец.

Всего было получено 356 колец для Конус-Винд и околоземных КА, и Конус-Винд и 
\textit{INTEGRAL}. На Рис.~\ref{img:dist_356} представлено распределение ошибок временных 
задержек и $3\sigma$ полуширин этих колец. Наименьшая неопределённость времени 
задержки составляет 2~мс, наибольшая -- 504~мс, средняя -- 43~мс и средняя 
геометрическая~--- 23~мс. Наименьшая $3\sigma$ полуширина кольца составляет 
$0\fdg027$~($1\farcm6$), наибольшая~--- $32\fdg2$, 
средняя~--- $1\fdg3$ и средняя геометрическая~--- $0\fdg43$.

В последующих подразделах приведены некоторые детали триангуляции с использованием 
KW и \textit{INTEGRAL} и околоземных КА.

\begin{figure}[h]
    \center{\includegraphics[width=0.75\textwidth]{dN_356_annuli}}
    \caption[Распределение ширин колец, построенных с использованием околоземных КА]
    {Распределение неопределенности времени задержки $d(\delta T) \equiv (\delta T_{+} + |\delta T_{-}|)/2$
    и $3\sigma$ полуширины для 356 триангуляционных колец, полученных с использованием данных
    KW и околоземного КА (или INTEGRAL).}
 \label{img:dist_356}  
\end{figure}
\FloatBarrier
\subsubsection{Триангуляции KW--\textit{CGRO}~(BATSE)}\label{sec:KW_BATSE_cc}
Эксперимент BATSE был установлен на обсерватории имени Комптона и предназначен 
для исследований в области астрофизики высоких энергий~\citep{Fishman_1992NASCP3137}. 
Его детекторы большой площади (Large Area Detectors) записывали временные истории 
гамма-всплесков в четырёх энергетических диапазонах: Ch1, Ch2, Ch3, Ch4 с номинальными 
границами каналов: 25--55~кэВ, 55--110~кэВ, 110--320~кэВ и $>320$~кэВ. 
Часы на борту \textit{CGRO} имели точность 100~мкс, которая проверялась при помощи 
тайминга пульсаров. Бортовое программное обеспечение увеличивало эту ошибку, 
давая неопределённость в триггерных временах BATSE до $\simeq 1$~мс.

BATSE зарегистрировал 52 коротких всплеска Конус-Винд: 44 в триггерном режиме 
и 8 в режиме реального времени (real-time mode), в котором ведётся непрерывная 
запись скорости счёта с разрешением 0.25, 0.5, 1 или 2~с в зависимости от скорости 
передачи информации. Триангуляционные кольца были получены для 44 всплесков, 
зарегистрированных в триггерном режиме, и 6 всплесков, зарегистрированных в 
режиме реального времени (эти всплески наблюдались только Конус-Винд и BATSE).

Для кросс-корреляции с триггерными всплесками BATSE использовались временные 
истории KW в диапазонах G2+G3 или G2 с временным разрешением 2 или 16~мс 
и объединённые временные истории BATSE (объединение типов данных DISCLA, PREB 
и DISCS~\citep{Fishman_1992NASCP3137}) в диапазонах Ch2+Ch3+Ch4 или Ch2+Ch3 
с временным разрешением 64~мс. Для нескольких всплесков такие временные истории 
были недоступны и были использованы другие типы данных BATSE. Обычно делались 
кросс-корреляции для различных комбинаций каналов для проверки согласия 
получаемых временных задержек и выбиралась та, для которой $\chi^2$ был наименьший. 
Кросскорреляционные кривые для различных диапазонов могут быть сдвинуты относите
льно друг друга (на несколько миллисекунд), но $3\sigma$ интервалы для 
кросскорреляционной задержки $\tau$ всегда согласуются хорошо.

Полученные $\chi^2_{r,\rmn{min}}$ находятся в диапазоне от 0.06 до 4.51 со 
средним 0.81. Максимальное $\chi^2_{r,\rmn{min}}=4.51$~(dof=6)~--- явный выброс 
в распределении всплесков по $\chi^2_{r,\rmn{min}}$.

Это значение соответствует особенно сильному всплеску GRB19970704\_T04097 
(триггер BATSE \#6293) с пиковой скоростью счёта $1.8\times10^5$~отсчётов~с$^{-1}$ 
на масштабе 2~мс на KW и $6.9\times10^5$~отсчётов~с$^{-1}$ на масштабе 64~мс у BATSE. 
Обе временные истории существенно искажены эффектами мёртвого времени и наложения 
импульсов (когда два фотона считаются как один с суммарной энергией). 
Полученная статистическая ошибка задержки для этого всплеска составила всего 3~мс, 
для учёта описанных эффектов была добавлена систематическая ошибка 6~мс.

Полученные ошибки временных задержек находятся в диапазоне от 5~мс до 84~мс со 
средним 24~мс и геометрическим средним 18~мс. Полученные $3\sigma$ полуширины  
колец находится в диапазоне от $0\fdg082$ до $11\fdg0$ 
со средним $1\fdg14$ и геометрическим средним $0\fdg60$. 
Наиболее широкое кольцо с $3\sigma$ полушириной $11\fdg0$ получено 
для GRB19991001\_T04950 (триггер BATSE \#7781), в это время \textit{Wind} находился 
всего в 0.34~световых секунды от Земли.

Расстояния между центральными линиями колец KW--BATSE и центрами локализаций 
BATSE находятся в диапазоне от $0\fdg007$ до $7\fdg7$ 
со средним $2\fdg23$ и геометрическим средним $0\fdg60$. 
Для 14 всплесков $1\sigma$ круговая область локализации BATSE не пересекает кольца 
KW--BATSE и расстояния от ближайшей границы кольца находятся в диапазоне 
от $1.02\sigma$ до $7.2\sigma$. Из 52 всплесков 16 наблюдались только Конус-Винд 
и BATSE и 12 наблюдались только Конус-Винд, BATSE и \textit{BeppoSAX}. Для этих 
всплесков область локализации была получена в виде сегмента кольца KW--BATSE 
с использованием следующего метода. В качестве центра сегмента выбиралась точка 
на центральной линии кольца ближайшая к центру локализации BATSE, и в качестве 
углов сегмента выбирались точки пересечения кольца и окружности с центром в этой 
точке и с радиусом равным сумме удвоенной $1\sigma$ ошибки локализации BATSE, 
систематической ошибки, взятой равной $2\fdg0$ и расстояния между 
центром локализации BATSE и центральной линией кольца. Иллюстрация метода приведена 
на Рис.~\ref{img:IPN_BATSE_loc}. Систематическая ошибка локализаций BATSE $\simeq 2^{\circ}$ 
была обнаружена в работе~\citep{Briggs_1999ApJS}.

\begin{figure}[h]
    \center{\includegraphics[width=0.5\textwidth]{IPN_BATSE_loc}}
    \caption[IPN/BATSE локализация GRB19960420\_T16844 (BATSE \#5439)]
    {IPN/BATSE локализация GRB19960420\_T16844 (BATSE \#5439). 
    Центр локализации BATSE 
    (R.A., decl.(J2000), Err = $ 234\fdg25$, $-27\fdg23$, $2\fdg37$)
    лежит в $3\fdg38$ от центральной линии кольца KW-BATSE.
    Полученный протяженный бокс показан сплошной черной линией и его центр 
    (то есть, ближайшая к центру локализации BATSE точка на центральной линии кольца) обозначена звездочкой. 
    Углы бокса образованы пересечением окружности с центром, отмеченным звездочкой, и
    радиусом $8\fdg12$, то есть суммой $2\sigma$ радиуса локализации BATSE
    и $3\fdg38$ систематической ошибки (пунктирная линия), и кольца KW-BATSE.
    }
 \label{img:IPN_BATSE_loc}  
\end{figure}
\FloatBarrier
\subsubsection{Триангуляции KW--\textit{Fermi}~(GBM)}
Инструмент GBM на борту обсерватории \textit{Fermi} предназначен для изучения 
гамма-всплесков в диапазоне $\sim8$~кэВ--40~МэВ~\citep{Meegan_2009ApJ}. 
Преимущества GBM состоят в высокой эффективной площади и возможности временной 
привязки каждого фотона (Time-tagged events, TTE данные). Часы на борту GBM имеют 
точность временной привязки превышающую 20~мкс. TTE данные содержат отсчёты в 128 
энергетических каналах от $\sim5$~кэВ--2~МэВ, что даёт возможность получить 
временную историю в тех же энергетических диапазонах что и на Конус-Винд.

Инструмент GBM наблюдал 34 коротких всплеска Конус-Винд, для всех из них были 
получены триангуляционные кольца. Для кросс-корреляции с триггерными всплесками 
BATSE использовались временные истории Конус-Винд в диапазонах G2+G3 или G2 с 
временным разрешением 2 или 16~мс и временные истории GBM с разрешением от 1 до 
16~мс созданные из TTE данных только NaI детекторов.

Полученные $\chi^2_{r,\textrm{min}}$ находятся в диапазоне от 0.16 до 2.10 со средним 0.9. 
Полученные ошибки времён задержки лежат в диапазоне 2.5~мс до 136~мс со средним 22~мс и 
геометрическим средним 15~мс. Полученные $3\sigma$ полуширины  колец находится в 
диапазоне от $0\fdg035$~($2\farcm1$) до $1\fdg65$ 
со средним $0\fdg35$ и геометрическим средним $0\fdg23$.

\subsubsection{Триангуляции KW--\textit{INTEGRAL}~(SPI-ACS)}
Помимо своего прямого назначения~--- отсечения фоновых событий германиевого 
спектрометра инструмента SPI, защита ACS используется как в качестве всенаправленного 
детектора гамма-всплесков~\citep{von_Kienlin_2003AA}. Инструмент измеряет временные 
истории гамма-всплесков с временным разрешением 50~мс в одном энергетическом 
диапазоне выше $\sim 80$~кэВ (подробнее см. в~\citep{Lichti_2000AIPC}). 
Систематическая ошибка $125\pm10$~мкс во временной привязке ACS была обнаружена~\citep{Rau_2004GCN} 
и начиная с апреля 2004~г. все временные истории SPI-ACS корректируются автоматически; 
корректировка для предшествующих данных была выполнена вручную.

Систематическая ошибка связана с тем что преобразование из бортового времени в UTC 
происходило приближенно при получении временных историй SPI-ACS в реальном времени 
(в пределах нескольких секунд после триггера \url{ftp://isdcarc.unige.ch/arc/FTP/ibas/spiacs/}). 
С другой стороны, преобразование времени, использовавшиеся для архивных и данных и данных, 
приходящих с задержкой, является точным. Также было показано, что дрейф часов ACS 
по отношению к часам германиевого детектора составляет в течении всей миссии 
составило $\sim 1$~мс~\citep{Zhang_2010int}, таким образом уменьшив систематическую 
ошибку временной привязки ACS с 10~мс до 1~мс.

Временные истории SPI-ACS, скорректированные на систематические сдвиги и имеющие 
высокую точность привязки (по крайней мере с точность вплоть до 1~мс), доступны 
в архиве данных \textit{INTEGRAL} начиная с версии 3. Архивные и данные, приходящие 
с задержкой, с одинаковой точностью временной привязки (приходящие в пределах 
часа после регистрации всплеска) доступны на ресурсе 
\url{http://isdc.unige.ch/~savchenk/spiacs-online/} и 
\url{http://www.isdc.unige.ch/heavens/}. 
Эти данные систематически используются для оперативной триангуляции.

Инструмент SPI-ACS зарегистрировал 139 коротких всплеска Конус-Винд, из них 
для 103 были получены триангуляционные кольца. Для кросс-корреляции использовались 
временные истории Конус-Винд в диапазонах G2+G3 или G3 с временным разрешением 2 или 16~мс.

Полученные $\chi^2_{r,\textrm{min}}$ находятся в диапазоне от 0.04 до 3.96 со средним 1.02. 
Полученные ошибки задержек лежат в диапазоне 4~мс до 175~мс со средним 24~мс 
и геометрическим средним 19~мс. Полученные $3\sigma$ полуширины колец находится 
в диапазоне от $0\fdg047$~($2\farcm8$) до $4\fdg3$ 
со средним $0\fdg41$ и геометрическим средним $0\fdg29$.

\subsubsection{Триангуляции KW--\textit{Suzaku}~(WAM)}
Инструмент WAM является активной защитой детектора жесткого рентгеновского 
излучения на борту миссии \textit{Suzaku}~\citep{Yamaoka_2009PASJ}. В триггерном 
режиме WAM записывает временные истории всплесков с временным разрешением 1/64~с 
в четырёх каналах в диапазоне $\approx50$--5000~кэВ. В режиме реального времени 
разрешение составляет 1~с. В работе~\citep{Yamaoka_2009PASJ} было показано, 
что систематическая ошибка временной привязки \textit{Suzaku}~(WAM) пренебрежимо мала.

Инструмент WAM зарегистрировал 61 короткий всплеск Конус-Винд: 51 в триггерном 
режиме и 10 в режиме реального времени. Кольца были получены для 45 триггерных всплесков.

Для кросс-корреляции использовались временные истории Конус-Винд в 
диапазонах G2+G3 или G3 с временным разрешением 2 или 16~мс и временные истории 
WAM в сумме четырёх диапазонов детектора с наиболее сильным откликом.

Полученные $\chi^2_{r,\textrm{min}}$ находятся в диапазоне от 0.21 до 1.78 со средним 1.03. 
Полученные ошибки задержек лежат в диапазоне от 4~мс до 104~мс со средним 20~мс 
и геометрическим средним 14~мс. Полученные $3\sigma$ полуширины колец находятся 
в диапазоне от $0\fdg060$~($3\farcm6$) до $2\fdg44$ 
со средним $0\fdg30$ и геометрическим средним $0\fdg21$.

\subsubsection{Триангуляции KW--\textit{BeppoSAX}~(GRBM)}
Инструмент \textit{BeppoSAX}~(GRBM) являлся защитой, работающей по принципу 
антисовпадения, системы детектирования гамма квантов PHOSWICH 
(PHOSphor sandWICH)~\citep{Feroci_1997SPIE, Frontera_1997AAS}. В триггерном режиме 
инструмент измерял временные истории с разрешением 7.8125~мс в диапазоне 40--700~кэВ; 
в режиме реального времени разрешение составляло 1~с.

Инструмент GRBM зарегистрировал 50 коротких всплесков Конус-Винд: 41 в триггерном 
режиме и 9 в режиме реального времени. Триангуляционные кольца были получены для 38 всплесков,
зарегистрированных в триггерном режиме и для одного всплеска, зарегистрированного 
в режиме реального времени (этот всплеск наблюдался только Конус-Винд и GRBM).

Для кросс-корреляции использовались временные истории Конус-Винд в диапазонах G2 
или G2+G3 с временным разрешением 2 или 16~мс и временные истории GRBM, 
приведённые к разрешению 32~мс.

Полученные $\chi^2_{r,\textrm{min}}$ находятся в диапазоне от 0.25 до 12.1 со 
средним 1.43. Максимальный $\chi^2_{r,\textrm{min}}$ равный 12.1 (dof=6) 
соответствует исключительно интенсивному событию GRB19970704\_T04097 с пиковой 
скоростью счёта $1.8\times10^5$~отсчётов~с$^{-1}$ на Конус-Винд на масштабе 2~мс 
и $1.5\times10^5$~отсчётов~с$^{-1}$ в GRBM на масштабе 32~мс. Обе временные истории 
существенно искажены эффектами мёртвого времени и наложения импульсов 
(когда два фотона считаются как один с суммарной энергией). Полученная статистическая 
ошибка задержки для этого всплеска составила всего 2~мс, для учёта описанных 
эффектов ошибка была увеличена до 6~мс.

Полученные ошибки задержек лежат в диапазоне от 4.5~мс до 216~мс со средним 32~мс 
и геометрическим средним 18~мс.

Сравнение первоначально полученных колец с другими кольцами IPN, так же как 
сравнение временных историй GRBM и BATSE выявило систематический сдвиг во 
временной привязке GRBM доходящий до 100~мс. Так как этот сдвиг варьируется от 
всплеска к всплеску, для триангуляции KW--GRBM была введена 100~мс 
систематическая ошибка. Это привело к существенному уширению колец. Таким образом, 
конечные $3\sigma$ полуширины колец находятся в диапазоне от $1\fdg23$ 
до $32\fdg2$ со средним $5\fdg30$ и геометрическим 
средним $3\fdg87$.

\subsubsection{Триангуляции KW--\textit{Swift}~(BAT)}
\textit{Swift}~(BAT) -- высокочувствительный телескоп с кодирующей маской с 
широким полем зрения, который регистрирует гамма-всплески в реальном 
времени~\citep{Barthelmy_2005SSRv}. Если всплеск происходит вне поля зрения, 
он не может быть локализован, но временная история BAT может быть использована 
для триангуляции. Для таких всплесков всегда доступна временная история 
с разрешением 64~мс в четырёх стандартных диапазонах BAT (15--25~кэВ, 25--50~кэВ, 
50--100~кэВ и 100-350~кэВ). Для некоторых всплесков доступны TTE данные, что 
даёт возможность получить временную историю с любым необходимым разрешением.

Инструмент BAT зарегистрировал 44 коротких всплеска Конус-Винд вне поля зрения, 
для 23 из них были получены триангуляционные кольца.

Для кросс-корреляции использовались временные истории Конус-Винд в диапазонах G2 
или G2+G3 с временным разрешением 2 или 16~мс и временные истории BAT с разрешением 
64~мс в большинстве случаев в диапазоне выше 50~кэВ, что обычно даёт лучшее 
отношение сигнал-шум и лучшее соответствует диапазону Конус-Винд.

Полученные $\chi^2_{r,\textrm{min}}$ находятся в диапазоне от 0.25 до 7.48 со 
средним 1.41. Максимальный $\chi^2_{r,\textrm{min}}$ равный 12.1 (dof=6) 
соответствует исключительно интенсивному всплеску GRB20060306\_T55358 с сильной 
спектральной эволюцией и пиковой скоростью счёта $1.9\times10^5$~отсчётов/с 
на Конус-Винд на масштабе 2~мс. Полученная статистическая ошибка задержки 
для этого всплеска составила всего 5~мс, для учёта описанных эффектов была 
добавлена систематическая ошибка 10~мс.

Полученные ошибки временных задержек лежат в диапазоне от 5~мс до 64~мс со 
средним 22~мс и геометрическим средним 18~мс. Полученные $3\sigma$ полуширины 
колец находятся в диапазоне от $0\fdg059$~($3\farcm5$) 
до $1\fdg18$ со средним $0\fdg41$ 
и геометрическим средним $0\fdg29$.

\subsubsection{Триангуляции KW--\textit{Коронас-Ф} (Геликон)}
Гамма спектрометр Геликон, установленный на солнечной обсерватории Коронас-Ф, 
имел схожие с  Конус-Винд характеристики детекторов и типы научных данных. 
Схожее устройство двух инструментов позволило получить хорошие кросс-корреляции 
временных историй всплесков.

Геликон зарегистрировал 14 коротких всплесков Конус-Винд, для всех из них были 
получены кольца KW--Геликон.

Полученные $\chi^2_{r,\textrm{min}}$ находятся в диапазоне от 0.25 до 2.67 со средним 1.02. 
Полученные ошибки временных задержек лежат в диапазоне от 4~мс до 80~мс со средним 25~мс 
и геометрическим средним 17~мс. Полученные $3\sigma$ полуширины колец находятся 
в диапазоне от $0\fdg045$~($2\farcm7$) до $1\fdg15$
со средним $0\fdg38$ и геометрическим средним $0\fdg25$.

\subsubsection{Триангуляции KW--\textit{Космос}~(Конус-А, А2, А3)}
Гамма спектрометры Конус-А, Конус-А2 и Конус-А3 были установлены на КА 
\textit{Космос} -2326, -2367 и -2421. Краткое описание инструмента Конус-А дано в~\citep{Aptekar_1998ApJ}, 
инструменты Конус-А2 и Конус-А3 имели схожее устройство и типы научных данных.

Этими инструментами было зарегистрировано в триггерном режиме пять коротких 
всплесков Конус-Винд. Триангуляционные кольца KW--\textit{Космос} были 
получены для четырёх из них.

Полученные $\chi^2_{r,\textrm{min}}$ находятся в диапазоне от 0.73 до 1.42 со средним 1.07. 
Полученные ошибки временных задержек лежат в диапазоне от 4~мс до 56~мс со средним 35~мс. 
Полученные $3\sigma$ полуширины колец находятся в диапазоне от $0\fdg15$ 
до $1\fdg20$ со средним $0\fdg79$.

\subsubsection{Триангуляции KW--\textit{RHESSI}}
Гамма спектрометр высокого разрешения \textit{RHESSI} предназначен для изучения 
излучения высоких энергий от солнечных вспышек в широком диапазоне энергий 
от 3~кэВ до 17~МэВ~\citep{Lin_2002SoPh, Smith_2002SoPh}. Данные накопленные в режиме TTE 
позволяют получить произвольную временную и спектральную группировку зарегистрированных квантов.

Инструмент \textit{RHESSI} зарегистрировал 58 коротких всплесков Конус-Винд, из 
них для 32 были получены кольца KW--\textit{RHESSI}.

Для кросс-корреляции использовались временные истории Конус-Винд 
в диапазонах G2, G1+G2 или G2+G3 с временным разрешением 2, 16, 64 или 256~мс, 
в зависимости от интенсивности всплеска.

Полученные $\chi^2_{r,\textrm{min}}$ находятся в диапазоне от 0.36 до 2.62 со 
средним 1.07. Полученные ошибки временных задержек лежат в диапазоне от 2~мс 
до 184~мс со средним 36~мс и геометрическим средним 20~мс. Полученные $3\sigma$ 
полуширины колец находятся в диапазоне от $0\fdg027$~($1\farcm6$) 
до $2\fdg71$ со средним $0\fdg53$ 
и геометрическим средним $0\fdg30$.

\subsubsection{Триангуляции KW--\textit{HETE-2} (FREGATE)}
Гамма спектрометр FREGATE на борту \textit{HETE-2} был предназначен для регистрации 
гамма-всплесков в диапазоне энергий 8--400~кэВ~\citep{Ricker_2003AIPC, Atteia_2003AIPC}. 
В триггерном режиме он записывал временные истории гамма-всплесков с временным 
разрешением 1/32~с в диапазоне 8--400~кэВ, помимо этого велась непрерывная запись 
скорости счёта с разрешением 0.1638~с.

Инструмент FREGATE зарегистрировал 16 коротких всплесков KW: 8 в триггерном режиме 
и 8 в режиме непрерывной записи. В большинстве случаев отклик FREGATE был существенно
слабее чем у других инструментов, установленных на КА с низкими околоземными орбитами, 
поэтому данные FREGATE использовались нескольких случаях, года ни один другой КА на низкой 
орбите не детектировал данный всплеск. Кольца KW--FREGATE были получены для четырёх 
всплесков, зарегистрированных в триггерном режиме.

Для кросс-корреляции использовались временные истории KW в диапазонах G2 
или G2+G3 с временным разрешением 2 или 16~мс.

Полученные $\chi^2_{r,\textrm{min}}$ находятся в диапазоне от 0.50 до 1.43 со средним 0.96. 
Полученные ошибки временных задержек лежат в диапазоне от 56~мс до 168~мс со средним 102~мс. 
Полученные $3\sigma$ полуширины колец находятся в диапазоне от $0\fdg95$ 
до $1\fdg47$ со средним $1\fdg10$.

\subsubsection{Триангуляции KW--\textit{AGILE}~(MCAL)}
Гамма спектрометр MCAL на борту миссии \textit{AGILE} чувствителен к гамма-квантам 
с энергией $\approx 0.35$--100~МэВ~\citep{Tavani_2009AA}. Запись временных историй 
гамма-всплесков в триггерном режиме ведётся в формате TTE.

Инструмент MCAL зарегистрировал 24 коротких всплесков KW: 22 в триггерном 
режиме и 2 в режиме непрерывной записи. Во многих случаях отклик MCAL оказывался 
слабым из-за его высокого энергетического порога и сильного экранирования инструментом GRID, 
поэтому MCAL использовался для триангуляции сильных всплесков. В сумме было получено девять колец
KW--MCAL.

Для кросс-корреляции использовались временные истории KW в диапазонах G3 
или G2+G3 с временным разрешением 2 или 16~мс.

Полученные $\chi^2_{r,\textrm{min}}$ находятся в диапазоне от 0.29 до 2.26 со 
средним 1.08. Полученные ошибки временных задержек лежат в диапазоне от 5~мс 
до 21~мс со средним 13~мс. Полученные $3\sigma$ полуширины колец находятся 
в диапазоне от $0\fdg071$~($4\farcm3$) до $0\fdg06$ 
со средним $0\fdg21$.

\subsubsection{Триангуляции \textit{INTEGRAL}--околоземные КА}
Даже без планетарных миссий, мини-сеть КА на низких околоземных орбитах, 
плюс \textit{INTEGRAL} и KW, часто позволяют получить область локализовать 
гамма-всплеска. Так как орбита \textit{INTEGRAL} расположена на расстояниях 
$\lesssim 0.5$~световых секунд, что гораздо меньше расстояния Земля--\textit{Wind} 
$\simeq 5$~световых секунд, кольца KW--\textit{INTEGRAL} и KW--околоземные~КА 
пересекаются под очень острым углом, образовывая одну или две вытянутых области локализации. 
В некоторых случаях пересечения колец \textit{INTEGRAL}--околоземный~КА и 
KW--околоземный~КА дают меньшую область локализации.

Суммарно было получено 11 колец \textit{INTEGRAL}--околоземные~КА. Полученные $3\sigma$ 
полуширины колец находятся в диапазоне от $1\fdg0$ до $14\fdg0$ 
со средним $5\fdg9$.

\subsection{Проверка достоверности триангуляционных колец}
Среди 271 короткого гамма всплеска KW, локализованного IPN, 17 были точно 
локализованы инструментами, способными строить изображения в рентгеновском или 
мягком гамма-диапазоне: 15~\textit{Swift}-BAT (один из них, GRB~090510, был 
так же локализован \textit{Fermi}-LAT), 1~\textit{HETE-2}~(WXM и SXC) и 1~\textit{INTEGRAL}~(IBIS/ISGRI).

Эти всплески были использованы для проверки полученных триангуляций. Для этих 17 всплесков 
было получено 21 кольцо KW--околоземные~КА и 12 колец KW--\textit{INTEGRAL}, 
при этом не использовалась временная история инструмента, строившего изображение, 
так как отклик инструмента на всплески в поле зрения отличен от отклика для всплесков 
вне поля зрения, чьи временные истории использовались для IPN триангуляции. 
Во всех случаях триангуляционные кольца согласовывались с известным положением источника. 
Подобная проверка не только подтверждает точность временной привязки и эфемерид космических
аппаратов, но и пригодность методики кросс-корреляции и метода получения колец.

На Рис.~\ref{img:dN_33_rel_offset} представлено распределение относительных расстояний источников 
от центральных линий колец, видно что все расстояния, по абсолютной величине, меньше $2\sigma$.
Наибольшее отрицательное отклонение составляет $-2\sigma$, 
наибольшее положительное~--- $1.9\sigma$, среднее~--- $0.04\sigma$ и стандартное отклонение $1.1\sigma$.
Помимо этой проверки, часто правильность триангуляции KW--околоземные~КА можно 
установить по согласию нескольких колец KW--околоземные КА между собой 
и с кольцами, полученными с использованием дальних КА.

\begin{figure}[h]
    \center{\includegraphics[width=0.5\textwidth]{dN_33_rel_offset}}
    \caption[Распределение расстояний от центра кольца до точного положения источника]
    {Распределение расстояний от точных позиций GRB до центральных линий 33 колец, 
    полученных с использованием  KW и околоземного КА (или INTEGRAL).
    }
 \label{img:dN_33_rel_offset}  
\end{figure}

\FloatBarrier
\subsection{Дополнительные ограничения локализаций}
Помимо триангуляционных колец, было получено ещё несколько типов локализационной информации. 
Они включают: диапазон эклиптических широт, автономные локализации, полученные 
\textit{CGRO}-BATSE, \textit{BeppoSAX}-GRBM и \textit{Fermi}-GBM, а также области 
затенённые Землёй или Марсом (\textit{MESSENGER} находится на вытянутой орбите вокруг Меркурия, из-за этого, 
затенения Меркурием редки). Эта дополнительная информация помогает ограничить 
положение источника, полученное триангуляционным методом, например, 
выбрать одну из областей локализации или исключить часть кольца.

\subsubsection{Эклиптические широты}
Эклиптические широты всплесков вычисляются на основе отношения скоростей счёта
в двух детекторах KW, измеренных в режиме фон с разрешением 1.472 или 2.944~c. Ось
детектора S2 направлена в северный полюс эклиптики, а ось детектора S1~--- в южный.
Помимо статистической ошибки, получаемая эклиптическая широта имеет систематическую 
ошибку, связанную, помимо прочего, с переменными рентгеновскими источниками,
затенениями другими инструментами на борту стабилизированного вращением КА~\textit{Wind}.
Ошибки полученных значений были взяты на уровне 95\%.

Диапазон эклиптических широт, а именно, наилучшая оценка $b$, верхний и нижний
пределы $b_{\rmn{min}}$, $b_{\rmn{max}}$, можно рассматривать как кольцо с центром в северном или южном
полюсе эклиптики с углом раствора $\theta = 90^\circ - |b|$ и полуширинами 
$d_{-}(\theta) = b_{\rmn{min}} - b$ и $d_{+}(\theta) = b_{\rmn{max}} - b$.

\subsubsection{Затенения планетами}
Затенения планетой задаётся прямым восхождением и склонением центра планеты и её
радиусом. При наблюдении всплеска на околоземном или околомарсианском КА планета
затеняет до $\approx 3.7$~ср~неба. Положение источника должно быть вне этой 
затенённой части неба.

Разрешенная часть неба может быть представлена как вырожденное кольцо с центром
в направлении, противоположном центру планеты с углом раскрытия $\theta =0$ и 
полу ширинами $d_{-}(\theta) = 0$ и $d_{+}(\theta)= \arcsin(R_{\rmn{planet}}/R)$, 
где $R$~--- радиус орбиты КА (здесь пренебрегается сплюснутостью планеты 
и поглощением излучения в её атмосфере).

\subsubsection{Автономные локализации}
Принцип автономной локализации всплесков с помощью системы детекторов с анизотропной 
чувствительностью был предложен в~\citep{Golenetskii_1974CosRe} и впервые реализован 
в системе Конус на КА Венера~11 и~12~\citep{Mazets_1981ApSS}. Схожие системы локализации 
с различным числом детекторов были установлены на KA \textit{CGRO} (BATSE), 
\textit{BeppoSAX} (GRBM) и \textit{Fermi} (GBM). Автономная локализация рассчитывается 
на основе соотношения скоростей счёта в детекторах. При этом на локализацию влияет 
альбедо Земли и поглощение излучения в конструкциях КА. Полученные локализации 
представляют собой плотность вероятности положения источника на небесной сфере. 
Локализации обычно имеют сложную форму и аппроксимируются кругом с центров в точке 
наиболее вероятного положения источника, радиус выбирается таким образом, 
чтобы площадь круга была равна соответствующей площади реальной области локализации 
для уровней значимости $1\sigma$ (BATSE, GBM) или 90\% (GRBM), с учётом только 
статистической неопределённости. Все локализации подвержены систематическим ошибкам 
не менее нескольких градусов.

Эти области также могут быть описаны вырожденным кольцом с центром в точке
наиболее вероятного положения источника, углом раскрытия $\theta =0$ и с ширинами 
$d_{-}(\theta) = 0$ и $d_{+}(\theta)= r$, где $r$~--- ошибка положения источника.

\section{Результаты локализации}
Информация по локализации 271 короткого всплеска Конус-Винд доступна на сайте 
ФТИ\footnote{\url{http://www.ioffe.ru/LEA/shortGRBs/Catalog2/Data/tables/}}.
Первая колонка содержит обозначение всплеска. Вторая колонка даёт 
число ограничений на локализацию всплеска (число строк с локализационной информацией 
для всплеска). Третья колонка содержит название источника локализации: 
КА1–КА2 (триангуляционное кольцо, полученное с использованием КА1 и КА2) или 
<<Ecl.Band>>\ (диапазон эклиптических широт), или <<Instrument>>\ (название инструмента, 
автономно локализовавшего всплеск), или <<Occ.sc>>\ (затенение планетой). 
Колонки (4)--(8) содержат прямые восхождения и склонения центров колец (J2000), 
угол раствора кольца $\theta$, и $3\sigma$ ошибки радиуса $d_{-}(\theta)$ и $d_{+}(\theta)$.

Затенения планетами даны только в случаях если они ограничивают локализацию. 
Диапазон эклиптических широт дан для всех всплесков. Также приведены все автономные 
локализации.

Локализации \textit{Swift}-BAT взяты из второго каталога \textit{Swift}-BAT, 
охватывающего период с 19~декабря 2004 по 21~декабря 2009~\citep{Sakamoto_2011ApJS}. 
Для более поздних всплесков локализации взяты из циркуляров GCN 
(\url{http://gcn.gsfc.nasa.gov/gcn3_archive.html}) с уточнённой локализацией BAT.

Локализация HETE-2 всплеска GRB~040924 (=GRB20040924\_T24735) взята из~\citep{Arimoto_2006GCN}.

Локализация IBIS/ISGRI всплеска GRB~070707 (=GRB20070707\_T58122) взята из~\citep{Gotz_2007GCN}.

Локализации BATSE взяты из каталога на сайте 
эксперимента\footnote{\url{http://www.batse.msfc.nasa.gov/batse/grb/catalog/current/}}, 
а также из каталога нетриггерных всплесков BATSE~\citep{Kommers_2000ApJ, Stern_2001ApJ}. 
Каталог~\citep{Stern_2001ApJ} содержит локализации всех восьми коротких всплесков Конус-Винд, 
зарегистрированных BATSE в режиме непрерывной записи, в каталоге~\citep{Kommers_2000ApJ} часть из них пропущена, 
приведённые локализации взяты исключительно из~\citep{Stern_2001ApJ}.

Локализации \textit{BeppoSAX} взяты как из каталога GRBM~\citep{Frontera_2009ApJS} 
так и из циркуляров IAU (\url{http://www.cbat.eps.harvard.edu/iauc/RecentIAUCs.html}) 
и GCN.

Локализации GBM взяты из первого каталога \textit{Fermi}-GBM, охватывающего период с
12~июля 2008 по 11~июля 2010~\citep{Paciesas_2012ApJS}, циркуляров GCN или из последних 
версий соответствующих glg\_tcat*.fit файлов из архива GBM 
(\url{ftp://legacy.gsfc.nasa.gov/fermi/data/gbm/}).

\subsection{Пересечения колец}
Для всплесков, зарегистрированных тремя или более достаточно удалёнными друг от 
друга КА, может быть получена область локализации (бокс) в виде четырёхугольника 
или более сложной формы.

В общем случае пересечение двух колец, построенных с использованием дальнего КА,
даёт маленький бокс с площадью вплоть до одной квадратной угловой минуты.

Пересечение двух колец, полученных с использованием дальнего КА, Конус-Винд и
околоземного КА обычно даёт вытянутый бокс, который тем не менее имеет небольшую
площадь в несколько сотен квадратных угловых минут. В некоторых случаях пересечение кольца, полученного
с использованием дальнего КА, Конус-Винд и околоземного КА может дать бокс меньшей
площади, чем с использованием двух дальних КА.

Протяженные боксы были получены для всплесков, не зарегистрированных ни одним дальним КА, 
но зарегистрированные KW, \textit{INTEGRAL} (SPI-ACS) и одним иди более околоземным КА. 
В этих случаях бокс образуют кольцо KW--околоземный КА и
кольцо \textit{INTEGRAL}--околоземный~КА или кольцо KW--околоземный КА 
и кольцо KW--\textit{INTEGRAL}, пересекающиеся под достаточно острым углом.
Во всех случаях, когда КА используемые для триангуляции, практически лежат на
одной линии кольца пересекаются под очень острыми углами, образуя протяженный бокс.
Сумаарно было получено 162 бокса для коротких всплесков Конус-Винд: 27 для всплесков, 
зарегистрированных двумя дальними КА; 84 для всплесков, зарегистрированных одним 
дальним КА, и, по крайней мере, одним околоземным КА и 51 для всплесков, 
зарегистрированных KW, INTEGRAL и, по крайней мере, одним околоземным КА. В некоторых
случаях полученные области локализации представляют собой дуги (значительные части
колец).

\subsection{Сегменты}
Для всплесков, зарегистрированных только KW и другим КА или KW 
и одним или более околоземных КА, полученная локализация представляет часть кольца 
(наиболее узкого при детектировании несколькими околоземными КА), удовлетворяющая 
дополнительными ограничениями. Эти локализации представляют собой целое кольцо 
(если оно полностью находится внутри ограничений на эклиптическую широту и 
отсутствуют другие ограничения), или один или два сегмента кольца, образованные 
пересечением кольца и диапазоном эклиптических широт, и/или исключением затенённой 
планетой части кольца, или часть кольца, ограниченная локализацией BATSE (см. раздел~\ref{sec:KW_BATSE_cc}).
Описанный тип локализации имеют 114 всплесков: 20~--- всплески, зарегистрированные 
KW и одним дальним КА (из них 3 зарегистрированы одним околоземным КА в режиме 
непрерывной записи, но для них не было получено кольцо KW--околоземный КА) и 94 
всплеска были зарегистрированы только KW и одним или более околоземным КА.

\subsection{Полученные области локализации}
Окончательные области локализации для 254 коротких всплесков KW 
(в список не включены 17 точно локализованных всплесков) приведены на ресурсе 
\url{http://vizier.cfa.harvard.edu/vizier/ftp/cats/J/ApJS/207/38/table3.dat}.
Девять колонок таблицы содержат: 
(1) обозначение всплеска; (2) число областей локализации всплеска $N_{\rmn{r}}$ (1 или 2); 
(3) число углов у области локализации $N_{\rmn{c}}$, 
(4) тип области: <<B>>\ (бокс), <<LB>>\ (протяженный бокс: бокс с наибольшим размером $> 10^\circ$),
<<S>>\ (сегмент) или <<A>>\ (кольцо);
(5) площадь области (или сумма площадей в случае двух областей локализации); 
(6) наибольший размер области (наибольшее угловое расстояние между двумя точками границы области; 
для сегментов больших чем половина кольца приведён внешний диаметр кольца);
(7) название локализаций, формирующих область;
(8) прямое восхождение центра области, в первой строке, и прямое восхождение 
всех углов области в последующих $N_\rmn{c}$ строках и 
(9) склонения центра области, в первой строке, и склонение всех углов области в последующих
$N_\rmn{c}$ строках. В случае двух областей локализации дополнительно даны $N_{\rmn{c}} + 1$ строк: 
центр второй области и её углы. 
Таким образом, для таких всплесков дано $2(N_{\rmn{c}} + 1)$ строк. Все координаты даны на эпоху J2000.
В общем случае область локализации не может быть полностью описана четырёхугольником, 
для точного представления должна учитываться кривизна колец. Только в случаях, 
когда максимальный размер области не превышает нескольких градусов, 
область может описываться только четырьмя углами. В других случаях, когда область переставляет
собой длинную дугу или сегмент кольца приведённые центр и углы области предназначены
для приблизительного указания положения области на небе. Рисунки, иллюстрирующие
IPN локализации (все вычисленные кольца и полученные области) можно найти на сайте
ФТИ им.~А.~Ф.~Иоффе (\url{http://www.ioffe.ru/LEA/ShortGRBs_IPN/}).

Распределение площадей полученных областей локализаций приведено на Рис.~\ref{img:dN_256_areas}.
Для всплесков, зарегистрированных дальними КА, площади находятся в диапазоне от 
$2.40 \times 10^{-4}$~deg$^2$ 0.86~arcmin$^2$) до 142~deg$^2$ со средним 3.49~deg$^2$ 
и геометрическим средним 0.141~deg$^2$. Для всплесков, не зарегистрированных дальними КА, 
площади находятся в диапазоне от 0.210~deg$^2$ до 4420~deg$^2$ со средним 242~deg$^2$
и геометрическим средним 46.2~deg$^2$.

\begin{figure}[h]
    \center{\includegraphics[width=0.5\textwidth]{dN_256_areas}}
    \caption[Распределение площадей 256 областей локализации]
    {Распределение площадей областей локализации для 123 коротких всплесков Конус-Винд
    наблюдавшихся, по крайней мере, одним дальним КА (красная пунктирная линия) и 
    131 всплеска, не наблюдавшегося ни одним дальним КА (синяя пунктирная линия). 
    Суммарное распределение 254 всплесков обозначено непрерывной линией 
    (17 точно локализованных всплесков не использовались при построении распределения).
    }
 \label{img:dN_256_areas}  
\end{figure}
\FloatBarrier
\section{Обсуждение особых событий}
GRB~051103 (=GRB20051103\_T33943) может являться гигантской вспышкой мягкого 
гамма-репитера (SGR) в близкой группе взаимодействующих галактик M81, что было 
предложено в работе~\citep{Frederiks_2007AstL}. Итоговая область локализации 
этого события совместно с дальнейшим исследованием предложенной природы этого 
события приведено в~\citep{Hurley_2010MNRAS}. В работе~\citep{Ofek_2006ApJ} 
приведены выводы, сделанные на основе оптических и радио наблюдений, последовавших 
за детектированием всплеска в гамма-диапазоне. В работе~\citep{Abadie_2012ApJ} обсуждается
не детектирование гравитационных волн во время этого события.

GRB~070201 (=GRB20070201\_T55390) вероятно является гигантской вспышкой SGR
в туманности Андромеды~\citep{Mazets_2008ApJ}. В работе~\citep{Abbott_2008ApJ} 
обсуждаются результаты недетектирования гравитационных волн от этого события 
и в~\citep{Ofek_2008ApJ} обсуждается недетектирование оптического послесвечения 
и периодического рентгеновского источника.

Поиск других кандидатов в гигантские вспышки SGR на основе полученных локализаций 
приведён в следующей главе.

GRB~000420 (=GRB20000420\_T42271) на основании кольца KW--\textit{NEAR} 
был отнесён в работе~\citep{Ofek_2007ApJ} к кандидатам в гигантские вспышки SGR 
в близкой галактике M74 (NGC~628). Положение этой галактики находится 
вне кольца KW--\textit{BeppoSAX}, таким образом исключая возможность 
происхождения этого всплеска в M74 (см.~Рис.~\ref{img:GRB000420_loc}).

\begin{figure}[h]
    \center{\includegraphics[width=0.5\textwidth]{GRB000420_loc}}
    \caption[Локализация всплеска GRB~000420]
    {IPN локализация GRB 000420 (= GRB20000420\_T42714).
    Узкое кольцо SAX-NEAR, шириной $1\farcm7$, проходит через близкую галактику M74,
    в то время как галактика лежит далеко за пределами широкого $3\sigma$ кольца KW-SAX.
    }
 \label{img:GRB000420_loc}  
\end{figure}

GRB~990405 (=GRB19990405\_T30059) изначально был классифицирован как всплеск
от SGR~1900$+$14 так как узкое кольцо \textit{BeppoSAX}--\textit{Ulysses} 
(с $3\sigma$ полушириной $0\fdg035$) проходило через положение 
этого SGR. Полученное кольцо KW--\textit{BeppoSAX} (с $3\sigma$ полушириной
$6\fdg4$) также согласуется положением этого SGR. Но этот всплеск 
существенно жестче, чем два необычно жестких всплеска из SGR~1900+14: 
981022 и 991001~\citep{Woods_1999ApJ}, делая ассоциацию этого всплеска 
с SGR~1900$+$14 сомнительной.
\FloatBarrier

\section{IPN локализация всплесков, наблюдаемых iPTF}
Короткие гамма-всплески являются наиболее вероятными источниками гравитационных
волн. Для детектирования грав. волн созданы детекторы  Advanced~LIGO~\citep{LIGO_2015CQGra} 
и Advanced~Virgo~\citep{Acernese_2015CQGra}, которые будут способны зарегистрировать сигнал от слияния
двух нейтронных звёзд на расстоянии в несколько сотен мегапарсек.
Область локализации источников гравитационных волн, в случае использования 
трёх детекторов может иметь размер до $\approx 300$~deg$^2$~\citep{Singer_2015PhDT}.

Основным инструментом для поиска оптических транзиентов, связанных с источниками 
гравитационных волн, в настоящее время является телескоп P48, входящий в состав 
системы телескопов для поиска транзиетов Паломарской обсерватории
(iPTF, \textit{Intermediate Palomar Transient Factory}~\citep{Rau_2009PASP}). 
Телескоп имеет диаметр $\approx 1.2$~м, поле зрения 7 квадратных градусов и 
чувствительность на уровне $5\sigma$ в фильтре $R$ (570--730~нм) $\approx 21$ зв.~величина за минутную экспозицию.
%Инструмент может выполнять съёмку заданной области на небе.
Для проверки возможностей инструмента проводится кампания по наблюдению локализаций
гамма-всплесков, зарегистрированных \textit{Fermi}-GBM и IPN.

Длинный гамма-всплеск GRB~120716A был зарегистрирован  \textit{Fermi}-GBM, Konus-\textit{Wind}, 
\textit{INTEGRAL}-SPI-ACS, \textit{Suzaku}-WAM и \textit{MESSENGER}-GRNS,
пощадь полученной локализации  составила $\approx 2$ квадратных градуса~\citep{Hurley_2012GCN13487}.
Оптическое послесвечение всплеска было обнаружено iPTF 
через $\sim 1.5$ дня~\citep{Cenko_2012GCN13489}, на расстоянии 3 угловые минуты от центра локализации.
% возможно стоит подробней написать про этот всплеск - длинный с коротким прекурсором.

За период с 2013 по 2014~гг. iPTF наблюдала локализации 35 гамма-всплесков, зарегистрированных GBM,
для восьми было обнаружено послесвечение. Из них, в четрёх случаях отбор кандидатов
был упрощён благодаря IPN локализации.

\section{Заключение}
В этой главе представлено продолжение серии каталогов локализаций гамма-всплесков 
триангуляционным методом с использованием КА сети IPN. Список предшествующих каталогов 
приведён в таблице~\ref{tab:IPN_cat}.

Полученные локализации могут быть использованы для большого круга задач, включающего: 
поиск гравитационных волн и нейтрино от слияния компактных объектов, 
поиск гамма-квантов сверхвысоких энергий из источников всплесков и поиск гигантских 
вспышек SGR в ближайших галактиках.

\begin{table} [h]
 \centering
 \caption{Каталоги IPN локализаций гамма-всплесков}\label{tab:IPN_cat}
\scriptsize
  \begin{center}
  \begin{tabular}{ccc}
  \hline
  \hline
  Период покрытия, годы & Число всплесков & Описание \\
  \hline
1990--1992 &16 &\textit{Ulysses}, \textit{Pioner Venus Orbiter}, WATCH, SIGMA, PHEBUS GRBs~\citep{Hurley_2000ApJ} \\
1990--1994 &56 &\textit{GRANAT}-WATCH supplement~\citep{Hurley_2000ApJS_WATCH} \\
1991--1992 &37 &\textit{Pioner Venus Orbiter}, \textit{CGRO}, \textit{Ulysses} GRBs~\citep{Laros_1998ApJS}\\
1991--1994 &218 &BATSE 3B supplement~\citep{Hurley_1999ApJSa}\\
1991--2000 &211 &BATSE untriggered GRBs supplement~\citep{Hurley_2005ApJS_BATSE_untrig}\\
1992--1993 &9 &\textit{Mars Observer} GRBs~\citep{Laros_1997ApJS}\\
1994--1996 &147 &BATSE 4Br supplement~\citep{Hurley_1999ApJS_BATSE_4Br}\\
1994--2010 &271 &короткие гамма-всплески Конус-Винд (эта работа)\\
1996--2000 &343 &BATSE 5B supplement~\citep{Hurley_2011ApJS_BATSE_5B}\\
1996--2002 &475 &\textit{BeppoSAX} supplement~\citep{Hurley_2010ApJS_BeppoSAX}\\
2000--2006 &226 &HETE-2 supplement~\citep{Hurley_2011ApJS_HETE2}\\
2008--2010 &146 &\textit{Fermi}-GBM supplement~\citep{Hurley_2013ApJS_GBM}\\
  \hline
  \hline
  \end{tabular}
  \end{center}
\end{table}

\FloatBarrier

В главе получены следующие результаты:
\begin{enumerate}
\item В главе получена наиболее полная локализационная информация для 271 короткого 
гамма-всплеска Конус-Винд. Для 254 всплесков были получены области локализации и 
для 17 всплесков с точно известной локализацией, полученной инструментами с 
возможностью построения изображений в жестком рентгеновском диапазоне, триангуляционные
кольца получены для проверки методики.

Суммарно было получено 517 триангуляционных колец, включая 150 колец с использованием 
дальних КА. Показано, что для многих коротких всплесков триангуляции 
KW--околоземные~КА (или \textit{INTEGRAL}) дают достаточно узкое кольцо 
с полушириной сравнимой или меньше чем при использовании дальних КА. Таким образом, 
даже в случае детек-тирования всплеска только одним дальним КА (и одним и более 
околоземным КА) можно получить область локализации с площадью несколько 
сотен квадратных угловых минут. Для всплесков, зарегистрированных только KW, 
\textit{INTEGRAL} (SPI-ACS) и одним или более околоземным КА, получаются 
протяженные области локализации.

\item С использованием описанной методики были получены локализации 146 гамма-всплесков,
зарегистрированных \textit{Fermi}~(GBM) за период с 12 июля 2008~г. по 11 июля 2010~г.
Было установлено, что IPN-триангуляции существенно улучшают локализации всплесков 
по сравнению с GBM, сокращая площадь области локализации всплеска в $\lesssim 180$~раз.

\item Описанная методика была успешно применена для подтверждения оптических послесвечений,
зарегистрированных системой телескопов для поиска транзиетов Паломарской обсерватории.
\end{enumerate}

По материалам Главы~\ref{IPN_catalog} на защиту выносится следующее положение:
\begin{itemize}
\item Каталог локализаций коротких гамма-всплесков, зарегистрированных в эксперименте
    Конус-Винд с 1994~г. по 2010~г.
\end{itemize}

Результаты, описанные в главе, отражены в следующих публикациях:
\begin{enumerate}
\item V.\,D.\,Pal'shin, K.\,Hurley, D.\,S.\,Svinkin et al. Interplanetary Network Localizations of
Konus Short Gamma-Ray Bursts // Astrophys.~J.~Suppl. 2013. Vol.~207. id~38;
\item K.\,Hurley, \dots\ , D.\,S.\,Svinkin et al. The~Interplanetary Network Supplement to 
the Fermi GBM Catalog of Cosmic Gamma-Ray Bursts // Astrophys.~J.~Suppl. 2013. Vol.~207. id~39;
\item Leo\,P.\,Singer, \dots\ , D.\,Svinkin, et al. The~Needle in the 100 deg$^2$ Haystack: 
Uncovering Afterglows of Fermi GRBs with the Palomar Transient Factory // 
Astrophys.~J. 2015. Vol.~806 p.~52.
\end{enumerate}

\clearpage

### 2025-10-23 disser/part4.tex
disser

\chapter{Поиск гигантских вспышек от мягких гамма-репитеров в ближайших галактиках 
         среди коротких всплесков Конус-Винд} \label{SGR_GF_search}

\section{Введение}
Мягкие гамма-репитеры (SGRs) относятся к редкому классу нейтронных звёзд, проявляющих 
два типа активности в жестком рентгеновском диапазоне ($\sim 10\textrm{--}1000$~кэВ). 
Во время периода активности SGRs испускают короткие ($\sim0.001\textrm{--}1$~c) жесткие рентгеновские всплески 
с пиковой светимостью $10^{38}\textrm{--}10^{42}$~эрг~с$^{-1}$. Фаза активности может длиться 
от дней до года, после чего наступает длительная фаза затишья. Значительно реже, 
возможно, один раз за время нахождения нейтронной звезды в стадии SGR, SGR может 
производить гигантские вспышки (GF), во время которых высвобождается значительная 
энергия $\sim(0.01\textrm{--}1)\times 10^{46}$~эрг. Гигантская вспышка начинается 
с короткого ($\sim 0.2\textrm{--}0.5$~с) жесткого импульса с быстрым нарастанием, 
порядка нескольких миллисекунд, и более медленным спаданием, который переходит 
в длинный затухающий хвост, модулированный вращением нейтронной звезды. 
Пиковая светимость начального импульса может достигать $\sim 10^{47}$~эрг~с$^{-1}$.
Подробное описание наблюдательных свойств SGRs дано в работе~\citep{Mereghetti2013}.

На конец 2015~г. известно 15 SGR~\citep{Olausen_Kaspi2014}, из которых 14 
находятся в нашей Галактике и один расположен в Большом Магеллановом Облаке.

Первая гигантская вспышка была зарегистрирована от SGR в Большом Магеллановом Облаке 
5~марта 1979~года приборами Конус на советских межпланетных станциях 
<<Венера-11 и -12>>~\citep{Golenetskii1979SvAL, Mazets1979} 
и сетью IPN~\citep{Barat1979, Cline1980, Evans1980, Cline1982}. 
На конец 2015~г. гигантские вспышки наблюдались только у трёх источников 
SGR~0526$-$66, SGR~1900$+$14 и SGR~1806$-$20, все они были зарегистрированы 
приборами Конус и сетью IPN. Две недавние вспышки от SGR~1900$+$14 и~SGR~1806$-$20 
сопровождались возрастанием вспышечной активности~\citep{Mazets1999a, Frederiks2007}.

Все известные SGR являются быстро замедляющимися рентгеновскими пульсарами 
с периодами 2--12~с и рентгеновскими светимостями в спокойном состоянии 
$(0.1\textrm{--}1)\times 10^{36}$~эрг~с$^{-1}$. Считается, что SGR принадлежат к 
более широкому классу магнетаров. Этот класс также включает аномальные 
рентгеновские пульсары (AXP) и радиопульсары с сильным магнитным полем (high-B radio pulsars).
Граница между SGR и AXP размыта из-за наличия аномально высокой рентгеновской светимости SGR,
не объясняемой в рамках существующих моделей остывания нейтронных звёзд, 
и регистрации жестких рентгеновских вспышек от AXPs. Примерно половина известных 
магнетаров отождествлена с областями активного звездообразования или остатками сверхновых.
Пространственное распределение магнетаров в Галактике схоже с распределением 
наиболее массивных звёзд класса O~\citep{Olausen_Kaspi2014}. Считается, что 
активность магнетаров связана с наличием у них сверхсильного магнитного поля,
оцененного из высокой скорости замедления SGR, 
$\sim 10^{13}\textrm{--}10^{15}$~Гс,~\citep{Duncan_and_Thompson_1992ApJ,Thompson_and_Duncan_1995MNRAS,Thompson_and_Duncan_1996ApJ}.
Альтернативные модели, описывающие активность SGR и AXP, приведены в работе~\citep{Bisnovatyi-Kogan_2014ARep}.

Благодаря огромной светимости начального импульса гигантские вспышки возможно  
регистрировать от SGR, расположенных в ближайших галактиках. При этом начальный 
импульс будет неотличим от короткого гамма-всплеска. Идея о возможности наблюдения 
гигантских вспышек в ближайших галактиках впервые была высказана 
в работах~\citep{Mazets1981,Mazets1982}. Оценки частоты гигантских вспышек от одного 
SGR и их доля среди коротких всплесков важны для понимания механизма генерации 
гигантских вспышек в рамках модели магнетара.

К настоящему времени обнаружено четыре коротких гамма-всплеска, которые могут 
являться GF в ближайших галактиках. У всплеска GRB~970110~\citep{Crider2006} 
по данным BATSE после начального импульса с длительностью $\approx 0.4$~с была 
обнаружена пульсирующая компонента с периодом 13.8~с на интервале 100~с. В область 
локализации этого всплеска попадает только одна близкая галактика NGC~6946 
на расстоянии 5.9~Мпк.

В работе~\citep{Levan2008} обсуждается возможность того, что всплеск GRB~050906, 
зарегистрированный \textit{Swift}, является GF в галактике 
IC~328 на расстоянии 138~Мпк. Свидетельством в пользу этой гипотезы авторы считают  
отсутствие детектирования спадающего рентгеновского и оптического послесвечений этого всплеска, 
однако, это не кажется удивительным, так как всплеск является наиболее слабым всплеском 
в каталоге \textit{Swift}-BAT~\citep{Sakamoto2011ApJS}. При этом 
авторами не отвергается возможность принадлежности источника всплеска к скоплению 
галактик на $z = 0.43$.

Всплеск GRB~051103, зарегистрированный сетью IPN, считается 
кандидатом в GF из группы галактик M81/M82, находящийся на расстоянии 
3.6~Мпк~\citep{Ofek_2006ApJ, Frederiks_2007AstLett, Hurley2010}. 
Этот всплеск представляет собой короткий импульс длительностью 170~мс с быстрым 
нарастанием ($<6$~мс) и экспоненциальным спадом с постоянной времени $\approx 55$~мс. 
Спектр наиболее интенсивной части всплеска описывается моделью 
CPL с $\alpha \approx 0.1$ и $E_\rmn{p} \approx 2.4$~МэВ. Локализация всплеска 
покрывает большое число рентгеновских источников в группе галактик M81/M82. Однако, 
в работе~\citep{Hurley2010} приводятся доводы против гипотезы о GF, в основном 
на основании гигантской пиковой светимости вспышки $\approx 4.7\times10^{48}$~эрг~с$^{-1}$,
если предположить, что источник находился в в группе галактик M81/M82.
Эта величина на порядок больше пиковой светимости вспышки от SGR~1806$-$20, которая 
составляла $(2\textrm{--}5)\times 10^{47}$~эрг~с$^{-1}$ в предположении 
расстояния до SGR 15~кпк.

Другой всплеск GRB~070201~\citep{Mazets_2008ApJ,Ofek_2008ApJ} имеет время нарастания $\approx 25$~мс 
и длительность $\approx 180$~мс. Спектр всплеска хорошо описывается степенным законом 
с экспоненциальным завалом~(формула~\ref{eq:CPL}) с $\alpha\approx-0.6$ и $E_\rmn{p}\approx 280$~кэВ. 
Область локализации этого всплеска накладывается на галактику M31 (0.78~Мпк). У всплеска обнаружено 
мягкое послесвечение в диапазоне 17--70~кэВ на интервале до 94~с после начального 
импульса, которое было интерпретировано как хвост GF.

Пределы на долю GF среди коротких гамма-всплесков были получены в нескольких 
работах~\citep{Lazzati2005,Palmer2005,Nakar2006ApJ,Popov2006,Ofek_2007ApJ,Tikhomirova2010AstL} 
и составили 1--15\%, см.~работу~\citep{Hurley2011} с подробным обзором предыдущих результатов.

Каталог локализаций коротких гамма-всплесков KW~\citep{Palshin2013}, 
описанный в предыдущей главе, содержит информацию о локализации 271 короткого 
всплеска, зарегистрированного за 16~лет практически непрерывных наблюдений всей 
небесной сферы. Этот каталог позволил осуществить поиск кандидатов в GF 
в наибольшем на 2014~г. наборе точно локализованных коротких всплесков.

В разделе~\ref{KW_sensitivity} обсуждается чувствительность KW и IPN к GF.
В разделе~\ref{Gal_sample} приводится набор близких галактик и обсуждаются их свойства. 
В разделе~\ref{GF_search} описывается поиск близких галактик в областях локализации 
коротких всплесков. На основании результатов этого поиска в разделе~\ref{GF_rate} 
вычисляется верхний предел на частоту гигантских вспышек от SGRs. Заключительные 
ремарки приведены в разделе~\ref{Summary}.

\section{Чувствительность KW и IPN}\label{KW_sensitivity}
Начальный импульс гигантской вспышки из близкой галактики вызовет срабатывание 
триггера на временном масштабе 140~мс. Среди трёх зарегистрированных гигантских 
вспышек ни одна не имеет достоверных прямых измерений параметров начального импульса 
из-за экстремальных потоков падающего излучения, перегружающих измерительные 
тракты всех гамма-детекторов. Значение потока, приводящего к насыщению детекторов 
KW составляет примерно $2.4 \times 10^{-2}$~эрг~см$^{-2}$~с$^{-1}$~\citep{Mazets1999a}.

Параметры начального импульса GF от SGR~1806$-$20 были определены благодаря 
регистрации излучения вспышки, отраженного от Луны~\citep{Frederiks2007}. 
Спектр излучения хорошо описывается моделью CPL с параметрами
$\alpha=-0.73_{-0.64}^{+0.47}$ и $E_{0}=666^{+1859}_{-368}$~кэВ 
(что соответствует энергии максимума $\nu F_{\nu}$ спектра~--- 
$E_{\rmn{p}} = (2-\alpha)E_0 = 850^{+1259}_{-303}$~кэВ). В работе~\citep{Terasawa2005} 
был восстановлен только временной профиль этой гигантской вспышки. 
Для двух других вспышек удалось получить только грубые оценки на $E_{\rmn{p}}$: 
$\sim 400\textrm{--}500$~кэВ (SGR~0526$-$66,~\citep{Golenetskii1979SvAL, Mazets1979}) 
и $>250$~кэВ (SGR~1900+14,~\citep{Hurley1999, Mazets1999}).

В предположении, что вся энергия начального импульса выделяется на коротком 
триггерном масштабе (140~мс), был оценен минимальный интегральный поток $S_\rmn{min}$ 
в диапазоне 20~кэВ--10~МэВ, который даст превышение скорости счёта над фоном на $9 \sigma$ 
в энергетическом диапазоне G2 (50--200~кэВ) при скорости счёта фона 400~отсч~с$^{-1}$. 
На основании полученного значения $S_{\rmn{min}}$, предельное расстояние регистрации 
вычислялось по формуле $d_{\rmn{max}} = \sqrt{Q/(4\pi S_{\rmn{min}})}$, где 
$Q$~[эрг]~--- энерговыделение начального импульса.

Минимальный поток $S_\rmn{min}$ сильно зависит от жесткости всплеска ($\alpha$ 
и $E_{\rmn{p}}$). Известный диапазон $E_\rmn{p}$ начальных импульсов 
GF~--- 200~кэВ--1~МэВ соответствует диапазону 
$S_\rmn{min}= (2.1\textrm{--}5.7)\times 10^{-7}$~эрг~см$^{-2}$, 
см.~рис.~\ref{img:KW_lim_distance}.

\begin{figure}[h] 
    \center
    \includegraphics [width=0.8\textwidth] {gLimDist18to30RU.eps}
    \caption[Зависимость минимального интегрального потока за 140~мс 
    (20~кэВ--10~МэВ) от пиковой энергии спектра.]
    {Зависимость минимального интегрального потока за 140~мс (20~кэВ--10~МэВ) 
и предельного расстояния, в предположении энерговыделения $Q=2.3\times 10^{46}$~эрг,
от параметров спектральной модели CPL $E_\rmn{p}$ и $\alpha$: сплошная линия~--- $\alpha=-1$, 
штриховая линия~--- $\alpha=-1.5$, пунктирная линия~--- $\alpha=-0.5$. 
Штрихованная область~--- диапазон предельных расстояний $\approx 18\textrm{--}30$~Мпк 
соответствующий диапазону $E_\rmn{p} = 250$~кэВ--1~МэВ.}
 \label{img:KW_lim_distance}
\end{figure}

Также была исследована зависимость площадей областей локализаций IPN коротких   %<--скольки областей?
всплесков от потока, измеренного KW в диапазоне 20~кэВ--10~МэВ на интервале 140~мс 
с наибольшей скоростью счёта, при этом не было обнаружено существенной корреляции, 
см.~рис.~\ref{img:IPN_box_area}. Таким образом, можно использовать полученный интервал 
значений $S_\rmn{min}$ для всей IPN.

\begin{figure}[h]
    \center
    \includegraphics [width=0.8\textwidth] {gAreaVsFluenceRU.eps}
    \caption{Площадь области локализации IPN в зависимости от интегрального потока 
    (20~кэВ--10~МэВ) на интервале 140~мс с максимальной скоростью счёта.}
    \label{img:IPN_box_area}
\end{figure}

Оценки расстояния до SGR~1806$-$20 находятся в диапазоне примерно от 6~кпк 
до~19~кпк~\citep{Tendulkar2012ApJ}, по самым последним оценкам~\citep{Svirski2011} 
диапазон расстояний составляет 9.4--18.6~кпк.

Изотропное энерговыделение начального импульса GF от SGR~1806$-$20 
($Q = 2.3\times 10^{46} d_{15}^2 $~эрг) предполагает предельное расстояние 
детектирования $d_\rmn{max} = (18\textrm{--}30)\times d_{15}$~Мпк, 
где $d_{15}=d/15$~кпк и $d$~--- расстояние до SGR~1806$-$20. 
Менее интенсивные GF с энерговыделением $Q \approx 10^{45}$~эрг 
(сравнимые со вспышкой 5 марта 1979~г. от SGR~0526$-$66) 
могут быть зарегистрированы до расстояний $d_\rmn{max} = (3.8\textrm{--}6.3) Q_{45}^{0.5}$~Мпк, 
где $Q_{45}$~--- энерговыделение вспышки в единицах $10^{45}$~эрг. 
Диапазон значений $d_\rmn{max}$ является трансляцией диапазона $S_\rmn{min}$.

\section{Набор близких галактик}\label{Gal_sample}
Наиболее полным каталогом, содержащим параметры близких галактик является 
каталог для поиска источников гравитационных волн (Gravitational Wave Galaxy Catalogue, 
GWGC~\citep{White2011CQGra}). Каталог содержит более 53000 галактик на расстояниях 
до~100~Мпк. Для поиска возможных родительских галактик гигантских вспышек SGR 
изначально был выбран набор из 8112 галактик на расстояниях до~30~Мпк. 
Неопределённость расстояний до галактик в наборе составляет 15--22\%.

Используя метод предложенный в работе~\citep{Ofek_2007ApJ}, была оценена полнота набора, 
связанная с затенением Галактикой. Было построено распределение галактик по Галактической 
широте с шагом $10^\circ$. Разность числа галактик в интервале 
$0^\circ\textrm{--}10^\circ$ и числа галактик в незатенённом интервале 
$10^\circ\textrm{--}20^\circ$, отнесённая к общему числу галактик, даёт долю 
потерянных галактик 6\% и полноту набора галактик $\epsilon_{\rmn{G}}=94$\%.

В предположении, что все SGR~--- молодые изолированные нейтронные звёзды, считалось, 
что число SGR пропорционально частоте вспышек сверхновых с коллапсом ядра 
(CCSN, типы Ib/c и~II) в галактике.

Следуя работам~\citep{Cappellaro1999, Boser2013} считалось, что частота вспышек 
сверхновых пропорциональна светимости галактики в фильтре 
$B$\footnote{Параметры фильтра $B$: $\lambda=445$~нм и $\rmn{FWHM}=94$~нм~\citep{Binney1998GalAstr}} 
($L_{B}$) $R_{\rmn{SN}} = k L_{B}$, 
где $k$ множитель зависящий от морфологического типа галактики данный 
в единицах SNu, см.~таб.~\ref{tab:RateCCSN}.

\begin{table} [h]
  \centering
  \scriptsize
  \parbox{15cm}{\caption{Ожидаемая частота вспышек CCSN в зависимости от типа галактики}
  \label{tab:RateCCSN}}
  \begin{tabular}{ccc}
  \hline
  \hline
  Тип галактики & Числовой тип &  Частота вспышек CCSN ($k$) в SNu \\
  по Хабблу     &  по Хабблу   &           \\     
  \hline
   E-S0           & от $-6$ до $-1$ &  $<0.05$ \\
   S0a-Sb         &  $0$--$3$   &  $0.89\pm0.33$ \\
   Sbc-Sd         &  $4$--$7$   &  $1.68\pm0.60$ \\
   Sm, Irr., Pec. &  $8$--$10$  &  $1.46\pm0.71$\\
  \hline
  \end{tabular}
\end{table}

Единица SNu соответствует $1\rmn{SN}(100\rmn{год})^{-1}(10^{10}L_{\odot B})^{-1}$, 
где $L_{\odot B} = 2.16\times10^{33}$~эрг~с$^{-1}$ светимость Солнца в фильтре~$B$. 
Светимость галактики $L_{B}$ вычислялась по формуле 
$L_{B}=10^{-0.4(M_B-M_{\odot B})} L_{\odot B}$, 
где $M_{\odot B}=5.48$ абсолютная звёздная величина Солнца в фильтре~$B$.

Исходный набор 8112 галактик содержит 790 галактик, для которых не дана $L_{B}$,
таким образом полнота набора по $L_{B}$ составляет $\epsilon_{L} \approx 90$\%. 
Среди оставшихся 7322 галактик с указанным $L_{B}$, для 2405 не указан морфологический тип. 
Яркость этих галактик в среднем меньше на три звёздные величины по сравнению 
с классифицированными галактиками, при этом они содержат меньше 7\% суммарной 
частоты вспышек сверхновых, поэтому эти галактики были исключены из дальнейшего 
рассмотрения.

В качестве итогового набора галактик был взяты 1896 галактики поздних типов 
(все кроме E и S0) с наибольшими $R_{\rmn{SN}}$, содержащие $\epsilon_{\rmn{SN}}=90$\% 
от суммарной частоты вспышек сверхновых. Плотность этих галактик на небесной сфере 
составляет 0.046~градус$^{-2}$. Суммарная частота вспышек сверхновых составляет 
$R_\rmn{SN}=22.8 \pm 0.4$~год$^{-1}$.

Для проверки методики определения $R_{\rmn{SN}}$, объемная плотность частоты 
вспышек сверхновых $R_\rmn{SN}(d)/(4/3\pi d^3)$, где $d$~--- расстояние, 
для набора 1896 галактик сравнивалась 
с нижним пределом $1.9_{-0.2}^{+0.4}\times 10^{-4}$~год$^{-1}$~Мпк$^{-3}$, 
полученным в обзоре сверхновых~\citep{Mattila2012} в галактиках ближе 15~Мпк. 
Зависимость плотности $R_{\rmn{SN}}$ от расстояния представлена на рис.~\ref{img:RateCCNvsDist}. 
Объёмная плотность, полученная на основе голубой светимости галактик согласуется в 
пределах $1 \sigma$ с наблюдаемой величиной, предполагая, что $\approx 19$\% близких CCSN 
были пропущены оптическими обзорами неба.

Объёмная плотность частоты вспышек CCSN демонстрирует значимый спад на расстояниях 
свыше $\approx 22$~Мпк, что может быть связано с падением полноты набора галактик 
на б\'{о}льших расстояниях. Для оценки объёмной частоты вспышек CCNS на расстояниях 
больших 22~Мпк было использовано среднее значение в диапазоне расстояний 
до 22~Мпк~--- $(2.74 \pm 0.18) \times 10^{-4}$~год$^{-1}$~Мпк$^{-3}$, 
ошибка дана на уровне значимости $1\sigma$.

Так же было обнаружено увеличение $R_{\rmn{SN}}/V$ внутри $\sim 10$~Мпк вплоть до 
$(9.3 \pm 0.16) \times 10^{-4}$~год$^{-1}$~Мпк$^{-3}$ внутри объёма 5~Мпк. 
При этом всего пять галактик содержат 25\% общей частоты CCSN:
PGC047885 на расстоянии $d = 5$~Мпк, IC~0342 на расстоянии 3.28~Мпк, NGC~6946 на 
расстоянии 5.9~Мпк, NGC~5457~(M101) на расстоянии 6.7~Мпк и  NGC~5194~(M51) 
на расстоянии 5.9~Мпк. Эти галактики являются наиболее вероятными  источниками 
гигантских вспышек SGR в ближайшей Вселенной в дополнение к предложенным ранее 
в работе~\citep{Popov2006}: M82 на расстоянии $d = 3.4$~Мпк, NGC~253 на расстоянии 2.5~Мпк, 
NGC~4945 на расстоянии 3.7~Мпк и M83 на расстоянии 3.7~Мпк.

\begin{figure}[h]
    \center
    \includegraphics [width=0.8\textwidth] {gRsn2VpubRU.eps}
    \caption[Удельная частота вспышек CCSN в зависимости от расстояния]
	{Удельная частота вспышек CCSN в зависимости от расстояния $d$. 
	Сплошная линия~--- $R_{\rmn{SN}}/V(d)$ для 1896 галактик, 
	содержащих 95\% $R_{\rmn{SN}}$ внутри 30~Мпк, из каталога GWGC. 
	Горизонтальные штриховые линии обозначают $\pm 1\sigma$ интервал для 
	локальной удельной частоты вспышек сверхновых 
	$2.3_{-0.2}^{+0.5}\times 10^{-4}$~год$^{-1}$~Мпк$^{-3}$ из работы~\citep{Mattila2012}, 
    предполагая, что доля пропущенных в обзоре сверхновых составляет 0.189.}
    \label{img:RateCCNvsDist}
\end{figure}

\section{Поиск гигантских вспышек среди коротких гамма-всплесков, 
зарегистрированных Конус-Винд}\label{GF_search}

Каталог локализаций коротких всплесков Конус-Винд, полученных при помощи 
сети IPN~\citep{Palshin2013}, содержит 271 всплеск, зарегистрированный 
по крайней мере одним КА IPN, что дало возможность локализовать их при помощи 
триангуляции (см.~Главу~\ref{IPN_catalog}). Этот набор содержит 30 всплесков, 
классифицированных как короткие гамма-всплески с продлённым излучением.

Процедура поиска наложений галактик и локализаций всплесков была проведена 
для нескольких поднаборов всплесков.  При поиске наложений галактика моделировалась 
кругом с центром, взятым из каталога, и диаметром, равным большой полуоси галактики. 
Оценка ожидаемого числа галактик в этом наборе областей локализации была выполнена 
методом Монте-Карло. Было сгенерировано 1000 реализаций набора галактик, в которых 
центры галактик выбирались случайным образом. Для каждого набора вычислялось число 
наложений галактик на локализации. Полученные числа сортировались по возрастанию 
и в качестве границ 95\% доверительного интервала для числа наложений 
бралось 25-е и~975-е число.

Наибольший набор включал 140 всплесков с площадями областей локализации 
меньше 10~кв.~градусов.
Было обнаружено, что на 12 из 140 IPN локализаций с общей площадью 217~кв.~градусов 
накладывается 20 галактик, при этом ни одна локализация всплеска с продлённым 
излучением не содержит галактики. Для этого набора локализаций число галактик, 
ожидаемое для случайного наложения, равно 22--44 на уровне значимости 95\%. 
Также было обнаружено, что локализация только одного всплеска (GRB~050312) 
накладывается на окраину скопления Девы, при этом локализация не накладывается 
ни на одну галактику из набора. Здесь скопление девы моделировалось кругом с центром в 
$\rmn{R.A.}=188^\circ$, $\rmn{Dec.}=12^\circ$ с радиусом $6^\circ$, 
параметры скопления были взяты из работы~\citep{Binggeli1987}.

Поиск наложений показал, что только два всплеска имеют малую вероятность случайного 
наложения на близкую галактику $P_\rmn{chance} \sim 1$\%. Эти всплески ранее были 
отнесены к внегалактическими GF: GRB~051103, чья локализация накладывается на группу галактик M81/M82 
(площадь бокса $4.3\times10^{-3}$~кв.~градусов) и GRB~070201, считающийся GF в галактике 
Андромеды (площадь бокса 0.123~кв.~градусов). Вероятность $P_{\rmn{chance}}$ 
соответствует обнаружению как минимум одной галактики в заданной локализации (боксе).

Затем процедура поиска была применена к набору 98 локализаций с площадью меньше 
1~кв.~градус, который содержит всплески, зарегистрированные по крайней мере одним 
удалённым космическим аппаратом~(см. главу~\ref{IPN_catalog}). Было обнаружено, 
что только локализации двух упомянутых выше всплесков содержат галактики.

Общей чертой всех известных GF является малая длительность начального импульса 
$\lesssim 500$~мс и малое время нарастания импульса $t_{\rmn{r}} \lesssim 25$~мс. 
Среди 296 коротких всплесков 40 имеют $t_{\rmn{r}}<25$~мс и длительность $<500$~мс. 
Ранее обнаруженные кандидаты GRB~051103 и GRB~070201 имеют $t_{\rmn{r}}=2$~мс 
и~$t_{\rmn{r}}=24$~мс соответственно. Среди коротких всплесков KW этим 
критериям удовлетворяют 17 событий с площадями областей локализации менее 10~кв.~градусов.
Было обнаружено, что четыре области локализации с общей площадью 47~кв.~градусов накладываются на пять галактик. 
Это число попадает в 95\% доверительный интервал для случайного наложения, 5--16 галактик. Из этих 
четырёх всплесков только GRB~051103 и GRB~070201 имеют малые вероятности случайного наложения.
Результаты поиска наложений для всех перечисленных наборов всплесков приведены 
в Таблице~\ref{tab:SearchResults}.

\begin{table}[h]
  \centering
  \scriptsize
  \caption{Результаты поиска кандидатов в гигантские вспышки SGR в 
  наборе коротких всплесков Конус-Винд}
  \label{tab:SearchResults}
  \begin{tabular}{cccc}
  \hline
  \hline
Описание & Число локализаций  & Число галактик & Ожидаемое число  \\
набора   &                    & в локализациях & наложений на уровне 95\% \\
\hline
Площадь локализации $<10$~кв.~градусов  & 140 & 20 & 22--44 \\
Площадь локализации $<1$~кв.~градусов   & 98 & 2 & 0--7\\
$t_{\rmn{r}} \leq 25$~мс, $T_{100}<500$~мс, & 17 & 5 & 5--16\\
и площадь локализации $<10$~кв.~градусов & & & \\
\hline
\end{tabular}
\end{table}

С учетом произведения факторов полноты набора галактик $\epsilon_{\rmn{G}} \epsilon_{\rmn{L}} \epsilon_{\rmn{SN}} \approx 76$\%, 
и предполагая, что было найдено два кандидата в GF среди 98 хорошо локализованных коротких всплесков, 
можно поставить верхний предел на долю GF среди коротких всплесков Конус-Винд 
$<8$\% (=6.296/98/0.76), где 6.296~--- 95\% односторонний верхний предел на 
число вспышек~\citep{Gehrels1986}. Благодаря непрерывному наблюдению всего неба IPN, 
этот предел может быть распространён на всю популяцию коротких гамма-всплесков с 
интегральными потоками выше $\sim 5\times 10^{-7}$~эрг~см$^{-2}$. Полученный верхний 
предел жестче чем полученный в работе~\citep{Ofek_2007ApJ}.

\section{Верхний предел на частоту гигантских вспышек}\label{GF_rate}
Предполагая, что только одна GF с энерговыделением $Q \gtrsim 10^{46}$~эрг наблюдалась 
в группе галактик M81/M82 внутри объёма $d \le 30$~Мпк, можно получить верхний предел 
на частоту подобных GF. Оценка была основана на предположении, что число активных SGR ($N_\rmn{SGR}(d)$) 
внутри сферы радиуса $d$ пропорционально частоте вспышек CCSN 
$R_\rmn{SN}(d)=4/3 \pi d^3 r_\rmn{SN}$, где $r_\rmn{SN}$~--- объёмная частота вспышек CCSN.
\begin{equation}\label{eq:NumSGR}
N_{\rmn{SGR}} (d) = \frac{N_{\rmn{SGR}, \rmn{MW+LMC}}}{ R_{\rmn{SN}, \rmn{MW+LMC}}} R_{\rmn{SN}}(d).
\end{equation}
Галактическая частота CCSN равна $R_{\rmn{SN}, \rmn{MW}} = 0.028\pm0.006$~в~год 
с систематической ошибкой $\sim 2$~раза~\citep{Li2011part3}, и частотой в Большом 
Магеллановом облаке (LMC)~$R_{\rmn{SN}, \rmn{LMC}} = 0.013\pm0.009$~в~год~\citep{Bergh1991}. 
Таким образом, суммарная частота равна $R_{\rmn{SGR}, \rmn{MW+LMC}} = 0.041\pm0.011$~в~год.

% Значение $R_{\rmn{SN}, \rmn{MW}} = 0.028\pm0.006$~в~год получено в Li et al., 2011
% экстраполяции R_SN полученной для галактик схожих с MW. Тип MW -Sbc 
% (спиральная с балджем и "неплотной намоткой рукавов").
% На самом деле R_SN в LMC ~1 в 1000 лет (0.001 yr-1) а не 0.013+/-0.009 yr-1, 
% определена по наблюдениям остатков сверхновых!!!

Наблюдаемая частота гигантских вспышек на SGR задаётся выражением
\begin{equation}\label{eq:RateGF}
R_{\rmn{GF}} = \frac{N_{\rmn{GF},\rmn{obs}}}{\Delta T N_{\rmn{SGR}}(d_{\rmn{max}})} ,
\end{equation}
где $N_{\rmn{GF},\rmn{obs}}$~--- число зарегистрированных GF, 
$\Delta T=16$~лет~--- время наблюдения KW на 2010~г. и $N_{\rmn{SGR}}(d_{\rmn{max}})$ 
задано уравнением~\ref{eq:NumSGR}. Для оценки верхнего предела на $R_{\rmn{GF}}$ 
в случае одной зарегистрированной GF использовался 95\% односторонний верхний предел
на $N_{\rmn{GF}, \rmn{obs}}=4.744$~\citep{Gehrels1986}.

Для частоты GF с энерговыделением $Q \gtrsim 10^{46}$~эрг в объеме $d\le30$~Мпк уравнение~\ref{eq:RateGF} 
даёт верхний предел ${(0.6\textrm{--}1.2)\times 10^{-4} Q_{46}^{-1.5}}$~год$^{-1}$~SGR$^{-1}$, 
где $Q_{46}$ энерговыделение GF в единицах $10^{46}$~эрг. Диапазон верхнего предела 
является трансляцией ошибок частоты вспышек CCSN. Регистрация только 
одной GF с энерговыделением $Q \gtrsim 10^{46}$~эрг за последние 35~лет 
с 1979~г от SGR~1806$-$20 предполагает частоту таких вспышек в галактике 
${(0.005\textrm{--}1)\times 10^{-2}}$~год$^{-1}$~SGR$^{-1}$ ($=1_{-0.98}^{+4.6} / 35 / 15 $) 
для одностороннего 95\% уровня значимости. Это величина согласуется с верхним пределом, 
вычисленным выше с учётом диапазона расстояний до SGR~1806$-$20 9.4--18.6~кпк, 
в основном из за больших неопределённостей в галактической частоте GF.

Для менее интенсивных GF с энерговыделением $Q \lesssim 10^{45}$~эрг, которые 
могут регистрироватся KW и IPN на расстояниях до 6.3~Мпк, считая что 
одна такая вспышка была зарегистрирована из галактики Андромеды, верхний предел 
на частоту вспышек составляет
${(0.9\textrm{--}1.7)\times 10^{-3}}$~год$^{-1}$~SGR$^{-1}$. Этот предел согласуется
с наблюдаемой галактической частотой таких вспышек 
${(0.5\textrm{--}1.4)\times 10^{-2}}$~год$^{-1}$~SGR$^{-1}$. Верхний предел и 
интервал галактической частоты GF дан на уровне значимости 95\%.

Дополнительно был вычислен верхний предел на частоту ярких GF с использованием 
данных \textit{Swift}-BAT. С момента запуска в ноябре 2004~г. \textit{Swift} наблюдал 
только один кандидат в GF, GRB~050906~\citep{Levan2008}, предположительно из 
галактики IC~328 на расстоянии $\approx 130$~Мпк. Этот всплеск имел наименьший 
интегральный поток в диапазоне 15--150~кэВ из всех зарегистрированных на 2011~г. всплесков, 
$S_{\rmn{min}} = 6.1\times10^{-9}$~эрг~см$^{-2}$~\citep{Sakamoto2011ApJS} и 
мягкий спектр, описываемый степенным законом с показателем~$-1.7$. Экстраполяция 
энерговыделения GF от SGR~1806$-$20 из диапазона 10~кэВ--10~МэВ в 15--150~кэВ, 
используя степенной закон с экспоненциальным завалом с параметрами $\alpha=-0.73$ 
и $E_{\rmn{p}}=850$~кэВ даёт $2.5\times10^{45}$~эрг, что соответствует предельному 
расстоянию детектирования 60~Мпк. Полученное предельное расстояние детектирования 
в совокупности с мягким спектром всплеска делает ассоциацию с SGR в IC~328 маловероятной.

Время наблюдения 90\% неба BAT в 2004--2010~гг. 
составило $7.25\times10^{6}$~с (0.23~года)~\citep{Baumgartner2013ApJS}, и экстраполяция 
этой величины на 2004--2013~гг. даёт 0.35~лет. Отсутствие зарегистрированных GF в 
объеме $d \lesssim 60$~Мпк в течение наблюдений \textit{Swift}-BAT даёт верхний предел 
на частоту GF на уровне 95\% $\sim 6 \times 10^{-4} Q_{45}^{-1.5} $~год$^{-1}$~на~SGR, 
где $Q_{45}$~--- энерговыделение GF в единицах $10^{45}$~эрг в диапазоне 15--150~кэВ. 
Таким образом, не смотря на высокую чувствительность BAT, полученный предел 
менее жесткий, чем полученный по данным KW и IPN из-за меньшей экспозиции всего неба.

\section{Заключение}\label{Summary}
В данной главе были получены следующие результаты:
\begin{enumerate}
\item Оценена чувствительность KW и IPN, и получено 
предельное расстояние регистрации гигантских вспышек SGR схожих с GF от SGR~1806$-$20 
равное $(18\textrm{--}30) d_{15}$~Мпк. Показано, что менее интенсивные GF, сравнимые 
с GF от SGR~1900+14 и SGR~0526$-$66 могут быть зарегистрированы IPN в галактиках 
не далее $\approx 6$~Мпк.

\item Произведён поиск близких галактик, находящихся ближе 30~Мпк, в локализациях 
коротких гамма-всплесков KW. Были обнаружены только два всплеска, ранее 
ассоциированые с группой галактик M81/M82 (GRB~051103) и галактикой Андромеды (GRB~070201),
локализации которых имеют малую вероятность случайного наложения на эти галактики.
Дополнительный поиск всплесков из скопления Девы не выявил возможных кандидатов в GF.

\item Получен верхний предел на частоту GF с энегрговыделением $Q \gtrsim 10^{46}$~эрг равный
${(0.6\textrm{--}1.2)\times 10^{-4} Q_{46}^{-1.5}}$~год$^{-1}$~на~SGR, который предполагает 
появление около одной GF с таким энерговыделением за время активности SGR, $10^3\textrm{--}10^5$~лет. 
Этот предел был вычислен на основе наибольшего на 2014~г.  
набора коротких всплесков и согласуется с ранее полученной в работе~\citep{Ofek_2007ApJ} оценкой.

Для GF, сопоставимых по энерговыделению со вспышкой 5 марта~1979~г. ($Q \lesssim 10^{45}$~эрг), 
полученный верхний предел на порядок выше~--- $(0.9\textrm{--}1.7)\times 10^{-3}$~год$^{-1}$~SGR$^{-1}$. 
Что может быть интерпретировано, как возможность наблюдать более одной подобной GF за время жизни SGR.
Полученные верхние пределы содержат неопределённость в порядок величины, связанную с
неопределённостью галактической частоты вспышек CCSN, расстояния до SGR~1806$-$20 и
предельного расстояния детектирования IPN.
Оцененные дипольные  магнитные поля SGR~1900+14 и SGR~0526$-$66, равные
$5.6\times10^{14}$~Гс и $7\times10^{14}$~Гс, соответственно~\citep{Olausen_Kaspi2014}, 
по-видимому, являются достаточными для генерации десятка GF с энерговыделением $Q \sim 10^{45}$~эрг,
что согласуется с полученными оценками на частоту появления GF.

\item Определены галактики, которые являются наиболее вероятными источниками GF 
из-за наибольшего оцененного количества SGR в этих галактиках. Это галактики
PGC047885, IC~0342, NGC~6946, NGC~5457 и NGC~5194, в дополнении к предложенным 
в работе~\citep{Popov2006}.
\end{enumerate}

По материалам Главы~\ref{SGR_GF_search} на защиту выносится следующее положение:
\begin{itemize}
\item Результаты поиска гигантских вспышек от мягких гамма-репитеров 
    в близлежащих галактиках по данным в эксперимента Конус-Винд. 
\end{itemize}

Результаты, полученные в главе, отражены публикации\\
D.~S. Svinkin, K. Hurley, R.~L. Aptekar, S.~V.~Golenetskii, D.~D.~Frederiks  
A~search for giant flares from soft gamma-repeaters in nearby galaxies in the 
Konus-Wind short burst sample // Mon.~Not.~R.~Astron.~Soc. 2015. Vol.~447,~1. p.~1028

\clearpage

### 2025-10-23 disser/part5.tex
disser

\chapter{Спектральный анализ коротких всплесков Конус-Винд} \label{sGRB_spectral_catalog}
\section{Введение}
Настоящая глава посвящена спектральному анализу 295 коротких всплесков Конус-Винд (KW),
локализация которых была описана в Главе~\ref{IPN_catalog}.

%В результате анализа было установлено, что для описания спектров некоторых всплесков необходима дополнительная
%степенная компонента, также обнаружено, что, вероятно, продлённое излучение значительной доли всплесков
%имеет достаточно жесткий спектр.

Глава организована следующим образом.
В разделе~\ref{sec:DATA ANALYSIS} дано описание методики спектрального анализа и
методики определения интегральных ($S$) и пиковых ($F_\rmn{peak}$) энергетических потоков.
В разделе~\ref{sec:RESULTS} приведены результаты анализа, представлены распределения параметров всплесков.
В разделе~\ref{sec:SUMMARY} полученные результаты обсуждаются в контексте физической 
классификации гамма-всплесков.

\section{Методика}\label{sec:DATA ANALYSIS}
Для каждого всплеска на основе его локализации\footnotemark вычислялась матрица отклика детектора
(см. Главу~\ref{KW_description}).
Типичный спектр короткого гамма-всплеска по данным KW состоит из подгруппы первых 
четырех спектров с временем накопления 64~мс на интервале 
от 0 до 0.256~с относительно триггера. Около 10\% всплесков имеют значимое 
излучение в пятом спектре (0.256--8.192~с). Для 4\% ярких событий
границы интервалов накопления спектров с 5 и далее были изменены системой адаптации.

\footnotetext{Углы падения для всплесков даны в таблице~5.1 по 
ссылке \url{http://www.ioffe.ru/LEA/shortGRBs/Catalog2/Data/tables}.}

Спектр фона для коротких всплесков без продлённого излучения, как правило, 
брался на интервале $T_0+25$~с со временем накопления около 100~с, он же использовался для калибровки.
Для примерно четверти всплесков анализ многоканальных спектров был не возможен, из-за того 
что основная часть события лежала до начала измерения спектров. 
Для таких всплесков строился трехканальный спектр, с использованием отсчетов временной истории, 
накопленных за полную длительность всплеска $T_{100}$.
Из-за малого числа отсчётов (количества фотонов) для большинства всплесков 
$F_\rmn{peak}$ определялся на основе интегрального спектра. 
Только для 18 достаточно интенсивных событий можно было выделить спектр вблизи 
максимальной скорости счёта.
Итоговой набор спектров состоял из 214 интегральных многоканальных спектров 
(без учёта 18 спектров вблизи пиковой скорости счёта) и 79 трёхканальных. 
Два слабых всплеска: GRB19960325\_T69892 и GRB19980614\_T31854 имели сбои во 
временных историях и спектрах и были исключены из рассмотрения.

\subsection{Многоканальные спектры}
Спектры коротких всплесков моделировались тремя функциями: простой степенью (PL), степенью с экспоненциальным 
обрезанием (CPL) и функцией Банда (BAND), см.~Главу~\ref{KW_description}.
Процедура определения параметров моделей, описанная в Главе~\ref{KW_description}, 
для многоканальных спектров производилась в программе 
\texttt{XSPEC}\citep{Arnaud_1996ASPC} версии~12.8.0.
Для определения параметров моделей использовалась статистика $\chi^2$. 
Спектральные каналы были сгруппированы для получения не менее 10 отсчётов в канале для обеспечения 
близкой к гауссовой статистики отсчётов. Группировка по 20 и более отсчётов на канал 
приводила к исключению из анализа старших каналов спектра, 
что в некоторых случаях значительно сужало энергетический диапазон. 
В качестве нормировки модели использовался поток в диапазоне 10~кэВ--10~МэВ, 
который вычислялся с использованием модели \texttt{cflux} в \texttt{XSPEC}.
Ошибки спектральных параметров определялись для уровня 
значимости 90\%.
После определения параметров каждой из трёх спектральных моделей для спектра, 
выбиралась наиболее подходящая модель на основе различия полученного минимального $\chi^2$. 
Критерием для предпочтения модели с одним дополнительным параметром было $\Delta \chi^2 \geq 5$ 
с соответствующей вероятностью $\approx 0.025$. Было обнаружено, что такое значение предпочтительно 
для нашего набора всплесков по сравнению с часто используемым $\Delta \chi^2 \geq 6$,
так как оно даёт в два раза меньше расходящихся PL моделей, в качестве наилучших,
что критично для вычисления энергетики всплесков.

\subsection{Трёхканальные спектры}
Определение параметров PL и CPL моделей для 79 трёхканальных спектров производилось 
специально созданной программой. При этом ошибки параметров определялись методом Монте-Карло.

Для проверки методики было проведено сравнение результатов многоканального и трёхканального спектрального анализа
на наборе 214 всплесков с многоканальными спектрами.
Для каждого всплеска был взят  трёхканальный спектр, накопленный в течение 
интервала измерения многоканального спектра.
Затем сравнивались параметры модели, которые лучше всего описывают многоканальный спектр
с параметрами и той же модели, полученными по трёхканальному спектру.
В случае PL полученные фотонные индексы согласуются для двух типов
спектрального анализа. Для CPL модели было обнаружено, что  $\alpha$ также в целом
согласуется между двумя типами спектров. То же самое верно и для $E_\rmn{p}$, но
только в случае если значение попадает в энергетический диапазон, охватываемый трёхканальным спектром
($\lesssim 1$~МэВ), в противном случае трёхканальный анализ давал завышенное и 
плохо ограниченное значение $E_\rmn{p}$. Это показывает, что на основе данных KW 
можно получать достаточно точные значения спектральных параметров и энергетики,
даже когда многоканальные спектральные данные отсутствуют.

Поскольку аппроксимация трёхканального спектра моделью CPL имеет нулевое число степеней свободы
(и, в случае сходимости, $\chi_\rmn{CPL}^2=0$), наиболее подходящая модель 
не может быть выбрана на основе критерия $\Delta \chi^2 \geq 5$.
Поэтому, для того чтобы не переоценить энергетику всплеска, для вычисления $S$ и $F_\rmn{peak}$
использовалась модель CPL в случаях когда $E_\rmn{p}$ была ограничена снизу.

\section{Результаты}\label{sec:RESULTS}
\subsection{Спектральные параметры}
В результате спектрального анализа было получено, что 201, 9 и 4 интегральных 
спектра наилучшим образом описываются CPL, BAND и PL, соответственно. 
Так как ни один метод основанный на использовании функции правдоподобия не даёт 
вероятности того, что конкретная модель хорошо описывает данные, в опубликованном 
каталоге были также приведены все модели параметры которых ограничены.
Для моделей CPL и BAND требовалось чтобы и $\alpha$ и $E_\rmn{p}$ имели ограниченные ошибки,
и для модели BAND дополнительно требовалось $\beta > -4$. Для исключения моделей
с большими невязками дополнительно было наложено условие на вероятность 
случайного получения заданного значения статистики $\chi^2$ 
(вероятностью нулевой гипотезы) $P> 10^{-6}$. 
Суммарно были получены параметры 473 моделей
для интегральных спектров (210~--- CPL, 117~--- BAND и 146~--- PL).
Для всех трёхканальных спектров кроме одного были получены параметры модели CPL, для GRB19961113\_T80522 с 
неограниченной $E_\rmn{p}$ оценки энергетики делались на основе модели PL\footnotemark.

\footnotetext{Параметры моделей для многоканальных и трёхканальных спектров даны 
в таблице~5.2 по ссылке \url{http://www.ioffe.ru/LEA/shortGRBs/Catalog2/Data/tables}.}

На Рис.~\ref{fig:par_dist}, представлены распределения интегральных спектров по $E_\rmn{p}$ и $\alpha$.
Фотонные индексы $\alpha$ наиболее подходящих спектральных моделей распределены 
вокруг значения~$-0.5$. 
Около 66\% фотонных индексов $\alpha > -2/3$, нарушают <<линию смерти>> для 
синхротронной модели излучения~\citep{Preece_1998ApJL}, в тоже время только около 
1\% (три фотонных индекса для PL моделей) фотонных индексов $\alpha< -3/2$, 
нарушают предел охлаждения в синхротронной модели излучения. 
Для четырёх спектров, которые описываются степенной моделью, фотонные индексы 
находятся на мягком конце распределения $\alpha$. Для 9 моделей BAND индексы $\beta$
распределены вблизи значения $-2.3$.
Распределение по $E_\rmn{p}$ для CPL моделей имеет максимум около 500~кэВ 
и покрывает около двух порядков величины. 
Было исследовано различие значений $E_\rmn{p}$ между моделями BAND и CPL в наборе 
моделей с ограниченными параметрами и было обнаружено, что для каждого спектра значения согласуются
в пределах ошибок на уровне 90\%.

% Раскрыть смысл <<линии смерти>> и предел охлаждения в синхротронной модели излучения!

\begin{figure}
	\begin{minipage}[h]{0.5\textwidth}
		\center{\includegraphics[width=1\textwidth]{gDistAlpha_ru.eps} \\ а)}
    \end{minipage}
    \hfill
    \begin{minipage}[h]{0.5\textwidth}
		\center{\includegraphics[width=1\textwidth]{gDistEp_ru.eps} \\ б)}
	\end{minipage}
    \caption{
    Распределение $\alpha$~(а) и $E_\rmn{p}$~(б), полученных 
    в результате моделирования интегральных спектров. Распределение для каждой модели
    показано отдельно.
    \label{fig:par_dist} }
\end{figure}

\subsection{Всплески с дополнительной спектральной компонентой}
Для трёх ярких всплесков была получена низкая достоверность моделирования 
интегральных спектров ($P<0.001$) для всех трёх использованных моделей. 
Детальное рассмотрение этих всплесков выявило причину малых значений $P$.
Всплеск GRB20060306\_T55358, с $P \approx 10^{-4}$, имеет сильную 
спектральную эволюцию~--- жесткость спектра уменьшается в течении всплеска, 
что приводит к систематическому расхождению с модельными спектрами (BAND и CPL).
Однако, модель BAND хорошо описывает ($P>0.05$) отдельные спектры этого яркого всплеска.
Было обнаружено, что интегральные спектры двух других всплесков GRB19960908\_T25028 и GRB20031214\_T36655
хорошо описываются суммой функций CPL и PL.
Помимо перечисленных всплесков, явные систематические отклонения модели от спектра
были обнаружены для интегрального спектра GRB19980205\_T19785 ($P=0.08$), 
который также хорошо аппроксимируется суммой CPL+PL.

Параметры CPL+PL аппроксимаций интегральных спектров описанных всплесков приведены в Таблице~\ref{tab:extra_comp}. 
Степенная компонента этих спектров, которая также детектируется в большинстве отдельных спектров,
является достаточно жесткой ($\alpha \sim -2$) и даёт основной вклад в спектр на  
энергиях ниже $\sim 50$--100~кэВ, и должна доминировать на энергиях $\gtrsim 10$~МэВ.
Жесткая CPL компонента имеет $E_\rmn{p}\sim(1.5\textrm{--}2)$~МэВ и достаточно пологий
фотонный индекс $\alpha > -1$).
Все перечисленные всплески находятся среди 10\% наиболее интенсивных всплесков
в смысле интегральных энергетических потоков, при этом
GRB20031214\_T36655 и GRB20060306\_T55358 имеют наибольшие потоки в наборе.
Иллюстрация временной истории и спектров GRB20031214\_T36655 приведена на рис.~\ref{fig:extra_comp}.

\input{tables/extra_comp_parameters}

\begin{figure}
    \begin{minipage}[b]{0.5\textwidth}
		\center{\includegraphics[width=1\textwidth]{GRB20031214_T36655_2ms_rig} \\ а)}
    \end{minipage}
    \hfill
    \begin{minipage}[b]{0.5\textwidth}
		\center{\includegraphics[width=1\textwidth]{GRB20031214_T36655_spectra.eps} \\ б)}
	\end{minipage}
\caption{
    Временн\'{а}я история GRB20031214\_T36655~(а) и результаты спектрального анализа~(б).
    Вертикальными штриховыми линиями на панели~(а) показаны границы накопления спектров.
    На панели~(б) показаны восстановленные спектры, где пунктиром показана каждая спектральная компонента.    
    \label{fig:extra_comp} }
\end{figure}

\FloatBarrier
\subsection{Интегральные и пиковые потоки}
Значения $S$ и $F_\rmn{peak}$ вычислялись с использованием энергетического потока,
полученного для наиболее подходящей спектральной модели в диапазоне 10~кэВ--10~МэВ.
Так как интервал накопления спектра обычно отличался от $T_{100}$, применялась 
коррекция, которая учитывала излучение лежащие вне интервала накопления спектра 
для вычисления $S$.
Для коротких всплесков с продлённым излучением интегральные энергетические потоки 
вычислялись отдельно для EE и начального импульса (см. Раздел~\ref{sec:EE}).
Значение $F_\rmn{peak}$ вычислялось на масштабе 16~мс с использованием наиболее 
подходящей спектральной модели для спектра вблизи максимальной скорости счёта.
Для получения $F_\rmn{peak}$ поток по модели умножался на отношение 
пиковой скорости счёта на масштабе 16~мс к средней скорости счёта за интервал 
накопления спектра. Обычно для коррекции использовались отсчёты в группах каналов
G2+G3, G1+G2 только G2 или G1+G2+G3 в зависимости от жесткости спектра.
Значения $S$ и $F_\rmn{peak}$ были получены для 293 всплесков.
Распределения всплесков по $S$ и $F_\rmn{peak}$ показаны на рисунке~\ref{fig:fl_pf_dist}.

Следует отметить, что для нескольких очень интенсивных и сильно переменных всплесков 
(GRB19970704\_T04097, GRB20031214\_T36655, GRB20051103\_T33943, GRB20060306\_T55358,
GRB20070201\_T55390 и GRB20070222\_T27115)
значения $F_\rmn{peak}$ (и в меньшей степени $S$) могут быть недооценены на основе описанного анализа
в $\sim1.5$--2 раз, из-за того что мёртвое время в спектрах оценивалось на основе предположения о 
постоянстве скорости счёта в интервале накопления.

\begin{figure}
    \begin{minipage}[h]{0.5\textwidth}
		\center{\includegraphics[width=1\textwidth]{gDistS_ru.eps} \\ а)}
    \end{minipage}
    \hfill
    \begin{minipage}[h]{0.5\textwidth}
		\center{\includegraphics[width=1\textwidth]{gDistPF_ru.eps} \\ б)}
	\end{minipage}
\caption{
    Распределение интегральных~(а) и пиковых энергетических потоков~(б)
    Серым на каждой панели показано суммарное по всем моделям распределение для 214
    многоканальных спектров, распределения для каждой модели показаны отдельно. 
    Штрихпунктирной линией показаны распределения для 79 трёхканальных спектров.
    \label{fig:fl_pf_dist} }
\end{figure}
\FloatBarrier
\subsection{Короткие всплески с продлённым излучением}\label{sec:EE}
Продлённое излучение (EE), которое следует после короткого начального импульса (IP),
наблюдалось у некоторой части коротких всплесков, зарегистрированных различными экспериментами:
\textit{CGRO}-BATSE~\citep{Burenin_2000AstL, Norris_and_Bonnel_2006ApJ, Bostanci_2013MNRAS}, 
KW~\citep{Mazets_2002astroph, Frederiks_2004ASPC}, 
\textit{INTEGRAL}-SPI-ACS~\citep{Minaev_2010AstL}, 
\textit{Swift}-BAT~\citep{Norris_2011ApJ, Sakamoto_2011ApJS}, 
и \textit{Fermi}-GBM~\citep{Kaneko_2015MNRAS}. 
Поиск кандидатов в короткие всплески с EE был произведён в Главе~\ref{KW_GRB_classification}, 
было обнаружено 31 событие. Хотя яркий начальный импульс GRB~070207~\citep{Golenetskii_2007GCN6089}
удовлетворяет критериям короткого всплеска ($T_{50}=0.010\pm0.004$~с) с $E_\rmn{p}\sim 300$~кэВ,
очень яркое и жесткое ($E_\rmn{p}\sim 1.5$~МэВ последующие излучение, которое лишь формально может
считаться EE, предполагает, что это событие~--- длинный/жесткий всплеск с коротким прекурсором,
схожий по морфологии с двумя другими всплесками KW, GRB~000115 
и GRB~001020~\citep{Hurley_2000GCN859}.

%We defined the following search criteria: 
%the burst initial pulse should meet our criteria for a short GRB, i.e. have $T_{50}<0.6$~s;
%and the remaining part of burst (EE) shouldn't exhibit peaks with prominent spectral evolution. 
%Applying these criteria to the full KW sample, we found 31 short GRBs with EE candidates. 
%Only for 22 events EE was bright enough to allow spectral analysis.
%The initial pulses of these events are classified in Table~\ref{tab:durations} as Iee or Iee/II.

Только для 21 события, из оставшихся 30, интенсивность EE была достаточной для спектрального анализа.
Для 15 всплесков наилучшей спектральной моделью для EE является PL и для 6~--- CPL.   
Фотонные индексы модели PL находятся в диапазоне от $-2.6$ до $-1.4$ с медианой $-1.6$,
и в диапазоне от $-1.4$ до $-0.3$ с медианой $-1.2$ для модели CPL.
Значения $E_\rmn{p}$ находятся в диапазоне от $\approx160$~кэВ до $\approx 2.2$~МэВ с медианой $\approx300$~кэВ
и средним геометрическим $\approx370$~кэВ. 
Для указанного 21 всплеска было вычислено отношение потоков EE/IP, 
которое изменяется в пределах от 0.06 до 15 с медианой 3.3. 
Из шести всплесков, у которых EE описывается моделью CPL, у четырёх $E_\rmn{p}$ EE 
меньше чем у начального импульса и для двух всплесков, GRB19950526\_T16613 и GRB20090720\_T61379, 
наблюдается обратная ситуация, причем второй всплеск имеет экстремально
жесткое EE ($E_\rmn{p} = 2.2(-1.0,+2.4)$~МэВ).

\section{Обсуждение результатов}\label{sec:SUMMARY}
В главе представлены результаты спектрального анализа 293 коротких гамма-всплесков,
зарегистрированных в эксперименте Конус-Винд, этот набор составляет $\sim 15$\% 
от полного числа всплесков, зарегистрированных за первые 15 лет работы инструмента.
Набор включает $\sim 70$\% всплесков Типа~I, $\sim 8$\% Типа~II и $\sim 12$\%
имеют неопределённый тип (I или~II). Доля коротких всплесков с продлённым 
излучением составляет $\sim 10$\%.

Суммарно было проанализировано 253 многоканальных спектров: 214 интегральных,
18 пиковых и 21 спектр продлённого излучения; и 79 трёхканальных спектров.
Таблица~\ref{tab:pardist} содержит медианы и 90\% интервалы распределений спектральных
параметров и энергетики, полученных в главе. В первой колонке дано название параметра;
во второй тип спектральных данных: многоканальный <<mult>> или трёхканальный <<3ch>> спектр;
последующие колонки содержат медианы и 90\% интервалы для распределений
$\alpha$, $E_\rmn{p}$, $S$ и $F_\rmn{peak}$. Для распределения по $\beta$
медиана и 90\% интервал составляют $-2.28$ и $[-3.15,-1.74]$, соответственно.

Наибольшие $E_\rmn{p}$ измеренные для коротких всплесков KW составляют $\sim 3$~МэВ: 
$E_\rmn{p} = 3.55(-0.71,+0.85)$~МэВ было получено для GRB20090510\_T01381
(GRB 090510~\citep{Ackermann_2010ApJ_716_1178A});
более мягкие всплески GRB19970704\_T04097 и GRB20080611\_T04742 имеют $E_\rmn{p} \approx 3.3$~МэВ.
Практически все всплески с $E_\rmn{p} \lesssim 200$~кэВ классифицированы как Тип~II или Тип~I/II
и, вероятно, представляют собой популяцию отличную от более жестких всплесков (см. обсуждение ниже).

\input{tables/distribution_parameters}

Результаты анализа подтверждают, что спектры большей части коротких всплесков хорошо
описываются моделью CPL с жестким $\alpha \sim -0.5$ и $E_\rmn{p}$ в диапазоне 100~кэВ--2~МэВ.
В представленном анализе только $\sim 4$\% коротких всплесков KW описываются моделью BAND.
Среди 5\%  всплесков с наибольшим $S$, 20\% описываются моделью BAND; для описания 
оставшихся 80\% требуется $\beta \lesssim -2.5$, в большинстве случаев неограниченное снизу.
Это предполагает, что отсутствие степенного поведения спектра на больших энергиях,
обнаруженное для большей части ярких коротких всплесков, по видимому является внутренним
свойством всплесков, а не связанно с низкой статистикой отсчётов на больших энергиях.

В данном разделе не ставилась задача проверки применимости более сложных моделей
для описания спектров коротких всплесков.
Однако, среди 214 всплесков с многоканальными спектрами было обнаружено три
события, для описания которых необходима дополнительная жесткая степенная 
спектральная компонента с фотонным индексом $\sim -2$. Эти всплески входят в 10\%
наиболее интенсивных событий из набора. Отношение энергетических потоков PL
компоненты к CPL находится в диапазоне от 0.03 для GRB20031214\_T366655 до
0.4 для GRB19980205\_T19785. Обнаруженная компонента может иметь ту же природу,
что и обнаруженная в GRB~081024B~\citep{Abdo_2010ApJ_712_558A} и 
GRB~090510~\citep{Ackermann_2010ApJ_716_1178A} на основе данных \textit{Fermi}-GBM и~LAT.
GRB~081024B не был зарегистрирован KW триггерном режиме, GRB~090510 (GRB20090510\_T01381)
содержится в нашем наборе. Для описания спектра последнего всплеска степенная 
компонента не требуется, верхний предел на поток степенной компоненты с фотонным 
индексом $-1.7$ составляет $\sim 1 \times 10^{-6}$~эрг~см$^{-2}$~с$^{-1}$ для 
уровня значимости 90\%. Соответствующий предел на отношение потока PL компоненты 
к потоку в основной компоненте составляет $\lesssim 0.02$ на уровне значимости 90\%.

\subsection{Сравнение коротких всплесков KW с BATSE и GBM}
Результаты представленного спектрального анализа были сопоставлены с данными
других экспериментов.
Наибольшие наборы данных по GRB в широком спектральном диапазоне получены экспериментами
\textit{CGRO}-BATSE\footnote{\url{http://heasarc.gsfc.nasa.gov/W3Browse/cgro/bat5bgrbsp.html}}
(в диапазоне 20~кэВ--2~МэВ~\citep{Goldstein_2013ApJS}) и 
\textit{Fermi}-GBM\footnote{\url{http://heasarc.gsfc.nasa.gov/W3Browse/fermi/fermigbrst.html}}
(в диапазоне 8~кэВ--40~МэВ~\citep{Gruber_2014ApJS}). 
Настоящий анализ содержит примерно в два раза коротких всплесков, чем каталог GBM, 
в более узком диапазоне энергий, и в $\sim 1.5 $ раза меньше чем набор BATSE, 
но в более широком диапазоне энергий.

Из каталога BATSE~5B было выбрано 427 всплесков с $T_{90}<2$~с и с интервалом 
накопления интегрального спектра короче 10~с. На основе критерия $\Delta \chi^2>6$
всплески имеют следующую статистику наиболее подходящих моделей:
11~--- BAND, 225~--- CPL и 191~--- PL.
Из второго каталога GBM было выбрано 146 всплесков с $T_{90}<2$~с. Наиболее подходящие модели,
по данным каталога распределены так: 3~--- BAND, 67~--- CPL и 76~--- PL;
всплески описываемые степенной моделью с изломом (около пяти событий) были 
исключены из сравнения.

Отношение числа всплесков, описываемых BAND и CPL моделями, мало ($\lesssim 5$\%)
для всех наборов. Было проверено согласуются ли распределения по $\alpha$ и 
$E_\rmn{p}$ модели CPL между инструментами. Двусторонние p-значение 
теста Колмогорова-Смирнова для двух наборов ($P_\rmn{KS}$) для распределений 
KW и GBM по $\alpha$ и $E_\rmn{p}$ равны 10\% и 25\%, соответственно, в то время 
как для сравнений KW и BATSE, и GBM и BATSE $P_\rmn{KS}<1$\%.
Набор BATSE имеет медиану $\alpha=-0.33$ в то время как медианы KW и GBM равны
$\alpha=-0.49$ и $\alpha=-0.50$, соответственно.
Медианы распределений по $E_\rmn{p}$ равны $\approx 400$~кэВ для BATSE и 
$\approx 550$~кэВ для KW и GBM.
Таким образом результаты KW хорошо согласуются с данными GBM и хуже с данными BATSE.

Доля всплесков, наилучшим образом описываемых моделью PL, сильно отличаются у KW и
других инструментов:
2\% (5\% используя критерий $\Delta \chi^2 > 6$)~--- KW, 52\%~--- GBM и 55\%~--- BATSE.
Причина достаточно большой доли PL моделей у BATSE и GBM была детально проанализирована.
Для более устойчивого сравнения были выбраны всплески с $S>5.5\times 10^{-7}$~эрг~см$^{-2}$,
данный порог примерно соответствует наименьшему $S$, полученному для всплесков KW с многоканальными
спектрами. Полученные поднаборы 138 (BATSE) и 49 (GBM) GRBs содержали 29 (21\%) и 3 (6\%) моделей PL, соответственно. 
Таким образом, в схожем диапазоне интегральных энергетических потоков доли PL моделей
согласуются для KW и GBM.
Среди моделей CPL с ограниченными параметрами для указанных 29 всплесков BATSE,
17 имеют $1\sigma$ верхний предел $E_\rmn{p}$, превышающий верхнюю границу 
спектрального диапазона  BATSE (2~МэВ). Оставшиеся 12 всплесков составляют 9\% от поднабора. 
Следовательно, можно считать, что основным источником избытка PL моделей для коротких 
всплесков BATSE и GBM является большое количество слабых всплесков (с низким $S$),
для которых более сложная модель не может быть предпочтительна из-за низкой статистики отсчётов.
Для всплесков BATSE дополнительным фактором, влияющим на увеличение доли PL моделей, 
является относительно узкий спектральный диапазон.

\subsection{Короткие всплески с EE}
Было обнаружено, что 30 всплесков из набора 1939 KW гамма-всплесков зарегистрированных
с 1994 по 2010 гг. могут быть классифицированы как короткие всплесков с EE, 
основываясь на длительности начального импульса и отсутствие в последующем излучении 
заметной спектральной эволюции. Из них, 21 всплеск имел интенсивность EE достаточную
для анализа многоканальных спектров. Для шести из этих всплесков спектр EE описывается
моделью с изломом (CPL) с достаточно высокой $E_\rmn{p} \sim 160$~кэВ--2.2~МэВ.
Начальные импульсы двух из них были классифицированы как типы Iee/II и они вероятно
являются длинными всплесками с короткими начальными импульсами. Оставшиеся четыре всплеска 
являются <<стандартными>> короткими всплесками с продлённым излучением на основании 
формы временного профиля. Подобные спектры был обнаружены у двух из 19 всплесков BATSE~\citep{Bostanci_2013MNRAS}
и у четырёх из 14 всплесков GBM~\citep{Kaneko_2015MNRAS}. 
Полученные соискателем результаты дают дополнительное свидетельство в пользу наличия 
достаточно жесткого продлённого излучения у коротких гамма-всплесков.

Набор коротких всплесков с продлённым излучением KW включает два всплеска, 
где EE наблюдалось на BATSE, и три, где EE наблюдалось GBM. 
Результаты спектрального анализа этих всплесков (включая GRB20090831\_T27393 с жетким EE, 
$E_\rmn{p} \approx 215$~кэВ) по данным KW согласуются с результатами упомянутых инструментов.
Всплеск GRB20090720\_T61379 с экстремально жестким EE ($E_\rmn{p} \approx 2.2$~МэВ) 
был зарегистрирован GBM. Хотя всплеск не был включён в набор~\citep{Kaneko_2015MNRAS},
параметры интегрального спектра этого всплеска по данным KW согласуются с полученными 
по данным GBM.

Яркий близкий всплеск GRB~060614, который был классифицирован как короткий 
всплеск с EE~\citep{Gehrels_2006Nature} был зарегистрирован KW ($T_0$(KW)=45831.590~с;~\citep{Golenetskii_GCN5264}).
Этот всплеск не был включён в наш набор коротких всплесков из-за большой 
длительности начального импульса $T_{50}=2.7 \pm 0.3$~с.

\subsection{Кандидаты в гигантские вспышки SGR}
Яркий начальный импульс гигантской вспышки (GF) источника мягких повторяющихся гамма-всплесков, 
произошедшей в близкой галактике, может выглядеть в данных гамма детектора как короткий гамма-всплеск. 
Верхней предел на частоту GF на основе локализаций коротких свсплесков KW был
получен в Главе~\ref{SGR_GF_search}. При этом было обнаружено только два кандидата:
GRB~051103 (GRB20051103\_T33943) в группе галактик M81/M82~\citep{Frederiks_2007AstLett} и 
GRB~070201 (GRB20070201\_T55390) в галактике M31~\citep{Mazets_2008ApJ}.
Оба события находятся среди 10\% наиболее интенсивных событий на основе интегрального 
энергетического потока. Спектральные параметры всплесков являются типичными для 
набора коротких всплесков KW.
Избыток отсчётов, наблюдаемый вплоть до 90~с после триггера GRB~070201, 
был интерпретирован как хвост гигантской вспышки~\citep{Mazets_2008ApJ}.
Значимость избытка составляет $4.3\sigma$, поэтому он не был обнаружен при 
поиске продлённого излучения (с порогом $5\sigma$).

\subsection{Неоднородность набора коротких всплесков}
На рис.~\ref{fig:EpT50} показано соотношение $E_\rmn{p}$, для наиболее подходящей CPL модели,
и длительности всплеска $T_{50}$. Всплески Типа~I, в среднем, короче и жестче 
($E_\rmn{p}\gtrsim 200$~кэВ), чем всплески Типа~II, что подтверждает классификацию,
полученную на основе распределения жесткость-длительность.
Среди 4 всплесков наилучшим образом описываемых моделью PL, два всплеска принадлежат к типам I и II,
и два имеют неопределённый тип (I/II). Явный недостаток всплесков с $E_\rmn{p}\lesssim 100$~кэВ,
скорее всего связан с эффектом селекции. 
Распределение по длительности начальных импульсов коротких всплесков типа~I с EE (Iee) согласуется с 
распределением для обычных коротких всплесков типа~I, о чем свидетельствует p-значение
теста Колмогорова-Смирнова $P_\rmn{KS}=0.5$. Также было обнаружено, что
начальные импульсы всплесков типа Iee, в среднем, жестче ($E_\rmn{p}$ в $\sim 1.5$ раза выше),
чем всплески типа~I ($P_\rmn{KS} = 0.01$). В заключении были сопоставлены распределения 
всплесков типов Iee и~I по $S$ и $F_\rmn{peak}$, для пар распределений 
была получена $P_\rmn{KS} \sim 0.01$, что свидетельствует о различие распределений,
при этом начальные импульсы всплесков с EE, в среднем, более интенсивные.

\begin{figure}
    \begin{minipage}[h]{0.5\textwidth}
		\center{\includegraphics[width=1\textwidth]{gEpvsT50TypesII_ru.eps} \\ а)}
    \end{minipage}
    \hfill
    \begin{minipage}[h]{0.5\textwidth}
		\center{\includegraphics[width=1\textwidth]{gEpvsT50ipEE_ru.eps} \\ б)}
	\end{minipage}

\caption{
    Соотношение $E_\rmn{p}$ и $T_{50}$ (только для модели CPL).
    На панели~(а) показаны всплески типа~I (серые точки), типа~II (красные треугольники)
    и всплески неопределённого типа, I или~II (синие окружности).
    На панели~(б) показаны всплески типа~I с EE (заполненные красные звёзды)
    и всплески неопределённого типа Iee/II (пустые красные звёзды)
    Для всплесков типа~I ошибки не показаны. 
    \label{fig:EpT50}}
\end{figure}

На рис.~\ref{fig:EpvsFPandFL} приведены соотношения $E_\rmn{p}$ с $S$ и $F_\rmn{peak}$.
Распределение 79 слабых всплесков с трёх-канальными спектрами по спектральным параметрам 
и $F_\rmn{peak}$ схоже с более интенсивными событиями, эти всплески являются продолжением
распределения коротких/жестких (Тип~I) всплесков в область низких $S$  на диаграмме $E_\rmn{p}$--$S$.
Явным выбросом из  $E_\rmn{p}$--$F_\rmn{peak}$ распределения является предполагаемая 
GF в галактике M31, что подкрепляет свидетельства в пользу отличной от GRB природы этого события.
Всплески типов I и~II занимают практически не пересекающиеся области на диаграмме $E_\rmn{p}$--$S$.
Всплески типа~I образуют вытянутое распределение, которое, в среднем, подчиняется 
соотношению $E_\rmn{p} \propto S^{1/2}$. Всплески типа II образуют небольшую группу событий
с низкой $E_\rmn{p}$, которая представляет собой малую часть распределения длинных всплесков.
На плоскости $E_\rmn{p}$--$F_\rmn{peak}$ всплески типа~II продлевают корреляцию 
жесткость-интенсивность в область низких $E_\rmn{p}$ и малых $F_\rmn{peak}$.
Так как только для девяти всплесков из набора были определены космологические 
красные смещения ($z\sim 0.1$--1.0, определённые как спектрометрически, так и
фотометрически), свойства всплесков в собственной системе отсчёта здесь не обсуждаются.
Хотя инструментальные эффекты селекции влияют на параметры представленного набора,
корреляции наблюдаемых параметров всё же могут отражать корреляции параметров 
в системе отсчёта источника всплеска $E_\rmn{p,rest}$--$E_\rmn{iso}$~--- 
соотношение Амати~\citep{Amati_2002AandA} и $E_\rmn{p,rest}$--$L_\rmn{iso}$~--- 
соотношение Йонетоку~\citep{Yonetoku_2004ApJ}
(см., к примеру,~\citep{Nava_2008MNRAS} и ссылки в ней).
Таким образом, описанные свойства распределения жесткость-интенсивность 
в системе наблюдателя, полученные для всплесков типов I и II из набора коротких 
всплесков KW, свидетельствуют в пользу того, что короткие/жесткие всплески подчиняются 
своему соотношению Амати на плоскости $E_\rmn{p,rest}$--$E_\rmn{iso}$
(см., к примеру,~\citep{Nava_2011MNRAS} и ссылки в ней).

\begin{figure}
    \begin{minipage}[h]{0.5\textwidth}
		\center{\includegraphics[width=1\textwidth]{gEpvsFl_ru.eps} \\ а)}
    \end{minipage}
    \hfill
    \begin{minipage}[h]{0.5\textwidth}
		\center{\includegraphics[width=1\textwidth]{gEpvsFlEE_ru.eps} \\ б)}
	\end{minipage}
    \vfill
    \begin{minipage}[h]{0.5\textwidth}
		\center{\includegraphics[width=1\textwidth]{gEpvsPF_ru.eps} \\ в)}
    \end{minipage}
    \hfill
    \begin{minipage}[h]{0.5\textwidth}
		\center{\includegraphics[width=1\textwidth]{gEpvsPFEE_ru.eps} \\ г)}
	\end{minipage}

\caption{\small
    Соотношение $E_\rmn{p}$ с $S$ и $F_\rmn{peak}$ для всплесков, описываемых моделью CPL.
    На панели~(а) показано соотношение $E_\rmn{p}$ и $S$ для
    всплесков типа~I с многоканальными спектрами (чёрные точки); 
    всплесков типа~I с трёхканальными спектрами (красные точки); 
    всплесков неопределённого типа (пустые точки), для обоих типов спектров;
    и для всплесков типа~II (треугольники). 
    На панели~(б) показаны всплески типов Iee (заполненные звёзды) и~Iee/II (пустые звёзды);
    остальные всплески из набора показаны серым.
    На панелях~(в) и~(г) показаны соотношение $E_\rmn{p}$ и $F_\rmn{peak}$ 
    для тех же наборов всплесков. Для всплесков типов~I и~I/II ошибки не показаны.
    Кандидаты в GF в близких галактиках обозначены ромбами. Звёздами показаны всплески с измеренным $z$.
    Пунктирная линия соответствует степенной аппроксимации соотношений для всплесков типа~I,
    в обоих случаях индекс близок к $0.5$.
    % $E_\rmn{p}$--$S$ $\lambda = 0.45 \pm 0.16$
    % $E_\rmn{p}$--$F_\rmn{peak}$ $\lambda = 0.51 \pm 0.22$
    \label{fig:EpvsFPandFL}}
\end{figure}

Интегральные распределения $\log N$--$\log S$ и $\log N$--$\log F_\rmn{peak}$
для набора 293 коротких всплесков KW, а так же степенные распределения с индексом $-3/2$, 
соответствующие однородному распределению источников всплесков в пространстве,
представлены на рис.~\ref{fig:logNlogS_PF}.
Распределение $\log N$--$\log S$ следует однородному распределению в узкой 
области интегральных потоков $(\sim 4\textrm{--}10)\times 10^{-6}$~эрг~см$^{-2}$.
В то время как недостаток слабых всплесков может быть объяснён эффектами селекции,
наблюдаемый избыток интенсивных всплесков в значительной мере обусловлен событиями 
не относящимися к <<классическим>> коротким/жестким всплескам. Среди 12 наиболее 
интенсивных всплесков в наборе KW с $S \gtrsim 10^{-5}$~эрг~см$^{-2}$ только 
четыре имеют Тип~I, остальные классифицированы как типы I/II, II или являются всплесками с EE.
При рассмотрении только всплесков Типа~I, соотношение $\log N$--$\log S$ хорошо 
описывается крутым показателем степени $-1.8 \pm 0.3$ выше $S\sim 5\times 10^{-6}$~эрг~см$^{-2}$.
Распределение $\log N$--$\log F_\rmn{peak}$ коротких всплесков KW также более пологое 
по сравнению с наклоном $-3/2$ и описывается показателем степени 
$-1.16 \pm 0.12$ для $F_\rmn{peak}$ лежащего в интервале
$(0.2\textrm{--}9.4)\times 10^{-4}$~эрг~см$^{-2}$~с$^{-1}$.
В том же диапазоне показатель степени для всплесков Типа~I имеет более крутой показатель 
$-1.42 \pm 0.16$, который согласуется с однородным распределением источников в пространстве.

\begin{figure}
    \begin{minipage}[h]{0.5\textwidth}
		\center{\includegraphics[width=1\textwidth]{glogNlogS_ru.eps} \\ а)}
    \end{minipage}
    \hfill
    \begin{minipage}[h]{0.5\textwidth}
		\center{\includegraphics[width=1\textwidth]{glogNlogPF_ru.eps} \\ б)}
	\end{minipage}
\caption{
    Интегральные распределения $\log N$--$\log S$~(а) и $\log N$--$\log F_\rmn{peak}$~(б).
    Чёрные гистограммы показывают интегральные распределения для всех всплесков, 
    красные~--- только для всплесков Типа~I. Сплошные прямые линии обозначают 
    степенное распределение с показателем степени $-3/2$ и различными нормировками, 
    ожидаемое для однородного расположения источников всплесков в Евклидовом пространстве.
    Пунктирные линии показывают степенную аппроксимацию распределений всплесков Типа~I.
    На панели~(б) штриховой линией показана аппроксимация распределения всех всплесков 
    (см. параметры и диапазоны аппроксимаций в тексте).
    \label{fig:logNlogS_PF} }
\end{figure}

Иллюстрации временных историй и спектров 214 коротких всплесков KW, а также таблицы размещены на сайте
ФТИ~им.~А.Ф.~Иоффе\footnote{\url{http://www.ioffe.ru/LEA/shortGRBs/Catalog2/}}.
\FloatBarrier

\subsection{Короткие гамма-всплески от испарения первичных чёрных дыр}
Считается, что помимо космологически удалённых коротких-гамма всплесков и гигантских вспышек SGR 
в ближайших галактиках в наборе коротких событий (с длительностями $\lesssim 1$~с) 
могут встречаться всплески, испущенные на финальной стадии испарением первичных чёрных дыр (ПЧД) 
в окрестностях Солнца (на расстояниях в несколько пк), см. работу~\citep{Ukwatta_2015arXiv151201264U} и ссылки в ней. 
Расстояние до источников таких всплесков можно оценить на основании данных 
сети IPN, если считать фронт излучения всплеска сферическим. 
При детектировании тремя KA, в принципе, можно исключить локальное происхождение всплеска,
но нельзя подтвердить, что всплеск является локальным.
Если всплеск был зарегистрирован четырьмя или более КА на межпланетных расстояниях,
то возможно получить точное положение источника в пространстве.

Для отбора кандидатов во всплески от испарения ПЧД были использованы следующие 
критерии отбора: 
(1)~событие зарегистрировано в период с октября 1990~г. по декабрь~2015~г. (24.25~лет);
(2)~выполнено условие детектирования несколькими КА, указанное выше 
(по этому критерию прошли только 839 GRBs из более чем 10000);
(3)~у всплеска не было обнаружено послесвечение;
(4)~всплеск имеет полную длительность $\leq 1$~с и не сопровождается продлённым излучением.
Условий на форму кривой блеска не накладывалось.
В итоговый набор всплесков вошли только 36 событий, из них для 34 длительности и 
энергетика были оценены соискателем на основании данных KW.

В результате авторами работы~\citep{Ukwatta_2015arXiv151201264U} были получены 
оценки минимального расстояния до источников всплесков
0.01--10~пк, при этом энергетика событий согласуется с моделью испарения ПЧД 
с массой $\sim 10^9\textrm{--}10^{14}$~г. В предположении, что все 36 выбранных всплесков связаны 
с испарением ПЧД был получен нижний предел на частоту таких 
событий $\sim 3$~всплеска~пк$^{-3}$~год$^{-1}$. При этом верхний предел на частоту 
всплесков от ПЧД составляет $\sim 10^5\textrm{--}10^6$~всплесков~пк$^{-3}$~год$^{-1}$ 
по данным наземных гамма-обсерваторий.

\section{Заключение}
В данной главе представлен спектральный анализ коротких гамма-всплесков Конус-Винд.

По материалам Главы~\ref{sGRB_spectral_catalog} на защиту выносится следующие положения:
\begin{itemize}
\item Каталог спектральных и временн\'{ы}х параметров коротких гамма-всплесков, 
    зарегистрированных в эксперименте Конус-Винд.
\item Обнаружение дополнительной степенной компоненты в спектрах трёх коротких GRB, 
    зарегистрированных в эксперименте Конус-Винд.
\item Временн\'{ы}е и спектральные характеристики коротких GRB 
    с продленным излучением, зарегистрированные в эксперименте Конус-Винд.
\end{itemize}

Результаты отражены в публикациях:
\begin{enumerate}
\item D.~S.~Svinkin, D.~D.~Frederiks, R.~L.~Aptekar, et al. 
The second Konus-\textit{Wind} catalog of short gamma-ray bursts //
arXiv:1603.06832, принято в ApJS;
\item T.~N.~Ukwatta, K.~Hurley, J.~H.~MacGibbon, D.~S.~Svinkin, et al.
Investigation of Primordial Black Hole Bursts using Interplanetary Network Gamma-ray Bursts // 
arXiv:1512.01264, отправлено в ApJ.
\end{enumerate}

%Автор лично произвел спектральный анализ описанного набора коротких гамма-всплесков.
%Интерпретация результатов была сделана совместно с соавторами при активном участии автора.

\clearpage

### 2025-10-23 disser/styles.tex
disser

%%% Макет страницы %%%
\geometry{a4paper,top=2cm,bottom=2cm,left=3cm,right=1cm}

\titleformat{\chapter}[display]
  {\normalfont\Large\filcenter\bfseries}{\chaptertitlename\ \thechapter}{20pt}{\LARGE}
\titlespacing*{\chapter}
  {0pt}{30pt}{20pt}

% см. http://tex.stackexchange.com/questions/134031/how-to-adjust-the-size-and-placement-of-chapter-heading-in-report-class
% и http://tex.stackexchange.com/questions/59726/change-size-of-section-subsection-subsubsection-paragraph-and-subparagraph-ti

%%% Кодировки и шрифты %%%
%\renewcommand{\rmdefault}{ftm} % Включаем Times New Roman

%%% Выравнивание и переносы %%%
\sloppy					% Избавляемся от переполнений
\clubpenalty=10000		% Запрещаем разрыв страницы после первой строки абзаца
\widowpenalty=10000		% Запрещаем разрыв страницы после последней строки абзаца
\linespread{1.3}

%%% Библиография %%%
\makeatletter
\bibliographystyle{ugost2008ns_v1.bst}	% Оформляем библиографию в соответствии с ГОСТ 7.0.5
%\bibliographystyle{utf8gost705u}
%\bibliographystyle{ugost2008}
\renewcommand{\@biblabel}[1]{#1.}	% Заменяем библиографию с квадратных скобок на точку:
\makeatother

%%% Изображения %%%
\graphicspath{{img_part1/}{img_part2/}{img_part3/}{img_part4/}{img_part5/}} % Пути к изображениям

%%% Цвета гиперссылок %%%
\definecolor{linkcolor}{rgb}{0.9,0,0}
\definecolor{citecolor}{rgb}{0,0,1}
\definecolor{urlcolor}{rgb}{0,0,1}
\hypersetup{
    colorlinks, linkcolor={linkcolor},
    citecolor={citecolor}, urlcolor={urlcolor}
}

%%% Оглавление %%%
\renewcommand{\cftchapdotsep}{\cftdotsep}

\newcommand{\kws}{\textit{KW}\ }
\newcommand{\rmn}[1]{{\mathrm{#1}}}
\newcommand\fdg{\mbox{$.\!\!^\circ$}}%
\newcommand\farcm{\mbox{$.\mkern-4mu^\prime$}}%
\newcommand\farcs{\mbox{$.\!\!^{\prime\prime}$}}%
\newcommand\nodata{ ~$\cdots$~ }%

### 2025-10-28 disser/part1.tex
bib

@@
 
-Последние наблюдения открыли новый класс объектов - быстрые голубые оптические транзиенты (FBOT)\cite{Drout2014, Margutti2014, Coppejans2020, Ho2020}.
+Последние наблюдения открыли новый класс объектов - быстрые голубые оптические транзиенты (FBOT)\cite{Drout2014, Margutti2014, Coppejans2020, Ho2020, YaoAt2020mrf, MatthewsAT2022tsd, CHrimesAT2023fhn}.
 Вместе со слабыми гамма-всплесками они возможно принадлежат к переходному классу объектов между нерелятивистскими сверхновыми и обычными гамма-всплесками.

### 2025-11-09 disser/part1.tex
disser

--- 2022-03-01 disser/part1.tex
@@
 
-Последние наблюдения открыли новый класс объектов - быстрые голубые оптические транзиенты (FBOT)\cite{Drout2014, Margutti2014, Coppejans2020, Ho2020}.
+Последние наблюдения открыли новый класс объектов - быстрые голубые оптические транзиенты (FBOT)\cite{Drout2014, Ho2019cow, Coppejans2020, Ho2020, PerleyCamel, YaoAt2020mrf, MatthewsAT2022tsd, ChrimesAT2023fhn, Perleyvth}.
+Они характеризуются большой светимостью $L > 10^{43} эрг/с$, смещением цветов в синюю сторону, коротким характерным временным масштабом порядка нескольких дней, низкой выброшенной массой эжекты и её высокими скоростями.
+Точная природа данных объектов на даный момент неизвестна.
 Вместе со слабыми гамма-всплесками они возможно принадлежат к переходному классу объектов между нерелятивистскими сверхновыми и обычными гамма-всплесками.
-Как показано в работе Шевалье и Ирвина \cite{Chevalier2011} наблюдаемые характеристики вспышки определяются длительностью действия центрального источника энергии выброса и его взаимодействием со внешними слоями взрывающейся звезды.
-При различных параметрах вспышка будет относиться к разным классам объектов.
-Известные на данный момент быстрые голубые оптические транзиенты AT2018cow \cite{Margutti2014}, CSS161010 \cite{Coppejans2020}, ZTF2018abvkwla \cite{Ho2020} и AT2020xnd \cite{Ho2021, Bright2021} характеризуются большой светимостью $L > 10^43 эрг/с$, коротким характерным временным масштабом порядка нескольких дней, низкой выброшенной массой эжекты и высокими скоростями.
-Рассматриваемы в данной главе объект CSS161010, расположенный в карликовой галактике на расстоянии 150 мегапарсек имеет по оценкам следующие\cite{Coppejans2020} характеристики: скорость эжекты 0.55 c на ранних этапах и около 0.3 c на поздних (порядка одного года), выброшенная масса порядка 0.01-0.1 солнечных масс.
-]textcolor{red}{Важнейшей особенностью этих объектов является вид их спектральной плотности излучения, имеющей выраженный максимум и степенные участки, говорящие о наличии синхротронного самопоглощения.}
+Так же возможно их объяснение событиями приливных разрывов.
+
+Рассматриваемый в данной главе объект CSS161010, расположенный в карликовой галактике на расстоянии 150 мегапарсек имеет по оценкам следующие\cite{Coppejans2020} характеристики: скорость эжекты 0.55 c на ранних этапах и около 0.3 c на поздних (порядка одного года), выброшенная масса порядка 0.01-0.1 солнечных масс.
+Он обладает наибольшей скоростью эжекты из всех известных на текущий момент FBOT объектов, и следовательно наиболее удобен для Particle-in-Cell моделирования в виду малого времени ускорения.
+\textcolor{red}{Важнейшей особенностью этих объектов является вид их спектральной плотности излучения, имеющей выраженный максимум и степенные участки, говорящие о наличии синхротронного самопоглощения.}
 %почему четвертый был отдельно в статье?
@@
 В данной главе будет проведено Particle-in-Cell моделирование ускорения частиц в транс-релятивистской ударной волне со скоростями, характерными для FBOT объектов.
-По имеющимся распределениям будет расчитано синхротронное излучение и с помощью фитирования наблюдательных данных будут подобраны такие параметры как магнитное поле, концентрация вещества и геометрические характеристики в ударной волне.
+По имеющимся распределениям будет расчитано синхротронное радио-излучение и с помощью фитирования наблюдательных данных транзиента CSS16 1010 будут подобраны такие параметры как магнитное поле, концентрация вещества и геометрические характеристики в ударной волне.
+Далее, с использованием широокого распределения частиц, полученного из Монте-Карло моделирования с соответствующими параметрами, будет расчитано рентгеновское излучение от данного объекта.
 
 \section{Излучение объектов с синхротронным самопоглощением}
-Процесс синхротронного излучения хороши известен и описан в классических работах.
-Но с точки зрения квантовой электродинамки, любому процессу излучения можно так же сопоставить процесс поглощения.
+Процесс синхротронного излучения хорошо известен и описан в классических работах.
+Но  любому процессу излучения можно так же сопоставить процесс поглощения.
 Сечение процесса синхротронного самопоглощения описано в работе Гизеллини и Свенсона \cite{Ghisellini1991}.
@@
 \end{equation}
-где $\phi$ это угол межде вектором магнитного поля и лучом зрения, $\displaystyle\nu_{c}$ критическая частота, определяемая выражением $\displaystyle\nu_{c} = 3 e^{2} B \sin(\phi) E^{2}/4\pi {m_{e}}^{3} c^{5}$, и~$K_{5/3}$ - функция МакДональда.
+где $F\left(E\right)$ - это функция распределения излучающих частиц, $\phi$ - угол межде вектором магнитного поля и лучом зрения, $\displaystyle\nu_{c}$  - критическая частота, определяемая выражением $\displaystyle\nu_{c} = 3 e^{2} B \sin(\phi) E^{2}/4\pi {m_{e}}^{3} c^{5}$, и~$K_{5/3}$ - функция МакДональда.
 Коэффициент поглощения для фотонов, распростроняющихся вдоль луча зрения равен
@@
 \end{equation}
-Используя эти формулы Шевалье \cite{Chevalier1998} построил модель излучения плоского однородного диска, расположенного перпендикулярно к лучу зрения, с радиусом $R$, толщиной $s$, однородным магнитным полем $B$ и степенной функцией распределения электронов $N(E) = N_0 E^{-p}$.
+Используя эти формулы Шевалье \cite{Chevalier1998} построил модель излучения плоского однородного диска, расположенного перпендикулярно к лучу зрения, с радиусом $R$, толщиной $s$, однородным магнитным полем $B$ и степенной функцией распределения электронов $F(E) = N_0 E^{-p}$.
 Вместо толщины обычно используется доля излучающего объема $f$, определенная так, что $\pi R^2 s = 4 \pi /3 R^3 f$.
@@
 где $D$ - расстояние до источника, $\epsilon_B, \epsilon_e$ - доли энергии в источнике, содержащиеся в магнитном поле и в электронах, $E1$ - энергия, с которой начинается степенное распределение электронов.
-Таким образом в работах  были оценены параметры источников AT2018cow \cite{Margutti2014}, CSS161010 \cite{Coppejans2020}, ZTF2018abvkwla \cite{Ho2020} и AT2020xnd \cite{Ho2021, Bright2021}.
-Так же, измеряя радиус источника $R$ в разные моменты времени можно оценить скорость распространения ударной волны.
+Таким образом в работах  были оценены параметры источников AT2018cow \cite{Ho2019cow}, CSS161010 \cite{Coppejans2020}, ZTF2018abvkwla \cite{Ho2020} и AT2020xnd \cite{Ho2021, Bright2021}.
+Так же, измеряя размер источника $R$ в разные моменты времени можно оценить скорость распространения ударной волны.
 \section{Particle-in-Cell моделирование }
@@
 
-\section{Результаты}
-Были проведены расчеты синхротронного излучения от оптического транзиента CSS161010 для момента времени $\displaystyle t = 357$ дней после взрыва, по данным работы \cite{Coppejans2020}.
-Параметры источника, определенные с помощью уравнений~(\ref{ChevR})
-и (\ref{ChevB}), в предположении равенства распределения энергий $\epsilon_e = \epsilon_B = 1/3$, получаются следующими: магнитное поле $\displaystyle B = 0.052$ Гс, радиус оболочки $\displaystyle R = 3.3\times10^{17}$~cm,
-концентрация $n = 1.9$~cm$^{-3}$, спектральный индекс электронов $s = 3.5$, который близок к полученному в PIC расчетах, и минимальна энергия электронов $\displaystyle E_{1} = 4~m_{e}c^{2}$.
+\section{Фитирование парметров радиоисточника}
+Были проведены расчеты синхротронного излучения и подбор параметров для оптического транзиента CSS161010 для моментов времени $\displaystyle t = 69, 99 и 357$ дней после взрыва, для четырех функций полученных из PIC моделирования - для скоростей потока 0.5 c и 0.75 c и для углов наклона 30 и 80 градусов.
+Полученные значения функционала и параметров для разных моделей для момента времени 99 дней после взрыва показаны в талблице \ref{parameters}.
+
+\textcolor{red}{написать про толщину 0.1 а не 0.5}
+
+\renewcommand{\arraystretch}{1.375}
+\begin{table}[h!]
+	\label{parameters}
+	\caption{Полученные значения параметров радиоисточника для различных моделей}
+	\begin{center}
+		\begin{tabular}{|c | c| c| c| c| c|}
+			\hline
+			model &$r$ & $B$, G& $n$,~$\rm cm^{-3}$&  $E_e,~\rm erg$& $E_p,~\rm erg$\\
+			\hline
+			A80 & 550 & 0.5 & 2$\times 10^6$  & 3.4$\times 10^{51}$ & 3$\times 10^{53}$\\
+			\hline
+			A30 & 140 & 0.1& 5$\times 10^5$&  8$\times 10^{51}$ & 7$\times 10^{53}$\\
+			\hline
+			B80 & 170 & 0.3& 540&  3$\times 10^{49}$& 3.7$\times 10^{51}$\\
+			\hline
+			B30 & 57 & 0.34& 17&  2.2$\times 10^{48}$& 1.4$\times 10^{50}$\\
+			\hline
+		\end{tabular}
+	\end{center}
+\end{table}
+
+Полученные значения магнитного поля и концентрации в лучше всего подходящей модели B30 близки к полученным авторами оригинальной работы \cite{Coppejans2020} с помощью уравнений~(\ref{ChevR})
+и (\ref{ChevB}), в предположении равенства распределения энергий $\epsilon_e = \epsilon_B = 1/3$, в то время как скорость и соответственно размер источника оказываются значитеьно большими.
+Для этой же модели B30 проведено фитирование параметров в другие моменты времени, получены следующие значения.
+На 69-ый день магнитно поле равно $0.4~\rm G$, концентрацияя $24~\rm cm^{-3}$ , а размер источника $R =1.3\times10^{17}~\rm{cm}$.
+На 357-ой день магнитное поле равно $0.04~\rm G$, концентрация $0.25~\rm cm^{-3}$, а размер источника $R= 6.9\times10^{17}~\rm{cm}$.
+Полученные значения магнитного поля и концентрации значительно отличаются от простых ожидаемых зависимостей от расстояния $B \propto r^{-1}$ and $n  \propto r^{-2}$, а так же попытка одновременного фитирования всех трех моментов времени со степенными зависимостями размера от времени и магнитного поля и концентрации от расстояния не удалась, что может говорит о сильной неоднородности и немонотонности плотности вокруг источника.
 
@@
 
+\section{Расчет рентгеновского излучения}
+На 99-ый день с момента детектирования имеются одновременные наблюдения в радио и в рентгеновском диапазоне.
+Объект обладает большой рентгеновской светимостью (поток $F_{X} = (1.33 \pm 0.76)\times10^{-15} ~\rm erg~s^{-1}~cm^{-2}$ и соответствующая изотропная светимость $L_{X}=(3.4\pm1.9)\times10^{39}~\rm erg~s^{-1}$ iв диапазоне $0.3-10~\rm keV$ \cite{Coppejans2020}).
+Последние исследования FBOT предполагают наличие центрального источника для объяснения рентгеновского излучения --- предполагается, что на ранних стадиях оно производится милисекундным магнетаром \citep{GottliebFBOTjet}, а на поздних стадиях аккреционным диском \citep{MiglioriFBOtlatestages}.
+Мы далее предлагаем модель объясняющую рентгеновское излучение FBOT (конкретно CSS161010) без центрального источника, по крайней мере на поздних стадиях.
+
+Если продлить степенное распределение электронов (с показателем спектра $\approx3.5$), полученное из PIC моделирования, до высоких энергий, то полученный уровень рентгеновского излучения за счет синхротронного механизма окажется на несколько порядков ниже наблюдаемого.
+Но как следует из Монте-Карло расчетов, на высоких энергиях спектр ускоренных частиц становится более жестким.
+Монте-Крло моделирование не рассматривает электроны, но в диффузионном приближении функция распределения на высоких энергиях определяется только коэффициентом диффузии, который мы берем зависящим только от гирорадиуса частцы, и являющегося одинаковым для пртонов и электронов с одной энергией.
+Поэтому функции распределения протонов и электронов будут подобны до тех пор, пока синхротронные потери не начнут играть роль.
+Ограничения за счет синхротронных потерь можно вычислить оценив максимальную достижимую энергию, сравнив темп набора энергии и темп синхротронных потерь.
+
+\begin{equation} \label{accTime}
+	t_{acc}=\int_{p_0}^p \frac{3}{v_1-v_2}\left(\frac{D_1}{v_1}+\frac{D_2}{v_2}\right)\frac{dp}{p}
+\end{equation}
+
+где $D_1, D_2$ коэффициенты диффузии, $v_1, v_2$ --- скорости потока перед и за фронтом ударной волны соответственно \citep{Drury83}.
+Для релятивистской частицы , ускоряющейся на сильной ударной волне с бомовским коэффициентом диффузии  $D = cE/3eB$, можно получить скорость набора энергии $dE_{acc}/dt = 3 v^2 e B/8c$.
+Приравняв это к скорости синхротронных потерь получим максимальную достижимую энергию электронов $E_{max}= (v/c) \cdot m^2 c^4/\sqrt{e^3 B}$.
+В нашей модели магнитное поле $B\approx0.3~\rm{G}$ и соответствующее значение максимальной энергии $E_{max}$ равно $\approx10^8~m_e c^2$.
+Таким образом, скорректируем функцию распределения электронов, домножив функцию распределения протонов на $\exp(-E/E_{max})$.
+Абсолютную величину высокоэнергичной части функции распределения электронов получим экстраполируя низкоэнегричную часть, полученную из PIC моделирования, до энергий больших энергии тепловых протонов.
+
+Была проведена серия Монте-Карло расчетов, со скоростью ударной волны и концентрациями, соответствующими модели B30 и различными значениями начального магнитного поля, так, чтобы в результате усиления магнитного поля, получилось значение так же соответствующее модели B30.
+Это происходит при величине начального поля $3\times10^{-5}~G$.
+Функция распределения из данного расчета была использована для получения композитных функция распределения протонов и электронов, изображенных на рисунке \ref{distributions}.
+Отличие в уровне инжекции протонов и электронов оказывается $\sim 2\times10^3$ что соотвтетствует результатам, полученным в работе \cite{ParkElectronIonShock}.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.4\textwidth]{./fig/distributions2_momentum.png} 
+	\caption{Функции распределения протонов и электронов, полученные объединением результатов PIC и МОнте-Карло моделирования со скоростью ударной волны 0.75 c.
+Пунктирная линия означает функцию распределения электронов без учета синхротронных потерь, а широкая голубая линия - область сшивки двух распределений.}
+	\label{distributions}
+\end{figure}
+
+Для расчета высокоэнергичного излучения необходимо учитывать пространственную неоднородность функции распределения за счет потерь при движении от фронта ударной волны.
+Пренебрежем диффузионным переносом и будет расчитывать аналитически конвекционный перенос с синхротронными потерями.
+Тогда на расстоянии $l$ от фронта ударной волны функция распределения будет выражаться через функцию распределения на фронте $f(E)$ как
+
+\begin{equation}
+	f_l(E)=f\left(\frac{E}{1-4e^4 B^2 E~l/9m^4 c^7 u_d}\right)\cdot\frac{1}{\left(1-4e^4 B^2 E~l/9m^4 c^7 u_d\right)^2}
+\end{equation}
+
+где $u_d$ - скорость потока в даунстриме, равная 0.2 c в нашей модели.
+Спектр синхротронного излучения источника в широком диапазоне энергий приведен на рисунке \ref{long_radiation}.
+Низкочастотный пик связан с синхротронным самопоглощение в радиодиапазоне, второй пик связан с укручением спектра электронов на высоких энергиях, а спад - с завалом функции распределения за счет синхротронных потерь.
+Полученное значение потока в диапазоне $0.3-10~\rm keV$ равно $8.9\times10^{-16}~\rm erg~cm^{-2} s^{-1}$.
+Это значение согласуется с наблюдаемым рентгеновским потоком от CSS161010 на 99-ый день после взрыва.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.4\textwidth]{./fig/long_radiation.png} 
+	\caption{Модельный спектр излучения синхротронного излучения с учетом композитной функции распределения электронов.
+Интегральны рентгеновский поток в диапазоне $0.3-10~\rm keV$ равен $8.9\times10^{-16}~\rm erg~cm^{-2} s^{-1}$ что соответствует наблюдаемому Чандрой значению $F_{X} = (1.33 \pm 0.76)\times10^{-15} ~\rm erg~s^{-1}~cm^{-2}$.}
+	\label{long_radiation}
+\end{figure}
+
 \FloatBarrier

### 2025-11-11 disser/part1.tex
pictures
disser

@@
 Поэтому для квазипоперечных ударных волн, у которых $c \cos(\theta')$ меньше скорости ударной волны $\displaystyle v_{sh}'$ ускорение невозможно.
-При углах наклона меньших критического, ударная волна способна эффективно ускорять частицы, формировать степенной хвост их функции распределения, как показано на рисунке~\ref{compareelectrons}.
-\begin{figure}
+При углах наклона меньших критического, ударная волна способна эффективно ускорять частицы и формировать степенной хвост их функции распределения.
+Проводились расчеты со скоростями натекающего потока 0.5 c, 0.75 c и углами наклона магнитного поля от 10 до 90 градусов.
+Полученные функции распределения протонов и электронов представлены на рисунках \ref{distributionsE} и \ref{distributionsP}.
+\begin{figure}[h]
+	\begin{minipage}{0.49\textwidth}
 	\centering
-	\includegraphics[width=10.5 cm]{compareelectrons2.eps} 
-	\caption{Функция распределения электронов за фронтом ударной волны при следующих параметрах: скорость потока \mbox{$v = 0.3~c,$} , магнетизация $\sigma = 0.0002$ при нескольких углах наклона.}
-	\label{compareelectrons}
+	\includegraphics[width=0.95\textwidth]{./img_part1/electrons_momentum.png} 
+	\caption{Функции распределения электронов за фронтом ударных волн с различными углами наклона магнитного поля  $\theta$ и скоростями $v$}
+	\label{distributionsE}
+    \end{minipage} \hfill
+    \begin{minipage}{0.49\textwidth}
+	\centering
+	\includegraphics[width=0.95\textwidth]{./img_part1/protons_momentum.png} 
+	\caption{Функции распределения протонов за фронтом ударных волн с различными углами наклона магнитного поля $\theta$ и скоростями $v$}
+	\label{distributionsP}
+    \end{minipage}
 \end{figure}
 
-Для дальнейшего моделирования излучения использовалась функция распределения, полученная для угла $\theta = 30^\circ$ градусов (квазипараллельная ударная волна).
+%Для дальнейшего моделирования излучения использовалась функция распределения, полученная для угла $\theta = 30^\circ$ градусов (квазипараллельная ударная волна).
 
-Из результатов моделирования можно извлечь параметры полученного распределения электронов.
+%Из результатов моделирования можно извлечь параметры полученного распределения электронов.
 За фронтом ударной волны функция распределения электронов имеет сложный вид, но можно выделить две основных компоненты - тепловой пик и степенной хвост на высоких энергиях.
-На низких энергиях ($\displaystyle E < 5 m_e c^2$) мы аппроксимируем фунцию распределения с помощью распределения Максвелла-Юттнера, итеративным процессом минимизируя функционал $\displaystyle f(T) = \int_{m_e c^{2}}^{5 m_e c^{2}} (F(E) - F_{mj}(E,T))^{2}dE$, где $\displaystyle F(E)$ функция распределения полученная из моделирования, $\displaystyle F_{mj}(E,T)$ - распределение Максвелла-Юттнера.
+%На низких энергиях ($\displaystyle E < 5 m_e c^2$) мы аппроксимируем фунцию распределения с помощью распределения Максвелла-Юттнера, итеративным процессом минимизируя функционал $\displaystyle f(T) = \int_{m_e c^{2}}^{5 m_e c^{2}} (F(E) - F_{mj}(E,T))^{2}dE$, где $\displaystyle F(E)$ функция распределения полученная из моделирования, $\displaystyle F_{mj}(E,T)$ - распределение Максвелла-Юттнера.
 Таким образом определяется температура, соответствующая тепловому пику.
@@
 Функция распределения электронов за фронтом ударной волны (от 2000 до 5000 ячеек за фронтом) в момент времени $\displaystyle t$ = 70,000~$\omega_{e}^{-1}$ при начальных параметрах $\displaystyle\sigma = 0.0002$, $\theta = 30^\circ$ и ее аппроксимации с \textcolor{red}{температурой $\displaystyle T_e = 5\times 10^{10}$~K} и спектральным индексом $\displaystyle p = 3.6$ изображены на рисунке ~\ref{electrons}.
-Так же с помощью Particle-in-Cell моделирования можно определить параметры $\epsilon_B, \epsilon_e$, определяющие связь между концентрацией и магнитным полем, которые необходимы для дальнейшего расчета излучения.
-\begin{figure}
-		\centering
-		\includegraphics[width=10.5 cm]{electrons.eps} 
-		\caption{Функция распределения электронов за фронтом ударной волны при начальных параметрах $v = 0.3~c,$ $\sigma = 2\times10^{-4}$ и $\theta = 30^{\circ}$.}
-		\label{electrons}
-\end{figure}
+%Так же с помощью Particle-in-Cell моделирования можно определить параметры $\epsilon_B, \epsilon_e$, определяющие связь между концентрацией и магнитным полем, которые необходимы для дальнейшего расчета излучения.
+%\begin{figure}
+%		\centering
+%		\includegraphics[width=10.5 cm]{electrons.eps} 
+%		\caption{Функция распределения электронов за фронтом ударной волны при начальных параметрах $v = 0.3~c,$ $\sigma = 2\times10^{-4}$ и $\theta = 30^{\circ}$.}
+%		\label{electrons}
+%\end{figure}
 
-\section{Расчёт излучения}
+%\section{Расчёт излучения}
+
+\section{Фитирование парметров радиоисточника}
 
 Зная точную функцию распределения и предположив геометрию источника в виде плоского диска, можно рассчитать наблюдаемые потоки с помощью известных формул для синхротронного излучения и самопоглощения, описанных например в работе \cite{Ghisellini}.
-\textcolor{red}{Мощность излучения на единицу частоты излучаемая единичным объемом равна}
-\begin{equation} \label{emission}
-	I(\nu)=\int_{E_{min}}^{E_{max}} dE \frac {\sqrt {3}{e}^{3}n F(E) B \sin ( \phi)}{{m_e}{c}^{2}}
-	\frac{\nu}{\nu_c}\int_{\frac {\nu}{\nu_c}}^{\infty }\it K_{5/3}(x)dx,
-\end{equation}
-где $\phi$ угол между магнитным полем и лучом зрения, $\displaystyle\nu_{c}$ критическая частота, определяемая выражением
-$\displaystyle\nu_{c} = 3 e^{2} B \sin(\phi) E^{2}/4\pi {m_{e}}^{3} c^{5}$, и~$K_{5/3}$ функция МакДональда. Коэффициент поглощения равен
-\begin{equation}\label{absorption}
-	k(\nu)=\int_{E_{min}}^{E_{max}}dE\frac {\sqrt {3}{e}^{3}}{8\pi m_e \nu^2}\frac{n B\sin(\phi)}{E^2}
-	\frac{d}{dE} E^2 F(E)\frac {\nu}{ \nu_c}\int_{\frac {\nu}{ \nu_c}}^{\infty }K_{5/3}(x) dx.
-\end{equation}
-Проинтегрировав излучательную способность по объему с учетом самопоглощения и зная расстояние $\displaystyle D$ до источника, можно определить спектральную плотность потока излучения.
+Проинтегрировав излучательную способность по объему с учетом самопоглощения (формулы \ref{emission} и \ref{absorption}) и зная расстояние $\displaystyle D$ до источника, можно определить спектральную плотность потока излучения.
 
-\section{Фитирование парметров радиоисточника}
 Были проведены расчеты синхротронного излучения и подбор параметров для оптического транзиента CSS161010 для моментов времени $\displaystyle t = 69, 99 и 357$ дней после взрыва, для четырех функций полученных из PIC моделирования - для скоростей потока 0.5 c и 0.75 c и для углов наклона 30 и 80 градусов.
@@
 	\centering
-	\includegraphics[width=0.4\textwidth]{./fig/distributions2_momentum.png} 
+	\includegraphics[width=0.8\textwidth]{./img_part1/distributions2_momentum.png} 
 	\caption{Функции распределения протонов и электронов, полученные объединением результатов PIC и МОнте-Карло моделирования со скоростью ударной волны 0.75 c.
@@
 	\centering
-	\includegraphics[width=0.4\textwidth]{./fig/long_radiation.png} 
+	\includegraphics[width=0.8\textwidth]{./img_part1/long_radiation.png} 
 	\caption{Модельный спектр излучения синхротронного излучения с учетом композитной функции распределения электронов.

### 2025-11-18 disser/part1.tex
fz

@@
 \chapter{Моделирование излучения FBOT}\label{FBOT}
-
-Последние наблюдения открыли новый класс объектов - быстрые голубые оптические транзиенты (FBOT)\cite{Drout2014, Ho2019cow, Coppejans2020, Ho2020, PerleyCamel, YaoAt2020mrf, MatthewsAT2022tsd, ChrimesAT2023fhn, Perleyvth}.
+\section{Введение}
+\subsection{Быстрые голубые оптические транзиенты}
+Последние наблюдения открыли новый класс объектов - быстрые голубые оптические транзиенты (FBOT)\cite{Margutti2019, Drout2014, Ho2019cow, Coppejans2020, Ho2020, PerleyCamel, YaoAt2020mrf, MatthewsAT2022tsd, ChrimesAT2023fhn, Perleyvth}.
 Они характеризуются большой светимостью $L > 10^{43} эрг/с$, смещением цветов в синюю сторону, коротким характерным временным масштабом порядка нескольких дней, низкой выброшенной массой эжекты и её высокими скоростями.
@@
 Так же возможно их объяснение событиями приливных разрывов.
+
+\begin{small}
+	\topcaption{Известные быстрые голубые оптические транзиенты}
+	\label{FBOTs}
+		\begin{xtabular}{|p{0.15\textwidth} | p{0.15\textwidth}| p{0.1\textwidth}| p{0.15\textwidth}| p{0.17\textwidth}| p{0.15\textwidth}|}
+			\hline
+			Название &дата & скорость эжекты & светимость, эрг/с&  рентгеновская светимость, эрг/с& расстояние, Мпк\\
+			\hline
+			AT2018cow  “The Cow”\cite{Ho2019cow}& 16.06.2018
+			 & 0.13c & 4$\times 10^44$  & $1\times 10^{43}$ & 62\\
+			\hline
+			ZTF18abvkwla “The Koala\cite{Ho2020}”& 12.09.2018
+			 & 0.3c& $1\times 10^44$&  ? & 1400\\
+			\hline
+			CSS161010\cite{Coppejans2020}
+			& 10.10.2016
+			 & 0.55c& ?&  $5\times 10^{39}$& 150\\
+			\hline
+			AT2020xnd “The Camel”\cite{PerleyCamel}& 12.10.2020
+			 & 0.2c& $1\times10^43$&  $7\times 10^{42}$& 1260\\
+			\hline
+			AT2020mrf\cite{YaoAt2020mrf}& 12.06.2020
+			 & 0.08c& ?&  $2\times 10^{43}$& 640\\
+			\hline
+			AT2022tsd
+			“Tasmanian devil”\cite{MatthewsAT2022tsd}& 7.09.2022
+			 & ?& ?&  $1\times 10^{44}$& 1300\\
+			\hline
+			AT2023fhn
+			“The Finch”\cite{ChrimesAT2023fhn}& 10.04.2023
+			 & ?& ?&  $1\times 10^{42}$& 1200\\
+			\hline
+			AT2023vth\cite{Perleyvth}& 17.11.2023
+			 & ?& 17&  ?& 340\\
+			\hline
+			AT2024wpp\cite{Nayana2024wpp}& 25.09.2024& 0.42c & $1\times10^45$& $1.5\times10^43$& 410\\
+			\hline
+		\end{xtabular}
+\end{small}
 
@@
 
-\section{Излучение объектов с синхротронным самопоглощением}
+\subsection{Модель излучения объектов с синхротронным самопоглощением}
 Процесс синхротронного излучения хорошо известен и описан в классических работах.
@@
 Так же, измеряя размер источника $R$ в разные моменты времени можно оценить скорость распространения ударной волны.
-\section{Particle-in-Cell моделирование }
+\section{Модель}
+\subsection{Particle-in-Cell моделирование ускорения частиц}
 Для более точного расчета синхротронного излучения оптических транзиентов необходимо знание функции распределения электронов.
@@
 
-\section{Фитирование парметров радиоисточника}
+\subsection{Фитирование радиоизлучения источника CSS161010}
 
@@
 
-\section{Расчет рентгеновского излучения}
+\subsection{Расчет рентгеновского излучения источника CSS161010}
 На 99-ый день с момента детектирования имеются одновременные наблюдения в радио и в рентгеновском диапазоне.
@@
 \FloatBarrier
-\section{Заключение}
+\section{Выводы}
 

### 2025-11-23 disser/part1.tex
disser

@@
 \subsection{Быстрые голубые оптические транзиенты}
-Последние наблюдения открыли новый класс объектов - быстрые голубые оптические транзиенты (FBOT)\cite{Margutti2019, Drout2014, Ho2019cow, Coppejans2020, Ho2020, PerleyCamel, YaoAt2020mrf, MatthewsAT2022tsd, ChrimesAT2023fhn, Perleyvth}.
-Они характеризуются большой светимостью $L > 10^{43} эрг/с$, смещением цветов в синюю сторону, коротким характерным временным масштабом порядка нескольких дней, низкой выброшенной массой эжекты и её высокими скоростями.
+Последние наблюдения открыли новый класс астрофизических объектов - быстрые голубые оптические транзиенты (FBOT)\cite{ Drout2014, Ho2019cow, Margutti2019}.
+Они характеризуются большой светимостью $L > 10^{43} эрг/с$, смещением цветов в синюю сторону и коротким характерным временным масштабом порядка нескольких дней.
+Из анализа их спектров и кривых блеска следуют так же маленькая масса выброшенной эжекты (сотые и тысячные доли солнечной массы) и её высокие скорости ( > 0.1 c) \cite{Margutti2019}.
+Всего имеется несколько десятков кандидатов в данный класс объектов.
+Иногда наиболее яркие объекты, обладающие большой светимостью как в оптическом, так и в радио и рентгеновском диапазоне выделяют в подкласс ярких транзиентов (LFBOT).
+Характеристики ярких быстрых голубых оптических транзиентов приведены в таблице \ref{FBOTs}.
+
 Точная природа данных объектов на даный момент неизвестна.
-Вместе со слабыми гамма-всплесками они возможно принадлежат к переходному классу объектов между нерелятивистскими сверхновыми и обычными гамма-всплесками.
-Так же возможно их объяснение событиями приливных разрывов.
+Некоторые модели предполагают коллапсирующие сверхновые с ассиметричной эжектой \cite{Margutti2019, Chrimes2025} или с формированием джета \cite{GottliebFBOTjet}.
+В других моделях рассматриваются события приливных разрывов \cite{Metzger2022, Kuin2019}, слияния различных компактных объектов \cite{Lyutikov2022} или аккреция остатка сверхновой на черную дыру - компаньона \cite{Lazzati2024}.
+В данной работе мы не касаемся вопроса о происхождении быстрых голубых оптических транзиентов, а рассматриваем процессы ускорения частиц и нетплового излучения, связанные с наличием трансрелятивистских потоков вещества.
 
@@
 			\hline
-			AT2024wpp\cite{Nayana2024wpp}& 25.09.2024& 0.42c & $1\times10^45$& $1.5\times10^43$& 410\\
+			AT2024wpp\cite{NayanaAT2024wpp}& 25.09.2024& 0.42c & $1\times10^45$& $1.5\times10^43$& 410\\
 			\hline
+			%AT2025wap\cite{Nayana2024wpp}& 21.08.2025& 0.42c %& $1\times10^45$& $1.5\times10^43$& 410\\
+			%\hline
 		\end{xtabular}

### 2025-11-25 disser/part1.tex
disser

@@
 
-Рассматриваемый в данной главе объект CSS161010, расположенный в карликовой галактике на расстоянии 150 мегапарсек имеет по оценкам следующие\cite{Coppejans2020} характеристики: скорость эжекты 0.55 c на ранних этапах и около 0.3 c на поздних (порядка одного года), выброшенная масса порядка 0.01-0.1 солнечных масс.
-Он обладает наибольшей скоростью эжекты из всех известных на текущий момент FBOT объектов, и следовательно наиболее удобен для Particle-in-Cell моделирования в виду малого времени ускорения.
-\textcolor{red}{Важнейшей особенностью этих объектов является вид их спектральной плотности излучения, имеющей выраженный максимум и степенные участки, говорящие о наличии синхротронного самопоглощения.}
-%почему четвертый был отдельно в статье?
+\subsection{Модель излучения объектов с синхротронным самопоглощением}
 
-В данной главе будет проведено Particle-in-Cell моделирование ускорения частиц в транс-релятивистской ударной волне со скоростями, характерными для FBOT объектов.
-По имеющимся распределениям будет расчитано синхротронное радио-излучение и с помощью фитирования наблюдательных данных транзиента CSS16 1010 будут подобраны такие параметры как магнитное поле, концентрация вещества и геометрические характеристики в ударной волне.
-Далее, с использованием широокого распределения частиц, полученного из Монте-Карло моделирования с соответствующими параметрами, будет расчитано рентгеновское излучение от данного объекта.
+Одной из особенностьей излучения быстрых оптических транзиентов является наличие характерного пика в спектре радиоизлучения, связанного с синхротронным самопоглощением.
+Именно благодаря этому удается вычислить скорость расширения данных объектов без непосредмтвенного наблюдения.
 
-\subsection{Модель излучения объектов с синхротронным самопоглощением}
 Процесс синхротронного излучения хорошо известен и описан в классических работах.
@@
 Так же, измеряя размер источника $R$ в разные моменты времени можно оценить скорость распространения ударной волны.
+
+Рассматриваемый в данной главе объект CSS161010, расположенный в карликовой галактике на расстоянии 150 мегапарсек имеет по оценкам следующие\cite{Coppejans2020} характеристики: скорость эжекты 0.55 c на ранних этапах и около 0.3 c на поздних (порядка одного года), выброшенная масса порядка 0.01-0.1 солнечных масс.
+Он обладает наибольшей скоростью эжекты из всех известных на текущий момент FBOT объектов, и следовательно наиболее удобен для Particle-in-Cell моделирования в виду малого времени ускорения.
+
+В данной главе будет проведено Particle-in-Cell моделирование ускорения частиц в транс-релятивистской ударной волне со скоростями, характерными для FBOT объектов.
+По имеющимся распределениям будет расчитано синхротронное радио-излучение и с помощью фитирования наблюдательных данных транзиента CSS16 1010 будут подобраны такие параметры как магнитное поле, концентрация вещества и геометрические характеристики в ударной волне.
+Далее, с использованием широокого распределения частиц, полученного из Монте-Карло моделирования с соответствующими параметрами, будет расчитано рентгеновское излучение от данного объекта.
 \section{Модель}

### 2025-12-01 disser/part1.tex
disser

@@
 	\label{parameters}
-	\caption{Полученные значения параметров радиоисточника для различных моделей}
+	\caption{Полученные значения параметров радиоисточника  на 99 день после взрыва для различных моделей}
 	\begin{center}
@@
 На 357-ой день магнитное поле равно $0.04~\rm G$, концентрация $0.25~\rm cm^{-3}$, а размер источника $R= 6.9\times10^{17}~\rm{cm}$.
+Результаты моделирования спектров излучения показаны на рисунке \ref{radiation}.
 Полученные значения магнитного поля и концентрации значительно отличаются от простых ожидаемых зависимостей от расстояния $B \propto r^{-1}$ and $n  \propto r^{-2}$, а так же попытка одновременного фитирования всех трех моментов времени со степенными зависимостями размера от времени и магнитного поля и концентрации от расстояния не удалась, что может говорит о сильной неоднородности и немонотонности плотности вокруг источника.
 
-Расчеты были проведены для трех конфигураций источника - степенной функцией распределения и параметрами из статьи \cite{Coppejans2020}, с функцией распределения, полученной из PIC моделирования, и с функцией с экстраполированным степенным хвостом до энергий $\displaystyle 500~m_{e} c^{2}$.
-В последних двух случаях размер источника и магнитное поле подбирались минимизацией функционала $\displaystyle g(B,R) = \sum (F(\nu_{i}, B, R) - F_{obs}(\nu_{i}))^{2}$, где $\nu_i$
-наблюдаемые частоты, $\displaystyle F_{obs}(\nu_{i})$ наблюдаемые потоки
-$\displaystyle F(\nu_{i}, B, R)$---рассчитанные потоки.
-Мы использовали шесть измерений из ~\cite{Coppejans2020}: четыре сделанные VLA на 357-ой день
-на частотах 1.5, 3.0, 6.05 и 10~GHz и два сделанные GMRT на 350-ый день на частотах 0.33 и 0.61 GHz. Результаты расчетов излучения показаны на рисунке \ref{radiation}.
-Как можно видеть из рисунка, модель с функцией распределения полученной непосредственно из PIC расчетов не может объяснить наблюдаемое излучение.
-Но так как время расчетов намного порядков меньше времени наблюдения источника, мы считаем возможным экстраполировать степенной хвост функции распределения до более высоких энергий.
-Такая экстраполированная функция распределения, как и чисто степенная, хорошо объясняют наблюдения.
-Но полученные физические параметры источника оказываются существенно различными.
-Магнитное поле $\displaystyle B = 0.069$~G, и радиус источника $\displaystyle R = 3.0 \times
-10^{17}$~cm близки к значениям из статьи \cite{Coppejans2020}, в то время как полученное значение концентрации $\displaystyle n = 210$~cm$^{-3}$ оказывается значительно больше.
 \begin{figure}
 	\centering
-	\includegraphics[width=10.5 cm]{radiation.eps} 
+	\includegraphics[width=10.5 cm]{./img_part1/radiation4.png} 
 	\caption{Наблюдаемые 
-		(зеленые круги) и расчетные спектральные плотности излучения CSS161010  на 357-ой день для разных моделей распределения электронов.} 
+		 и расчетные спектральные плотности излучения CSS161010  на 69, 99 и 357-ой день для модели с углом наклона магнитного поля 30 и скоростью потока 0.75 c.} 
 	\label{radiation} 

### 2025-12-01 disser/part2.tex
disser

@@
-\chapter{Классификация гамма-всплесков, зарегистрированных в эксперименте Конус-Винд}\label{KW_GRB_classification}
+\chapter{Моделирование морфологии и высокоэнергичного излучения W50}\label{SS433chapter}
 
+\FloatBarrier
 \section{Введение}
-Первым свидетельством наличия двух классов всплесков было обнаружение их бимодального 
-распределения по длительности~\citep{Mazets_1981_part_1,Norris_1984,Kouveliotou_1993,Aptekar_1998}. 
-В работе \citep{Kouveliotou_1993} были введены две меры длительности $T_{50}$ и $T_{90}$, 
-которые соответствуют временам накопления 50\% и 90\% от суммарного числа отсчётов 
-всплеска (интервалы от 25\% до 75\% и от 5\% до 95\% отсчётов, соответственно), 
-и определена граница между длинными и короткими всплесками $T_{90}=2$~с 
-на основе локального минимума бимодального распределения. 
-В работе \citep{McBreen_1994} обсуждается, что распределение всплесков по 
-длительности $T_{90}$ хорошо описывается суммой двух логнормальных компонент. 
-Более детальный анализ распределения по $T_{90}$ всплесков из третьего каталога 
-эксперимента BATSE~\citep{Horvath_2002} показал, что распределение 
-хорошо аппроксимируется тремя логнормальными компонентами. 
-Третья компонента располагалась между короткими и длинными всплесками, 
-что свидетельствовало о наличие третьего класса всплесков с промежуточной длительностью.
-Доля промежуточных всплесков составила 6\%.
 
-Систематические ошибки величин $T_{50}$ и $T_{90}$, которые искажают наблюдаемое 
-распределение всплесков, были рассмотрены в работе~\citep{Norris_and_Bonnel_2006ApJ} 
-на примере всплесков, зарегистрированных в эксперименте \textit{CGRO}-BATSE~\citep{Fishman_1992NASCP3137}. 
-Было показано, что как минимум четыре фактора влияют на измеренное значение длительности 
-всплеска и размывают границы между классами коротких и длинных всплесков: 
-(1)~различие в отношении сигнал-шум (S/N) между самым интенсивным и самым слабым 
-всплеском  при условии, что эти всплески не отличаются по другим параметрам, 
-приводит к различию их длительностей $T_{90}$ в $\sim 2$~раза~\citep{Bonnell_1997}; 
-(2)~длительность космологически удалённых длинных всплесков (с красными смещениями источников 
-$z \sim 2 \textrm{--}10$) увеличивается в 2--5 раз по сравнению со всплесками, 
-пришедшими с $z=1$, в это же время поправка длительности 
-для коротких всплесков не превышает двух раз, так как их источники в основном 
-расположены на $z<1$ (этот фактор увеличивает обособленность распределений 
-коротких и длинных всплесков); (3)~эффект зависимости длительности отдельных импульсов 
-всплесков от энергии $\propto E^{-0.35}$~\citep{Fenimore_1995}, а так же модификация 
-спектра за счёт космологического расширения приводит к зависимости длительности 
-всплеска от энергетического диапазона наблюдений; (4)~распределение по длительностям 
-обрезано со стороны коротких всплесков вследствие наблюдательной селекции триггерного 
-алгоритма.
+Туманность W50 - протяженная структура, видимая в радио лучах и создаваемая компактным источником SS 433 - двойной системой с аккрецирующей черной дырой (или нейтронной звездой) \cite{Margon1984, Fabrika2004, Cherepashchuk2025} показана на рисунке \ref{W50image}.
+Аккреция в данной системе вероятно в течение долгого времени поддерживается в сверх-эдингтоновском режиме \cite{Heuvel2017, Cherepashchuk2023}.
+Полная энергия, выделенная источником в окружающую среду сопоставима с энергией взрыва сверхновой.
+Поэтому SS 433 сильно воздействует на межзвездную среду, создавая туманность протяженностью больше 100 парсек \cite{Begelman1980}.
 
-Кроме $T_{50}$ и $T_{90}$ была предложена и другая мера длительности~--- время 
-эмиссии~\citep{Mitrofanov_1999}, которая вычисляется как суммарная длительность 
-бинов временной истории, в которых интенсивность излучения превышает заданный 
-уровень от пиковой интенсивности. Этот уровень задается таким образом, чтобы 
-суммарное число отсчетов в бинах составляло заданную долю от общего числа отсчётов 
-всплеска (использовались уровни 30\% и 50\%, соответствующие меры длительности 
-были обозначены $\tau_{30}$ и $\tau_{50}$). Было показано, что данная мера 
-длительности гораздо менее чувствительна и к отношению сигнал-шум и к количеству 
-импульсов во всплеске, разделенных значительным промежутком фона, так как этот 
-промежуток не входит, по определению, в определяемое время эмиссии. Ясно, однако, 
-что эта мера чувствительна к временному разрешению. Распределения всплесков 
-по $\tau_{30}$ и $\tau_{50}$  также являются бимодальными. Эти меры длительности 
-не получили широкого распространения, поэтому классификация всплесков 
-на их основе в данной работе не рассматривается.
+В процессе аккреции SS~433 производит транс-релятивистские джеты со скоростью потока $\sim 0.26\,c$ и углом растовора менее двух градусов \citep[e.g.
+][]{Begelman1980, Calvani1981}.
+Эти узкие джеты видны в рентгеновском и оптическом диапазоне как пары линий сдвинутых в синюю и красную стороны, излучающиеся в пространственных областях $\lesssim 10^{12}$ см и $\lesssim 10^{15}$ см от источника соответственнои \citep[][]{Marshall2002, Fabrika2004,Khabibulin2016}.
+Однако сами джеты невидимы на расстояниях больших $\sim0.1$ пк от центрального источника \citep[e.g.
+][]{Hjellming1981,Blundell2004}.
+Но так как на больших масштабах (от 40 до 110 парсек) окружающая SS~433 яркая радио и ${\rm H}\alpha$ туманность W50 имеет вытянутую структуру с соотношением 3:1 в направлении оси прецессии джета, предполагается, что именно джеты надувают туманность W50  \citep[][]{Begelman1980,Zealey1980,Eichler1983,Peter1993,Zavala2008, Panferov2017}.
 
-Короткие и длинные всплески, в среднем, отличаются жёсткостью спектра.
-В работе \citep{Kouveliotou_1993} было обнаружено, что короткие всплески BATSE, в 
-среднем, имеют большую жёсткость ($\rmn{HR}_{32}$), чем длинные, где $\rmn{HR}_{32}$ определялась 
-как отношение числа отсчетов, накопленных за $T_{90}$ в энергетических 
-диапазонах 100--300~кэВ и 50--100~кэВ (диапазоны 3 и 2) эксперимента \textit{CGRO}-BATSE. 
-Однако, в последующих работах было отмечено, что жёсткость не является обязательной чертой 
-коротких всплесков~\citep[см., например,][]{Sakamoto_2006_proc, Norris_and_Bonnel_2006ApJ}.
+There is no direct evidence of the termination of the narrow jets in the radio, optical and X-ray maps of W50 \citep[][]{1998AJ....116.1842D,2011MNRAS.414.2838G,2017MNRAS.467.4777F,2018MNRAS.475.5360B}, while a
+broader and more extended X-ray structures are seen on both sides of SS~433  aligned with the narrow optical jet precession axis \cite[e.g.,][]{1983ApJ...273..688W,1994PASJ...46L.109Y,Safi-Harb1997,1999ApJ...512..784S,Brinkmann2007,Safi-Harb2022}.
+These extended X-ray jets appear at a distance of $\sim 20\,$pc from the compact source and have an opening angle of $\sim 20^\circ$ which is twice smaller than the precession amplitude of the narrow jets.
+The X-ray structures are probably of synchrotron origin \citep[e.g.,][]{Brinkmann2007}, which is confirmed by the detection of X-ray polarization by the IXPE telescope \citep{IXPE2024SS433}.
+The very high energy (VHE) gamma-ray emission from the source was predicted in a number of models  \citep[see, e.g.,][]{Safi-Harb1997, Reynoso2008, 1998NewAR..42..579A,
+	2020ApJ...904..188K, 2020ApJ...889..146S}.
+Recently, the extended very high energy gamma-ray emission was detected from W50 \citep[][]{2018Natur.562...82A,HESS2024SS433,LHAASO2024SS433}, indicating the acceleration of PeV-regime particles in the source.
+The physical origin of these extended structures and their connection to the narrow transrelativistic jets currently launched by the system remains unclear \citep[e.g.
+][ and references therein]{2020MNRAS.495L..51C}
 
-На основе распределения на плоскости жёсткость-длительность также был обнаружен 
-третий класс всплесков ~\citep{Mukherjee_1998, Hakkila_2000}. Анализ влияния 
-систематических эффектов на наличие третьего класса всплесков был проведён 
-в~\citep{Hakkila_2003}. Авторы показали, что третий класс всплесков не является 
-физическим и может являться следствием селекции триггерного алгоритма. Также было 
-показано, что сам по себе эффект зависимости длительности от яркости 
-всплеска~\citep{Bonnell_1997} не может привести к появлению третьего класса всплесков. 
-Однако, в последующей работе~\citep{Horvath_2006} были приведены доводы в пользу 
-реальности третьего класса всплесков и исследовалось количество и значимость 
-классов всплесков на основе параметров распределения на плоскости 
-$\log T_{90}$--$\log \rmn{HR}_{32}$ для 1956 всплесков, зарегистрированных 
-в эксперименте BATSE. При этом распределение аппроксимировалось 
-суммой двумерных нормальных распределений. Было установлено, что распределение 
-наилучшим образом описывается тремя компонентами (классами). При этом доля всплесков 
-из класса с промежуточной длительностью составляет 10--15\%. Также было показано, что поправка набора 
-всплесков на селекцию триггерного алгоритма сохраняет полученное разделение на классы. 
-Аналогичный метод был применен к 222 всплескам \textit{Swift}-BAT~\citep{Horvath_2010}, 
-при этом также было обнаружено три класса всплесков. При этом доля промежуточного 
-класса всплесков составила~30\%.
+The `minimalist' model \citep{Churazov} explains the extended jet as a result of the powerful two-component anisotropic outflow produced by a supercritical accretion disk.
+In this model, a more collimated outflow (polar wind), aligned with the rotation axis of the binary, coexists with a more isotropic wind.
+The extended X-ray jets observed are associated with the recollimation shocks, which appear right after the collision of the collimated outflow with the isotropic wind termination surface.
+The structure of  MHD outflow in the interaction region of the collimated wind with the termination surface of the isotropic wind was simulated with the PLUTO MHD code \cite{Mignone2007}.
+We present a model of very high-energy particle acceleration in the extended jets of the SS~433/W50 system following the two-component outflow scenario.
 
-Помимо жесткости и длительности, еще одним параметром для классификации всплесков 
-может служить спектральная задержка. Этот параметр характеризует запаздывание 
-излучения в более мягком диапазоне по отношению к излучению в более жестком диапазоне. 
-Значение задержки ($\tau_\rmn{lag}$) соответствует положению  максимума 
-кросскорреляционной функции временных историй в двух энергетических диапазонах~\citep{Norris_2000}. 
-Анализ спектральных задержек коротких всплесков BATSE и \textit{Swift}-BAT показал, 
-что они распределены вблизи нуля с разбросом 
-порядка 25~мс \citep{Norris_and_Bonnel_2006ApJ, Norris_2011ApJ}. В то же время длинные 
-всплески могут иметь значительные задержки. Идея о том, что различие спектральных 
-задержек длинных и коротких всплесков отражает различие физических свойств источников 
-всплесков, была выдвинута в работе~\citep{Gehrels_2006_Nature}.
+The results of MHD modeling are described in Section \ref{sec:morphological}.
+Using the spatial distributions of the accelerated particles and magnetic field spectra obtained in the model, we construct profiles and spectra of non-thermal X-ray and gamma radiation of this object.
+The modeling includes the following steps, split across several sections.
 
-%Существует несколько моделей, объясняющих появление спектральной задержки. 
-%В работе~\citep{Ioka_and_Nakamura_2001} появление задержки объясняется, тем что 
-%узкий джет наблюдается под разными углами для разных всплесков. Большая задержка 
-%соответствует большему углу наблюдения относительно оси джета. 
-%Авторы работы~\citep{Schaefer_2004_lag} считают, что спектральная задержка в 
-%каждом импульсе всплеска характеризует время потери энергии излучающими частицами.
+Section \ref{sec:MC}.
+Based on the plane-parallel Monte Carlo model of particle acceleration and magnification of magnetic fields near the shock fronts, the spectra of accelerated particles and magnetic fields near the shock waves obtained in the MHD model are found.
 
-После начала эры многоволновых наблюдений гамма-всплесков стало возможным 
-классифицировать всплески на основании параметров послесвечений и родительских галактик. 
-В работах~\citep{Zhang_2006, Zhang_2007, Zhang_2009} была предложена схема классификации всплесков, 
-основанная на этих параметрах, на физические типы: I и~II. Считается, что всплески типа~I, 
-генерируются при слиянии компактных объектов, а всплески типа~II образуются 
-при коллапсе молодых массивных звёзд. В работе~\citep{Zhang_2009} было показано, 
-что всплески типа II на плоскости жёсткость-длительность располагаются в области 
-длинных/мягких всплесков, при этом всплески типа~I~--- в области коротких/жестких событий. 
-Авторы также отмечают, что всплески типа I обладают незначительной спектральной задержкой.
+Section \ref{sec:CorSh}.
+Based on MHD modeling of the passage of turbulence with significant density fluctuations through the shock front, profiles of the longitudinal $B_{\parallel}$  and transverse $B_{\perp}$ components of the magnetic field behind the shock front are obtained using the PLUTO code.
 
-Неоднозначность соответствия коротких всплесков и всплесков типа~I связана, в том числе, с 
-существованием коротких гамма-всплесков, сопровождающихся слабым сравнительно мягким 
-излучением в гамма-диапазоне, длящимся десятки секунд после начального короткого импульса. 
-Это явление получило название продленного излучения (EE).
-Впервые этот класс всплесков был обнаружен в данных BATSE~\citep{Burenin_2000AstL} и 
-Конус~\citep{Mazets_2002astro_ph}. В работе~\cite{Frederiks_2004ASPC} 
-было обнаружено 11 всплесков с продлённым излучением среди 125 коротких гамма-всплесков KW. 
-Для оставшихся всплесков был обнаружен избыток излучения на интервале 5--100~с в сумме временных 
-профилей с вычтенным фоном. Такой же избыток был выявлен в данных 
-BATSE~\citep{Lazzati_2001AandA, Connaughton_2002ApJ}, \textit{BeppoSAX}-GRBM~\citep{Montanari_2005ApJ} и 
-\textit{INTEGRAL}-SPI-ACS~\citep{Minaev_2010AstL}. 
-%написать про всплески с EE на BAT и GBM
-Основную сложность представляет разделение коротких всплесков с EE 
-и длинных всплесков. Авторы работы~\citep{Norris_and_Bonnel_2006ApJ} использовали визуальные признаки 
-для выделения коротких всплесков с продлённым излучением: длительность начального 
-импульса $T_{90}<2$~с, отсутствие заметной спектральной эволюции в начальном 
-импульсе и EE. Авторы обнаружили в данных BATSE 8 всплесков, 
-удовлетворяющих этим критериям, и показали, что их спектральные задержки близки к нулю. 
-Также в работе отмечается наличие у некоторых всплесков интервала с практически нулевой интенсивностью излучения 
-между начальным импульсом и продлённым излучением. Сравнение интенсивностей продлённого 
-излучения для этих 8 всплесков и интенсивности излучения, обнаруженного в 
-суммарном временном профиле в работе~\citep{Lazzati_2001AandA}, показывает, 
-что динамический диапазон интенсивности продлённого излучения составляет $\sim 10^4$, 
-так же как и динамический диапазон отношения интенсивности начального импульса и продлённого излучения.
-В работе~\citep{Norris_2010ApJ} на основе 12 всплесков с продлённым излучением, зарегистрированных 
-\textit{Swift}-BAT, был оценен физический порог отношения интенсивностей продлённого излучения 
-и начального пика как единицы $\times 10^3$ и было показано, что для начальных импульсов 
-всплесков без продлённого излучения характерны в $\sim 2\textrm{--}3$ раза меньшие длительности
- по сравнению с всплесками с EE, также всплески без продлённого излучения 
-имеют меньшее число пиков в начальном импульсе.
+Section \ref{sec:Spectrum}.
+Next, the spectra and profiles of the X-ray synchrotron emission are calculated based on the spectra of accelerated electrons and magnetic field profiles.
+Gamma Rays in the TeV range are also calculated.
+The simulation parameters are selected to match the model with the observational data \citep[][]{Brinkmann2007, Safi-Harb2022, HESS2024SS433, LHAASO2024SS433}.
+The anisotropy of the magnetic field in the shock downstream leads to the observed polarization of the X-ray radiation.
+The X-ray polarization obtained in the model near the shock wave is compared with the observational data from the IXPE observatory \citep{IXPE2024SS433}.
 
-Различие природы источников коротких гамма-всплесков с продлённым излучением и 
-без него в настоящее время не подтверждено. В работе~\citep{Troja_2008MNRAS} 
-было установлено, что смещения относительно центра родительской галактики 
-коротких всплесков с продлённым излучением в среднем меньше, чем для всплесков без него: 
-единицы и десятки килопарсек, соответственно. 
-Исследование~\citep{Fong_2010ApJ} с использованием данных наблюдений 
-телескопа \textit{Hubble Space Telescope} опровергло этот результат.
+High energy particle acceleration in microquasars as a potential contributor to the galactic cosmic ray pool was widely discussed (see e.g.
+\cite{2002A&A...390..751H, 2005A&A...432..609B,2017SSRv..207....5R}), while recently the interest to SS 433 as a pevatron was raised in \cite{Churazov,2024arXiv241108762P,2025arXiv250620193Z,2025arXiv250622550C,2025arXiv250721048W} by VHE gamma-ray observations reported in \cite{HESS2024SS433,LHAASO2024SS433}.
 
-В данной главе анализируются временные истории всплесков, зарегистрированных в 
-эксперименте Конус-Винд. Всплески классифицируются на основе длительности, жесткости и 
-спектральной задержки, и определяется набор коротких всплесков для дальнейшей локализации и спектрального анализа. 
-При этом обсуждается принадлежность исследуемых всплесков к типам~I и~II. 
-В разделе~\ref{sec:GRB_sample} приводится описание использованного набора всплесков. 
-В разделе~\ref{sec:Durations} рассматривается распределение всплесков по длительности, 
-определяется граница между длинными и короткими всплесками на основе 
-длительности $T_{50}$ и определяется итоговый набор коротких всплесков. 
-В разделе~\ref{sec:Hardness} рассматриваются жесткости всплесков и производится 
-классификация всплесков на основе соотношения жёсткость-длительность. 
-В разделе~\ref{sec:Lags} анализируются спектральные задержки коротких всплесков. 
-Сравнение классификации на физические типы~I и~II и классификации на основе жесткости, 
-длительности и спектральной задержки приведено в разделе~\ref{sec:Phys_Classification}. 
-Обобщение и обсуждение результатов приведено в разделе~\ref{sec:Conclision}.
-
-\section{Набор всплесков}\label{sec:GRB_sample}
-За период с ноября 1994~г. по декабрь 2010~г. KW зарегистрировал 2008 гамма-всплесков 
-в триггерном режиме. Среди них 69 всплесков содержат сбои, представляющие собой 
-промежутки в данных во время всплеска, или значительные флуктуации фона, 
-накладывающиеся на всплеск. Всего 1939 всплесков без сбоев.
-
-Для вычисления параметров временных историй использовалась 
-сшивка фоновой и триггерной записи. Длительность всплесков, у которых значительная 
-часть события лежит в бине фоновой записи, предшествующем триггерной записи, может быть 
-искажена из-за низкого разрешения фоновой записи. Из дальнейшего рассмотрения 
-было исключено 105 таких событий. Таким образом, для анализа использовались временные 
-истории 1834 гамма-всплеска.
-
-\section{Длительности}\label{sec:Durations}
-Для вычисления длительностей мы использовали сумму временных историй всплеска 
-в энергетических диапазонах G2 и G3. Такой выбор обоснован тем, что  пиковая 
-энергия $E F_{E}$ спектра ($E_\rmn{p}$) абсолютного большинства всплесков лежит в 
-диапазоне $E_\rmn{p}>80$~кэВ, и, следовательно, отсчёты, соответствующие основному энерговыделению 
-во всплеске, регистрируются в диапазонах G2 и G3. Фон в G2 и G3 стабилен на более 
-длительных интервалах времени, чем в G1. Это позволяет с гораздо меньшей вероятностью 
-неверно определять начало и конец всплеска из-за флуктуаций фона. Также в диапазон G1 
-может попадать начало рентгеновского послесвечения, учёт которого исказит 
-длительность начального импульса всплеска.
-
-Уровни фона для вычисления длительностей определялись по методике, описанной в 
-разделе~\ref{sec:Bg_rate} Главы~\ref{KW_description}. 
-В результате, из всего набора 1834 всплесков у 96\% всплесков длительность 
-интервала для определения фона составила 750~с и только у $<1$\% всплесков длительность этого 
-интервала составила 30--100~c.
-
-\subsection{Автоматическая процедура определения длительности}
-Основной задачей при вычислении любой меры длительности является определение момента 
-начала и конца всплеска. В работе~\citep{Koshut_1996}, описывающей методику вычисления 
-длительностей $T_{90}$ и $T_{50}$  для второго каталога BATSE, моменты начала и 
-конца всплеска $t_{0}$ и $t_{100}$ определялись визуально на основании интеграла 
-от временной истории с вычтенным фоном всплеска, где $t_{x}$ соответствует времени накопления заданной 
-доли $x$  (0\%, 5\%, \dots, 100\%) от полного числа отсчётов всплеска. 
-В работе~\citep{Bonnell_1997} использовалась автоматическая процедура поиска 
-$t_{0}$ и $t_{100}$, которая заключалась в поиске положительной флуктуации числа 
-отсчетов с вычетом фона со значимостью от $3\sigma$ до $6\sigma$ на временных 
-масштабах от 0.5~с до 16~с в диапазоне~$>25$~кэВ.
-% что для Swift-BAT и Fermi-GBM?
-
-Для вычисления длительности всплесков KW использовалась автоматическая 
-процедура поиска начала и конца всплеска. Поиск производился на интервале от 
-$T_0-250$~с до $T_0+230$~с (конец триггерной записи), где $T_0$~--- время триггера KW. 
-Для некоторых всплесков границы интервала вычисления длительности отличались от указанных, 
-к примеру, в случае присутствия солнечной вспышки в данных. Для поиска начала 
-всплеска производился поиск положительной флуктуации числа отсчетов с вычетом 
-фона на заданном уровне значимости. При этом поиск превышения производился на 
-интервале с длительностью от временного разрешения текущего участка записи до 100~с. 
-Началом всплеска $t_0$ считается начало интервала, на котором было обнаружено превышение, 
-если вероятность его случайного обнаружения была меньше пороговой. 
-Иллюстрация алгоритма поиска начала всплеска приведена на рис.~\ref{fig:T100_calculation}.
-Времена начала интервалов поиска превышения брались последовательно, начиная от $T_0-250$~с. 
-Конец всплеска~--- $t_{100}$ определяется аналогичного, при этом поиск ведётся в 
-обратном направлении по времени, начиная от $T_0 + 230$~с. Длительности всплесков 
-вычислялись для порогов, соответствующих вероятности случайного превышения 
-значений $4\sigma$, $5\sigma$ и $6\sigma$ для распределения Гаусса. При этом на 
-интервалах, где число отсчётов от фона меньше 20, для вычисления вероятности 
-случайного превышения использовалась статистика Пуассона. Значения $T_{90}$ и $T_{50}$ 
-и их ошибки определялись методом, описанном в работе~\citep{Koshut_1996}.
-
-\begin{figure}[h]
-  \begin{minipage}[h]{0.5\textwidth}
-    \center{\includegraphics[width=1.0\textwidth]{gDurStartRU.eps} \\ а)}
-  \end{minipage}
-  \hfill
-  \begin{minipage}[h]{0.5\textwidth}
-    \center{\includegraphics[width=1.0\textwidth]{gDurT5RU.eps} \\ б)}
-  \end{minipage}  
-  \caption{
-  (а) Иллюстрация определения времени начала всплеска $t_0$ для GRB20031214\_T36655. 
-  Сплошная линия~--- интегральное число отсчетов всплеска с вычетом фона. 
-  Число отсчётов $N_\rmn{acc}$ (горизонтальные штриховые линии), 
-  накопленное на интервале $\Delta T_\rmn{acc}$ (вертикальные штриховые линии), 
-  соответствует превышению $5\sigma$ над фоном. Уровень $1\sigma$ относительно интегрального 
-  числа отсчётов на момент $t_0$  изображен горизонтальной штрихпунктирной линией. 
-  Горизонтальная точечная линия обозначает нулевой уровень.
-  (б) Иллюстрация определения начала интервала $T_{90}$ ($t_5$) для того же всплеска.
-  Число отсчётов, соответствующие 5\% от общего числа отсчётов $N_\rmn{tot}$ 
-  обозначено горизонтальными штриховыми линиями, время начала всплеска $t_0$ и время $t_5$ 
-  обозначены вертикальными штриховыми линиями. \label{fig:T100_calculation} }
+\begin{figure}
+	\centering
+	\includegraphics[width=0.9\textwidth]{./img_part2/w50.png} 
+	\caption{Изображение туманности W50 из работы \cite{Safi-Harb2022}.
+Красный цвет - радио \cite{Dubner1998}, зеленый - оптическое излучение \cite{Boumis2007}, желтый - мягкий рентген (0.5-1 кэВ), фиолетовый - средний рентген (1-2 кэВ), голубой - жесткий рентген (2-12 кэВ)\cite{Safi-Harb2022, Moldowan2005}.}
+	\label{W50image} 
 \end{figure}
 
-\subsection{Распределения по длительностям}
-Для 1834 всплесков KW были построены гистограммы распределений всплесков по $T_{90}$ и $T_{50}$ 
-для порогов значимости поиска начала и конца всплеска $4\sigma$, $5\sigma$ и $6\sigma$.
+\section{Моделирование структуры туманности}
+\section{Расчет высокоэнергичного излучения}
+\subsection{Рентгеновское излучение}
+\subsection{Гамма излучение}
+\section{Выводы} \label{SS433conclision}
 
-Распределения  аппроксимировались методом минимизации $\chi^2$ суммой двух логнормальных распределений
-\begin{equation}
-f(x) = \sum_{l=1}^{2} A_l f_l(x) \mbox{ ,}
-\end{equation}
-где
-\begin{equation}
-f_l(x) = \frac{1}{w_l \sqrt{\pi/2}} \exp\left(\frac{-2(x-x_{cl})^2}{w_l^2}\right) \mbox{ ,}
-\end{equation}
-с параметрами $A_l$~--- площадь под распределением, $w=2\sigma$~--- ширина, $x_{cl}$~--- среднее значение и 
-$x=\log T$. При этом число всплесков в бине по модели равно  интегралу от аппроксимирующей функции в границах бина.
-
-Доверительные интервалы параметров распределения на уровне 68\% вычислялись 
-методом Mонте-Карло для 1000 реализаций распределения. При этом значение в каждом 
-бине распределения разыгрывалось как пуассоновская случайной величиной со средним, равным 
-числу отсчётов в соответствующем бине исходного распределения. Полученный набор 
-значений каждого параметра распределения сортировался по возрастанию. В качестве нижней и верхней 
-границы доверительного интервала на уровне 68\% из этого отсортированного массива 
-параметров брались значения с индексами 160 и 840, соответственно. 
-В качестве значения параметра бралась медиана распределения.
-
-Из-за эффектов селекции число очень коротких всплесков ($T_{90} \lesssim 100$~мс) и 
-число длинных всплесков ($T_{90} \gtrsim 200$~с) в наборе KW недооценено. Так же неверно 
-считать ошибки числа отсчётов в бине гауссовыми при числе отсчётов $<10$, поэтому 
-бины с $\log T_{90} \leq -1.2$ ($T_{90} \leq 0.063$~с) и $\log T_{90} \geq 2.4$ ($T_{90} \geq 251$~с) 
-не учитывались при аппроксимации гистограммы распределения по $T_{90}$,  для аппроксимации 
-использовалось 18 бинов. При аппроксимации гистограммы распределения по $T_{50}$ игнорировались 
-бины c $\log T_{50} \leq -1.8$ ($T_{50} \leq 0.016$~с) и  $\log T_{50} \geq 2.0$ 
-($T_{50} \geq 100$~с), для аппроксимации использовалось 19 бинов.
-
-Параметры распределений представлены в таблицах~\ref{tab:T90_distr} и~\ref{tab:T50_distr}. 
-Из представленных результатов видно, что при изменении порога от $4\sigma$ до $6\sigma$ 
-параметры распределения по $T_{50}$ и $T_{90}$ для полного набора 1834 всплесков изменяются. 
-Распределение длинных всплесков смещается в сторону меньших длительностей, такое же смещение, 
-но в меньшей степени, наблюдается для распределение коротких всплесков. 
-При этом изменение параметров распределения по $T_{90}$ гораздо существенней, 
-чем для распределения по $T_{50}$, см. таб.~\ref{tab:Txx_vs_thd},
-где $T_{\rmn{int}}$ соответствует пересечению распределений и 
-$\delta T =|T_{4\sigma}-T_{6\sigma}| / T_{4\sigma}$. 
-Так как вариации параметров распределения по $T_{50}$ существенно меньше, 
-чем для распределения по $T_{90}$ для классификации всплесков на основе длительностей 
-мы выбрали величину~$T_{50}$.
-
-\input{tables/duration_vs_thd}
-
-Вследствие наблюдательной селекции триггерной системы по отношению сигнал-шум 
-наш набор всплесков неоднороден. Как видно из диаграммы $T_{50}$--$\rmn{S/N}$~(рис.~\ref{img:SNvsT50}) 
-в наборе отсутствуют всплески с отношением сигнал-шум $\rmn{S/N} < 10$ и $T_{50} \lesssim 100$~мс. 
-Здесь S/N определено как отношение числа отсчётов от источника к стандартному отклонению 
-числа отсчётов фона на интервале 64~мс с максимальной скоростью счёта. 
-Для дальнейшего рассмотрения был выбран однородный набор всплесков с $\rmn{S/N}\geq 10$, 
-который содержит 1168 всплесков. 
-Результаты аппроксимации распределений набора по $T_{50}$ и $T_{90}$ приведены в 
-таблицах \ref{tab:T90_distr} и~\ref{tab:T50_distr}, и на рис.~\ref{img:T90andT50s5}. 
-Относительные изменения параметров скорректированных распределений при изменении 
-порога от $4\sigma$ до $6\sigma$ приведены в таб.~\ref{tab:Txx_vs_thd}.
-
-В качестве порога значимости был выбран уровень $5\sigma$ так как было замечено, 
-что при пороге $4\sigma$ в качестве начала и конца всплеска часто выбираются 
-флуктуации фона или эпизоды излучения транзиентов, не относящиеся ко всплеску  
-(установить отношение подобных слабых эпизодов к триггерному всплеску в 
-большинстве случаев не представляется возможным). При пороге $6\sigma$ часто 
-игнорируется <<хвост>>, непосредственно прилегающий к основному пику, тем самым полное 
-число отсчетов во всплеске может недооцениваться на величину до $\sim 70$\%.
-Такая ситуация возможна, к примеру, для коротких всплесков с продлённым излучением.
-
-В качестве границы между длинными и короткими всплесками была выбрана точка пересечения 
-двух логнормальных распределений по $T_{50}$ для порога значимости $5\sigma$ 
-набора 1168 всплесков $T_{50\rmn{,int}} = 0.6$~с.
-
-Для указанного распределения и границы между длинными и короткими всплесками 
-можно оценить долю <<засорения>>\ набора коротких всплесков длинными и наоборот. 
-Доля <<засорения>>\ набора коротких всплесков длинными равна отношению интеграла 
-от распределения длинных всплесков и интеграла от распределения коротких всплесков 
-на интервале $-\infty$ до $T_{50\rmn{int}}$) и составляет $\approx 7$\%. 
-При этом доля <<засорения>>\ длинных всплесков короткими составляет~$\approx 2$\%.
-
-Границы диапазонов KW менялись со временем, что могло внести систематический 
-сдвиг в длительности всплесков, зарегистрированных в различные периоды, и 
-привести к изменению со временем границы между длинными и короткими всплесками. 
-Для анализа дрейфа границы между длинными и короткими всплесками набор 1168 
-всплесков был разделён на три поднабора: 583 GRBs ноябрь 1994~-- май 2003; 
-585 GRBs июнь 2003~-- декабрь 2010; 582 GRBs  май 1999~-- январь 2007. 
-Полученные для поднаборов границы $T_{50\rmn{int}}$ равны: 
-$0.41_{-0.09}^{+0.19}$~c, 
-$0.74_{-0.16}^{+0.22}$~c и 
-$0.62_{-0.17}^{+0.28}$~c, соответственно. 
-Наблюдаемый дрейф границы лежит в пределах $1\sigma$ интервала для границы $T_{50\rmn{int}} = 0.6$~с, 
-полученной с использованием полного набора всплесков.
-
-\begin{figure} [h] 
-  \center
-  \includegraphics [width=0.75\textwidth] {gSNvsT50_ru}
-  \caption[Соотношение S/N--$T_{50}$]{Соотношение S/N--$T_{50}$ для 1834 всплесков. 
-  Пунктирная линия соответствует $\rmn{S/N}=10$. Штрихованная область наиболее подвержена 
-  селекции триггерного алгоритма. Отношение сигнал-шум $\rmn{S/N}\geq 10$ имеют 1168 всплесков.} 
-  \label{img:SNvsT50}  
-\end{figure}
-
-\begin{figure}[h]
-  \begin{minipage}[h]{0.5\textwidth}
-    \center{\includegraphics[width=1.0\textwidth]{dT90s5simple} \\ а)}
-  \end{minipage}
-  \hfill
-  \begin{minipage}[h]{0.5\textwidth}
-    \center{\includegraphics[width=1.0\textwidth]{dT90s5simple_SN10} \\ б)}
-  \end{minipage}
-  \vfill
-  \begin{minipage}[h]{0.5\textwidth}
-    \center{\includegraphics[width=1.0\textwidth]{dT50s5simple} \\ в)}
-  \end{minipage}
-  \hfill
-  \begin{minipage}[h]{0.5\textwidth}
-    \center{\includegraphics[width=1.0\textwidth]{dT50s5simple_SN10} \\ г)}
-  \end{minipage}
-  \caption[Распределения 1834 и 1168 всплесков с $\rmn{S/N}\geq 10$ по $T_{90}$ и~$T_{50}$]
-  {Распределения 1834 и 1168 всплесков с $\rmn{S/N}\geq 10$ по $T_{90}$~(а и б, соответственно) и 
-  $T_{50}$ (в и г, соответственно) для порога $5\sigma$ (непрерывная гистограмма), 
-  игнорированные при аппроксимации бины гисторграммы показаны пунктиром. 
-  Аппроксимация распределения суммой двух логнормальных распределений  показана 
-  непрерывной линией, распределение коротких и длинных всплесков показаны 
-  штрихпунктирными линиями.  Вертикальные штриховые лини обозначают средние значения 
-  распределений коротких и длинных всплесков и точку их пересечения. 
-  На нижней панели каждого рисунка показаны невязки аппроксимации.}
-  \label{img:T90andT50s5}  
-\end{figure}
-
-\begin{landscape}
-\input{tables/duration_tables_mc_diss}
-\end{landscape}
-
-\subsection{Сравнение длительностей определенных по данным BATSE и KW}
-Среди 1939 всплесков KW без сбоев 267 всплесков наблюдались BATSE. 
-Соотношение длительностей $T_{50}$ и $T_{90}$, вычисленных по данным BATSE~\citep{Paciesas_1999} 
-и KW, изображено на рис.~\ref{img:T90andT50_KWvsBATSE}.
-
-Большинство всплесков имеют б\'{о}льшую длительность по данным BATSE, 
-в особенности $T_{90}$ по сравнению с KW, что связано с большей  эффективной площадью детекторов 
-BATSE $\sim 10^3$~см$^2$ против $(0.80\textrm{--}1.6)\times 10^2$~см$^2$ у KW. 
-Другим фактором является то, что длительность всплесков BATSE определяется в 
-диапазоне $>25$~кэВ, что позволяет учитывать продолжительные мягкие хвосты всплесков. 
-Более редкой ситуацией является существенное превышение длительностей по данным KW, 
-наблюдаемое в 10 всплесках. Шесть всплесков имеют плавное спадание/нарастание интенсивности, 
-которое не было учтено при вычислении  длительностей по данным BATSE, возможно, 
-из-за недостаточно точной аппроксимации фона. В трех всплесках по данным KW
-в интервал $T_{90}$ попадает слабый импульс, отделенный от основного пика. 
-При этом эклиптические широты основного пика и слабого импульса согласуются, 
-что позволяет отнести их к одному источнику. Во всплеске GRB~950904, 
-$T_{0\rmn{,BATSE}} = 52777$~s~UT ($T_{90\rmn{,KW}} = 99 \pm 4$~с, $T_{90\rmn{,BATSE}} = 11.1\pm 1.8$~с) 
-длительности BATSE даны для второго импульса, так как в~\citep{Hurley_2005} полагается, 
-что два импульса являются различными всплесками.
-
-Среди общих всплесков KW и BATSE 46 имеют $T_{50\rmn{KW}} \leq 0.6$~с, 
-из них 9 входят в набор 105 всплесков с искаженной длительностью (см.~Раздел~\ref{sec:GRB_sample}). 
-Из оставшихся 37, 8 имеют $T_{90\rmn{BATSE}} > 2$~с, 
-из них один является короткими всплеском с продлённым излучением. Таким образом, 
-засорение нашего списка длинными всплесками, связанное с низким S/N всплесков KW
-по сравнению с BATSE, составляет 19\% (=7/37). % без учета всплесков с продлённым излучением. 
-Ранее в работе~\citep{Ofek_2007ApJ} была получена доля 
-<<засорения>> коротких всплесков KW длинными $\approx 60$\%. В этой работе
-длительность всплесков в данных KW оценивалась визуально, что могло 
-стать причиной получения высокой доли засорения.
-
-\begin{figure}[h]
-  \begin{minipage}[h]{0.5\textwidth}
-    \center{\includegraphics[width=1.0\textwidth]{gT90_KWvsBATSE_ru} \\ а)}
-  \end{minipage}
-  \hfill
-  \begin{minipage}[h]{0.5\textwidth}
-    \center{\includegraphics[width=1.0\textwidth]{gT50_KWvsBATSE_ru} \\ б)}
-  \end{minipage}
-  \caption{Соотношение длительностей $T_{90}$~(a) и $T_{50}$~(б), определённых по 
-  данным BATSE и KW для 267 гамма-всплесков.}
-  \label{img:T90andT50_KWvsBATSE}  
-\end{figure}
-
-\subsection{Набор коротких всплесков}
-Набор из 1834 всплесков KW содержит 277 событий с $T_{50} \leq 0.6$~с.    %<-- число 277 проверено
-Среди них было обнаружено: пять очевидно длинных всплесков, в которых $T_{50}$ 
-% (ID = 1122, 2196, 2922, 2184) и ID=1228 $T_{90} = 8$~с в данных SAX
-определялось наиболее сильным импульсом, при этом полная длительность всплеска была 
-значительно больше 2~с; и 12 кандидатов в короткие всплески с продлённым излучением.
-Визуальный анализ показал, что среди 69 всплесков со сбоями пять имеют $T_{50} \leq 0.6$~с, 
-эти всплески были включены в итоговый набор. 
-% ID = 243, 704, 882 (солн. вспышка в данных), 934, 3319 (пропущен ранее)
-Таким образом, набор коротких всплесков KW содержит 265 событий без продлённого излучения.
-
-Кроме того, было обнаружено 19 кандидатов во всплески с продлённым излучением среди всплесков с $T_{50} > 0.6$~с. 
-К этому типу всплесков были отнесены события, имеющие короткий начальный импульс с $T_{50} \leq 0.6$~с, 
-за которым следует эпизод излучения, не содержащий импульсов с заметной спектральной эволюцией. 
-В некоторых случаях начальный пик и продлённое излучение были разделены интервалом, 
-на котором интенсивность излучения мала. Таким образом, полученный набор коротких всплесков с 
-продлённым излучением насчитывает 31 событие.
-
-Объединённый набор коротких всплесков с учётом всплесков с продлённым излучением 
-содержит 296 событий.
-\FloatBarrier
-
-\section{Жесткости}\label{sec:Hardness}
-Для анализа спектральных различий всплесков использовалось две величины: интегральная 
-жёсткость ($\rmn{HR}_{32}$)~--- отношение числа отсчетов, накопленных в каналах G3 и G2 
-за интервал $T_{100}$ и пиковая жёсткость ($\rmn{HR}_{32\rmn{pk}}$)~--- отношение числа отсчетов, 
-накопленных за интервал 64~мс с пиковой скоростью счета. 
-Отношение числа отсчётов в каналах G3 и G2 было выбрано, так как 
-оно наименее чувствительно к углу падения на детектор и более чувствительно 
-к пиковой энергии $E_\rmn{p}$.
-
-При вычислении жесткости $\rmn{HR}_{32}$ был учтён временн\'{о}й дрейф границ диапазонов. 
-Для этого трёхканальный спектр отсчётов аппроксимировался степенной функцией 
-с экспоненциальным завалом $dN/dE \propto E^{\alpha} \exp(-E/E_0)$. 
-С использованием полученной аппроксимации, вычислялось количество отсчётов в номинальных 
-границах диапазонов: 13--50~кэВ~(G1), 50--200~кэВ~(G2) и 200--750~кэВ~(G3). 
-Реальные границы диапазонов были вычислены с использованием калибровки, полученной 
-по многоканальному спектру (см. Главу~\ref{KW_description}).
-
-При рассмотрении интегральной жесткости использовался однородный набор из 1168 
-всплесков с $\rmn{S/N} \geq 10$. Набор содержит восемь всплесков со сбоями в G1, 
-для которых невозможно провести аппроксимацию трёхканального спектра. 
-Из оставшихся всплесков, 17 имеют значительную ошибку $\rmn{HR}_{32}$ 
-($\sigma\rmn{HR}_{32}/\rmn{HR}_{32} >0.3$), эти всплески были исключены из рассмотрения. 
-В итоге для исследования соотношения жёсткость-длительность использовалось 1143 всплеска.
-
-Распределение этого набора всплесков по $\rmn{HR}_{32}$ хорошо описывается суммой двух 
-лог-нормальных распределений. Добавление третьей компоненты не даёт существенного 
-улучшения аппроксимации.
-
-Аналогично работе~\citep{Horvath_2006} распределения всплесков на 
-плоскости $\log T_{50}$--$\log \rmn{HR}_{32}$ аппроксимировалось суммой двух и трёх 
-гауссовых компонент, при помощи метода наибольшего правдоподобия. Каждая компонента имеет вид: 
-\begin{equation}
-p(x,y| l) = \frac{1}{2\pi \sigma_x \sigma_y \sqrt{1-r^2}} \times 
-\exp\left[ -\frac{1}{2(1-r^2)}\left( \frac{(x-a_x)^2}{\sigma_x^2} + 
-\frac{(y-a_y)^2}{\sigma_y^2} -\frac{C}{\sigma_x \sigma_y}\right)\right]\mbox{ ,}
-\end{equation}
-где
-\begin{equation}
-C = 2r(x-a_x)(y-a_y)\mbox{ ,} \nonumber
-\end{equation}
-где $x=\log T_{50}$ и $y=\log \mbox{HR}_{32}$,  $a_x$, $a_y$~--- средние, 
-$\sigma_x$, $\sigma_y$~--- дисперсии, $r$~--- коэффициент корреляции, 
-и $l$~--- номер компоненты. При этом функция правдоподобия определяется следующим образом:
-\begin{equation}
-L = \sum_i \ln p(x_i, y_i)\mbox{ ,}
-\end{equation}
-где
-\begin{equation}
-p(x,y) = \sum_l  p(x, y|l)p_l \nonumber
-\end{equation}
-Аппроксимация производилась с помощью алгоритма 
-expectation maximization~\citep{Horvath_2006, Balazs_2003AA} реализованного в пакете 
-Scikit-learn~\citep{scikit-learn}. 
-Ошибки параметров вычислялись таким же методом, как и для распределения по длительностям. 
-На каждой из 1000 итераций генерировалось распределение на плоскости  
-$\log T_{50}$--$\log\rmn{HR}_{32}$ на основе ошибок $T_{50}$ и $\rmn{HR}_{32}$. 
-Затем из полученного массива параметров в качестве центрального значения и ширины доверительного 
-интервала бралась медиана и ширина на уровне 68\%. Значения параметров гауссовых компонент 
-для случая двух и трёх компонент представлены в таблицах~\ref{tab:clusters2_HR} 
-и~\ref{tab:clusters3_HR}, расположение компонент представлено на рис.~\ref{img:HRvsT50}.
-
-Различие значений функций правдоподобия для двух компонент $L_2 = -1243$ и трёх 
-компонент $L_3 = -1219$ ($\Delta L = 24$). Как показано в~\citep{Horvath_2006} 
-в этом случае $2\Delta L$ распределено как $\chi^2_6$ и вероятность случайного 
-отклонения $2\Delta L = 48$ равна $10^{-8}$. Функция правдоподобия для четырёх 
-компонент $L_4 = -1211$, $2(L_4 - L_3) = 16$ вероятность получения этой величины случайным образом равна $0.01$. 
-Это свидетельствует о наличии не более трёх классов всплесков. Третий класс всплесков 
-значительно перекрывается с классом длинных/мягких всплесков (см. Рис.~\ref{img:HRvsT50}), поэтому нельзя утверждать, 
-что он соответствует реальному физически выделенному классу всплесков, а не вызван эффектами селекции. 
-Также, учитывая, что распределения по $T_{50}$ и $\rmn{HR}_{32}$ сами по себе описываются 
-только двумя компонентами, для дальнейшего описания набора всплесков KW была выбрана
-сумма только двух компонент, которые далее именуются короткие/жесткие и длинные/мягкие всплески.
-
-Принадлежность всплеска к классу с номером $l$ определяется на основании значения индикаторной функции
-\begin{equation}
-I_l =\frac{p_l p(x,y|l)}{\sum_l  p(x, y|l)}
-\end{equation}
-Используя индикаторную функцию кластера коротких/жестких всплесков $I$, 
-всплески были классифицированы следующим образом: 
-$I \geq 0.9$~--- Тип~I, $0.1 \leq I < 0.9$~--- Тип~I/II (неопределённый тип), 
-$I < 0.1$~--- Тип~II. Названия типов выбраны по аналогии с физической классификацией. 
-Результаты классификации представлены на Рис~\ref{img:HRvsT50Types}.
-
-В наборе 1143 всплесков доли всплесков разных типов составляют: Тип~I~--- 18\%,
-Тип~I/II~--- 4\% и Тип~II~--- 78\%. Для всех всплесков типа~I длительность согласуется 
-с ограничением $T_{50} \leq 0.6$~с. Доля всплесков типа~II среди коротких всплесков 
-составляет 7\% (19\% если всплески типа~I/II относятся к типу~II), что согласуется с долей
-<<засорения>> коротких всплесков длинными, вычисленной только на основе распределения 
-по длительностям. Для временных историй коротких всплесков типа~II характерно 
-наличие одного импульса с плавным нарастанием и спадом.
-Из 31-го кандидата в короткие всплески с EE начальные импульсы 21-го события 
-имеют Тип~I (далее Iee) и 10 имеют неопределённый тип (далее Iee/II).
-При этом начальные импульсы двух событий, отнесённых к всплескам с EE, имеют тип~II.   % ID = 554, 2225
-
-Также был произведён анализ распределения длительность-пиковая жёсткость. 
-Для этого использовались 1104 GRBs с $\rmn{S/N} \geq 10$ и $\sigma \rmn{HR}_{32\rmn{pk}}/\rmn{HR}_{32\rmn{pk}} <0.3$. 
-Результаты аппроксимации двумя и тремя гауссовыми компонентами приведены на рис.~\ref{img:HRpkvsT50}. 
-Параметры распределений приведены в таблице~\ref{tab:tab_2clusters_HRpk} и~\ref{tab:tab_3clusters_HRpk}. 
-Различие в жесткостях длинных и коротких всплесков составляет примерно 1.7 раза для пиковых 
-жесткостей и 2.4 раза для интегральных.
-
-\input{tables/cluster_tables_diss}
-
-\begin{figure}[h]
-  \begin{minipage}[h]{0.5\textwidth}
-    \center{\includegraphics[width=1.0\textwidth]{list1143calHR_2} \\ а)}
-  \end{minipage}
-  \hfill
-  \begin{minipage}[h]{0.5\textwidth}
-    \center{\includegraphics[width=1.0\textwidth]{list1143calHR_3} \\ б)}
-  \end{minipage}
-  \caption[Аппроксимация распределения $\log T_{50}$--$\log \mbox{HR}_{32}$.]
-  {Аппроксимация распределения $\log T_{50}$--$\log \mbox{HR}_{32}$ 
-  для 1143 ярких ($\rmn{S/N} \geq 10$) всплесков двумя~(a) и тремя~(б) гауссовыми компонентами. 
-  Эллипсами для каждого распределения отмечены области $1\sigma$ и~$3\sigma$. 
-  Пунктирная вертикальная линия~--- $T_{50} = 0.6$~с. Кресты~--- короткие/жесткие всплески, 
-  круги~--- длинные/мягкие всплески, треугольники~--- промежуточные всплески. 
-  Изображенное разделение сделано на основе $I_l$, но отличается от указанного в тексте для большей наглядности.
-  Всплеск отнесён к кластеру $l$ если значение $I_l$ для этого кластера, 
-  превышает значения для других кластеров.
-  \label{img:HRvsT50}}
-\end{figure}
-
-\begin{figure} [h] 
-  \center
-  \includegraphics [width=0.75\textwidth] {gHRT50_Types_ru}
-  \caption{Классификация 1143 ярких всплесков KW.
-  Черные точки~--- Тип~I, круги~--- Тип~I/II, треугольники~--- Тип~II,
-  см. описание методики классификации в тексте.
-  } 
-  \label{img:HRvsT50Types}  
-\end{figure}
-
-\begin{figure}[h]
-  \begin{minipage}[h]{0.5\textwidth}
-    \center{\includegraphics[width=1.0\textwidth]{list1104calHRpcr_2} \\ а)}
-  \end{minipage}
-  \hfill
-  \begin{minipage}[h]{0.5\textwidth}
-    \center{\includegraphics[width=1.0\textwidth]{list1104calHRpcr_3} \\ б)}
-  \end{minipage}
-  \caption[Аппроксимация распределения $\log T_{50}\textrm{--}\log \mbox{HR}_{32\rmn{pk}}$.]
-  {Аппроксимация распределения $\log T_{50}\textrm{--}\log \mbox{HR}_{32\rmn{pk}}$ 
-  для 1104 ярких всплесков ($\rmn{S/N} \geq 10$) двумя~(a) и тремя~(б) гауссовыми компонентами. 
-  Эллипсами для каждого распределения отмечены области $1\sigma$ и~$3\sigma$. 
-  Пунктирная вертикальная линия~--- $T_{50} = 0.6$~с. Кресты~--- короткие/жесткие всплески, 
-  круги~--- длинные/мягкие всплески, треугольники~--- промежуточные всплески.
-  Изображенное разделение сделано на основе $I_l$, но отличается от указанного в тексте для большей наглядности.
-  Всплеск отнесён к кластеру $l$ если значение $I_l$ для этого кластера, 
-  превышает значения для других кластеров.
-  }
-  \label{img:HRpkvsT50}  
-\end{figure}
-
-\FloatBarrier
-
-\section{Спектральные задержки}\label{sec:Lags}
-Одной из характеристик спектральной эволюции гамма-всплесков является спектральная задержка.
-Она представляет собой количественную характеристику запаздывания излучения в 
-мягком спектральном диапазоне по сравнению с более жестким~\citep[см., например,][Глава~3]{Minaev_PhD}.
-Обзор моделей, описывающих появление спектральной задержки, смотри там же.
-В данном разделе спектральная задержка используется исключительно в роли дополнительного 
-параметра классификации гамма-всплесков, без обсуждения физической причины её возникновения.
-
-\subsection{Методика вычисления спектральных задержек для кривых блеска KW}
-Для вычисления спектральных задержек всплесков KW использовался 
-кросскорреляционный анализ временных историй~\citep{Band_1997ApJ, Norris_2000}. 
-В этом методе задержка ($\tau_\rmn{lag}$) соответствует положению максимума 
-кросс-корреляционной функции (ККФ) временных историй в различных каналах детектора.
-Определение спектральной задержки между временными историями в двух диапазонах ($\tau_\rmn{lag}$) 
-включало выбор разрешения временных историй, выбор интервала кросс-корреляции 
-и вычисление $\tau_\rmn{lag}$ и её ошибки.
-
-Временное разрешение выбралось таким образом, чтобы для временных историй 
-в обоих диапазонах бин с наибольшей скоростью счета имел отношение сигнал-шум~$\rmn{S/N} \geq 8$. 
-Порог $\rmn{S/N} \geq 8$ был выбран произвольно, чтобы исключить из рассмотрения 
-слабые всплески с большими ошибками $\tau_\rmn{lag}$. Возможные значения разрешения 
-временной истории составляли от 4~мс до 1024~мс.
-В качестве интервала кросс-корреляции брался наиболее широкий интервал, ограниченный
-бинами, в которых было обнаружено превышение $5\sigma$ над фоном хотя бы в одном из каналов.
-
-Значение $\tau_\rmn{lag}$ и его ошибка определялись методом Монте-Карло на основе 
-100 реализаций исходных кривых блеска. 
-На каждой итерации генерировались искусственные временн\'{ы}е истории 
-при помощи добавления пуассоновского шума к исходным данным.
-На каждой итерации значения ККФ вычислялось аналогично формуле~(10) из~\citep{Band_1997ApJ}, 
-ошибки ККФ вычислялись аналогично формуле~(5) из~\citep{Fenimore_1995}.
-После построения ККФ производился поиск интервала, содержащего основной пик ККФ
-при этом использовались только значения $>0.1$.  
-После чего ККФ аппроксимировалась полиномом 4-й степени на выбранном интервале
-и в качестве значения $\tau_\rmn{lag}$ брался максимум полинома.
-Если $p$-значение аппроксимации (\textit{null hypothesis probability}) 
-оказывалось меньше порогового значения 1\%, 
-то из ККФ удалялись две крайние точки, и процедура повторялась до превышения порога. 
-Если в результате уменьшения интервала аппроксимации число степеней свободы сокращалось до нуля,
-то текущая итерация считалась неудачной. 
-В случае, если итоговое число удачных итераций было больше $50$, то значение $\tau_\rmn{lag}$ 
-вычислялось как среднее по всем удачным итерация, в качестве ошибки бралась дисперсия значений.
-Иначе значение $\tau_\rmn{lag}$ считалось определенным ненадёжно и дальше не использовалось.
-Примеры вычисления лага на одной из итераций представлены рис.~\ref{fig:lag_calculation}.
-
-\begin{figure}[h]
-  \center{\includegraphics[width=0.75\textwidth]{GRB20070512_T21972_lag_31}}
-  \caption{Иллюстрация вычисления лага между кривыми блеска в G3 и G1 
-   для короткого всплеска GRB20070512\_T21972 со значительной задержкой, $\tau_\rmn{lag}=277\pm16$~мс. 
-  Вертикальные линии на нижней панели обозначают границы интервала кросс-корреляции. 
-  Непрерывная линия на верхней панели показывает аппроксимацию ККФ полиномом 4-й степени. 
-  Вертикальная пунктирная линия на верхней панели соответствует значению лага.
-  \label{fig:lag_calculation}}  
-\end{figure}
-
-\subsection{Спектральные задержки коротких всплесков}
-Спектральные задержки вычислялись между парами каналов G3--G1, G2--G1 и G3--G2, 
-где $\tau_\rmn{lag}$ соответствует сдвигу временной истории в канале, указанном вторым
-по отношению к каналу, указанному первым. Для коротких всплесков было получено
-42, 66 и 158 задержек для пар каналов G3--G1, G2--G1 и G3--G2, соответственно.
-
-На рис.~\ref{img:LagDistrs} представлены распределения коротких всплесков без EE и 
-начальных импульсов коротких всплесков с EE по спектральным задержкам. 
-Всплески Типа~I имеют, в основном, незначительные $\tau_\rmn{lag}\lesssim 25$~мс для 
-пар каналов G3--G1 и G2--G1, при этом значительная часть всплесков типов~I/II и~II 
-имеют задержки $\tau_\rmn{lag}\gtrsim 25$~мс для тех же пар каналов.
-
-Среди коротких всплесков с продлённым излучением два события имеют существенные  % (ID=554, 1531)
-спектральные задержки начального импульса в G3--G1 и были классифицированы Тип~Iee/II в предыдущем разделе. 
-Таким образом, можно заключить, что значительная спектральная задержка подкрепляет классификацию 
-на основе распределения на плоскости жёсткость-длительность.
-
-\begin{figure}[h]
-  \begin{minipage}[h]{0.5\textwidth}
-    \center{\includegraphics[width=1.0\textwidth]{gDistLags} \\ а)}
-  \end{minipage}
-  \hfill
-  \begin{minipage}[h]{0.5\textwidth}
-    \center{\includegraphics[width=1.0\textwidth]{gDistLagsEE} \\ б)}
-  \end{minipage}
-  \caption{Распределение коротких всплесков без EE~(а) и начальных импульсов 
-  всплесков с EE~(б) по спектральной задержке.
-  На панели~(а), заполненная гистограмма~--- всплески Типа~I, заштрихованные 
-  гистограммы~--- всплески Типа~I/II и~II.  На панели~(б), заполненная 
-  гистограмма~--- всплески Типа~I, заштрихованные гистограммы~--- всплески Типа~Iee и~Iee/II.}
-  \label{img:LagDistrs}  
-\end{figure}
-
-\FloatBarrier
-
-\section{Сравнение классификаций на физические типы I и II всплесков KW 
-с определенными красными смещениями}\label{sec:Phys_Classification}
-Набор всплесков KW, зарегистрированных с ноября 1994~г. по июнь 2014~г., 
-содержит 126 гамма-всплесков с измеренными красными смещениями~\citep{Tsvetkova_KW_GRBs_with_z}. 
-Из них пять имеют сбои во временной истории, далее рассматривается набор из 121 всплесков. 
-Указанный набор содержит 11 коротких и 110 длинных всплесков.
-
-Короткие всплески KW имеют красные смещения в диапазоне 0.1--1.0, длинные~--- 0.1--5.0.
-Для $\sim 50$\% длинных и практически всех коротких всплесков из набора были 
-определены физические типы на основе данных о послесвечении и родительских 
-галактиках~\citep{Zhang_2009, Kann_2010ApJ, Kann_2011ApJ}.
-При этом стоит выделить два всплеска. 
-Длинный всплеск GRB~060614 относят к коротким всплескам с продлённым излучением~\citep{Gehrels_2006_Nature}. 
-Однако, по данным KW его нельзя отнести к этому классу, так как на основе полной 
-длительности $T_{50} = 49 \pm 5$~с, длительности его начального импульса 
-$T_{50} = 2.7\pm 0.3$~с и жесткости этот всплеск относится к типу~II.
-Всплеск GRB~040924 можно отнести к коротким всплескам на основе длительности $T_{50} = 0.34\pm0.06$~c, 
-при этом он имеет значительную спектральную задержку, $120 \pm 30$~мс между каналами G2 и G1, 
-и является достаточно мягким. По данным KW этот всплеск классифицируется как Тип~II, 
-так же как и по данным многоволновых наблюдений~\citep{Zhang_2009}.
-
-Было также проанализировано распределение жёсткость-длительность в системе 
-отсчёта источника всплеска. При этом $T_{50}$ уменьшается в $1+z$~раз, а жёсткость 
-вычислялась в каналах, чьи границы соответствуют номинальным умноженным на $1/(1+z)$. 
-Распределения на плоскости $\log T_{50}$--$\log \mbox{HR}_{32}$ в системе отсчёта наблюдателя 
-и собственной системе отсчёта представлены на рис.~\ref{img:T50HRzCorr}. 
-Для сравнения распределений по жесткости и длительности всплесков типов I и II
-в системе наблюдателя и собственной системе был использован тест Колмогорова-Смирнова для двух наборов. 
-Вероятности (p-значения) того, что жесткости или длительности 
-121-го всплеска являются выборкой из общего распределения малы ($<1$\%) 
-как в случае системы наблюдателя, так и источника. Хотя в последнем случае
-граница между типами становится размытой, рассмотренные распределения существенно отличаются,
-что свидетельствует различие параметров коротких/жестких и длинных/мягких всплесков как 
-в системе наблюдателя, так и системе источника всплеска.
-
-\begin{figure}[h]
-  \begin{minipage}[h]{0.5\textwidth}
-    \center{\includegraphics[width=1.0\textwidth]{gHRvsT50obs_ru} \\ а)}
-  \end{minipage}
-  \hfill
-  \begin{minipage}[h]{0.5\textwidth}
-    \center{\includegraphics[width=1.0\textwidth]{gHRvsT50src_ru} \\ б)}
-  \end{minipage}
-  \caption{Распределение всплесков на плоскости $\log T_{50}$--$\log \rmn{HR}_{32}$ 
-  в системе отсчёта наблюдателя~(а) и источника~(б), с поправкой на космологическое красное смещение. 
-  Круги~--- всплески типа~I, треугольники~--- всплески типа~II, 
-  серые точки~--- набор 1143 ярких всплесков.
-  }
-  \label{img:T50HRzCorr}  
-\end{figure}
-
-\section{Заключение} \label{sec:Conclision}
-В данной главе описана методика классификации всплесков, зарегистрированных в эксперименте
-Конус-Винд и получены следующие результаты:
-
+Получены следующие результаты:
 \begin{enumerate}
-\item Для набора 1834 всплесков KW были вычислены длительности $T_{50}$ и $T_{90}$, жесткости 
-и спектральные задержки. Показано, что распределения 
-всплесков по $T_{50}$ и $T_{90}$ хорошо аппроксимируются двумя логнормальными 
-распределениями. Обнаружено, что параметры аппроксимации распределения $T_{50}$ 
-более устойчивы к выбору порога поиска начала и конца всплеска, поэтому длительность 
-$T_{50}$ более предпочтительна для классификации всплесков. В качестве границы между 
-длинными и короткими всплесками была выбрана точка пересечения логнормальных компонент 
-для порога значимости $5\sigma$, $T_{50} = 0.6$~с.
-
-\item Выбран набор коротких всплесков (без продлённого излучения), включающий 265 событий, 
-что составляет 14\% от общего числа проанализированных всплесков. Для сравнения, 
-доля коротких ($T_{90}<2$~с) всплесков в каталоге BATSE равна 24\%~\citep{Meegan_2001}, %(=500/2014)
-в каталоге \textit{Fermi}-GBM 18\%~\citep{Paciesas_2012}, 
-в каталоге \textit{Swift}-BAT 8\%~\citep{Sakamoto_2011ApJS}. 
-Доля коротких всплесков, оцененная из площадей гауссовых компонент 
-распределения $T_{50}$, равна 30\%. Соответствующая доля в наборе всплесков BATSE равна 32\%~\citep{Horvath_2002} 
-и 8\% в наборе всплесков \textit{Swift}-BAT~\citep{Horvath_2008}.
-
-\item Обнаружен 31 всплеск, который можно классифицировать как короткий 
-всплеск с продлённым излучением. Таким образом, доля всплесков с продлённым 
-излучением среди коротких всплесков составляет 10\%. Соответствующая доля 
-в выборке \textit{Swift}-BAT равна 20\%~\citep{Sakamoto_2011ApJS}. 
-Спектральный анализ продлённого излучения представлен в главе~\ref{sGRB_spectral_catalog}.
-
-\item Аппроксимация распределения всплесков на плоскости $\log T_{50}$--$\log \rmn{HR}_{32}$ 
-набором гауссовых компонент показала наличие двух классов всплесков, коротких/жестких 
-и длинных/мягких. При этом доля коротких/жестких всплесков, определённая на основе 
-аппроксимации, составляет 21\%. По данным  BATSE доля коротких/жестких всплесков 
-составляет 28\%~\citep{Horvath_2002}.  
-Добавление третьей компоненты даёт значимое улучшения аппроксимации, однако эта 
-компонента существенно перекрывается с компонентой, описывающей длинные всплески, 
-и не представляет физического смысла. Дополнительный довод в пользу использования
-только двух классов всплесков связан с тем, что для описания распределений 
-по $T_{50}$ и $\rmn{HR}_{32}$ достаточно только двух компонент.
-На основании полученной аппроксимации 7\% всплесков с $T_{50} < 0.6$~с относятся к типу~II.  (0.07=18/260)
-При этом к типу~I относятся только всплески с $T_{50} < 0.6$~с 
-(исключение составляет жесткий всплеск GRB20061006\_T31422 с пограничной  длительностью $T_{50} = 0.620\pm 0.049$~с, 
-который не входит в набор коротких всплесков). Неопределенную классификацию (I/II) имеют 
-около 4\% всплесков с длительностями $T_{50}$ от 0.04~с до 1.65~с.
-
-На основании аппроксимации распределения $\log T_{50}$--$\log \rmn{HR}_{32}$ 
-для 1143 всплесков, был классифицирован набор 265 коротких всплесков без продлённого излучения. 
-Набор включает $\sim 70$\% всплесков Типа~I, $\sim 8$\% Типа~II и $\sim 12$\%
-всплесков неопределённого типа (I или~II). Доля коротких всплесков с продлённым 
-излучением составляет $\sim 10$\%.
-Среди начальных импульсов всплесков, отнесённых на основе морфологии временной 
-истории к коротким всплескам с продлённым излучением, 21 (68\%) классифицированы как Тип~I 
-7 как неопределённый тип (I/II) и 3 как Тип~II.
-
-\item Анализ спектральных задержек коротких всплесков показал, что большинство коротких всплесков 
-Типа~I имеют незначительную по абсолютному значению спектральную задержку $\tau_\rmn{lag} \lesssim 25$~мс, 
-в то же время значительная доля всплески типов~II и I/II имеет $\tau_\rmn{lag} > 25$~мс. 
-Два всплеска с продленным излучением имеют задержку начального импульса $> 100$~мс
-и были классифицированы как тип~II, что свидетельствует о том, что
-эти всплески относятся к популяции длинных всплесков.
-
-\item Сравнение классификаций на физические типы~I и~II с классификацией на основе 
-длительности, жесткости и спектральной задержки подтвердило, что всплески Типа~I 
-относятся к коротким/жестким всплескам с малой спектральной задержкой, а всплески 
-Типа~II, в основном,~--- длинные мягкие с заметной спектральной задержкой. 
-Сравнение распределений $\log T_{50}$--$\log \rmn{HR}_{32}$ в системе отсчёта наблюдателя 
-и в собственной системе отсчёта показывает, что различие в жесткости и длительности
-всплесков типа~I и~II становится менее значимым, но сохраняется.
+	\item 
+	\item 
 \end{enumerate}
 
-Для дальнейшего анализа выбран полный набор 296 коротких всплесков. 
-Различия между короткими всплесками разных типов будут подробно исследованы на
-основе спектрального анализа всплесков в главе~\ref{sGRB_spectral_catalog}.
-
-По материалам Главы~\ref{KW_GRB_classification} на защиту выносится следующее положение:
-\begin{itemize}
-\item Метод классификации гамма-всплесков по данным эксперимента Конус-Винд на основе
-    длительности и жесткости излучения всплеска, а также величин спектральных задержек.
-\end{itemize}
-
 \clearpage

### 2025-12-02 disser/part2.tex
disser

@@
 
-There is no direct evidence of the termination of the narrow jets in the radio, optical and X-ray maps of W50 \citep[][]{1998AJ....116.1842D,2011MNRAS.414.2838G,2017MNRAS.467.4777F,2018MNRAS.475.5360B}, while a
-broader and more extended X-ray structures are seen on both sides of SS~433  aligned with the narrow optical jet precession axis \cite[e.g.,][]{1983ApJ...273..688W,1994PASJ...46L.109Y,Safi-Harb1997,1999ApJ...512..784S,Brinkmann2007,Safi-Harb2022}.
-These extended X-ray jets appear at a distance of $\sim 20\,$pc from the compact source and have an opening angle of $\sim 20^\circ$ which is twice smaller than the precession amplitude of the narrow jets.
-The X-ray structures are probably of synchrotron origin \citep[e.g.,][]{Brinkmann2007}, which is confirmed by the detection of X-ray polarization by the IXPE telescope \citep{IXPE2024SS433}.
-The very high energy (VHE) gamma-ray emission from the source was predicted in a number of models  \citep[see, e.g.,][]{Safi-Harb1997, Reynoso2008, 1998NewAR..42..579A,
-	2020ApJ...904..188K, 2020ApJ...889..146S}.
-Recently, the extended very high energy gamma-ray emission was detected from W50 \citep[][]{2018Natur.562...82A,HESS2024SS433,LHAASO2024SS433}, indicating the acceleration of PeV-regime particles in the source.
-The physical origin of these extended structures and their connection to the narrow transrelativistic jets currently launched by the system remains unclear \citep[e.g.
-][ and references therein]{2020MNRAS.495L..51C}
-
-The `minimalist' model \citep{Churazov} explains the extended jet as a result of the powerful two-component anisotropic outflow produced by a supercritical accretion disk.
-In this model, a more collimated outflow (polar wind), aligned with the rotation axis of the binary, coexists with a more isotropic wind.
-The extended X-ray jets observed are associated with the recollimation shocks, which appear right after the collision of the collimated outflow with the isotropic wind termination surface.
-The structure of  MHD outflow in the interaction region of the collimated wind with the termination surface of the isotropic wind was simulated with the PLUTO MHD code \cite{Mignone2007}.
-We present a model of very high-energy particle acceleration in the extended jets of the SS~433/W50 system following the two-component outflow scenario.
-
-The results of MHD modeling are described in Section \ref{sec:morphological}.
-Using the spatial distributions of the accelerated particles and magnetic field spectra obtained in the model, we construct profiles and spectra of non-thermal X-ray and gamma radiation of this object.
-The modeling includes the following steps, split across several sections.
-
-Section \ref{sec:MC}.
-Based on the plane-parallel Monte Carlo model of particle acceleration and magnification of magnetic fields near the shock fronts, the spectra of accelerated particles and magnetic fields near the shock waves obtained in the MHD model are found.
-
-Section \ref{sec:CorSh}.
-Based on MHD modeling of the passage of turbulence with significant density fluctuations through the shock front, profiles of the longitudinal $B_{\parallel}$  and transverse $B_{\perp}$ components of the magnetic field behind the shock front are obtained using the PLUTO code.
-
-Section \ref{sec:Spectrum}.
-Next, the spectra and profiles of the X-ray synchrotron emission are calculated based on the spectra of accelerated electrons and magnetic field profiles.
-Gamma Rays in the TeV range are also calculated.
-The simulation parameters are selected to match the model with the observational data \citep[][]{Brinkmann2007, Safi-Harb2022, HESS2024SS433, LHAASO2024SS433}.
-The anisotropy of the magnetic field in the shock downstream leads to the observed polarization of the X-ray radiation.
-The X-ray polarization obtained in the model near the shock wave is compared with the observational data from the IXPE observatory \citep{IXPE2024SS433}.
-
-High energy particle acceleration in microquasars as a potential contributor to the galactic cosmic ray pool was widely discussed (see e.g.
-\cite{2002A&A...390..751H, 2005A&A...432..609B,2017SSRv..207....5R}), while recently the interest to SS 433 as a pevatron was raised in \cite{Churazov,2024arXiv241108762P,2025arXiv250620193Z,2025arXiv250622550C,2025arXiv250721048W} by VHE gamma-ray observations reported in \cite{HESS2024SS433,LHAASO2024SS433}.
-
-\begin{figure}
+\begin{figure}[h]
 	\centering
@@
 \end{figure}
+
+Нет наблюдаемых свидетельств остановки джетов и в то же время наблюдаются широкие протяженные рентгеновские структуры по обе стороны от SS~433, сонаправленные с осью прецессии джетов \cite[e.g.,][]{Watson1983,Yamauchi1994,Safi-Harb1997,Safi-Harb1999,Brinkmann2007,Safi-Harb2022}.
+Эти расширенные джеты появляются на расстояниях около $\sim 20\,$пк от центрального источника и имеют угол раствора$\sim 20^\circ$, что в два раза меньше чем угол прецессии узких джетов.
+Рентгеновское излучение в этих структурах имеет, вероятно, синхротронное происхождение \citep[e.g.,][]{Brinkmann2007}, что подтверждается измерянной рентгеновской поляризацией телескопом IXPE \citep{IXPE2024SS433}.
+Высокоэнергичной гамма-излучения от источника предсказывалось многими моделями \citep[see, e.g.,][]{Safi-Harb1997, Reynoso2008, Aharonian1998, Kimura2020, Sudoh2020}.
+Недавно оно было задетектировано телескопами HAWC, H.E.S.S.
+и LHAASO \citep[][]{Abeysekara2018,HESS2024SS433,LHAASO2024SS433}, что указывает на наличие частиц, ускоренных до энергий порядка петаэлектрон-вольт.
+
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=0.9\textwidth]{./img_part2/HESS.png} 
+	\caption{Изображение туманности W50 в гамма лучах по данным телескопа HESS \cite{HESS2024SS433}.
+Приведена значимость сигнала в диапазонах: панель A - 0.8-2.5 ТэВ, панель B - 2.5-10 ТэВ, панель C - выше 10 ТэВ}
+	\label{W50HESS}
+
+	\centering
+	\includegraphics[width=0.9\textwidth]{./img_part2/LHAASO.png} 
+	\caption{Изображение туманности W50 в гамма лучах по данным телескопа LHAASO \cite{LHAASO2024SS433}.
+Приведена значимость сигнала в диапазонах: панель a - 1-25 ТэВ, панель b - 25-100 ТэВ, панель c - выше 100 ТэВ.
+Голубым цветом отмечена область излучающая гамма лучи с энергиями выше 10 ТэВ по данным HESS.}
+	\label{W50LHASSOimage} 
+\end{figure}
+
+Модель, предложенная в работе \citep{Churazov}, объяснят структуру туманности W50 как результат взаимодействия мощного двухкомпонентного ветра, создаваемого аккреционным диском.
+Транс-релятивистский поток, сонаправленный с осью прецесси узких джетов взаимодействует с более медленным квазисферическим ветром.
+В результате этого взаимодействия в полярном ветре возникают реколлимационные ударные волны, приводящие к ускорению космических лучей и фысокоэнергичному излучению.
+В данной главе мы проведем моделирование структуры туманности W50 в рамках вышеописанной модели с помощью магнито-гидродинамического кода PLUTO \cite{Mignone2007}, моделирование ускорения космических лучей полученными ударными волнами, а так же проведем расчет высокоэнергичного излучений и сравнение с наблюдательными данными.
 

### 2025-12-03 disser/part2.tex
fz

@@
 \section{Моделирование структуры туманности}
+\begin{figure}
+	\includegraphics[width=0.9\textwidth]{./img_part2/Scheme.png}
+	\caption{Schematic representation of the SS433+W50 system adopted here.
+The structures with bright non-thermal X-ray emission in the extended jet of SS 433 which we model here are marked as  `Cone", `Head", `e1' and `e2".
+They were detected and discussed in the papers \cite{Safi-Harb1997,Brinkmann2007,Safi-Harb2022}.}
+	\label{Scheme}
+\end{figure}
+
+In this section, we perform axisymmetric MHD modeling of the W50 nebula around the microquasar SS 433.
+It has a characteristic shape: a central spherical part and two opposite lobes to the east and west of the central source.
+The X-ray emission of the nebula is very inhomogeneous, with several bright X-ray regions - `e1", `e2", and `e3' to the east and `w1' and `w2' to the west \cite{Safi-Harb1997}.
+Following \cite{Churazov}, we model a wind from the central microquasar with two components - quasi-isotropic slow wind and a more collimated outflow, the polar wind.
+The kinetic power of each component is about $10^{39}$ erg/s, and the velocities are $\approx3000$ km/s for the isotropic wind and $\approx0.2~c$ for the jet.
+Interactions of the collimated outflow, isotropic wind, and ambient medium cause several recollimation shocks in the jet, which explains the existence of different X-ray emitting regions.
+The sketch of the hydrodynamical model with the emission regions shown from \cite{Safi-Harb2022, IXPE2024SS433} is presented in Figure \ref{Scheme}.
+
+We perform numerical simulations of the W50 nebula using the MHD code PLUTO \cite{Derouillat}.
+We used a 2D setup in cylindrical coordinates.
+In the central region, at the sphere with radius of 5 pc, we set internal boundary conditions with fixed density, pressure, and radial velocity, depending on the polar angle $\theta$ of the current grid point.
+For small angles, the outflow parameters correspond to the trans-relativistic collimated outflow, and for the larger angles to the isotropic wind.
+The ambient medium has constant density and pressure.
+
+The magnetic field was set to 100 $\mu$G in the jet base and 1 $\mu$G in the ambient medium and isotropic wind, and directed along the z-axis.
+We managed to obtain an elongated structure with two strong recollimation shocks with noticeable Mach diamonds, located approximately at distances 32 and 48 pc from the central source, close to observed bright X-ray emitting regions, named `e1' and `e2' in the eastern jet \cite{Safi-Harb1997}.
+The structure of hydrodynamical flow is shown in Fig.~\ref{velocity} - large-scale velocity map in the top panel, and zoomed maps of velocity, density, and magnetic field in the region between the central source and `e2' in the lower panels.
+The parameters for this setup are: kinetic power of the isotropic wind $P_i$ is equal to kinetic power of the collimated outflow $P_j = 2.1\times10^{39}$ erg/s, the velocity of isotropic wind is $v_i=3000$ km/s, the velocity of collimated outflow is $v_j=0.2~c$, its half-width angle is $5^{\circ}$, the ambient number density is $\rho_{amb} = 0.005~\rm{cm^{-3}}$ and the temperature $T_{amb} = 8\times10^4~\rm{K}$.
+We assumed that the accretion disk outflows which produced the extended X-ray jet are propagating through a nebula of low plasma density created previously  by the supernova event and modified by the winds from the binary (see e.g.
+a model of the nebula in \citep{2021ApJ...910..149O}).
+
+The ambient matter parameters used above are not unique.
+It is possible to obtain a similar structure of the extended X-ray jets and nebulae with higher ambient density and lower temperature e.g.
+($\rho_{amb}=0.05~\rm{cm^{-3}}$, $T_{amb} = 8.6\times10^3~\rm{K}$).
+In this case, the structure  with strong internal shocks,  similar to that shown in Fig.
+\ref{MHD_profile} appears at later times of about 100, 000 years.
+We left the detailed study of the parameter space of the ambient matter allowed by multiwavelength observations of W50 for future study, and here we model the VHE particle acceleration and nonthermal emission of the extended X-ray jets using the main setup described above.
+
+The first recollimation shock occurs close to the termination shock of the isotropic wind.
+The later behavior of the outflow - will it propagate further with a series of recollimation shocks, or stop after the first one, is determined by the pressure of the ambient medium.
+It should be low enough for the collimated outflow to propagate a large distance.
+In addition, relatively low ambient pressure is necessary to form Mach shocks with plasma flow normal to the shock front, instead of a system of crossing oblique shocks, as shown in \cite{Marti2018}.
+In numerical simulations, spatial resolution should be high enough to resolve Mach diamonds.
+
+The one-dimensional  density, velocity, and entropy profiles of the hydrodynamic flow near the regions `e1' and `e2' are shown in Figure~\ref{MHD_profile}.
+The density jumps at the two shocks are close to 4, which is a sign of the strong hydrodynamic shocks.
+Entropy jumps are relatively large as well.
+Also, there is a weak shock with smaller jumps between two strong shocks.
+Kinetic Monte Carlo collisionless shock models, which take into account the energy fluxes of accelerated particles, showed that the compression at the shock front can significantly exceed the adiabatic value 4 (see e.g.
+simulation in \cite{Bykov3inst2014}).
+
+\begin{figure}
+	\includegraphics[width=0.9\textwidth]{./img_part2/velocity4.png}
+	\caption{Morphology of the W50 nebula in simulations: a) large-scale velocity profile in the entire computational domain, b) velocity near e1 and e2 regions, c) density near e1 and e2 regions, d) magnetic field near e1 and e2 regions.}
+	\label{velocity}
+\end{figure}
+
+\begin{figure}
+	\includegraphics[width=0.9\textwidth]{./img_part2/profile_1d_window2.png}
+	\caption{Density, velocity, and entropy profiles in the MHD run shown in Fig.~\ref{velocity}.} 
+	\label{MHD_profile}
+\end{figure}
+
+In the minimalists' scenario, the main goal was reproducing the overall morphology of the W50 radio emission and the emergence of the `extended X-ray jets' as a result of recollimation shocks in the anisotropic wind scenario.
+This model has several free parameters, including overall energetics, age of the system, density of the ambient medium, velocities and densities of the isotropic and polar winds, and the opening angle of the polar component.
+Various combinations of these parameters can lead to a qualitatively similar morphology of the nebula (see Fig.A1 in \cite{Churazov}).
+Here, our primary goal is to find parameters that lead to the efficient acceleration of the particles that give rise to the observed emission of EXJs.
+Knowing that accelerating electrons to $\gsim$ 100 TeV energies is especially efficient for strong shocks with velocities $\gtrsim$ 0.2c \cite{Churazov}, we ran the MHD version of a set-up with this value of the polar component velocity, but did not require the morphology of the simulated nebula boundary to match the observed one perfectly.
+We leave the task of finding a parameter combination that satisfies all observational requirements best for a future study.
 \section{Расчет высокоэнергичного излучения}

### 2025-12-05 disser/part2.tex
fz

@@
 
-In this section, we perform axisymmetric MHD modeling of the W50 nebula around the microquasar SS 433.
-It has a characteristic shape: a central spherical part and two opposite lobes to the east and west of the central source.
-The X-ray emission of the nebula is very inhomogeneous, with several bright X-ray regions - `e1", `e2", and `e3' to the east and `w1' and `w2' to the west \cite{Safi-Harb1997}.
-Following \cite{Churazov}, we model a wind from the central microquasar with two components - quasi-isotropic slow wind and a more collimated outflow, the polar wind.
-The kinetic power of each component is about $10^{39}$ erg/s, and the velocities are $\approx3000$ km/s for the isotropic wind and $\approx0.2~c$ for the jet.
-Interactions of the collimated outflow, isotropic wind, and ambient medium cause several recollimation shocks in the jet, which explains the existence of different X-ray emitting regions.
-The sketch of the hydrodynamical model with the emission regions shown from \cite{Safi-Harb2022, IXPE2024SS433} is presented in Figure \ref{Scheme}.
+Туманность W50 имеет специфическую форму - у нее есть сферическая центральная часть и две вытянутые структуры к востоку и западу от ентрального источника.
+Рентгеновское излучение туманности неоднородно, а происходит от нескольких ярких областей - `e1", `e2", и `e3' на восточной стороне, `w1' и `w2' на западной \cite{Safi-Harb1997}.
+Следуя работе \cite{Churazov}, мы инициализируем течение из микроквазара в виде двух компонент - квазисферический медленный ветер и направленный транс-релятивистский полярный ветер (джет).
+Кинетическая мощность каждой компоненты составляет $2\times10^{39}$ erg/s, скорость медленного ветра - $\approx3000$ km/s, скорость полярного ветра $\approx0.2~c$.
+В результате взаимодействия направленного ветра, изотроного ветра и межзвездной среды образуются реколлимационные ударные волны в направленном ветре, что позволяет объяснить наличие нескольких ярких рентгеновских областей.
+Схема гидродинамической модели с излучающими областями, использованными в работах  \cite{Safi-Harb2022, IXPE2024SS433} показана на рисунке \ref{Scheme}.
 

### 2025-12-07 disser/part2.tex
disser

--- 2025-12-02 disser/part2.tex
@@
 \section{Моделирование структуры туманности}
+\begin{figure}
+	\includegraphics[width=0.9\textwidth]{./img_part2/Scheme.pdf}
+	\caption{Схематичное изображение системы SS433+W50 system adopted here.
+Яркие рентгеновские области в расширенном джете обозначены как `Cone', `Head', `e1' и `e2'.
+Они были обнаружени и обсуждались в работах \cite{Safi-Harb1997,Brinkmann2007,Safi-Harb2022}.}
+	\label{Scheme}
+\end{figure}
+
+Туманность W50 имеет специфическую форму - у нее есть сферическая центральная часть и две вытянутые структуры к востоку и западу от ентрального источника.
+Рентгеновское излучение туманности неоднородно, а происходит от нескольких ярких областей - `e1", `e2", и `e3' на восточной стороне, `w1' и `w2' на западной \cite{Safi-Harb1997}.
+Следуя работе \cite{Churazov}, мы инициализируем течение из микроквазара в виде двух компонент - квазисферический медленный ветер и направленный транс-релятивистский полярный ветер (расширенный джет).
+Кинетическая мощность каждой компоненты составляет $2\times10^{39}$ erg/s, скорость медленного ветра - $\approx3000$ km/s, скорость полярного ветра $\approx0.2~c$.
+В результате взаимодействия направленного ветра, изотроного ветра и межзвездной среды образуются реколлимационные ударные волны в направленном ветре, что позволяет объяснить наличие нескольких ярких рентгеновских областей.
+Схема гидродинамической модели с излучающими областями, использованными в работах  \cite{Safi-Harb2022, IXPE2024SS433} показана на рисунке \ref{Scheme}.
+
+Мы провели численное моделирование туманности W50 используя магнито-гидродинамический код PLUTO \cite{Mignone2007}.
+Мы использовали двумерные расчеты в цилиндрических координатах.
+В центральной области - сфере радиуса 5 пк, задавались внутренние граничные условия с заданными плотностью, давлением и скоростью радиального течения, которые зависели от полярного угла $\theta$ данной пространственной ячейки.
+Для малых углов параметры соответствовали транс-релятивистскому направленному потоку, а для больших - изотропному ветру.
+Внешняя межзвездная среда задавалась с постоянными плотностью и давлением.
+
+Магнитное поле задавалось равным 100 $\mu$G в основании расширенного джета и 1 $\mu$G в окружающей среде и изотропном ветре, и направлено по оси z.
+В результате моделирования удалось получить протяженную структуру со сферической центральной частью и двумя сильными реколлимационными ударными волнами с заметными дисками Маха, расположенными на расстоянии примерно 32 и 48 парсек от центрального источника, что близко к наблюдаемым ярким рентгеновским областям `e1' и `e2' в восточном джете \cite{Safi-Harb1997}.
+Структура гидродинамического течения изображена на рисунке ~\ref{velocity} - крупномасштабная карта скорости на верхней панели и увеличенные карты скорости, плотности и магнитного поля в области между центральным источником и `e2' на нижних панелях.
+Для данного сетапа были выбраны следующие параметры: кинетическая можность изотропного ветра $P_i$ и равная ей кинетическая мощность расширенного джета $P_j = 2.1\times10^{39}$ эрг/с, скорость изотропного ветра  $v_i=3000$ км/с, скорость расширенного джета $v_j=0.2~c$, его угол полураствора $5^{\circ}$, плотность межзвездной среды $\rho_{amb} = 0.005~\rm{cm^{-3}}$ и ее температура $T_{amb} = 8\times10^4~\rm{K}$.
+Время моделирования составило 30 000 лет.
+
+Подобную структуру можно и получить и при других параметрах среды, при большей плотности и меньшей температуре, например
+ ($\rho_{amb}=0.05~\rm{cm^{-3}}$, $T_{amb} = 8.6\times10^3~\rm{K}$).
+Но в этом случае подобная структура с сильными реколлимационными ударными волнами, расположенными в местах соответствующих областям `e1' и `e2' будет наблюдаться на более поздних временах, примерно 100 000 лет.
+
+Первая реколлимационная ударная волна случается на расстоянии соответствующем ударной волне остановки изотропного ветра.
+Дальнейшее поведение расширенного джета - будет ли он распространяться с серией реколимационных ударных волн, или остановится после первой, определяется давлением внешней средыю Оно должно быть достаточно малым, чтобы направленный поток распространился на большие расстояния.
+К тому же, как показано в работе \cite{Marti2018}, относительно низкое внешнее давление необходимо для формирование перпендикулярных ударных волн (дисков Маха), а не системы косых ударных волн, менее пригодных для ускорения частиц.
+В дополнение, при численном моделировании необходимо достаточно высокое пространственное разрешение, чтобы было возможно наблюдать диски Маха.
+
+Одномерные профили плотности, скорости и энтропии гидродинамического потока вблизи областей `e1' and `e2' показаны на рисунке~\ref{MHD_profile}.
+Скачок плотности на обоих ударных волнах близок к 4, что говорит о сильной ударной волне.
+Скачки энтропии тоже достаточно большие.
+Так же имеются признаки более слабых ударных волн с меньшими скачками плотности.
+Кинетическое Монте-Карло моделирование показывает, что при учете давления ускоренных частиц, сжатие на фронте ударной волны может значительно превосходить адиабатическиое предельное значение 4 (см.
+например \cite{Bykov3inst2014}).
+
+\begin{figure}
+	\includegraphics[width=0.9\textwidth]{./img_part2/velocity4.png}
+	\caption{Моделированная структура туманности W50: a) крупномасштабная карта скорости b) карта скорости вблизи областей e1 и e2 c) карта плотности d) карта магнитного поля.} 
+	\label{velocity}
+\end{figure}
+
+\begin{figure}
+	\includegraphics[width=0.9\textwidth]{./img_part2/profile_1d_window2.png}
+	\caption{Одномерные профили плотности, скорости и энтропии, соответствующие расчетам, показанным на рисунке~\ref{velocity}.} 
+	\label{MHD_profile}
+\end{figure}
+
+\textcolor{red}{In the minimalists' scenario, the main goal was reproducing the overall morphology of the W50 radio emission and the emergence of the `extended X-ray jets' as a result of recollimation shocks in the anisotropic wind scenario.
+This model has several free parameters, including overall energetics, age of the system, density of the ambient medium, velocities and densities of the isotropic and polar winds, and the opening angle of the polar component.
+Various combinations of these parameters can lead to a qualitatively similar morphology of the nebula (see Fig.A1 in \cite{Churazov}).
+Here, our primary goal is to find parameters that lead to the efficient acceleration of the particles that give rise to the observed emission of EXJs.
+Knowing that accelerating electrons to above 100 TeV energies is especially efficient for strong shocks with velocities $\gtrsim$ 0.2c \cite{Churazov}, we ran the MHD version of a set-up with this value of the polar component velocity, but did not require the morphology of the simulated nebula boundary to match the observed one perfectly.
+We leave the task of finding a parameter combination that satisfies all observational requirements best for a future study.}
 \section{Расчет высокоэнергичного излучения}
+Туманность W50 является ярким источником высокоэнергичного электромагнитного излучения.
+Она видна в рентгеновском диапазоне, с четко различимыми излучающими областями \cite{Safi-Harb1997, Safi-Harb2022}, а так же в гамма-излучении, кторое носит более диффузный характер \cite{HESS2024SS433, LHAASO2024SS433}.
+Для рождения фотонов с такими высокими энергиями (несколько сотен ТэВ) необходимо ускорение частиц до высоких энергий, порядка 1 ПэВ.
+Мы использовали спектры ускоренных частиц, полученные методом Монте-Карло моделирования, описанного в работах \citep[][]{EBJ96, Bykov3inst2014}, задавая в расчетах параметры ударных волн, соответствующие ударным волнам, полученным из гидродинамического моделирования.
+Использованный метод Монте-Карло моделирования не может корректно расчитывать ускорение электронов, но в диффузионной модели с коэффициентом диффузии зависящим только от заряда и энергии частицы движение электронов и протонов с одинаковыми энергиями будет одинаковым.
+Вследствие чего для достаточно больших энергий функции распределения протонов и электронов будут подобны.
+Функции распределения ускоренных частиц на фронтах ударных волн, с параметрами соответствующими ударным волнам, приведенным на рисунке \ref{velocity} показаны на рисунке \ref{pdf_shock_Q_esc}.
+
+\begin{figure}
+	\includegraphics[width=0.9\textwidth]{./img_part2/pdf_shock_Q_esc.pdf}
+	\caption{Функции распределения ускоренных частиц на фронтах ударных волн, полученные Монте-Карло моделированием.
+Красная и зеленая кривая - протоны, синяя и фиолетовая - электроны.
+Пунктирные лини показывают низкоэнергичную часть электронной функции распределения, которая не может быть точно получена из метода Монте-Карло.}
+	\label{pdf_shock_Q_esc}
+\end{figure}
+
 \subsection{Рентгеновское излучение}
+X-ray emission from SS433/W50 nebula has several specific features - inhomogeneous spatial profile with bright regions, energy spectra, and polarization.
+In this section, we present a model describing observed X-ray emission from the eastern lobe of W50 \citep{Brinkmann2007,Safi-Harb2022}, via synchrotron radiation of electrons, accelerated on two trans-relativistic shocks, in the inhomogeneous anisotropic turbulent magnetic field.
+Assumption of inhomogeneous profile of turbulent magnetic field is supported by the fact that narrow peaks in observed profiles in the `Head' region in energy ranges 0.3-10 keV and 3-30 keV \cite{Safi-Harb2022} have similar widths, which can be explained by the structure of the magnetic field.
+On the contrary, under the assumption that the peak width is a result of synchrotron losses,  peaks should have different widths.
+
+\begin{figure}[th!]
+	\includegraphics[width=0.9\textwidth]{./img_part2/W50synch.png}
+	\caption{Modeled spectrum of X-ray synchrotron radiation in `Head' and `Cone' regions and observational data from \cite{Safi-Harb2022}.
+The estimated model fluxes of the GHz synchrotron radio emission from the two regions are about a few mJy.
+}
+	\label{synchrotron}
+\end{figure}
+
+The exact description of VHE particle transport in the collimated outflow and the surrounding plasma cocoon would require a multiscale modeling of stochastic magnetic fields, which in turn would include different sources of turbulence associated with a complex MHD dynamics with recollimation shocks, shear flows, and cosmic-ray-driven instabilities.
+Such a model is unfeasible now.
+Instead, we parameterized the diffusion coefficient of relativistic particles in the system.
+
+To evaluate the electron distribution function in the downstream flow, we use the advection - diffusion equation, neglecting stochastic acceleration and adiabatic losses:
+
+\begin{eqnarray}
+	&& \frac{\partial f({\bf r},E)}{\partial t} + {\bf v_{\alpha}(r)}\frac{\partial f({\bf r},E)}{\partial r_\alpha} = \nonumber \\ &=& \frac{\partial}{\partial r_\alpha} D_{\alpha \beta}({\bf r},E)\frac{\partial f({\bf r}, E)}{\partial r_\beta}  - \frac{\partial [a({\bf r},E) f({\bf r},E)]}{\partial E}.
+\end{eqnarray}
+
+Here $D_{\alpha \beta}({\bf r},E)$ is the diffusion coefficient,  $a({\bf r},E)=\frac{4\sigma_T}{3m_e^2c^3}\left(B({\bf r})^2/8\pi + U_{ph}\right)E^2$ is the electron synchrotron and Compton loss rates, $\sigma_T$ is the Thompson cross section, $B$ is the magnetic field and $U_{ph}$ is the energy density of the photon field.
+
+We simulated a set of models with different assumptions about the diffusion coefficient of VHE particles.
+If $D < 3\times 10^{28} см^2/c$, which is about 100 times larger than the Bohm diffusion coefficient at 100 TeV, the advection dominates the transport along the X-ray jet.
+Then, in a stationary regime, one can solve this equation by calculating the energy losses of the single electron during its advection into the downstream.
+
+\begin{equation}
+	E(z) = \frac{E\left(0\right)}{1+E\left(0\right)k(z)}
+\end{equation}
+
+where $k(z) = \int_0^z \frac{4\sigma_T}{3m_e^2c^3}\left(B(z)^2/8\pi + U_{ph}\right) \frac{dx}{v(z)}$. 
+The photon field is taken as cosmic microwave background radiation, scattering on all other possible components is suppressed due to the Klein-Nishina regime in the considered energy range (electrons with energies higher than 100 TeV).
+The turbulent magnetic field was established according to Section \ref{sec:CorSh}.
+We used two models of the magnetic
+field. In the first model, we calculate
+only a small-scale turbulent magnetic field. 
+In this model, we assume that the field decay stops at some distance
+from the shock, so the magnetic field remains constant onward. 
+In our calculation, we assume that the critical value at which the field decay stops is $8$ $\mu G$.
+In Model 2, we assume a superposition of the simulated short- and long-scale magnetic fields calculated
+for $l_{*}=3\cdot10^{17}$~cm and $l_{*}=3\cdot10^{19}$~cm. 
+Using the results of MHD magnetic field simulations discussed in Section \ref{sec:CorSh}, we constructed a smooth analytical approximation for square-averaged magnetic fields, neglecting small-scale fluctuations, as was done in \cite{2024PhRvD.110b3041B} for Tycho SNR.
+The averaging
+is performed on slices of the simulated MHD datasets at fixed coordinates $z$. 
+The square average magnetic field profiles for models 1 and 2
+are shown in Fig.~\ref{fig:Bprof}. 
+The approximation has a different
+functional dependence on the distance from the shock for normal ($B_{\parallel}$) and transverse (${B_{\perp}}$) field components.
+
+Thus, in the advection model, the electron distribution at a distance $z$ from the shock in the downstream can be evaluated as
+
+\begin{equation}
+	f(z,E)=f\left(0,\frac{E}{1-k(z)E}\right)\frac{1}{\left(1-k(z)E\right)^2}
+\end{equation}
+
+where $f(0,E)$ is the distribution function of electrons at the shock front, shown in Figure~\ref{pdf_shock_Q_esc}. 
+The modeled level of synchrotron radiation derived with the electron-to-proton ratio fixed to be 3.3$ \times 10^{-3}$ at 1 GeV described above fits well with the observed data.
+The energy spectrum of the `Head' and `Cone' regions in the short-scale Bell's turbulent field model with characteristic length $l_\ast = 3\times10^{17}~\rm{cm}$, maximum field on the front $15~\rm{\mu G}$ and constant level $8~\rm{\mu G}$, constant velocity $v = 10.3\times10^8~\rm{cm/s}$ (consistent with Monte Carlo simulation), and observational data \cite{Safi-Harb2022} are shown in Fig.~\ref{synchrotron}.
+The modeled spectra are consistent with the power-law functions obtained in the observations.
+
+\begin{figure}[h!]
+	\includegraphics[width=0.9\textwidth]{./img_part2/profile.png}
+	\caption{Spatial profiles of synchrotron emission in e1 region in the energy range 0.3-10 keV predicted by the model in comparison with the XMM data \cite{Safi-Harb2022} (top panel), in the energy range 3-30 keV and the NuSTAR data \cite{Safi-Harb2022} (middle panel), and the profile of magnetic field used in the model (bottom panel).}
+	\label{profile_e1}
+\end{figure}
+
+\begin{figure}[h!]
+	\includegraphics[width=0.9\textwidth]{./img_part2/profile_Brinkmann.png}
+	\caption{Modeled spatial profile of synchrotron radiation in e2 region in energy range 0.3-10 keV and XMM data \cite{Brinkmann2007} (top panel), and profile of magnetic field used for modeling radiation (bottom panel).
+}
+	\label{profile_Brinkmann}
+\end{figure}
+
+Spatial profile of the X-ray luminosity for the same model is shown in Figure~\ref{profile_e1}.
+It is also consistent with observations.
+The profile for the `e2' region is described in \cite{Brinkmann2007}, and was studied only by XMM.
+To explain this case, we use a turbulent field with the maximum value of $9~\rm{\mu G}$, constant level of $3~\rm{\mu G}$, characteristic length $2\times10^{18}~\rm{cm}$ and velocity $9.3\times10^8~\rm{cm/s}$, results of simulation are shown in Fig.~\ref{profile_Brinkmann}.
+
+\begin{figure}[h!]
+	\includegraphics[scale=0.4]{B_test_model1}
+	\includegraphics[scale=0.4]{B_test_model2}
+
+	\caption{\label{fig:Bprof} The upper panel shows the downstream rms magnetic field spatial profile approximations for model 1 described in the text, while
+		the lower panel shows it for model 2. }
+
+\end{figure}
 \subsection{Гамма излучение}

### 2025-12-09 disser/part2.tex
disser

@@
 
-\section{Моделирование структуры туманности}
+\section{Моделирование структуры туманности} \label{W50morphology}
 \begin{figure}
@@
 \subsection{Рентгеновское излучение}
-X-ray emission from SS433/W50 nebula has several specific features - inhomogeneous spatial profile with bright regions, energy spectra, and polarization.
-In this section, we present a model describing observed X-ray emission from the eastern lobe of W50 \citep{Brinkmann2007,Safi-Harb2022}, via synchrotron radiation of electrons, accelerated on two trans-relativistic shocks, in the inhomogeneous anisotropic turbulent magnetic field.
-Assumption of inhomogeneous profile of turbulent magnetic field is supported by the fact that narrow peaks in observed profiles in the `Head' region in energy ranges 0.3-10 keV and 3-30 keV \cite{Safi-Harb2022} have similar widths, which can be explained by the structure of the magnetic field.
-On the contrary, under the assumption that the peak width is a result of synchrotron losses,  peaks should have different widths.
+Рентгеновское излучение SS433/W50 имеет несколько особенностей - неоднородное пространственное распределение с несколькими яркими областями, спектр с разными степенными индексами в разных диапазонах и поляризацию.
+В этом разделе представлена модель объясняющая рентгеновское излучение от восточной части туманности W50 как синхротронное излучение электронов, ускоренных на транс-релятивистских ударных волнах, в неоднородном турбулентном магнитном поле.
+Предположение о неоднородности магнитного поля подтверждается тем, что узкие пространственные пики яркости излучения в области, обозначенной как `Head' в диапазонах энергии 0.3-10 кэВ и 3-30 кэВ \cite{Safi-Harb2022} имеют приблизительно одинаковую ширину, в то время как если бы ширина пиков определялась бы синхротронными потерями в однородном поле, они имели бы разную ширину.
 
@@
 	\includegraphics[width=0.9\textwidth]{./img_part2/W50synch.png}
-	\caption{Modeled spectrum of X-ray synchrotron radiation in `Head' and `Cone' regions and observational data from \cite{Safi-Harb2022}.
-The estimated model fluxes of the GHz synchrotron radio emission from the two regions are about a few mJy.
-}
+	\caption{Моделированный спектр рентгеновского излучения в областях `Head' и `Cone' и наблюдательные данные из работы\cite{Safi-Harb2022}.} 
 	\label{synchrotron}
@@
 
-The exact description of VHE particle transport in the collimated outflow and the surrounding plasma cocoon would require a multiscale modeling of stochastic magnetic fields, which in turn would include different sources of turbulence associated with a complex MHD dynamics with recollimation shocks, shear flows, and cosmic-ray-driven instabilities.
-Such a model is unfeasible now.
-Instead, we parameterized the diffusion coefficient of relativistic particles in the system.
+Полноценное описание переноса частиц в сложном гидродинамическом потоке, описанном в разделе \ref{W50morphology},  требует моделирования стохастического магнитного поля.
+Что в свою очередь требует расчета турбулентности создаваемой реколимационными ударными волнами, сдвиговыми течениями и током космическич лучей.
+Это задача недоступная численному моделированию на современных компьютерах, поэтому для расчета распространения ускоренных электронов внутри расширенного джета мы решаем одномерное конвекционно-диффузионное уравнение с добавленным членом для учета синхротронных потерь в заданном профиле усредненного магнитного поля \ref{convection-diffusion}.
 
-To evaluate the electron distribution function in the downstream flow, we use the advection - diffusion equation, neglecting stochastic acceleration and adiabatic losses:
-
-\begin{eqnarray}
+\begin{eqnarray}\label{convection-diffusion}
 	&& \frac{\partial f({\bf r},E)}{\partial t} + {\bf v_{\alpha}(r)}\frac{\partial f({\bf r},E)}{\partial r_\alpha} = \nonumber \\ &=& \frac{\partial}{\partial r_\alpha} D_{\alpha \beta}({\bf r},E)\frac{\partial f({\bf r}, E)}{\partial r_\beta}  - \frac{\partial [a({\bf r},E) f({\bf r},E)]}{\partial E}.
@@
 
-Here $D_{\alpha \beta}({\bf r},E)$ is the diffusion coefficient,  $a({\bf r},E)=\frac{4\sigma_T}{3m_e^2c^3}\left(B({\bf r})^2/8\pi + U_{ph}\right)E^2$ is the electron synchrotron and Compton loss rates, $\sigma_T$ is the Thompson cross section, $B$ is the magnetic field and $U_{ph}$ is the energy density of the photon field.
+Где $D_{\alpha \beta}({\bf r},E)$ - коэффициент диффузии,  $a({\bf r},E)=\frac{4\sigma_T}{3m_e^2c^3}\left(B({\bf r})^2/8\pi + U_{ph}\right)E^2$ - потери электронов на синхротронное и комптоновское излучение, $\sigma_T$ - томпсоновское сечение, $B$ - магнитное поле и $U_{ph}$ - плотность энергии фотонного поля.
 
-We simulated a set of models with different assumptions about the diffusion coefficient of VHE particles.
-If $D < 3\times 10^{28} см^2/c$, which is about 100 times larger than the Bohm diffusion coefficient at 100 TeV, the advection dominates the transport along the X-ray jet.
-Then, in a stationary regime, one can solve this equation by calculating the energy losses of the single electron during its advection into the downstream.
+Чтобы еще упростить уравнение, мы оценили вклад диффузионного члена, проведя Монте-Карло моделирование распространения частиц в соответствии с уравнением \ref{convection-diffusion}.
+Было установлено, что для коэфициентов диффузии $D < 3\times 10^{28} см^2/c$, что в 100 раз меньше бомовского коэффициента для энергии 100 ТэВ, конвекционный перенос вдоль оси джета доминирует доминирует над диффузионным на интересующих нас масштабах.
+Таким образом, дифузионным членом можно пренебречь.
+Тогда в приближении стационарности уравнение \ref{convection-diffusion} можно решить аналитически.
+Энергия частицы при движении вдоль оси z меняется как
 
@@
 
-where $k(z) = \int_0^z \frac{4\sigma_T}{3m_e^2c^3}\left(B(z)^2/8\pi + U_{ph}\right) \frac{dx}{v(z)}$. 
-The photon field is taken as cosmic microwave background radiation, scattering on all other possible components is suppressed due to the Klein-Nishina regime in the considered energy range (electrons with energies higher than 100 TeV).
-The turbulent magnetic field was established according to Section \ref{sec:CorSh}.
-We used two models of the magnetic
+где $k(z) = \int_0^z \frac{4\sigma_T}{3m_e^2c^3}\left(B(z)^2/8\pi + U_{ph}\right) \frac{dx}{v(z)}$.
+
+А функция распределения при движения за фронт ударной волны определяется уравнением
+
+\begin{equation}
+	f(z,E)=f\left(0,\frac{E}{1-k(z)E}\right)\frac{1}{\left(1-k(z)E\right)^2}
+\end{equation}
+
+где $f(0,E)$ - функция распределения электронов на фронте ударной волны, показанная на рисунке~\ref{pdf_shock_Q_esc}.
+
+В качестве фотонного поля бралось реликтовое излучение, так как для интересующих нас наиболее высоких энрегий (выше 100 ТэВ), рассеяние на более высокочастотных фотонах подавлено в Клейн-Нишиновском режиме.
+
+ The turbulent magnetic field was established according to Section \ref{sec:CorSh}. We used two models of the magnetic
 field. In the first model, we calculate
@@
 from the shock, so the magnetic field remains constant onward.
+
+When MHD turbulence containing strong density fluctuations passes through the shock,  strong oscillations of the shock front occur.
+These oscillations lead to significant turbulent fluctuations in the downstream flow velocity.
+Moreover, turbulent velocity fluctuations amplify the magnetic field downstream by the mechanism of a small-scale dynamo.
+It appears that in such a system, the longitudinal component of the magnetic field relative to the shock normal is amplified more strongly than the transverse components \citep{inoue13,Zirakashvili_2008pol,lazarian_MFA22,federath23,2024PhRvD.110b3041B}.
+Thus, there is a region downstream where the longitudinal component of the turbulent magnetic field is greater than the transverse one.
+Synchrotron radiation of electrons from this region should be highly polarized with polarization direction transverse to the shock normal.
+
+We simulate the passage of MHD turbulence amplified by the Bell instability through the shock to determine the anisotropy of the magnetic field downstream in a way similar to that in the paper \citep{2024PhRvD.110b3041B}.
+The 3D calculations are performed on the basis of the MHD module of the open-source code PLUTO.
+The calculation is performed in two stages: in the first stage, the amplification of MHD turbulence by the Bell instability is simulated in a cubic box (512x512x512 cells), and in the second stage, the passage of turbulence through the shock front is simulated (see \citep{2024PhRvD.110b3041B}).
+The sizes of the simulation box at the first stage in all directions vary from $-l_{\ast}$ to $l_{\ast}$, where $l_{\ast}$ is the normalization length of the PLUTO code.
+The size of the simulation box at the second stage along the z-axis is from 0 to $10l_{\ast}$, and along the x and y axes is from $-l_{\ast}$ to $l_{\ast}$ (512x512x2560 cells).
+The box of the first stage is used at the second stage as a boundary condition on the right boundary along the z-axis.
+The boundary condition on the left boundary of the z-axis is an outflow.
+The boundary conditions along the x and y axes are periodic.
+We vary the length of $l_{\ast}$ to achieve agreement with the observational data.
+The energy-carrying scale of turbulence $l_{0}$ (calculated at the first stage) passing the shock is approximately equal to $l_{\ast}/3$.
+At the initial moment, the shock is set perpendicular to the z-axis at the point $z=2 l_{\ast}$ and remains at rest in the absence of perturbations.
+
+The simulation parameters in the first stage of the calculations are taken in accordance with the conditions upstream of the first shock described in Section \ref{sec:MC}.
+The ratio of the rms deviation of the density to the average density is $\delta \rho/\rho\approx 0.6$ at the end of the first-stage simulation.
+
+The results of the simulations for turbulence with strong density fluctuations and magnetic fields amplified by the Bell instability are shown in  Fig.~\ref{B2_Bell_art}.
+The origin of the z coordinate in Fig.~\ref{B2_Bell_art} and Fig.~\ref{B2_all_longscale_art} is shifted to the location of the shock front.
+These figures show the profiles of the magnetic field components averaged over slices with fixed coordinate z.
+In the figures, $B_{\parallel}$ is the RMS of the magnetic field component
+parallel to the shock normal of the undisturbed shock.
+$B_{\perp}=\sqrt{B_{x}^{2}+B_{y}^{2}}$ is the RMS perpendicular component of the magnetic field, where $B_{x}$ is the RMS value of the x projection of the magnetic field, $B_{y}$ is the RMS value of the y projection of the magnetic field, and $B=\sqrt{B_{\parallel}^{2}+B_{\perp}^{2}}$.
+The blue curves in the figures show the same data after filtering out small-scale fluctuations.
+They are used in subsequent calculations.
+
 In our calculation, we assume that the critical value at which the field decay stops is $8$ $\mu G$.
@@
 
-Thus, in the advection model, the electron distribution at a distance $z$ from the shock in the downstream can be evaluated as
-
-\begin{equation}
-	f(z,E)=f\left(0,\frac{E}{1-k(z)E}\right)\frac{1}{\left(1-k(z)E\right)^2}
-\end{equation}
-
-where $f(0,E)$ is the distribution function of electrons at the shock front, shown in Figure~\ref{pdf_shock_Q_esc}. 
 The modeled level of synchrotron radiation derived with the electron-to-proton ratio fixed to be 3.3$ \times 10^{-3}$ at 1 GeV described above fits well with the observed data.
@@
 
-\begin{figure}[h!]
-	\includegraphics[scale=0.4]{B_test_model1}
-	\includegraphics[scale=0.4]{B_test_model2}
+\begin{figure}[h]
+	\begin{minipage}{0.49\textwidth}
+		\centering
+		\includegraphics[width=0.95\textwidth]{./img_part2/B2_Bell_art.pdf} 
+		\caption{Profiles of RMS components of the magnetic field in the shock downstream.
+The turbulence is produced by the passage of fluctuations amplified in the upstream by Bell's instability of accelerated cosmic rays.
+The green curve is the parallel magnetic field component.
+The red curve is the perpendicular magnetic field component.
+The dark green curve is the total magnetic field.
+Blue curves are fits to the data.}
+		\label{B2_Bell_art}
+	\end{minipage} \hfill
+	\begin{minipage}{0.49\textwidth}
+		\centering
+		\includegraphics[width=0.95\textwidth]{./img_part2/B2_all_longscale_art.pdf} 
+		\caption{Profiles of RMS components of the magnetic field in the shock downstream produced by the shock amplification of the preexisting long-wavelength fluctuations in the jet.
+The green curve is the parallel magnetic field component.
+The red curve is the perpendicular magnetic field component.
+The dark green curve is the total magnetic field.
+Blue curves are fits to the data.}
+\label{B2_all_longscale_art}
+	\end{minipage}
+\end{figure}
 
-	\caption{\label{fig:Bprof} The upper panel shows the downstream rms magnetic field spatial profile approximations for model 1 described in the text, while
-		the lower panel shows it for model 2. }
-
-\end{figure}
 \subsection{Гамма излучение}

### 2025-12-13 disser/part2.tex
w50 rentgen

@@
 
- The turbulent magnetic field was established according to Section \ref{sec:CorSh}. We used two models of the magnetic
-field. In the first model, we calculate
-only a small-scale turbulent magnetic field. 
-In this model, we assume that the field decay stops at some distance
-from the shock, so the magnetic field remains constant onward.
+Магнитное поле, необходимое для сильного синхротронного излучения, обеспечивается за счет создания и усиления турбулентности ударной волной.
+Известно, что неустойивости, создаваемые током космических лучей, создают турбулентное поле в предфронте ударной волны \cite{Bell2004,Bykov3inst2014}.
+При переходе этого поля через фронт, оно дополнительно усиливается за счет сжатия на ударной волне \cite{Zirakashvili2008pol, Inoue13}.
+Для расчета синхротронного излучения мы брали пространственный профиль турбулентного магнитного поля за фронтом ударной волны в соответствии с процедурой, описанной в работе \cite{BykovTycho}.
+С помощью кода PLUTO исследовалось развитие мелкомасштабной неустойчивостей заданным током космических лучей, а потом прохождение созданной турбулентности через фронт ударной волны.
+Так же исследовалась крупномасштабная турбулентность, связанная с гидродиномическим течением.
+Полученные профили среднеквадратичного магнитного поля для различных видов турбулентности показаны на рисунках \ref{B2_Bell_art} и \ref{B2_all_longscale_art}.
 
-When MHD turbulence containing strong density fluctuations passes through the shock,  strong oscillations of the shock front occur.
-These oscillations lead to significant turbulent fluctuations in the downstream flow velocity.
-Moreover, turbulent velocity fluctuations amplify the magnetic field downstream by the mechanism of a small-scale dynamo.
-It appears that in such a system, the longitudinal component of the magnetic field relative to the shock normal is amplified more strongly than the transverse components \citep{inoue13,Zirakashvili_2008pol,lazarian_MFA22,federath23,2024PhRvD.110b3041B}.
-Thus, there is a region downstream where the longitudinal component of the turbulent magnetic field is greater than the transverse one.
-Synchrotron radiation of electrons from this region should be highly polarized with polarization direction transverse to the shock normal.
+\begin{figure}[h]
+	\begin{minipage}[t]{0.49\textwidth}
+		\centering
+		\includegraphics[width=0.95\textwidth]{./img_part2/B2_Bell_art.pdf} 
+		\caption{Профиль среднеквадратичных компонент магнитного поля при прохождении турбулентности, создаваемой Белловской неустойчивостью, через ударную волну.} 
+		\label{B2_Bell_art}
+	\end{minipage} \hfill
+	\begin{minipage}[t]{0.49\textwidth}
+		\centering
+		\includegraphics[width=0.95\textwidth]{./img_part2/B2_all_longscale_art.pdf} 
+		\caption{Профиль среднеквадратичных компонент магнитного поля при прохождении длинноволновой турбулентности через ударную волну.} 
+		\label{B2_all_longscale_art}
+	\end{minipage}
+\end{figure}
 
-We simulate the passage of MHD turbulence amplified by the Bell instability through the shock to determine the anisotropy of the magnetic field downstream in a way similar to that in the paper \citep{2024PhRvD.110b3041B}.
-The 3D calculations are performed on the basis of the MHD module of the open-source code PLUTO.
-The calculation is performed in two stages: in the first stage, the amplification of MHD turbulence by the Bell instability is simulated in a cubic box (512x512x512 cells), and in the second stage, the passage of turbulence through the shock front is simulated (see \citep{2024PhRvD.110b3041B}).
-The sizes of the simulation box at the first stage in all directions vary from $-l_{\ast}$ to $l_{\ast}$, where $l_{\ast}$ is the normalization length of the PLUTO code.
-The size of the simulation box at the second stage along the z-axis is from 0 to $10l_{\ast}$, and along the x and y axes is from $-l_{\ast}$ to $l_{\ast}$ (512x512x2560 cells).
-The box of the first stage is used at the second stage as a boundary condition on the right boundary along the z-axis.
-The boundary condition on the left boundary of the z-axis is an outflow.
-The boundary conditions along the x and y axes are periodic.
-We vary the length of $l_{\ast}$ to achieve agreement with the observational data.
-The energy-carrying scale of turbulence $l_{0}$ (calculated at the first stage) passing the shock is approximately equal to $l_{\ast}/3$.
-At the initial moment, the shock is set perpendicular to the z-axis at the point $z=2 l_{\ast}$ and remains at rest in the absence of perturbations.
-
-The simulation parameters in the first stage of the calculations are taken in accordance with the conditions upstream of the first shock described in Section \ref{sec:MC}.
-The ratio of the rms deviation of the density to the average density is $\delta \rho/\rho\approx 0.6$ at the end of the first-stage simulation.
-
-The results of the simulations for turbulence with strong density fluctuations and magnetic fields amplified by the Bell instability are shown in  Fig.~\ref{B2_Bell_art}.
-The origin of the z coordinate in Fig.~\ref{B2_Bell_art} and Fig.~\ref{B2_all_longscale_art} is shifted to the location of the shock front.
-These figures show the profiles of the magnetic field components averaged over slices with fixed coordinate z.
-In the figures, $B_{\parallel}$ is the RMS of the magnetic field component
-parallel to the shock normal of the undisturbed shock.
-$B_{\perp}=\sqrt{B_{x}^{2}+B_{y}^{2}}$ is the RMS perpendicular component of the magnetic field, where $B_{x}$ is the RMS value of the x projection of the magnetic field, $B_{y}$ is the RMS value of the y projection of the magnetic field, and $B=\sqrt{B_{\parallel}^{2}+B_{\perp}^{2}}$.
-The blue curves in the figures show the same data after filtering out small-scale fluctuations.
-They are used in subsequent calculations.
-
-In our calculation, we assume that the critical value at which the field decay stops is $8$ $\mu G$.
-In Model 2, we assume a superposition of the simulated short- and long-scale magnetic fields calculated
-for $l_{*}=3\cdot10^{17}$~cm and $l_{*}=3\cdot10^{19}$~cm. 
-Using the results of MHD magnetic field simulations discussed in Section \ref{sec:CorSh}, we constructed a smooth analytical approximation for square-averaged magnetic fields, neglecting small-scale fluctuations, as was done in \cite{2024PhRvD.110b3041B} for Tycho SNR.
-The averaging
-is performed on slices of the simulated MHD datasets at fixed coordinates $z$. 
-The square average magnetic field profiles for models 1 and 2
-are shown in Fig.~\ref{fig:Bprof}. 
-The approximation has a different
-functional dependence on the distance from the shock for normal ($B_{\parallel}$) and transverse (${B_{\perp}}$) field components.
-
-The modeled level of synchrotron radiation derived with the electron-to-proton ratio fixed to be 3.3$ \times 10^{-3}$ at 1 GeV described above fits well with the observed data.
-The energy spectrum of the `Head' and `Cone' regions in the short-scale Bell's turbulent field model with characteristic length $l_\ast = 3\times10^{17}~\rm{cm}$, maximum field on the front $15~\rm{\mu G}$ and constant level $8~\rm{\mu G}$, constant velocity $v = 10.3\times10^8~\rm{cm/s}$ (consistent with Monte Carlo simulation), and observational data \cite{Safi-Harb2022} are shown in Fig.~\ref{synchrotron}.
-The modeled spectra are consistent with the power-law functions obtained in the observations.
+Для объяснения наблюдаемых профилей рентгеновской яркости в близи первой реколлимационной ударной волны (областях `Head' и `Cone') мы использовали следующую многокомпонентную модель магнитного поля - сумма описанных выше двух турбулентных полей, с равными на фронте значениями магнитного поля $7.5$ мкГс и характерными масштабами $3\times10^{17}$ см для короткомасштабной неустойчивости и $3\times10^{19}$ см для крупномасштабной.
+Так же считалось, что среднеквадратичное перестает спадать с удалением от фронта при достижении значения $8$ мкГс.
+Скорость течения за фронтом ударной волны бралась равной $v = 10.3\times10^8$ см/с (согласованно с Монте-Карло моделированием спектра ускоренных частиц).
+Результаты моделирования синхротронного излучения, а так же наблюдаетельные данные \cite{Safi-Harb2022} представлены на рисунке \ref{synchrotron}.
+Полученные спектры согласуются с данными наблюдений
 
 \begin{figure}[h!]
-	\includegraphics[width=0.9\textwidth]{./img_part2/profile.png}
-	\caption{Spatial profiles of synchrotron emission in e1 region in the energy range 0.3-10 keV predicted by the model in comparison with the XMM data \cite{Safi-Harb2022} (top panel), in the energy range 3-30 keV and the NuSTAR data \cite{Safi-Harb2022} (middle panel), and the profile of magnetic field used in the model (bottom panel).}
+	\includegraphics[width=0.8\textwidth]{./img_part2/profile.png}
+	\caption{Моделированные пространственные профили синхротронного излучения вблизи области e1 в диапазоне 0.3-10 кэВ и данные наблюдений XMM (верхняя панель), в диапазоне 3-30 кэВ и данные наблюдений NuSTAR (средняя панель) и профиль используемого магнитного поля (нижняя панель).}
 	\label{profile_e1}
@@
 \begin{figure}[h!]
-	\includegraphics[width=0.9\textwidth]{./img_part2/profile_Brinkmann.png}
-	\caption{Modeled spatial profile of synchrotron radiation in e2 region in energy range 0.3-10 keV and XMM data \cite{Brinkmann2007} (top panel), and profile of magnetic field used for modeling radiation (bottom panel).
-}
+	\includegraphics[width=0.8\textwidth]{./img_part2/profile_Brinkmann.png}
+	\caption{Моделированный пространственный профиль синхротронного излучения в диапазоне 0ю3-10 кэВ и данные наблюдений XMM (верхняя панель) и использованный профиль магнитного поля (нижняя панель).} 
 	\label{profile_Brinkmann}
@@
 
-Spatial profile of the X-ray luminosity for the same model is shown in Figure~\ref{profile_e1}.
-It is also consistent with observations.
-The profile for the `e2' region is described in \cite{Brinkmann2007}, and was studied only by XMM.
-To explain this case, we use a turbulent field with the maximum value of $9~\rm{\mu G}$, constant level of $3~\rm{\mu G}$, characteristic length $2\times10^{18}~\rm{cm}$ and velocity $9.3\times10^8~\rm{cm/s}$, results of simulation are shown in Fig.~\ref{profile_Brinkmann}.
-
-\begin{figure}[h]
-	\begin{minipage}{0.49\textwidth}
-		\centering
-		\includegraphics[width=0.95\textwidth]{./img_part2/B2_Bell_art.pdf} 
-		\caption{Profiles of RMS components of the magnetic field in the shock downstream.
-The turbulence is produced by the passage of fluctuations amplified in the upstream by Bell's instability of accelerated cosmic rays.
-The green curve is the parallel magnetic field component.
-The red curve is the perpendicular magnetic field component.
-The dark green curve is the total magnetic field.
-Blue curves are fits to the data.}
-		\label{B2_Bell_art}
-	\end{minipage} \hfill
-	\begin{minipage}{0.49\textwidth}
-		\centering
-		\includegraphics[width=0.95\textwidth]{./img_part2/B2_all_longscale_art.pdf} 
-		\caption{Profiles of RMS components of the magnetic field in the shock downstream produced by the shock amplification of the preexisting long-wavelength fluctuations in the jet.
-The green curve is the parallel magnetic field component.
-The red curve is the perpendicular magnetic field component.
-The dark green curve is the total magnetic field.
-Blue curves are fits to the data.}
-\label{B2_all_longscale_art}
-	\end{minipage}
-\end{figure}
+Аналогично было проведено моделирование рентгеновского синхротронного излучения в области `e2', наблюдаемый пространственный профиль которого описан в работе \cite{Brinkmann2007}.
+В этом случае мы взяли турбулентное магнитное поле с максимальным значением $9$ мкГс на фронте ударной волны, постоянным уровнем $3$ мкГс, характерной длиной турбулентности $2\times10^{18}$ см и скоростью за фронтом $9.3\times10^8$ см/с.
+Результаты моделирования показаны на рисунке \ref{profile_Brinkmann}.
 
 \subsection{Гамма излучение}
+In this Section, we model the gamma radiation of W50, which is observed by H.E.S.S.
+\cite{HESS2024SS433} and LHAASO \cite{LHAASO2024SS433}.
+The leptonic model assumes that it is produced via inverse Compton scattering of electrons accelerated by recollimation shocks in the collimated outflow.
+
+In Fig.
+\ref{compton}, the red curve shows the gamma radiation produced within the outflow by accelerated VHE electrons as they are advected through the downstream of the first shock to the second one.
+This corresponds to pure advection of VHE electrons along the X-ray outflow with the turbulent magnetic field and the distribution function of electrons, which were used to simulate the synchrotron X-ray radiation discussed above.
+The red curve is significantly below the observational data, and it has a different spectral shape at low energies.
+This is because the radiating electrons in the X-ray jet are in the thin target regime.
+On the other hand, the electrons accelerated at the first shock diffuse in the direction transverse to the X-ray outflow.
+These electrons left the axial X-ray outflow and diffused through the surrounding cocoon, comprising a spatially broad component.
+If their diffusion coefficient in the cocoon is $\sim 10^{28} см^2/c$ at 100 TeV, they would radiate gamma-rays in the thick target regime in a wide cocoon of a size $\sim$ 30 pc around the axial jet.
+Electrons that have flown through the X-ray axial jet with a magnetic field amplified by cosmic ray-driven instabilities at the shock and escape into the cocoon downstream would also radiate in the thick regime.
+In Fig.
+\ref{compton}, the gamma-ray spectrum integrated over the wide cocoon is shown by the blue curve, which is in reasonably good agreement with the LHAASO data points taken from \cite{LHAASO2024SS433}.
+The gamma-ray spectrum integrated over a somewhat narrower region of the cocoon corresponds to the intermediate regime between the thin and the thick targets.
+It is shown by a brown curve in Fig.~\ref{compton} and is compared with the data reported by H.E.S.S.
+\cite{HESS2024SS433}.
+
+The LHAASO images above 100 TeV \cite{LHAASO2024SS433} show a more extended emission region compared to the lower energy bands.
+This morphology can be understood within our diffusive shock acceleration model, where the highest energy particles escape into the shock upstream, where the low magnetic field allows the escaped VHE leptons to form an extended gamma-ray emission region by the inverse Compton process.
+
+In our model, $\sim$ 30 $\%$ of the electrons accelerated at the shock front had escaped from the jet and diffused in the cocoon.
+This allows us to explain the observational data from LHAASO as shown in Figure \ref{compton}.
+H.E.S.S.
+data fall between the model for LHAASO and radiation from the jet itself, which can be explained by a smaller region size, observed by H.E.S.S.
+
+There are recent studies of the acceleration of PeV protons in microquasars \citep{2024A&A...691A..93A,2024arXiv241108762P,2025arXiv250510620K}.
+A spherically symmetric semi-analytical model of particle acceleration to energies of several PeV by the termination wind shock was discussed in \cite{2024arXiv241108762P}.
+The acceleration of particles in the shear flows of jet-cocoon-structured microquasars is suggested in \citep{2025arXiv250620193Z}.
+Shear acceleration in the model requires the injection of particles pre-accelerated up to TeV regime energies.
+Hadronic interactions of accelerated protons are producing a substantial fraction of the observed VHE gamma-ray emission in the models \citep{2024arXiv241108762P,2025arXiv250510620K}.
+
+The recollimation shocks in the diffusive shock acceleration model considered here efficiently accelerate protons up to PeV energies, as is seen in Fig. \ref{pdf_shock_Q_esc}.
+In our model, $\gsim$ 10\% of the shocked jet power is transferred to accelerated protons while the VHE emission comes from the inverse Compton radiation of VHE electrons accelerated at shocks, producing polarized synchrotron X-ray emission of the extended X-ray jet.
+The leptonic gamma-ray emission is produced in the thick target regime within the cocoon surrounding the extended X-ray jet.
+The VHE gamma rays can be produced by hadronic interactions of the protons with ambient matter.
+The very low number density of the plasma in the extended jet and in the surrounding cocoon would require a very small diffusion coefficient for the accelerated VHE protons to contribute to the detected gamma-ray emission from the cocoon, given by the gamma-ray significance maps of H.E.S.S.
+and LHAASO.
+Therefore, most of the accelerated protons would escape the accelerator in the SS433/W50 extended X-ray jet and then contribute to the gamma-ray emission from the denser gas in the vicinity of W50 and supply the galactic cosmic-ray population.
 \section{Выводы} \label{SS433conclision}

### 2025-12-13 disser/part2.tex
disser

@@
 Результаты моделирования синхротронного излучения, а так же наблюдаетельные данные \cite{Safi-Harb2022} представлены на рисунке \ref{synchrotron}.
-Полученные спектры согласуются с данными наблюдений
+Полученные спектры согласуются с данными наблюдений.
+Уровень рентгеновского излучение так же позволяет оценить отношение инжекций электронов и протонов в процесс ускорения.
+Как уже говорилось выше, Монте-Карло моделирование не способно корректно расчитывать инжекцию электронов, и нормировка нетеплового хвоста их распределения произвольна.
+Вычисление абсолютной величины потоков синхротронного излучение позволяет определить абсолютную величину функции распределения электронов.
+В данном случае правильная величина потоков получается при отношении инжекций $3.3\times10^{-3}$, что соответствует теоретическим оценкам из работы \cite{ParkElectronIonShock}.
 
@@
 \subsection{Гамма излучение}
-In this Section, we model the gamma radiation of W50, which is observed by H.E.S.S.
-\cite{HESS2024SS433} and LHAASO \cite{LHAASO2024SS433}.
-The leptonic model assumes that it is produced via inverse Compton scattering of electrons accelerated by recollimation shocks in the collimated outflow.
 
-In Fig.
-\ref{compton}, the red curve shows the gamma radiation produced within the outflow by accelerated VHE electrons as they are advected through the downstream of the first shock to the second one.
-This corresponds to pure advection of VHE electrons along the X-ray outflow with the turbulent magnetic field and the distribution function of electrons, which were used to simulate the synchrotron X-ray radiation discussed above.
-The red curve is significantly below the observational data, and it has a different spectral shape at low energies.
-This is because the radiating electrons in the X-ray jet are in the thin target regime.
-On the other hand, the electrons accelerated at the first shock diffuse in the direction transverse to the X-ray outflow.
-These electrons left the axial X-ray outflow and diffused through the surrounding cocoon, comprising a spatially broad component.
-If their diffusion coefficient in the cocoon is $\sim 10^{28} см^2/c$ at 100 TeV, they would radiate gamma-rays in the thick target regime in a wide cocoon of a size $\sim$ 30 pc around the axial jet.
-Electrons that have flown through the X-ray axial jet with a magnetic field amplified by cosmic ray-driven instabilities at the shock and escape into the cocoon downstream would also radiate in the thick regime.
-In Fig.
-\ref{compton}, the gamma-ray spectrum integrated over the wide cocoon is shown by the blue curve, which is in reasonably good agreement with the LHAASO data points taken from \cite{LHAASO2024SS433}.
-The gamma-ray spectrum integrated over a somewhat narrower region of the cocoon corresponds to the intermediate regime between the thin and the thick targets.
-It is shown by a brown curve in Fig.~\ref{compton} and is compared with the data reported by H.E.S.S.
-\cite{HESS2024SS433}.
+Туманность W50 является ярким источником гамма-излучения.
+Она наблюдалась телескопами HAWC \cite{Abeysekara2018}, H.E.S.S \cite{HESS2024SS433} и LHAASO \cite{LHAASO2024SS433}.
+Мы предположили лептонную модель происхождения данного излучения - в результате обратного комптоновского рассеяния электронов, ускоренных на реколлимационных ударных волнах.
 
-The LHAASO images above 100 TeV \cite{LHAASO2024SS433} show a more extended emission region compared to the lower energy bands.
-This morphology can be understood within our diffusive shock acceleration model, where the highest energy particles escape into the shock upstream, where the low magnetic field allows the escaped VHE leptons to form an extended gamma-ray emission region by the inverse Compton process.
+Спектры, расчитанные в различных моделях, показаны на рисунке \ref{compton}.
+Красная кривая соответствует излучению ускоренных электронов, конвективно переносимых вдоль джета от первой реколлимационной ударной волны, как описано в предыдущем разделе.
+С учетом нормировки, полученной из оценки рентгеновского излучения, оказывается, что излучения этих электронов не хватает для объяснения наблюдаемого гамма излучения.
+Так же видно, что на низких энергиях степенной индекс намного жестче, чем наблюдаемое значение, что указывает на то, что наблюдаемое излучение происходит в режиме толстой мишени даже на низких энергиях.
+Отсюда можно предположить, что часть ускоренных электронов покидает джет за счет поперечной диффузии и попадает в область слабого магнитного поля в коконе, где излучает преимущественно за счет обратного комптоновского рассеяния.
+Если коэффициент диффузии на энергиях около 100 ТэВ меньше $10^{28} см^2/c$, то такие электроны могут удерживаться в коконе шириной около 30 пк на временах порядка десятков тысяч лет и излучать в режиме толстой мишени, теряя всю свою энергию на рождение гамма квантов.
+Если доля таких электронов, перешедших из джета в кокон, не теряя энергию на синхротронное излучение составит $30\%$ от общего числа ускоренных электронов, то суммарное излучение, показанное розовой кривой на рисунке \ref{compton} позволит объяснить данные H.E.S.S и высокоэнергичную часть данных LHAASO.
 
-In our model, $\sim$ 30 $\%$ of the electrons accelerated at the shock front had escaped from the jet and diffused in the cocoon.
-This allows us to explain the observational data from LHAASO as shown in Figure \ref{compton}.
-H.E.S.S.
-data fall between the model for LHAASO and radiation from the jet itself, which can be explained by a smaller region size, observed by H.E.S.S.
+\begin{figure}
+	\includegraphics[width=0.8\textwidth]{W50compton3.png}
+	\caption{Моделированный спектр гамма излучения W50 и наблюдательные данные\cite{HESS2024SS433, LHAASO2024SS433}. }
+	\label{compton}
+\end{figure}
 
-There are recent studies of the acceleration of PeV protons in microquasars \citep{2024A&A...691A..93A,2024arXiv241108762P,2025arXiv250510620K}.
-A spherically symmetric semi-analytical model of particle acceleration to energies of several PeV by the termination wind shock was discussed in \cite{2024arXiv241108762P}.
-The acceleration of particles in the shear flows of jet-cocoon-structured microquasars is suggested in \citep{2025arXiv250620193Z}.
-Shear acceleration in the model requires the injection of particles pre-accelerated up to TeV regime energies.
-Hadronic interactions of accelerated protons are producing a substantial fraction of the observed VHE gamma-ray emission in the models \citep{2024arXiv241108762P,2025arXiv250510620K}.
+Для объяснения подъема на низкоэнергичной части спектра LHAASO предположим, что электроны, полностью прошедшие вдоль джета и потерявшие значительную часть энергии на синхротронное излучение, в конце концов тоже попадают в кокон и свободно диффундируют в нем, так же излучая обратным комптоновским рассеянием в режиме толстой мишени.
+Сумарный спектр трех компонент даст голубую кривую на рисунке \ref{compton}, хорошо совпадающую с данными LHAASO.
+Предположении о наличии большего числа электронов, видимых LHAASO чем HESS, косвенно подтверждается большими размерами областей со значимым сигналом, которые видит LHAASO.
 
-The recollimation shocks in the diffusive shock acceleration model considered here efficiently accelerate protons up to PeV energies, as is seen in Fig. \ref{pdf_shock_Q_esc}.
-In our model, $\gsim$ 10\% of the shocked jet power is transferred to accelerated protons while the VHE emission comes from the inverse Compton radiation of VHE electrons accelerated at shocks, producing polarized synchrotron X-ray emission of the extended X-ray jet.
-The leptonic gamma-ray emission is produced in the thick target regime within the cocoon surrounding the extended X-ray jet.
-The VHE gamma rays can be produced by hadronic interactions of the protons with ambient matter.
-The very low number density of the plasma in the extended jet and in the surrounding cocoon would require a very small diffusion coefficient for the accelerated VHE protons to contribute to the detected gamma-ray emission from the cocoon, given by the gamma-ray significance maps of H.E.S.S.
-and LHAASO.
-Therefore, most of the accelerated protons would escape the accelerator in the SS433/W50 extended X-ray jet and then contribute to the gamma-ray emission from the denser gas in the vicinity of W50 and supply the galactic cosmic-ray population.
 \section{Выводы} \label{SS433conclision}
